### 2024-09-16 ms/ms.tex
copernicus Rmd-template for manuscript

%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
%% ---------------------------------
%% This template should be used for copernicus.cls
%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
%% For further assistance please contact Copernicus Publications at: production@copernicus.org
%% https://publications.copernicus.org/for_authors/manuscript_preparation.html

%% copernicus_rticles_template (flag for rticles template detection - do not remove!)

%% Please use the following documentclass and journal abbreviations for discussion papers and final revised papers.

%% 2-column papers and discussion papers
\documentclass[gc, manuscript]{copernicus}

%% Journal abbreviations (please use the same for preprints and final revised papers)

% Advances in Geosciences (adgeo)
% Advances in Radio Science (ars)
% Advances in Science and Research (asr)
% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
% Aerosol Research (ar)
% Annales Geophysicae (angeo)
% Archives Animal Breeding (aab)
% Atmospheric Chemistry and Physics (acp)
% Atmospheric Measurement Techniques (amt)
% Biogeosciences (bg)
% Climate of the Past (cp)
% DEUQUA Special Publications (deuquasp)
% Earth Surface Dynamics (esurf)
% Earth System Dynamics (esd)
% Earth System Science Data (essd)
% E&G Quaternary Science Journal (egqsj)
% EGUsphere (egusphere) | This is only for EGUsphere preprints submitted without relation to an EGU journal.
% European Journal of Mineralogy (ejm)
% Fossil Record (fr)
% Geochronology (gchron)
% Geographica Helvetica (gh)
% Geoscience Communication (gc)
% Geoscientific Instrumentation, Methods and Data Systems (gi)
% Geoscientific Model Development (gmd)
% History of Geo- and Space Sciences (hgss)
% Hydrology and Earth System Sciences (hess)
% Journal of Bone and Joint Infection (jbji)
% Journal of Micropalaeontology (jm)
% Journal of Sensors and Sensor Systems (jsss)
% Magnetic Resonance (mr)
% Mechanical Sciences (ms)
% Natural Hazards and Earth System Sciences (nhess)
% Nonlinear Processes in Geophysics (npg)
% Ocean Science (os)
% Polarforschung - Journal of the German Society for Polar Research (polf)
% Primate Biology (pb)
% Proceedings of the International Association of Hydrological Sciences (piahs)
% Safety of Nuclear Waste Disposal (sand)
% Scientific Drilling (sd)
% SOIL (soil)
% Solid Earth (se)
% State of the Planet (sp)
% The Cryosphere (tc)
% Weather and Climate Dynamics (wcd)
% Web Ecology (we)
% Wind Energy Science (wes)

% Pandoc citation processing

% The "Technical instructions for LaTex" by Copernicus require _not_ to insert any additional packages.
% 
% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%
%% \usepackage commands included in the copernicus.cls:
%\usepackage[german, english]{babel}
%\usepackage{tabularx}
%\usepackage{cancel}
%\usepackage{multirow}
%\usepackage{supertabular}
%\usepackage{algorithmic}
%\usepackage{algorithm}
%\usepackage{amsthm}
%\usepackage{float}
%\usepackage{subfig}
%\usepackage{rotating}

\begin{document}

\title{Template for preparing your manuscript submission to Copernicus
journals using RMarkdown}

\Author[1, *][daniel.nuest@uni-muenster.de]{Daniel}{Nüst}
\Author[2]{Josiah}{Carberry}
\Author[1, *]{Markus}{Konkol}
\Author[3, †]{Nikolaus}{Copernicus}

\affil[1]{Institute for Geoinformatics, University of Münster, 48149
Münster, Germany}
\affil[2]{Psychoceramics, Wesleyan University, Middletown, CT, United
States}
\affil[3]{University of Ferrara, Ferrara, Italy}
\affil[†]{deceased, 24 May 1543}
\affil[*]{These authors contributed equally to this work.}

\runningtitle{R Markdown Template for Copernicus}

\runningauthor{Nüst et al.}

\correspondence{Daniel\ Nüst\ (daniel.nuest@uni-muenster.de)}

\received{}
\pubdiscuss{} %% only important for two-stage journals
\revised{}
\accepted{}
\published{}

%% These dates will be inserted by Copernicus Publications during the typesetting process.

\firstpage{1}

\maketitle

\begin{abstract}
The abstract goes here. It can also be on \emph{multiple lines}.
\end{abstract}

\copyrightstatement{The author's copyright for this publication is
transferred to institution/company.}

\introduction[Introduction]

Introduction text goes here. You can change the name of the section if
necessary using
\texttt{\textbackslash{}introduction{[}modified\ heading{]}}.

The following settings can or must be configured in the header of this
file and are bespoke for Copernicus manuscripts:

\begin{itemize}
\item
  The \texttt{journal} you are submitting to using the official
  abbreviation. You can use the function
  \texttt{rticles::copernicus\_journal\_abbreviations(name\ =\ \textquotesingle{}...\textquotesingle{})}
  to search the existing journals.
\item
  Specific sections of the manuscript:

  \begin{itemize}
  \item
    \texttt{running} with \texttt{title} and \texttt{author}
  \item
    \texttt{competinginterests}
  \item
    \texttt{copyrightstatement} (optional)
  \item
    \texttt{availability} (strongly recommended if any used), one of
    \texttt{code}, \texttt{data}, or \texttt{codedata}
  \item
    \texttt{authorcontribution}
  \item
    \texttt{disclaimer}
  \item
    \texttt{acknowledgements}
  \end{itemize}
\end{itemize}

See the defaults and examples from the skeleton and the official
Copernicus documentation for details.

\textbf{Please note:} Per
\href{https://publications.copernicus.org/for_authors/manuscript_preparation.html}{their
guidelines}, Copernicus does not support additional \LaTeX{} packages or
new \LaTeX{} commands than those defined in their \texttt{.cls} file.
This means that you cannot add any extra dependencies and a warning will
be thrown if so.\\
This extends to syntax highlighting of source code. Therefore this
template sets the parameter \texttt{highlight} in the YAML header to
\texttt{NULL} to deactivate Pandoc syntax highlighter. This prevent
addition of external packages for highlighting inserted by Pandoc.
However, it might be desirable to have syntax highlight available in
preprints or for others reasons. Please see
\texttt{?rmarkdown::pdf\_document} for available options to activate
highlighting.

\textbf{Important}: Always double-check with the official manuscript
preparation guidelines at
\url{https://publications.copernicus.org/for_authors/manuscript_preparation.html},
especially the sections ``Technical instructions for LaTeX'' and
``Manuscript composition''. Please contact Daniel Nüst,
\texttt{daniel.nuest@uni-muenster.de}, with any problems.

\section{Content section one}

\subsection{Subsection Heading Here}

Subsection text here.

\subsubsection{Subsubsection Heading Here}

Subsubsection text here.

\section{Content section with citations}

See the
\href{http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html}{R
Markdown docs for bibliographies and citations}.

Copernicus supports biblatex and a sample bibliography is in file
\texttt{sample.bib}. Read \citep{Feynman1963118}, and
\citep[see][]{Dirac1953888}.

\section{Content section with R code chunks}

You should always use \texttt{echo\ =\ FALSE} on R Markdown code blocks
as they add formatting and styling not desired by Copernicus. The hidden
workflow results in 42.

You can add verbatim code snippets without extra styles by using
\texttt{\textasciigrave{}\textasciigrave{}\textasciigrave{}} without
additional instructions.

\begin{verbatim}
sum <- 1 + 41
\end{verbatim}

\section{Content section with list}

If you want to insert a list, you must

\begin{itemize}
\item
  leave
\item
  empty lines
\item
  between each list item
\end{itemize}

because the \texttt{\textbackslash{}tightlist} format used by R Markdown
is not supported in the Copernicus template. Example:

\begin{verbatim}
- leave

- empty lines

- between each list item
\end{verbatim}

\section{Examples from the official template}

\subsection{FIGURES}

When figures and tables are placed at the end of the MS (article in
one-column style), please add \clearpage between bibliography and first
table and/or figure as well as between each table and/or figure.

\subsubsection{ONE-COLUMN FIGURES}

Include a 12cm width figure of Nikolaus Copernicus from
\href{https://en.wikipedia.org/wiki/File:Nikolaus_Kopernikus.jpg}{Wikipedia}
with caption using R Markdown.

\begin{figure}
\includegraphics[width=8.3cm]{Nikolaus_Kopernikus} \caption{one column figure}\label{fig:unnamed-chunk-2}
\end{figure}

\subsubsection{TWO-COLUMN FIGURES}

You can also include a larger figure.

\begin{figure}
\includegraphics[width=12cm]{Nikolaus_Kopernikus} \caption{two column figure}\label{fig:unnamed-chunk-3}
\end{figure}

\subsection{TABLES}

You can ad \LaTeX table in an R Markdown document to meet the template
requirements.

\subsubsection{ONE-COLUMN TABLE}

\begin{table}[t]
\caption{TEXT}
\begin{tabular}{l c r}
\tophline

a & b & c \\
\middlehline
1 & 2 & 3 \\

\bottomhline
\end{tabular}
\belowtable{Table Footnotes}
\end{table}

\subsubsection{TWO-COLUMN TABLE}

\begin{table*}[t]
\caption{TEXT}
\begin{tabular}{l c r}
\tophline

a & b & c \\
\middlehline
1 & 2 & 3 \\

\bottomhline
\end{tabular}
\belowtable{Table footnotes}
\end{table*}

\subsection{MATHEMATICAL EXPRESSIONS}

All papers typeset by Copernicus Publications follow the math
typesetting regulations given by the IUPAC Green Book (IUPAC:
Quantities, Units and Symbols in Physical Chemistry, 2nd Edn., Blackwell
Science, available at:
http://old.iupac.org/publications/books/gbook/green\_book\_2ed.pdf,
1993).

Physical quantities/variables are typeset in italic font (t for time, T
for Temperature)

Indices which are not defined are typeset in italic font (x, y, z, a, b,
c)

Items/objects which are defined are typeset in roman font (Car A, Car B)

Descriptions/specifications which are defined by itself are typeset in
roman font (abs, rel, ref, tot, net, ice)

Abbreviations from 2 letters are typeset in roman font (RH, LAI)

Vectors are identified in bold italic font using \vec{x}

Matrices are identified in bold roman font

Multiplication signs are typeset using the LaTeX commands
\texttt{\textbackslash{}times} (for vector products, grids, and
exponential notations) or \texttt{\textbackslash{}cdot}

The character * should not be applied as mutliplication sign

\subsection{EQUATIONS}

\subsubsection{Single-row equation}

Unnumbered equations (i.e.~using \texttt{\$\$} and getting inline
preview in RStudio) are not supported by Copernicus.

\begin{equation}
1 \times 1 \cdot 1 = 42
\end{equation}

\begin{equation}
A = \pi r^2
\end{equation}

\begin{equation}
x=\frac{2b\pm\sqrt{b^{2}-4ac}}{2c}.  
\end{equation}

\subsubsection{Multiline equation}

\begin{align}
& 3 + 5 = 8\\
& 3 + 5 = 8\\
& 3 + 5 = 8
\end{align}

\subsection{MATRICES}

\[
\begin{matrix}
x & y & z\\
x & y & z\\
x & y & z\\
\end{matrix}
\]

\subsection{ALGORITHM/PROGRAMMING CODE}

If you want to use algorithms, you need to make sure yourself that the
\LaTeX packages \texttt{algorithms} and \texttt{algorithmicx} are
installed so that \texttt{algorithm.sty} respectively
\texttt{algorithmic.sty} can be loaded by the Copernicus template. Both
need to be available through your preferred \LaTeX{} distribution. With
TinyTeX (or TeX Live), you can do so by running
\texttt{tinytex::tlmgr\_install(c("algorithms",\ "algorithmicx"))}

\begin{verbatim}
## tlmgr update --all --self
\end{verbatim}

\begin{verbatim}
## tlmgr install algorithms algorithmicx
\end{verbatim}

Copernicus staff will no accept any additional packages from your LaTeX
source code, so please stick to these two acceptable packages. They are
needed to use the example below

\begin{algorithm}
\caption{Algorithm Caption}
\label{a1}
\begin{algorithmic}
\STATE $i\gets 10$
\IF {$i\geq 5$} 
        \STATE $i\gets i-1$
\ELSE
        \IF {$i\leq 3$}
                \STATE $i\gets i+2$
        \ENDIF
\ENDIF
\end{algorithmic}
\end{algorithm}

\subsection{CHEMICAL FORMULAS AND REACTIONS}

For formulas embedded in the text, please use
\texttt{\textbackslash{}chem\{\}}, e.g.~\chem{A \rightarrow B}.

The reaction environment creates labels including the letter R,
i.e.~(R1), (R2), etc.

\begin{itemize}
\item
  \texttt{\textbackslash{}rightarrow} should be used for normal
  (one-way) chemical reactions
\item
  \texttt{\textbackslash{}rightleftharpoons} should be used for
  equilibria
\item
  \texttt{\textbackslash{}leftrightarrow} should be used for resonance
  structures
\end{itemize}

\begin{reaction}
A \rightarrow B \\
\end{reaction}
\begin{reaction}
Coper \rightleftharpoons nicus \\
\end{reaction}
\begin{reaction}
Publi \leftrightarrow cations
\end{reaction}

\subsection{PHYSICAL UNITS}

Please use \texttt{\textbackslash{}unit\{\}} (allows to save the
math/\texttt{\$} environment) and apply the exponential notation, for
example \(3.14\,\unit{km\,h^{-1}}\) (using LaTeX mode:
\texttt{\textbackslash{}(\ 3.14\textbackslash{},\textbackslash{}unit\{...\}\ \textbackslash{})})
or \unit{0.872\,m\,s^{-1}} (using only
\texttt{\textbackslash{}unit\{0.872\textbackslash{},m\textbackslash{},s\^{}\{-1\}\}}).

\conclusions[Conclusions]

The conclusion goes here. You can modify the section name with
\texttt{\textbackslash{}conclusions{[}modified\ heading\ if\ necessary{]}}.

\codedataavailability{use this to add a statement when having data sets
and software code
available} %% use this section when having data sets and software code available

\sampleavailability{use this section when having geoscientific samples
available} %% use this section when having geoscientific samples available

\videosupplement{use this section when having video supplements
available} %% use this section when having geoscientific samples available

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\section{Figures and tables in appendices}

Regarding figures and tables in appendices, the following two options
are possible depending on your general handling of figures and tables in
the manuscript environment:

\subsection{Option 1}

If you sorted all figures and tables into the sections of the text,
please also sort the appendix figures and appendix tables into the
respective appendix sections. They will be correctly named
automatically.

\subsection{Option 2}

If you put all figures after the reference list, please insert appendix
tables and figures after the normal tables and figures.

To rename them correctly to A1, A2, etc., please add the following
commands in front of them: \texttt{\textbackslash{}appendixfigures}
needs to be added in front of appendix figures
\texttt{\textbackslash{}appendixtables} needs to be added in front of
appendix tables

Please add \texttt{\textbackslash{}clearpage} between each table and/or
figure. Further guidelines on figures and tables can be found below.
\noappendix

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontribution{Daniel wrote the package. Josiah thought about
poterry. Markus filled in for a second author.} %% optional section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\competinginterests{The authors declare no competing
interests.} %% this section is mandatory even if you declare that no competing interests are present

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\disclaimer{We like Copernicus.} %% optional section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgements}
Thanks to the rticles contributors!
\end{acknowledgements}

%% REFERENCES
%% DN: pre-configured to BibTeX for rticles

%% The reference list is compiled as follows:
%%
%% \begin{thebibliography}{}
%%
%% \bibitem[AUTHOR(YEAR)]{LABEL1}
%% REFERENCE 1
%%
%% \bibitem[AUTHOR(YEAR)]{LABEL2}
%% REFERENCE 2
%%
%% \end{thebibliography}

%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
%% authors experienced with BibTeX only have to include the following two lines:
%%
\bibliographystyle{copernicus}
\bibliography{sample.bib}
%%
%% URLs and DOIs can be entered in your BibTeX file as:
%%
%% URL = {http://www.xyz.org/~jones/idx_g.htm}
%% DOI = {10.5194/xyz}

%% LITERATURE CITATIONS
%%
%% command                        & example result
%% \citet{jones90}|               & Jones et al. (1990)
%% \citep{jones90}|               & (Jones et al., 1990)
%% \citep{jones90,jones93}|       & (Jones et al., 1990, 1993)
%% \citep[p.~32]{jones90}|        & (Jones et al., 1990, p.~32)
%% \citep[e.g.,][]{jones90}|      & (e.g., Jones et al., 1990)
%% \citep[e.g.,][p.~32]{jones90}| & (e.g., Jones et al., 1990, p.~32)
%% \citeauthor{jones90}|          & Jones et al.
%% \citeyear{jones90}|            & 1990

\end{document}

### 2024-09-17 ms/ms.Rmd
barebones adapted to submitting guidelines and workflow

---
title: Wall-to-wall mapping of peat depth from Lidar terrain and airborne radiometrics in Norwegian landscapes
journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'soil')`"
author:
  - given_name: Julien
    surname: Vollering
    affiliation: 1
    email: julien.vollering@hvl.no
    corresponding: true
  - given_name: Naomi
    surname: Gatis
    affiliation: 2
  - given_name: Mette
    surname: Kusk Gillespie
    affiliation: 1
  - given_name: Karl-Kristian
    surname: Muggerud
    affiliation: 1
  - given_name: Sigurd Daniel
    surname: Nerhus
    affiliation: 1
  - given_name: Knut
    surname: Rydgren
    affiliation: 1
  - given_name: Mikko
    surname: Sparf
    affiliation: 1
affiliation:
  - code: 1
    address: Department of Civil Engineering and Environmental Sciences, Western Norway University of Applied Sciences, Norway
  - code: 2
    address: Department of Geography, University of Exeter, United Kingdom
abstract: |
  The abstract goes here.
  It can also be on _multiple lines_.
bibliography: ms.bib
running:
  title: Peat depth from terrain and radiometrics
  author: Vollering et al.
# This section is mandatory even if you declare that no competing interests are present.
competinginterests: |
  *Competing interests.* The authors declare that they have no conflict of interest.
# See https://publications.copernicus.org/for_authors/licence_and_copyright.html, normally used for transferring the copyright, if needed. 
# Note: additional copyright statements for affiliated software or data need to be placed in the data availability section. 
### The following commands are for the statements about the availability of data sets and/or software code corresponding to the manuscript.
### It is strongly recommended to make use of these sections in case data sets and/or software code have been part of your research the article is based on.
### Note: unless stated otherwise, software and data affiliated with the manuscript are assumed to be published under the same licence as the article (currently Creative Commons 4.0)
availability:
  #code: |
  #  use this to add a statement when having only software code available
  #data: |
  #  use this to add a statement when having only data sets available
  codedata: |
    *Code and data availability.* Use this to add a statement when having data sets and software code available
authorcontribution: |
  *Author contributions.*
  JV: Conceptualization, Investigation, Data curation, Formal analysis, Writing – original draft.
  NG: Conceptualization, Methodology, Writing - review & editing.
  MKG: Investigation, Writing - review & editing.
  KKM: Investigation, Data curation, Writing - review & editing.
  SDN: Investigation, Writing - review & editing.
  KR: Conceptualization, Investigation, Writing - review & editing.
  MS: Investigation, Data curation, Writing - review & editing.
disclaimer: |
  *Disclaimer.* The authors declare that the results, discussions, and interpretations presented in this study are solely their own. The views expressed herein do not necessarily reflect those of their respective institutions or funding agencies.
acknowledgements: |
  *Acknowledgements.* We thank the Norwegian Public Roads Administration for sharing data from ground-penetrating radar surveys. We also thank Vikas Baranwal from the Geological Survey of Norway for helping us access the radiometric data from Skrim. 
appendix: |
  \section{For submission}
  "Appendices: all material required to understand the essential aspects of the paper
  such as experimental methods, data, and interpretation 
  should preferably be included in the main text. 
  Additional figures, tables, as well as technical and theoretical developments 
  which are not critical to support the conclusion of the paper,
  but which provide extra detail and/or support useful for experts in the field
  and whose inclusion in the main text would disrupt the flow of descriptions or demonstrations 
  may be presented as appendices.
  These should be labelled with capital letters: Appendix A, Appendix B etc.
  Equations, figures and tables should be numbered as (A1), Fig. B5 or Table C6, respectively.
  Please keep in mind that appendices are part of the manuscript
  whereas supplements (see below) are published along with the manuscript."

  \section{Figures and tables in appendices}
  Please also sort the appendix figures and appendix tables into the respective appendix sections.
  They will be correctly named automatically.

  \section{Copernicus from Rmarkdown}
  **Please note:** Per [their guidelines](https://publications.copernicus.org/for_authors/manuscript_preparation.html),
  Copernicus does not support additional \LaTeX{} packages 
  or new \LaTeX{} commands than those defined in their `.cls` file. 
  This means that you cannot add any extra dependencies 
  and a warning will be thrown if so.
  **Important**: Always double-check with the official manuscript preparation guidelines 
  at [https://publications.copernicus.org/for_authors/manuscript_preparation.html](https://publications.copernicus.org/for_authors/manuscript_preparation.html),
  especially the sections "Technical instructions for LaTeX" and "Manuscript composition". 
  Please contact Daniel Nüst, `daniel.nuest@uni-muenster.de`, with any problems.
output:
  bookdown::pdf_book:
    base_format: rticles::copernicus_article # for using bookdown features like \@ref()
---

# Introduction

Introduction text goes here.
You can change the name of the section if necessary using `\introduction[modified heading]`.

Lorem ipsum dolor sit amet, consectetur adipiscing elit.
Sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
Excepteur sint occaecat cupidatat non proident,
sunt in culpa qui officia deserunt mollit anim id est laborum.

Nulla facilisi.
Maecenas vel nunc nec purus tincidunt congue.
Proin auctor, lectus eu pharetra malesuada,
nisi nunc bibendum nunc,
eget tincidunt nunc nisi id nunc.
Sed euismod, nunc sit amet aliquam tincidunt,
nunc nunc tincidunt nunc,
nec tincidunt nunc nunc nec nunc.
Donec auctor, nunc sit amet aliquam tincidunt,
nunc nunc tincidunt nunc,
nec tincidunt nunc nunc nec nunc.

# Materials and methods

## Subsection Heading Here

Read Gatis et al. [-@gatisMappingUplandPeat2019] 
and related work [@minasnyDigitalMappingPeatlands2019].

### Subsubsection Heading Here

Subsubsection text here.

# Results

Include a 12cm width figure of Nikolaus Copernicus from [Wikipedia](https://en.wikipedia.org/wiki/File:Nikolaus_Kopernikus.jpg) with caption using R Markdown (Figure \@ref(fig:portrait)).

```{r portrait, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
knitr::include_graphics("Nikolaus_Kopernikus.jpg")
```

## Tables

You can add \LaTeX table in an R Markdown document to meet the template requirements (Table \ref{tab:latextable}).

\begin{table}[t]
\caption{TEXT}
\begin{tabular}{l c r}
\tophline

a & b & c \\
\middlehline
1 & 2 & 3 \\

\bottomhline
\end{tabular}
\belowtable{Table Footnotes}
\label{tab:latextable}
\end{table}

Or you can use markdown to create the table with booktabs = FALSE (<https://github.com/rstudio/rticles/issues/558#issuecomment-1907981541>).

See Table \@ref(tab:test).

```{r test, echo=FALSE}
df = mtcars[1:3, 1:3]
knitr::kable(df, booktabs = FALSE, caption = "My caption")
```

# Discussion

Lorem ipsum dolor sit amet, consectetur adipiscing elit.
Sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
Excepteur sint occaecat cupidatat non proident,
sunt in culpa qui officia deserunt mollit anim id est laborum.

# Conclusions

Nulla facilisi.
Maecenas vel nunc nec purus tincidunt congue.
Proin auctor, lectus eu pharetra malesuada,
nisi nunc bibendum nunc,
eget tincidunt nunc nisi id nunc.
Sed euismod, nunc sit amet aliquam tincidunt,
nunc nunc tincidunt nunc,
nec tincidunt nunc nunc nec nunc.
Donec auctor, nunc sit amet aliquam tincidunt,
nunc nunc tincidunt nunc,
nec tincidunt nunc nunc nec nunc.

### 2024-09-17 ms/ms.tex
barebones adapted to submitting guidelines and workflow

@@
 %% 2-column papers and discussion papers
-\documentclass[gc, manuscript]{copernicus}
+\documentclass[soil, manuscript]{copernicus}
 
@@
 % The "Technical instructions for LaTex" by Copernicus require _not_ to insert any additional packages.
+% % % From pandoc table feature
+% \usepackage{longtable,booktabs,array}
+% % \usepackage{calc} % for calculating minipage widths
+% % Correct order of tables after \paragraph or \subparagraph
+% \usepackage{etoolbox}
+% \makeatletter
+% \patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+% \makeatother
+% % Allow footnotes in longtable head/foot
+% \IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+% \makesavenoteenv{longtable}
 % 
@@
 
-\title{Template for preparing your manuscript submission to Copernicus
-journals using RMarkdown}
+\title{Wall-to-wall mapping of peat depth from Lidar terrain and airborne radiometrics in Norwegian landscapes}
 
-\Author[1, *][daniel.nuest@uni-muenster.de]{Daniel}{Nüst}
-\Author[2]{Josiah}{Carberry}
-\Author[1, *]{Markus}{Konkol}
-\Author[3, †]{Nikolaus}{Copernicus}
+\Author[1][julien.vollering@hvl.no]{Julien}{Vollering}
+\Author[2]{Naomi}{Gatis}
+\Author[1]{Mette}{Kusk Gillespie}
+\Author[1]{Karl-Kristian}{Muggerud}
+\Author[1]{Sigurd Daniel}{Nerhus}
+\Author[1]{Knut}{Rydgren}
+\Author[1]{Mikko}{Sparf}
 
-\affil[1]{Institute for Geoinformatics, University of Münster, 48149
-Münster, Germany}
-\affil[2]{Psychoceramics, Wesleyan University, Middletown, CT, United
-States}
-\affil[3]{University of Ferrara, Ferrara, Italy}
-\affil[†]{deceased, 24 May 1543}
-\affil[*]{These authors contributed equally to this work.}
+\affil[1]{Department of Civil Engineering and Environmental Sciences, Western Norway University of Applied Sciences, Norway}
+\affil[2]{Department of Geography, University of Exeter, United Kingdom}
 
-\runningtitle{R Markdown Template for Copernicus}
+\runningtitle{Peat depth from terrain and radiometrics}
 
-\runningauthor{Nüst et al.}
+\runningauthor{Vollering et al.}
 
-\correspondence{Daniel\ Nüst\ (daniel.nuest@uni-muenster.de)}
+\correspondence{Julien\ Vollering\ (julien.vollering@hvl.no)}
 
@@
 \begin{abstract}
-The abstract goes here. It can also be on \emph{multiple lines}.
+The abstract goes here.
+It can also be on \emph{multiple lines}.
 \end{abstract}
 
-\copyrightstatement{The author's copyright for this publication is
-transferred to institution/company.}
+\section{Introduction}
 
-\introduction[Introduction]
+Introduction text goes here.
+You can change the name of the section if necessary using \texttt{\textbackslash{}introduction{[}modified\ heading{]}}.
 
-Introduction text goes here. You can change the name of the section if
-necessary using
-\texttt{\textbackslash{}introduction{[}modified\ heading{]}}.
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident,
+sunt in culpa qui officia deserunt mollit anim id est laborum.
 
-The following settings can or must be configured in the header of this
-file and are bespoke for Copernicus manuscripts:
+Nulla facilisi.
+Maecenas vel nunc nec purus tincidunt congue.
+Proin auctor, lectus eu pharetra malesuada,
+nisi nunc bibendum nunc,
+eget tincidunt nunc nisi id nunc.
+Sed euismod, nunc sit amet aliquam tincidunt,
+nunc nunc tincidunt nunc,
+nec tincidunt nunc nunc nec nunc.
+Donec auctor, nunc sit amet aliquam tincidunt,
+nunc nunc tincidunt nunc,
+nec tincidunt nunc nunc nec nunc.
 
-\begin{itemize}
-\item
-  The \texttt{journal} you are submitting to using the official
-  abbreviation. You can use the function
-  \texttt{rticles::copernicus\_journal\_abbreviations(name\ =\ \textquotesingle{}...\textquotesingle{})}
-  to search the existing journals.
-\item
-  Specific sections of the manuscript:
-
-  \begin{itemize}
-  \item
-    \texttt{running} with \texttt{title} and \texttt{author}
-  \item
-    \texttt{competinginterests}
-  \item
-    \texttt{copyrightstatement} (optional)
-  \item
-    \texttt{availability} (strongly recommended if any used), one of
-    \texttt{code}, \texttt{data}, or \texttt{codedata}
-  \item
-    \texttt{authorcontribution}
-  \item
-    \texttt{disclaimer}
-  \item
-    \texttt{acknowledgements}
-  \end{itemize}
-\end{itemize}
-
-See the defaults and examples from the skeleton and the official
-Copernicus documentation for details.
-
-\textbf{Please note:} Per
-\href{https://publications.copernicus.org/for_authors/manuscript_preparation.html}{their
-guidelines}, Copernicus does not support additional \LaTeX{} packages or
-new \LaTeX{} commands than those defined in their \texttt{.cls} file.
-This means that you cannot add any extra dependencies and a warning will
-be thrown if so.\\
-This extends to syntax highlighting of source code. Therefore this
-template sets the parameter \texttt{highlight} in the YAML header to
-\texttt{NULL} to deactivate Pandoc syntax highlighter. This prevent
-addition of external packages for highlighting inserted by Pandoc.
-However, it might be desirable to have syntax highlight available in
-preprints or for others reasons. Please see
-\texttt{?rmarkdown::pdf\_document} for available options to activate
-highlighting.
-
-\textbf{Important}: Always double-check with the official manuscript
-preparation guidelines at
-\url{https://publications.copernicus.org/for_authors/manuscript_preparation.html},
-especially the sections ``Technical instructions for LaTeX'' and
-``Manuscript composition''. Please contact Daniel Nüst,
-\texttt{daniel.nuest@uni-muenster.de}, with any problems.
-
-\section{Content section one}
+\section{Materials and methods}
 
@@
 
-Subsection text here.
+Read Gatis et al. \citeyearpar{gatisMappingUplandPeat2019}
+and related work \citep{minasnyDigitalMappingPeatlands2019}.
 
@@
 
-\section{Content section with citations}
+\section{Results}
 
-See the
-\href{http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html}{R
-Markdown docs for bibliographies and citations}.
-
-Copernicus supports biblatex and a sample bibliography is in file
-\texttt{sample.bib}. Read \citep{Feynman1963118}, and
-\citep[see][]{Dirac1953888}.
-
-\section{Content section with R code chunks}
-
-You should always use \texttt{echo\ =\ FALSE} on R Markdown code blocks
-as they add formatting and styling not desired by Copernicus. The hidden
-workflow results in 42.
-
-You can add verbatim code snippets without extra styles by using
-\texttt{\textasciigrave{}\textasciigrave{}\textasciigrave{}} without
-additional instructions.
-
-\begin{verbatim}
-sum <- 1 + 41
-\end{verbatim}
-
-\section{Content section with list}
-
-If you want to insert a list, you must
-
-\begin{itemize}
-\item
-  leave
-\item
-  empty lines
-\item
-  between each list item
-\end{itemize}
-
-because the \texttt{\textbackslash{}tightlist} format used by R Markdown
-is not supported in the Copernicus template. Example:
-
-\begin{verbatim}
-- leave
-
-- empty lines
-
-- between each list item
-\end{verbatim}
-
-\section{Examples from the official template}
-
-\subsection{FIGURES}
-
-When figures and tables are placed at the end of the MS (article in
-one-column style), please add \clearpage between bibliography and first
-table and/or figure as well as between each table and/or figure.
-
-\subsubsection{ONE-COLUMN FIGURES}
-
-Include a 12cm width figure of Nikolaus Copernicus from
-\href{https://en.wikipedia.org/wiki/File:Nikolaus_Kopernikus.jpg}{Wikipedia}
-with caption using R Markdown.
+Include a 12cm width figure of Nikolaus Copernicus from \href{https://en.wikipedia.org/wiki/File:Nikolaus_Kopernikus.jpg}{Wikipedia} with caption using R Markdown (Figure \ref{fig:portrait}).
 
 \begin{figure}
-\includegraphics[width=8.3cm]{Nikolaus_Kopernikus} \caption{one column figure}\label{fig:unnamed-chunk-2}
+\includegraphics[width=8.3cm]{Nikolaus_Kopernikus} \caption{one column figure}\label{fig:portrait}
 \end{figure}
 
-\subsubsection{TWO-COLUMN FIGURES}
+\subsection{Tables}
 
-You can also include a larger figure.
-
-\begin{figure}
-\includegraphics[width=12cm]{Nikolaus_Kopernikus} \caption{two column figure}\label{fig:unnamed-chunk-3}
-\end{figure}
-
-\subsection{TABLES}
-
-You can ad \LaTeX table in an R Markdown document to meet the template
-requirements.
-
-\subsubsection{ONE-COLUMN TABLE}
+You can add \LaTeX table in an R Markdown document to meet the template requirements (Table \ref{tab:latextable}).
 
@@
 \belowtable{Table Footnotes}
+\label{tab:latextable}
 \end{table}
 
-\subsubsection{TWO-COLUMN TABLE}
+Or you can use markdown to create the table with booktabs = FALSE (\url{https://github.com/rstudio/rticles/issues/558\#issuecomment-1907981541}).
 
-\begin{table*}[t]
-\caption{TEXT}
-\begin{tabular}{l c r}
-\tophline
+See Table \ref{tab:test}.
 
-a & b & c \\
-\middlehline
-1 & 2 & 3 \\
+\begin{table}
 
-\bottomhline
+\caption{\label{tab:test}My caption}
+\centering
+\begin{tabular}[t]{l|r|r|r}
+\hline
+  & mpg & cyl & disp\\
+\hline
+Mazda RX4 & 21.0 & 6 & 160\\
+\hline
+Mazda RX4 Wag & 21.0 & 6 & 160\\
+\hline
+Datsun 710 & 22.8 & 4 & 108\\
+\hline
 \end{tabular}
-\belowtable{Table footnotes}
-\end{table*}
+\end{table}
 
-\subsection{MATHEMATICAL EXPRESSIONS}
+\section{Discussion}
 
-All papers typeset by Copernicus Publications follow the math
-typesetting regulations given by the IUPAC Green Book (IUPAC:
-Quantities, Units and Symbols in Physical Chemistry, 2nd Edn., Blackwell
-Science, available at:
-http://old.iupac.org/publications/books/gbook/green\_book\_2ed.pdf,
-1993).
+Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
+Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
+Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
+Excepteur sint occaecat cupidatat non proident,
+sunt in culpa qui officia deserunt mollit anim id est laborum.
 
-Physical quantities/variables are typeset in italic font (t for time, T
-for Temperature)
+\section{Conclusions}
 
-Indices which are not defined are typeset in italic font (x, y, z, a, b,
-c)
+Nulla facilisi.
+Maecenas vel nunc nec purus tincidunt congue.
+Proin auctor, lectus eu pharetra malesuada,
+nisi nunc bibendum nunc,
+eget tincidunt nunc nisi id nunc.
+Sed euismod, nunc sit amet aliquam tincidunt,
+nunc nunc tincidunt nunc,
+nec tincidunt nunc nunc nec nunc.
+Donec auctor, nunc sit amet aliquam tincidunt,
+nunc nunc tincidunt nunc,
+nec tincidunt nunc nunc nec nunc.
 
-Items/objects which are defined are typeset in roman font (Car A, Car B)
-
-Descriptions/specifications which are defined by itself are typeset in
-roman font (abs, rel, ref, tot, net, ice)
-
-Abbreviations from 2 letters are typeset in roman font (RH, LAI)
-
-Vectors are identified in bold italic font using \vec{x}
-
-Matrices are identified in bold roman font
-
-Multiplication signs are typeset using the LaTeX commands
-\texttt{\textbackslash{}times} (for vector products, grids, and
-exponential notations) or \texttt{\textbackslash{}cdot}
-
-The character * should not be applied as mutliplication sign
-
-\subsection{EQUATIONS}
-
-\subsubsection{Single-row equation}
-
-Unnumbered equations (i.e.~using \texttt{\$\$} and getting inline
-preview in RStudio) are not supported by Copernicus.
-
-\begin{equation}
-1 \times 1 \cdot 1 = 42
-\end{equation}
-
-\begin{equation}
-A = \pi r^2
-\end{equation}
-
-\begin{equation}
-x=\frac{2b\pm\sqrt{b^{2}-4ac}}{2c}.  
-\end{equation}
-
-\subsubsection{Multiline equation}
-
-\begin{align}
-& 3 + 5 = 8\\
-& 3 + 5 = 8\\
-& 3 + 5 = 8
-\end{align}
-
-\subsection{MATRICES}
-
-\[
-\begin{matrix}
-x & y & z\\
-x & y & z\\
-x & y & z\\
-\end{matrix}
-\]
-
-\subsection{ALGORITHM/PROGRAMMING CODE}
-
-If you want to use algorithms, you need to make sure yourself that the
-\LaTeX packages \texttt{algorithms} and \texttt{algorithmicx} are
-installed so that \texttt{algorithm.sty} respectively
-\texttt{algorithmic.sty} can be loaded by the Copernicus template. Both
-need to be available through your preferred \LaTeX{} distribution. With
-TinyTeX (or TeX Live), you can do so by running
-\texttt{tinytex::tlmgr\_install(c("algorithms",\ "algorithmicx"))}
-
-\begin{verbatim}
-## tlmgr update --all --self
-\end{verbatim}
-
-\begin{verbatim}
-## tlmgr install algorithms algorithmicx
-\end{verbatim}
-
-Copernicus staff will no accept any additional packages from your LaTeX
-source code, so please stick to these two acceptable packages. They are
-needed to use the example below
-
-\begin{algorithm}
-\caption{Algorithm Caption}
-\label{a1}
-\begin{algorithmic}
-\STATE $i\gets 10$
-\IF {$i\geq 5$} 
-        \STATE $i\gets i-1$
-\ELSE
-        \IF {$i\leq 3$}
-                \STATE $i\gets i+2$
-        \ENDIF
-\ENDIF
-\end{algorithmic}
-\end{algorithm}
-
-\subsection{CHEMICAL FORMULAS AND REACTIONS}
-
-For formulas embedded in the text, please use
-\texttt{\textbackslash{}chem\{\}}, e.g.~\chem{A \rightarrow B}.
-
-The reaction environment creates labels including the letter R,
-i.e.~(R1), (R2), etc.
-
-\begin{itemize}
-\item
-  \texttt{\textbackslash{}rightarrow} should be used for normal
-  (one-way) chemical reactions
-\item
-  \texttt{\textbackslash{}rightleftharpoons} should be used for
-  equilibria
-\item
-  \texttt{\textbackslash{}leftrightarrow} should be used for resonance
-  structures
-\end{itemize}
-
-\begin{reaction}
-A \rightarrow B \\
-\end{reaction}
-\begin{reaction}
-Coper \rightleftharpoons nicus \\
-\end{reaction}
-\begin{reaction}
-Publi \leftrightarrow cations
-\end{reaction}
-
-\subsection{PHYSICAL UNITS}
-
-Please use \texttt{\textbackslash{}unit\{\}} (allows to save the
-math/\texttt{\$} environment) and apply the exponential notation, for
-example \(3.14\,\unit{km\,h^{-1}}\) (using LaTeX mode:
-\texttt{\textbackslash{}(\ 3.14\textbackslash{},\textbackslash{}unit\{...\}\ \textbackslash{})})
-or \unit{0.872\,m\,s^{-1}} (using only
-\texttt{\textbackslash{}unit\{0.872\textbackslash{},m\textbackslash{},s\^{}\{-1\}\}}).
-
-\conclusions[Conclusions]
-
-The conclusion goes here. You can modify the section name with
-\texttt{\textbackslash{}conclusions{[}modified\ heading\ if\ necessary{]}}.
-
-\codedataavailability{use this to add a statement when having data sets
-and software code
-available} %% use this section when having data sets and software code available
-
-\sampleavailability{use this section when having geoscientific samples
-available} %% use this section when having geoscientific samples available
-
-\videosupplement{use this section when having video supplements
-available} %% use this section when having geoscientific samples available
+\codedataavailability{\emph{Code and data availability.} Use this to add a statement when having data sets and software code available} %% use this section when having data sets and software code available
 
@@
 \appendix
+\section{For submission}
+
+``Appendices: all material required to understand the essential aspects of the paper
+such as experimental methods, data, and interpretation
+should preferably be included in the main text.
+Additional figures, tables, as well as technical and theoretical developments
+which are not critical to support the conclusion of the paper,
+but which provide extra detail and/or support useful for experts in the field
+and whose inclusion in the main text would disrupt the flow of descriptions or demonstrations
+may be presented as appendices.
+These should be labelled with capital letters: Appendix A, Appendix B etc.
+Equations, figures and tables should be numbered as (A1), Fig. B5 or Table C6, respectively.
+Please keep in mind that appendices are part of the manuscript
+whereas supplements (see below) are published along with the manuscript.''
+
 \section{Figures and tables in appendices}
 
-Regarding figures and tables in appendices, the following two options
-are possible depending on your general handling of figures and tables in
-the manuscript environment:
+Please also sort the appendix figures and appendix tables into the respective appendix sections.
+They will be correctly named automatically.
 
-\subsection{Option 1}
+\section{Copernicus from Rmarkdown}
 
-If you sorted all figures and tables into the sections of the text,
-please also sort the appendix figures and appendix tables into the
-respective appendix sections. They will be correctly named
-automatically.
-
-\subsection{Option 2}
-
-If you put all figures after the reference list, please insert appendix
-tables and figures after the normal tables and figures.
-
-To rename them correctly to A1, A2, etc., please add the following
-commands in front of them: \texttt{\textbackslash{}appendixfigures}
-needs to be added in front of appendix figures
-\texttt{\textbackslash{}appendixtables} needs to be added in front of
-appendix tables
-
-Please add \texttt{\textbackslash{}clearpage} between each table and/or
-figure. Further guidelines on figures and tables can be found below.
+\textbf{Please note:} Per \href{https://publications.copernicus.org/for_authors/manuscript_preparation.html}{their guidelines},
+Copernicus does not support additional \LaTeX{} packages
+or new \LaTeX{} commands than those defined in their \texttt{.cls} file.
+This means that you cannot add any extra dependencies
+and a warning will be thrown if so.
+\textbf{Important}: Always double-check with the official manuscript preparation guidelines
+at \url{https://publications.copernicus.org/for_authors/manuscript_preparation.html},
+especially the sections ``Technical instructions for LaTeX'' and ``Manuscript composition''.
+Please contact Daniel Nüst, \texttt{daniel.nuest@uni-muenster.de}, with any problems.
 \noappendix
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\authorcontribution{Daniel wrote the package. Josiah thought about
-poterry. Markus filled in for a second author.} %% optional section
+\authorcontribution{\emph{Author contributions.}
+JV: Conceptualization, Investigation, Data curation, Formal analysis, Writing -- original draft.
+NG: Conceptualization, Methodology, Writing - review \& editing.
+MKG: Investigation, Writing - review \& editing.
+KKM: Investigation, Data curation, Writing - review \& editing.
+SDN: Investigation, Writing - review \& editing.
+KR: Conceptualization, Investigation, Writing - review \& editing.
+MS: Investigation, Data curation, Writing - review \& editing.} %% optional section
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\competinginterests{The authors declare no competing
-interests.} %% this section is mandatory even if you declare that no competing interests are present
+\competinginterests{\emph{Competing interests.} The authors declare that they have no conflict of interest.} %% this section is mandatory even if you declare that no competing interests are present
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\disclaimer{We like Copernicus.} %% optional section
+\disclaimer{\emph{Disclaimer.} The authors declare that the results, discussions, and interpretations presented in this study are solely their own.
+The views expressed herein do not necessarily reflect those of their respective institutions or funding agencies.} %% optional section
 
@@
 \begin{acknowledgements}
-Thanks to the rticles contributors!
+\emph{Acknowledgements.} We thank the Norwegian Public Roads Administration for sharing data from ground-penetrating radar surveys.
+We also thank Vikas Baranwal from the Geological Survey of Norway for helping us access the radiometric data from Skrim.
 \end{acknowledgements}
@@
 \bibliographystyle{copernicus}
-\bibliography{sample.bib}
+\bibliography{ms.bib}
 %%

### 2024-09-17 R/Skrim/createMasks.R
Skrim: ar5 composition

@@ -10,6 +10,8 @@ sa <- st_read("data/Skrim/Skrim-site.gpkg", "fieldsite_outline_utm") |>
   st_geometry() |> 
   st_transform(crssite)
 plot(sa)
+st_area(sa) |> 
+  units::set_units("km^2")
 
 # AR5 myr ####
 

### 2024-09-18 R/Skrim/createMasks.R
Skrim: ar5 composition

@@ -18,6 +18,13 @@ st_area(sa) |>
 ar5 <- st_read("data/Skrim/Basisdata_3303_Kongsberg_25832_FKB-AR5_FGDB.gdb", 
                layer="fkb_ar5_omrade")
 ar5 <- st_transform(ar5, crssite)
+st_intersection(ar5, sa) |> 
+  mutate(area = st_area(SHAPE)) |> 
+  st_drop_geometry() |> 
+  group_by(arealtype) |> 
+  summarize(area = units::set_units(sum(area), "km^2")) |> 
+  mutate(percent = area / sum((area)) * 100)
+
 ar5.myr <- dplyr::filter(ar5, arealtype == 60) |> 
   st_union() |> 
   st_cast("POLYGON") |> 

### 2024-09-18 R/createMasks.R
ar5 composition

@@ -14,12 +14,21 @@ sea_mask <- filter(ar5, arealtype == 82) |>
 sa <- st_difference(st_intersection(RAD_10m_mask, DTM_mask), sea_mask) |> 
   st_geometry()
 plot(sa)
+st_area(sa) |> 
+  units::set_units("km^2")
 st_write(sa, "data/Orskogfjellet-site.gpkg", "mask_studyarea", 
          append = FALSE)
 
 # AR5 myr ####
 
 ar5 <- st_transform(ar5, st_crs(RAD_10m_mask))
+st_intersection(ar5, sa) |> 
+  mutate(area = st_area(geom)) |> 
+  st_drop_geometry() |> 
+  group_by(arealtype) |> 
+  summarize(area = units::set_units(sum(area), "km^2")) |> 
+  mutate(percent = area / sum((area)) * 100)
+
 ar5.myr <- dplyr::filter(ar5, arealtype == 60) |> 
   st_union() |> 
   st_cast("POLYGON") |> 

### 2024-09-18 R/Skrim/exploratoryDataAnalysis.R
add EDA of study area (summary of preds)

@@ -4,6 +4,14 @@ library(terra)
 library(gstat)
 library(clhs)
 
+# Study area
+
+predictors <- rast("output/Skrim/predictors.tif")
+sa <- st_read("data/Skrim/Skrim-site.gpkg", "fieldsite_outline_utm") |> 
+  st_transform(crs(predictors))
+extract(predictors, sa, ID = FALSE) |> 
+  summary()
+
 # Spatial structure in peat depth ####
 
 depthpts <- read_csv("data/Skrim/depth_all.csv") |> 

### 2024-09-18 R/exploratoryDataAnalysis.R

@@ -4,6 +4,13 @@ library(terra)
 library(gstat)
 library(clhs)
 
+# Study area
+
+predictors <- rast("output/predictors.tif")
+sa <- st_read("data/Orskogfjellet-site.gpkg", "mask_studyarea")
+extract(predictors, sa, ID = FALSE) |> 
+  summary()
+
 # Spatial structure in peat depth ####
 
 depthpts <- read_csv("data/depth/all.csv") |> 

### 2024-09-18 ms/ms.Rmd
qualitative site descriptions

@@
 Introduction text goes here.
-You can change the name of the section if necessary using `\introduction[modified heading]`.
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident,
-sunt in culpa qui officia deserunt mollit anim id est laborum.
-
-Nulla facilisi.
-Maecenas vel nunc nec purus tincidunt congue.
-Proin auctor, lectus eu pharetra malesuada,
-nisi nunc bibendum nunc,
-eget tincidunt nunc nisi id nunc.
-Sed euismod, nunc sit amet aliquam tincidunt,
-nunc nunc tincidunt nunc,
-nec tincidunt nunc nunc nec nunc.
-Donec auctor, nunc sit amet aliquam tincidunt,
-nunc nunc tincidunt nunc,
-nec tincidunt nunc nunc nec nunc.
+Read Gatis et al. [-@gatisMappingUplandPeat2019] and related work [@minasnyDigitalMappingPeatlands2019].
 
@@
 
-## Subsection Heading Here
+## Sites
 
-Read Gatis et al. [-@gatisMappingUplandPeat2019] 
-and related work [@minasnyDigitalMappingPeatlands2019].
+We assessed how well we could predict peat depth at two sites with conspicuously different physical geography: Skrimfjella in eastern Norway and Ørskogfjellet in western Norway (*insert figure ref*).
+These sites were chosen because they were covered by radiometric data from airborne surveys, relatively little built-up area, and road access.
 
-### Subsubsection Heading Here
+At Skrimfjella we delineated a study area of \unit{34\,km^{2}} based on radiometric coverage and accessibility.   
+The landscape within our delineation is classified as *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
+It is almost without human infrastructure, dominated by forest, and borders on a large nature reserve.
+The study area has a mean elevation of \unit{438\,m} above sea level (range 223--711, IQR 351--509), and its mean slope at \unit{10\,m} resolution is 10.8° (IQR 4.6--15.1°).
+In Norway's AR5 national land dataset, \unit{1.5\,km^{2}} (4.5 %) of the study area is classified as peatland.
+
+At Ørskogfjellet we defined a study area of \unit{124\,km^{2}} which basically followed the footprint of the radiometric survey. 
+This study area comprises a wide range of major landscape types: *coastal plains*, *coastal fjord*, *inland valleys*, as well as *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
+It is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
+Its mean elevation is \unit{211\,m} above sea level (range 0--807, IQR 73--310), and its mean slope at \unit{10\,m} resolution is 13.0° (IQR 4.7--18.3°).
+The AR5 data set counts \unit{15.3\,km^{2}} (12.4 %) of the study area as peatland.
+
+## Measuring peat depth
+
+### Skrimfjella
+
+### Ørskogfjellet
 

### 2024-09-18 ms/ms.tex
qualitative site descriptions

@@
 Introduction text goes here.
-You can change the name of the section if necessary using \texttt{\textbackslash{}introduction{[}modified\ heading{]}}.
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident,
-sunt in culpa qui officia deserunt mollit anim id est laborum.
-
-Nulla facilisi.
-Maecenas vel nunc nec purus tincidunt congue.
-Proin auctor, lectus eu pharetra malesuada,
-nisi nunc bibendum nunc,
-eget tincidunt nunc nisi id nunc.
-Sed euismod, nunc sit amet aliquam tincidunt,
-nunc nunc tincidunt nunc,
-nec tincidunt nunc nunc nec nunc.
-Donec auctor, nunc sit amet aliquam tincidunt,
-nunc nunc tincidunt nunc,
-nec tincidunt nunc nunc nec nunc.
+Read Gatis et al. \citeyearpar{gatisMappingUplandPeat2019} and related work \citep{minasnyDigitalMappingPeatlands2019}.
 
@@
 
-\subsection{Subsection Heading Here}
+\subsection{Sites}
 
-Read Gatis et al. \citeyearpar{gatisMappingUplandPeat2019}
-and related work \citep{minasnyDigitalMappingPeatlands2019}.
+We assessed how well we could predict peat depth at two sites with conspicuously different physical geography: Skrimfjella in eastern Norway and Ørskogfjellet in western Norway (\emph{insert figure ref}).
+These sites were chosen because they were covered by radiometric data from airborne surveys, relatively little built-up area, and road access.
 
-\subsubsection{Subsubsection Heading Here}
+At Skrimfjella we delineated a study area of \unit{34\,km^{2}} based on radiometric coverage and accessibility.\\
+The landscape within our delineation is classified as \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
+It is almost without human infrastructure, dominated by forest, and borders on a large nature reserve.
+The study area has a mean elevation of \unit{438\,m} above sea level (range 223--711, IQR 351--509), and its mean slope at \unit{10\,m} resolution is 10.8° (IQR 4.6--15.1°).
+In Norway's AR5 national land dataset, \unit{1.5\,km^{2}} (4.5 \%) of the study area is classified as peatland.
+
+At Ørskogfjellet we defined a study area of \unit{124\,km^{2}} which basically followed the footprint of the radiometric survey.
+This study area comprises a wide range of major landscape types: \emph{coastal plains}, \emph{coastal fjord}, \emph{inland valleys}, as well as \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
+It is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
+Its mean elevation is \unit{211\,m} above sea level (range 0--807, IQR 73--310), and its mean slope at \unit{10\,m} resolution is 13.0° (IQR 4.7--18.3°).
+The AR5 data set counts \unit{15.3\,km^{2}} (12.4 \%) of the study area as peatland.
+
+\subsection{Measuring peat depth}
+
+\subsubsection{Skrimfjella}
+
+\subsubsection{Ørskogfjellet}
 

### 2024-09-18 ms/figures.R
Create sites map inset

@@ -1,0 +1,63 @@
+library(tidyverse)
+library(patchwork)
+library(ggtext)
+library(ggrepel)
+library(sf)
+library(rnaturalearth)
+
+# Sites map ####
+
+ne_download(scale = 50, type = 'populated_places', category = 'cultural', 
+            destdir = "ms/figures/data", load = FALSE)
+# ne_download(scale = 10, type = 'minor_islands', category = 'physical', 
+#             destdir = here("data", "raw", "cartographic"), load = FALSE)
+# 
+cities <- ne_load(scale = 50, type = 'populated_places', category = 'cultural', 
+                  destdir = "ms/figures/data", returnclass = 'sf')
+norcities <- filter(cities, NAME %in% c("Oslo", "Trondheim", "Bergen"))
+# norislands <- ne_load(scale = 10, type = 'minor_islands', category = 'physical', 
+#                 destdir = here("data", "raw", "cartographic"), returnclass = 'sf') %>% 
+#   st_crop(xmin = 4, xmax = 32, ymin = 58, ymax = 72) %>% 
+#   st_transform(crs = 25833) # ETRS89 / UTM zone 33N
+
+nor50 <- ne_countries(scale = 50, country = c("Norway"), returnclass = "sf") %>% 
+  st_transform(crs = 25833)
+orskog <- st_read("data/Orskogfjellet-site.gpkg", "mask_studyarea")
+skrim <- st_read("data/Skrim/Skrim-site.gpkg", "fieldsite_outline_utm")
+both <- bind_rows(orskog,skrim)
+labels <- both |> 
+  st_centroid() |> 
+  mutate(site = c("Ørskogfjellet", "Skrimfjella"))
+
+CBdark2.3class <- c('#d95f02', '#7570b3', '#1b9e77')
+
+g1 <- ggplot() +
+  geom_sf(data = nor50, fill = "white") + 
+  geom_sf(data = norcities, color = "grey50") +
+  geom_sf_text(data = norcities, aes(label = NAME), color = "grey50",
+               nudge_y = c(-1.5e4,1.5e4,2e4), 
+               nudge_x = c(0, 4e4, 0), size = 2) +
+  geom_sf(data = both,
+          mapping = aes(color = 'red', fill = 'red')) +
+  geom_text_repel(
+    data = labels,
+    aes(label = site, geometry = geom),
+    stat = "sf_coordinates",
[…]

### 2024-09-20 R/Skrim/exploratoryDataAnalysis.R
remove oversize pairs plot

@@ -194,9 +194,6 @@ df |>
   complete.cases() |> 
   all()
 
-select(df, !any_of(c(metadata, auxiliary))) |> 
-  slice_sample(n = 1000) |> 
-  GGally::ggpairs()
 corrplot::corrplot(cor(select(df, !any_of(c(metadata, auxiliary)))), 
                    method = 'ellipse', diag = F)
 

### 2024-09-20 ms/ms.Rmd
Add sites figure to ms

@@
 
-We assessed how well we could predict peat depth at two sites with conspicuously different physical geography: Skrimfjella in eastern Norway and Ørskogfjellet in western Norway (*insert figure ref*).
+We assessed how well we could predict peat depth at two sites with conspicuously different physical geography: Skrimfjella in eastern Norway and Ørskogfjellet in western Norway (Fig.
+\@ref(fig:sites)c).
 These sites were chosen because they were covered by radiometric data from airborne surveys, relatively little built-up area, and road access.
 
-At Skrimfjella we delineated a study area of \unit{34\,km^{2}} based on radiometric coverage and accessibility.   
+```{r sites, out.height="90%", echo = FALSE, fig.cap = "Study areas at Ørskogfjellet (a) and Skrimfjella (b) within southern Norway (c).
+Land cover shown here is from the AR50 national land resource database and has simplified geometry with respect to the AR5 database used in the study."}
+knitr::include_graphics("figures/sites-patchwork.pdf")
+```
+
+At Skrimfjella we delineated a study area of \unit{34\,km^{2}} based on radiometric coverage and accessibility (Fig. \@ref(fig:sites)b).   
 The landscape within our delineation is classified as *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
@@
 
-At Ørskogfjellet we defined a study area of \unit{124\,km^{2}} which basically followed the footprint of the radiometric survey. 
+At Ørskogfjellet we defined a study area of \unit{124\,km^{2}} which basically followed the footprint of the radiometric survey (Fig. \@ref(fig:sites)a). 
 This study area comprises a wide range of major landscape types: *coastal plains*, *coastal fjord*, *inland valleys*, as well as *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
@@
 
-Include a 12cm width figure of Nikolaus Copernicus from [Wikipedia](https://en.wikipedia.org/wiki/File:Nikolaus_Kopernikus.jpg) with caption using R Markdown (Figure \@ref(fig:portrait)).
+Include a 12cm width figure of Nikolaus Copernicus from [Wikipedia](https://en.wikipedia.org/wiki/File:Nikolaus_Kopernikus.jpg) with caption using R Markdown (Fig. \@ref(fig:portrait)).
 

### 2024-09-20 ms/ms.tex
Add sites figure to ms

@@
 
-We assessed how well we could predict peat depth at two sites with conspicuously different physical geography: Skrimfjella in eastern Norway and Ørskogfjellet in western Norway (\emph{insert figure ref}).
+We assessed how well we could predict peat depth at two sites with conspicuously different physical geography: Skrimfjella in eastern Norway and Ørskogfjellet in western Norway (Fig.
+\ref{fig:sites}c).
 These sites were chosen because they were covered by radiometric data from airborne surveys, relatively little built-up area, and road access.
 
-At Skrimfjella we delineated a study area of \unit{34\,km^{2}} based on radiometric coverage and accessibility.\\
+\begin{figure}
+\includegraphics[height=0.9\textheight]{figures/sites-patchwork} \caption{Study areas at Ørskogfjellet (a) and Skrimfjella (b) within southern Norway (c).
+Land cover shown here is from the AR50 national land resource database and has simplified geometry with respect to the AR5 database used in the study.}\label{fig:sites}
+\end{figure}
+
+At Skrimfjella we delineated a study area of \unit{34\,km^{2}} based on radiometric coverage and accessibility (Fig. \ref{fig:sites}b).\\
 The landscape within our delineation is classified as \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
@@
 
-At Ørskogfjellet we defined a study area of \unit{124\,km^{2}} which basically followed the footprint of the radiometric survey.
+At Ørskogfjellet we defined a study area of \unit{124\,km^{2}} which basically followed the footprint of the radiometric survey (Fig. \ref{fig:sites}a).
 This study area comprises a wide range of major landscape types: \emph{coastal plains}, \emph{coastal fjord}, \emph{inland valleys}, as well as \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
@@
 
-Include a 12cm width figure of Nikolaus Copernicus from \href{https://en.wikipedia.org/wiki/File:Nikolaus_Kopernikus.jpg}{Wikipedia} with caption using R Markdown (Figure \ref{fig:portrait}).
+Include a 12cm width figure of Nikolaus Copernicus from \href{https://en.wikipedia.org/wiki/File:Nikolaus_Kopernikus.jpg}{Wikipedia} with caption using R Markdown (Fig. \ref{fig:portrait}).
 

### 2024-09-20 ms/figures.R

@@ -1,42 +1,40 @@
+library(sf)
+library(terra)
+library(tidyterra)
 library(tidyverse)
 library(patchwork)
 library(ggtext)
 library(ggrepel)
-library(sf)
+library(ggspatial)
 library(rnaturalearth)
 
 # Sites map ####
 
-ne_download(scale = 50, type = 'populated_places', category = 'cultural', 
-            destdir = "ms/figures/data", load = FALSE)
-# ne_download(scale = 10, type = 'minor_islands', category = 'physical', 
-#             destdir = here("data", "raw", "cartographic"), load = FALSE)
-# 
+## Inset ####
+
+# ne_download(scale = 50, type = 'populated_places', category = 'cultural', 
+#             destdir = "ms/figures/data", load = FALSE)
 cities <- ne_load(scale = 50, type = 'populated_places', category = 'cultural', 
                   destdir = "ms/figures/data", returnclass = 'sf')
 norcities <- filter(cities, NAME %in% c("Oslo", "Trondheim", "Bergen"))
-# norislands <- ne_load(scale = 10, type = 'minor_islands', category = 'physical', 
-#                 destdir = here("data", "raw", "cartographic"), returnclass = 'sf') %>% 
-#   st_crop(xmin = 4, xmax = 32, ymin = 58, ymax = 72) %>% 
-#   st_transform(crs = 25833) # ETRS89 / UTM zone 33N
 
 nor50 <- ne_countries(scale = 50, country = c("Norway"), returnclass = "sf") %>% 
   st_transform(crs = 25833)
-orskog <- st_read("data/Orskogfjellet-site.gpkg", "mask_studyarea")
-skrim <- st_read("data/Skrim/Skrim-site.gpkg", "fieldsite_outline_utm")
-both <- bind_rows(orskog,skrim)
+orskog <- st_read("data/Orskogfjellet-site.gpkg", "mask_studyarea") |> 
+  st_transform(25833)
+skrim <- st_read("data/Skrim/Skrim-site.gpkg", "fieldsite_outline_utm") |> 
+  st_transform(25833)
+both <- bind_rows(orskog, skrim)
 labels <- both |> 
   st_centroid() |> 
   mutate(site = c("Ørskogfjellet", "Skrimfjella"))
-
-CBdark2.3class <- c('#d95f02', '#7570b3', '#1b9e77')
 
 g1 <- ggplot() +
[…]

### 2024-09-26 ms/ms.Rmd
ms: skrimfjella depth and occurrence locations

@@
 The study area has a mean elevation of \unit{438\,m} above sea level (range 223--711, IQR 351--509), and its mean slope at \unit{10\,m} resolution is 10.8° (IQR 4.6--15.1°).
-In Norway's AR5 national land dataset, \unit{1.5\,km^{2}} (4.5 %) of the study area is classified as peatland.
+In Norway's AR5 national land capability dataset [@ahlstromAR5Klassifikasjonssystem2019], \unit{1.5\,km^{2}} (4.5 %) of the study area is classified as mire --- defined as areas with mire vegetation and at least \unit{30\,cm} of peat depth.
 
@@
 Its mean elevation is \unit{211\,m} above sea level (range 0--807, IQR 73--310), and its mean slope at \unit{10\,m} resolution is 13.0° (IQR 4.7--18.3°).
-The AR5 data set counts \unit{15.3\,km^{2}} (12.4 %) of the study area as peatland.
+The AR5 dataset counts \unit{15.3\,km^{2}} (12.4 %) of the study area as mire.
 
-## Measuring peat depth
+## Peat depth measurements
+
+At both study sites, our measurements of peat depth were made for the purpose of training a machine learning model of peat depth, and we designed our sampling with this in mind [@brusSamplingDigitalSoil2019].
+Broadly, we aimed for a sample that was representative the predictor space of the most important predictors of peat depth [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
+A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones [@brusSamplingDigitalSoil2019].
+We chose for our sampling and modelling a spatial resolution of \unit{10\,m}.
+We considered this a reasonable compromise between digital terrain model resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.
 
 ### Skrimfjella
+
+We measured peat depth in selected locations (\unit{10\,m} raster cells) at Skrimfjella.
+The locations were chosen only from areas delineated as mire in the AR5 national land capability dataset.
+Within this mire area, we stratified our sample across values of elevation, slope, and potassium ground concentration [from processed airborne gamma ray spectrometry, @baranwalHelicopterborneMagneticElectromagnetic2013].
+Specifically, we used the *eSample* function in the *iSDM* R package (v.1.0) to chose an environmentally systematic sample.
+This function defines the environmental space as a two-dimensional convex hull around the ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest [@hattabUnifiedFrameworkModel2017].
+Elevation was extracted from the \unit{10\,m} national digital terrain model, slope calculated in degrees, and potassium ground concentration downscaled with bilinear resampling.
+We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and with these parameters *eSample* returned 105 raster cells.
+
+In addition to the peat depth locations, we had another arm of our sampling design for measuring peatland occurrence, as binary variable.
+We wanted to measure peatland occurrence outside of mapped mire areas because the AR5 dataset is known to underestimate peatland coverage [especially in forests, @brynLandCoverNorway2018], and because airborne radiometrics may help identify unmapped peatland [@gatisMappingUplandPeat2019; @olearyDigitalSoilMapping2022].
+The occurrence locations were sampled from the part of the study area that (1) was mapped as something other than mire in the AR5 database and (2) had a slope < 20°. 
+We performed environmentally systematic sampling of this population with the same procedure as for the depth locations, and *eSample* returned 106 raster cells.
+
+Field work at Skrimfjella was conducted in August 2020.
+We navigated to the centers of the raster cells in the depth and occurrence samples by handheld GPS, checking that positional error was below \unit{3\,m}.
+For each depth sample location, we measured peat depth three times (at the vertices of a triangle with \unit{2\,m} sides) to get a more representative value for the \unit{10\,m} raster cell, and to dampen the effect of outlying measurements [@parryEvaluatingApproachesEstimating2014].
+We used a metal probe pushed downward until resistance indicated the base of the peat column.
+Probe locations were adjusted up to \unit{20\,cm} if the base of the peat column seemed to be blocked by an obvious artifact.
+For each occurrence sample location, we recorded the presence or absence of peatland --- primarily by digging and examining the top \unit{20\,cm} of soil (where this was possible).
+We judged whether the soil was a peat soil based on its density, texture, and color.
+Occasionally, when the soil itself was difficult to judge, we made our determination also based on the presence or absence of mire vegetation.
+Although peat soil is strictly defined by organic content (which we did not analyse), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland [@minasnyMappingMonitoringPeatland2023].
+
+GPR data.
 
@@
 Subsubsection text here.
+
+## Peat depth predictors
+
+## Predictive models of peat depth
 

### 2024-09-26 ms/ms.tex
ms: skrimfjella depth and occurrence locations

@@
 The study area has a mean elevation of \unit{438\,m} above sea level (range 223--711, IQR 351--509), and its mean slope at \unit{10\,m} resolution is 10.8° (IQR 4.6--15.1°).
-In Norway's AR5 national land dataset, \unit{1.5\,km^{2}} (4.5 \%) of the study area is classified as peatland.
+In Norway's AR5 national land capability dataset \citep{ahlstromAR5Klassifikasjonssystem2019}, \unit{1.5\,km^{2}} (4.5 \%) of the study area is classified as mire --- defined as areas with mire vegetation and at least \unit{30\,cm} of peat depth.
 
@@
 Its mean elevation is \unit{211\,m} above sea level (range 0--807, IQR 73--310), and its mean slope at \unit{10\,m} resolution is 13.0° (IQR 4.7--18.3°).
-The AR5 data set counts \unit{15.3\,km^{2}} (12.4 \%) of the study area as peatland.
+The AR5 dataset counts \unit{15.3\,km^{2}} (12.4 \%) of the study area as mire.
 
-\subsection{Measuring peat depth}
+\subsection{Peat depth measurements}
+
+At both study sites, our measurements of peat depth were made for the purpose of training a machine learning model of peat depth, and we designed our sampling with this in mind \citep{brusSamplingDigitalSoil2019}.
+Broadly, we aimed for a sample that was representative the predictor space of the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
+A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones \citep{brusSamplingDigitalSoil2019}.
+We chose for our sampling and modelling a spatial resolution of \unit{10\,m}.
+We considered this a reasonable compromise between digital terrain model resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.
 
 \subsubsection{Skrimfjella}
+
+We measured peat depth in selected locations (\unit{10\,m} raster cells) at Skrimfjella.
+The locations were chosen only from areas delineated as mire in the AR5 national land capability dataset.
+Within this mire area, we stratified our sample across values of elevation, slope, and potassium ground concentration \citep[from processed airborne gamma ray spectrometry,][]{baranwalHelicopterborneMagneticElectromagnetic2013}.
+Specifically, we used the \emph{eSample} function in the \emph{iSDM} R package (v.1.0) to chose an environmentally systematic sample.
+This function defines the environmental space as a two-dimensional convex hull around the ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest \citep{hattabUnifiedFrameworkModel2017}.
+Elevation was extracted from the \unit{10\,m} national digital terrain model, slope calculated in degrees, and potassium ground concentration downscaled with bilinear resampling.
+We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and with these parameters \emph{eSample} returned 105 raster cells.
+
+In addition to the peat depth locations, we had another arm of our sampling design for measuring peatland occurrence, as binary variable.
+We wanted to measure peatland occurrence outside of mapped mire areas because the AR5 dataset is known to underestimate peatland coverage \citep[especially in forests,][]{brynLandCoverNorway2018}, and because airborne radiometrics may help identify unmapped peatland \citep{gatisMappingUplandPeat2019, olearyDigitalSoilMapping2022}.
+The occurrence locations were sampled from the part of the study area that (1) was mapped as something other than mire in the AR5 database and (2) had a slope \textless{} 20°.
+We performed environmentally systematic sampling of this population with the same procedure as for the depth locations, and \emph{eSample} returned 106 raster cells.
+
+Field work at Skrimfjella was conducted in August 2020.
+We navigated to the centers of the raster cells in the depth and occurrence samples by handheld GPS, checking that positional error was below \unit{3\,m}.
+For each depth sample location, we measured peat depth three times (at the vertices of a triangle with \unit{2\,m} sides) to get a more representative value for the \unit{10\,m} raster cell, and to dampen the effect of outlying measurements \citep{parryEvaluatingApproachesEstimating2014}.
+We used a metal probe pushed downward until resistance indicated the base of the peat column.
+Probe locations were adjusted up to \unit{20\,cm} if the base of the peat column seemed to be blocked by an obvious artifact.
+For each occurrence sample location, we recorded the presence or absence of peatland --- primarily by digging and examining the top \unit{20\,cm} of soil (where this was possible).
+We judged whether the soil was a peat soil based on its density, texture, and color.
+Occasionally, when the soil itself was difficult to judge, we made our determination also based on the presence or absence of mire vegetation.
+Although peat soil is strictly defined by organic content (which we did not analyse), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland \citep{minasnyMappingMonitoringPeatland2023}.
+
+GPR data.
 
@@
 Subsubsection text here.
+
+\subsection{Peat depth predictors}
+
+\subsection{Predictive models of peat depth}
 

### 2024-09-27 ms/ms.Rmd
ms: skrim GPR data including new wave speed calibration

@@
 
-GPR data.
+Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively-chosen, individual mires, using ground-penetrating radar (GPR).
+We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with its \unit{500\,MHz} shielded antenna mounted in a plastic sledge, and its control unit connected to a GNSS receiver.
+At each of the three mires we recorded GPR traces along walking transects that covered the extent of the mire, mostly in traversing, zigzag patterns with between \unit{5\,m} and \unit{20\,m} spacing at their vertices.
+Along the GPR transects we also probed peat depth at marked trace locations, to be able to calibrate the GPR wave speed velocity.
+We processed the GPR data with Reflex2DQuick software (v.3.0; Sandmeier Scientific Software, Germany), applying a time-zero correction, a dewow filter, and a gain filter based on observed energy decay.
+Then we picked strong reflectors in the radargrams that we interpreted as the base of the peat column.
+We used picks at marked trace locations to calibrate wave speed velocity; we pooled calibration points across the three mires and fitted a linear regression of depth on one-way travel time with the intercept fixed at zero.
+In total we had 46 calibration points along \unit{3.5\,km} of GPR transects.
+Finally, we used the calculated wave velocity (\unit{0.0387\,m\,ns^{-1}}, $R^2 = 0.874$) to convert the travel times of all picks to calibrated peat depths.
 

### 2024-09-27 ms/ms.tex
ms: skrim GPR data including new wave speed calibration

@@
 
-GPR data.
+Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively-chosen, individual mires, using ground-penetrating radar (GPR).
+We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with its \unit{500\,MHz} shielded antenna mounted in a plastic sledge, and its control unit connected to a GNSS receiver.
+At each of the three mires we recorded GPR traces along walking transects that covered the extent of the mire, mostly in traversing, zigzag patterns with between \unit{5\,m} and \unit{20\,m} spacing at their vertices.
+Along the GPR transects we also probed peat depth at marked trace locations, to be able to calibrate the GPR wave speed velocity.
+We processed the GPR data with Reflex2DQuick software (v.3.0; Sandmeier Scientific Software, Germany), applying a time-zero correction, a dewow filter, and a gain filter based on observed energy decay.
+Then we picked strong reflectors in the radargrams that we interpreted as the base of the peat column.
+We used picks at marked trace locations to calibrate wave speed velocity; we pooled calibration points across the three mires and fitted a linear regression of depth on one-way travel time with the intercept fixed at zero.
+In total we had 46 calibration points along \unit{3.5\,km} of GPR transects.
+Finally, we used the calculated wave velocity (\unit{0.0387\,m\,ns^{-1}}, \(R^2 = 0.874\)) to convert the travel times of all picks to calibrated peat depths.
 

### 2024-09-27 R/Skrim/concatenateDepthMeasurements.R

@@ -14,12 +14,51 @@ filter(probe, is.na(depth_cm))
 
 # GPR data ####
 
+## Calibrate wave velocity ####
+caldata <- read_csv("data/Skrim/GPRcalibration.csv")
+
+m0 <- lm(depth ~ OWTT + 0, data = caldata)
+summary(m0) # OWTT  0.03867
+
+modelfit <- data.frame(OWTT = seq(0, max(caldata$OWTT)*1.05, length.out = 100))
+modelfit <- modelfit %>% 
+  bind_cols(predict(m0, modelfit, se.fit = TRUE, interval = "confidence")$fit) %>% 
+  rename(depth = fit)
+
+calplot <- caldata %>% 
+  ggplot(aes(x=OWTT, y=depth)) +
+  geom_point(pch=16, size=2) +
+  geom_ribbon(data=modelfit, aes(ymin=lwr, ymax=upr), alpha=0.4) +
+  geom_line(data=modelfit) +
+  geom_abline(slope = 0.03, intercept = 0, color="red", lty=2) +
+  annotate("text", x = 60, y = 3.5, label = "R-squared = 0.874") +
+  annotate("text", x = 60, y = 1.7, label = "0.03 m/ns (fresh water)", color="red", angle=22) +
+  annotate("text", x = 60, y = 2.5, label = "0.039 m/ns", angle=28) +
+  labs(x = "one-way travel time (ns)", y = "depth (m)") +
+  coord_cartesian(expand = FALSE) +
+  theme_bw()
+
+ggsave('output/Skrim/calibrationGPR.svg', calplot, width = 119, height = 84, units = "mm")
+v <- units::set_units(m0$coefficients[[1]], "m/ns")
+
+## TWTT to depth ####
 mire1 <- st_read("data/Skrim/GPRpicks.gpkg", "mire1")
 mire2 <- st_read("data/Skrim/GPRpicks.gpkg", "mire2")
 mire3 <- st_read("data/Skrim/GPRpicks.gpkg", "mire3")
 gpr <- bind_rows(mire1, mire2, mire3) |> 
-  mutate(depth_cm = Depth*100) |> 
+  mutate(depth_cm = TWT/2*units::drop_units(v)*100) |> 
   select(depth_cm)
+
+gprlines <- bind_rows(mire1, mire2, mire3) %>% 
+  group_by(Name) |> 
+  arrange(Trace) |>
+  select(Name) |>
[…]

### 2024-09-30 ms/ms.Rmd
ms: Ørskog sample size

@@
 
-Subsubsection text here.
+At Ørskogfjellet we also measured peat depth in a sample of \unit{10\,m} raster cells, selected from the part of the study area classified in the AR5 dataset as mire.
+Before selecting locations, we determined a minimal sample size that would adequately capture the terrain and radiometric properties of the entire mire area.
+Specifically, we aimed to identify the size at which adding locations produced diminishing decreases in divergence between sample and population distributions --- i.e.
+the elbow point in a curve of similarity between sample and population [@maloneMethodsImproveUtility2019].
+This approach has been found to identify sample sizes that correspond with diminishing returns in predictive model performance on external evaluation data [@sauretteDivergenceMetricsDetermining2023].
+We defined a sequence of increasing sample sizes (50--500) and for ten replicate samples at each size [drawn by conditioned latin hypercube sampling, @minasnyConditionedLatinHypercube2006; @roudierClhsPackageConditioned2011], we calculated the mean Kullback-Leibler divergence between sample and population distributions [@maloneMethodsImproveUtility2019; @sauretteDivergenceMetricsDetermining2023].
+The variables in the divergence calculation were terrain slope and four radiometrics: potassium, thorium, uranium, and total count.
+Next, we fitted a asymptotic regression of mean divergence on sample size, and identified the sample size at which the curve reached 95 % of the fitted asymptote.
+Through this procedure we found that we could adequately capture the population distribution with a sample of 160 locations.
 

### 2024-09-30 ms/ms.tex
ms: Ørskog sample size

@@
 
-Subsubsection text here.
+At Ørskogfjellet we also measured peat depth in a sample of \unit{10\,m} raster cells, selected from the part of the study area classified in the AR5 dataset as mire.
+Before selecting locations, we determined a minimal sample size that would adequately capture the terrain and radiometric properties of the entire mire area.
+Specifically, we aimed to identify the size at which adding locations produced diminishing decreases in divergence between sample and population distributions --- i.e.~the elbow point in a curve of similarity between sample and population \citep{maloneMethodsImproveUtility2019}.
+This approach has been found to identify sample sizes that correspond with diminishing returns in predictive model performance on external evaluation data \citep{sauretteDivergenceMetricsDetermining2023}.
+We defined a sequence of increasing sample sizes (50--500) and for ten replicate samples at each size \citep[drawn by conditioned latin hypercube sampling,][]{minasnyConditionedLatinHypercube2006, roudierClhsPackageConditioned2011}, we calculated the mean Kullback-Leibler divergence between sample and population distributions \citep{maloneMethodsImproveUtility2019, sauretteDivergenceMetricsDetermining2023}.
+The variables in the divergence calculation were terrain slope and four radiometrics: potassium, thorium, uranium, and total count.
+Next, we fitted a asymptotic regression of mean divergence on sample size, and identified the sample size at which the curve reached 95 \% of the fitted asymptote.
+Through this procedure we found that we could adequately capture the population distribution with a sample of 160 locations.
 

### 2024-09-30 ms/ms.Rmd
ms: add bedrock to sites

@@
 At Skrimfjella we delineated a study area of \unit{34\,km^{2}} based on radiometric coverage and accessibility (Fig. \@ref(fig:sites)b).
+The study area has a diverse bedrock, with 32 % alkali feltspat granite, 26 % mergelstein, 10 % granite, and eight other types with > 1 % coverage (NGU, 1:250 000 dataset).
 The landscape within our delineation is classified as *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
@@
 At Ørskogfjellet we defined a study area of \unit{124\,km^{2}} which basically followed the footprint of the radiometric survey (Fig. \@ref(fig:sites)a).
+According to the Geological Survey of Norway, bedrock in the area is 84 % granitic gneiss,  11 % granite, and 5 % aluminium silicate gneiss (NGU, 1:250 000 dataset). 
 This study area comprises a wide range of major landscape types: *coastal plains*, *coastal fjord*, *inland valleys*, as well as *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].

### 2024-09-30 ms/ms.tex
ms: add bedrock to sites

@@
 
-At Skrimfjella we delineated a study area of \unit{34\,km^{2}} based on radiometric coverage and accessibility (Fig. \ref{fig:sites}b).\\
+At Skrimfjella we delineated a study area of \unit{34\,km^{2}} based on radiometric coverage and accessibility (Fig. \ref{fig:sites}b).
+The study area has a diverse bedrock, with 32 \% alkali feltspat granite, 26 \% mergelstein, 10 \% granite, and eight other types with \textgreater{} 1 \% coverage (NGU, 1:250 000 dataset).
 The landscape within our delineation is classified as \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
@@
 At Ørskogfjellet we defined a study area of \unit{124\,km^{2}} which basically followed the footprint of the radiometric survey (Fig. \ref{fig:sites}a).
+According to the Geological Survey of Norway, bedrock in the area is 84 \% granitic gneiss, 11 \% granite, and 5 \% aluminium silicate gneiss (NGU, 1:250 000 dataset).
 This study area comprises a wide range of major landscape types: \emph{coastal plains}, \emph{coastal fjord}, \emph{inland valleys}, as well as \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.

### 2024-09-30 R/Skrim/createMasks.R

@@ -32,6 +32,22 @@ ar5.myr <- dplyr::filter(ar5, arealtype == 60) |>
 st_write(ar5.myr, "data/Skrim/Skrim-site.gpkg", "mask_ar5myr", 
          append = FALSE)
 
+# Bedrock ####
+
+bedrock <- st_read("data/Skrim/Geologi_33_BerggrunnN250/BerggrunnN250.gdb", 
+                   layer="BergartFlate_N250") |> 
+  select(contains("bergart"))
+bedrock <- st_transform(bedrock, crssite) |> 
+  group_by(hovedbergart, hovedbergart_navn) |>
+  summarize()
+st_intersection(bedrock, sa) |> 
+  mutate(area = st_area(SHAPE)) |> 
+  st_drop_geometry() |> 
+  group_by(hovedbergart, hovedbergart_navn) |> 
+  summarize(area = units::set_units(sum(area), "km^2"), .groups = "drop") |> 
+  mutate(percent = area / sum((area)) * 100) |> 
+  arrange(desc(percent))
+
 # Model predictive domain ####
 
 predictivedomain <- st_intersection(sa, ar5.myr) |> 
@@ -41,3 +57,7 @@ predictivedomain <- st_intersection(sa, ar5.myr) |>
 plot(predictivedomain)
 st_write(predictivedomain, "data/Skrim/Skrim-site.gpkg", "mask_predictivedomain", 
          append = FALSE)
+
+# sessionInfo ####
+
+sessioninfo::session_info()

### 2024-09-30 R/createMasks.R

@@ -36,6 +36,22 @@ ar5.myr <- dplyr::filter(ar5, arealtype == 60) |>
 st_write(ar5.myr, "data/Orskogfjellet-site.gpkg", "mask_ar5myr", 
          append = FALSE)
 
+# Bedrock ####
+
+bedrock <- st_read("data/Geologi_15_BerggrunnN250/BerggrunnN250.gdb", 
+                   layer="BergartFlate_N250") |> 
+  select(contains("bergart"))
+bedrock <- st_transform(bedrock, st_crs(RAD_10m_mask)) |> 
+  group_by(hovedbergart, hovedbergart_navn) |>
+  summarize()
+st_intersection(bedrock, sa) |> 
+  mutate(area = st_area(SHAPE)) |> 
+  st_drop_geometry() |> 
+  group_by(hovedbergart, hovedbergart_navn) |> 
+  summarize(area = units::set_units(sum(area), "km^2"), .groups = "drop") |> 
+  mutate(percent = area / sum((area)) * 100) |> 
+  arrange(desc(percent))
+
 # Model predictive domain ####
 
 predictivedomain <- st_intersection(sa, ar5.myr) |> 
@@ -45,3 +61,7 @@ predictivedomain <- st_intersection(sa, ar5.myr) |>
 plot(predictivedomain)
 st_write(predictivedomain, "data/Orskogfjellet-site.gpkg", "mask_predictivedomain", 
          append = FALSE)
+
+# sessionInfo ####
+
+sessioninfo::session_info()

### 2024-10-01 ms/ms.Rmd
ms: add data license info for DTM, AR5, DMK, berggrunn

@@
 We also thank Vikas Baranwal from the Geological Survey of Norway for helping us access the radiometric data from Skrim.
+This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, © Kartverket, (2) *Norge digitalt* license, Norwegian Institute of Bioeconomy Research (NIBIO), © Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway (NGU).
 appendix: |

### 2024-10-01 ms/ms.tex
ms: add data license info for DTM, AR5, DMK, berggrunn

@@
 We also thank Vikas Baranwal from the Geological Survey of Norway for helping us access the radiometric data from Skrim.
+This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, © Kartverket, (2) \emph{Norge digitalt} license, Norwegian Institute of Bioeconomy Research (NIBIO), © Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway (NGU).
 \end{acknowledgements}

### 2024-10-01 ms/ms.Rmd
ms: FSCS at Ørskogfjellet

@@
 
-At both study sites, our measurements of peat depth were made for the purpose of training a machine learning model of peat depth, and we designed our sampling with this in mind [@brusSamplingDigitalSoil2019].
+At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest model of peat depth, and we designed our sampling with this in mind [@brusSamplingDigitalSoil2019].
 Broadly, we aimed for a sample that was representative the predictor space of the most important predictors of peat depth [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
@@
 Through this procedure we found that we could adequately capture the population distribution with a sample of 160 locations.
+
+To choose 160 locations, we performed feature space coverage sampling.
+This approach has been found to produce higher accuracy in Random Forest models than conditioned latin hypercube sampling [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
+Feature space coverage sampling aims to disperse samples as uniformly as possible in multidimensional predictor space, and is implemented by choosing locations that are closest to cluster centers in a k-means clustering of the standardized predictor space [@brusSamplingDigitalSoil2019].
+Feature space coverage sampling works best when all dimensions are important predictors of the outcome [@wadouxSamplingDesignOptimization2019], and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
+We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
+Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
+To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly-accessible roads.
+Cost distance was calculated using GRASS's r.walk function, with friction costs defined by AR5 land classes [@GRASSv8-2].
+After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
+Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 % (with respect to the ideal sample), but distance in our sample was still only 46 % of the mean distance to cluster centers --- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
 

### 2024-10-01 ms/ms.tex
ms: FSCS at Ørskogfjellet

@@
 
-At both study sites, our measurements of peat depth were made for the purpose of training a machine learning model of peat depth, and we designed our sampling with this in mind \citep{brusSamplingDigitalSoil2019}.
+At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest model of peat depth, and we designed our sampling with this in mind \citep{brusSamplingDigitalSoil2019}.
 Broadly, we aimed for a sample that was representative the predictor space of the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
@@
 Through this procedure we found that we could adequately capture the population distribution with a sample of 160 locations.
+
+To choose 160 locations, we performed feature space coverage sampling.
+This approach has been found to produce higher accuracy in Random Forest models than conditioned latin hypercube sampling \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
+Feature space coverage sampling aims to disperse samples as uniformly as possible in multidimensional predictor space, and is implemented by choosing locations that are closest to cluster centers in a k-means clustering of the standardized predictor space \citep{brusSamplingDigitalSoil2019}.
+Feature space coverage sampling works best when all dimensions are important predictors of the outcome \citep{wadouxSamplingDesignOptimization2019}, and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
+We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
+Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
+To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly-accessible roads.
+Cost distance was calculated using GRASS's r.walk function, with friction costs defined by AR5 land classes \citep{GRASSv8-2}.
+After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
+Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 \% (with respect to the ideal sample), but distance in our sample was still only 46 \% of the mean distance to cluster centers --- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
 

### 2024-10-01 R/sampling/03_makeFSCS.R

@@ -190,6 +190,9 @@ distadj <- diag(adjrdist)
 #Restricting access by r.walk.cost and manually increases distances by %: 
 #(using per cluster minimums and means as references)
 (sum(distadj) - sum(mindist))/(sum(meandist) - sum(mindist))
+sum(distadj - mindist)/sum(meandist - mindist)
+sum(distadj)/sum(mindist) #1.776308
+sum(distadj)/sum(meandist) #0.4621858
 
 plot(meandist, ylim=c(0,max(meandist)))
 points(mindist, col='blue')

### 2024-10-02 ms/ms.Rmd
ms: minor edits

@@
 This approach has been found to identify sample sizes that correspond with diminishing returns in predictive model performance on external evaluation data [@sauretteDivergenceMetricsDetermining2023].
-We defined a sequence of increasing sample sizes (50--500) and for ten replicate samples at each size [drawn by conditioned latin hypercube sampling, @minasnyConditionedLatinHypercube2006; @roudierClhsPackageConditioned2011], we calculated the mean Kullback-Leibler divergence between sample and population distributions [@maloneMethodsImproveUtility2019; @sauretteDivergenceMetricsDetermining2023].
+We defined a sequence of sample sizes (50--500) and for ten replicate samples at each size [drawn by conditioned latin hypercube sampling, @minasnyConditionedLatinHypercube2006; @roudierClhsPackageConditioned2011], we calculated the mean Kullback-Leibler divergence between sample and population distributions [@maloneMethodsImproveUtility2019; @sauretteDivergenceMetricsDetermining2023].
 The variables in the divergence calculation were terrain slope and four radiometrics: potassium, thorium, uranium, and total count.
@@
 To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly-accessible roads.
-Cost distance was calculated using GRASS's r.walk function, with friction costs defined by AR5 land classes [@GRASSv8-2].
+Cost distance was calculated using GRASS's *r.walk* function, with friction costs defined by AR5 land classes [@GRASSv8-2].
 After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.

### 2024-10-02 ms/ms.Rmd
ms: Orskog field work

@@
 
+Field work at Ørskogfjellet was conducted in August 2023.
+We navigated to the centers of the raster cells in the sample using real time kinematic differential GNSS (Topcon Positioning Systems, USA), to ensure sub-meter positional accuracy.
+At each location we measured peat depth three times by manual probing, with probe locations spaced approximately \unit{2.5\,m} apart.
+In areas with dense sampling locations, we also measured peat depth with GPR along snaking transects passing through the centers of the sampling cells (seven transects, \unit{6.2\,km} total length).
+We used the same GPR system as at Skrimfjella, but with a \unit{100\,MHz} Malå rough terrain antenna (Guideline Geo AB, Sweden) at some transects.
+To navigate the GPR transects, we placed flags at the cell centers of the sample locations, and used a handheld GNSS receiver to guide the GPR operator.
+At sampling locations crossed by a GPR transect, we arranged the manual probe positions along the transect (for better calibration of the GPR wave speed velocity), while other locations were probed in a triangular pattern around the cell center like at Skrimfjella.
+
+We processed the GPR data with Reflexw software (v.8.5; Sandmeier Scientific Software, Germany), applying a dewow filter, time-zero correction, bandpass filter, gain filter, and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennae.
+The last correction is important for the rough terrain antenna, which has an antenna separation (\unit{2.2\,m}) --- comparable to typical peat depths.
+As with the data from Skrimfjella, we picked the base of the peat column from strong reflectors in the radargram, and calibrated wave velocity with manual probe measurements in a linear regression.
+The points in the regression were created by joining to each probe measurement the travel time of the nearest pick, but only if these were within \unit{2\,m} of each other.
+In total we had 78 calibration points along \unit{7.8\,km} of interpretable GPR traces (transect length exceeded because of extra GPR data).
+Finally, we used the calculated wave velocity (\unit{0.0427\,m\,ns^{-1}}, $R^2 = 0.946$) to convert the travel times of all picks to calibrated peat depths.
+
 ## Peat depth predictors

### 2024-10-02 ms/ms.tex
ms: Orskog field work

@@
 This approach has been found to identify sample sizes that correspond with diminishing returns in predictive model performance on external evaluation data \citep{sauretteDivergenceMetricsDetermining2023}.
-We defined a sequence of increasing sample sizes (50--500) and for ten replicate samples at each size \citep[drawn by conditioned latin hypercube sampling,][]{minasnyConditionedLatinHypercube2006, roudierClhsPackageConditioned2011}, we calculated the mean Kullback-Leibler divergence between sample and population distributions \citep{maloneMethodsImproveUtility2019, sauretteDivergenceMetricsDetermining2023}.
+We defined a sequence of sample sizes (50--500) and for ten replicate samples at each size \citep[drawn by conditioned latin hypercube sampling,][]{minasnyConditionedLatinHypercube2006, roudierClhsPackageConditioned2011}, we calculated the mean Kullback-Leibler divergence between sample and population distributions \citep{maloneMethodsImproveUtility2019, sauretteDivergenceMetricsDetermining2023}.
 The variables in the divergence calculation were terrain slope and four radiometrics: potassium, thorium, uranium, and total count.
@@
 To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly-accessible roads.
-Cost distance was calculated using GRASS's r.walk function, with friction costs defined by AR5 land classes \citep{GRASSv8-2}.
+Cost distance was calculated using GRASS's \emph{r.walk} function, with friction costs defined by AR5 land classes \citep{GRASSv8-2}.
 After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
 Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 \% (with respect to the ideal sample), but distance in our sample was still only 46 \% of the mean distance to cluster centers --- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
+
+Field work at Ørskogfjellet was conducted in August 2023.
+We navigated to the centers of the raster cells in the sample using real time kinematic differential GNSS (Topcon Positioning Systems, USA), to ensure sub-meter positional accuracy.
+At each location we measured peat depth three times by manual probing, with probe locations spaced approximately \unit{2.5\,m} apart.
+In areas with dense sampling locations, we also measured peat depth with GPR along snaking transects passing through the centers of the sampling cells (seven transects, \unit{6.2\,km} total length).
+We used the same GPR system as at Skrimfjella, but with a \unit{100\,MHz} Malå rough terrain antenna (Guideline Geo AB, Sweden) at some transects.
+To navigate the GPR transects, we placed flags at the cell centers of the sample locations, and used a handheld GNSS receiver to guide the GPR operator.
+At sampling locations crossed by a GPR transect, we arranged the manual probe positions along the transect (for better calibration of the GPR wave speed velocity), while other locations were probed in a triangular pattern around the cell center like at Skrimfjella.
+
+We processed the GPR data with Reflexw software (v.8.5; Sandmeier Scientific Software, Germany), applying a dewow filter, time-zero correction, bandpass filter, gain filter, and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennae.
+The last correction is important for the rough terrain antenna, which has an antenna separation (\unit{2.2\,m}) --- comparable to typical peat depths.
+As with the data from Skrimfjella, we picked the base of the peat column from strong reflectors in the radargram, and calibrated wave velocity with manual probe measurements in a linear regression.
+The points in the regression were created by joining to each probe measurement the travel time of the nearest pick, but only if these were within \unit{2\,m} of each other.
+In total we had 78 calibration points along \unit{7.8\,km} of interpretable GPR traces (transect length exceeded because of extra GPR data).
+Finally, we used the calculated wave velocity (\unit{0.0427\,m\,ns^{-1}}, \(R^2 = 0.946\)) to convert the travel times of all picks to calibrated peat depths.
 

### 2024-10-02 R/concatenateDepthMeasurements.R

@@ -29,6 +29,36 @@ gprpicks <- st_read("data/depth/GPRpicks.gpkg", "picks-round2")
 gprpicks <- select(gprpicks, file = field_1, trace =field_4, TWTT = field_6) |> 
   st_transform(st_crs(probe))
 
+gprlines <- gprpicks %>% 
+  group_by(file) |> 
+  arrange(trace) |>
+  select(file) |>
+  summarize(file = first(file), do_union=FALSE) |> 
+  st_cast("LINESTRING")
+gprlines |> 
+  mutate(length = st_length(gprlines)) |> 
+  pull(length) |> 
+  sum()
+
+# Distance threshold to divide trace sequences into separate lines
+threshold_distance <- 10
+buffers_sf <- st_buffer(gprpicks, dist = threshold_distance/2) |> 
+  st_union() |> 
+  st_cast("POLYGON") |> 
+  st_sf() |> 
+  rowid_to_column("group_id")
+gprlines <- gprpicks |> 
+  st_join(buffers_sf) |> 
+  arrange(trace) |>
+  group_by(file, group_id) |>
+  summarize(file = first(file), group_id = first(group_id), 
+            do_union=FALSE, .groups = "drop") |> 
+  st_cast("LINESTRING")
+gprlines |> 
+  mutate(length = st_length(gprlines)) |> 
+  pull(length) |> 
+  sum()
+
 ## Convert GPR TWTT to depth ####
 
 ### Spatially join probe to GPR ####
@@ -142,4 +172,5 @@ all |>
   group_by(source) |> 
   summarize(n = n(), meandepth=mean(depth_cm))
 
+st_delete("data/depth/all.csv")
 st_write(all, "data/depth/all.csv", layer_options = "GEOMETRY=AS_XY", append=FALSE)

### 2024-10-02 ms/ms.Rmd
ms: Orskog legacy data

@@
 
+We also used two sets of existing depth measurements from Ørskogfjellet.
+The first set was provided by the Norwegian Public Roads Administration, who commissioned GPR surveys of particular peatland areas in 2020 and 2021.
+The surveys were conducted with a dual channel system (\unit{70\,MHz} and \unit{300\,MHz}; ImpulseRadar AB, Sweden), connected to GNSS with CPOS correction.
+We used interpreted and calibrated traces from these surveys, and discarded some data where multiple depths were interpreted for the same locations.
+This summed to \unit{7.4\,km} of interpreted traces.
+The second set of existing depth data we extracted from a paper map made by the Norwegian Soil and Mire Company in 1984.
+This map presents 44 borehole depths (in decimeters) across a \unit{9\,ha} peatland area.
+We georeferenced the map and digitized the borehole locations and depths.
+
 ## Peat depth predictors

### 2024-10-02 ms/ms.tex
ms: Orskog legacy data

@@
 
+We also used two sets of existing depth measurements from Ørskogfjellet.
+The first set was provided by the Norwegian Public Roads Administration, who commissioned GPR surveys of particular peatland areas in 2020 and 2021.
+The surveys were conducted with a dual channel system (\unit{70\,MHz} and \unit{300\,MHz}; ImpulseRadar AB, Sweden), connected to GNSS with CPOS correction.
+We used interpreted and calibrated traces from these surveys, and discarded some data where multiple depths were interpreted for the same locations.
+This summed to \unit{7.4\,km} of interpreted traces.
+The second set of existing depth data we extracted from a paper map made by the Norwegian Soil and Mire Company in 1984.
+This map presents 44 borehole depths (in decimeters) across a \unit{9\,ha} peatland area.
+We georeferenced the map and digitized the borehole locations and depths.
+
 \subsection{Peat depth predictors}

### 2024-10-02 R/concatenateDepthMeasurements.R

@@ -1,6 +1,8 @@
 library(tidyverse)
 library(sf)
 
+sa <- st_read("data/Orskogfjellet-site.gpkg", "mask_studyarea")
+
 # Probe data ####
 
 probe <- read_csv("data/depth/concatenatedRawProbe.csv")
@@ -151,6 +153,28 @@ wisen2021 <- mutate(wisen2021, hasSpatDuplicate = map_lgl(dupcheck, \(x) length(
 
 # Filter out duplicates, because some show discrepancies suggestive of error (@wisenE39Romsdalsfjorden202020212021)
 wisen2021 <- filter(wisen2021, !hasSpatDuplicate)
+
+# Sum up profile length
+wisen2021.32632 <- wisen2021 |> 
+  st_transform(st_crs(sa))
+wisen2021.32632 <- wisen2021.32632[sa,]
+threshold_distance <- 10
+buffers_sf <- st_buffer(wisen2021.32632, dist = threshold_distance/2) |> 
+  st_union() |> 
+  st_cast("POLYGON") |> 
+  st_sf() |> 
+  rowid_to_column("group_id")
+wisen2021lines <- wisen2021.32632 |> 
+  st_join(buffers_sf) |> 
+  arrange(file, line, trace) |>
+  group_by(file, line) |>
+  summarize(file = first(file), line = first(line), 
+            do_union=FALSE, .groups = "drop") |> 
+  st_cast("LINESTRING")
+wisen2021lines |> 
+  mutate(length = st_length(wisen2021lines)) |> 
+  pull(length) |> 
+  sum()
 
 # Myrarkivet ####
 

### 2024-10-02 ms/ms.Rmd
ms: predictor classes + radiometric surveys

@@
 
+We created the same suite of 25 quantitative peat depth predictors for both sites.
+All predictors were derived either from an airborne radiometric survey or from a digital terrain model.
+From the radiometric surveys we simply used the four variables produced by the surveyors (Geological Survey of Norway): ground concentration of Potassium, Thorium, Uranium, as well as total count.
+From the digital terrain models we calculated several simple terrain variables and a series of more complex geomorphometric and hydrological variables.
+
+The radiometric survey covering Skrimfjella was conducted in 2008--2011. 
+The survey was flown at an average altitude of \unit{75\,m} and average speed of \unit{108\,km\,h^{-1}}, with flight lines spaced \unit{200\,m} apart.
+Spectrometer count rates were calibrated annually to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
+The Geological Survey of Norway processed data from the spectrometer following standard procedures outlined by the International Atomic Energy Association, and the processing included: correction for aircraft and cosmic background radiation, correction for radon in the air, window stripping of the gamma ray spectrum, correction for flying height, conversion of count rates to ground concentrations, and finally gridding to \unit{50\,m} resolution with micro-leveling.
+Further details about the survey and data processing are provided in Baranwal et al. [-@baranwalHelicopterborneMagneticElectromagnetic2013].
+
+A very similar radiometric survey covering Ørskogfjellet was conducted in December 2014 and January 2015. 
+This survey was flown at an average altitude of \unit{80\,m} and average speed of \unit{88\,km\,h^{-1}}, with flight lines also spaced \unit{200\,m} apart.
+Spectrometer count rates were calibrated in 2013 to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
+The spectrometer data were processed following the same procedure as for the survey at Skrimfjella, except that a convolution filter was added to smoothen the gridded data.
+Further details about the survey and data processing are provided in Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015].
+
 ## Predictive models of peat depth

### 2024-10-02 ms/ms.tex
ms: predictor classes + radiometric surveys

@@
 
+We created the same suite of 25 quantitative peat depth predictors for both sites.
+All predictors were derived either from an airborne radiometric survey or from a digital terrain model.
+From the radiometric surveys we simply used the four variables produced by the surveyors (Geological Survey of Norway): ground concentration of Potassium, Thorium, Uranium, as well as total count.
+From the digital terrain models we calculated several simple terrain variables and a series of more complex geomorphometric and hydrological variables.
+
+The radiometric survey covering Skrimfjella was conducted in 2008--2011.
+The survey was flown at an average altitude of \unit{75\,m} and average speed of \unit{108\,km\,h^{-1}}, with flight lines spaced \unit{200\,m} apart.
+Spectrometer count rates were calibrated annually to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
+The Geological Survey of Norway processed data from the spectrometer following standard procedures outlined by the International Atomic Energy Association, and the processing included: correction for aircraft and cosmic background radiation, correction for radon in the air, window stripping of the gamma ray spectrum, correction for flying height, conversion of count rates to ground concentrations, and finally gridding to \unit{50\,m} resolution with micro-leveling.
+Further details about the survey and data processing are provided in Baranwal et al. \citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013}.
+
+A very similar radiometric survey covering Ørskogfjellet was conducted in December 2014 and January 2015.
+This survey was flown at an average altitude of \unit{80\,m} and average speed of \unit{88\,km\,h^{-1}}, with flight lines also spaced \unit{200\,m} apart.
+Spectrometer count rates were calibrated in 2013 to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
+The spectrometer data were processed following the same procedure as for the survey at Skrimfjella, except that a convolution filter was added to smoothen the gridded data.
+Further details about the survey and data processing are provided in Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.
+
 \subsection{Predictive models of peat depth}

### 2024-10-03 ms/ms.Rmd
ms: add predictors table

@@
 
-We created the same suite of 25 quantitative peat depth predictors for both sites.
+We created the same suite of 25 quantitative peat depth predictors for both sites (Table \@ref(tab:preds)).
 All predictors were derived either from an airborne radiometric survey or from a digital terrain model.
@@
 From the digital terrain models we calculated several simple terrain variables and a series of more complex geomorphometric and hydrological variables.
+
+```{r preds, echo=FALSE, message=FALSE}
+df <- readr::read_csv("tables/predictors.csv")
+knitr::kable(df, booktabs = FALSE, caption = "Quantitative predictors of peat depth.")
+```
 

### 2024-10-03 ms/ms.tex
ms: add predictors table

@@
 
-We created the same suite of 25 quantitative peat depth predictors for both sites.
+We created the same suite of 25 quantitative peat depth predictors for both sites (Table \ref{tab:preds}).
 All predictors were derived either from an airborne radiometric survey or from a digital terrain model.
@@
 From the digital terrain models we calculated several simple terrain variables and a series of more complex geomorphometric and hydrological variables.
+
+\begin{table}
+
+\caption{\label{tab:preds}Quantitative predictors of peat depth.}
+\centering
+\begin{tabular}[t]{l|l}
+\hline
+name & description\\
+\hline
+radK & Potassium ground concentration\\
+\hline
+radTh & Thorium ground concentration\\
+\hline
+radU & Uranium ground concentration\\
+\hline
+radTC & Total count of gamma radiation\\
+\hline
+elevation & Mean elevation\\
+\hline
+slope1m & Mean of 1 m slope\\
+\hline
+TPI1m & Mean of 1 m topographic position index\\
+\hline
+TRI1m & Mean of 1 m terrain ruggedness index\\
+\hline
+roughness1m & Mean of 1 m roughness\\
+\hline
+slope10m & 10 m slope\\
+\hline
+TPI10m & 10 m topographic position index\\
+\hline
+TRI10m & 10 m terrain ruggedness index\\
+\hline
+roughness10m & 10 m roughness\\
+\hline
+MRVBF & Multi-resolution valley bottom flatness\\
+\hline
+TWI5m & Mean of 5 m topographic wetness index\\
+\hline
+TWI10m & 10 m topographic wetness index\\
+\hline
+TWI20m & Bilinear interpolation of 20 m topographic wetness index\\
+\hline
+TWI50m & Bilinear interpolation of 50 m topographic wetness index\\
+\hline
+DTW2500 & Depth-to-water index, flow initiation area of 2500 m\textasciicircum{}2\\
+\hline
+DTW5000 & Depth-to-water index, flow initiation area of 5000 m\textasciicircum{}2\\
+\hline
+DTW10000 & Depth-to-water index, flow initiation area of 10000 m\textasciicircum{}2\\
+\hline
+DTW20000 & Depth-to-water index, flow initiation area of 20000 m\textasciicircum{}2\\
+\hline
+DTW40000 & Depth-to-water index, flow initiation area of 40000 m\textasciicircum{}2\\
+\hline
+DTW80000 & Depth-to-water index, flow initiation area of 80000 m\textasciicircum{}2\\
+\hline
+DTW160000 & Depth-to-water index, flow initiation area of 160000 m\textasciicircum{}2\\
+\hline
+\end{tabular}
+\end{table}
 

### 2024-10-03 ms/tables.R

@@ -1,0 +1,38 @@
+library(tidyverse)
+
+# Predictors table ####
+
+rast <- terra::rast("output/predictors.tif")
+names(rast)
+tribble(
+  ~name, ~description,
+  "radK", "Potassium ground concentration",        
+  "radTh", "Thorium ground concentration",
+  "radU" , "Uranium ground concentration",        
+  "radTC", "Total count of gamma radiation",        
+  "elevation", "Mean elevation",
+  "slope1m", "Mean of 1 m slope",      
+  "TPI1m", "Mean of 1 m topographic position index",
+  "TRI1m", "Mean of 1 m terrain ruggedness index",        
+  "roughness1m", "Mean of 1 m roughness",  
+  "slope10m", "10 m slope",    
+  "TPI10m", "10 m topographic position index",
+  "TRI10m", "10 m terrain ruggedness index",
+  "roughness10m" , "10 m roughness",
+  "MRVBF", "Multi-resolution valley bottom flatness",
+  "TWI5m", "Mean of 5 m topographic wetness index",
+  "TWI10m", "10 m topographic wetness index",
+  "TWI20m", "Bilinear interpolation of 20 m topographic wetness index",
+  "TWI50m", "Bilinear interpolation of 50 m topographic wetness index",
+  "DTW2500", "Depth-to-water index, flow initiation area of 2500 m^2",
+  "DTW5000", "Depth-to-water index, flow initiation area of 5000 m^2",
+  "DTW10000", "Depth-to-water index, flow initiation area of 10000 m^2",
+  "DTW20000", "Depth-to-water index, flow initiation area of 20000 m^2",
+  "DTW40000", "Depth-to-water index, flow initiation area of 40000 m^2",
+  "DTW80000", "Depth-to-water index, flow initiation area of 80000 m^2",
+  "DTW160000", "Depth-to-water index, flow initiation area of 160000 m^2") |> 
+  write_csv("ms/tables/predictors.csv")
+
+# sessionInfo ####
+
+sessioninfo::session_info()

### 2024-10-04 R/exploratoryDataAnalysis.R
remove minor bug

@@ -178,12 +178,13 @@ df <- st_read("output/modeling.gpkg", "dataframe") |>
   st_drop_geometry() |> 
   as_tibble()
 metadata <- c("sourceProbe", "sourceGPR", "sourceWisen2021", "sourceMyrarkivet")
-any(!complete.cases(df))
+auxiliary <- c("ar5cover", "ar5soil", "dmkdepth")
+sum(!complete.cases(df))
 
-select(df, !any_of(metadata)) |> 
+select(df, !any_of(c(metadata, auxiliary))) |> 
   slice_sample(n = 1000) |> 
   pairs(pch='.')
-corrplot::corrplot(cor(select(df, !any_of(metadata)) ), 
+corrplot::corrplot(cor(select(df, !any_of(c(metadata, auxiliary))) ), 
                    method = 'ellipse', diag = F)
 
 dflong <- df |> 

### 2024-10-04 ms/ms.Rmd
ms: simple terrain preds

@@
 We chose for our sampling and modelling a spatial resolution of \unit{10\,m}.
-We considered this a reasonable compromise between digital terrain model resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.
+We considered this a reasonable compromise between digital terrain model (DTM) resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.
 
@@
 This function defines the environmental space as a two-dimensional convex hull around the ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest [@hattabUnifiedFrameworkModel2017].
-Elevation was extracted from the \unit{10\,m} national digital terrain model, slope calculated in degrees, and potassium ground concentration downscaled with bilinear resampling.
+Elevation was extracted from the \unit{10\,m} national DTM, slope calculated in degrees, and potassium ground concentration downscaled with bilinear resampling.
 We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and with these parameters *eSample* returned 105 raster cells.
@@
 Feature space coverage sampling works best when all dimensions are important predictors of the outcome [@wadouxSamplingDesignOptimization2019], and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
+The radiometric variables were downscaled to \unit{10\,m} resolution with cubic B-spline resampling in QGIS [v.3, @QGISsoftware].
 We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
@@
 We created the same suite of 25 quantitative peat depth predictors for both sites (Table \@ref(tab:preds)).
-All predictors were derived either from an airborne radiometric survey or from a digital terrain model.
+All quantitative predictors were derived either from an airborne radiometric survey or from a DTM.
 From the radiometric surveys we simply used the four variables produced by the surveyors (Geological Survey of Norway): ground concentration of Potassium, Thorium, Uranium, as well as total count.
-From the digital terrain models we calculated several simple terrain variables and a series of more complex geomorphometric and hydrological variables.
-
+From the DTMs we calculated several land surface parameters, ranging from simple terrain indices to more complex geomorphometric and hydrological variables [@maxwellLandsurfaceParametersSpatial2022].
 ```{r preds, echo=FALSE, message=FALSE}
@@
 Further details about the survey and data processing are provided in Baranwal et al. [-@baranwalHelicopterborneMagneticElectromagnetic2013].
+We downscaled the processed data to \unit{10\,m} resolution by cubic spline resampling, using the *terra* package in R.
 
@@
 Spectrometer count rates were calibrated in 2013 to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
-The spectrometer data were processed following the same procedure as for the survey at Skrimfjella, except that a convolution filter was added to smoothen the gridded data.
+The spectrometer data were processed following the same procedure as for the survey at Skrimfjella, except that a convolution filter was added to smooth the gridded data.
 Further details about the survey and data processing are provided in Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015].
+The \unit{10\,m} resolution predictors from this survey were identical to the layers used in the sampling design (resampled from \unit{50\,m} resolution with cubic B-splines).
+
+For terrain-derived variables, we obtained \unit{1\,m} resolution DTMs from the Norwegian Mapping Authority.
+The DTM for Skrimfjella was produced from airborne laser scanning surveys in 2015 and 2022, with laser point density of \unit{5\,pts\,m^{-2}}.
+For Ørskogfjellet, the DTM was produced from a 2015 survey with \unit{2\,pts\,m^{-2}}.
+Where necessary, DTMs were resampled to the coordinate reference system of the radiometric data.
+We used the *terra* R package to calculate: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
+These variables were derived at two scales to produce different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
+This kind of multiscale feature engineering of land surface parameters has been found to greatly improve machine learning predictions of soil properties [@millerImpactMultiscalePredictor2015; @newmanAssessingSpatiallyHeterogeneous2023].
+We chose \unit{1\,m} and \unit{10\,m} resolutions because we know that peat depth tends to vary at fine scales in Norway [@maxwellLandsurfaceParametersSpatial2022].
+
+We also explored one categorical predictor --- peat depth class --- derived from historical map data.
 

### 2024-10-04 ms/ms.tex
ms: simple terrain preds

@@
 This function defines the environmental space as a two-dimensional convex hull around the ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest \citep{hattabUnifiedFrameworkModel2017}.
-Elevation was extracted from the \unit{10\,m} national digital terrain model, slope calculated in degrees, and potassium ground concentration downscaled with bilinear resampling.
+Elevation was extracted from the \unit{10\,m} national DTM, slope calculated in degrees, and potassium ground concentration downscaled with bilinear resampling.
 We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and with these parameters \emph{eSample} returned 105 raster cells.
@@
 Feature space coverage sampling works best when all dimensions are important predictors of the outcome \citep{wadouxSamplingDesignOptimization2019}, and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
+The radiometric variables were downscaled to \unit{10\,m} resolution with cubic B-spline resampling in QGIS \citep[v.3,][]{QGISsoftware}.
 We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
@@
 We created the same suite of 25 quantitative peat depth predictors for both sites (Table \ref{tab:preds}).
-All predictors were derived either from an airborne radiometric survey or from a digital terrain model.
+All quantitative predictors were derived either from an airborne radiometric survey or from a DTM.
 From the radiometric surveys we simply used the four variables produced by the surveyors (Geological Survey of Norway): ground concentration of Potassium, Thorium, Uranium, as well as total count.
-From the digital terrain models we calculated several simple terrain variables and a series of more complex geomorphometric and hydrological variables.
+From the DTMs we calculated several land surface parameters, ranging from simple terrain indices to more complex geomorphometric and hydrological variables \citep{maxwellLandsurfaceParametersSpatial2022}.
 
@@
 Further details about the survey and data processing are provided in Baranwal et al. \citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013}.
+We downscaled the processed data to \unit{10\,m} resolution by cubic spline resampling, using the \emph{terra} package in R.
 
@@
 Spectrometer count rates were calibrated in 2013 to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
-The spectrometer data were processed following the same procedure as for the survey at Skrimfjella, except that a convolution filter was added to smoothen the gridded data.
+The spectrometer data were processed following the same procedure as for the survey at Skrimfjella, except that a convolution filter was added to smooth the gridded data.
 Further details about the survey and data processing are provided in Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.
+The \unit{10\,m} resolution predictors from this survey were identical to the layers used in the sampling design (resampled from \unit{50\,m} resolution with cubic B-splines).
+
+For terrain-derived variables, we obtained \unit{1\,m} resolution DTMs from the Norwegian Mapping Authority.
+The DTM for Skrimfjella was produced from airborne laser scanning surveys in 2015 and 2022, with laser point density of \unit{5\,pts\,m^{-2}}.
+For Ørskogfjellet, the DTM was produced from a 2015 survey with \unit{2\,pts\,m^{-2}}.
+Where necessary, DTMs were resampled to the coordinate reference system of the radiometric data.
+We used the \emph{terra} R package to calculate: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
+These variables were derived at two scales to produce different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
+This kind of multiscale feature engineering of land surface parameters has been found to greatly improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, newmanAssessingSpatiallyHeterogeneous2023}.
+We chose \unit{1\,m} and \unit{10\,m} resolutions because we know that peat depth tends to vary at fine scales in Norway \citep{maxwellLandsurfaceParametersSpatial2022}.
+
+We also explored one categorical predictor --- peat depth class --- derived from historical map data.
 

### 2024-10-04 R/Skrim/createPredictorStack.R

@@ -26,6 +26,18 @@ rad10m <- rad |>
 plot(rad10m)
 
 # Simple terrain ####
+
+# Lidar point density of DTM
+sa <- st_read("data/Skrim/Skrim-site.gpkg", "fieldsite_outline_utm") |> 
+  st_transform(crssite)
+metadata <- st_read("data/Skrim/eksport_926374_20240909/dtm1/metadata/dtm1_Metadata.shp") |> 
+  select(Pkttetthet, Aarstall)
+st_intersection(sa, metadata) |> 
+  mutate(area = units::drop_units(st_area(geom)),
+         areaprop = area/sum(area)) |> 
+  st_drop_geometry()
+
+# Inspect data
 plot(dtm1m)
 origin(dtm1m)
 

### 2024-10-04 R/createMasks.R

@@ -19,6 +19,16 @@ st_area(sa) |>
 st_write(sa, "data/Orskogfjellet-site.gpkg", "mask_studyarea", 
          append = FALSE)
 
+# Study area land ####
+
+water <- ar5 |> 
+  filter(arealtype == 81 | arealtype == 82) |> 
+  st_union() |> 
+  st_transform(st_crs(RAD_10m_mask))
+st_difference(sa, water) |> 
+  st_write("data/Orskogfjellet-site.gpkg", "mask_studyarealand", 
+           append = FALSE)
+
 # AR5 myr ####
 
 ar5 <- st_transform(ar5, st_crs(RAD_10m_mask))

### 2024-10-04 R/createPredictorStack.R

@@ -27,6 +27,14 @@ probecells <- extract(radK10m, probe, cells = TRUE, xy=TRUE, ID=FALSE) |>
   st_as_sf(coords = c('x','y'), crs=crs(radK10m))
 
 # Simple terrain ####
+
+# Lidar point density of DTM
+saland <- st_read("data/Orskogfjellet-site.gpkg", "mask_studyarealand")
+laserdensity <- rast("data/Haram Skodje Ørskog Vestnes 2pkt 2015/metadata/Haram Skodje Ørskog Vestnes 2pkt 2015_Punkttetthet_DTM.tif")
+extract(laserdensity, saland, ID=FALSE) |> 
+  summary()
+
+# Inspect data
 dtm1m <- rast("data/DTMmerged.tif")
 plot(dtm1m)
 origin(dtm1m)

### 2024-10-04 ms/ms.Rmd
ms: MRVBF and TWI preds

@@
 Along the GPR transects we also probed peat depth at marked trace locations, to be able to calibrate the GPR wave speed velocity.
-We processed the GPR data with Reflex2DQuick software (v.3.0; Sandmeier Scientific Software, Germany), applying a time-zero correction, a dewow filter, and a gain filter based on observed energy decay.
+We processed the GPR data with Reflex2DQuick software (v.3.0, Sandmeier Scientific Software, Germany), applying a time-zero correction, a dewow filter, and a gain filter based on observed energy decay.
 Then we picked strong reflectors in the radargrams that we interpreted as the base of the peat column.
@@
 
-We processed the GPR data with Reflexw software (v.8.5; Sandmeier Scientific Software, Germany), applying a dewow filter, time-zero correction, bandpass filter, gain filter, and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennae.
+We processed the GPR data with Reflexw software (v.8.5, Sandmeier Scientific Software, Germany), applying a dewow filter, time-zero correction, bandpass filter, gain filter, and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennae.
 The last correction is important for the rough terrain antenna, which has an antenna separation (\unit{2.2\,m}) --- comparable to typical peat depths.
@@
 Where necessary, DTMs were resampled to the coordinate reference system of the radiometric data.
-We used the *terra* R package to calculate: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
+
+We used the *terra* R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
 These variables were derived at two scales to produce different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
@@
 We chose \unit{1\,m} and \unit{10\,m} resolutions because we know that peat depth tends to vary at fine scales in Norway [@maxwellLandsurfaceParametersSpatial2022].
+We also calculated the the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm [@gallantMultiresolutionIndexValley2003].
+We calculated this index in SAGA GIS [v.9.3.2, @conradSystemAutomatedGeoscientific2015] with default parameters.
+
+Next, we calculated the Topographic Wetness Index [@quinnPredictionHillslopeFlow1991].
+This index is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate-to-coarse resolution DTMs [@agrenEvaluatingDigitalTerrain2014; @riihimakiTopographicWetnessIndex2021], so we calculated it from \unit{5\,m}, \unit{10\,m}, \unit{20\,m}, and \unit{50\,m} DTM resolution.
+The calculations were performed with Whitebox software [@lindsayWhiteboxGATCase2016], accessed through the *whitebox* R package [v2.4, @wuWhiteboxWhiteboxToolsFrontend2022].
+We filled depressions in the DTM with the algorithm in Wang & Liu [-@wangEfficientMethodIdentifying2006], and used the deterministic infinity flow accumulation algorithm [@tarbotonNewMethodDetermination1997].
 

### 2024-10-04 ms/ms.tex
ms: MRVBF and TWI preds

@@
 Where necessary, DTMs were resampled to the coordinate reference system of the radiometric data.
-We used the \emph{terra} R package to calculate: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
+
+We used the \emph{terra} R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
 These variables were derived at two scales to produce different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
@@
 We chose \unit{1\,m} and \unit{10\,m} resolutions because we know that peat depth tends to vary at fine scales in Norway \citep{maxwellLandsurfaceParametersSpatial2022}.
+We also calculated the the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
+We calculated this index in SAGA GIS \citep[v.9.3.2,][]{conradSystemAutomatedGeoscientific2015} with default parameters.
+
+Next, we calculated the Topographic Wetness Index \citep{quinnPredictionHillslopeFlow1991}.
+This index is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate-to-coarse resolution DTMs \citep{agrenEvaluatingDigitalTerrain2014, riihimakiTopographicWetnessIndex2021}, so we calculated it from \unit{5\,m}, \unit{10\,m}, \unit{20\,m}, and \unit{50\,m} DTM resolution.
+The calculations were performed with Whitebox software \citep{lindsayWhiteboxGATCase2016}, accessed through the \emph{whitebox} R package \citep[v2.4,][]{wuWhiteboxWhiteboxToolsFrontend2022}.
+We filled depressions in the DTM with the algorithm in Wang \& Liu \citeyearpar{wangEfficientMethodIdentifying2006}, and used the deterministic infinity flow accumulation algorithm \citep{tarbotonNewMethodDetermination1997}.
 

### 2024-10-07 ms/ms.Rmd
ms: DTW preds

@@
 
+The last terrain-based predictor we included was the depth-to-water index [@murphyMappingWetlandsComparison2007].
+This index approximates a location's vertical height above the surface water feature that it is likely to drain towards.
+It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature [eq. 5 in @murphyTopographicModellingSoil2009].
+We calculated unitless slope from the \unit{1\,m} DTM using the Whitebox software.
+Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell [@schonauerSpatiotemporalPredictionSoil2021; @schonauerRcodeCalculatingDepthtowater2021].
+This catchments area layer was then thresholded at seven different levels (*flow initiation areas* from \unit{0.25\,ha} to \unit{16\,ha}) to estimate surface water features under moisture scenarios varying from wet to dry [@murphyModellingMappingTopographic2011; @agrenEvaluatingDigitalTerrain2014; @schonauerSpatiotemporalPredictionSoil2021].
+In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
+For each of the seven surface water layers, we derived the depth-to-water index using the *Distance Accumulation* tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.
+
 We also explored one categorical predictor --- peat depth class --- derived from historical map data.

### 2024-10-07 ms/ms.tex
ms: DTW preds

@@
 \hline
-DTW2500 & Depth-to-water index, flow initiation area of 2500 m\textasciicircum{}2\\
+DTW2500 & Depth-to-water index, flow initiation area of 0.25 ha\\
 \hline
-DTW5000 & Depth-to-water index, flow initiation area of 5000 m\textasciicircum{}2\\
+DTW5000 & Depth-to-water index, flow initiation area of 0.5 ha\\
 \hline
-DTW10000 & Depth-to-water index, flow initiation area of 10000 m\textasciicircum{}2\\
+DTW10000 & Depth-to-water index, flow initiation area of 1 ha\\
 \hline
-DTW20000 & Depth-to-water index, flow initiation area of 20000 m\textasciicircum{}2\\
+DTW20000 & Depth-to-water index, flow initiation area of 2 ha\\
 \hline
-DTW40000 & Depth-to-water index, flow initiation area of 40000 m\textasciicircum{}2\\
+DTW40000 & Depth-to-water index, flow initiation area of 4 ha\\
 \hline
-DTW80000 & Depth-to-water index, flow initiation area of 80000 m\textasciicircum{}2\\
+DTW80000 & Depth-to-water index, flow initiation area of 8 ha\\
 \hline
-DTW160000 & Depth-to-water index, flow initiation area of 160000 m\textasciicircum{}2\\
+DTW160000 & Depth-to-water index, flow initiation area of 16 ha\\
 \hline
@@
 We filled depressions in the DTM with the algorithm in Wang \& Liu \citeyearpar{wangEfficientMethodIdentifying2006}, and used the deterministic infinity flow accumulation algorithm \citep{tarbotonNewMethodDetermination1997}.
+
+The last terrain-based predictor we included was the depth-to-water index \citep{murphyMappingWetlandsComparison2007}.
+This index approximates a location's vertical height above the surface water feature that it is likely to drain towards.
+It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature \citep[eq. 5 in][]{murphyTopographicModellingSoil2009}.
+We calculated unitless slope from the \unit{1\,m} DTM using the Whitebox software.
+Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell \citep{schonauerSpatiotemporalPredictionSoil2021, schonauerRcodeCalculatingDepthtowater2021}.
+This catchments area layer was then thresholded at seven different levels (\emph{flow initiation areas} from \unit{0.25\,ha} to \unit{16\,ha}) to estimate surface water features under moisture scenarios varying from wet to dry \citep{murphyModellingMappingTopographic2011, agrenEvaluatingDigitalTerrain2014, schonauerSpatiotemporalPredictionSoil2021}.
+In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
+For each of the seven surface water layers, we derived the depth-to-water index using the \emph{Distance Accumulation} tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.
 

### 2024-10-07 ms/tables.R

@@ -24,13 +24,13 @@ tribble(
   "TWI10m", "10 m topographic wetness index",
   "TWI20m", "Bilinear interpolation of 20 m topographic wetness index",
   "TWI50m", "Bilinear interpolation of 50 m topographic wetness index",
-  "DTW2500", "Depth-to-water index, flow initiation area of 2500 m^2",
-  "DTW5000", "Depth-to-water index, flow initiation area of 5000 m^2",
-  "DTW10000", "Depth-to-water index, flow initiation area of 10000 m^2",
-  "DTW20000", "Depth-to-water index, flow initiation area of 20000 m^2",
-  "DTW40000", "Depth-to-water index, flow initiation area of 40000 m^2",
-  "DTW80000", "Depth-to-water index, flow initiation area of 80000 m^2",
-  "DTW160000", "Depth-to-water index, flow initiation area of 160000 m^2") |> 
+  "DTW2500", "Depth-to-water index, flow initiation area of 0.25 ha",
+  "DTW5000", "Depth-to-water index, flow initiation area of 0.5 ha",
+  "DTW10000", "Depth-to-water index, flow initiation area of 1 ha",
+  "DTW20000", "Depth-to-water index, flow initiation area of 2 ha",
+  "DTW40000", "Depth-to-water index, flow initiation area of 4 ha",
+  "DTW80000", "Depth-to-water index, flow initiation area of 8 ha",
+  "DTW160000", "Depth-to-water index, flow initiation area of 16 ha") |> 
   write_csv("ms/tables/predictors.csv")
 
 # sessionInfo ####

### 2024-10-07 ms/ms.Rmd
ms: DMK predictor

@@
 
-We also explored one categorical predictor --- peat depth class --- derived from historical map data.
+In addition to the quantitative predictors described above, we prepared one categorical predictor --- peat depth class --- from a historical national map dataset called *DMK* [@ahlstromAR5Klassifikasjonssystem2019].
+The DMK peat depth classes are: < \unit{1\,m} (*shallow*), > \unit{1\,m} (*deep*), and *unknown*.
+This dataset stems from field measurements made in 1964--2007 as part of a wider land cover mapping in Norway.
+The mapping was primarily for identifying agricultural and silvicultural resources, so it covers productive areas below the tree line and has greater coverage of peat depth class in peatlands judged to be arable or afforestable [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007; @ahlstromAR5Klassifikasjonssystem2019].
+Mappers generally assigned peat depth classes to polygons of at least \unit{0.5\,ha}, although delineating polygons down to \unit{0.2\,ha} was allowed if peat depth showed a "particularly marked difference" [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
+We rasterized the peat depth class attribute to our \unit{10\,m} grid.
 

### 2024-10-07 ms/ms.tex
ms: DMK predictor

@@
 
-We also explored one categorical predictor --- peat depth class --- derived from historical map data.
+In addition to the quantitative predictors described above, we prepared one categorical predictor --- peat depth class --- from a historical national map dataset called \emph{DMK} \citep{ahlstromAR5Klassifikasjonssystem2019}.
+The DMK peat depth classes are: \textless{} \unit{1\,m} (\emph{shallow}), \textgreater{} \unit{1\,m} (\emph{deep}), and \emph{unknown}.
+This dataset stems from field measurements made in 1964--2007 as part of a wider land cover mapping in Norway.
+The mapping was primarily for identifying agricultural and silvicultural resources, so it covers productive areas below the tree line and has greater coverage of peat depth class in peatlands judged to be arable or afforestable \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007, ahlstromAR5Klassifikasjonssystem2019}.
+Mappers generally assigned peat depth classes to polygons of at least \unit{0.5\,ha}, although delineating polygons down to \unit{0.2\,ha} was allowed if peat depth showed a ``particularly marked difference'' \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
+We rasterized the peat depth class attribute to our \unit{10\,m} grid.
 

### 2024-10-07 ms/ms.Rmd
ms: minor edits

@@
 At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest model of peat depth, and we designed our sampling with this in mind [@brusSamplingDigitalSoil2019].
-Broadly, we aimed for a sample that was representative the predictor space of the most important predictors of peat depth [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
+Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
 A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones [@brusSamplingDigitalSoil2019].
@@
 These variables were derived at two scales to produce different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
-This kind of multiscale feature engineering of land surface parameters has been found to greatly improve machine learning predictions of soil properties [@millerImpactMultiscalePredictor2015; @newmanAssessingSpatiallyHeterogeneous2023].
+This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties [@millerImpactMultiscalePredictor2015; @dornikOptimalScalingPredictors2022; @newmanAssessingSpatiallyHeterogeneous2023].
 We chose \unit{1\,m} and \unit{10\,m} resolutions because we know that peat depth tends to vary at fine scales in Norway [@maxwellLandsurfaceParametersSpatial2022].

### 2024-10-07 ms/ms.Rmd
ms: models rf + rk

@@
 
+We used Random Forests (RF) to predict peat depth at both sites.
+RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees [@breimanRandomForests2001].
+We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in digital soil mapping applications than other algorithms [@beguinPredictingSoilProperties2017; @nussbaumEvaluationDigitalSoil2018; @lamichhaneDigitalSoilMapping2019].
+It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance [@khaledianSelectingAppropriateMachine2020].
+Evaluating variable importance in a maximally-predictive model aligns with the aim of this study.
+
+RF by itself is not a spatial model, and it will only predict spatial structure in the outcome to the degree that the structure is captured by predictors.
+We considered using regression kriging, a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals [@henglGenericFrameworkSpatial2004].
+The spatial component in regression kriging often improves map accuracy compared to a non-spatial model [@beguinPredictingSoilProperties2017; @lamichhaneDigitalSoilMapping2019; @mollaMachineLearningGeostatistical2023], but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations [@henglGenericFrameworkSpatial2004].
+If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.
+We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that simple RF rather regression kriging was justified at both sites.
+
 # Results

### 2024-10-07 ms/ms.tex
ms: models rf + rk

@@
 At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest model of peat depth, and we designed our sampling with this in mind \citep{brusSamplingDigitalSoil2019}.
-Broadly, we aimed for a sample that was representative the predictor space of the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
+Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
 A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones \citep{brusSamplingDigitalSoil2019}.
@@
 These variables were derived at two scales to produce different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
-This kind of multiscale feature engineering of land surface parameters has been found to greatly improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, newmanAssessingSpatiallyHeterogeneous2023}.
+This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, dornikOptimalScalingPredictors2022, newmanAssessingSpatiallyHeterogeneous2023}.
 We chose \unit{1\,m} and \unit{10\,m} resolutions because we know that peat depth tends to vary at fine scales in Norway \citep{maxwellLandsurfaceParametersSpatial2022}.
@@
 \subsection{Predictive models of peat depth}
+
+We used Random Forests (RF) to predict peat depth at both sites.
+RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees \citep{breimanRandomForests2001}.
+We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in digital soil mapping applications than other algorithms \citep{beguinPredictingSoilProperties2017, nussbaumEvaluationDigitalSoil2018, lamichhaneDigitalSoilMapping2019}.
+It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance \citep{khaledianSelectingAppropriateMachine2020}.
+Evaluating variable importance in a maximally-predictive model aligns with the aim of this study.
+
+RF by itself is not a spatial model, and it will only predict spatial structure in the outcome to the degree that the structure is captured by predictors.
+We considered using regression kriging, a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals \citep{henglGenericFrameworkSpatial2004}.
+The spatial component in regression kriging often improves map accuracy compared to a non-spatial model \citep{beguinPredictingSoilProperties2017, lamichhaneDigitalSoilMapping2019, mollaMachineLearningGeostatistical2023}, but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations \citep{henglGenericFrameworkSpatial2004}.
+If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.
+We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that simple RF rather regression kriging was justified at both sites.
 

### 2024-10-08 ms/ms.Rmd
ms: minor edits - added refs

@@
 We considered using regression kriging, a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals [@henglGenericFrameworkSpatial2004].
-The spatial component in regression kriging often improves map accuracy compared to a non-spatial model [@beguinPredictingSoilProperties2017; @lamichhaneDigitalSoilMapping2019; @mollaMachineLearningGeostatistical2023], but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations [@henglGenericFrameworkSpatial2004].
+The spatial component in regression kriging often improves map accuracy compared to a non-spatial model [@beguinPredictingSoilProperties2017; @lamichhaneDigitalSoilMapping2019; @mollaMachineLearningGeostatistical2023], but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations [@henglGenericFrameworkSpatial2004; @szaboMappingSoilHydraulic2019; @takoutsingComparingPredictionPerformance2022].
 If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.

### 2024-10-08 ms/ms.Rmd
ms: high-level modelling approach

@@
 
+For each site we followed the same modelling workflow, so the description below applies to Skrimfjella and Ørskogfjellet.
+First we trained a RF to predict peat depth based on our 25 quantitative predictors, to simulate applications where radiometric and terrain data are available, but no maps of peat depth.
+Next we trained a RF with all 26 predictors (DMK peat depth class included), to fully leverage data available at these sites and check the added value of the national DMK dataset.
+Finally, we fit a simple linear model to only DMK depth class, to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth calibrated on the same data.
+We implemented models in the *tidymodels* framework in R [@kuhnTidymodelsCollectionPackages2020], with the *ranger* R package for RFs [v.0.16, @wrightRangerFastImplementation2017].
+RFs were fit with 1000 trees, minimum node size of 5, and the number of predictors randomly sampled at each split was the square root of the total number of predictors (*ranger* default).
+
+In the absence of additional field work to collect a model-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance.
+
 # Results

### 2024-10-10 ms/ms.Rmd
ms: kNNDM

@@
 
-In the absence of additional field work to collect a model-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance.
+In the absence of additional field work to collect a design-based independent validation set, we used a spatially stratified cross-validation scheme to evaluate model performance [@wadouxSpatialCrossvalidationNot2021; @meyerMachineLearningbasedGlobal2022].
+Specifically, we used k-Means Nearest Neighbor Distance Matching (kNNDM), which creates cross-validation folds that mimic the spatial prediction task that is defined as the goal [@linnenbrinkKNNDMCVKfold2024].
+In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
+That way, the spatial separation between folds is similar to the separation between training and prediction locations --- which increases the quality of the map accuracy estimate [@linnenbrinkKNNDMCVKfold2024].
+For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatially stratified cross-validation [@wadouxSpatialCrossvalidationNot2021].
+We implemented the kNNDM with the *CAST* R package [v.1.0.2, @meyerCASTPackageTraining2024], setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
+In general, the performance of digital soil mapping must be evaluated with reference to a specific purpose [i.e., map vs.
+model validation, interpolation vs.
+extrapolation, @robertsCrossvalidationStrategiesData2017; @milaNearestNeighbourDistance2022], and using kNNDM with a defined prediction domain increases the likelihood of quality accuracy estimates for a particular map.
 

### 2024-10-10 ms/ms.tex
ms: kNNDM

@@
 We considered using regression kriging, a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals \citep{henglGenericFrameworkSpatial2004}.
-The spatial component in regression kriging often improves map accuracy compared to a non-spatial model \citep{beguinPredictingSoilProperties2017, lamichhaneDigitalSoilMapping2019, mollaMachineLearningGeostatistical2023}, but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations \citep{henglGenericFrameworkSpatial2004}.
+The spatial component in regression kriging often improves map accuracy compared to a non-spatial model \citep{beguinPredictingSoilProperties2017, lamichhaneDigitalSoilMapping2019, mollaMachineLearningGeostatistical2023}, but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations \citep{henglGenericFrameworkSpatial2004, szaboMappingSoilHydraulic2019, takoutsingComparingPredictionPerformance2022}.
 If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.
 We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that simple RF rather regression kriging was justified at both sites.
+
+For each site we followed the same modelling workflow, so the description below applies to Skrimfjella and Ørskogfjellet.
+First we trained a RF to predict peat depth based on our 25 quantitative predictors, to simulate applications where radiometric and terrain data are available, but no maps of peat depth.
+Next we trained a RF with all 26 predictors (DMK peat depth class included), to fully leverage data available at these sites and check the added value of the national DMK dataset.
+Finally, we fit a simple linear model to only DMK depth class, to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth calibrated on the same data.
+We implemented models in the \emph{tidymodels} framework in R \citep{kuhnTidymodelsCollectionPackages2020}, with the \emph{ranger} R package for RFs \citep[v.0.16,][]{wrightRangerFastImplementation2017}.
+RFs were fit with 1000 trees, minimum node size of 5, and the number of predictors randomly sampled at each split was the square root of the total number of predictors (\emph{ranger} default).
+
+In the absence of additional field work to collect a design-based independent validation set, we used a spatially stratified cross-validation scheme to evaluate model performance \citep{wadouxSpatialCrossvalidationNot2021, meyerMachineLearningbasedGlobal2022}.
+Specifically, we used k-Means Nearest Neighbor Distance Matching (kNNDM), which creates cross-validation folds that mimic the spatial prediction task that is defined as the goal \citep{linnenbrinkKNNDMCVKfold2024}.
+In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
+That way, the spatial separation between folds is similar to the separation between training and prediction locations --- which increases the quality of the map accuracy estimate \citep{linnenbrinkKNNDMCVKfold2024}.
+For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatially stratified cross-validation \citep{wadouxSpatialCrossvalidationNot2021}.
+We implemented the kNNDM with the \emph{CAST} R package \citep[v.1.0.2,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
+In general, the performance of digital soil mapping must be evaluated with reference to a specific purpose \citep[i.e., map vs.~model validation, interpolation vs.~extrapolation,][]{robertsCrossvalidationStrategiesData2017, milaNearestNeighbourDistance2022}, and using kNNDM with a defined prediction domain increases the likelihood of quality accuracy estimates for a particular map.
 

### 2024-10-15 ms/ms.tex
ms: variable importance methods

@@
 We calculated unitless slope from the \unit{1\,m} DTM using the Whitebox software.
-Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell \citep{schonauerSpatiotemporalPredictionSoil2021, schonauerRcodeCalculatingDepthtowater2021}.
+Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell \citep{schonauerSpatiotemporalPredictionSoil2021, schonauerRcodeCalculatingDepthwater2021}.
 This catchments area layer was then thresholded at seven different levels (\emph{flow initiation areas} from \unit{0.25\,ha} to \unit{16\,ha}) to estimate surface water features under moisture scenarios varying from wet to dry \citep{murphyModellingMappingTopographic2011, agrenEvaluatingDigitalTerrain2014, schonauerSpatiotemporalPredictionSoil2021}.
@@
 In general, the performance of digital soil mapping must be evaluated with reference to a specific purpose \citep[i.e., map vs.~model validation, interpolation vs.~extrapolation,][]{robertsCrossvalidationStrategiesData2017, milaNearestNeighbourDistance2022}, and using kNNDM with a defined prediction domain increases the likelihood of quality accuracy estimates for a particular map.
+
+We quantified global variable importance for the RFs with the 25 quantitative predictors.
+Global variable importance measures the influence of a given predictor on the output of the model, aggregated across all locations.
+We calculated variable importance with the \emph{vip} R package (v.0.4.1), by three different methods: \emph{FIRM}, \emph{permutation}, and \emph{Shapley} \citep{greenwellVariableImportancePlots2020}.
+\emph{Permutation} values were aggregated from ten iterations with RMSE as the performance measure.
 

### 2024-10-15 R/Skrim/modelDepth.R

@@ -52,33 +52,52 @@ fit_remotesensing <-
 # Variable importance ####
 
 fit_remotesensing
-import_perm <- fit_remotesensing %>% 
+import_perm_ranger <- fit_remotesensing %>% 
   extract_fit_parsnip() %>% 
-  vip::vi()
-vip::vip(import_perm, num_features = 25)
-
-ftnames <- fit_remotesensing |> 
-  extract_recipe() |> 
-  summary() |> 
-  filter(role == "predictor") |> 
-  pull(variable)
-import_firm <- fit_remotesensing %>% 
-  extract_fit_parsnip() |> 
-  vip::vi(method = "firm", feature_names = ftnames, 
-          train = st_drop_geometry(frame))
-vip::vip(import_firm, num_features = 25)
+  vip::vi(scale = TRUE)
+vip::vip(import_perm_ranger, num_features = 25)
 
 pfun <- function(object, newdata) {  # needs to return a numeric vector
   predict(object, new_data = newdata)$.pred  
 }
 # pfun(fit_remotesensing, frame)
+ftnames <- fit_remotesensing |> 
+  extract_recipe() |> 
+  summary() |> 
+  filter(role == "predictor") |> 
+  pull(variable)
+
+set.seed(403)  
+import_perm_vip <- fit_remotesensing %>% 
+  extract_fit_parsnip() |> 
+  vip::vi(method = "permute", feature_names = ftnames, 
+          train = st_drop_geometry(frame), 
+          target = "depth_cm", metric = "rmse",
+          pred_wrapper = pfun, nsim = 10, scale = TRUE)
+vip::vip(import_perm_vip, geom = "boxplot")
+vip::vip(import_perm_vip, num_features = 25)
+
+import_firm <- fit_remotesensing %>% 
+  extract_fit_parsnip() |> 
+  vip::vi(method = "firm", ice = TRUE, feature_names = ftnames, 
+          train = st_drop_geometry(frame), scale = TRUE)
+vip::vip(import_firm, num_features = 25)
+
 import_shap <- fit_remotesensing %>% 
   extract_fit_parsnip() |> 
   vip::vi(method = "shap", 
           pred_wrapper = pfun,
           feature_names = ftnames, 
-          train = st_drop_geometry(frame))
+          train = st_drop_geometry(frame),
+          scale = TRUE)
 vip::vip(import_shap, num_features = 25)
+
[…]

### 2024-10-15 R/modelDepth.R

@@ -52,33 +52,52 @@ fit_remotesensing <-
 # Variable importance ####
 
 fit_remotesensing
-import_perm <- fit_remotesensing %>% 
+import_perm_ranger <- fit_remotesensing %>% 
   extract_fit_parsnip() %>% 
-  vip::vi()
-vip::vip(import_perm, num_features = 25)
-
-ftnames <- fit_remotesensing |> 
-  extract_recipe() |> 
-  summary() |> 
-  filter(role == "predictor") |> 
-  pull(variable)
-import_firm <- fit_remotesensing %>% 
-  extract_fit_parsnip() |> 
-  vip::vi(method = "firm", feature_names = ftnames, 
-          train = st_drop_geometry(frame))
-vip::vip(import_firm, num_features = 25)
+  vip::vi(scale = TRUE)
+vip::vip(import_perm_ranger, num_features = 25)
 
 pfun <- function(object, newdata) {  # needs to return a numeric vector
   predict(object, new_data = newdata)$.pred  
 }
 # pfun(fit_remotesensing, frame)
+ftnames <- fit_remotesensing |> 
+  extract_recipe() |> 
+  summary() |> 
+  filter(role == "predictor") |> 
+  pull(variable)
+
+set.seed(403)  
+import_perm_vip <- fit_remotesensing %>% 
+  extract_fit_parsnip() |> 
+  vip::vi(method = "permute", feature_names = ftnames, 
+          train = st_drop_geometry(frame), 
+          target = "depth_cm", metric = "rmse",
+          pred_wrapper = pfun, nsim = 10, scale = TRUE)
+vip::vip(import_perm_vip, geom = "boxplot")
+vip::vip(import_perm_vip, num_features = 25)
+
+import_firm <- fit_remotesensing %>% 
+  extract_fit_parsnip() |> 
+  vip::vi(method = "firm", ice = TRUE, feature_names = ftnames, 
+          train = st_drop_geometry(frame), scale = TRUE)
+vip::vip(import_firm, num_features = 25)
+
 import_shap <- fit_remotesensing %>% 
   extract_fit_parsnip() |> 
   vip::vi(method = "shap", 
           pred_wrapper = pfun,
           feature_names = ftnames, 
-          train = st_drop_geometry(frame))
+          train = st_drop_geometry(frame),
+          scale = TRUE)
 vip::vip(import_shap, num_features = 25)
+
+bind_rows(perm.ranger = import_perm_ranger, 
[…]

### 2024-10-15 ms/figures.R

@@ -190,6 +190,53 @@ ggsave(filename = 'sites-patchwork.pdf', path = "ms/figures",
 #                               type='https://opencache.statkart.no/gatekeeper/gk/gk.open_gmaps?layers=topo4&zoom={z}&x={x}&y={y}')
 # plot(map)
 
+# Variable importance ####
+
+orskog <- read_csv("output/variable_importance.csv") %>% 
+  filter(type != "perm.ranger")
+skrim <- read_csv("output/Skrim/variable_importance.csv") %>% 
+  filter(type != "perm.ranger")
+
+orskog.rank <- orskog %>% 
+  group_by(Variable) %>% 
+  summarize(Imp.median = median(Importance))
+skrim.rank <- skrim %>% 
+  group_by(Variable) %>% 
+  summarize(Imp.median = median(Importance))
+
+g1 <- left_join(orskog, orskog.rank, by = join_by(Variable)) %>% 
+  ggplot() +
+  geom_point(aes(x = fct_reorder(Variable, Imp.median), 
+                 y = Importance, 
+                 color = type)) +
+  coord_flip() +
+  scale_color_discrete(labels = c(firm = "FIRM", 
+                                  perm.vip = "permutation", 
+                                  shap = "Shapley")) +
+  theme_minimal() +
+  theme(axis.title.y = element_blank(),
+        legend.title = element_blank())
+
+g2 <- left_join(skrim, skrim.rank, by = join_by(Variable)) %>% 
+  ggplot() +
+  geom_point(aes(x = fct_reorder(Variable, Imp.median), 
+                 y = Importance, 
+                 color = type)) +
+  coord_flip() +
+  scale_color_discrete(labels = c(firm = "FIRM", 
+                                  perm.vip = "permutation", 
+                                  shap = "Shapley")) +
+  theme_minimal() +
+  theme(axis.title.y = element_blank(),
+        legend.title = element_blank())
+
+## Combined ####
+
+g1 + g2 + plot_layout(ncol = 1, guides = 'collect', axes = 'collect') +
+  plot_annotation(tag_levels = 'a', tag_prefix = '(', tag_suffix = ')') 
+ggsave(filename = 'variable_importance.pdf', path = "ms/figures",
+       width =(210-30)/1.5, height = (240-40)/1.25, units = 'mm') #copernicus.cls page 210x240
+
 # sessionInfo ####
 
 sessioninfo::session_info()

### 2024-10-16 R/Skrim/modelDepth.R
model metrics results

@@ -49,6 +49,12 @@ fit_remotesensing <-
   workflow_remotesensing %>% 
   fit(frame)
 
+fit_remotesensing %>% 
+  augment(new_data = frame) %>%
+  ggplot(aes(depth_cm, .pred)) +
+  geom_point() +
+  geom_abline()
+
 # Variable importance ####
 
 fit_remotesensing
@@ -101,7 +107,7 @@ bind_rows(perm.ranger = import_perm_ranger,
 
 # Error estimation ####
 
-evaluation_metrics <- metric_set(rmse, mae, rsq)
+evaluation_metrics <- metric_set(rmse, mae, rsq, ccc)
 
 # With kNNDM from CAST into tidymodels workflow
 library(CAST) #v.1.0.2
@@ -150,8 +156,17 @@ fit_remotesensing_knndm <-
   workflow_remotesensing %>% 
   fit_resamples(
     resamples = folds,
-    metrics = evaluation_metrics)
+    metrics = evaluation_metrics,
+    control = control_resamples(save_pred = TRUE))
 collect_metrics(fit_remotesensing_knndm)
+
+preds <- fit_remotesensing_knndm %>% 
+  collect_predictions() |> 
+  select(id, depth_cm, .pred)
+preds %>% 
+  ggplot(aes(depth_cm, .pred)) +
+  geom_point() +
+  geom_abline()
 
 ### DMK-only model ####
 
@@ -193,7 +208,7 @@ frame |>
 ### Leveraging model ####
 
 recipe_leveraging <- 
-  recipe(formula = depth_cm ~ ., data = select(frame, !starts_with("source"))) |> 
+  recipe(formula = depth_cm ~ ., data = select(frame, !starts_with(c("source", "ar5")))) |> 
   remove_role(geom, old_role = "predictor") |> 
   step_unknown(dmkdepth) |> 
   step_dummy(dmkdepth)
@@ -216,6 +231,13 @@ collect_metrics(fit_leveraging_knndm)
 collect_predictions(fit_leveraging_knndm) |> 
   drop_na(.pred) |> 
   nrow()
+
+bind_rows(
+  remotesensing = collect_metrics(fit_remotesensing_knndm),
+  dmkintercept = collect_metrics(fit_dmkintercept_knndm),
+  leveraging = collect_metrics(fit_leveraging_knndm), 
+  .id = "model") |> 
+  readr::write_csv("output/Skrim/modelmetrics.csv")
 
 ## Extrapolating beyond mire ####
 

### 2024-10-16 R/modelDepth.R

@@ -49,6 +49,12 @@ fit_remotesensing <-
   workflow_remotesensing %>% 
   fit(frame)
 
+fit_remotesensing %>% 
+  augment(new_data = frame) %>%
+  ggplot(aes(depth_cm, .pred)) +
+  geom_point() +
+  geom_abline()
+
 # Variable importance ####
 
 fit_remotesensing
@@ -101,7 +107,7 @@ bind_rows(perm.ranger = import_perm_ranger,
 
 # Error estimation ####
 
-evaluation_metrics <- metric_set(rmse, mae, rsq)
+evaluation_metrics <- metric_set(rmse, mae, rsq, ccc)
 
 # With kNNDM from CAST into tidymodels workflow
 library(CAST) #v.1.0.2
@@ -150,8 +156,17 @@ fit_remotesensing_knndm <-
   workflow_remotesensing %>% 
   fit_resamples(
     resamples = folds,
-    metrics = evaluation_metrics)
+    metrics = evaluation_metrics,
+    control = control_resamples(save_pred = TRUE))
 collect_metrics(fit_remotesensing_knndm)
+
+preds <- fit_remotesensing_knndm %>% 
+  collect_predictions() |> 
+  select(id, depth_cm, .pred)
+preds %>% 
+  ggplot(aes(depth_cm, .pred)) +
+  geom_point() +
+  geom_abline()
 
 ### DMK-only model ####
 
@@ -190,7 +205,7 @@ frame |>
 ### Leveraging model ####
 
 recipe_leveraging <- 
-  recipe(formula = depth_cm ~ ., data = select(frame, !starts_with("source"))) |> 
+  recipe(formula = depth_cm ~ ., data = select(frame, !starts_with(c("source", "ar5")))) |> 
   remove_role(geom, old_role = "predictor") |> 
   step_unknown(dmkdepth) |> 
   step_dummy(dmkdepth)
@@ -209,6 +224,13 @@ fit_leveraging_knndm <-
     resamples = folds,
     metrics = evaluation_metrics)
 collect_metrics(fit_leveraging_knndm)
+
+bind_rows(
+  remotesensing = collect_metrics(fit_remotesensing_knndm),
+  dmkintercept = collect_metrics(fit_dmkintercept_knndm),
+  leveraging = collect_metrics(fit_leveraging_knndm), 
+  .id = "model") |> 
+  readr::write_csv("output/modelmetrics.csv")
 
 ## Extrapolating beyond mire ####
 

### 2024-10-16 ms/figures.R

@@ -1,3 +1,5 @@
+# Sites map ####
+
 library(sf)
 library(terra)
 library(tidyterra)
@@ -7,8 +9,6 @@ library(ggtext)
 library(ggrepel)
 library(ggspatial)
 library(rnaturalearth)
-
-# Sites map ####
 
 ## Inset ####
 
@@ -27,7 +27,7 @@ skrim <- st_read("data/Skrim/Skrim-site.gpkg", "fieldsite_outline_utm") |>
 both <- bind_rows(orskog, skrim)
 labels <- both |> 
   st_centroid() |> 
-  mutate(site = c("Ørskogfjellet", "Skrimfjella"))
+  mutate(site = c("\u00D8rskogfjellet", "Skrimfjella"))
 
 g1 <- ggplot() +
   geom_sf(data = nor50, fill = "white") + 
@@ -190,7 +190,58 @@ ggsave(filename = 'sites-patchwork.pdf', path = "ms/figures",
 #                               type='https://opencache.statkart.no/gatekeeper/gk/gk.open_gmaps?layers=topo4&zoom={z}&x={x}&y={y}')
 # plot(map)
 
+# Model metrics ####
+
+library(tidyverse)
+
+orskog <- read_csv("output/modelmetrics.csv") 
+skrim <- read_csv("output/Skrim/modelmetrics.csv")
+plotting <- bind_rows(orskog = orskog, skrim = skrim, .id = 'site') %>% 
+  filter(.metric != 'mae') %>% 
+  mutate(
+    site = fct_relevel(site, "orskog"),
+    .metric = fct_relevel(.metric, "ccc", "rsq", "rmse"),
+    metric = case_when(
+      .metric == "ccc" ~ "Concordance \ncorrelation coefficient",
+      .metric == "rsq" ~ "R-squared",
+      .metric == "rmse" ~ "RMSE"),
+    model = case_when(
+      model == "dmkintercept" ~ "DMK peat depth class (1)",
+      model == "remotesensing" ~ "Quantitative predictors (25)",
+      model == "leveraging" ~ "All predictors (26)"),
+    model = fct_reorder(model, mean)
+    )
+str(plotting)
+
+g1 <- ggplot(plotting) +
+  geom_linerange(aes(y = model, 
+                     x = mean, 
+                     xmin = mean - std_err, 
+                     xmax = mean + std_err, 
+                     color = site),
+                 position = position_dodge2(width= 0.2, reverse = TRUE)) +
+  geom_point(aes(y = model, x = mean, color = site),
[…]

### 2024-10-16 ms/ms.Rmd
ms: minor edits in methods-models

@@
   bookdown::pdf_book:
+    keep_md: true
     base_format: rticles::copernicus_article # for using bookdown features like \@ref()
@@
 We calculated unitless slope from the \unit{1\,m} DTM using the Whitebox software.
-Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell [@schonauerSpatiotemporalPredictionSoil2021; @schonauerRcodeCalculatingDepthtowater2021].
+Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell [@schonauerSpatiotemporalPredictionSoil2021; @schonauerRcodeCalculatingDepthwater2021].
 This catchments area layer was then thresholded at seven different levels (*flow initiation areas* from \unit{0.25\,ha} to \unit{16\,ha}) to estimate surface water features under moisture scenarios varying from wet to dry [@murphyModellingMappingTopographic2011; @agrenEvaluatingDigitalTerrain2014; @schonauerSpatiotemporalPredictionSoil2021].
@@
 RF by itself is not a spatial model, and it will only predict spatial structure in the outcome to the degree that the structure is captured by predictors.
-We considered using regression kriging, a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals [@henglGenericFrameworkSpatial2004].
+We considered using regression kriging --- a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals [@henglGenericFrameworkSpatial2004].
 The spatial component in regression kriging often improves map accuracy compared to a non-spatial model [@beguinPredictingSoilProperties2017; @lamichhaneDigitalSoilMapping2019; @mollaMachineLearningGeostatistical2023], but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations [@henglGenericFrameworkSpatial2004; @szaboMappingSoilHydraulic2019; @takoutsingComparingPredictionPerformance2022].
 If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.
-We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that simple RF rather regression kriging was justified at both sites.
+We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that (non-spatial) RF rather regression kriging was justified at both sites.
 
-For each site we followed the same modelling workflow, so the description below applies to Skrimfjella and Ørskogfjellet.
+For each site we followed the same modelling workflow.
 First we trained a RF to predict peat depth based on our 25 quantitative predictors, to simulate applications where radiometric and terrain data are available, but no maps of peat depth.
-Next we trained a RF with all 26 predictors (DMK peat depth class included), to fully leverage data available at these sites and check the added value of the national DMK dataset.
+Next we trained a RF with all 26 predictors (DMK peat depth class included), to fully leverage the relevant available data and check the added value of the national DMK dataset.
 Finally, we fit a simple linear model to only DMK depth class, to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth calibrated on the same data.
@@
 RFs were fit with 1000 trees, minimum node size of 5, and the number of predictors randomly sampled at each split was the square root of the total number of predictors (*ranger* default).
+We did not tune these hyperparameters because RFs are relatively insensitive to tuning [@probstHyperparametersTuningStrategies2019], and because it would require nested spatial cross-validation to prevent data leakage [@schratzHyperparameterTuningPerformance2019].
 
-In the absence of additional field work to collect a design-based independent validation set, we used a spatially stratified cross-validation scheme to evaluate model performance [@wadouxSpatialCrossvalidationNot2021; @meyerMachineLearningbasedGlobal2022].
+The performance of digital soil mapping must be evaluated with reference to a specific purpose [i.e., map vs.
+model validation, interpolation vs.
+extrapolation, @robertsCrossvalidationStrategiesData2017; @milaNearestNeighbourDistance2022], and here we aimed to evaluate maps covering the study areas.
+In the absence of additional field work to collect a design-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance [@wadouxSpatialCrossvalidationNot2021; @meyerMachineLearningbasedGlobal2022].
 Specifically, we used k-Means Nearest Neighbor Distance Matching (kNNDM), which creates cross-validation folds that mimic the spatial prediction task that is defined as the goal [@linnenbrinkKNNDMCVKfold2024].
@@
 That way, the spatial separation between folds is similar to the separation between training and prediction locations --- which increases the quality of the map accuracy estimate [@linnenbrinkKNNDMCVKfold2024].
-For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatially stratified cross-validation [@wadouxSpatialCrossvalidationNot2021].
+For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation [@wadouxSpatialCrossvalidationNot2021].
 We implemented the kNNDM with the *CAST* R package [v.1.0.2, @meyerCASTPackageTraining2024], setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
-In general, the performance of digital soil mapping must be evaluated with reference to a specific purpose [i.e., map vs.
-model validation, interpolation vs.
-extrapolation, @robertsCrossvalidationStrategiesData2017; @milaNearestNeighbourDistance2022], and using kNNDM with a defined prediction domain increases the likelihood of quality accuracy estimates for a particular map.
+From the cross-validation we quantified *root mean squared error* (accuracy), *R^2^* (correlation), and *Lin's concordance correlation coefficient* (accuracy + correlation).
+
+We quantified global variable importance for the RFs with the 25 quantitative predictors.
+Global variable importance measures the influence of a given predictor on the output of the model, aggregated across all locations.
+We calculated variable importance with the *vip* R package (v.0.4.1), by three different methods: *FIRM*, *permutation*, and *Shapley* [@greenwellVariableImportancePlots2020].
+*Permutation* values were aggregated from ten iterations with RMSE as the performance measure.
 

### 2024-10-16 ms/ms.tex
ms: minor edits in methods-models

@@
 RF by itself is not a spatial model, and it will only predict spatial structure in the outcome to the degree that the structure is captured by predictors.
-We considered using regression kriging, a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals \citep{henglGenericFrameworkSpatial2004}.
+We considered using regression kriging --- a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals \citep{henglGenericFrameworkSpatial2004}.
 The spatial component in regression kriging often improves map accuracy compared to a non-spatial model \citep{beguinPredictingSoilProperties2017, lamichhaneDigitalSoilMapping2019, mollaMachineLearningGeostatistical2023}, but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations \citep{henglGenericFrameworkSpatial2004, szaboMappingSoilHydraulic2019, takoutsingComparingPredictionPerformance2022}.
 If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.
-We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that simple RF rather regression kriging was justified at both sites.
+We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that (non-spatial) RF rather regression kriging was justified at both sites.
 
-For each site we followed the same modelling workflow, so the description below applies to Skrimfjella and Ørskogfjellet.
+For each site we followed the same modelling workflow.
 First we trained a RF to predict peat depth based on our 25 quantitative predictors, to simulate applications where radiometric and terrain data are available, but no maps of peat depth.
-Next we trained a RF with all 26 predictors (DMK peat depth class included), to fully leverage data available at these sites and check the added value of the national DMK dataset.
+Next we trained a RF with all 26 predictors (DMK peat depth class included), to fully leverage the relevant available data and check the added value of the national DMK dataset.
 Finally, we fit a simple linear model to only DMK depth class, to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth calibrated on the same data.
@@
 RFs were fit with 1000 trees, minimum node size of 5, and the number of predictors randomly sampled at each split was the square root of the total number of predictors (\emph{ranger} default).
+We did not tune these hyperparameters because RFs are relatively insensitive to tuning \citep{probstHyperparametersTuningStrategies2019}, and because it would require nested spatial cross-validation to prevent data leakage \citep{schratzHyperparameterTuningPerformance2019}.
 
-In the absence of additional field work to collect a design-based independent validation set, we used a spatially stratified cross-validation scheme to evaluate model performance \citep{wadouxSpatialCrossvalidationNot2021, meyerMachineLearningbasedGlobal2022}.
+The performance of digital soil mapping must be evaluated with reference to a specific purpose \citep[i.e., map vs.~model validation, interpolation vs.~extrapolation,][]{robertsCrossvalidationStrategiesData2017, milaNearestNeighbourDistance2022}, and here we aimed to evaluate maps covering the study areas.
+In the absence of additional field work to collect a design-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance \citep{wadouxSpatialCrossvalidationNot2021, meyerMachineLearningbasedGlobal2022}.
 Specifically, we used k-Means Nearest Neighbor Distance Matching (kNNDM), which creates cross-validation folds that mimic the spatial prediction task that is defined as the goal \citep{linnenbrinkKNNDMCVKfold2024}.
@@
 That way, the spatial separation between folds is similar to the separation between training and prediction locations --- which increases the quality of the map accuracy estimate \citep{linnenbrinkKNNDMCVKfold2024}.
-For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatially stratified cross-validation \citep{wadouxSpatialCrossvalidationNot2021}.
+For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation \citep{wadouxSpatialCrossvalidationNot2021}.
 We implemented the kNNDM with the \emph{CAST} R package \citep[v.1.0.2,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
-In general, the performance of digital soil mapping must be evaluated with reference to a specific purpose \citep[i.e., map vs.~model validation, interpolation vs.~extrapolation,][]{robertsCrossvalidationStrategiesData2017, milaNearestNeighbourDistance2022}, and using kNNDM with a defined prediction domain increases the likelihood of quality accuracy estimates for a particular map.
+From the cross-validation we quantified \emph{root mean squared error} (accuracy), \emph{R\textsuperscript{2}} (correlation), and \emph{Lin's concordance correlation coefficient} (accuracy + correlation).
 

### 2024-10-18 R/modelDepth.R
orskog: add QRF uncertainty

@@ -50,10 +50,94 @@ fit_remotesensing <-
   fit(frame)
 
 fit_remotesensing %>% 
-  augment(new_data = frame) %>%
-  ggplot(aes(depth_cm, .pred)) +
-  geom_point() +
-  geom_abline()
+  augment(new_data = frame) %>% 
+  probably::cal_plot_regression(depth_cm, .pred, smooth = FALSE)
+
+## Uncertainty ####
+
+### Outside tidymodels ####
+train.dat <- frame %>% 
+  st_drop_geometry() %>% 
+  select(c(1,6:30))
+set.seed(345)
+ranger.qrf <- ranger::ranger(formula = depth_cm ~ ., 
+       data = train.dat, 
+       num.trees = 1000,
+       mtry = floor(sqrt(ncol(train.dat)-1)), 
+       min.node.size = 5,
+       quantreg = TRUE,
+       seed = 345)
+ranger.quantiles <- predict(ranger.qrf, train.dat, 
+                type = "quantiles",
+                quantiles = c(0.05, 0.95))$predictions %>% 
+  as_tibble() %>% 
+  mutate(observed = train.dat$depth_cm, .before = 1)
+  
+ranger.quantiles %>% 
+  arrange(observed) %>% 
+  rowid_to_column() %>%
+  pivot_longer(cols = !matches("rowid"), names_to = "what", values_to = "value") %>% 
+  ggplot(aes(x= rowid, y = value, color = what)) +
+  geom_point() 
+ranger.quantiles <- ranger.quantiles %>% 
+  mutate(coverage = ifelse(observed >= `quantile= 0.05` & 
+                             observed <= `quantile= 0.95`, 1, 0))
+ranger.quantiles %>% 
+  pull(coverage) %>%
+  mean()
+
+### In tidymodels ####
+# Note: no tidymodels-native syntax for quantile predictions yet (https://github.com/tidymodels/parsnip/issues/119)
+mod_qrf <- 
+  rand_forest(mtry = NULL, min_n = 5, trees = 1000) %>% 
+  set_engine("ranger", quantreg = TRUE, seed = 345) %>% 
+  set_mode("regression")
+workflow_remotesensing_uncertainty <- 
+  workflow() %>% 
+  add_model(mod_qrf) %>% 
+  add_recipe(recipe_remotesensing)
+set.seed(345)
+fit_remotesensing_uncertainty <- 
+  workflow_remotesensing_uncertainty %>% 
+  fit(frame)
[…]

### 2024-10-29 ms/ms.Rmd
Model interpretation with noncollinear features Added pdp Harmonized modeling across sites Ms: model approach, performance, interpretation

@@
 
-We created the same suite of 25 quantitative peat depth predictors for both sites (Table \@ref(tab:preds)).
-All quantitative predictors were derived either from an airborne radiometric survey or from a DTM.
+We created the same suite of peat depth predictors for both sites (25 continuous and 1 categorical; Table \@ref(tab:preds)).
+All continuous predictors were derived either from an airborne radiometric survey or from a DTM.
 From the radiometric surveys we simply used the four variables produced by the surveyors (Geological Survey of Norway): ground concentration of Potassium, Thorium, Uranium, as well as total count.
 From the DTMs we calculated several land surface parameters, ranging from simple terrain indices to more complex geomorphometric and hydrological variables [@maxwellLandsurfaceParametersSpatial2022].
+The categorical predictor was peat depth class, from a national map dataset.
+Complete descriptions of predictors follow below.
 ```{r preds, echo=FALSE, message=FALSE}
 df <- readr::read_csv("tables/predictors.csv")
-knitr::kable(df, booktabs = FALSE, caption = "Quantitative predictors of peat depth.")
+knitr::kable(df, booktabs = FALSE, caption = "Candidate predictors of peat depth.")
 ```
+
+### Radiometric
 
@@
 
+### Terrain
+
 For terrain-derived variables, we obtained \unit{1\,m} resolution DTMs from the Norwegian Mapping Authority.
@@
 This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties [@millerImpactMultiscalePredictor2015; @dornikOptimalScalingPredictors2022; @newmanAssessingSpatiallyHeterogeneous2023].
-We chose \unit{1\,m} and \unit{10\,m} resolutions because we know that peat depth tends to vary at fine scales in Norway [@maxwellLandsurfaceParametersSpatial2022].
+We know that peat depth tends to vary at fine scales in Norway, which is why we chose \unit{1\,m} and \unit{10\,m} resolutions [@maxwellLandsurfaceParametersSpatial2022].
 We also calculated the the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm [@gallantMultiresolutionIndexValley2003].
@@
 
-In addition to the quantitative predictors described above, we prepared one categorical predictor --- peat depth class --- from a historical national map dataset called *DMK* [@ahlstromAR5Klassifikasjonssystem2019].
+### Peat depth class
+
+We prepared one categorical predictor --- peat depth class --- from a historical national map dataset called *DMK* [@ahlstromAR5Klassifikasjonssystem2019].
 The DMK peat depth classes are: < \unit{1\,m} (*shallow*), > \unit{1\,m} (*deep*), and *unknown*.
@@
 ## Predictive models of peat depth
+
+### Modelling approach
 
@@
 
-For each site we followed the same modelling workflow.
-First we trained a RF to predict peat depth based on our 25 quantitative predictors, to simulate applications where radiometric and terrain data are available, but no maps of peat depth.
-Next we trained a RF with all 26 predictors (DMK peat depth class included), to fully leverage the relevant available data and check the added value of the national DMK dataset.
-Finally, we fit a simple linear model to only DMK depth class, to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth calibrated on the same data.
 We implemented models in the *tidymodels* framework in R [@kuhnTidymodelsCollectionPackages2020], with the *ranger* R package for RFs [v.0.16, @wrightRangerFastImplementation2017].
@@
 
+### Model performance
+
+For both sites we compared the performance of models with five different configurations of three groups of predictors. 
+Specifically, we trained models with:
+
+  1. only DMK peat depth class (2 binary indicator variables from 3 levels)
+  2. only terrain (21 continuous variables)
+  3. terrain and DMK peat depth class (23 variables)
+  4. radiometric and terrain predictors (25 continuous variables)
+  5. all predictors (26 variables)
+
+These different configurations simulate different scenarios of data availability that are common in Norway or internationally.
+Comparing the different configurations allowed us to isolate the added value of each of the predictor groups.
+We did not explore configurations comprising radiometrics without terrain, because Lidar terrain surveys typically precede airborne radiometric surveys.
+The models with only DMK peat depth class were simple linear models rather than RFs, and served to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth, calibrated on the same data.
+
 The performance of digital soil mapping must be evaluated with reference to a specific purpose [i.e., map vs.
 model validation, interpolation vs.
-extrapolation, @robertsCrossvalidationStrategiesData2017; @milaNearestNeighbourDistance2022], and here we aimed to evaluate maps covering the study areas.
+extrapolation, @robertsCrossvalidationStrategiesData2017; @milaNearestNeighbourDistance2022], and here we aimed to evaluate maps of peat depth across the study areas.
 In the absence of additional field work to collect a design-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance [@wadouxSpatialCrossvalidationNot2021; @meyerMachineLearningbasedGlobal2022].
@@
 We implemented the kNNDM with the *CAST* R package [v.1.0.2, @meyerCASTPackageTraining2024], setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
-From the cross-validation we quantified *root mean squared error* (accuracy), *R^2^* (correlation), and *Lin's concordance correlation coefficient* (accuracy + correlation).
+From the cross-validation we quantified *Root Mean Squared Error* (accuracy, original scale), *R^2^* (correlation, standardized scale), and *Lin's concordance correlation coefficient* (accuracy and correlation, standardized scale).
 
-We quantified global variable importance for the RFs with the 25 quantitative predictors.
+Uncertainty.
+
+Extrapolation.
+
+### Model interpretation
+
+We quantified global variable importance and examined partial dependence plots for the best-performing predictor configuration at each site.
 Global variable importance measures the influence of a given predictor on the output of the model, aggregated across all locations.
+Partial dependence plots depict the shape of the fitted relationship between a given predictor and the outcome.
+Both are useful for understanding the mechanisms behind the model's predictions and the roles of the predictors in the model.
+
+For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration --- because correlation between predictors degrades variable importance measures [@stroblConditionalVariableImportance2008; @biauRandomForestGuided2016] and can produce misleading visualizations of predictor-outcome relationships [@biecekExplanatoryModelAnalysis2021; @dwivediExplainableAIXAI2023].
+Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation above 0.7.
+Thus, highly-correlated sets of predictors are represented by a single variable for the purposes of model interpretation.
+
 We calculated variable importance with the *vip* R package (v.0.4.1), by three different methods: *FIRM*, *permutation*, and *Shapley* [@greenwellVariableImportancePlots2020].
-*Permutation* values were aggregated from ten iterations with RMSE as the performance measure.
+*FIRM* values measure the flatness of the partial dependence plot, *permutation* values measure the decrease in model performance when the predictor is permuted, and *Shapley* values are aggregated from local, game-theoretical measures of variable importance [@greenwellVariableImportancePlots2020].
+*Permutation* values were obtained from ten iterations, with RMSE as the performance measure.
+
+We calculated partial dependence with the *pdp* R package [v.0.8.1, @greenwellPdpPackageConstructing2017].
+For the top six most important variables, we plotted both partial dependence and individual conditional expectation (ICE), to show the average effect of the predictor on the outcome and the variation in the effect across observations, repectively [@goldsteinPeekingBlackBox2015].
+Non-parallel ICE lines indicate the presence of interactions between predictors.
 

### 2024-10-29 ms/ms.tex
Model interpretation with noncollinear features Added pdp Harmonized modeling across sites Ms: model approach, performance, interpretation

@@
 
-We created the same suite of 25 quantitative peat depth predictors for both sites (Table \ref{tab:preds}).
-All quantitative predictors were derived either from an airborne radiometric survey or from a DTM.
+We created the same suite of peat depth predictors for both sites (25 continuous and 1 categorical; Table \ref{tab:preds}).
+All continuous predictors were derived either from an airborne radiometric survey or from a DTM.
 From the radiometric surveys we simply used the four variables produced by the surveyors (Geological Survey of Norway): ground concentration of Potassium, Thorium, Uranium, as well as total count.
 From the DTMs we calculated several land surface parameters, ranging from simple terrain indices to more complex geomorphometric and hydrological variables \citep{maxwellLandsurfaceParametersSpatial2022}.
+The categorical predictor was peat depth class, from a national map dataset.
+Complete descriptions of predictors follow below.
 
@@
 
-\caption{\label{tab:preds}Quantitative predictors of peat depth.}
+\caption{\label{tab:preds}Candidate predictors of peat depth.}
 \centering
@@
 \hline
+DMK & DMK peat depth class, categorical with 3 levels\\
+\hline
 \end{tabular}
 \end{table}
+
+\subsubsection{Radiometric}
 
@@
 
+\subsubsection{Terrain}
+
 For terrain-derived variables, we obtained \unit{1\,m} resolution DTMs from the Norwegian Mapping Authority.
@@
 This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, dornikOptimalScalingPredictors2022, newmanAssessingSpatiallyHeterogeneous2023}.
-We chose \unit{1\,m} and \unit{10\,m} resolutions because we know that peat depth tends to vary at fine scales in Norway \citep{maxwellLandsurfaceParametersSpatial2022}.
+We know that peat depth tends to vary at fine scales in Norway, which is why we chose \unit{1\,m} and \unit{10\,m} resolutions \citep{maxwellLandsurfaceParametersSpatial2022}.
 We also calculated the the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
@@
 
-In addition to the quantitative predictors described above, we prepared one categorical predictor --- peat depth class --- from a historical national map dataset called \emph{DMK} \citep{ahlstromAR5Klassifikasjonssystem2019}.
+\subsubsection{Peat depth class}
+
+We prepared one categorical predictor --- peat depth class --- from a historical national map dataset called \emph{DMK} \citep{ahlstromAR5Klassifikasjonssystem2019}.
 The DMK peat depth classes are: \textless{} \unit{1\,m} (\emph{shallow}), \textgreater{} \unit{1\,m} (\emph{deep}), and \emph{unknown}.
@@
 \subsection{Predictive models of peat depth}
+
+\subsubsection{Modelling approach}
 
@@
 
-For each site we followed the same modelling workflow.
-First we trained a RF to predict peat depth based on our 25 quantitative predictors, to simulate applications where radiometric and terrain data are available, but no maps of peat depth.
-Next we trained a RF with all 26 predictors (DMK peat depth class included), to fully leverage the relevant available data and check the added value of the national DMK dataset.
-Finally, we fit a simple linear model to only DMK depth class, to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth calibrated on the same data.
 We implemented models in the \emph{tidymodels} framework in R \citep{kuhnTidymodelsCollectionPackages2020}, with the \emph{ranger} R package for RFs \citep[v.0.16,][]{wrightRangerFastImplementation2017}.
@@
 
-The performance of digital soil mapping must be evaluated with reference to a specific purpose \citep[i.e., map vs.~model validation, interpolation vs.~extrapolation,][]{robertsCrossvalidationStrategiesData2017, milaNearestNeighbourDistance2022}, and here we aimed to evaluate maps covering the study areas.
+\subsubsection{Model performance}
+
+For both sites we compared the performance of models with five different configurations of three groups of predictors.
+Specifically, we trained models with:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  only DMK peat depth class (2 binary indicator variables from 3 levels)
+\item
+  only terrain (21 continuous variables)
+\item
+  terrain and DMK peat depth class (23 variables)
+\item
+  radiometric and terrain predictors (25 continuous variables)
+\item
+  all predictors (26 variables)
+\end{enumerate}
+
+These different configurations simulate different scenarios of data availability that are common in Norway or internationally.
+Comparing the different configurations allowed us to isolate the added value of each of the predictor groups.
+We did not explore configurations comprising radiometrics without terrain, because Lidar terrain surveys typically precede airborne radiometric surveys.
+The models with only DMK peat depth class were simple linear models rather than RFs, and served to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth, calibrated on the same data.
+
+The performance of digital soil mapping must be evaluated with reference to a specific purpose \citep[i.e., map vs.~model validation, interpolation vs.~extrapolation,][]{robertsCrossvalidationStrategiesData2017, milaNearestNeighbourDistance2022}, and here we aimed to evaluate maps of peat depth across the study areas.
 In the absence of additional field work to collect a design-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance \citep{wadouxSpatialCrossvalidationNot2021, meyerMachineLearningbasedGlobal2022}.
@@
 We implemented the kNNDM with the \emph{CAST} R package \citep[v.1.0.2,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
-From the cross-validation we quantified \emph{root mean squared error} (accuracy), \emph{R\textsuperscript{2}} (correlation), and \emph{Lin's concordance correlation coefficient} (accuracy + correlation).
+From the cross-validation we quantified \emph{Root Mean Squared Error} (accuracy, original scale), \emph{R\textsuperscript{2}} (correlation, standardized scale), and \emph{Lin's concordance correlation coefficient} (accuracy and correlation, standardized scale).
 
-We quantified global variable importance for the RFs with the 25 quantitative predictors.
+Uncertainty.
+
+Extrapolation.
+
+\subsubsection{Model interpretation}
+
+We quantified global variable importance and examined partial dependence plots for the best-performing predictor configuration at each site.
 Global variable importance measures the influence of a given predictor on the output of the model, aggregated across all locations.
+Partial dependence plots depict the shape of the fitted relationship between a given predictor and the outcome.
+Both are useful for understanding the mechanisms behind the model's predictions and the roles of the predictors in the model.
+
+For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration --- because correlation between predictors degrades variable importance measures \citep{stroblConditionalVariableImportance2008, biauRandomForestGuided2016} and can produce misleading visualizations of predictor-outcome relationships \citep{biecekExplanatoryModelAnalysis2021, dwivediExplainableAIXAI2023}.
+Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation above 0.7.
+Thus, highly-correlated sets of predictors are represented by a single variable for the purposes of model interpretation.
+
 We calculated variable importance with the \emph{vip} R package (v.0.4.1), by three different methods: \emph{FIRM}, \emph{permutation}, and \emph{Shapley} \citep{greenwellVariableImportancePlots2020}.
-\emph{Permutation} values were aggregated from ten iterations with RMSE as the performance measure.
+\emph{FIRM} values measure the flatness of the partial dependence plot, \emph{permutation} values measure the decrease in model performance when the predictor is permuted, and \emph{Shapley} values are aggregated from local, game-theoretical measures of variable importance \citep{greenwellVariableImportancePlots2020}.
+\emph{Permutation} values were obtained from ten iterations, with RMSE as the performance measure.
+
+We calculated partial dependence with the \emph{pdp} R package \citep[v.0.8.1,][]{greenwellPdpPackageConstructing2017}.
+For the top six most important variables, we plotted both partial dependence and individual conditional expectation (ICE), to show the average effect of the predictor on the outcome and the variation in the effect across observations, repectively \citep{goldsteinPeekingBlackBox2015}.
+Non-parallel ICE lines indicate the presence of interactions between predictors.
 

### 2024-10-29 R/Skrim/modelDepth.R

@@ -1,4 +1,4 @@
-library(terra)
+library(tidyverse)
 library(sf)
 library(tidymodels)
 
@@ -7,13 +7,9 @@ library(tidymodels)
 frame <- st_read("output/Skrim/modeling.gpkg", layer="dataframe") |> 
   mutate(across(.cols = c(ar5cover, ar5soil, dmkdepth), .fns = as.factor)) |> 
   filter(!(ar5cover %in% c(11,12)))
-predictors <- rast("output/Skrim/predictors.tif")
-
-plot(predictors$elevation)
-plot(frame[,"depth_cm"],add=T)
+predictors <- terra::rast("output/Skrim/predictors.tif")
 
 predictivedomain <- st_read("data/Skrim/Skrim-site.gpkg", "mask_predictivedomain")
-plot(predictivedomain)
 
 # Make CV folds ####
 
@@ -21,14 +17,14 @@ plot(predictivedomain)
 library(CAST) #v.1.0.2
 
 predictorspts <- predictors |> 
-  as.points(values = FALSE) |> 
+  terra::as.points(values = FALSE) |> 
   st_as_sf()
 predpts <- predictorspts[predictivedomain,] |> 
   st_transform(st_crs(frame))
 st_crs(frame) == st_crs(predpts)
 
 set.seed(123)
-predptssample <- dplyr::slice_sample(predpts, n=1e3)
+predptssample <- slice_sample(predpts, n=1e3)
 # Vary number of folds
 knndm_list <- c(20,15,10,5) |>
   purrr::map(\(x) knndm(tpoints = frame, 
@@ -38,7 +34,7 @@ knndm_list <- c(20,15,10,5) |>
 knndm <- knndm_list[[which.min(map_dbl(knndm_list, \(x) x$W))]]
 knndm
 plot(knndm, type = "simple")
-ggplot() + geom_sf(data = dplyr::mutate(frame, cvfold = as.factor(knndm$clusters)), 
+ggplot() + geom_sf(data = mutate(frame, cvfold = as.factor(knndm$clusters)), 
                    aes(color=cvfold), size=0.5, shape=3)
 
 # Creating rsample splits
@@ -67,23 +63,20 @@ evaluation_metrics <- metric_set(rmse, mae, rsq, ccc)
 ### preds: radiometric + terrain ####
 
 recipe_RT <- 
-  recipe(formula = depth_cm ~ ., data = select(frame, !starts_with("source"))) |> 
-  remove_role(ar5cover, old_role = "predictor") |> 
[…]

### 2024-10-29 R/exploratoryDataAnalysis.R

@@ -52,23 +52,23 @@ sqrt(760*2) # Nugget at 10 m of 40 cm
 # Sample attributes ####
 
 depthcells <- st_read("output/modeling.gpkg", "dataframe", stringsAsFactors = TRUE)
-count(depthcells, ar5cover)
-count(depthcells, ar5cover, ar5soil)
-count(depthcells, dmkdepth)
+dplyr::count(depthcells, ar5cover)
+dplyr::count(depthcells, ar5cover, ar5soil)
+dplyr::count(depthcells, dmkdepth)
 depthcells |> 
   st_drop_geometry() |> 
   group_by(ar5cover) |> 
-  summarize(n = n(), depth_cm = mean(depth_cm))
+  dplyr::summarize(n = n(), depth_cm = mean(depth_cm))
 depthcells |> 
   st_drop_geometry() |> 
   filter(ar5cover %in% c(30,50,60)) |> 
   group_by(ar5cover, ar5soil) |> 
-  summarize(n = n(), depth_cm = mean(depth_cm))
+  dplyr::summarize(n = n(), depth_cm = mean(depth_cm))
 depthcells |> 
   st_drop_geometry() |> 
   filter(ar5cover == 60) |> 
   group_by(dmkdepth) |> 
-  summarize(n = n(), depth_cm = mean(depth_cm))
+  dplyr::summarize(n = n(), depth_cm = mean(depth_cm))
   
 # Representativeness of the sample ####
 # Representativeness cf. mapped peatland in the study area
@@ -179,7 +179,10 @@ df <- st_read("output/modeling.gpkg", "dataframe") |>
   as_tibble()
 metadata <- c("sourceProbe", "sourceGPR", "sourceWisen2021", "sourceMyrarkivet")
 auxiliary <- c("ar5cover", "ar5soil", "dmkdepth")
-sum(!complete.cases(df))
+df |> 
+  select(!any_of(c(metadata, auxiliary))) |> 
+  complete.cases() |> 
+  all()
 
 svg("output/exploratoryDataAnalysis-corr.svg", width = 12, height = 12)
 corrplot::corrplot(cor(select(df, !any_of(c(metadata, auxiliary)))), 

### 2024-10-29 R/modelDepth.R

@@ -1,4 +1,4 @@
-library(terra)
+library(tidyverse)
 library(sf)
 library(tidymodels)
 
@@ -7,13 +7,9 @@ library(tidymodels)
 frame <- st_read("output/modeling.gpkg", layer="dataframe") |> 
   mutate(across(.cols = c(ar5cover, ar5soil, dmkdepth), .fns = as.factor)) |> 
   filter(!(ar5cover %in% c(11,12)))
-predictors <- rast("output/predictors.tif")
-
-plot(predictors$elevation)
-plot(frame[,"depth_cm"],add=T)
+predictors <- terra::rast("output/predictors.tif")
 
 predictivedomain <- st_read("data/Orskogfjellet-site.gpkg", "mask_predictivedomain")
-plot(predictivedomain)
 
 # Make CV folds ####
 
@@ -21,14 +17,14 @@ plot(predictivedomain)
 library(CAST) #v.1.0.2
 
 predictorspts <- predictors |> 
-  as.points(values = FALSE) |> 
+  terra::as.points(values = FALSE) |> 
   st_as_sf()
 predpts <- predictorspts[predictivedomain,] |> 
   st_transform(st_crs(frame))
 st_crs(frame) == st_crs(predpts)
 
 set.seed(123)
-predptssample <- dplyr::slice_sample(predpts, n=1e3)
+predptssample <- slice_sample(predpts, n=1e3)
 # Vary number of folds
 knndm_list <- c(20,15,10,5) |>
   purrr::map(\(x) knndm(tpoints = frame, 
@@ -38,7 +34,7 @@ knndm_list <- c(20,15,10,5) |>
 knndm <- knndm_list[[which.min(map_dbl(knndm_list, \(x) x$W))]]
 knndm
 plot(knndm, type = "simple")
-ggplot() + geom_sf(data = dplyr::mutate(frame, cvfold = as.factor(knndm$clusters)), 
+ggplot() + geom_sf(data = mutate(frame, cvfold = as.factor(knndm$clusters)), 
                    aes(color=cvfold), size=0.5, shape=3)
 
 # Creating rsample splits
@@ -67,23 +63,20 @@ evaluation_metrics <- metric_set(rmse, mae, rsq, ccc)
 ### preds: radiometric + terrain ####
 
 recipe_RT <- 
-  recipe(formula = depth_cm ~ ., data = select(frame, !starts_with("source"))) |> 
-  remove_role(ar5cover, old_role = "predictor") |> 
[…]

### 2024-10-29 ms/figures.R

@@ -164,6 +164,8 @@ g2 + g3 + g1 + plot_layout(design = layout) +
   plot_annotation(tag_levels = 'a', tag_prefix = '(', tag_suffix = ')') 
 ggsave(filename = 'sites-patchwork.pdf', path = "ms/figures",
        width =210-30, height = 240-40, units = 'mm') #copernicus.cls page 210x240
+# ggsave(filename = 'sites-patchwork.svg', path = "ms/figures",
+#        width =210-30, height = 240-40, units = 'mm')
 
 ## With tmap instead of ggplot2 ####
 # library(tmap)
@@ -206,17 +208,17 @@ plotting <- bind_rows(orskog = orskog, skrim = skrim, .id = 'site') %>%
       .metric == "rsq" ~ "R-squared",
       .metric == "rmse" ~ "RMSE"),
     model = case_when(
-      model == "DMK" ~ "DMK depth class (1)",
-      model == "Terrain" ~ "Terrain (21)",
-      model == "TerrainDMK" ~ "Terrain + DMK (22)",
-      model == "RadiometricTerrain" ~ "Radiometric + Terrain (25)",
-      model == "RadiometricTerrainDMK" ~ "All predictors (26)"),
+      model == "DMK" ~ "DMK class (1)",
+      model == "Terrain" ~ "terrain (21)",
+      model == "TerrainDMK" ~ "terrain + DMK class (22)",
+      model == "RadiometricTerrain" ~ "radiometric + terrain (25)",
+      model == "RadiometricTerrainDMK" ~ "all predictors (26)"),
     model = fct_relevel(model,
-                        "DMK depth class (1)",
-                        "Terrain (21)",
-                        "Terrain + DMK (22)",
-                        "Radiometric + Terrain (25)",
-                        "All predictors (26)"))
+                        "DMK class (1)",
+                        "terrain (21)",
+                        "terrain + DMK class (22)",
+                        "radiometric + terrain (25)",
+                        "all predictors (26)"))
 str(plotting)
 
 g1 <- ggplot(plotting) +
@@ -244,6 +246,8 @@ g1 <- ggplot(plotting) +
         legend.key.spacing.y = unit(-2, "mm"),)
 ggsave(filename = 'modelmetrics.pdf', path = "ms/figures",
        width =210-30, height = (240-40)/2.5, units = 'mm') #copernicus.cls page 210x240
[…]

### 2024-10-29 ms/tables.R

@@ -30,7 +30,8 @@ tribble(
   "DTW20000", "Depth-to-water index, flow initiation area of 2 ha",
   "DTW40000", "Depth-to-water index, flow initiation area of 4 ha",
   "DTW80000", "Depth-to-water index, flow initiation area of 8 ha",
-  "DTW160000", "Depth-to-water index, flow initiation area of 16 ha") |> 
+  "DTW160000", "Depth-to-water index, flow initiation area of 16 ha",
+  "DMK", "DMK peat depth class, categorical with 3 levels") |> 
   write_csv("ms/tables/predictors.csv")
 
 # sessionInfo ####

### 2024-10-29 ms/ms.Rmd
ms: model performance - uncertainty

@@
 RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees [@breimanRandomForests2001].
-We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in digital soil mapping applications than other algorithms [@beguinPredictingSoilProperties2017; @nussbaumEvaluationDigitalSoil2018; @lamichhaneDigitalSoilMapping2019].
+We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in DSM applications than other algorithms [@beguinPredictingSoilProperties2017; @nussbaumEvaluationDigitalSoil2018; @lamichhaneDigitalSoilMapping2019].
 It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance [@khaledianSelectingAppropriateMachine2020].
@@
 
-The performance of digital soil mapping must be evaluated with reference to a specific purpose [i.e., map vs.
+The performance of DSM must be evaluated with reference to a specific purpose [i.e., map vs.
 model validation, interpolation vs.
@@
 
-Uncertainty.
+DSM products have much more value when their predictions are accompanied by uncertainty estimates, and all DSM should strive to assess uncertainty [@arrouaysImpressionsDigitalSoil2020; @wadouxMachineLearningDigital2020].
+Moreover, the quality of uncertainty estimates should be evaluated, just as the quality of predictions are   [@heuvelinkSpatialStatisticsSoil2022].
+Therefore, we produced prediction intervals with quantile regression forests [@meinshausenQuantileRegressionForests2006], and used the same spatial cross-validation to evaluate the prediction interval coverage probability [@shresthaMachineLearningApproaches2006].
+The quantile regression forests were trained with predictor configuration that showed the highest performance at each site (under the assumption that these models would be put into production) and we extracted 90 % prediction intervals.
 

### 2024-10-29 ms/ms.tex
ms: model performance - uncertainty

@@
 RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees \citep{breimanRandomForests2001}.
-We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in digital soil mapping applications than other algorithms \citep{beguinPredictingSoilProperties2017, nussbaumEvaluationDigitalSoil2018, lamichhaneDigitalSoilMapping2019}.
+We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in DSM applications than other algorithms \citep{beguinPredictingSoilProperties2017, nussbaumEvaluationDigitalSoil2018, lamichhaneDigitalSoilMapping2019}.
 It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance \citep{khaledianSelectingAppropriateMachine2020}.
@@
 
-The performance of digital soil mapping must be evaluated with reference to a specific purpose \citep[i.e., map vs.~model validation, interpolation vs.~extrapolation,][]{robertsCrossvalidationStrategiesData2017, milaNearestNeighbourDistance2022}, and here we aimed to evaluate maps of peat depth across the study areas.
+The performance of DSM must be evaluated with reference to a specific purpose \citep[i.e., map vs.~model validation, interpolation vs.~extrapolation,][]{robertsCrossvalidationStrategiesData2017, milaNearestNeighbourDistance2022}, and here we aimed to evaluate maps of peat depth across the study areas.
 In the absence of additional field work to collect a design-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance \citep{wadouxSpatialCrossvalidationNot2021, meyerMachineLearningbasedGlobal2022}.
@@
 
-Uncertainty.
+DSM products have much more value when their predictions are accompanied by uncertainty estimates, and all DSM should strive to assess uncertainty \citep{arrouaysImpressionsDigitalSoil2020, wadouxMachineLearningDigital2020}.
+Moreover, the quality of uncertainty estimates should be evaluated, just as the quality of predictions are \citep{heuvelinkSpatialStatisticsSoil2022}.
+Therefore, we produced prediction intervals with quantile regression forests \citep{meinshausenQuantileRegressionForests2006}, and used the same spatial cross-validation to evaluate the prediction interval coverage probability \citep{shresthaMachineLearningApproaches2006}.
+The quantile regression forests were trained with predictor configuration that showed the highest performance at each site (under the assumption that these models would be put into production) and we extracted 90 \% prediction intervals.
 

### 2024-10-29 ms/ms.Rmd
MS: model performance - peatland extrapolation

@@
 
-Extrapolation.
+While the primary aim of the DSM was to predict peat depth within peatlands (where peat > 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth outside of mapped peatland.
+In other words, we tested if predictions of deep peat indicated unmapped peatland occurrence in the wider landscape.
+It is practically difficult to obtain suitable data for this problem, because locations with non-zero peat depth outside of mapped peatland are unknown and relatively rare.
+Therefore, we made do with the same data that we used for the peat depth predictions, which included a fraction of locations not mapped as peatland.
+We used the same spatial cross-validation as for the peat depth predictions, but this time we trained the models only on locations mapped as peatland, and evaluated them only on locations not mapped as peatland.
+In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
+We evaluated models with radiometric and terrain predictors, or terrain predictors alone (since DMK peat depth class is always undefined outside known peatlands).
+From the cross-validation we quantified *Root Mean Squared Error* and *Mean Absolute Error*, and we tested the strength of the correlation between predicted and observed peat depth.
 

### 2024-10-29 ms/ms.tex
MS: model performance - peatland extrapolation

@@
 
-Extrapolation.
+While the primary aim of the DSM was to predict peat depth within peatlands (where peat \textgreater{} 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth outside of mapped peatland.
+In other words, we tested if predictions of deep peat indicated unmapped peatland occurrence in the wider landscape.
+It is practically difficult to obtain suitable data for this problem, because locations with non-zero peat depth outside of mapped peatland are unknown and relatively rare.
+Therefore, we made do with the same data that we used for the peat depth predictions, which included a fraction of locations not mapped as peatland.
+We used the same spatial cross-validation as for the peat depth predictions, but this time we trained the models only on locations mapped as peatland, and evaluated them only on locations not mapped as peatland.
+In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
+We evaluated models with radiometric and terrain predictors, or terrain predictors alone (since DMK peat depth class is always undefined outside known peatlands).
+From the cross-validation we quantified \emph{Root Mean Squared Error} and \emph{Mean Absolute Error}, and we tested the strength of the correlation between predicted and observed peat depth.
 

### 2024-10-30 ms/ms.Rmd
ms: minor edits - model performance

@@
   3. terrain and DMK peat depth class (23 variables)
-  4. radiometric and terrain predictors (25 continuous variables)
+  4. terrain and radiometric predictors (25 continuous variables)
   5. all predictors (26 variables)
@@
 Therefore, we made do with the same data that we used for the peat depth predictions, which included a fraction of locations not mapped as peatland.
-We used the same spatial cross-validation as for the peat depth predictions, but this time we trained the models only on locations mapped as peatland, and evaluated them only on locations not mapped as peatland.
+We used the same spatial cross-validation folds, but this time we trained the models only on locations mapped as peatland (in the training folds), and evaluated them only on locations not mapped as peatland (in the test fold).
 In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
-We evaluated models with radiometric and terrain predictors, or terrain predictors alone (since DMK peat depth class is always undefined outside known peatlands).
+Since DMK peat depth class is always undefined outside known peatlands, we evaluated two predictor configurations: terrain predictors alone or terrain and radiometric predictors.
 From the cross-validation we quantified *Root Mean Squared Error* and *Mean Absolute Error*, and we tested the strength of the correlation between predicted and observed peat depth.

### 2024-10-30 ms/ms.tex
ms: minor edits - model performance

@@
 \item
-  radiometric and terrain predictors (25 continuous variables)
+  terrain and radiometric predictors (25 continuous variables)
 \item
@@
 Therefore, we made do with the same data that we used for the peat depth predictions, which included a fraction of locations not mapped as peatland.
-We used the same spatial cross-validation as for the peat depth predictions, but this time we trained the models only on locations mapped as peatland, and evaluated them only on locations not mapped as peatland.
+We used the same spatial cross-validation folds, but this time we trained the models only on locations mapped as peatland (in the training folds), and evaluated them only on locations not mapped as peatland (in the test fold).
 In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
-We evaluated models with radiometric and terrain predictors, or terrain predictors alone (since DMK peat depth class is always undefined outside known peatlands).
+Since DMK peat depth class is always undefined outside known peatlands, we evaluated two predictor configurations: terrain predictors alone or terrain and radiometric predictors.
 From the cross-validation we quantified \emph{Root Mean Squared Error} and \emph{Mean Absolute Error}, and we tested the strength of the correlation between predicted and observed peat depth.

### 2024-10-30 ms/ms.Rmd
add Skrimfjella occurrence evaluation

--- 2024-10-29 ms/ms.Rmd
@@
 
-At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest model of peat depth, and we designed our sampling with this in mind [@brusSamplingDigitalSoil2019].
+At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest (RF) model of peat depth, and we designed our sampling with this in mind [@brusSamplingDigitalSoil2019].
 Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
@@
 We considered this a reasonable compromise between digital terrain model (DTM) resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.
+The point measurements of peat depth described below were ultimately aggregated to \unit{10\,m} resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
 
@@
 To choose 160 locations, we performed feature space coverage sampling.
-This approach has been found to produce higher accuracy in Random Forest models than conditioned latin hypercube sampling [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
+This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
 Feature space coverage sampling aims to disperse samples as uniformly as possible in multidimensional predictor space, and is implemented by choosing locations that are closest to cluster centers in a k-means clustering of the standardized predictor space [@brusSamplingDigitalSoil2019].
@@
 
-We used Random Forests (RF) to predict peat depth at both sites.
+We used RF to predict peat depth at both sites.
 RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees [@breimanRandomForests2001].
@@
 
-Extrapolation.
+While the primary aim of the DSM was to predict peat depth within mapped peatlands (where peat depth > 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth or occurrence outside of mapped peatland.
+In other words, we tested whether our peatland-trained models of peat depth could indicate which other locations were most likely to have peat.
+It is generally impractical to obtain good data for this purpose, because locations with non-zero peat depth outside of mapped peatland are unknown and relatively rare.
+For Skrimfjella, we had independent data from the arm of our sampling design that measured peatland occurrence outside of mapped mire areas.
+For Ørskogfjelle we did not have such data.
+Therefore, we first --- for both sites --- performed an evaluation with just our peat depth measurements, which included a fraction of locations not mapped as peatland.
+For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on locations mapped as peatland (in the training folds), and evaluated them only on locations not mapped as peatland (in the test fold).
+In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
+From the cross-validation we quantified *Root Mean Squared Error* and *Mean Absolute Error*, and we tested the strength of the correlation between predicted and observed peat depth.
+Second --- for Skrimfjella --- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
+Since DMK peat depth class is always undefined outside known peatlands, we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.
 

### 2024-10-30 ms/ms.tex
add Skrimfjella occurrence evaluation

--- 2024-10-29 ms/ms.tex
@@
 
-At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest model of peat depth, and we designed our sampling with this in mind \citep{brusSamplingDigitalSoil2019}.
+At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest (RF) model of peat depth, and we designed our sampling with this in mind \citep{brusSamplingDigitalSoil2019}.
 Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
@@
 To choose 160 locations, we performed feature space coverage sampling.
-This approach has been found to produce higher accuracy in Random Forest models than conditioned latin hypercube sampling \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
+This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
 Feature space coverage sampling aims to disperse samples as uniformly as possible in multidimensional predictor space, and is implemented by choosing locations that are closest to cluster centers in a k-means clustering of the standardized predictor space \citep{brusSamplingDigitalSoil2019}.
@@
 
-We used Random Forests (RF) to predict peat depth at both sites.
+We used RF to predict peat depth at both sites.
 RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees \citep{breimanRandomForests2001}.
@@
 
-Extrapolation.
+While the primary aim of the DSM was to predict peat depth within mapped peatlands (where peat depth \textgreater{} 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth or occurrence outside of mapped peatland.
+In other words, we tested whether our peatland-trained models of peat depth could indicate which other locations were most likely to have peat.
+It is generally impractical to obtain good data for this purpose, because locations with non-zero peat depth outside of mapped peatland are unknown and relatively rare.
+For Skrimfjella, we had independent data from the arm of our sampling design that measured peatland occurrence outside of mapped mire areas.
+For Ørskogfjelle we did not have such data.
+Therefore, we first --- for both sites --- performed an evaluation with just our peat depth measurements, which included a fraction of locations not mapped as peatland.
+For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on locations mapped as peatland (in the training folds), and evaluated them only on locations not mapped as peatland (in the test fold).
+In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
+From the cross-validation we quantified \emph{Root Mean Squared Error} and \emph{Mean Absolute Error}, and we tested the strength of the correlation between predicted and observed peat depth.
+Second --- for Skrimfjella --- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
+Since DMK peat depth class is always undefined outside known peatlands, we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.
 

### 2024-10-30 R/RUNALL.R

@@ -12,3 +12,4 @@ render("R/Skrim/createPredictorStack.R",output_format="html_document", knit_root
 render("R/Skrim/createDataFrame.R",output_format="html_document", knit_root_dir = "../../")
 render("R/Skrim/exploratoryDataAnalysis.R",output_format="html_document", knit_root_dir = "../../")
 render("R/Skrim/modelDepth.R", output_format="html_document", knit_root_dir = "../../")
+render("R/Skrim/modelDepth-evaluateOccurrence.R", output_format="html_document", knit_root_dir = "../../")

### 2024-10-30 R/Skrim/createDataFrame.R

@@ -4,12 +4,15 @@ library(sf)
 
 depthpts <- read_csv("data/Skrim/depth_all.csv") |> 
   st_as_sf(coords = c('X', 'Y'), crs = 25833)
+occpts <- st_read("data/Skrim/Skrim-site.gpkg", "occurrence")
 
 # Raster origo and resolution ####
 radK10m <- rast("output/Skrim/predictors.tif")[[1]] 
 plot(radK10m)
 
-# Depth per cell ####
+# Depth frame ####
+
+## Depth per cell ####
 
 crs(depthpts) == crs(radK10m)
 depthcells <- extract(radK10m, depthpts, cells = TRUE, xy = TRUE, ID = FALSE)
@@ -44,18 +47,18 @@ celldepth <- depth |>
   select(-cell, -pts, -depthMean, -depthIDW) |> 
   st_as_sf(coords = c('x','y'), crs = st_crs(radK10m))
 
-# Write cell depths ####
+## Write cell depths ####
 
 write_sf(celldepth, "output/Skrim/modeling.gpkg", layer="celldepth", append = FALSE)
 
-# Join predictors ####
+## Join predictors ####
 
 predictors <- rast("output/Skrim/predictors.tif")
 frame <- celldepth |> 
   bind_cols(extract(predictors, celldepth, ID=FALSE, cell=FALSE)) |> 
   select(depth_cm, starts_with("source"), geometry, everything())
 
-# Join auxiliary attributes ####
+## Join auxiliary attributes ####
 
 ar5 <- st_read("data/Skrim/Basisdata_3303_Kongsberg_25832_FKB-AR5_FGDB.gdb", 
                layer="fkb_ar5_omrade") |> 
@@ -76,6 +79,37 @@ frame <- st_join(frame, dmk[,"myr"]) |>
 frame <- frame |> 
   mutate(across(.cols = c(ar5cover, ar5soil, dmkdepth), .fns = as.factor))
 
-# Write data frame ####
+## Write data frame ####
 
 write_sf(frame, "output/Skrim/modeling.gpkg", layer="dataframe", append = FALSE)
+
+# Occurrence frame ####
+
+## Occurrence per cell ####
+
+crs(occpts) == crs(radK10m)
+occcells <- extract(radK10m, occpts, cells = TRUE, xy = TRUE, ID = FALSE)
+occcells %>% 
+  pull(cell) %>% 
+  duplicated() %>% 
+  any()
+
[…]

### 2024-10-30 R/Skrim/modelDepth-evaluateOccurrence.R

@@ -1,0 +1,89 @@
+library(tidyverse)
+library(sf)
+library(tidymodels)
+
+# Reading data ####
+
+frame.depth <- st_read("output/Skrim/modeling.gpkg", layer="dataframe") |> 
+  mutate(across(.cols = c(ar5cover, ar5soil, dmkdepth), .fns = as.factor)) |> 
+  filter(!(ar5cover %in% c(11,12)))
+frame.occ <- st_read("output/Skrim/modeling.gpkg", layer="dataframe.occurrence") |> 
+  mutate(across(.cols = c(occurrence, ar5cover, ar5soil), .fns = as.factor)) |> 
+  filter(!(ar5cover %in% c(11,12)))
+
+# Evaluation with independent data ####
+
+mod_rf <- 
+  rand_forest(mtry = NULL, min_n = 5, trees = 1000) %>% 
+  set_engine("ranger", importance = "permutation",
+             scale.permutation.importance	= TRUE) %>% 
+  set_mode("regression")
+
+## test ####
+
+### preds: radiometric + terrain ####
+
+recipe_RT <- 
+  recipe(formula = depth_cm ~ 
+           radK + radTh + radU + radTC +
+           elevation + 
+           slope1m + TPI1m + TRI1m + roughness1m + 
+           slope10m + TPI10m + TRI10m + roughness10m + 
+           MRVBF + 
+           TWI5m + TWI10m + TWI20m + TWI50m + 
+           DTW2500 + DTW5000 + DTW10000 + DTW20000 + DTW40000 + DTW80000 + DTW160000,
+         data = frame.depth)
+
+workflow_RT <- 
+  workflow() %>% 
+  add_model(mod_rf) %>% 
+  add_recipe(recipe_RT)
+
+set.seed(456)
+fit_RT <- 
+  workflow_RT %>% 
+  fit(frame.depth)
+
+eval_RT <- fit_RT %>% 
+  augment(new_data = frame.occ)
+
+eval_RT %>% 
+  roc_curve(truth = occurrence, .pred) %>% 
+  autoplot()
+eval_RT %>% 
+  roc_auc(truth = occurrence, .pred)
+
+### preds: terrain ####
+
+recipe_T <- 
+  recipe(formula = depth_cm ~ 
+           elevation + 
+           slope1m + TPI1m + TRI1m + roughness1m + 
[…]

### 2024-10-30 ms/ms.Rmd
minor style edit

@@
 The study area has a mean elevation of \unit{438\,m} above sea level (range 223--711, IQR 351--509), and its mean slope at \unit{10\,m} resolution is 10.8° (IQR 4.6--15.1°).
-In Norway's AR5 national land capability dataset [@ahlstromAR5Klassifikasjonssystem2019], \unit{1.5\,km^{2}} (4.5 %) of the study area is classified as mire --- defined as areas with mire vegetation and at least \unit{30\,cm} of peat depth.
+In Norway's AR5 national land capability dataset [@ahlstromAR5Klassifikasjonssystem2019], \unit{1.5\,km^{2}} (4.5 %) of the study area is classified as mire -- defined as areas with mire vegetation and at least \unit{30\,cm} of peat depth.
 
@@
 Probe locations were adjusted up to \unit{20\,cm} if the base of the peat column seemed to be blocked by an obvious artifact.
-For each occurrence sample location, we recorded the presence or absence of peatland --- primarily by digging and examining the top \unit{20\,cm} of soil (where this was possible).
+For each occurrence sample location, we recorded the presence or absence of peatland -- primarily by digging and examining the top \unit{20\,cm} of soil (where this was possible).
 We judged whether the soil was a peat soil based on its density, texture, and color.
@@
 Before selecting locations, we determined a minimal sample size that would adequately capture the terrain and radiometric properties of the entire mire area.
-Specifically, we aimed to identify the size at which adding locations produced diminishing decreases in divergence between sample and population distributions --- i.e.
+Specifically, we aimed to identify the size at which adding locations produced diminishing decreases in divergence between sample and population distributions -- i.e.
 the elbow point in a curve of similarity between sample and population [@maloneMethodsImproveUtility2019].
@@
 After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
-Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 % (with respect to the ideal sample), but distance in our sample was still only 46 % of the mean distance to cluster centers --- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
+Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 % (with respect to the ideal sample), but distance in our sample was still only 46 % of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
 
@@
 We processed the GPR data with Reflexw software (v.8.5, Sandmeier Scientific Software, Germany), applying a dewow filter, time-zero correction, bandpass filter, gain filter, and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennae.
-The last correction is important for the rough terrain antenna, which has an antenna separation (\unit{2.2\,m}) --- comparable to typical peat depths.
+The last correction is important for the rough terrain antenna, which has an antenna separation (\unit{2.2\,m}) -- comparable to typical peat depths.
 As with the data from Skrimfjella, we picked the base of the peat column from strong reflectors in the radargram, and calibrated wave velocity with manual probe measurements in a linear regression.
@@
 
-We prepared one categorical predictor --- peat depth class --- from a historical national map dataset called *DMK* [@ahlstromAR5Klassifikasjonssystem2019].
+We prepared one categorical predictor -- peat depth class -- from a historical national map dataset called *DMK* [@ahlstromAR5Klassifikasjonssystem2019].
 The DMK peat depth classes are: < \unit{1\,m} (*shallow*), > \unit{1\,m} (*deep*), and *unknown*.
@@
 RF by itself is not a spatial model, and it will only predict spatial structure in the outcome to the degree that the structure is captured by predictors.
-We considered using regression kriging --- a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals [@henglGenericFrameworkSpatial2004].
+We considered using regression kriging -- a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals [@henglGenericFrameworkSpatial2004].
 The spatial component in regression kriging often improves map accuracy compared to a non-spatial model [@beguinPredictingSoilProperties2017; @lamichhaneDigitalSoilMapping2019; @mollaMachineLearningGeostatistical2023], but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations [@henglGenericFrameworkSpatial2004; @szaboMappingSoilHydraulic2019; @takoutsingComparingPredictionPerformance2022].
@@
 In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
-That way, the spatial separation between folds is similar to the separation between training and prediction locations --- which increases the quality of the map accuracy estimate [@linnenbrinkKNNDMCVKfold2024].
+That way, the spatial separation between folds is similar to the separation between training and prediction locations -- which increases the quality of the map accuracy estimate [@linnenbrinkKNNDMCVKfold2024].
 For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation [@wadouxSpatialCrossvalidationNot2021].
@@
 For Ørskogfjelle we did not have such data.
-Therefore, we first --- for both sites --- performed an evaluation with just our peat depth measurements, which included a fraction of locations not mapped as peatland.
+Therefore, we first -- for both sites -- performed an evaluation with just our peat depth measurements, which included a fraction of locations not mapped as peatland.
 For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on locations mapped as peatland (in the training folds), and evaluated them only on locations not mapped as peatland (in the test fold).
@@
 From the cross-validation we quantified *Root Mean Squared Error* and *Mean Absolute Error*, and we tested the strength of the correlation between predicted and observed peat depth.
-Second --- for Skrimfjella --- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
+Second -- for Skrimfjella -- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
 Since DMK peat depth class is always undefined outside known peatlands, we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.
@@
 
-For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration --- because correlation between predictors degrades variable importance measures [@stroblConditionalVariableImportance2008; @biauRandomForestGuided2016] and can produce misleading visualizations of predictor-outcome relationships [@biecekExplanatoryModelAnalysis2021; @dwivediExplainableAIXAI2023].
+For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration -- because correlation between predictors degrades variable importance measures [@stroblConditionalVariableImportance2008; @biauRandomForestGuided2016] and can produce misleading visualizations of predictor-outcome relationships [@biecekExplanatoryModelAnalysis2021; @dwivediExplainableAIXAI2023].
 Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation above 0.7.

### 2024-10-30 ms/ms.tex
minor style edit

@@
 The study area has a mean elevation of \unit{438\,m} above sea level (range 223--711, IQR 351--509), and its mean slope at \unit{10\,m} resolution is 10.8° (IQR 4.6--15.1°).
-In Norway's AR5 national land capability dataset \citep{ahlstromAR5Klassifikasjonssystem2019}, \unit{1.5\,km^{2}} (4.5 \%) of the study area is classified as mire --- defined as areas with mire vegetation and at least \unit{30\,cm} of peat depth.
+In Norway's AR5 national land capability dataset \citep{ahlstromAR5Klassifikasjonssystem2019}, \unit{1.5\,km^{2}} (4.5 \%) of the study area is classified as mire -- defined as areas with mire vegetation and at least \unit{30\,cm} of peat depth.
 
@@
 Before selecting locations, we determined a minimal sample size that would adequately capture the terrain and radiometric properties of the entire mire area.
-Specifically, we aimed to identify the size at which adding locations produced diminishing decreases in divergence between sample and population distributions --- i.e.~the elbow point in a curve of similarity between sample and population \citep{maloneMethodsImproveUtility2019}.
+Specifically, we aimed to identify the size at which adding locations produced diminishing decreases in divergence between sample and population distributions -- i.e.~the elbow point in a curve of similarity between sample and population \citep{maloneMethodsImproveUtility2019}.
 This approach has been found to identify sample sizes that correspond with diminishing returns in predictive model performance on external evaluation data \citep{sauretteDivergenceMetricsDetermining2023}.
@@
 After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
-Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 \% (with respect to the ideal sample), but distance in our sample was still only 46 \% of the mean distance to cluster centers --- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
+Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 \% (with respect to the ideal sample), but distance in our sample was still only 46 \% of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
 
@@
 
-We prepared one categorical predictor --- peat depth class --- from a historical national map dataset called \emph{DMK} \citep{ahlstromAR5Klassifikasjonssystem2019}.
+We prepared one categorical predictor -- peat depth class -- from a historical national map dataset called \emph{DMK} \citep{ahlstromAR5Klassifikasjonssystem2019}.
 The DMK peat depth classes are: \textless{} \unit{1\,m} (\emph{shallow}), \textgreater{} \unit{1\,m} (\emph{deep}), and \emph{unknown}.
@@
 RF by itself is not a spatial model, and it will only predict spatial structure in the outcome to the degree that the structure is captured by predictors.
-We considered using regression kriging --- a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals \citep{henglGenericFrameworkSpatial2004}.
+We considered using regression kriging -- a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals \citep{henglGenericFrameworkSpatial2004}.
 The spatial component in regression kriging often improves map accuracy compared to a non-spatial model \citep{beguinPredictingSoilProperties2017, lamichhaneDigitalSoilMapping2019, mollaMachineLearningGeostatistical2023}, but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations \citep{henglGenericFrameworkSpatial2004, szaboMappingSoilHydraulic2019, takoutsingComparingPredictionPerformance2022}.
@@
 In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
-That way, the spatial separation between folds is similar to the separation between training and prediction locations --- which increases the quality of the map accuracy estimate \citep{linnenbrinkKNNDMCVKfold2024}.
+That way, the spatial separation between folds is similar to the separation between training and prediction locations -- which increases the quality of the map accuracy estimate \citep{linnenbrinkKNNDMCVKfold2024}.
 For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation \citep{wadouxSpatialCrossvalidationNot2021}.
@@
 From the cross-validation we quantified \emph{Root Mean Squared Error} and \emph{Mean Absolute Error}, and we tested the strength of the correlation between predicted and observed peat depth.
-Second --- for Skrimfjella --- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
+Second -- for Skrimfjella -- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
 Since DMK peat depth class is always undefined outside known peatlands, we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.
@@
 
-For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration --- because correlation between predictors degrades variable importance measures \citep{stroblConditionalVariableImportance2008, biauRandomForestGuided2016} and can produce misleading visualizations of predictor-outcome relationships \citep{biecekExplanatoryModelAnalysis2021, dwivediExplainableAIXAI2023}.
+For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration -- because correlation between predictors degrades variable importance measures \citep{stroblConditionalVariableImportance2008, biauRandomForestGuided2016} and can produce misleading visualizations of predictor-outcome relationships \citep{biecekExplanatoryModelAnalysis2021, dwivediExplainableAIXAI2023}.
 Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation above 0.7.

### 2024-10-30 ms/ms.Rmd
table style for clarity

@@
 df <- readr::read_csv("tables/predictors.csv")
-knitr::kable(df, booktabs = FALSE, caption = "Candidate predictors of peat depth.")
+knitr::kable(df, booktabs = FALSE, vline = "", linesep = "",
+             col.names = c("Name", "Description"),
+             caption = "Candidate predictors of peat depth.")
 ```
@@
 We calculated partial dependence with the *pdp* R package [v.0.8.1, @greenwellPdpPackageConstructing2017].
-For the top six most important variables, we plotted both partial dependence and individual conditional expectation (ICE), to show the average effect of the predictor on the outcome and the variation in the effect across observations, repectively [@goldsteinPeekingBlackBox2015].
+For the top six most important variables, we plotted both partial dependence and individual conditional expectation (ICE), to show the average effect of the predictor on the outcome and the variation in the effect across observations, respectively [@goldsteinPeekingBlackBox2015].
 Non-parallel ICE lines indicate the presence of interactions between predictors.

### 2024-10-30 ms/ms.tex
table style for clarity

@@
 \centering
-\begin{tabular}[t]{l|l}
+\begin{tabular}[t]{ll}
 \hline
-name & description\\
+Name & Description\\
 \hline
 radK & Potassium ground concentration\\
-\hline
 radTh & Thorium ground concentration\\
-\hline
 radU & Uranium ground concentration\\
-\hline
 radTC & Total count of gamma radiation\\
-\hline
 elevation & Mean elevation\\
-\hline
 slope1m & Mean of 1 m slope\\
-\hline
 TPI1m & Mean of 1 m topographic position index\\
-\hline
 TRI1m & Mean of 1 m terrain ruggedness index\\
-\hline
 roughness1m & Mean of 1 m roughness\\
-\hline
 slope10m & 10 m slope\\
-\hline
 TPI10m & 10 m topographic position index\\
-\hline
 TRI10m & 10 m terrain ruggedness index\\
-\hline
 roughness10m & 10 m roughness\\
-\hline
 MRVBF & Multi-resolution valley bottom flatness\\
-\hline
 TWI5m & Mean of 5 m topographic wetness index\\
-\hline
 TWI10m & 10 m topographic wetness index\\
-\hline
 TWI20m & Bilinear interpolation of 20 m topographic wetness index\\
-\hline
 TWI50m & Bilinear interpolation of 50 m topographic wetness index\\
-\hline
 DTW2500 & Depth-to-water index, flow initiation area of 0.25 ha\\
-\hline
 DTW5000 & Depth-to-water index, flow initiation area of 0.5 ha\\
-\hline
 DTW10000 & Depth-to-water index, flow initiation area of 1 ha\\
-\hline
 DTW20000 & Depth-to-water index, flow initiation area of 2 ha\\
-\hline
 DTW40000 & Depth-to-water index, flow initiation area of 4 ha\\
-\hline
 DTW80000 & Depth-to-water index, flow initiation area of 8 ha\\
-\hline
 DTW160000 & Depth-to-water index, flow initiation area of 16 ha\\
-\hline
 DMK & DMK peat depth class, categorical with 3 levels\\
@@
 We calculated partial dependence with the \emph{pdp} R package \citep[v.0.8.1,][]{greenwellPdpPackageConstructing2017}.
-For the top six most important variables, we plotted both partial dependence and individual conditional expectation (ICE), to show the average effect of the predictor on the outcome and the variation in the effect across observations, repectively \citep{goldsteinPeekingBlackBox2015}.
+For the top six most important variables, we plotted both partial dependence and individual conditional expectation (ICE), to show the average effect of the predictor on the outcome and the variation in the effect across observations, respectively \citep{goldsteinPeekingBlackBox2015}.
 Non-parallel ICE lines indicate the presence of interactions between predictors.

### 2024-10-30 ms/ms.Rmd
ms: add depthsByClass table

@@
 
-Include a 12cm width figure of Nikolaus Copernicus from [Wikipedia](https://en.wikipedia.org/wiki/File:Nikolaus_Kopernikus.jpg) with caption using R Markdown (Fig. \@ref(fig:portrait)).
+We obtained depth measurements for 372 cells (XX % of mapped peatland) at Skrimfjella and 1889 cells (XX % of mapped peatland) at Ørskogfjellet.
+The mean depths of these \unit{10\,m} cells differed by only \unit{6\,cm} between Skrimfjella and Ørskogfjellet (Table \ref{tab:depthsByClass}).
 
-```{r portrait, out.width = "8.3cm", echo = FALSE, fig.cap = "one column figure"}
-knitr::include_graphics("Nikolaus_Kopernikus.jpg")
-```
-
-## Tables
-
-You can add \LaTeX table in an R Markdown document to meet the template requirements (Table \ref{tab:latextable}).
-
-\begin{table}[t]
-\caption{TEXT}
-\begin{tabular}{l c r}
-\tophline
-
-a & b & c \\
-\middlehline
-1 & 2 & 3 \\
-
-\bottomhline
+\begin{table}[tbp]
+\caption{Mean peat depth (cm) in \unit{10\,m} cells at Skrimfjella and Ørskogfjellet. The cells are also shown stratified by AR5 land class and DMK peat depth class.}
+\begin{tabular}{llllllll}
+\hline
+    &                            & \multicolumn{3}{c}{Skrimfjella} & \multicolumn{3}{c}{Ørskogfjellet} \\ \cline{3-8} 
+    &                            & n        & \%      & depth      & n          & \%      & depth      \\ \hline
+    &                            & 372      & 100     & 119        & 1889       & 100     & 125        \\
+AR5 & Fully cultivated           &          &         &            & 21         & 1       & 30         \\
+    & Forest                     & 52       & 14      & 65         & 272        & 14      & 71         \\
+    & Open firm ground           & 19       & 5       & 98         & 134        & 7       & 39         \\
+    & Peatland                   & 301      & 81      & 130        & 1451       & 77      & 145        \\
+DMK & deep (\textgreater 100 cm) & 94       & 25      & 100        & 659        & 35      & 219        \\
+    & shallow (\textless 100 cm) & 6        & 2       & 50         & 839        & 44      & 82         \\
+    & unknown                    & 272      & 73      & 127        & 391        & 21      & 59         \\ \hline
 \end{tabular}
-\belowtable{Table Footnotes}
-\label{tab:latextable}
+\label{tab:depthsByClass}
 \end{table}
-
-Or you can use markdown to create the table with booktabs = FALSE (<https://github.com/rstudio/rticles/issues/558#issuecomment-1907981541>).
-
-See Table \@ref(tab:test).
-
-```{r test, echo=FALSE}
-df = mtcars[1:3, 1:3]
-knitr::kable(df, booktabs = FALSE, caption = "My caption")
-```
 

### 2024-10-30 ms/ms.tex
ms: add depthsByClass table

@@
 
-Include a 12cm width figure of Nikolaus Copernicus from \href{https://en.wikipedia.org/wiki/File:Nikolaus_Kopernikus.jpg}{Wikipedia} with caption using R Markdown (Fig. \ref{fig:portrait}).
+We obtained depth measurements for 372 cells (XX \% of mapped peatland) at Skrimfjella and 1889 cells (XX \% of mapped peatland) at Ørskogfjellet.
+The mean depths of these \unit{10\,m} cells differed by only \unit{6\,cm} between Skrimfjella and Ørskogfjellet (Table \ref{tab:depthsByClass}).
 
-\begin{figure}
-\includegraphics[width=8.3cm]{Nikolaus_Kopernikus} \caption{one column figure}\label{fig:portrait}
-\end{figure}
-
-\subsection{Tables}
-
-You can add \LaTeX table in an R Markdown document to meet the template requirements (Table \ref{tab:latextable}).
-
-\begin{table}[t]
-\caption{TEXT}
-\begin{tabular}{l c r}
-\tophline
-
-a & b & c \\
-\middlehline
-1 & 2 & 3 \\
-
-\bottomhline
+\begin{table}[tbp]
+\caption{Mean peat depth (cm) in \unit{10\,m} cells at Skrimfjella and Ørskogfjellet. The cells are also shown stratified by AR5 land class and DMK peat depth class.}
+\begin{tabular}{llllllll}
+\hline
+    &                            & \multicolumn{3}{c}{Skrimfjella} & \multicolumn{3}{c}{Ørskogfjellet} \\ \cline{3-8} 
+    &                            & n        & \%      & depth      & n          & \%      & depth      \\ \hline
+    &                            & 372      & 100     & 119        & 1889       & 100     & 125        \\
+AR5 & Fully cultivated           &          &         &            & 21         & 1       & 30         \\
+    & Forest                     & 52       & 14      & 65         & 272        & 14      & 71         \\
+    & Open firm ground           & 19       & 5       & 98         & 134        & 7       & 39         \\
+    & Peatland                   & 301      & 81      & 130        & 1451       & 77      & 145        \\
+DMK & deep (\textgreater 100 cm) & 94       & 25      & 100        & 659        & 35      & 219        \\
+    & shallow (\textless 100 cm) & 6        & 2       & 50         & 839        & 44      & 82         \\
+    & unknown                    & 272      & 73      & 127        & 391        & 21      & 59         \\ \hline
 \end{tabular}
-\belowtable{Table Footnotes}
-\label{tab:latextable}
-\end{table}
-
-Or you can use markdown to create the table with booktabs = FALSE (\url{https://github.com/rstudio/rticles/issues/558\#issuecomment-1907981541}).
-
-See Table \ref{tab:test}.
-
-\begin{table}
-
-\caption{\label{tab:test}My caption}
-\centering
-\begin{tabular}[t]{l|r|r|r}
-\hline
-  & mpg & cyl & disp\\
-\hline
-Mazda RX4 & 21.0 & 6 & 160\\
-\hline
-Mazda RX4 Wag & 21.0 & 6 & 160\\
-\hline
-Datsun 710 & 22.8 & 4 & 108\\
-\hline
-\end{tabular}
+\label{tab:depthsByClass}
 \end{table}

### 2024-10-30 ms/tables.R

@@ -1,6 +1,6 @@
-library(tidyverse)
-
 # Predictors table ####
+
+library(tidyverse)
 
 rast <- terra::rast("output/predictors.tif")
 names(rast)
@@ -33,6 +33,44 @@ tribble(
   "DTW160000", "Depth-to-water index, flow initiation area of 16 ha",
   "DMK", "DMK peat depth class, categorical with 3 levels") |> 
   write_csv("ms/tables/predictors.csv")
+
+# Summary cell depths by attributes ####
+
+library(tidyverse)
+
+skrim <- sf::st_read("output/Skrim/modeling.gpkg", layer="dataframe") |> 
+  sf::st_drop_geometry()
+orskog <- sf::st_read("output/modeling.gpkg", layer="dataframe") |> 
+  sf::st_drop_geometry()
+sites <- bind_rows(skrim = skrim, orskog = orskog, .id = "site") |> 
+  mutate(dmkdepth = case_when(
+    dmkdepth == "djup myr" ~ "djup",
+    dmkdepth == "grunn myr" ~ "grunn",
+    TRUE ~ dmkdepth
+  ))
+overall <- sites |> 
+  group_by(site) |> 
+  summarize(n = n(), depth_cm = mean(depth_cm)) |> 
+  pivot_wider(names_from = 'site', values_from = c(n, depth_cm), 
+              names_vary = "slowest")
+byAR5 <- sites |> 
+  group_by(site, ar5cover) |> 
+  summarize(n = n(), depth_cm = mean(depth_cm)) |> 
+  mutate(prop = n / sum(n)) |>
+  filter(!(ar5cover %in% c("11", "12"))) |>
+  pivot_wider(names_from = 'site', values_from = c(n, prop, depth_cm), 
+              names_vary = "slowest")
+byDMK <- sites |> 
+  group_by(site, dmkdepth) |> 
+  summarize(n = n(), depth_cm = mean(depth_cm)) |> 
+  mutate(prop = n / sum(n)) |>
+  pivot_wider(names_from = 'site', values_from = c(n, prop, depth_cm), 
+              names_vary = "slowest")
+bind_rows(overall, byAR5, byDMK) |> 
+  select(ar5cover, dmkdepth, n_skrim, prop_skrim, depth_cm_skrim,
+         n_orskog, prop_orskog, depth_cm_orskog) |> 
+  write_csv("ms/tables/celldepth-attribute.csv")
+
 
 # sessionInfo ####
 

### 2024-10-31 R/Skrim/concatenateDepthMeasurements.R
ms: right-censored probes

@@ -12,6 +12,9 @@ probe <- pivot_longer(probe, cols = starts_with("probe"),
                       values_to = "depth_cm")
 filter(probe, is.na(depth_cm)) 
 
+probe |> 
+  filter(grepl("plus|\\+", note))
+
 # GPR data ####
 
 ## Calibrate wave velocity ####

### 2024-10-31 R/concatenateDepthMeasurements.R

@@ -23,6 +23,9 @@ probeclean <- probe |>
   filter(!(from == "dgps2" & ptname == 276 & comment == "fyllmasse, ikke torv")) |> 
   filter(!(from == "dgps2" & ptname == 251 & comment == "road, fyllmasse paa veg"))
 
+probe |> 
+  filter(grepl("plus|\\+", comment, ignore.case = TRUE))
+
 arrange(probeclean, desc(HRMS))
 
 # GPR data ####

### 2024-10-31 ms/ms.Rmd
ms: right-censored probes
use child document for table

@@
 Probe locations were adjusted up to \unit{20\,cm} if the base of the peat column seemed to be blocked by an obvious artifact.
+A single probe measurement was right-censored because the peat column was deeper than the probe.
 For each occurrence sample location, we recorded the presence or absence of peatland -- primarily by digging and examining the top \unit{20\,cm} of soil (where this was possible).
@@
 At each location we measured peat depth three times by manual probing, with probe locations spaced approximately \unit{2.5\,m} apart.
+For a tiny fraction of these measurements (5 total), it was not possible to reach the base of the peat column, and a right-censored depth was recorded. 
 In areas with dense sampling locations, we also measured peat depth with GPR along snaking transects passing through the centers of the sampling cells (seven transects, \unit{6.2\,km} total length).
@@
 
-\begin{table}[tbp]
-\caption{Mean peat depth (cm) in \unit{10\,m} cells at Skrimfjella and Ørskogfjellet. The cells are also shown stratified by AR5 land class and DMK peat depth class.}
-\begin{tabular}{llllllll}
-\hline
-    &                            & \multicolumn{3}{c}{Skrimfjella} & \multicolumn{3}{c}{Ørskogfjellet} \\ \cline{3-8} 
-    &                            & n        & \%      & depth      & n          & \%      & depth      \\ \hline
-    &                            & 372      & 100     & 119        & 1889       & 100     & 125        \\
-AR5 & Fully cultivated           &          &         &            & 21         & 1       & 30         \\
-    & Forest                     & 52       & 14      & 65         & 272        & 14      & 71         \\
-    & Open firm ground           & 19       & 5       & 98         & 134        & 7       & 39         \\
-    & Peatland                   & 301      & 81      & 130        & 1451       & 77      & 145        \\
-DMK & deep (\textgreater 100 cm) & 94       & 25      & 100        & 659        & 35      & 219        \\
-    & shallow (\textless 100 cm) & 6        & 2       & 50         & 839        & 44      & 82         \\
-    & unknown                    & 272      & 73      & 127        & 391        & 21      & 59         \\ \hline
-\end{tabular}
-\label{tab:depthsByClass}
-\end{table}
+```{r depthsByClass, echo=FALSE, results='asis'}
+depthsByClass <- readr::read_file("tables/depthsByClass.tex")
+cat(depthsByClass)
+```
 

### 2024-10-31 ms/ms.Rmd
ms: results - depthByClass description

@@
 
-We obtained depth measurements for 372 cells (XX % of mapped peatland) at Skrimfjella and 1889 cells (XX % of mapped peatland) at Ørskogfjellet.
-The mean depths of these \unit{10\,m} cells differed by only \unit{6\,cm} between Skrimfjella and Ørskogfjellet (Table \ref{tab:depthsByClass}).
+We obtained depth measurements for 372 cells at Skrimfjella (area equivalent to 2.4 % of mapped peatland) and 1878 cells at Ørskogfjellet (area equivalent to 1.2 % of mapped peatland).
+Roughly 80 % of these \unit{10\,m} cells were within mapped peatland, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
+Coverage of DMK peat depth classes was higher at Ørskogfjellet than at Skrimfjella (79 % versus 27 % of cells), and at Ørskogfjellet the *deep* and *shallow* classes showed a larger difference in measured depth.
+Overall mean peat depths were similar at Skrimfjella and Ørskogfjellet:  \unit{119\,cm} and \unit{126\,cm}, respectively.
 

### 2024-10-31 ms/ms.tex
ms: right-censored probes
ms: results - depthByClass description

@@
 At each location we measured peat depth three times by manual probing, with probe locations spaced approximately \unit{2.5\,m} apart.
+For a tiny fraction of these measurements (5 total), it was not possible to reach the base of the peat column, and a right-censored depth was recorded.
 In areas with dense sampling locations, we also measured peat depth with GPR along snaking transects passing through the centers of the sampling cells (seven transects, \unit{6.2\,km} total length).
@@
 
-We obtained depth measurements for 372 cells (XX \% of mapped peatland) at Skrimfjella and 1889 cells (XX \% of mapped peatland) at Ørskogfjellet.
-The mean depths of these \unit{10\,m} cells differed by only \unit{6\,cm} between Skrimfjella and Ørskogfjellet (Table \ref{tab:depthsByClass}).
+We obtained depth measurements for 372 cells at Skrimfjella (area equivalent to 2.4 \% of mapped peatland) and 1878 cells at Ørskogfjellet (area equivalent to 1.2 \% of mapped peatland).
+Roughly 80 \% of these \unit{10\,m} cells were within mapped peatland, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
+Coverage of DMK peat depth classes was higher at Ørskogfjellet than at Skrimfjella (79 \% versus 27 \% of cells), and at Ørskogfjellet the \emph{deep} and \emph{shallow} classes showed a larger difference in measured depth.
+Overall mean peat depths were similar at Skrimfjella and Ørskogfjellet: \unit{119\,cm} and \unit{126\,cm}, respectively.
 
@@
     &                            & n        & \%      & depth      & n          & \%      & depth      \\ \hline
-    &                            & 372      & 100     & 119        & 1889       & 100     & 125        \\
-AR5 & Fully cultivated           &          &         &            & 21         & 1       & 30         \\
-    & Forest                     & 52       & 14      & 65         & 272        & 14      & 71         \\
-    & Open firm ground           & 19       & 5       & 98         & 134        & 7       & 39         \\
+    &                            & 372      & 100     & 119        & 1878       & 100     & 126        \\
+AR5 & Agricultural               &          &         &            & 21         & 1       & 30         \\
+    & Forest                     & 52       & 14      & 65         & 272        & 15      & 71         \\
+    & Open upland                & 19       & 5       & 98         & 134        & 7       & 39         \\
     & Peatland                   & 301      & 81      & 130        & 1451       & 77      & 145        \\
 DMK & deep (\textgreater 100 cm) & 94       & 25      & 100        & 659        & 35      & 219        \\
-    & shallow (\textless 100 cm) & 6        & 2       & 50         & 839        & 44      & 82         \\
-    & unknown                    & 272      & 73      & 127        & 391        & 21      & 59         \\ \hline
+    & shallow (\textless 100 cm) & 6        & 2       & 50         & 838        & 45      & 82         \\
+    & unknown                    & 272      & 73      & 127        & 381        & 20      & 60         \\ \hline
 \end{tabular}

### 2024-10-31 R/Skrim/modelDepth.R

@@ -22,6 +22,7 @@ predictorspts <- predictors |>
 predpts <- predictorspts[predictivedomain,] |> 
   st_transform(st_crs(frame))
 st_crs(frame) == st_crs(predpts)
+nrow(frame)/nrow(predpts)
 
 set.seed(123)
 predptssample <- slice_sample(predpts, n=1e3)

### 2024-10-31 ms/tables.R

@@ -47,7 +47,8 @@ sites <- bind_rows(skrim = skrim, orskog = orskog, .id = "site") |>
     dmkdepth == "djup myr" ~ "djup",
     dmkdepth == "grunn myr" ~ "grunn",
     TRUE ~ dmkdepth
-  ))
+  )) |> 
+  filter(!(ar5cover %in% c("11", "12")))
 overall <- sites |> 
   group_by(site) |> 
   summarize(n = n(), depth_cm = mean(depth_cm)) |> 
@@ -57,7 +58,6 @@ byAR5 <- sites |>
   group_by(site, ar5cover) |> 
   summarize(n = n(), depth_cm = mean(depth_cm)) |> 
   mutate(prop = n / sum(n)) |>
-  filter(!(ar5cover %in% c("11", "12"))) |>
   pivot_wider(names_from = 'site', values_from = c(n, prop, depth_cm), 
               names_vary = "slowest")
 byDMK <- sites |> 
@@ -70,7 +70,6 @@ bind_rows(overall, byAR5, byDMK) |>
   select(ar5cover, dmkdepth, n_skrim, prop_skrim, depth_cm_skrim,
          n_orskog, prop_orskog, depth_cm_orskog) |> 
   write_csv("ms/tables/celldepth-attribute.csv")
-
 
 # sessionInfo ####
 

### 2024-10-31 ms/ms.Rmd
ms: results - first description of model metrics Term consistency: predictors/variables

@@
 
-In addition to the peat depth locations, we had another arm of our sampling design for measuring peatland occurrence, as binary variable.
+In addition to the peat depth locations, we had another arm of our sampling design for measuring peatland occurrence, as a binary variable.
 We wanted to measure peatland occurrence outside of mapped mire areas because the AR5 dataset is known to underestimate peatland coverage [especially in forests, @brynLandCoverNorway2018], and because airborne radiometrics may help identify unmapped peatland [@gatisMappingUplandPeat2019; @olearyDigitalSoilMapping2022].
@@
 We defined a sequence of sample sizes (50--500) and for ten replicate samples at each size [drawn by conditioned latin hypercube sampling, @minasnyConditionedLatinHypercube2006; @roudierClhsPackageConditioned2011], we calculated the mean Kullback-Leibler divergence between sample and population distributions [@maloneMethodsImproveUtility2019; @sauretteDivergenceMetricsDetermining2023].
-The variables in the divergence calculation were terrain slope and four radiometrics: potassium, thorium, uranium, and total count.
+The predictors in the divergence calculation were terrain slope and four radiometrics: potassium, thorium, uranium, and total count.
 Next, we fitted a asymptotic regression of mean divergence on sample size, and identified the sample size at which the curve reached 95 % of the fitted asymptote.
@@
 Feature space coverage sampling works best when all dimensions are important predictors of the outcome [@wadouxSamplingDesignOptimization2019], and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
-The radiometric variables were downscaled to \unit{10\,m} resolution with cubic B-spline resampling in QGIS [v.3, @QGISsoftware].
+The radiometric predictors were downscaled to \unit{10\,m} resolution with cubic B-spline resampling in QGIS [v.3, @QGISsoftware].
 We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
@@
 
-For terrain-derived variables, we obtained \unit{1\,m} resolution DTMs from the Norwegian Mapping Authority.
+For terrain-derived predictors, we obtained \unit{1\,m} resolution DTMs from the Norwegian Mapping Authority.
 The DTM for Skrimfjella was produced from airborne laser scanning surveys in 2015 and 2022, with laser point density of \unit{5\,pts\,m^{-2}}.
@@
 We used the *terra* R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
-These variables were derived at two scales to produce different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
+These were derived at two scales to produce eight different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
 This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties [@millerImpactMultiscalePredictor2015; @dornikOptimalScalingPredictors2022; @newmanAssessingSpatiallyHeterogeneous2023].
@@
 
-  1. only DMK peat depth class (2 binary indicator variables from 3 levels)
+  1. only DMK peat depth class (2 indicator variables from 3 levels)
   2. only terrain (21 continuous variables)
@@
   4. terrain and radiometric predictors (25 continuous variables)
-  5. all predictors (26 variables)
+  5. all predictors (27 variables)
 
@@
 
+None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \@ref(fig:modelMetrics)).
+For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an *R^2^* of 0.34, and an RMSE of \unit{78\,cm}.
+For Ørskogfjellet, the same values were 0.39, 0.33, and \unit{75\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
+
+```{r modelMetrics, echo = FALSE, fig.cap = "Performance of models with different predictor configurations, evaluated via spatial cross-validation.
+Parentheses denote the number of variables in each predictor configuration, and point estimates are shown +/- their standard error."}
+knitr::include_graphics("figures/modelmetrics.pdf")
+```
+
 # Discussion

### 2024-10-31 ms/ms.tex
ms: results - first description of model metrics Term consistency: predictors/variables

@@
 
-In addition to the peat depth locations, we had another arm of our sampling design for measuring peatland occurrence, as binary variable.
+In addition to the peat depth locations, we had another arm of our sampling design for measuring peatland occurrence, as a binary variable.
 We wanted to measure peatland occurrence outside of mapped mire areas because the AR5 dataset is known to underestimate peatland coverage \citep[especially in forests,][]{brynLandCoverNorway2018}, and because airborne radiometrics may help identify unmapped peatland \citep{gatisMappingUplandPeat2019, olearyDigitalSoilMapping2022}.
@@
 We defined a sequence of sample sizes (50--500) and for ten replicate samples at each size \citep[drawn by conditioned latin hypercube sampling,][]{minasnyConditionedLatinHypercube2006, roudierClhsPackageConditioned2011}, we calculated the mean Kullback-Leibler divergence between sample and population distributions \citep{maloneMethodsImproveUtility2019, sauretteDivergenceMetricsDetermining2023}.
-The variables in the divergence calculation were terrain slope and four radiometrics: potassium, thorium, uranium, and total count.
+The predictors in the divergence calculation were terrain slope and four radiometrics: potassium, thorium, uranium, and total count.
 Next, we fitted a asymptotic regression of mean divergence on sample size, and identified the sample size at which the curve reached 95 \% of the fitted asymptote.
@@
 Feature space coverage sampling works best when all dimensions are important predictors of the outcome \citep{wadouxSamplingDesignOptimization2019}, and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
-The radiometric variables were downscaled to \unit{10\,m} resolution with cubic B-spline resampling in QGIS \citep[v.3,][]{QGISsoftware}.
+The radiometric predictors were downscaled to \unit{10\,m} resolution with cubic B-spline resampling in QGIS \citep[v.3,][]{QGISsoftware}.
 We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
@@
 We used the \emph{terra} R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
-These variables were derived at two scales to produce different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
+These were derived at two scales to produce eight different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
 This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, dornikOptimalScalingPredictors2022, newmanAssessingSpatiallyHeterogeneous2023}.
@@
 \item
-  only DMK peat depth class (2 binary indicator variables from 3 levels)
+  only DMK peat depth class (2 indicator variables from 3 levels)
 \item
@@
 \item
-  all predictors (26 variables)
+  all predictors (27 variables)
 \end{enumerate}
@@
 \end{table}
+
+None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \ref{fig:modelMetrics}).
+For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an \emph{R\textsuperscript{2}} of 0.34, and an RMSE of \unit{78\,cm}.
+For Ørskogfjellet, the same values were 0.39, 0.33, and \unit{75\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
+
+\begin{figure}
+\centering
+\includegraphics{figures/modelmetrics.pdf}
+\caption{\label{fig:modelMetrics}Performance of models with different predictor configurations, evaluated via spatial cross-validation.
+Parentheses denote the number of variables in each predictor configuration, and point estimates are shown +/- their standard error.}
+\end{figure}
 

### 2024-10-31 ms/figures.R

@@ -208,17 +208,17 @@ plotting <- bind_rows(orskog = orskog, skrim = skrim, .id = 'site') %>%
       .metric == "rsq" ~ "R-squared",
       .metric == "rmse" ~ "RMSE"),
     model = case_when(
-      model == "DMK" ~ "DMK class (1)",
+      model == "DMK" ~ "DMK class (2)",
       model == "Terrain" ~ "terrain (21)",
-      model == "TerrainDMK" ~ "terrain + DMK class (22)",
+      model == "TerrainDMK" ~ "terrain + DMK class (23)",
       model == "RadiometricTerrain" ~ "radiometric + terrain (25)",
-      model == "RadiometricTerrainDMK" ~ "all predictors (26)"),
+      model == "RadiometricTerrainDMK" ~ "all predictors (27)"),
     model = fct_relevel(model,
-                        "DMK class (1)",
+                        "DMK class (2)",
                         "terrain (21)",
-                        "terrain + DMK class (22)",
+                        "terrain + DMK class (23)",
                         "radiometric + terrain (25)",
-                        "all predictors (26)"))
+                        "all predictors (27)"))
 str(plotting)
 
 g1 <- ggplot(plotting) +

### 2024-11-02 ms/ms.Rmd
ms: results - modelmetrics and variable importance Variable importance figure adapted to show correlations

@@
 For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an *R^2^* of 0.34, and an RMSE of \unit{78\,cm}.
-For Ørskogfjellet, the same values were 0.39, 0.33, and \unit{75\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
+For Ørskogfjellet the same values were 0.39, 0.33, and \unit{75\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
 
@@
 knitr::include_graphics("figures/modelmetrics.pdf")
+```
+
+For Skrimfjella, the best predictor configuration was *all predictors*, followed closely by *terrain and radiometric*.
+The performance gap between the *terrain and DMK* configuration and the *terrain-only* configuration was similarly small.
+Compared to the *terrain-only* configuration, the *terrain and radiometric* configuration improved concordance correlation by 0.04, R^2^ by 0.04, and RMSE by \unit{2\,cm}.
+*DMK class* alone -- although calibrated to measured depths -- was a very poor predictor of peat depth, with a concordance correlation coefficient of 0.008.
+
+For Ørskogfjellet, the best predictor configuration was *terrain and DMK*, followed by *terrain-only*.
+Adding radiometric predictors to these configurations worsened model performance, especially in terms of concordance correlation and R^2^.
+By itself, *DMK class* produced a concordance correlation coefficient of 0.17, but the worst RMSE of any model: \unit{98\,cm}.
+
+For the purpose of model interpretation, the *all predictors* configuration for Skrimfjella was reduced from 27 variables to 11 non-collinear variables.
+Similarly, the *terrain and DMK* configuration for Ørskogfjellet was reduced from 23 variables to 11 non-collinear variables.
+At both sites, *elevation* and *MRVBF* were important predictors (Fig. \@ref(fig:varImp)).
+At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet *elevation* was more important than *MRVBF*.
+*DMK* was also important -- the shallow class in particular -- but only at Ørskogfjellet.
+Some realizations of the hydrological predictors *TWI* and *DTW* showed considerable importance, while others showed little -- with no consistency between sites.
+For example, *TWI5m*, *DTW40000*, and *DTW2500* rounded out the top five most important variables at Skrimfjella, while *TWI20m* and *TWI50m* were least important, other than *DMK*.
+At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices *slope*, *TRI*, *TPI*, and *roughness*.
+The radiometric variable *radU* showed moderate importance at Skrimfjella.
+
+```{r varImp, echo = FALSE, fig.cap = "Global variable importance in the best-performing models at Ørskogfjellet (a) and Skrimfjella (b), as measured by three different metrics.
+Variables removed due to collinearity are shown to the right of that with which they are most correlated."}
+knitr::include_graphics("figures/variable_importance.pdf")
 ```

### 2024-11-02 ms/ms.tex
ms: results - modelmetrics and variable importance Variable importance figure adapted to show correlations

@@
 For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an \emph{R\textsuperscript{2}} of 0.34, and an RMSE of \unit{78\,cm}.
-For Ørskogfjellet, the same values were 0.39, 0.33, and \unit{75\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
+For Ørskogfjellet the same values were 0.39, 0.33, and \unit{75\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
 
@@
 Parentheses denote the number of variables in each predictor configuration, and point estimates are shown +/- their standard error.}
+\end{figure}
+
+For Skrimfjella, the best predictor configuration was \emph{all predictors}, followed closely by \emph{terrain and radiometric}.
+The performance gap between the \emph{terrain and DMK} configuration and the \emph{terrain-only} configuration was similarly small.
+Compared to the \emph{terrain-only} configuration, the \emph{terrain and radiometric} configuration improved concordance correlation by 0.04, R\textsuperscript{2} by 0.04, and RMSE by \unit{2\,cm}.
+\emph{DMK class} alone -- although calibrated to measured depths -- was a very poor predictor of peat depth, with a concordance correlation coefficient of 0.008.
+
+For Ørskogfjellet, the best predictor configuration was \emph{terrain and DMK}, followed by \emph{terrain-only}.
+Adding radiometric predictors to these configurations worsened model performance, especially in terms of concordance correlation and R\textsuperscript{2}.
+By itself, \emph{DMK class} produced a concordance correlation coefficient of 0.17, but the worst RMSE of any model: \unit{98\,cm}.
+
+For the purpose of model interpretation, the \emph{all predictors} configuration for Skrimfjella was reduced from 27 variables to 11 non-collinear variables.
+Similarly, the \emph{terrain and DMK} configuration for Ørskogfjellet was reduced from 23 variables to 11 non-collinear variables.
+At both sites, \emph{elevation} and \emph{MRVBF} were important predictors (Fig. \ref{fig:varImp}).
+At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet \emph{elevation} was more important than \emph{MRVBF}.
+\emph{DMK} was also important -- the shallow class in particular -- but only at Ørskogfjellet.
+Some realizations of the hydrological predictors \emph{TWI} and \emph{DTW} showed considerable importance, while others showed little -- with no consistency between sites.
+For example, \emph{TWI5m}, \emph{DTW40000}, and \emph{DTW2500} rounded out the top five most important variables at Skrimfjella, while \emph{TWI20m} and \emph{TWI50m} were least important, other than \emph{DMK}.
+At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices \emph{slope}, \emph{TRI}, \emph{TPI}, and \emph{roughness}.
+The radiometric variable \emph{radU} showed moderate importance at Skrimfjella.
+
+\begin{figure}
+\centering
+\includegraphics{figures/variable_importance.pdf}
+\caption{\label{fig:varImp}Global variable importance in the best-performing models at Ørskogfjellet (a) and Skrimfjella (b), as measured by three different metrics.
+Variables removed due to collinearity are shown to the right of that with which they are most correlated.}
 \end{figure}

### 2024-11-02 R/Skrim/modelDepth.R

@@ -398,8 +398,14 @@ quantiles_RTD %>%
 
 # Feature selection ####
 
+cormat <- bake(prep(recipe_RTD, frame), frame) |> 
+  select(!depth_cm) |> 
+  cor()
+corrplot::corrplot(cormat, method = 'number', type = 'upper', diag = TRUE,
+                   order = 'hclust', number.cex = 0.5, addCoefasPercent=TRUE)
+
 recipe_RTD_uncorr <- recipe_RTD %>% 
-  step_corr(all_predictors(), threshold = 0.7, method = "pearson")
+  step_corr(all_predictors(), threshold = 0.7, method = "pearson") # Of two variables with super-cutoff correlation, that with smallest mean absolute correlation is kept
 
 workflow_RTD_uncorr <- 
   workflow() %>% 
@@ -410,6 +416,15 @@ set.seed(456)
 fit_RTD_uncorr <- 
   workflow_RTD_uncorr %>% 
   fit(frame)
+
+vars_sel <- extract_fit_engine(fit_RTD_uncorr)$forest$independent.variable.names
+vars_diff <- setdiff(colnames(cormat), vars_sel)
+vars_corr <- map_chr(vars_diff, function(x) {
+  correlations <- cormat[x, vars_sel, drop = FALSE]
+  colnames(correlations[,which.max(abs(correlations)), drop = FALSE])})
+corrKey <- tibble(removed = vars_diff, correlated = vars_corr) |> 
+  group_by(correlated) |>
+  summarize(removed = paste(removed, collapse = ", "), .groups = 'drop')
 
 ## Variable importance ####
 
@@ -453,6 +468,7 @@ bind_rows(perm.vip = import_perm_vip,
           firm = import_firm, 
           shap = import_shap,
           .id = 'type') |> 
+  left_join(corrKey, by = c("Variable" = "correlated")) |>
   readr::write_csv("output/Skrim/variable_importance.csv")
 
 ## Partial dependence plots ####

### 2024-11-02 R/modelDepth.R

@@ -398,8 +398,14 @@ quantiles_TD %>%
 
 # Feature selection ####
 
+cormat <- bake(prep(recipe_TD, frame), frame) |> 
+  select(!depth_cm) |> 
+  cor()
+corrplot::corrplot(cormat, method = 'number', type = 'upper', diag = TRUE,
+                   order = 'hclust', number.cex = 0.5, addCoefasPercent=TRUE)
+
 recipe_TD_uncorr <- recipe_TD %>% 
-  step_corr(all_predictors(), threshold = 0.7, method = "pearson")
+  step_corr(all_predictors(), threshold = 0.7, method = "pearson") # Of two variables with super-cutoff correlation, that with smallest mean absolute correlation is kept
 
 workflow_TD_uncorr <- 
   workflow() %>% 
@@ -410,6 +416,15 @@ set.seed(456)
 fit_TD_uncorr <- 
   workflow_TD_uncorr %>% 
   fit(frame)
+
+vars_sel <- extract_fit_engine(fit_TD_uncorr)$forest$independent.variable.names
+vars_diff <- setdiff(colnames(cormat), vars_sel)
+vars_corr <- map_chr(vars_diff, function(x) {
+  correlations <- cormat[x, vars_sel, drop = FALSE]
+  colnames(correlations[,which.max(abs(correlations)), drop = FALSE])})
+corrKey <- tibble(removed = vars_diff, correlated = vars_corr) |> 
+  group_by(correlated) |>
+  summarize(removed = paste(removed, collapse = ", "), .groups = 'drop')
 
 ## Variable importance ####
 
@@ -453,6 +468,7 @@ bind_rows(perm.vip = import_perm_vip,
           firm = import_firm, 
           shap = import_shap,
           .id = 'type') |> 
+  left_join(corrKey, by = c("Variable" = "correlated")) |>
   readr::write_csv("output/variable_importance.csv")
 
 ## Partial dependence plots ####

### 2024-11-02 ms/figures.R

@@ -211,13 +211,13 @@ plotting <- bind_rows(orskog = orskog, skrim = skrim, .id = 'site') %>%
       model == "DMK" ~ "DMK class (2)",
       model == "Terrain" ~ "terrain (21)",
       model == "TerrainDMK" ~ "terrain + DMK class (23)",
-      model == "RadiometricTerrain" ~ "radiometric + terrain (25)",
+      model == "RadiometricTerrain" ~ "terrain + radiometric (25)",
       model == "RadiometricTerrainDMK" ~ "all predictors (27)"),
     model = fct_relevel(model,
                         "DMK class (2)",
                         "terrain (21)",
                         "terrain + DMK class (23)",
-                        "radiometric + terrain (25)",
+                        "terrain + radiometric (25)",
                         "all predictors (27)"))
 str(plotting)
 
@@ -266,7 +266,14 @@ skrim.rank <- skrim %>%
   group_by(Variable) %>% 
   summarize(Imp.median = median(Importance))
 
-g1 <- left_join(orskog, orskog.rank, by = join_by(Variable)) %>% 
+plot.orskog <- left_join(orskog, orskog.rank, by = join_by(Variable))
+labs.orskog <- plot.orskog |> 
+  distinct(Variable, removed, Imp.median) |>
+  replace_na(list(removed = "")) |>
+  arrange(Imp.median) |> 
+  pull(removed) |> 
+  stringr::str_wrap(width = 40)
+g1 <- plot.orskog %>% 
   mutate(Variable = case_when(
     Variable == "dmkdepth_grunn.myr" ~ "DMK_shallow",
     Variable == "dmkdepth_unknown" ~ "DMK_unknown",
@@ -277,15 +284,27 @@ g1 <- left_join(orskog, orskog.rank, by = join_by(Variable)) %>%
                group = type,
                fill = type),
            position = position_dodge2()) +
+  scale_x_discrete(sec.axis = dup_axis(labels = labs.orskog)) +
   coord_flip() +
   scale_fill_discrete(labels = c(firm = "FIRM", 
                                   perm.vip = "permutation", 
                                   shap = "Shapley")) +
   theme_minimal() +
   theme(axis.title.y = element_blank(),
-        legend.title = element_blank())
+        legend.title = element_blank(),
+        axis.text.y.right = element_text(size = 8, color = "grey50"),
[…]

### 2024-11-03 ms/ms.Rmd
ms: results - pdps

@@
 
+Many of the the most important predictors in the best performing models showed non-monotonic effects on peat depth (Fig. \@ref(fig:pdps)).
+At Ørskogfjellet for example, increasing *elevation* was predictive of deeper peat up to about 75 meters above sea level, after which a further increase was predictive of shallower peat.
+At Skrimfjella the partial dependence on *elevation* had the opposite shape, with the shallowest peats predicted at intermediate elevations, around 350 meters above sea level.
+*TWI50m* at Ørskogfjellet and *DTW4000* and *radU* at Skrimfjella were other predictors that showed considerable fluctuations in their predictive effects on peat depth.
+The radiometric predictor in particular displayed an idiosyncratic effect, with a marked dip in predicted depth at intermediate values of *radU*.
+On the other hand, the partial effects of some important predictors were more straightforward.
+The partial dependence on *MRVBF* was quite similar across sites, with the deepest peats predicted in the very flattest valley bottoms.
+Also, *TWI5m* and *DTW2500* at Skrimfjella showed monotonically positive and negative predictive effects, respectively.
+
+Individual conditional expectation lines indicated some interactions between predictors (Fig. \@ref(fig:pdps)).
+For example, the magnitude of the increase in depth with *elevation* that the model expected at Ørskogfjellet was different for different observations; some depth predictions increased by only \unit{40\,cm} while others increased by more than \unit{100\,cm}, over the same elevation gain.
+Similarly, ICE lines of *radU* at Skrimfjella were non-parallel, with some locations showing monotonically increasing peat depth predictions with uranium concentration (unlike the average effect).
+Nevertheless, most ICE lines were generally parallel -- indicating that the average effects of the predictors were good representations of their overall effects.
+
+```{r pdps, echo = FALSE, fig.cap = "Partial dependence plots of the six most important variables in the best-performing models at Ørskogfjellet (a) and Skrimfjella (b).
+The average effect of the predictor on the outcome (red line) overlays the variation in the effect across observations (grey lines)."}
+knitr::include_graphics("figures/partial_dependence.pdf")
+```
+
 # Discussion

### 2024-11-03 ms/ms.tex
ms: results - pdps

@@
 
+Many of the the most important predictors in the best performing models showed non-monotonic effects on peat depth (Fig. \ref{fig:pdps}).
+At Ørskogfjellet for example, increasing \emph{elevation} was predictive of deeper peat up to about 75 meters above sea level, after which a further increase was predictive of shallower peat.
+At Skrimfjella the partial dependence on \emph{elevation} had the opposite shape, with the shallowest peats predicted at intermediate elevations, around 350 meters above sea level.
+\emph{TWI50m} at Ørskogfjellet and \emph{DTW4000} and \emph{radU} at Skrimfjella were other predictors that showed considerable fluctuations in their predictive effects on peat depth.
+The radiometric predictor in particular displayed an idiosyncratic effect, with a marked dip in predicted depth at intermediate values of \emph{radU}.
+On the other hand, the partial effects of some important predictors were more straightforward.
+The partial dependence on \emph{MRVBF} was quite similar across sites, with the deepest peats predicted in the very flattest valley bottoms.
+Also, \emph{TWI5m} and \emph{DTW2500} at Skrimfjella showed monotonically positive and negative predictive effects, respectively.
+
+Individual conditional expectation lines indicated some interactions between predictors (Fig. \ref{fig:pdps}).
+For example, the magnitude of the increase in depth with \emph{elevation} that the model expected at Ørskogfjellet was different for different observations; some depth predictions increased by only \unit{40\,cm} while others increased by more than \unit{100\,cm}, over the same elevation gain.
+Similarly, ICE lines of \emph{radU} at Skrimfjella were non-parallel, with some locations showing monotonically increasing peat depth predictions with uranium concentration (unlike the average effect).
+Nevertheless, most ICE lines were generally parallel -- indicating that the average effects of the predictors were good representations of their overall effects.
+
+\begin{figure}
+\centering
+\includegraphics{figures/partial_dependence.pdf}
+\caption{\label{fig:pdps}Partial dependence plots of the six most important variables in the best-performing models at Ørskogfjellet (a) and Skrimfjella (b).
+The average effect of the predictor on the outcome (red line) overlays the variation in the effect across observations (grey lines).}
+\end{figure}
+
 \section{Discussion}

### 2024-11-03 ms/figures.R

@@ -394,9 +394,9 @@ g2 <- ggplot(data = skrim, aes(x = .borders, y = .value, group = .id)) +
 g1 + g2 + plot_layout(ncol = 1, guides = 'collect', axes = 'collect') +
   plot_annotation(tag_levels = 'a', tag_prefix = '(', tag_suffix = ')') 
 ggsave(filename = 'partial_dependence.pdf', path = "ms/figures",
-       width =(210-30), height = (240-40), units = 'mm') #copernicus.cls page 210x240
+       width =(210-30), height = (240-60), units = 'mm') #copernicus.cls page 210x240
 # ggsave(filename = 'partial_dependence.svg', path = "ms/figures",
-#        width =(210-30), height = (240-40), units = 'mm')
+#        width =(210-30), height = (240-60), units = 'mm')
   
 # sessionInfo ####
 

### 2024-11-06 ms/ms.Rmd
ms: results - extrapolation outside mapped peatland

@@
 In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
-From the cross-validation we quantified *Root Mean Squared Error* and *Mean Absolute Error*, and we tested the strength of the correlation between predicted and observed peat depth.
+From the cross-validation we quantified *Root Mean Squared Error*.
 Second -- for Skrimfjella -- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
+In this analysis, the model's prediction of peat depth is treated as a rank index of peatland likelihood.
 Since DMK peat depth class is always undefined outside known peatlands, we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.
@@
 
-```{r modelMetrics, echo = FALSE, fig.cap = "Performance of models with different predictor configurations, evaluated via spatial cross-validation.
+```{r modelMetrics, echo = FALSE, fig.cap = "Performance of peatland depth models with different predictor configurations, evaluated via spatial cross-validation.
 Parentheses denote the number of variables in each predictor configuration, and point estimates are shown +/- their standard error."}
@@
 By itself, *DMK class* produced a concordance correlation coefficient of 0.17, but the worst RMSE of any model: \unit{98\,cm}.
+
+When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced marginally better RMSE than just assuming a constant \unit{30\,cm} depth (Fig.
+\@ref(fig:modelMetricsExtrapolation)).
+Compared to our error estimates with the full dataset, RMSE at Skrimfjella was nearly unchanged, while at Ørskogfjellet it was lower outside of mapped peatland.
+With the independent occurrence data at Skrimfjella, we found that neither the *terrain-only* nor *terrain and radiometric* configurations were able to discriminate between peat presence and absence (area under the curve of the receiver operating characteristic 0.44 and 0.52 respectively, where 0.5 indicates random guessing).
+
+```{r modelMetricsExtrapolation, echo = FALSE, fig.cap = "Performance of models that extrapolate from training data inside of mapped peatland to test data outside of mapped peatland, evaluated via spatial cross-validation.
+Parentheses denote the number of variables in each model, and point estimates are shown +/- their standard error."}
+knitr::include_graphics("figures/modelmetrics-extrapolation.pdf")
+```
 

### 2024-11-06 ms/ms.tex
ms: results - extrapolation outside mapped peatland

@@
 In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
-From the cross-validation we quantified \emph{Root Mean Squared Error} and \emph{Mean Absolute Error}, and we tested the strength of the correlation between predicted and observed peat depth.
+From the cross-validation we quantified \emph{Root Mean Squared Error}.
 Second -- for Skrimfjella -- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
+In this analysis, the model's prediction of peat depth is treated as a rank index of peatland likelihood.
 Since DMK peat depth class is always undefined outside known peatlands, we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.
@@
 \includegraphics{figures/modelmetrics.pdf}
-\caption{\label{fig:modelMetrics}Performance of models with different predictor configurations, evaluated via spatial cross-validation.
+\caption{\label{fig:modelMetrics}Performance of peatland depth models with different predictor configurations, evaluated via spatial cross-validation.
 Parentheses denote the number of variables in each predictor configuration, and point estimates are shown +/- their standard error.}
@@
 By itself, \emph{DMK class} produced a concordance correlation coefficient of 0.17, but the worst RMSE of any model: \unit{98\,cm}.
+
+When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced marginally better RMSE than just assuming a constant \unit{30\,cm} depth (Fig.
+\ref{fig:modelMetricsExtrapolation}).
+Compared to our error estimates with the full dataset, RMSE at Skrimfjella was nearly unchanged, while at Ørskogfjellet it was lower outside of mapped peatland.
+With the independent occurrence data at Skrimfjella, we found that neither the \emph{terrain-only} nor \emph{terrain and radiometric} configurations were able to discriminate between peat presence and absence (area under the curve of the receiver operating characteristic 0.44 and 0.52 respectively, where 0.5 indicates random guessing).
+
+\begin{figure}
+\centering
+\includegraphics{figures/modelmetrics-extrapolation.pdf}
+\caption{\label{fig:modelMetricsExtrapolation}Performance of models that extrapolate from training data inside of mapped peatland to test data outside of mapped peatland, evaluated via spatial cross-validation.
+Parentheses denote the number of variables in each model, and point estimates are shown +/- their standard error.}
+\end{figure}
 

### 2024-11-06 R/Skrim/modelDepth.R

@@ -285,14 +285,25 @@ cor.test(T_extrap$depth_cm, T_extrap$.pred)
 
 ### preds: none (assume 30cm) ####
 
-frame |> 
+fit_null_extrap <- frame |> 
   st_drop_geometry() |> 
   filter(ar5cover != 60) |> 
   mutate(.pred = 30,
          .resid = .pred - depth_cm,
          absoluteError = abs(.resid)) |>
   summarise(mae = mean(absoluteError),
-            rmse = sqrt(mean(.resid^2))) 
+            rmse = sqrt(mean(.resid^2))) |> 
+  pivot_longer(cols = everything(), names_to = ".metric", values_to = "mean")
+fit_null_extrap
+
+### collect metrics ####
+
+modelmetrics.extrap <- bind_rows(
+  RadiometricTerrain = collect_metrics(fit_RT_extrap),
+  Terrain = collect_metrics(fit_T_extrap),
+  null = fit_null_extrap,
+  .id = "model")
+readr::write_csv(modelmetrics.extrap, "output/Skrim/modelmetrics-extrapolation.csv")
 
 # Uncertainty with CV ####
 

### 2024-11-06 R/modelDepth.R

@@ -285,14 +285,25 @@ cor.test(T_extrap$depth_cm, T_extrap$.pred)
 
 ### preds: none (assume 30cm) ####
 
-frame |> 
+fit_null_extrap <- frame |> 
   st_drop_geometry() |> 
   filter(ar5cover != 60) |> 
   mutate(.pred = 30,
          .resid = .pred - depth_cm,
          absoluteError = abs(.resid)) |>
   summarise(mae = mean(absoluteError),
-            rmse = sqrt(mean(.resid^2))) 
+            rmse = sqrt(mean(.resid^2))) |> 
+  pivot_longer(cols = everything(), names_to = ".metric", values_to = "mean")
+fit_null_extrap
+
+### collect metrics ####
+
+modelmetrics.extrap <- bind_rows(
+  RadiometricTerrain = collect_metrics(fit_RT_extrap),
+  Terrain = collect_metrics(fit_T_extrap),
+  null = fit_null_extrap,
+  .id = "model")
+readr::write_csv(modelmetrics.extrap, "output/modelmetrics-extrapolation.csv")
 
 # Uncertainty with CV ####
 

### 2024-11-06 ms/figures.R

@@ -249,6 +249,55 @@ ggsave(filename = 'modelmetrics.pdf', path = "ms/figures",
 # ggsave(filename = 'modelmetrics.svg', path = "ms/figures",
 #        width =210-30, height = (240-40)/2.5, units = 'mm')
 
+
+# Model metrics extrapolation####
+
+library(tidyverse)
+
+orskog <- read_csv("output/modelmetrics-extrapolation.csv") 
+skrim <- read_csv("output/Skrim/modelmetrics-extrapolation.csv")
+plotting <- bind_rows(orskog = orskog, skrim = skrim, .id = 'site') %>% 
+  filter(.metric != 'mae') %>% 
+  mutate(
+    site = fct_relevel(site, "orskog"),
+    metric = case_when(
+      .metric == "rmse" ~ "RMSE"),
+    model = case_when(
+      model == "null" ~ "null, assume 30 cm",
+      model == "Terrain" ~ "terrain (21)",
+      model == "RadiometricTerrain" ~ "terrain + radiometric (25)"),
+    model = fct_relevel(model,
+                        "null, assume 30 cm",
+                        "terrain (21)",
+                        "terrain + radiometric (25)"))
+str(plotting)
+
+g1 <- ggplot(plotting) +
+  geom_linerange(aes(y = model, 
+                     x = mean, 
+                     xmin = mean - std_err, 
+                     xmax = mean + std_err, 
+                     color = site),
+                 position = position_dodge2(width= 0.3, reverse = TRUE)) +
+  geom_point(aes(y = model, x = mean, color = site),
+             position = position_dodge2(width= 0.3, reverse = TRUE)) +
+  geom_text(aes(y = model, x = mean, label = signif(mean, 2), group = site),
+            position = position_dodge2(width= 1, reverse = TRUE),
+            color = "grey50", size = 2.5) +
+  facet_wrap(~metric, nrow = 1, scales = "free_x", axes ="margins") +
+  scale_color_discrete(labels = c(orskog = "\u00D8rskogfjellet", 
+                                  skrim = "Skrimfjella")) +
+  theme_bw() +
+  theme(axis.title = element_blank(),
+        legend.title = element_blank(),
+        panel.grid = element_blank(),
+        axis.ticks.y = element_blank(),
+        legend.position = "inside",
+        legend.position.inside = c(-0.4, 1.06),
+        legend.key.spacing.y = unit(-2, "mm"),)
[…]

### 2024-11-06 ms/ms.Rmd
ms: remove template appendices

@@
 This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, © Kartverket, (2) *Norge digitalt* license, Norwegian Institute of Bioeconomy Research (NIBIO), © Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway (NGU).
-appendix: |
-  \section{For submission}
-  "Appendices: all material required to understand the essential aspects of the paper
-  such as experimental methods, data, and interpretation 
-  should preferably be included in the main text. 
-  Additional figures, tables, as well as technical and theoretical developments 
-  which are not critical to support the conclusion of the paper,
-  but which provide extra detail and/or support useful for experts in the field
-  and whose inclusion in the main text would disrupt the flow of descriptions or demonstrations 
-  may be presented as appendices.
-  These should be labelled with capital letters: Appendix A, Appendix B etc.
-  Equations, figures and tables should be numbered as (A1), Fig. B5 or Table C6, respectively.
-  Please keep in mind that appendices are part of the manuscript
-  whereas supplements (see below) are published along with the manuscript."
-
-  \section{Figures and tables in appendices}
-  Please also sort the appendix figures and appendix tables into the respective appendix sections.
-  They will be correctly named automatically.
-
-  \section{Copernicus from Rmarkdown}
-  **Please note:** Per [their guidelines](https://publications.copernicus.org/for_authors/manuscript_preparation.html),
-  Copernicus does not support additional \LaTeX{} packages 
-  or new \LaTeX{} commands than those defined in their `.cls` file. 
-  This means that you cannot add any extra dependencies 
-  and a warning will be thrown if so.
-  **Important**: Always double-check with the official manuscript preparation guidelines 
-  at [https://publications.copernicus.org/for_authors/manuscript_preparation.html](https://publications.copernicus.org/for_authors/manuscript_preparation.html),
-  especially the sections "Technical instructions for LaTeX" and "Manuscript composition". 
-  Please contact Daniel Nüst, `daniel.nuest@uni-muenster.de`, with any problems.
 output:

### 2024-11-06 ms/ms.tex
ms: remove template appendices

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-\section{For submission}
-
-``Appendices: all material required to understand the essential aspects of the paper
-such as experimental methods, data, and interpretation
-should preferably be included in the main text.
-Additional figures, tables, as well as technical and theoretical developments
-which are not critical to support the conclusion of the paper,
-but which provide extra detail and/or support useful for experts in the field
-and whose inclusion in the main text would disrupt the flow of descriptions or demonstrations
-may be presented as appendices.
-These should be labelled with capital letters: Appendix A, Appendix B etc.
-Equations, figures and tables should be numbered as (A1), Fig. B5 or Table C6, respectively.
-Please keep in mind that appendices are part of the manuscript
-whereas supplements (see below) are published along with the manuscript.''
-
-\section{Figures and tables in appendices}
-
-Please also sort the appendix figures and appendix tables into the respective appendix sections.
-They will be correctly named automatically.
-
-\section{Copernicus from Rmarkdown}
-
-\textbf{Please note:} Per \href{https://publications.copernicus.org/for_authors/manuscript_preparation.html}{their guidelines},
-Copernicus does not support additional \LaTeX{} packages
-or new \LaTeX{} commands than those defined in their \texttt{.cls} file.
-This means that you cannot add any extra dependencies
-and a warning will be thrown if so.
-\textbf{Important}: Always double-check with the official manuscript preparation guidelines
-at \url{https://publications.copernicus.org/for_authors/manuscript_preparation.html},
-especially the sections ``Technical instructions for LaTeX'' and ``Manuscript composition''.
-Please contact Daniel Nüst, \texttt{daniel.nuest@uni-muenster.de}, with any problems.
-\noappendix
 

### 2024-11-20 ms/ms.Rmd
ms: add intro

--- 2024-11-06 ms/ms.Rmd
@@
 This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, © Kartverket, (2) *Norge digitalt* license, Norwegian Institute of Bioeconomy Research (NIBIO), © Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway (NGU).
-appendix: |
-  \section{For submission}
-  "Appendices: all material required to understand the essential aspects of the paper
-  such as experimental methods, data, and interpretation 
-  should preferably be included in the main text. 
-  Additional figures, tables, as well as technical and theoretical developments 
-  which are not critical to support the conclusion of the paper,
-  but which provide extra detail and/or support useful for experts in the field
-  and whose inclusion in the main text would disrupt the flow of descriptions or demonstrations 
-  may be presented as appendices.
-  These should be labelled with capital letters: Appendix A, Appendix B etc.
-  Equations, figures and tables should be numbered as (A1), Fig. B5 or Table C6, respectively.
-  Please keep in mind that appendices are part of the manuscript
-  whereas supplements (see below) are published along with the manuscript."
-
-  \section{Figures and tables in appendices}
-  Please also sort the appendix figures and appendix tables into the respective appendix sections.
-  They will be correctly named automatically.
-
-  \section{Copernicus from Rmarkdown}
-  **Please note:** Per [their guidelines](https://publications.copernicus.org/for_authors/manuscript_preparation.html),
-  Copernicus does not support additional \LaTeX{} packages 
-  or new \LaTeX{} commands than those defined in their `.cls` file. 
-  This means that you cannot add any extra dependencies 
-  and a warning will be thrown if so.
-  **Important**: Always double-check with the official manuscript preparation guidelines 
-  at [https://publications.copernicus.org/for_authors/manuscript_preparation.html](https://publications.copernicus.org/for_authors/manuscript_preparation.html),
-  especially the sections "Technical instructions for LaTeX" and "Manuscript composition". 
-  Please contact Daniel Nüst, `daniel.nuest@uni-muenster.de`, with any problems.
+Large language models have been used during the drafting and editing of this manuscript, with author oversight.
+We maintain full responsibility for the scientific output, as per the European Commission's *Living guidelines on the responsible use of generative AI in research*.
 output:
@@
 
-Introduction text goes here.
-Read Gatis et al. [-@gatisMappingUplandPeat2019] and related work [@minasnyDigitalMappingPeatlands2019].
+Peat soils are a terrestrial carbon stock of global importance. 
+They store about 30 % of global soil carbon despite covering only 2-3 % of Earth's land [@xuPEATMAPRefiningEstimates2018; @friedlingsteinGlobalCarbonBudget2020; @unepGlobalPeatlandsAssessment2022]. 
+In other words, they are extremely carbon dense -- denser than any other ecosystem per square meter [@temminkRecoveringWetlandBiogeomorphic2022]. 
+This makes peatlands crucial to climate change mitigation. 
+Intact peatlands sequester carbon and produce a negative temperature forcing, overall [@joostenRolePeatlandsClimate2016]. 
+When disturbed, often by conversion to another land use, they can produce large greenhouse gas emissions [@maGloballyRobustRelationship2022].
+
+One of the keys to the areal density of peatland carbon stocks lies in the third dimension: their depth. 
+Peat soils range from zero to over ten meters deep, so deep peats comprise a large volume. 
+Their depth results from the accumulation of organic matter over thousands of years [@loiselDatabaseSynthesisNorthern2014; @joostenRolePeatlandsClimate2016]. 
+In the anoxic and acidic conditions created by a high water table, plant material decay is slightly outweighed by new growth, and the surplus carbon is laid down as peat. 
+There it remains sequestered as long as the water table stays high and prevents oxidation.
+
+Peatlands are unevenly distributed globally, and in regions with high cover they are frequently converted to human land use [@unepGlobalPeatlandsAssessment2022].
+They are attractive for agriculture and forestry because they are flat, treeless, and have developed soils.
+However, other land uses also displace peatlands. 
+In Norway, where 9% of the land area is peatland [@brynLandCoverNorway2018], construction and development have become important drivers of loss, after lawmakers restricted conversion to forest and farmland in recent decades.
+
+The spatial distribution of peat depth is often overlooked in land use planning and carbon accounting because peat depth is not mapped with sufficient coverage, resolution, or accuracy.
+Maps are crucial because they link high-level targets to specific management decisions, unlike spatially aggregated estimates [@oecdOECDEnvironmentalPerformance2022].
+For example, Norwegian regulations restrict conversion to farmland more strictly where peat depth exceeds \unit{1\,m}, and the distribution of peat depth on a farmer's property determines whether he is allowed to convert peatland (Forskrift om nydyrking, 1997, § 5a).
+Maps also make it possible to quantify the effect of specific management decisions and thereby understand how local outcomes contribute to regional and national outcomes [@oecdOECDEnvironmentalPerformance2022].
+
+Measuring peat depth on the ground is straightforward, and a field survey can map a small area at low cost. 
+However, surveying large areas is impractical when depth varies over short distances -- as in most peatlands. 
+A complementary approach from soil science is digital soil mapping (DSM). 
+DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest. 
+This approach has grown important with the availability of remotely sensed data and the advancement of methods for identifying pattern, especially through machine learning [@minasnyDigitalMappingPeatlands2019; @wadouxMachineLearningDigital2020].
+
+The crux for DSM of peat depth are the relationships between peat depth and the other, mapped variables. 
+For DSM to work, these relationships must be strong and consistent over the area of interest.
+They can be purely correlative rather than causal, but mechanistic relationships are stronger and more consistent. 
+The *scorpan* framework for DSM suggests seven predictor classes to explore: other soil properties, climate, organisms, relief (topography), parent material, age, and spatial position [@mcbratneyDigitalSoilMapping2003].
+
+The most practical and widespread *scorpan* factors for peat depth are relief and spatial position. 
+Spatial position is unique because it is always known (with varying accuracy). 
+However, the short range of spatial autocorrelation in peat depth limits its mapping value [@henglGenericFrameworkSpatial2004]. 
+Relief, or topography, is widely and accurately mapped in digital terrain models. 
+For example, most of mainland Norway has at least one elevation measurement per square meter from airborne LiDAR surveys. 
+Moreover, mechanisms of peat formation are linked to topography. 
+For example, a steep slope is unlikely to have a high water table, thereby limiting accumulation.
+
+Studies have consistently shown relationships between peat depth and topography, although the specific patterns vary. 
+One of the most robust relationships is a negative correlation between depth and terrain slope [e.g. @holdenEstimatingCarbonStock2011; @parryMethodModellingPeat2012; @gatisMappingUplandPeat2019]. 
+Peat depth also changes with elevation in many contexts, but with inconsistent directionality [e.g.
+@holdenEstimatingCarbonStock2011; @parryMethodModellingPeat2012; @rudiyantoDigitalMappingCosteffective2016; @rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024].
+Also more complex derivations of topography, such as the Topographic Wetness Index (TWI) and the Multi-Resolution Valley Bottom Flatness (MRVBF) index, have shown associations with peat depth in some studies [e.g.
+@rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024].
+Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic variables and the resolution of the peat depth analysis.
+
+Another set of variables related to peat depth are measurements of natural radioactivity from the ground surface. 
+Gamma-ray spectrometry can survey the activity (decay counts per second) from radioactive isotopes in the earth's crust: potassium-40, uranium-238, and thorium-232 [@reinhardtGammaraySpectrometryVersatile2019].
+These exist in bedrock (and mineral soils) and a peat overburden attenuates the radiation intensity at the surface. 
+The degree of attenuation relates to properties of the overburden, especially composition and depth. 
+Deep soil with high water content attenuates most [@beamishGammaRayAttenuation2013; @reinhardtGammaraySpectrometryVersatile2019]. 
+Thus, gamma-ray radiometrics integrate the *scorpan* factors soil, parent material, and age.
+
+Although theory suggests that one meter of peat may fully attenuate radiometric signal [@beamishGammaRayAttenuation2013; @reinhardtGammaraySpectrometryVersatile2019], empirical investigations show that the association between peat depth and radioactivity can extend beyond the first meter [@gatisMappingUplandPeat2019; @kogantiMappingPeatDepth2023].
+Radiometric data are increasingly available over large areas, presenting a big opportunity insofar as they are predictive. 
+Airborne surveys are used for various applications, and some countries have high spatial coverage of such data  [@minasnyDigitalMappingPeatlands2019; @baranwalAirborneGeophysicalSurveys2020].
+
+The effectiveness of DSM depends not only on the methods and data, but also on the characteristics of the mapping area. 
+Norway may be instructive in this respect because its peatlands vary widely across climates and topographies.
+There are many mire massif types, with different hydrology, formation, and development -- from topogeneous or soligeneous fens to blanket or raised bogs [@lyngstadBeskrivelserAvTorvmassivenheter2023].
+These types have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric variables will vary by landscape.
+
+The little we know about the depths of Norwegian peatlands comes from surveys meant to identify arable land. 
+After scattered surveys in the early 20th century, a comprehensive round of surveying was completed 1964-2001 as part of a wider land cover mapping in Norway  [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
+Because of its agricultural and silvicultural focus, it covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable [@ahlstromAR5Klassifikasjonssystem2019].
+Field surveyors carried a one-meter probe, so they measured peat depth as shallow (< 1 m), or deep (> 1 m). 
+These classes were assigned to whole mire polygons, so spatial resolution is on the order of hectares.
+
+Soil science needs more studies of relationships between peat depth and topographic or radiometric data to determine which variables are most predictive, how predictive they are, and how consistent their associations.
+The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify rich peatland carbon stocks and avoid their conversion.
+Although small areas can be mapped accurately on the ground, having landscape-scale maps *before* detailed investigation increases the option space for spatial planning and land management. 
+For example, the Norwegian Public Roads Administration routinely measures peat depth during geotechnical work, but by that time, the route of the road is already set. 
+With a digital peat depth map, planners could better compare routes and climate mitigation measures would have more leverage.
+
+We assess how well remotely sensed topographic and radiometric data can predict peat depth at the landscape scale, with a view towards revising regional and national maps in Norway. 
+Specifically, we: (1) quantify the accuracy of predictions from topographic and radiometric variables, (2) identify key predictive variables, and (3) compare these relationships across two different Norwegian landscapes.
+To our knowledge, this is the first study to predict peat depth from airborne radiometric data using machine learning algorithms. 
+Where airborne radiometric data have been used to predict peat depth, it has been through modeling techniques less suited for prediction and spatial extrapolation [e.g., @keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @siemonAirborneElectromagneticRadiometric2020].
+Where machine learning algorithms have been used on airborne radiometric data, it has been to predict peat extent rather than depth [e.g., @olearyDigitalSoilMapping2022].
 
@@
 The DMK peat depth classes are: < \unit{1\,m} (*shallow*), > \unit{1\,m} (*deep*), and *unknown*.
-This dataset stems from field measurements made in 1964--2007 as part of a wider land cover mapping in Norway.
-The mapping was primarily for identifying agricultural and silvicultural resources, so it covers productive areas below the tree line and has greater coverage of peat depth class in peatlands judged to be arable or afforestable [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007; @ahlstromAR5Klassifikasjonssystem2019].
 Mappers generally assigned peat depth classes to polygons of at least \unit{0.5\,ha}, although delineating polygons down to \unit{0.2\,ha} was allowed if peat depth showed a "particularly marked difference" [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
@@
 For Skrimfjella, we had independent data from the arm of our sampling design that measured peatland occurrence outside of mapped mire areas.
-For Ørskogfjelle we did not have such data.
+For Ørskogfjellet we did not have such data.
 Therefore, we first -- for both sites -- performed an evaluation with just our peat depth measurements, which included a fraction of locations not mapped as peatland.

### 2024-11-20 ms/ms.tex
ms: add intro

@@
 
-Introduction text goes here.
-Read Gatis et al. \citeyearpar{gatisMappingUplandPeat2019} and related work \citep{minasnyDigitalMappingPeatlands2019}.
+Peat soils are a terrestrial carbon stock of global importance.
+They store about 30 \% of global soil carbon despite covering only 2-3 \% of Earth's land \citep{xuPEATMAPRefiningEstimates2018, friedlingsteinGlobalCarbonBudget2020, unepGlobalPeatlandsAssessment2022}.
+In other words, they are extremely carbon dense -- denser than any other ecosystem per square meter \citep{temminkRecoveringWetlandBiogeomorphic2022}.
+This makes peatlands crucial to climate change mitigation.
+Intact peatlands sequester carbon and produce a negative temperature forcing, overall \citep{joostenRolePeatlandsClimate2016}.
+When disturbed, often by conversion to another land use, they can produce large greenhouse gas emissions \citep{maGloballyRobustRelationship2022}.
+
+One of the keys to the areal density of peatland carbon stocks lies in the third dimension: their depth.
+Peat soils range from zero to over ten meters deep, so deep peats comprise a large volume.
+Their depth results from the accumulation of organic matter over thousands of years \citep{loiselDatabaseSynthesisNorthern2014, joostenRolePeatlandsClimate2016}.
+In the anoxic and acidic conditions created by a high water table, plant material decay is slightly outweighed by new growth, and the surplus carbon is laid down as peat.
+There it remains sequestered as long as the water table stays high and prevents oxidation.
+
+Peatlands are unevenly distributed globally, and in regions with high cover they are frequently converted to human land use \citep{unepGlobalPeatlandsAssessment2022}.
+They are attractive for agriculture and forestry because they are flat, treeless, and have developed soils.
+However, other land uses also displace peatlands.
+In Norway, where 9\% of the land area is peatland \citep{brynLandCoverNorway2018}, construction and development have become important drivers of loss, after lawmakers restricted conversion to forest and farmland in recent decades.
+
+The spatial distribution of peat depth is often overlooked in land use planning and carbon accounting because peat depth is not mapped with sufficient coverage, resolution, or accuracy.
+Maps are crucial because they link high-level targets to specific management decisions, unlike spatially-aggregated estimates \citep{oecdOECDEnvironmentalPerformance2022}.
+For example, Norwegian regulations restrict conversion to farmland more strictly where peat depth exceeds \unit{1\,m}, and the distribution of peat depth on a farmer's property determines whether he is allowed to convert peatland (Forskrift om nydyrking, 1997, § 5a).
+Maps also make it possible to quantify the effect of specific management decisions and thereby understand how local outcomes contribute to regional and national outcomes \citep{oecdOECDEnvironmentalPerformance2022}.
+
+Measuring peat depth on the ground is straightforward, and a field survey can map a small area at low cost.
+However, surveying large areas is impractical when depth varies over short distances -- as in most peatlands.
+A complementary approach from soil science is digital soil mapping (DSM).
+DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest.
+This approach has grown important with the availability of remotely sensed data and the advancement of methods for identifying pattern, especially through machine learning \citep{minasnyDigitalMappingPeatlands2019, wadouxMachineLearningDigital2020}.
+
+The crux for DSM of peat depth are the relationships between peat depth and the other, mapped variables.
+For DSM to work, these relationships must be strong and consistent over the area of interest.
+They can be purely correlative rather than causal, but mechanistic relationships are stronger and more consistent.
+The \emph{scorpan} framework for DSM suggests seven predictor classes to explore: other soil properties, climate, organisms, relief (topography), parent material, age, and spatial position \citep{mcbratneyDigitalSoilMapping2003}.
+
+The most practical and widespread \emph{scorpan} factors for peat depth are relief and spatial position.
+Spatial position is unique because it is always known (with varying accuracy).
+However, the short range of spatial autocorrelation in peat depth limits its mapping value \citep{henglGenericFrameworkSpatial2004}.
+Relief, or topography, is widely and accurately mapped in digital terrain models.
+For example, most of mainland Norway has at least one elevation measurement per square meter from airborne LiDAR surveys.
+Moreover, mechanisms of peat formation are linked to topography.
+For example, a steep slope is unlikely to have a high water table, thereby limiting accumulation.
+
+Studies have consistently shown relationships between peat depth and topography, although the specific patterns vary.
+One of the most robust relationships is a negative correlation between depth and terrain slope \citep[e.g.][]{holdenEstimatingCarbonStock2011, parryMethodModellingPeat2012, gatisMappingUplandPeat2019}.
+Peat depth also changes with elevation in many contexts, but with inconsistent directionality \citep[e.g.][]{holdenEstimatingCarbonStock2011, parryMethodModellingPeat2012, rudiyantoDigitalMappingCosteffective2016, rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024}.
+Also more complex derivations of topography, such as the Topographic Wetness Index (TWI) and the Multi-Resolution Valley Bottom Flatness (MRVBF) index, have shown associations with peat depth in some studies \citep[e.g.][]{rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024}.
+Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic variables and the resolution of the peat depth analysis.
+
+Another set of variables related to peat depth are measurements of natural radioactivity from the ground surface.
+Gamma-ray spectrometry can survey the activity (decay counts per second) from radioactive isotopes in the earth's crust: potassium-40, uranium-238, and thorium-232 \citep{reinhardtGammaraySpectrometryVersatile2019}.
+These exist in bedrock (and mineral soils) and a peat overburden attenuates the radiation intensity at the surface.
+The degree of attenuation relates to properties of the overburden, especially composition and depth.
+Deep soil with high water content attenuates most \citep{beamishGammaRayAttenuation2013, reinhardtGammaraySpectrometryVersatile2019}.
+Thus, gamma-ray radiometrics integrate the \emph{scorpan} factors soil, parent material, and age.
+
+Although theory suggests that one meter of peat may fully attenuate radiometric signal \citep{beamishGammaRayAttenuation2013, reinhardtGammaraySpectrometryVersatile2019}, empirical investigations show that the association between peat depth and radioactivity can extend beyond the first meter \citep{gatisMappingUplandPeat2019, kogantiMappingPeatDepth2023}.
+Radiometric data are increasingly available over large areas, presenting a big opportunity insofar as they are predictive.
+Airborne surveys are used for various applications, and some countries have high spatial coverage of such data \citep{minasnyDigitalMappingPeatlands2019, baranwalAirborneGeophysicalSurveys2020}.
+
+The effectiveness of DSM depends not only on the methods and data, but also on the characteristics of the mapping area.
+Norway may be instructive in this respect because its peatlands vary widely across climates and topographies.
+There are many mire massif types, with different hydrology, formation, and development -- from topogeneous or soligeneous fens to blanket or raised bogs \citep{lyngstadBeskrivelserAvTorvmassivenheter2023}.
+These types have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric variables will vary by landscape.
+
+The little we know about the depths of Norwegian peatlands comes from surveys meant to identify arable land.
+After scattered surveys in the early 20th century, a comprehensive round of surveying was completed 1964-2001 as part of a wider land cover mapping in Norway \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
+Because of its agricultural and silvicultural focus, it covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable \citep{ahlstromAR5Klassifikasjonssystem2019}.
+Field surveyors carried a one-meter probe, so they measured peat depth as shallow (\textless{} 1 m), or deep (\textgreater{} 1 m).
+These classes were assigned to whole mire polygons, so spatial resolution is on the order of hectares.
+
+Soil science needs more studies of relationships between peat depth and topographic or radiometric data to determine which variables are most predictive, how predictive they are, and how consistent their associations.
+The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify rich peatland carbon stocks and avoid their conversion.
+Although small areas can be mapped accurately on the ground, having landscape-scale maps \emph{before} detailed investigation increases the option space for spatial planning and land management.
+For example, the Norwegian Public Roads Administration routinely measures peat depth during geotechnical work, but by that time, the route of the road is already set.
+With a digital peat depth map, planners could better compare routes and climate mitigation measures would have more leverage.
+
+We assess how well remotely sensed topographic and radiometric data can predict peat depth at the landscape scale, with a view towards revising regional and national maps in Norway.
+Specifically, we: (1) quantify the accuracy of predictions from topographic and radiometric variables, (2) identify key predictive variables, and (3) compare these relationships across two different Norwegian landscapes.
+To our knowledge, this is the first study to predict peat depth from airborne radiometric data using machine learning algorithms.
+Where airborne radiometric data have been used to predict peat depth, it has been through modeling techniques less suited for prediction and spatial extrapolation \citep[e.g.,][]{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, siemonAirborneElectromagneticRadiometric2020}.
+Where machine learning algorithms have been used on airborne radiometric data, it has been to predict peat extent rather than depth \citep[e.g.,][]{olearyDigitalSoilMapping2022}.
 
@@
 The DMK peat depth classes are: \textless{} \unit{1\,m} (\emph{shallow}), \textgreater{} \unit{1\,m} (\emph{deep}), and \emph{unknown}.
-This dataset stems from field measurements made in 1964--2007 as part of a wider land cover mapping in Norway.
-The mapping was primarily for identifying agricultural and silvicultural resources, so it covers productive areas below the tree line and has greater coverage of peat depth class in peatlands judged to be arable or afforestable \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007, ahlstromAR5Klassifikasjonssystem2019}.
 Mappers generally assigned peat depth classes to polygons of at least \unit{0.5\,ha}, although delineating polygons down to \unit{0.2\,ha} was allowed if peat depth showed a ``particularly marked difference'' \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
@@
 This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, © Kartverket, (2) \emph{Norge digitalt} license, Norwegian Institute of Bioeconomy Research (NIBIO), © Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway (NGU).
+Large language models have been used during the drafting and editing of this manuscript, with author oversight.
+We maintain full responsibility for the scientific output, as per the European Commission's \emph{Living guidelines on the responsible use of generative AI in research}.
 \end{acknowledgements}

### 2024-11-20 ms/ms.Rmd
ms: results - add PICP

@@
 
+Although the difficulty of predicting peat depth caused prediction intervals to be wide, these uncertainty estimates were well-calibrated.
+At Skrimfjella the prediction interval coverage probability was 91 %, and at Ørskogfjellet it was 84 % (both compared to the target value of 90 %).
+Observations outside of the prediction intervals were evenly distributed across each of the study areas.
+
 For Skrimfjella, the best predictor configuration was *all predictors*, followed closely by *terrain and radiometric*.

### 2024-11-20 ms/ms.tex
ms: results - add PICP

@@
 The spatial distribution of peat depth is often overlooked in land use planning and carbon accounting because peat depth is not mapped with sufficient coverage, resolution, or accuracy.
-Maps are crucial because they link high-level targets to specific management decisions, unlike spatially-aggregated estimates \citep{oecdOECDEnvironmentalPerformance2022}.
+Maps are crucial because they link high-level targets to specific management decisions, unlike spatially aggregated estimates \citep{oecdOECDEnvironmentalPerformance2022}.
 For example, Norwegian regulations restrict conversion to farmland more strictly where peat depth exceeds \unit{1\,m}, and the distribution of peat depth on a farmer's property determines whether he is allowed to convert peatland (Forskrift om nydyrking, 1997, § 5a).
@@
 
+Although the difficulty of predicting peat depth caused prediction intervals to be wide, these uncertainty estimates were well-calibrated.
+At Skrimfjella the prediction interval coverage probability was 91 \%, and at Ørskogfjellet it was 84 \% (both compared to the target value of 90 \%).
+Observations outside of the prediction intervals were evenly distributed across each of the study areas.
+
 For Skrimfjella, the best predictor configuration was \emph{all predictors}, followed closely by \emph{terrain and radiometric}.

### 2024-11-21 ms/ms.Rmd
ms: minor edit

@@
 
-Soil science needs more studies of relationships between peat depth and topographic or radiometric data to determine which variables are most predictive, how predictive they are, and how consistent their associations.
+Soil science needs more studies of relationships between peat depth and topographic or radiometric data to determine how predictive they are, which variables are most predictive, and how consistent their associations.
 The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify rich peatland carbon stocks and avoid their conversion.

### 2024-11-21 ms/ms.Rmd
ms: results - minor edit

@@
 Adding radiometric predictors to these configurations worsened model performance, especially in terms of concordance correlation and R^2^.
-By itself, *DMK class* produced a concordance correlation coefficient of 0.17, but the worst RMSE of any model: \unit{98\,cm}.
+By itself, *DMK class* produced a concordance correlation coefficient of 0.17 (compared to 0.008 at Skrimfjella), but the worst RMSE of any model at either site: \unit{98\,cm}.
 

### 2024-11-21 ms/ms.tex
ms: results - minor edit

@@
 Adding radiometric predictors to these configurations worsened model performance, especially in terms of concordance correlation and R\textsuperscript{2}.
-By itself, \emph{DMK class} produced a concordance correlation coefficient of 0.17, but the worst RMSE of any model: \unit{98\,cm}.
+By itself, \emph{DMK class} produced a concordance correlation coefficient of 0.17 (compared to 0.008 at Skrimfjella), but the worst RMSE of any model at either site: \unit{98\,cm}.
 

### 2024-11-21 ms/ms.Rmd
ms: results - minor edit extrapolation RMSE

@@
 \@ref(fig:modelMetricsExtrapolation)).
-Compared to our error estimates with the full dataset, RMSE at Skrimfjella was nearly unchanged, while at Ørskogfjellet it was lower outside of mapped peatland.
+At Skrimfjella RMSE in the extrapolation was nearly the same as RMSE across all land cover types.
+At Ørskogfjellet RMSE was lower in the extrapolation.
 With the independent occurrence data at Skrimfjella, we found that neither the *terrain-only* nor *terrain and radiometric* configurations were able to discriminate between peat presence and absence (area under the curve of the receiver operating characteristic 0.44 and 0.52 respectively, where 0.5 indicates random guessing).

### 2024-11-21 ms/ms.tex
ms: results - minor edit extrapolation RMSE

@@
 \ref{fig:modelMetricsExtrapolation}).
-Compared to our error estimates with the full dataset, RMSE at Skrimfjella was nearly unchanged, while at Ørskogfjellet it was lower outside of mapped peatland.
+At Skrimfjella RMSE in the extrapolation was nearly the same as RMSE across all land cover types.
+At Ørskogfjellet RMSE was lower in the extrapolation.
 With the independent occurrence data at Skrimfjella, we found that neither the \emph{terrain-only} nor \emph{terrain and radiometric} configurations were able to discriminate between peat presence and absence (area under the curve of the receiver operating characteristic 0.44 and 0.52 respectively, where 0.5 indicates random guessing).

### 2024-11-22 ms/ms.Rmd
switch from RMSE to MAE

--- 2024-11-21 ms/ms.Rmd
@@
 We implemented the kNNDM with the *CAST* R package [v.1.0.2, @meyerCASTPackageTraining2024], setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
-From the cross-validation we quantified *Root Mean Squared Error* (accuracy, original scale), *R^2^* (correlation, standardized scale), and *Lin's concordance correlation coefficient* (accuracy and correlation, standardized scale).
+From the cross-validation we quantified mean absolute error (accuracy, original scale), R^2^ (correlation, standardized scale), and Lin's concordance correlation coefficient (accuracy and correlation, standardized scale).
 
@@
 In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
-From the cross-validation we quantified *Root Mean Squared Error*.
+From the cross-validation we quantified mean absolute error.
 Second -- for Skrimfjella -- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
@@
 None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \@ref(fig:modelMetrics)).
-For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an *R^2^* of 0.34, and an RMSE of \unit{78\,cm}.
-For Ørskogfjellet the same values were 0.39, 0.33, and \unit{75\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
+For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an *R^2^* of 0.34, and a mean absolute error of \unit{60\,cm}.
+For Ørskogfjellet the same values were 0.39, 0.33, and \unit{56\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
 
@@
 The performance gap between the *terrain and DMK* configuration and the *terrain-only* configuration was similarly small.
-Compared to the *terrain-only* configuration, the *terrain and radiometric* configuration improved concordance correlation by 0.04, R^2^ by 0.04, and RMSE by \unit{2\,cm}.
+Compared to the *terrain-only* configuration, the *terrain and radiometric* configuration improved concordance correlation by 0.04, R^2^ by 0.04, and mean absolute error by \unit{1\,cm}.
 *DMK class* alone -- although calibrated to measured depths -- was a very poor predictor of peat depth, with a concordance correlation coefficient of 0.008.
@@
 Adding radiometric predictors to these configurations worsened model performance, especially in terms of concordance correlation and R^2^.
-By itself, *DMK class* produced a concordance correlation coefficient of 0.17, but the worst RMSE of any model: \unit{98\,cm}.
+By itself, *DMK class* produced a concordance correlation coefficient of 0.17 (compared to 0.008 at Skrimfjella), but the worst mean absolute error of any model at either site: \unit{77\,cm}.
 
-When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced marginally better RMSE than just assuming a constant \unit{30\,cm} depth (Fig.
+When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced worse mean absolute error than just assuming a constant \unit{30\,cm} depth (Fig.
 \@ref(fig:modelMetricsExtrapolation)).
-Compared to our error estimates with the full dataset, RMSE at Skrimfjella was nearly unchanged, while at Ørskogfjellet it was lower outside of mapped peatland.
 With the independent occurrence data at Skrimfjella, we found that neither the *terrain-only* nor *terrain and radiometric* configurations were able to discriminate between peat presence and absence (area under the curve of the receiver operating characteristic 0.44 and 0.52 respectively, where 0.5 indicates random guessing).

### 2024-11-22 ms/ms.tex
switch from RMSE to MAE

--- 2024-11-21 ms/ms.tex
@@
 We implemented the kNNDM with the \emph{CAST} R package \citep[v.1.0.2,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
-From the cross-validation we quantified \emph{Root Mean Squared Error} (accuracy, original scale), \emph{R\textsuperscript{2}} (correlation, standardized scale), and \emph{Lin's concordance correlation coefficient} (accuracy and correlation, standardized scale).
+From the cross-validation we quantified mean absolute error (accuracy, original scale), R\textsuperscript{2} (correlation, standardized scale), and Lin's concordance correlation coefficient (accuracy and correlation, standardized scale).
 
@@
 In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
-From the cross-validation we quantified \emph{Root Mean Squared Error}.
+From the cross-validation we quantified mean absolute error.
 Second -- for Skrimfjella -- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
@@
 None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \ref{fig:modelMetrics}).
-For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an \emph{R\textsuperscript{2}} of 0.34, and an RMSE of \unit{78\,cm}.
-For Ørskogfjellet the same values were 0.39, 0.33, and \unit{75\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
+For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an \emph{R\textsuperscript{2}} of 0.34, and a mean absolute error of \unit{60\,cm}.
+For Ørskogfjellet the same values were 0.39, 0.33, and \unit{56\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
 
@@
 The performance gap between the \emph{terrain and DMK} configuration and the \emph{terrain-only} configuration was similarly small.
-Compared to the \emph{terrain-only} configuration, the \emph{terrain and radiometric} configuration improved concordance correlation by 0.04, R\textsuperscript{2} by 0.04, and RMSE by \unit{2\,cm}.
+Compared to the \emph{terrain-only} configuration, the \emph{terrain and radiometric} configuration improved concordance correlation by 0.04, R\textsuperscript{2} by 0.04, and mean absolute error by \unit{1\,cm}.
 \emph{DMK class} alone -- although calibrated to measured depths -- was a very poor predictor of peat depth, with a concordance correlation coefficient of 0.008.
@@
 Adding radiometric predictors to these configurations worsened model performance, especially in terms of concordance correlation and R\textsuperscript{2}.
-By itself, \emph{DMK class} produced a concordance correlation coefficient of 0.17, but the worst RMSE of any model: \unit{98\,cm}.
+By itself, \emph{DMK class} produced a concordance correlation coefficient of 0.17 (compared to 0.008 at Skrimfjella), but the worst mean absolute error of any model at either site: \unit{77\,cm}.
 
-When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced marginally better RMSE than just assuming a constant \unit{30\,cm} depth (Fig.
+When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced worse mean absolute error than just assuming a constant \unit{30\,cm} depth (Fig.
 \ref{fig:modelMetricsExtrapolation}).
-Compared to our error estimates with the full dataset, RMSE at Skrimfjella was nearly unchanged, while at Ørskogfjellet it was lower outside of mapped peatland.
 With the independent occurrence data at Skrimfjella, we found that neither the \emph{terrain-only} nor \emph{terrain and radiometric} configurations were able to discriminate between peat presence and absence (area under the curve of the receiver operating characteristic 0.44 and 0.52 respectively, where 0.5 indicates random guessing).

### 2024-11-22 ms/figures.R

@@ -199,14 +199,17 @@ library(tidyverse)
 orskog <- read_csv("output/modelmetrics.csv") 
 skrim <- read_csv("output/Skrim/modelmetrics.csv")
 plotting <- bind_rows(orskog = orskog, skrim = skrim, .id = 'site') %>% 
-  filter(.metric != 'mae') %>% 
+  filter(.metric != 'rmse') %>% 
   mutate(
     site = fct_relevel(site, "orskog"),
-    .metric = fct_relevel(.metric, "ccc", "rsq", "rmse"),
     metric = case_when(
       .metric == "ccc" ~ "Concordance \ncorrelation coefficient",
       .metric == "rsq" ~ "R-squared",
-      .metric == "rmse" ~ "RMSE"),
+      .metric == "mae" ~ "Mean \nabsolute error"),
+    metric = fct_relevel(metric,
+                         "Concordance \ncorrelation coefficient",
+                         "R-squared",
+                         "Mean \nabsolute error"),
     model = case_when(
       model == "DMK" ~ "DMK class (2)",
       model == "Terrain" ~ "terrain (21)",
@@ -257,11 +260,11 @@ library(tidyverse)
 orskog <- read_csv("output/modelmetrics-extrapolation.csv") 
 skrim <- read_csv("output/Skrim/modelmetrics-extrapolation.csv")
 plotting <- bind_rows(orskog = orskog, skrim = skrim, .id = 'site') %>% 
-  filter(.metric != 'mae') %>% 
+  filter(.metric != 'rmse') %>% 
   mutate(
     site = fct_relevel(site, "orskog"),
     metric = case_when(
-      .metric == "rmse" ~ "RMSE"),
+      .metric == "mae" ~ "Mean \nabsolute error"),
     model = case_when(
       model == "null" ~ "null, assume 30 cm",
       model == "Terrain" ~ "terrain (21)",
@@ -293,7 +296,7 @@ g1 <- ggplot(plotting) +
         panel.grid = element_blank(),
         axis.ticks.y = element_blank(),
         legend.position = "inside",
-        legend.position.inside = c(-0.4, 1.06),
+        legend.position.inside = c(-0.4, 1.14),
         legend.key.spacing.y = unit(-2, "mm"),)
 ggsave(filename = 'modelmetrics-extrapolation.pdf', path = "ms/figures",
        width =(210-30)/2, height = (240-40)/4, units = 'mm') #copernicus.cls page 210x240

### 2024-11-22 ms/ms.Rmd
add calibration plots

@@
 
+The best models at both sites strongly underpredicted the deepest peats (Fig. \@ref(fig:calPlots)).
+
+```{r calPlots, echo = FALSE, fig.cap = "Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions from spatial cross-validation.
+Blue smoother lines are local polynomial regressions.
+Marginal distributions are shaded by quartile."}
+knitr::include_graphics("figures/calibration_plots.pdf")
+```
+
 When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced worse mean absolute error than just assuming a constant \unit{30\,cm} depth (Fig.

### 2024-11-22 ms/ms.tex
add calibration plots

@@
 
+The best models at both sites strongly underpredicted the deepest peats (Fig. \ref{fig:calPlots}).
+
+\begin{figure}
+\centering
+\includegraphics{figures/calibration_plots.pdf}
+\caption{\label{fig:calPlots}Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions from spatial cross-validation.
+Blue smoother lines are local polynomial regressions.
+Marginal distributions are shaded by quartile.}
+\end{figure}
+
 When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced worse mean absolute error than just assuming a constant \unit{30\,cm} depth (Fig.

### 2024-11-22 R/Skrim/modelDepth.R

@@ -226,6 +226,7 @@ preds <- fit_RTD_knndm %>%
   select(id, depth_cm, .pred)
 preds %>% 
   probably::cal_plot_regression(depth_cm, .pred, smooth = FALSE)
+readr::write_csv(preds, "output/Skrim/calibrationplot.csv")
 
 ## Extrapolating beyond mire ####
 

### 2024-11-22 R/modelDepth.R

@@ -226,6 +226,7 @@ preds <- fit_TD_knndm %>%
   select(id, depth_cm, .pred)
 preds %>% 
   probably::cal_plot_regression(depth_cm, .pred, smooth = FALSE)
+readr::write_csv(preds, "output/calibrationplot.csv")
 
 ## Extrapolating beyond mire ####
 

### 2024-11-22 ms/figures.R

@@ -301,6 +301,56 @@ g1 <- ggplot(plotting) +
 ggsave(filename = 'modelmetrics-extrapolation.pdf', path = "ms/figures",
        width =(210-30)/2, height = (240-40)/4, units = 'mm') #copernicus.cls page 210x240
 
+# Calibration plots ####
+
+library(tidyverse)
+library(patchwork)
+
+orskog <- read_csv("output/calibrationplot.csv")
+skrim <- read_csv("output/Skrim/calibrationplot.csv")
+
+probs <- c(0.25, 0.5, 0.75)
+
+scatter1 <- ggplot(orskog, aes(x = depth_cm, y = .pred)) +
+  geom_point(size = 0.8, alpha=0.5) +
+  geom_smooth(method = 'loess', formula= y ~ x, se = FALSE) +
+  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "red") +
+  tune::coord_obs_pred(ratio = 1) +
+  labs(x = "Observed depth (cm)", y = "Predicted depth (cm)") +
+  theme_bw()
+dens <- density(orskog$depth_cm)
+df <- data.frame(x = dens$x, y = dens$y)
+quantiles <- quantile(orskog$depth_cm, prob=probs)
+df$quant <- factor(findInterval(df$x, quantiles))
+density1 <- ggplot(df, aes(x,y)) + 
+  geom_area(aes(y = y, fill=quant)) +
+  geom_line() +
+  scale_fill_brewer(guide="none", palette = "Greys") +
+  theme_void()
+
+scatter2 <- ggplot(skrim, aes(x = depth_cm, y = .pred)) +
+  geom_point(size = 0.8, alpha=0.5) +
+  geom_smooth(method = 'loess', formula= y ~ x, se = FALSE) +
+  geom_abline(intercept = 0, slope = 1, linetype = 2, color = "red") +
+  tune::coord_obs_pred(ratio = 1) +
+  labs(x = "Observed depth (cm)", y = "Predicted depth (cm)") +
+  theme_bw()
+dens <- density(skrim$depth_cm)
+df <- data.frame(x = dens$x, y = dens$y)
+quantiles <- quantile(skrim$depth_cm, prob=probs)
+df$quant <- factor(findInterval(df$x, quantiles))
+density2 <- ggplot(df, aes(x,y)) + 
+  geom_area(aes(y = y, fill=quant)) +
+  geom_line() +
[…]

### 2024-11-22 R/Skrim/modelDepth.R
add best model's mean error

@@ -226,6 +226,7 @@ preds <- fit_RTD_knndm %>%
   select(id, depth_cm, .pred)
 preds %>% 
   probably::cal_plot_regression(depth_cm, .pred, smooth = FALSE)
+mean(preds$.pred - preds$depth_cm)
 readr::write_csv(preds, "output/Skrim/calibrationplot.csv")
 
 ## Extrapolating beyond mire ####

### 2024-11-22 R/modelDepth.R

@@ -226,6 +226,7 @@ preds <- fit_TD_knndm %>%
   select(id, depth_cm, .pred)
 preds %>% 
   probably::cal_plot_regression(depth_cm, .pred, smooth = FALSE)
+mean(preds$.pred - preds$depth_cm)
 readr::write_csv(preds, "output/calibrationplot.csv")
 
 ## Extrapolating beyond mire ####

### 2024-11-26 ms/ms.Rmd
add best model's mean error
ms: Intro - add ref

@@
 
-Although theory suggests that one meter of peat may fully attenuate radiometric signal [@beamishGammaRayAttenuation2013; @reinhardtGammaraySpectrometryVersatile2019], empirical investigations show that the association between peat depth and radioactivity can extend beyond the first meter [@gatisMappingUplandPeat2019; @kogantiMappingPeatDepth2023].
+Although theory suggests that one meter of peat may fully attenuate radiometric signal [@beamishGammaRayAttenuation2013; @reinhardtGammaraySpectrometryVersatile2019], empirical investigations show that the association between peat depth and radioactivity can extend beyond the first meter [@keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @kogantiMappingPeatDepth2023].
 Radiometric data are increasingly available over large areas, presenting a big opportunity insofar as they are predictive. 
@@
 
-The best models at both sites strongly underpredicted the deepest peats (Fig. \@ref(fig:calPlots)).
+The best models at both sites overpredicted shallow peats and strongly underpredicted very deep peats (Fig. \@ref(fig:calPlots)).
+The mean error (bias) of these models was \unit{10\,cm} at Skrimfjella and \unit{-4\,cm} at Ørskogfjellet.
 

### 2024-11-26 ms/ms.tex
add best model's mean error
ms: Intro - add ref

@@
 
-Although theory suggests that one meter of peat may fully attenuate radiometric signal \citep{beamishGammaRayAttenuation2013, reinhardtGammaraySpectrometryVersatile2019}, empirical investigations show that the association between peat depth and radioactivity can extend beyond the first meter \citep{gatisMappingUplandPeat2019, kogantiMappingPeatDepth2023}.
+Although theory suggests that one meter of peat may fully attenuate radiometric signal \citep{beamishGammaRayAttenuation2013, reinhardtGammaraySpectrometryVersatile2019}, empirical investigations show that the association between peat depth and radioactivity can extend beyond the first meter \citep{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, kogantiMappingPeatDepth2023}.
 Radiometric data are increasingly available over large areas, presenting a big opportunity insofar as they are predictive.
@@
 
-The best models at both sites strongly underpredicted the deepest peats (Fig. \ref{fig:calPlots}).
+The best models at both sites overpredicted shallow peats and strongly underpredicted very deep peats (Fig. \ref{fig:calPlots}).
+The mean error (bias) of these models was \unit{10\,cm} at Skrimfjella and \unit{-4\,cm} at Ørskogfjellet.
 

### 2024-12-02 ms/ms.Rmd
ms: minor edits Methods/Results

@@
 We implemented the kNNDM with the *CAST* R package [v.1.0.2, @meyerCASTPackageTraining2024], setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
-From the cross-validation we quantified mean absolute error (accuracy, original scale), R^2^ (correlation, standardized scale), and Lin's concordance correlation coefficient (accuracy and correlation, standardized scale).
+From the cross-validation we quantified mean absolute error (error magnitude, original scale), R^2^ (explained variation, standardized scale), and Lin's concordance correlation coefficient (error magnitude and explained variation, standardized scale).
 
@@
 For Ørskogfjellet the same values were 0.39, 0.33, and \unit{56\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
+These values were derived from kNNDM spatial cross-validation with 20 folds at Skrimfjella and 10 folds at Ørskogfjellet.
 

### 2024-12-02 ms/ms.tex
ms: minor edits Methods/Results

@@
 We implemented the kNNDM with the \emph{CAST} R package \citep[v.1.0.2,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
-From the cross-validation we quantified mean absolute error (accuracy, original scale), R\textsuperscript{2} (correlation, standardized scale), and Lin's concordance correlation coefficient (accuracy and correlation, standardized scale).
+From the cross-validation we quantified mean absolute error (error magnitude, original scale), R\textsuperscript{2} (explained variation, standardized scale), and Lin's concordance correlation coefficient (error magnitude and explained variation, standardized scale).
 

### 2024-12-04 ms/ms.Rmd
ms: add Discussion

@@
 None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \@ref(fig:modelMetrics)).
-For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an *R^2^* of 0.34, and a mean absolute error of \unit{60\,cm}.
+For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an R^2^ of 0.34, and a mean absolute error of \unit{60\,cm}.
 For Ørskogfjellet the same values were 0.39, 0.33, and \unit{56\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
@@
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident,
-sunt in culpa qui officia deserunt mollit anim id est laborum.
+## Can we improve Norway’s peat depth maps?
 
-# Conclusions
+### Remotely sensed variables are weak (but not useless!) predictors of peat depth
 
-Nulla facilisi.
-Maecenas vel nunc nec purus tincidunt congue.
-Proin auctor, lectus eu pharetra malesuada,
-nisi nunc bibendum nunc,
-eget tincidunt nunc nisi id nunc.
-Sed euismod, nunc sit amet aliquam tincidunt,
-nunc nunc tincidunt nunc,
-nec tincidunt nunc nunc nec nunc.
-Donec auctor, nunc sit amet aliquam tincidunt,
-nunc nunc tincidunt nunc,
-nec tincidunt nunc nunc nec nunc.
+Our ability to predict peat depth in the study sites based on terrain and radiometric data was limited.
+Mean absolute errors of 60 and 56 cm at the two sites — relative to mean depths of 119 and 126 cm — illustrate the practical limitations of these maps.
+Since any given 10 m cell will miss by about 60 cm, applications requiring detailed peat depth in a small area (e.g.
+< 1 ha) would benefit from measuring depth on the ground rather than relying on the DSM alone.
+
+On the other hand, we improved on the best available map of peat depth (DMK depth class), which is based on field measurements only.
+This highlights the general value of remotely sensed data, whose complete coverage can improve maps even when their association with the variable of interest is weak.
+Since remotely sensed data are widely available, improvements to soil maps as shown here are low hanging fruit.
+This point is recognized and reflected in the rise of DSM.
+
+DMK peat depth classes were a worse predictor of peat depth than our models even though we calibrated them with the same data.
+That is: we gave DMK and the DSM equal footing for a fair comparison.
+If we had taken the DMK peat depth classes at face value and assumed depths according to their class definitions (< 1 m, > 1 m), they would have performed worse and the advantage of the DSM would be greater.
+The advantage of the DSM was not large in absolute terms (9 cm and 21 cm improvements in mean absolute error), but it explained much more of the variation in depth (improvements in R^2^ of 0.16 and 0.19).
+We attribute this result to the poor spatial and thematic resolution of DMK peat depth, which precludes a robust correlation with peat depths varying from 0-8 m at fine spatial scales.
+
+The performance of our terrain- and radiometric-based maps could have been improved with a spatially explicit mapping approach like regression kriging.
+By ignoring spatial autocorrelation in peat depth (i.e. the n in scorpan), we have not extracted all of the information about peat depth in the study area out of the training data.
+If we were to put our DSM approach into production for published maps, we would harness the spatial component, but the actual maps for Ørskogfjellet and Skrimfjella are not of primary interest in this study.
+Moreover, residuals of the RF predictions showed weak spatial structure at Skrimfjella, and only up to a range of 150 m at Ørskogfjellet.
+Therefore, the improvement from regression kriging would be small overall and limited to parts of the maps close to measurements.
+
+### Similar error but less explanatory power in Norwegian peatlands
+
+Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth but generally produced better or comparable error magnitude (R^2^ vs.
+MAE/RMSE; @wadouxIntegratedApproachEvaluation2022).
+It is important to keep in mind that differences in peat depth distributions, spatial scales, and evaluation methods make direct performance comparisons precarious. 
+More standardized reporting would help but not eliminate this consideration. 
+For example, R^2^ is sensitive to high-leverage extreme values, so it will evaluate a right-skewed distribution differently than a symmetrical distribution. 
+We evaluated our models with respect to the explicit purpose of creating peat depth maps across the study areas, but not all studies tailored evaluation to match an explicitly formulated problem [@milaNearestNeighbourDistance2022].
+
+Gatis et al. [-@gatisMappingUplandPeat2019] used similar predictors and the same spatial grain, finding a much stronger correlation between predicted and observed peat depth (R^2^ = 0.68). 
+Although their random evaluation data partition could make performance estimates too optimistic [@robertsCrossvalidationStrategiesData2017; @wadouxSpatialCrossvalidationNot2021], the confounding effect of spatial structure is probably small because they used linear regression and few predictors.
+Their model had limited opportunity to overfit to the spatial structure in peat depth. 
+The most salient difference in Gatis et al. [-@gatisMappingUplandPeat2019] compared to our study is the character of the study area. 
+They study a flatter area with a higher proportion of peatland cover, and their peatland is primarily blanket bog. 
+The predominance of fens and smaller peatland extent may have contributed to worse performance in our study.
+
+Marchant [-@marchantUsingRemoteSensors2021] examined a subset of the area studied in Gatis et al.
+[-@gatisMappingUplandPeat2019] at 100 m resolution, using splines to relax linearity between radiometry/terrain and peat depth.
+He found that radK alone predicted peat depth with much higher concordance than our models (CCC = 0.76) and that elevation alone produced comparable performance (CCC = 0.27). 
+The RMSE from these univariate models was 46-68 cm (cf. 78 and 75 cm for Skrimfjella and Ørskogfjellet).
+
+Koganti et al. [-@kogantiMappingPeatDepth2023] had a peat depth distribution and predictors similar to ours, but at much smaller spatial grain and extent. 
+Their training and validation points are closer than the range of spatial autocorrelation in peat depth, so our results are best compared to their non-spatial models. 
+They accounted for more variability in peat depth (adjusted $R^2 = 0.71$) but had larger errors (RSME = 110 cm). 
+Koganti et al.’s [-@kogantiMappingPeatDepth2023] linear regression models produced negative predictions, and it is unclear whether the values quoted above include these. 
+If we disregard the negative predictions, their model showed the same pattern as ours in overpredicting shallow peats and underpredicting deep peats, although their underprediction was less severe. 
+An important difference between Koganti et al.
+[-@kogantiMappingPeatDepth2023] and our study (besides spatial scale) is that they measured radiometrics on the ground, rather than using airborne survey data.
+
+### Asymmetries in depth predictions for land use planning and carbon accounting
+
+Our models erred most for the deepest peats. 
+Where overprediction occurred, it was smaller. 
+This is not unexpected for the right-skewed distributions of peat depth, but it has management implications. 
+Identifying the deepest peats will require additional field work in candidate areas, which could be defined by an upper quantile of predicted depth. 
+Map users should not trust the maps to identify all large carbon stocks, but they can trust that identified large stocks really are large. 
+That makes the map more suited for “red-lighting” than “green-lighting” peatland conversion, for example. 
+Where it does not prohibit conversion (i.e. predicts shallow peat), a ground survey should be done before conversion is allowed (assuming depth is the only consideration). 
+Although this recommendation aligns with the precautionary principle, here we make it on technical grounds based on the maps’ characteristics.
+
+### Depth predictions do not necessarily extrapolate to peat occurrence
+
+That we measured more than 30 cm of peat in areas not mapped as peatland was expected, because AR5 underrepresents peatland extent by about one third [@brynLandCoverNorway2018]. 
+For that reason, maps of peatland occurrence also need revising. 
+Unfortunately, our models did not predict peat occurrence outside mapped peatlands. 
+In these areas, assuming a constant 30 cm of peat (the depth threshold used in AR5) produced less error than our models. 
+Moreover, the performance gap would likely have been larger had our evaluation data been representative of the non-peatland areas. 
+Instead they were near mapped peatlands. 
+Most importantly, the independent occurrence data at Skrimfjella showed the model completely failing to predict occurrence, and these data were suited for testing this ability. 
+Although the Ørskogfjellet model may have done better than the Skrimfjella model on a similar independent evaluation set (since it predicted depth better), the improvement would probably be small.
+
+Why wasn't peat occurrence predicted well? 
+The models were trained on fundamentally different populations of locations (~80 % mapped peatland in training vs.
+0 % in prediction), so it is not surprising that the associations they learned did not transfer well.
+Moreover, the ~20 % of training data from outside mapped peatland were incidentally collected (near peatland) and not representative of other land cover classes. 
+In short, the models were blind to the fact that most of both study areas have no peat.
+
+Peatland extent is probably best mapped using different remotely sensed data than we used here [@bakkestuenDelineationWetlandAreas2023], and this study’s purpose was not to map extent. 
+Nevertheless — as long as the relevant peatland definition includes a depth component — depth predictions (or predictors with subsurface information) should help delineate peatland extent [@olearyDigitalSoilMapping2022; @beamishDetailedMappingPeat2024].
+We return to this point in our discussion of implications for digital soil mapping.
+
+## Which variables predict peat depth?
+
+### Airborne radiometrics do not predict Norwegian peat depth
+
+Radiometric data had no predictive value at Ørskogfjellet, while at Skrimfjella they had minor influence in a relatively weak model. 
+The bedrock is more homogeneous at Ørskogfjellet than at Skrimfjella, so uneven radiogenesis is not a viable explanation for the differences between sites nor the poor performance in general [@beamishEnvironmentalRadioactivityUK2014; @reinhardtGammaraySpectrometryVersatile2019].
+To the extent that radiometrics had predictive value at Skrimfjella, it appears that they were most valuable near the extremes of the depth distribution, since their inclusion improved R^2^ more than mean absolute error.
+All four variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value. 
+This contrasts with Koganti et al. [-@kogantiMappingPeatDepth2023], who found that radTC was a much better predictor than radK.
+
+We suspect that the primary reason for the poor predictive value of the radiometric data was the large footprint of the detector in the airborne survey. 
+With an average flight altitude of 75 m, less than half of the radiation reaching the detector comes from inside the 100 m diameter circle directly below it [@beamishEnhancingResolutionAirborne2016; @beamishDetailedMappingPeat2024].
+The rest of the measured activity integrates a much wider area. 
+For comparison, empirical variograms of peat depth at Skrimfjella and Ørskogfjellet showed no spatial autocorrelation beyond 50 and 75 m (among GPR data) or 110 and 230 m (among 10 m cells). 
+Basically, the airborne radiometric data will not capture large variation over short (< 100 m) distances; the instrument’s field of view has a large smoothing effect on the data [@beamishEnhancingResolutionAirborne2016; @reinhardtGammaraySpectrometryVersatile2019].
+Different landforms and the changes they cause in the geometry between the radioactive source and the detector can also distort airborne measurements [@reinhardtGammaraySpectrometryVersatile2019].
+
+Studies comparing airborne and ground radiometric surveys confirm that they are poorly correlated in low-activity areas like peatlands [@kockComparisonAirborneTerrestrial2011; @karjalainenComparisonTwoGammaray2024].
+Karjalainen et al. [-@karjalainenComparisonTwoGammaray2024] found that ground-based measurements predicted peat depth better than airborne measurements. 
+Nevertheless, a large radiometric footprint did not prevent strong associations with peat depth in Gatis et al.
+[-@gatisMappingUplandPeat2019] and Marchant [-@marchantUsingRemoteSensors2021], likely due to the extensive blanket bog landscape with more gradual changes in depth [@lindsayBogsEcologyClassification1995; @lyngstadBeskrivelserAvTorvmassivenheter2023].
+We are unsure whether short-range depth changes explain the weak associations that Siemon et al. [-@siemonAirborneElectromagneticRadiometric2020] found in a large raised bog.
+
+Weather conditions varied during the Ørskogfjellet radiometric survey and affected its data [@ofstadHelicopterborneMagneticRadiometric2015]. 
+Thus, uneven snow cover and air moisture may also have masked the soil signal in these data.
+
+We do not believe that the poor predictive value of the radiometric data in this study was caused by fully attenuated radioactivity — at least not in large part. 
+The RF algorithm’s flexibility means that radiometrics could be used for shallower peats if they provided predictive value for that part of the depth distribution, but there is no evidence of that in our results.
+In the partial dependence plot of radU at Skrimfjella, the expected negative relationship between depth and radU cannot be found by ignoring the left (highly-attenuated) side of the distribution. 
+About a quarter of the peats in our study were less than a meter deep, and full attenuation is unlikely for these [@beamishGammaRayAttenuation2013].
+
+Although we do not believe full attenuation is the primary reason for poor performance in our study, it may limit peat depth mapping under other circumstances. 
+First-principle calculations suggest that radiation should be 90 % attenuated after about 50-60 cm of typical, wet peat or 85 cm of unnaturally dry peat [@beamishGammaRayAttenuation2013; @beamishDetailedMappingPeat2024], and some field tests support these values [@billenEignungGammaspektrometrieKartieren2015].
+It is remarkable that particular studies detected radiation differences up to several meters deep [@gatisMappingUplandPeat2019; @kogantiMappingPeatDepth2023], but these may be the exceptions rather than the rule.
+Perhaps relatively deeper water tables in these study sites [blanket bog, drained fen, @pricePeatlandRestorationHydrology2016] contributed to better penetration.
+
+### Terrain-based variables can predict peat depth
+
+At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables. 
+Elevation was the most important predictor at both sites, and peat depth showed non-monotonic responses to changes in elevation. 
+We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g.
+via climate) to explain the observed patterns.
+Indeed, elevation at Ørskogfjellet seems to interact with other variables (non-parallel ICE lines), complicating its interpretation. 
+Moreover, relationships between elevation and peat depth have previously shown opposite shapes in different areas [@finlaysonEstimatingOrganicSurface2021]. 
+Nonetheless, a relationship that is not generalizable beyond the mapping area is still useful for DSM, as long as it is evaluated to demonstrate its robustness for the predictive task (e.g.
+through kNNDM spatial cross validation).
+
+One interesting feature of the elevational relationships we found may be generalizable: a steep increase in peat depth near the upper marine limit after the last ice age. 
+At Ørskogfjellet, the upper marine limit is about 75 meters above today’s sea level [Geological Survey of Norway, @hogaasDatabaseRegistreringAv2012], where the partial dependence plot of elevation shows a sharp increase in peat depth.
+In areas under the upper marine limit there has been less time for peat accumulation since the ice sheets retreated, and it is plausible that this makes peats there shallower, all else being equal. 
+We cannot evaluate this effect at Skrimfjella, where the upper marine limit is below our study area (at 175 m.a.s.l.).
+
+Another influential terrain-based predictor was MRVBF. 
+Unlike elevation, it showed a monotonic effect on peat depth: greater valley bottom flatness was always associated with increases in peat depth. 
+Delineating a valley bottom involves ambiguity, but the MRVBF index is a pragmatic approach that considers a location a valley bottom if it is sufficiently low and flat at a particular scale [@gallantMultiresolutionIndexValley2003].
+The multiresolution nature of the index allows small elevated but flat areas (including saddles) to be characterized as having high valley bottom flatness [@gallantMultiresolutionIndexValley2003]. 
+Our results suggest that MRVBF is a robust indicator of high water tables (and peat accumulation) over millennial time scales, corroborating other studies [@rudiyantoOpenDigitalMapping2018; @deragonMappingMaximumPeat2023].
+
+Other terrain-derived predictors with predictive value in our study are hydrological (Topographic Wetness Index and Depth-To-Water index). 
+Notably, terrain slope was inferior to (Ørskogfjellet) or highly correlated with (Skrimfjella) these hydrological indices. 
+Mappers of peat depth should not assume that slope is the best predictor in its class, despite its prevalence in the literature. 
+Wetter locations (high TWI and low DTW) were generally associated with deeper peat, but these relationships were not as strong or consistent as with MRVBF. 
+The optimal scale for Topographic Wetness Index and Depth-To-Water index varied, and likely depends on both the dominant peat formation processes and the typical size of peatland features in a landscape.
+Including multiple scales of these variables allows the model to capture different hydrological mechanisms operating at different spatial scales.
+
+### Legacy depth maps have inconsistent predictive value
+
+DMK peat depth class proved an inconsistent predictor of peat depth. 
+At Skrimfjella, it barely improved model performance. 
+At Ørskogfjellet, it increased performance more, and both indicator variables were among the most important in the model. 
+We suspect the discrepancy between sites is due to different levels of effort and coverage during the historical surveys; more lowland peatland near agriculture at Ørskogfjellet may have caused more purposeful surveying.
+73 % of cells measured at Skrimfjella had unknown depth in DMK, compared to 20 % at Ørskogfjellet. 
+Still, the 25 % of cells classified as deep at Skrimfjella were not meaningfully deeper than the rest, according to our model. 
+Interactions between DMK and other variables underline the inconsistency of DMK depth maps, even within a site. 
+That a peat at Ørskogfjellet was classified as shallow rather than deep did not uniformly lower the model’s expectation of depth there. 
+Rather, the veracity of the deep-indication was contingent on other factors.
+
+## Implications for digital soil mapping of peat depth
+
+The performance gap between the best models and the DMK-only models shows that peat depth in Norway should be mapped digitally. 
+Anywhere we have some calibrating measurements, we can get better maps than DMK peat depth classes, at low cost. 
+Moreover, DSM methodology can align map products with open science principles by making their production transparent, reproducible, and updatable. 
+The large difference we found between the coverage and quality of DMK peat depth at Skrimfjella versus Ørskogfjellet underlines these advantages. 
+With DSM we can apply the same approach across different areas and make maps with full spatial coverage, continuous values, and validated uncertainty. 
+The rest of this section discusses recommendations and needs for more extensive peat depth DSM.
+
+### Peat depth measurements should be organized
+
+High-quality digital terrain models are available for mainland Norway, making peat depth measurements the critical training data need. 
+For a given area, some minimum number of measurements is necessary to create meaningful improvement over DMK peat depth. 
+The proportion of peatlands sampled at Skrimfjella was twice that at Ørskogfjellet, but the Skrimfjella model performed worse, which shows that the size of the depth dataset is not all-important.
+We had the luxury of stratifying our measurements over candidate predictors, and performance may suffer where locations are opportunistic or tied to a sampling design with a different purpose. 
+The magnitude of this penalty will depend on the dataset characteristics, but having enough depth measurements is probably more consequential [@wadouxSamplingDesignOptimization2019].
+
+Depth measurements are foundational, and better infrastructure to collect, preserve, and make FAIR these data would help DSM and other applications. 
+Geoportal access and data exchange standards [like Natural England’s for peat surveys, -@naturalenglandDataExchangeStandard2023] are important. 
+Peat data often fall through the cracks between geology-oriented and ecology-oriented archives, but increased awareness of peatland importance is a good impetus to remedy this situation. 
+Peat depth is quick and easy to measure, so integrating its measurement into existing national field programs, like Norway’s spatially representative nature monitoring (ANO) or national forest inventory, would be advantageous (although not sufficient for regional or local mapping).
+Municipalities and local actors in Norway are increasingly measuring peat depth and can contribute to a growing data foundation [@kyrkjeeideCalculatorLocalPeatland2023]. 
+All of this lays the groundwork for renewed peatland maps.
+
+### Spatial scale affects model performance and utility
+
+The scarcity of depth measurements and their short spatial autocorrelation mean that mapping peat depth at broad scales is not an exercise in spatial interpolation [@henglGenericFrameworkSpatial2004].
+Peat depths in our GPR data showed spatial autocorrelation to a range of 50-100 m, and getting measurements at such fine grain is only realistic for small areas, not across landscapes. 
+Therefore, we anticipate that spatially-explicit DSM is of limited value in all but the most intensively sampled landscapes (currently nowhere in Norway). 
+Fine-scale variation in peat depth raises the possibility of mapping at very fine resolution (1 m) — even if land use planning and carbon accounting do not operate at this grain. 
+Terrain-depth relationships might be stronger in 1m cells than in our 10 m cells, considering the hummock-hollow microtopography of many peatlands.
+
+Choosing a spatial extent for DSM can be tricky. 
+A natural starting point is the bounding area around a spatial cluster of depth measurements. 
+For a given cluster, kNNDM can define how far the boundary can extend beyond the point measurements; if a cross-validation nearest neighbor distribution cannot match the sample-to-prediction nearest neighbor distribution, the boundary is too expansive [@meyerMachineLearningbasedGlobal2022; @linnenbrinkKNNDMCVKfold2024].
+However, it is unclear how big (N) any cluster of depths should be, and the tradeoff between multiple small-extent DSM and fewer large-extent DSM needs research. 
+Bohn and Miller [-@bohnLocallyEnhancedDigital2024] advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions, where the composition of peatland types is constant [e.g.
+@moenVegetasjonstyper1998].
+Depth varies strongly between peats formed by terrestrialization versus paludification [@buffamFillingHolesRegional2010], and probably along other axes of peatland typology. 
+Although most Norwegian landscapes contain multiple, unknown types, it makes sense to limit their diversity within a single DSM.
+
+### Machine learning approaches can build on success
+
+The DSM literature and our results support using flexible machine learning algorithms like RF to predict peat depth. 
+RF avoids negative predictions [c.f.
+@kogantiMappingPeatDepth2023] and produces good uncertainty estimates [our study, @vaysseUsingQuantileRegression2017;@takoutsingComparingPredictionPerformance2022].
+As depth data become more abundant, we may move from pixel-based learners to deep learning approaches like convolutional neural networks (CNNs). 
+Bakkestuen et al.
+[-@bakkestuenDelineationWetlandAreas2023] successfully predicted peatland extent using CNNs, and their ability to automatically learn multi-scale spatial features would reduce the need to manually engineer these.
+MRVBF's success as a predictor of peat depth demonstrates that multi-scale spatial patterns matter for peat depth and CNNs are designed to learn such patterns. 
+The kind of relationship described in Buffam et al. [-@buffamFillingHolesRegional2010], where peat depth in basins related to terrain slope at the basin edge, is also something a CNN could learn. 
+However, since this approach is data-hungry, we should build soil knowledge into the DSM where we can [@minasnySoilScienceInformedMachine2024]. 
+For example, if further research confirms the effect of the upper marine boundary that we found, then it is better to include the marine boundary as a predictor than to make the algorithm learn this pattern from elevation independently.
+
+### Peat occurrence and depth should be mapped together
+
+Finally, we want to highlight the need for research on peatland occurrence and peat depth to be better integrated. 
+Since peatland occurrence/extent requires non-zero peat depth [the specific threshold varies by definition, @minasnyMappingMonitoringPeatland2023], they are fundamentally linked. 
+The goal, therefore, should be a unified prediction framework for occurrence and depth. 
+We caution against reducing continuous depth predictions to arbitrary classes [as in @ivanovsModelingGeospatialDistribution2024; @karjalainenComparisonTwoGammaray2024], since classes can be derived from continuous predictions.
+The distribution of peat depths across landscapes is zero-inflated, and research is needed to determine whether it is more efficient to parameterize a single model of peat depth (with a larger, generalized dataset) or to break down the problem by classifying zero depth and then regressing non-zero depths (with smaller, specialized datasets).
+Coupling occurrence and depth will reduce the prevalence of incoherence that we found: deep peat outside the peatland extent and zero depth inside it. 
+A key challenge going forward will be obtaining training data that represents both zero and non-zero components of the depth distribution, since sampling designs often focus on known peatlands.

### 2024-12-04 ms/ms.tex
ms: add Discussion

@@
 None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \ref{fig:modelMetrics}).
-For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an \emph{R\textsuperscript{2}} of 0.34, and a mean absolute error of \unit{60\,cm}.
+For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an R\textsuperscript{2} of 0.34, and a mean absolute error of \unit{60\,cm}.
 For Ørskogfjellet the same values were 0.39, 0.33, and \unit{56\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
@@
 
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur.
-Excepteur sint occaecat cupidatat non proident,
-sunt in culpa qui officia deserunt mollit anim id est laborum.
+\subsection{Can we improve Norway's peat depth maps?}
 
-\section{Conclusions}
+\subsubsection{Remotely sensed variables are weak (but not useless!) predictors of peat depth}
 
-Nulla facilisi.
-Maecenas vel nunc nec purus tincidunt congue.
-Proin auctor, lectus eu pharetra malesuada,
-nisi nunc bibendum nunc,
-eget tincidunt nunc nisi id nunc.
-Sed euismod, nunc sit amet aliquam tincidunt,
-nunc nunc tincidunt nunc,
-nec tincidunt nunc nunc nec nunc.
-Donec auctor, nunc sit amet aliquam tincidunt,
-nunc nunc tincidunt nunc,
-nec tincidunt nunc nunc nec nunc.
+Our ability to predict peat depth in the study sites based on terrain and radiometric data was limited.
+Mean absolute errors of 60 and 56 cm at the two sites --- relative to mean depths of 119 and 126 cm --- illustrate the practical limitations of these maps.
+Since any given 10 m cell will miss by about 60 cm, applications requiring detailed peat depth in a small area (e.g.~\textless{} 1 ha) would benefit from measuring depth on the ground rather than relying on the DSM alone.
+
+On the other hand, we improved on the best available map of peat depth (DMK depth class), which is based on field measurements only.
+This highlights the general value of remotely sensed data, whose complete coverage can improve maps even when their association with the variable of interest is weak.
+Since remotely sensed data are widely available, improvements to soil maps as shown here are low hanging fruit.
+This point is recognized and reflected in the rise of DSM.
+
+DMK peat depth classes were a worse predictor of peat depth than our models even though we calibrated them with the same data.
+That is: we gave DMK and the DSM equal footing for a fair comparison.
+If we had taken the DMK peat depth classes at face value and assumed depths according to their class definitions (\textless{} 1 m, \textgreater{} 1 m), they would have performed worse and the advantage of the DSM would be greater.
+The advantage of the DSM was not large in absolute terms (9 cm and 21 cm improvements in mean absolute error), but it explained much more of the variation in depth (improvements in R\textsuperscript{2} of 0.16 and 0.19).
+We attribute this result to the poor spatial and thematic resolution of DMK peat depth, which precludes a robust correlation with peat depths varying from 0-8 m at fine spatial scales.
+
+The performance of our terrain- and radiometric-based maps could have been improved with a spatially explicit mapping approach like regression kriging.
+By ignoring spatial autocorrelation in peat depth (i.e.~the n in scorpan), we have not extracted all of the information about peat depth in the study area out of the training data.
+If we were to put our DSM approach into production for published maps, we would harness the spatial component, but the actual maps for Ørskogfjellet and Skrimfjella are not of primary interest in this study.
+Moreover, residuals of the RF predictions showed weak spatial structure at Skrimfjella, and only up to a range of 150 m at Ørskogfjellet.
+Therefore, the improvement from regression kriging would be small overall and limited to parts of the maps close to measurements.
+
+\subsubsection{Similar error but less explanatory power in Norwegian peatlands}
+
+Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth but generally produced better or comparable error magnitude (R\textsuperscript{2} vs.~MAE/RMSE; \citet{wadouxIntegratedApproachEvaluation2022}).
+It is important to keep in mind that differences in peat depth distributions, spatial scales, and evaluation methods make direct performance comparisons precarious.
+More standardized reporting would help but not eliminate this consideration.
+For example, R\textsuperscript{2} is sensitive to high-leverage extreme values, so it will evaluate a right-skewed distribution differently than a symmetrical distribution.
+We evaluated our models with respect to the explicit purpose of creating peat depth maps across the study areas, but not all studies tailored evaluation to match an explicitly formulated problem \citep{milaNearestNeighbourDistance2022}.
+
+Gatis et al.
+\citeyearpar{gatisMappingUplandPeat2019} used similar predictors and the same spatial grain, finding a much stronger correlation between predicted and observed peat depth (R\textsuperscript{2} = 0.68).
+Although their random evaluation data partition could make performance estimates too optimistic \citep{robertsCrossvalidationStrategiesData2017, wadouxSpatialCrossvalidationNot2021}, the confounding effect of spatial structure is probably small because they used linear regression and few predictors.
+Their model had limited opportunity to overfit to the spatial structure in peat depth.
+The most salient difference in Gatis et al. \citeyearpar{gatisMappingUplandPeat2019} compared to our study is the character of the study area.
+They study a flatter area with a higher proportion of peatland cover, and their peatland is primarily blanket bog.
+The predominance of fens and smaller peatland extent may have contributed to worse performance in our study.
+
+Marchant \citeyearpar{marchantUsingRemoteSensors2021} examined a subset of the area studied in Gatis et al.
+\citeyearpar{gatisMappingUplandPeat2019} at 100 m resolution, using splines to relax linearity between radiometry/terrain and peat depth.
+He found that radK alone predicted peat depth with much higher concordance than our models (CCC = 0.76) and that elevation alone produced comparable performance (CCC = 0.27).
+The RMSE from these univariate models was 46-68 cm (cf.~78 and 75 cm for Skrimfjella and Ørskogfjellet).
+
+Koganti et al. \citeyearpar{kogantiMappingPeatDepth2023} had a peat depth distribution and predictors similar to ours, but at much smaller spatial grain and extent.
+Their training and validation points are closer than the range of spatial autocorrelation in peat depth, so our results are best compared to their non-spatial models.
+They accounted for more variability in peat depth (adjusted \(R^2 = 0.71\)) but had larger errors (RSME = 110 cm).
+Koganti et al.'s \citeyearpar{kogantiMappingPeatDepth2023} linear regression models produced negative predictions, and it is unclear whether the values quoted above include these.
+If we disregard the negative predictions, their model showed the same pattern as ours in overpredicting shallow peats and underpredicting deep peats, although their underprediction was less severe.
+An important difference between Koganti et al.
+\citeyearpar{kogantiMappingPeatDepth2023} and our study (besides spatial scale) is that they measured radiometrics on the ground, rather than using airborne survey data.
+
+\subsubsection{Asymmetries in depth predictions for land use planning and carbon accounting}
+
+Our models erred most for the deepest peats.
+Where overprediction occurred, it was smaller.
+This is not unexpected for the right-skewed distributions of peat depth, but it has management implications.
+Identifying the deepest peats will require additional field work in candidate areas, which could be defined by an upper quantile of predicted depth.
+Map users should not trust the maps to identify all large carbon stocks, but they can trust that identified large stocks really are large.
+That makes the map more suited for ``red-lighting'' than ``green-lighting'' peatland conversion, for example.
+Where it does not prohibit conversion (i.e.~predicts shallow peat), a ground survey should be done before conversion is allowed (assuming depth is the only consideration).
+Although this recommendation aligns with the precautionary principle, here we make it on technical grounds based on the maps' characteristics.
+
+\subsubsection{Depth predictions do not necessarily extrapolate to peat occurrence}
+
+That we measured more than 30 cm of peat in areas not mapped as peatland was expected, because AR5 underrepresents peatland extent by about one third \citep{brynLandCoverNorway2018}.
+For that reason, maps of peatland occurrence also need revising.
+Unfortunately, our models did not predict peat occurrence outside mapped peatlands.
+In these areas, assuming a constant 30 cm of peat (the depth threshold used in AR5) produced less error than our models.
+Moreover, the performance gap would likely have been larger had our evaluation data been representative of the non-peatland areas.
+Instead they were near mapped peatlands.
+Most importantly, the independent occurrence data at Skrimfjella showed the model completely failing to predict occurrence, and these data were suited for testing this ability.
+Although the Ørskogfjellet model may have done better than the Skrimfjella model on a similar independent evaluation set (since it predicted depth better), the improvement would probably be small.
+
+Why wasn't peat occurrence predicted well?
+The models were trained on fundamentally different populations of locations (\textasciitilde80 \% mapped peatland in training vs.~0 \% in prediction), so it is not surprising that the associations they learned did not transfer well.
+Moreover, the \textasciitilde20 \% of training data from outside mapped peatland were incidentally collected (near peatland) and not representative of other land cover classes.
+In short, the models were blind to the fact that most of both study areas have no peat.
+
+Peatland extent is probably best mapped using different remotely sensed data than we used here \citep{bakkestuenDelineationWetlandAreas2023}, and this study's purpose was not to map extent.
+Nevertheless --- as long as the relevant peatland definition includes a depth component --- depth predictions (or predictors with subsurface information) should help delineate peatland extent \citep{olearyDigitalSoilMapping2022, beamishDetailedMappingPeat2024}.
+We return to this point in our discussion of implications for digital soil mapping.
+
+\subsection{Which variables predict peat depth?}
+
+\subsubsection{Airborne radiometrics do not predict Norwegian peat depth}
+
+Radiometric data had no predictive value at Ørskogfjellet, while at Skrimfjella they had minor influence in a relatively weak model.
+The bedrock is more homogeneous at Ørskogfjellet than at Skrimfjella, so uneven radiogenesis is not a viable explanation for the differences between sites nor the poor performance in general \citep{beamishEnvironmentalRadioactivityUK2014, reinhardtGammaraySpectrometryVersatile2019}.
+To the extent that radiometrics had predictive value at Skrimfjella, it appears that they were most valuable near the extremes of the depth distribution, since their inclusion improved R\textsuperscript{2} more than mean absolute error.
+All four variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value.
+This contrasts with Koganti et al. \citeyearpar{kogantiMappingPeatDepth2023}, who found that radTC was a much better predictor than radK.
+
+We suspect that the primary reason for the poor predictive value of the radiometric data was the large footprint of the detector in the airborne survey.
+With an average flight altitude of 75 m, less than half of the radiation reaching the detector comes from inside the 100 m diameter circle directly below it \citep{beamishEnhancingResolutionAirborne2016, beamishDetailedMappingPeat2024}.
+The rest of the measured activity integrates a much wider area.
+For comparison, empirical variograms of peat depth at Skrimfjella and Ørskogfjellet showed no spatial autocorrelation beyond 50 and 75 m (among GPR data) or 110 and 230 m (among 10 m cells).
+Basically, the airborne radiometric data will not capture large variation over short (\textless{} 100 m) distances; the instrument's field of view has a large smoothing effect on the data \citep{beamishEnhancingResolutionAirborne2016, reinhardtGammaraySpectrometryVersatile2019}.
+Different landforms and the changes they cause in the geometry between the radioactive source and the detector can also distort airborne measurements \citep{reinhardtGammaraySpectrometryVersatile2019}.
+
+Studies comparing airborne and ground radiometric surveys confirm that they are poorly correlated in low-activity areas like peatlands \citep{kockComparisonAirborneTerrestrial2011, karjalainenComparisonTwoGammaray2024}.
+Karjalainen et al. \citeyearpar{karjalainenComparisonTwoGammaray2024} found that ground-based measurements predicted peat depth better than airborne measurements.
+Nevertheless, a large radiometric footprint did not prevent strong associations with peat depth in Gatis et al.
+\citeyearpar{gatisMappingUplandPeat2019} and Marchant \citeyearpar{marchantUsingRemoteSensors2021}, likely due to the extensive blanket bog landscape with more gradual changes in depth \citep{lindsayBogsEcologyClassification1995, lyngstadBeskrivelserAvTorvmassivenheter2023}.
+We are unsure whether short-range depth changes explain the weak associations that Siemon et al. \citeyearpar{siemonAirborneElectromagneticRadiometric2020} found in a large raised bog.
+
+Weather conditions varied during the Ørskogfjellet radiometric survey and affected its data \citep{ofstadHelicopterborneMagneticRadiometric2015}.
+Thus, uneven snow cover and air moisture may also have masked the soil signal in these data.
+
+We do not believe that the poor predictive value of the radiometric data in this study was caused by fully attenuated radioactivity --- at least not in large part.
+The RF algorithm's flexibility means that radiometrics could be used for shallower peats if they provided predictive value for that part of the depth distribution, but there is no evidence of that in our results.
+In the partial dependence plot of radU at Skrimfjella, the expected negative relationship between depth and radU cannot be found by ignoring the left (highly-attenuated) side of the distribution.
+About a quarter of the peats in our study were less than a meter deep, and full attenuation is unlikely for these \citep{beamishGammaRayAttenuation2013}.
+
+Although we do not believe full attenuation is the primary reason for poor performance in our study, it may limit peat depth mapping under other circumstances.
+First-principle calculations suggest that radiation should be 90 \% attenuated after about 50-60 cm of typical, wet peat or 85 cm of unnaturally dry peat \citep{beamishGammaRayAttenuation2013, beamishDetailedMappingPeat2024}, and some field tests support these values \citep{billenEignungGammaspektrometrieKartieren2015}.
+It is remarkable that particular studies detected radiation differences up to several meters deep \citep{gatisMappingUplandPeat2019, kogantiMappingPeatDepth2023}, but these may be the exceptions rather than the rule.
+Perhaps relatively deeper water tables in these study sites \citep[blanket bog, drained fen,][]{pricePeatlandRestorationHydrology2016} contributed to better penetration.
+
+\subsubsection{Terrain-based variables can predict peat depth}
+
+At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables.
+Elevation was the most important predictor at both sites, and peat depth showed non-monotonic responses to changes in elevation.
+We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g.~via climate) to explain the observed patterns.
+Indeed, elevation at Ørskogfjellet seems to interact with other variables (non-parallel ICE lines), complicating its interpretation.
+Moreover, relationships between elevation and peat depth have previously shown opposite shapes in different areas \citep{finlaysonEstimatingOrganicSurface2021}.
+Nonetheless, a relationship that is not generalizable beyond the mapping area is still useful for DSM, as long as it is evaluated to demonstrate its robustness for the predictive task (e.g.~through kNNDM spatial cross validation).
+
+One interesting feature of the elevational relationships we found may be generalizable: a steep increase in peat depth near the upper marine limit after the last ice age.
+At Ørskogfjellet, the upper marine limit is about 75 meters above today's sea level \citep[Geological Survey of Norway,][]{hogaasDatabaseRegistreringAv2012}, where the partial dependence plot of elevation shows a sharp increase in peat depth.
+In areas under the upper marine limit there has been less time for peat accumulation since the ice sheets retreated, and it is plausible that this makes peats there shallower, all else being equal.
+We cannot evaluate this effect at Skrimfjella, where the upper marine limit is below our study area (at 175 m.a.s.l.).
+
+Another influential terrain-based predictor was MRVBF.
+Unlike elevation, it showed a monotonic effect on peat depth: greater valley bottom flatness was always associated with increases in peat depth.
+Delineating a valley bottom involves ambiguity, but the MRVBF index is a pragmatic approach that considers a location a valley bottom if it is sufficiently low and flat at a particular scale \citep{gallantMultiresolutionIndexValley2003}.
+The multiresolution nature of the index allows small elevated but flat areas (including saddles) to be characterized as having high valley bottom flatness \citep{gallantMultiresolutionIndexValley2003}.
+Our results suggest that MRVBF is a robust indicator of high water tables (and peat accumulation) over millennial time scales, corroborating other studies \citep{rudiyantoOpenDigitalMapping2018, deragonMappingMaximumPeat2023}.
+
+Other terrain-derived predictors with predictive value in our study are hydrological (Topographic Wetness Index and Depth-To-Water index).
+Notably, terrain slope was inferior to (Ørskogfjellet) or highly correlated with (Skrimfjella) these hydrological indices.
+Mappers of peat depth should not assume that slope is the best predictor in its class, despite its prevalence in the literature.
+Wetter locations (high TWI and low DTW) were generally associated with deeper peat, but these relationships were not as strong or consistent as with MRVBF.
+The optimal scale for Topographic Wetness Index and Depth-To-Water index varied, and likely depends on both the dominant peat formation processes and the typical size of peatland features in a landscape.
+Including multiple scales of these variables allows the model to capture different hydrological mechanisms operating at different spatial scales.
+
+\subsubsection{Legacy depth maps have inconsistent predictive value}
+
+DMK peat depth class proved an inconsistent predictor of peat depth.
+At Skrimfjella, it barely improved model performance.
+At Ørskogfjellet, it increased performance more, and both indicator variables were among the most important in the model.
+We suspect the discrepancy between sites is due to different levels of effort and coverage during the historical surveys; more lowland peatland near agriculture at Ørskogfjellet may have caused more purposeful surveying.
+73 \% of cells measured at Skrimfjella had unknown depth in DMK, compared to 20 \% at Ørskogfjellet.
+Still, the 25 \% of cells classified as deep at Skrimfjella were not meaningfully deeper than the rest, according to our model.
+Interactions between DMK and other variables underline the inconsistency of DMK depth maps, even within a site.
+That a peat at Ørskogfjellet was classified as shallow rather than deep did not uniformly lower the model's expectation of depth there.
+Rather, the veracity of the deep-indication was contingent on other factors.
+
+\subsection{Implications for digital soil mapping of peat depth}
+
+The performance gap between the best models and the DMK-only models shows that peat depth in Norway should be mapped digitally.
+Anywhere we have some calibrating measurements, we can get better maps than DMK peat depth classes, at low cost.
+Moreover, DSM methodology can align map products with open science principles by making their production transparent, reproducible, and updatable.
+The large difference we found between the coverage and quality of DMK peat depth at Skrimfjella versus Ørskogfjellet underlines these advantages.
+With DSM we can apply the same approach across different areas and make maps with full spatial coverage, continuous values, and validated uncertainty.
+The rest of this section discusses recommendations and needs for more extensive peat depth DSM.
+
+\subsubsection{Peat depth measurements should be organized}
+
+High-quality digital terrain models are available for mainland Norway, making peat depth measurements the critical training data need.
+For a given area, some minimum number of measurements is necessary to create meaningful improvement over DMK peat depth.
+The proportion of peatlands sampled at Skrimfjella was twice that at Ørskogfjellet, but the Skrimfjella model performed worse, which shows that the size of the depth dataset is not all-important.
+We had the luxury of stratifying our measurements over candidate predictors, and performance may suffer where locations are opportunistic or tied to a sampling design with a different purpose.
+The magnitude of this penalty will depend on the dataset characteristics, but having enough depth measurements is probably more consequential \citep{wadouxSamplingDesignOptimization2019}.
+
+Depth measurements are foundational, and better infrastructure to collect, preserve, and make FAIR these data would help DSM and other applications.
+Geoportal access and data exchange standards \citeyearpar[like Natural England's for peat surveys,][]{naturalenglandDataExchangeStandard2023} are important.
+Peat data often fall through the cracks between geology-oriented and ecology-oriented archives, but increased awareness of peatland importance is a good impetus to remedy this situation.
+Peat depth is quick and easy to measure, so integrating its measurement into existing national field programs, like Norway's spatially representative nature monitoring (ANO) or national forest inventory, would be advantageous (although not sufficient for regional or local mapping).
+Municipalities and local actors in Norway are increasingly measuring peat depth and can contribute to a growing data foundation \citep{kyrkjeeideCalculatorLocalPeatland2023}.
+All of this lays the groundwork for renewed peatland maps.
+
+\subsubsection{Spatial scale affects model performance and utility}
+
+The scarcity of depth measurements and their short spatial autocorrelation mean that mapping peat depth at broad scales is not an exercise in spatial interpolation \citep{henglGenericFrameworkSpatial2004}.
+Peat depths in our GPR data showed spatial autocorrelation to a range of 50-100 m, and getting measurements at such fine grain is only realistic for small areas, not across landscapes.
+Therefore, we anticipate that spatially-explicit DSM is of limited value in all but the most intensively sampled landscapes (currently nowhere in Norway).
+Fine-scale variation in peat depth raises the possibility of mapping at very fine resolution (1 m) --- even if land use planning and carbon accounting do not operate at this grain.
+Terrain-depth relationships might be stronger in 1m cells than in our 10 m cells, considering the hummock-hollow microtopography of many peatlands.
+
+Choosing a spatial extent for DSM can be tricky.
+A natural starting point is the bounding area around a spatial cluster of depth measurements.
+For a given cluster, kNNDM can define how far the boundary can extend beyond the point measurements; if a cross-validation nearest neighbor distribution cannot match the sample-to-prediction nearest neighbor distribution, the boundary is too expansive \citep{meyerMachineLearningbasedGlobal2022, linnenbrinkKNNDMCVKfold2024}.
+However, it is unclear how big (N) any cluster of depths should be, and the tradeoff between multiple small-extent DSM and fewer large-extent DSM needs research.
+Bohn and Miller \citeyearpar{bohnLocallyEnhancedDigital2024} advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions, where the composition of peatland types is constant \citep[e.g.][]{moenVegetasjonstyper1998}.
+Depth varies strongly between peats formed by terrestrialization versus paludification \citep{buffamFillingHolesRegional2010}, and probably along other axes of peatland typology.
+Although most Norwegian landscapes contain multiple, unknown types, it makes sense to limit their diversity within a single DSM.
+
+\subsubsection{Machine learning approaches can build on success}
+
+The DSM literature and our results support using flexible machine learning algorithms like RF to predict peat depth.
+RF avoids negative predictions \citep[c.f.][]{kogantiMappingPeatDepth2023} and produces good uncertainty estimates \citep[our study,][]{vaysseUsingQuantileRegression2017, takoutsingComparingPredictionPerformance2022}.
+As depth data become more abundant, we may move from pixel-based learners to deep learning approaches like convolutional neural networks (CNNs).
+Bakkestuen et al.
+\citeyearpar{bakkestuenDelineationWetlandAreas2023} successfully predicted peatland extent using CNNs, and their ability to automatically learn multi-scale spatial features would reduce the need to manually engineer these.
+MRVBF's success as a predictor of peat depth demonstrates that multi-scale spatial patterns matter for peat depth and CNNs are designed to learn such patterns.
+The kind of relationship described in Buffam et al.
+\citeyearpar{buffamFillingHolesRegional2010}, where peat depth in basins related to terrain slope at the basin edge, is also something a CNN could learn.
+However, since this approach is data-hungry, we should build soil knowledge into the DSM where we can \citep{minasnySoilScienceInformedMachine2024}.
+For example, if further research confirms the effect of the upper marine boundary that we found, then it is better to include the marine boundary as a predictor than to make the algorithm learn this pattern from elevation independently.
+
+\subsubsection{Peat occurrence and depth should be mapped together}
+
+Finally, we want to highlight the need for research on peatland occurrence and peat depth to be better integrated.
+Since peatland occurrence/extent requires non-zero peat depth \citep[the specific threshold varies by definition,][]{minasnyMappingMonitoringPeatland2023}, they are fundamentally linked.
+The goal, therefore, should be a unified prediction framework for occurrence and depth.
+We caution against reducing continuous depth predictions to arbitrary classes \citep[as in][]{ivanovsModelingGeospatialDistribution2024, karjalainenComparisonTwoGammaray2024}, since classes can be derived from continuous predictions.
+The distribution of peat depths across landscapes is zero-inflated, and research is needed to determine whether it is more efficient to parameterize a single model of peat depth (with a larger, generalized dataset) or to break down the problem by classifying zero depth and then regressing non-zero depths (with smaller, specialized datasets).
+Coupling occurrence and depth will reduce the prevalence of incoherence that we found: deep peat outside the peatland extent and zero depth inside it.
+A key challenge going forward will be obtaining training data that represents both zero and non-zero components of the depth distribution, since sampling designs often focus on known peatlands.
 

### 2024-12-05 ms/ms.Rmd
ms: minor Discussion edits

@@
 Peat data often fall through the cracks between geology-oriented and ecology-oriented archives, but increased awareness of peatland importance is a good impetus to remedy this situation. 
-Peat depth is quick and easy to measure, so integrating its measurement into existing national field programs, like Norway’s spatially representative nature monitoring (ANO) or national forest inventory, would be advantageous (although not sufficient for regional or local mapping).
+Peat depth is quick and easy to measure, so integrating its measurement into existing national field programs, like Norway’s spatially representative nature monitoring (ANO) or national forest inventory, would be helpful (although not sufficient for regional or local mapping).
 Municipalities and local actors in Norway are increasingly measuring peat depth and can contribute to a growing data foundation [@kyrkjeeideCalculatorLocalPeatland2023]. 
-All of this lays the groundwork for renewed peatland maps.
+Low-altitude, drone-mounted GPR may prove an efficient approach for collecting many, accurate depth data in a landscape, by combining the advantages of airborne deployment and active sensing [@pelletierPeatAnalysesHudson1991; @ruolsDevelopmentDronebasedGroundpenetrating2023].
+All of the above can lay the groundwork for renewed peatland maps.
 

### 2024-12-05 ms/ms.tex
ms: minor Discussion edits

@@
 Peat data often fall through the cracks between geology-oriented and ecology-oriented archives, but increased awareness of peatland importance is a good impetus to remedy this situation.
-Peat depth is quick and easy to measure, so integrating its measurement into existing national field programs, like Norway's spatially representative nature monitoring (ANO) or national forest inventory, would be advantageous (although not sufficient for regional or local mapping).
+Peat depth is quick and easy to measure, so integrating its measurement into existing national field programs, like Norway's spatially representative nature monitoring (ANO) or national forest inventory, would be helpful (although not sufficient for regional or local mapping).
 Municipalities and local actors in Norway are increasingly measuring peat depth and can contribute to a growing data foundation \citep{kyrkjeeideCalculatorLocalPeatland2023}.
-All of this lays the groundwork for renewed peatland maps.
+Low-altitude, drone-mounted GPR may prove an efficient approach for collecting many, accurate depth data in a landscape, by combining the advantages of airborne deployment and active sensing \citep{pelletierPeatAnalysesHudson1991, ruolsDevelopmentDronebasedGroundpenetrating2023}.
+All of the above can lay the groundwork for renewed peatland maps.
 

### 2024-12-05 ms/ms.Rmd
ms: add abstract

@@
 abstract: |
-  The abstract goes here.
-  It can also be on _multiple lines_.
+  Peatlands are Earth's most carbon-dense terrestrial ecosystems and crucial stores of soil carbon. 
+  Their carbon density results largely from their depth, which can exceed ten meters. 
+  However, peat depth is poorly mapped at scales relevant for land management and carbon accounting. 
+  We evaluated whether digital soil mapping using remotely sensed data could improve existing maps of peat depth in Norway. 
+  Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data at two contrasting sites. 
+  We measured peat depth at 372 and 1878 locations using manual probing and ground-penetrating radar, then trained Random Forest models to predict depth at 10 m resolution from various combinations of predictors.
+  The best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth. 
+  Terrain variables were more important predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth. 
+  Adding radiometric predictors improved model performance at one site but degraded it at the other. 
+  Our models improved upon Norway's existing peat depth maps, but the weak relationships between peat depth and remotely sensed variables demonstrate limitations to mapping without extensive field measurements.
+  Models trained only on known peatlands could not reliably identify peat presence elsewhere, highlighting the need to map peat occurrence and depth together.
 bibliography: ms.bib

### 2024-12-05 ms/ms.tex
ms: add abstract

@@
 \begin{abstract}
-The abstract goes here.
-It can also be on \emph{multiple lines}.
+Peatlands are Earth's most carbon-dense terrestrial ecosystems and crucial stores of soil carbon.
+Their carbon density results largely from their depth, which can exceed ten meters.
+However, peat depth is poorly mapped at scales relevant for land management and carbon accounting.
+We evaluated whether digital soil mapping using remotely sensed data could improve existing maps of peat depth in Norway.
+Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data at two contrasting sites.
+We measured peat depth at 372 and 1878 locations using manual probing and ground-penetrating radar, then trained Random Forest models to predict depth at 10 m resolution from various combinations of predictors.
+The best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth.
+Terrain variables were more important predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth.
+Adding radiometric predictors improved model performance at one site but degraded it at the other.
+Our models improved upon Norway's existing peat depth maps, but the weak relationships between peat depth and remotely sensed variables demonstrate limitations to mapping without extensive field measurements.
+Models trained only on known peatlands could not reliably identify peat presence elsewhere, highlighting the need to map peat occurrence and depth together.
 \end{abstract}

### 2024-12-05 ms/ms.Rmd
ms: add groups to preds table

@@
 df <- readr::read_csv("tables/predictors.csv")
+options(knitr.kable.NA = '')
 knitr::kable(df, booktabs = FALSE, vline = "", linesep = "",
-             col.names = c("Name", "Description"),
+             col.names = c("Group", "Name", "Description"),
              caption = "Candidate predictors of peat depth.")

### 2024-12-05 ms/ms.tex
ms: add groups to preds table

@@
 \centering
-\begin{tabular}[t]{ll}
+\begin{tabular}[t]{lll}
 \hline
-Name & Description\\
+Group & Name & Description\\
 \hline
-radK & Potassium ground concentration\\
-radTh & Thorium ground concentration\\
-radU & Uranium ground concentration\\
-radTC & Total count of gamma radiation\\
-elevation & Mean elevation\\
-slope1m & Mean of 1 m slope\\
-TPI1m & Mean of 1 m topographic position index\\
-TRI1m & Mean of 1 m terrain ruggedness index\\
-roughness1m & Mean of 1 m roughness\\
-slope10m & 10 m slope\\
-TPI10m & 10 m topographic position index\\
-TRI10m & 10 m terrain ruggedness index\\
-roughness10m & 10 m roughness\\
-MRVBF & Multi-resolution valley bottom flatness\\
-TWI5m & Mean of 5 m topographic wetness index\\
-TWI10m & 10 m topographic wetness index\\
-TWI20m & Bilinear interpolation of 20 m topographic wetness index\\
-TWI50m & Bilinear interpolation of 50 m topographic wetness index\\
-DTW2500 & Depth-to-water index, flow initiation area of 0.25 ha\\
-DTW5000 & Depth-to-water index, flow initiation area of 0.5 ha\\
-DTW10000 & Depth-to-water index, flow initiation area of 1 ha\\
-DTW20000 & Depth-to-water index, flow initiation area of 2 ha\\
-DTW40000 & Depth-to-water index, flow initiation area of 4 ha\\
-DTW80000 & Depth-to-water index, flow initiation area of 8 ha\\
-DTW160000 & Depth-to-water index, flow initiation area of 16 ha\\
-DMK & DMK peat depth class, categorical with 3 levels\\
+radiometric & radK & Potassium ground concentration\\
+ & radTh & Thorium ground concentration\\
+ & radU & Uranium ground concentration\\
+ & radTC & Total count of gamma radiation\\
+terrain & elevation & Mean elevation\\
+ & slope1m & Mean of 1 m slope\\
+ & TPI1m & Mean of 1 m topographic position index\\
+ & TRI1m & Mean of 1 m terrain ruggedness index\\
+ & roughness1m & Mean of 1 m roughness\\
+ & slope10m & 10 m slope\\
+ & TPI10m & 10 m topographic position index\\
+ & TRI10m & 10 m terrain ruggedness index\\
+ & roughness10m & 10 m roughness\\
+ & MRVBF & Multi-resolution valley bottom flatness\\
+ & TWI5m & Mean of 5 m topographic wetness index\\
+ & TWI10m & 10 m topographic wetness index\\
+ & TWI20m & Bilinear interpolation of 20 m topographic wetness index\\
+ & TWI50m & Bilinear interpolation of 50 m topographic wetness index\\
+ & DTW2500 & Depth-to-water index, flow initiation area of 0.25 ha\\
+ & DTW5000 & Depth-to-water index, flow initiation area of 0.5 ha\\
+ & DTW10000 & Depth-to-water index, flow initiation area of 1 ha\\
+ & DTW20000 & Depth-to-water index, flow initiation area of 2 ha\\
+ & DTW40000 & Depth-to-water index, flow initiation area of 4 ha\\
+ & DTW80000 & Depth-to-water index, flow initiation area of 8 ha\\
+ & DTW160000 & Depth-to-water index, flow initiation area of 16 ha\\
+existing map & DMK & DMK peat depth class, categorical with 3 levels\\
 \hline

### 2024-12-05 ms/ms.Rmd
ms: title and minor edits

@@
 ---
-title: Wall-to-wall mapping of peat depth from Lidar terrain and airborne radiometrics in Norwegian landscapes
+title: Terrain is a stronger predictor of peat depth than airborne radiometrics in Norwegian landscapes
 journal: "`r rticles::copernicus_journal_abbreviations(journal_name = 'soil')`"
@@
   Adding radiometric predictors improved model performance at one site but degraded it at the other. 
-  Our models improved upon Norway's existing peat depth maps, but the weak relationships between peat depth and remotely sensed variables demonstrate limitations to mapping without extensive field measurements.
+  Our models improved upon the national peat depth map, but the weak relationships between peat depth and remotely sensed variables demonstrate limitations to mapping without extensive field measurements.
   Models trained only on known peatlands could not reliably identify peat presence elsewhere, highlighting the need to map peat occurrence and depth together.

### 2024-12-05 ms/ms.tex
ms: title and minor edits

@@
 
-\title{Wall-to-wall mapping of peat depth from Lidar terrain and airborne radiometrics in Norwegian landscapes}
+\title{Terrain is a stronger predictor of peat depth than airborne radiometrics in Norwegian landscapes}
 
@@
 Adding radiometric predictors improved model performance at one site but degraded it at the other.
-Our models improved upon Norway's existing peat depth maps, but the weak relationships between peat depth and remotely sensed variables demonstrate limitations to mapping without extensive field measurements.
+Our models improved upon the national peat depth map, but the weak relationships between peat depth and remotely sensed variables demonstrate limitations to mapping without extensive field measurements.
 Models trained only on known peatlands could not reliably identify peat presence elsewhere, highlighting the need to map peat occurrence and depth together.
@@
  & DTW160000 & Depth-to-water index, flow initiation area of 16 ha\\
-existing map & DMK & DMK peat depth class, categorical with 3 levels\\
+national map & DMK & DMK peat depth class, categorical with 3 levels\\
 \hline

### 2024-12-05 ms/ms.Rmd
ms: revise abstract

@@
 abstract: |
-  Peatlands are Earth's most carbon-dense terrestrial ecosystems and crucial stores of soil carbon. 
-  Their carbon density results largely from their depth, which can exceed ten meters. 
-  However, peat depth is poorly mapped at scales relevant for land management and carbon accounting. 
-  We evaluated whether digital soil mapping using remotely sensed data could improve existing maps of peat depth in Norway. 
+  Peatlands are Earth's most carbon-dense terrestrial ecosystems and their carbon density varies with the depth of the peat layer. 
+  Accurate mapping of peat depth is crucial for carbon accounting and land management, yet existing maps lack the resolution and accuracy needed for these applications. 
+  This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in Norway. 
   Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data at two contrasting sites. 
-  We measured peat depth at 372 and 1878 locations using manual probing and ground-penetrating radar, then trained Random Forest models to predict depth at 10 m resolution from various combinations of predictors.
+  We measured peat depth at 372 and 1878 locations and trained Random Forest models to predict depth at 10 m resolution. 
   The best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth. 
-  Terrain variables were more important predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth. 
-  Adding radiometric predictors improved model performance at one site but degraded it at the other. 
-  Our models improved upon the national peat depth map, but the weak relationships between peat depth and remotely sensed variables demonstrate limitations to mapping without extensive field measurements.
-  Models trained only on known peatlands could not reliably identify peat presence elsewhere, highlighting the need to map peat occurrence and depth together.
+  Terrain variables were better predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth. 
+  Radiometric data showed inconsistent value, improving predictions at one site while degrading performance at the other. 
+  Our models improved upon the national peat depth map, but weak relationships with remotely sensed variables make peat depth hard to predict. 
+  These findings suggest that while digital soil mapping can enhance existing peat depth maps, reliable carbon stock assessments still require strategic field sampling. 
+  Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat occurrence and depth.
 bibliography: ms.bib

### 2024-12-05 ms/ms.tex
ms: revise abstract

@@
 \begin{abstract}
-Peatlands are Earth's most carbon-dense terrestrial ecosystems and crucial stores of soil carbon.
-Their carbon density results largely from their depth, which can exceed ten meters.
-However, peat depth is poorly mapped at scales relevant for land management and carbon accounting.
-We evaluated whether digital soil mapping using remotely sensed data could improve existing maps of peat depth in Norway.
+Peatlands are Earth's most carbon-dense terrestrial ecosystems and their carbon density varies with the depth of the peat layer.
+Accurate mapping of peat depth is crucial for carbon accounting and land management, yet existing maps lack the resolution and accuracy needed for these applications.
+This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in Norway.
 Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data at two contrasting sites.
-We measured peat depth at 372 and 1878 locations using manual probing and ground-penetrating radar, then trained Random Forest models to predict depth at 10 m resolution from various combinations of predictors.
+We measured peat depth at 372 and 1878 locations and trained Random Forest models to predict depth at 10 m resolution.
 The best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth.
-Terrain variables were more important predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth.
-Adding radiometric predictors improved model performance at one site but degraded it at the other.
-Our models improved upon the national peat depth map, but the weak relationships between peat depth and remotely sensed variables demonstrate limitations to mapping without extensive field measurements.
-Models trained only on known peatlands could not reliably identify peat presence elsewhere, highlighting the need to map peat occurrence and depth together.
+Terrain variables were better predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth.
+Radiometric data showed inconsistent value, improving predictions at one site while degrading performance at the other.
+Our models improved upon the national peat depth map, but weak relationships with remotely sensed variables make peat depth hard to predict.
+These findings suggest that while digital soil mapping can enhance existing peat depth maps, reliable carbon stock assessments still require strategic field sampling.
+Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat occurrence and depth.
 \end{abstract}

### 2024-12-05 ms/ms.Rmd
ms: hyphens and abbreviations

@@
   Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data at two contrasting sites. 
-  We measured peat depth at 372 and 1878 locations and trained Random Forest models to predict depth at 10 m resolution. 
+  We measured peat depth at 372 and 1878 locations and trained random forest models to predict depth at 10 m resolution. 
   The best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth. 
@@
 Peat soils are a terrestrial carbon stock of global importance. 
-They store about 30 % of global soil carbon despite covering only 2-3 % of Earth's land [@xuPEATMAPRefiningEstimates2018; @friedlingsteinGlobalCarbonBudget2020; @unepGlobalPeatlandsAssessment2022]. 
+They store about 30 % of global soil carbon despite covering only 2--3 % of Earth's land [@xuPEATMAPRefiningEstimates2018; @friedlingsteinGlobalCarbonBudget2020; @unepGlobalPeatlandsAssessment2022]. 
 In other words, they are extremely carbon dense -- denser than any other ecosystem per square meter [@temminkRecoveringWetlandBiogeomorphic2022]. 
@@
 However, the short range of spatial autocorrelation in peat depth limits its mapping value [@henglGenericFrameworkSpatial2004]. 
-Relief, or topography, is widely and accurately mapped in digital terrain models. 
-For example, most of mainland Norway has at least one elevation measurement per square meter from airborne LiDAR surveys. 
+Relief, or topography, is widely and accurately mapped in digital terrain models (DTMs). 
+For example, most of mainland Norway has at least one elevation measurement per square meter from airborne Light Detection and Ranging (LiDAR) surveys. 
 Moreover, mechanisms of peat formation are linked to topography. 
@@
 The little we know about the depths of Norwegian peatlands comes from surveys meant to identify arable land. 
-After scattered surveys in the early 20th century, a comprehensive round of surveying was completed 1964-2001 as part of a wider land cover mapping in Norway  [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
+After scattered surveys in the early 20th century, a comprehensive round of surveying was completed 1964--2001 as part of a wider land cover mapping in Norway  [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
 Because of its agricultural and silvicultural focus, it covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable [@ahlstromAR5Klassifikasjonssystem2019].
-Field surveyors carried a one-meter probe, so they measured peat depth as shallow (< 1 m), or deep (> 1 m). 
+Field surveyors carried a 1 m probe, so they measured peat depth as shallow (< 1 m), or deep (> 1 m). 
 These classes were assigned to whole mire polygons, so spatial resolution is on the order of hectares.
@@
 We chose for our sampling and modelling a spatial resolution of \unit{10\,m}.
-We considered this a reasonable compromise between digital terrain model (DTM) resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.
+We considered this a reasonable compromise between DTM resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.
 The point measurements of peat depth described below were ultimately aggregated to \unit{10\,m} resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
@@
 Field work at Skrimfjella was conducted in August 2020.
-We navigated to the centers of the raster cells in the depth and occurrence samples by handheld GPS, checking that positional error was below \unit{3\,m}.
+We navigated to the centers of the raster cells in the depth and occurrence samples with a handheld global navigation satellite system (GNSS) receiver, checking that positional error was below \unit{3\,m}.
 For each depth sample location, we measured peat depth three times (at the vertices of a triangle with \unit{2\,m} sides) to get a more representative value for the \unit{10\,m} raster cell, and to dampen the effect of outlying measurements [@parryEvaluatingApproachesEstimating2014].
@@
 Occasionally, when the soil itself was difficult to judge, we made our determination also based on the presence or absence of mire vegetation.
-Although peat soil is strictly defined by organic content (which we did not analyse), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland [@minasnyMappingMonitoringPeatland2023].
+Although peat soil is strictly defined by organic content (which we did not analyze), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland [@minasnyMappingMonitoringPeatland2023].
 
-Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively-chosen, individual mires, using ground-penetrating radar (GPR).
+Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively chosen, individual mires, using ground-penetrating radar (GPR).
 We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with its \unit{500\,MHz} shielded antenna mounted in a plastic sledge, and its control unit connected to a GNSS receiver.
@@
 This approach has been found to identify sample sizes that correspond with diminishing returns in predictive model performance on external evaluation data [@sauretteDivergenceMetricsDetermining2023].
-We defined a sequence of sample sizes (50--500) and for ten replicate samples at each size [drawn by conditioned latin hypercube sampling, @minasnyConditionedLatinHypercube2006; @roudierClhsPackageConditioned2011], we calculated the mean Kullback-Leibler divergence between sample and population distributions [@maloneMethodsImproveUtility2019; @sauretteDivergenceMetricsDetermining2023].
+We defined a sequence of sample sizes (50--500) and for ten replicate samples at each size [drawn by conditioned latin hypercube sampling, @minasnyConditionedLatinHypercube2006; @roudierClhsPackageConditioned2011], we calculated the mean Kullback--Leibler divergence between sample and population distributions [@maloneMethodsImproveUtility2019; @sauretteDivergenceMetricsDetermining2023].
 The predictors in the divergence calculation were terrain slope and four radiometrics: potassium, thorium, uranium, and total count.
@@
 Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
-To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly-accessible roads.
+To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly accessible roads.
 Cost distance was calculated using GRASS's *r.walk* function, with friction costs defined by AR5 land classes [@GRASSv8-2].
@@
 We know that peat depth tends to vary at fine scales in Norway, which is why we chose \unit{1\,m} and \unit{10\,m} resolutions [@maxwellLandsurfaceParametersSpatial2022].
-We also calculated the the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm [@gallantMultiresolutionIndexValley2003].
+We also calculated the MRVBF index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm [@gallantMultiresolutionIndexValley2003].
 We calculated this index in SAGA GIS [v.9.3.2, @conradSystemAutomatedGeoscientific2015] with default parameters.
@@
 Next, we calculated the Topographic Wetness Index [@quinnPredictionHillslopeFlow1991].
-This index is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate-to-coarse resolution DTMs [@agrenEvaluatingDigitalTerrain2014; @riihimakiTopographicWetnessIndex2021], so we calculated it from \unit{5\,m}, \unit{10\,m}, \unit{20\,m}, and \unit{50\,m} DTM resolution.
+This index is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate to coarse resolution DTMs [@agrenEvaluatingDigitalTerrain2014; @riihimakiTopographicWetnessIndex2021], so we calculated it from \unit{5\,m}, \unit{10\,m}, \unit{20\,m}, and \unit{50\,m} DTM resolution.
 The calculations were performed with Whitebox software [@lindsayWhiteboxGATCase2016], accessed through the *whitebox* R package [v2.4, @wuWhiteboxWhiteboxToolsFrontend2022].
@@
 
-The last terrain-based predictor we included was the depth-to-water index [@murphyMappingWetlandsComparison2007].
+The last terrain-based predictor we included was the depth-to-water (DTW) index [@murphyMappingWetlandsComparison2007].
 This index approximates a location's vertical height above the surface water feature that it is likely to drain towards.
@@
 In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
-For each of the seven surface water layers, we derived the depth-to-water index using the *Distance Accumulation* tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.
+For each of the seven surface water layers, we derived DTW using the *Distance Accumulation* tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.
 
@@
 It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance [@khaledianSelectingAppropriateMachine2020].
-Evaluating variable importance in a maximally-predictive model aligns with the aim of this study.
+Evaluating variable importance in a maximally predictive model aligns with the aim of this study.
 
 RF by itself is not a spatial model, and it will only predict spatial structure in the outcome to the degree that the structure is captured by predictors.
-We considered using regression kriging -- a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals [@henglGenericFrameworkSpatial2004].
+We considered using regression kriging -- a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically interpolated surface of RF residuals [@henglGenericFrameworkSpatial2004].
 The spatial component in regression kriging often improves map accuracy compared to a non-spatial model [@beguinPredictingSoilProperties2017; @lamichhaneDigitalSoilMapping2019; @mollaMachineLearningGeostatistical2023], but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations [@henglGenericFrameworkSpatial2004; @szaboMappingSoilHydraulic2019; @takoutsingComparingPredictionPerformance2022].
@@
 Comparing the different configurations allowed us to isolate the added value of each of the predictor groups.
-We did not explore configurations comprising radiometrics without terrain, because Lidar terrain surveys typically precede airborne radiometric surveys.
+We did not explore configurations comprising radiometrics without terrain, because LiDAR terrain surveys typically precede airborne radiometric surveys.
 The models with only DMK peat depth class were simple linear models rather than RFs, and served to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth, calibrated on the same data.
@@
 
-For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration -- because correlation between predictors degrades variable importance measures [@stroblConditionalVariableImportance2008; @biauRandomForestGuided2016] and can produce misleading visualizations of predictor-outcome relationships [@biecekExplanatoryModelAnalysis2021; @dwivediExplainableAIXAI2023].
+For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration -- because correlation between predictors degrades variable importance measures [@stroblConditionalVariableImportance2008; @biauRandomForestGuided2016] and can produce misleading visualizations of predictor--outcome relationships [@biecekExplanatoryModelAnalysis2021; @dwivediExplainableAIXAI2023].
 Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation above 0.7.
-Thus, highly-correlated sets of predictors are represented by a single variable for the purposes of model interpretation.
+Thus, highly correlated sets of predictors are represented by a single variable for the purposes of model interpretation.
 
@@
 
-Although the difficulty of predicting peat depth caused prediction intervals to be wide, these uncertainty estimates were well-calibrated.
+Although the difficulty of predicting peat depth caused prediction intervals to be wide, these uncertainty estimates were well calibrated.
 At Skrimfjella the prediction interval coverage probability was 91 %, and at Ørskogfjellet it was 84 % (both compared to the target value of 90 %).
@@
 *DMK* was also important -- the shallow class in particular -- but only at Ørskogfjellet.
-Some realizations of the hydrological predictors *TWI* and *DTW* showed considerable importance, while others showed little -- with no consistency between sites.
+Some realizations of the hydrological predictors TWI and DTW showed considerable importance, while others showed little -- with no consistency between sites.
 For example, *TWI5m*, *DTW40000*, and *DTW2500* rounded out the top five most important variables at Skrimfjella, while *TWI20m* and *TWI50m* were least important, other than *DMK*.
-At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices *slope*, *TRI*, *TPI*, and *roughness*.
+At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices slope, TRI, TPI, and roughness.
 The radiometric variable *radU* showed moderate importance at Skrimfjella.
@@
 The advantage of the DSM was not large in absolute terms (9 cm and 21 cm improvements in mean absolute error), but it explained much more of the variation in depth (improvements in R^2^ of 0.16 and 0.19).
-We attribute this result to the poor spatial and thematic resolution of DMK peat depth, which precludes a robust correlation with peat depths varying from 0-8 m at fine spatial scales.
+We attribute this result to the poor spatial and thematic resolution of DMK peat depth, which precludes a robust correlation with peat depths varying from 0--8 m at fine spatial scales.
 
@@
 More standardized reporting would help but not eliminate this consideration. 
-For example, R^2^ is sensitive to high-leverage extreme values, so it will evaluate a right-skewed distribution differently than a symmetrical distribution. 
+For example, R^2^ is sensitive to high leverage, extreme values, so it will evaluate a right-skewed distribution differently than a symmetrical distribution. 
 We evaluated our models with respect to the explicit purpose of creating peat depth maps across the study areas, but not all studies tailored evaluation to match an explicitly formulated problem [@milaNearestNeighbourDistance2022].
@@
 He found that radK alone predicted peat depth with much higher concordance than our models (CCC = 0.76) and that elevation alone produced comparable performance (CCC = 0.27). 
-The RMSE from these univariate models was 46-68 cm (cf. 78 and 75 cm for Skrimfjella and Ørskogfjellet).
+The RMSE from these univariate models was 46--68 cm (cf. 78 and 75 cm for Skrimfjella and Ørskogfjellet).
 
@@
 Map users should not trust the maps to identify all large carbon stocks, but they can trust that identified large stocks really are large. 
-That makes the map more suited for “red-lighting” than “green-lighting” peatland conversion, for example. 
+That makes the map more suited for "red-lighting" than "green-lighting" peatland conversion, for example. 
 Where it does not prohibit conversion (i.e. predicts shallow peat), a ground survey should be done before conversion is allowed (assuming depth is the only consideration). 
@@
 The RF algorithm’s flexibility means that radiometrics could be used for shallower peats if they provided predictive value for that part of the depth distribution, but there is no evidence of that in our results.
-In the partial dependence plot of radU at Skrimfjella, the expected negative relationship between depth and radU cannot be found by ignoring the left (highly-attenuated) side of the distribution. 
+In the partial dependence plot of radU at Skrimfjella, the expected negative relationship between depth and radU cannot be found by ignoring the left (highly attenuated) side of the distribution. 
 About a quarter of the peats in our study were less than a meter deep, and full attenuation is unlikely for these [@beamishGammaRayAttenuation2013].
@@
 Although we do not believe full attenuation is the primary reason for poor performance in our study, it may limit peat depth mapping under other circumstances. 
-First-principle calculations suggest that radiation should be 90 % attenuated after about 50-60 cm of typical, wet peat or 85 cm of unnaturally dry peat [@beamishGammaRayAttenuation2013; @beamishDetailedMappingPeat2024], and some field tests support these values [@billenEignungGammaspektrometrieKartieren2015].
+First-principle calculations suggest that radiation should be 90 % attenuated after about 50--60 cm of typical, wet peat or 85 cm of unnaturally dry peat [@beamishGammaRayAttenuation2013; @beamishDetailedMappingPeat2024], and some field tests support these values [@billenEignungGammaspektrometrieKartieren2015].
 It is remarkable that particular studies detected radiation differences up to several meters deep [@gatisMappingUplandPeat2019; @kogantiMappingPeatDepth2023], but these may be the exceptions rather than the rule.
@@
 Delineating a valley bottom involves ambiguity, but the MRVBF index is a pragmatic approach that considers a location a valley bottom if it is sufficiently low and flat at a particular scale [@gallantMultiresolutionIndexValley2003].
-The multiresolution nature of the index allows small elevated but flat areas (including saddles) to be characterized as having high valley bottom flatness [@gallantMultiresolutionIndexValley2003]. 
+The multiscale nature of the index allows small elevated but flat areas (including saddles) to be characterized as having high valley bottom flatness [@gallantMultiresolutionIndexValley2003]. 
 Our results suggest that MRVBF is a robust indicator of high water tables (and peat accumulation) over millennial time scales, corroborating other studies [@rudiyantoOpenDigitalMapping2018; @deragonMappingMaximumPeat2023].
 
-Other terrain-derived predictors with predictive value in our study are hydrological (Topographic Wetness Index and Depth-To-Water index). 
+Other terrain-derived predictors with predictive value in our study are hydrological (TWI and DTW). 
 Notably, terrain slope was inferior to (Ørskogfjellet) or highly correlated with (Skrimfjella) these hydrological indices. 
@@
 Wetter locations (high TWI and low DTW) were generally associated with deeper peat, but these relationships were not as strong or consistent as with MRVBF. 
-The optimal scale for Topographic Wetness Index and Depth-To-Water index varied, and likely depends on both the dominant peat formation processes and the typical size of peatland features in a landscape.
+The optimal scale for TWI and DTW varied, and likely depends on both the dominant peat formation processes and the typical size of peatland features in a landscape. 
 Including multiple scales of these variables allows the model to capture different hydrological mechanisms operating at different spatial scales.
@@
 That a peat at Ørskogfjellet was classified as shallow rather than deep did not uniformly lower the model’s expectation of depth there.
-Rather, the veracity of the deep-indication was contingent on other factors.
 
@@
 
-High-quality digital terrain models are available for mainland Norway, making peat depth measurements the critical training data need. 
+High quality DTMs are available for mainland Norway, making peat depth measurements the critical training data need. 
 For a given area, some minimum number of measurements is necessary to create meaningful improvement over DMK peat depth. 
@@
 The scarcity of depth measurements and their short spatial autocorrelation mean that mapping peat depth at broad scales is not an exercise in spatial interpolation [@henglGenericFrameworkSpatial2004].
-Peat depths in our GPR data showed spatial autocorrelation to a range of 50-100 m, and getting measurements at such fine grain is only realistic for small areas, not across landscapes. 
-Therefore, we anticipate that spatially-explicit DSM is of limited value in all but the most intensively sampled landscapes (currently nowhere in Norway). 
+Peat depths in our GPR data showed spatial autocorrelation to a range of 50--100 m, and getting measurements at such fine grain is only realistic for small areas, not across landscapes. 
+Therefore, we anticipate that spatially explicit DSM is of limited value in all but the most intensively sampled landscapes (currently nowhere in Norway). 
 Fine-scale variation in peat depth raises the possibility of mapping at very fine resolution (1 m) — even if land use planning and carbon accounting do not operate at this grain. 
-Terrain-depth relationships might be stronger in 1m cells than in our 10 m cells, considering the hummock-hollow microtopography of many peatlands.
+Terrain--depth relationships might be stronger in 1m cells than in our 10 m cells, considering the hummock--hollow microtopography of many peatlands.
 
@@
 Bakkestuen et al.
-[-@bakkestuenDelineationWetlandAreas2023] successfully predicted peatland extent using CNNs, and their ability to automatically learn multi-scale spatial features would reduce the need to manually engineer these.
-MRVBF's success as a predictor of peat depth demonstrates that multi-scale spatial patterns matter for peat depth and CNNs are designed to learn such patterns. 
+[-@bakkestuenDelineationWetlandAreas2023] successfully predicted peatland extent using CNNs, and their ability to automatically learn multiscale spatial features would reduce the need to manually engineer these.
+MRVBF's success as a predictor of peat depth demonstrates that multiscale spatial patterns matter for peat depth and CNNs are designed to learn such patterns. 
 The kind of relationship described in Buffam et al. [-@buffamFillingHolesRegional2010], where peat depth in basins related to terrain slope at the basin edge, is also something a CNN could learn. 

### 2024-12-05 ms/ms.tex
ms: hyphens and abbreviations

@@
 Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data at two contrasting sites.
-We measured peat depth at 372 and 1878 locations and trained Random Forest models to predict depth at 10 m resolution.
+We measured peat depth at 372 and 1878 locations and trained random forest models to predict depth at 10 m resolution.
 The best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth.
@@
 Peat soils are a terrestrial carbon stock of global importance.
-They store about 30 \% of global soil carbon despite covering only 2-3 \% of Earth's land \citep{xuPEATMAPRefiningEstimates2018, friedlingsteinGlobalCarbonBudget2020, unepGlobalPeatlandsAssessment2022}.
+They store about 30 \% of global soil carbon despite covering only 2--3 \% of Earth's land \citep{xuPEATMAPRefiningEstimates2018, friedlingsteinGlobalCarbonBudget2020, unepGlobalPeatlandsAssessment2022}.
 In other words, they are extremely carbon dense -- denser than any other ecosystem per square meter \citep{temminkRecoveringWetlandBiogeomorphic2022}.
@@
 However, the short range of spatial autocorrelation in peat depth limits its mapping value \citep{henglGenericFrameworkSpatial2004}.
-Relief, or topography, is widely and accurately mapped in digital terrain models.
-For example, most of mainland Norway has at least one elevation measurement per square meter from airborne LiDAR surveys.
+Relief, or topography, is widely and accurately mapped in digital terrain models (DTMs).
+For example, most of mainland Norway has at least one elevation measurement per square meter from airborne Light Detection and Ranging (LiDAR) surveys.
 Moreover, mechanisms of peat formation are linked to topography.
@@
 The little we know about the depths of Norwegian peatlands comes from surveys meant to identify arable land.
-After scattered surveys in the early 20th century, a comprehensive round of surveying was completed 1964-2001 as part of a wider land cover mapping in Norway \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
+After scattered surveys in the early 20th century, a comprehensive round of surveying was completed 1964--2001 as part of a wider land cover mapping in Norway \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
 Because of its agricultural and silvicultural focus, it covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable \citep{ahlstromAR5Klassifikasjonssystem2019}.
-Field surveyors carried a one-meter probe, so they measured peat depth as shallow (\textless{} 1 m), or deep (\textgreater{} 1 m).
+Field surveyors carried a 1 m probe, so they measured peat depth as shallow (\textless{} 1 m), or deep (\textgreater{} 1 m).
 These classes were assigned to whole mire polygons, so spatial resolution is on the order of hectares.
@@
 Field work at Skrimfjella was conducted in August 2020.
-We navigated to the centers of the raster cells in the depth and occurrence samples by handheld GPS, checking that positional error was below \unit{3\,m}.
+We navigated to the centers of the raster cells in the depth and occurrence samples with a handheld global navigation satellite system (GNSS) receiver, checking that positional error was below \unit{3\,m}.
 For each depth sample location, we measured peat depth three times (at the vertices of a triangle with \unit{2\,m} sides) to get a more representative value for the \unit{10\,m} raster cell, and to dampen the effect of outlying measurements \citep{parryEvaluatingApproachesEstimating2014}.
@@
 Occasionally, when the soil itself was difficult to judge, we made our determination also based on the presence or absence of mire vegetation.
-Although peat soil is strictly defined by organic content (which we did not analyse), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland \citep{minasnyMappingMonitoringPeatland2023}.
+Although peat soil is strictly defined by organic content (which we did not analyze), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland \citep{minasnyMappingMonitoringPeatland2023}.
 
-Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively-chosen, individual mires, using ground-penetrating radar (GPR).
+Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively chosen, individual mires, using ground-penetrating radar (GPR).
 We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with its \unit{500\,MHz} shielded antenna mounted in a plastic sledge, and its control unit connected to a GNSS receiver.
@@
 This approach has been found to identify sample sizes that correspond with diminishing returns in predictive model performance on external evaluation data \citep{sauretteDivergenceMetricsDetermining2023}.
-We defined a sequence of sample sizes (50--500) and for ten replicate samples at each size \citep[drawn by conditioned latin hypercube sampling,][]{minasnyConditionedLatinHypercube2006, roudierClhsPackageConditioned2011}, we calculated the mean Kullback-Leibler divergence between sample and population distributions \citep{maloneMethodsImproveUtility2019, sauretteDivergenceMetricsDetermining2023}.
+We defined a sequence of sample sizes (50--500) and for ten replicate samples at each size \citep[drawn by conditioned latin hypercube sampling,][]{minasnyConditionedLatinHypercube2006, roudierClhsPackageConditioned2011}, we calculated the mean Kullback--Leibler divergence between sample and population distributions \citep{maloneMethodsImproveUtility2019, sauretteDivergenceMetricsDetermining2023}.
 The predictors in the divergence calculation were terrain slope and four radiometrics: potassium, thorium, uranium, and total count.
@@
 Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
-To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly-accessible roads.
+To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly accessible roads.
 Cost distance was calculated using GRASS's \emph{r.walk} function, with friction costs defined by AR5 land classes \citep{GRASSv8-2}.
@@
 We know that peat depth tends to vary at fine scales in Norway, which is why we chose \unit{1\,m} and \unit{10\,m} resolutions \citep{maxwellLandsurfaceParametersSpatial2022}.
-We also calculated the the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
+We also calculated the MRVBF index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
 We calculated this index in SAGA GIS \citep[v.9.3.2,][]{conradSystemAutomatedGeoscientific2015} with default parameters.
@@
 Next, we calculated the Topographic Wetness Index \citep{quinnPredictionHillslopeFlow1991}.
-This index is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate-to-coarse resolution DTMs \citep{agrenEvaluatingDigitalTerrain2014, riihimakiTopographicWetnessIndex2021}, so we calculated it from \unit{5\,m}, \unit{10\,m}, \unit{20\,m}, and \unit{50\,m} DTM resolution.
+This index is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate to coarse resolution DTMs \citep{agrenEvaluatingDigitalTerrain2014, riihimakiTopographicWetnessIndex2021}, so we calculated it from \unit{5\,m}, \unit{10\,m}, \unit{20\,m}, and \unit{50\,m} DTM resolution.
 The calculations were performed with Whitebox software \citep{lindsayWhiteboxGATCase2016}, accessed through the \emph{whitebox} R package \citep[v2.4,][]{wuWhiteboxWhiteboxToolsFrontend2022}.
@@
 
-The last terrain-based predictor we included was the depth-to-water index \citep{murphyMappingWetlandsComparison2007}.
+The last terrain-based predictor we included was the depth-to-water (DTW) index \citep{murphyMappingWetlandsComparison2007}.
 This index approximates a location's vertical height above the surface water feature that it is likely to drain towards.
@@
 In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
-For each of the seven surface water layers, we derived the depth-to-water index using the \emph{Distance Accumulation} tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.
+For each of the seven surface water layers, we derived DTW using the \emph{Distance Accumulation} tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.
 
@@
 It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance \citep{khaledianSelectingAppropriateMachine2020}.
-Evaluating variable importance in a maximally-predictive model aligns with the aim of this study.
+Evaluating variable importance in a maximally predictive model aligns with the aim of this study.
 
 RF by itself is not a spatial model, and it will only predict spatial structure in the outcome to the degree that the structure is captured by predictors.
-We considered using regression kriging -- a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically-interpolated surface of RF residuals \citep{henglGenericFrameworkSpatial2004}.
+We considered using regression kriging -- a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically interpolated surface of RF residuals \citep{henglGenericFrameworkSpatial2004}.
 The spatial component in regression kriging often improves map accuracy compared to a non-spatial model \citep{beguinPredictingSoilProperties2017, lamichhaneDigitalSoilMapping2019, mollaMachineLearningGeostatistical2023}, but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations \citep{henglGenericFrameworkSpatial2004, szaboMappingSoilHydraulic2019, takoutsingComparingPredictionPerformance2022}.
@@
 
-For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration -- because correlation between predictors degrades variable importance measures \citep{stroblConditionalVariableImportance2008, biauRandomForestGuided2016} and can produce misleading visualizations of predictor-outcome relationships \citep{biecekExplanatoryModelAnalysis2021, dwivediExplainableAIXAI2023}.
+For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration -- because correlation between predictors degrades variable importance measures \citep{stroblConditionalVariableImportance2008, biauRandomForestGuided2016} and can produce misleading visualizations of predictor--outcome relationships \citep{biecekExplanatoryModelAnalysis2021, dwivediExplainableAIXAI2023}.
 Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation above 0.7.
-Thus, highly-correlated sets of predictors are represented by a single variable for the purposes of model interpretation.
+Thus, highly correlated sets of predictors are represented by a single variable for the purposes of model interpretation.
 
@@
 
-Although the difficulty of predicting peat depth caused prediction intervals to be wide, these uncertainty estimates were well-calibrated.
+Although the difficulty of predicting peat depth caused prediction intervals to be wide, these uncertainty estimates were well calibrated.
 At Skrimfjella the prediction interval coverage probability was 91 \%, and at Ørskogfjellet it was 84 \% (both compared to the target value of 90 \%).
@@
 \emph{DMK} was also important -- the shallow class in particular -- but only at Ørskogfjellet.
-Some realizations of the hydrological predictors \emph{TWI} and \emph{DTW} showed considerable importance, while others showed little -- with no consistency between sites.
+Some realizations of the hydrological predictors TWI and DTW showed considerable importance, while others showed little -- with no consistency between sites.
 For example, \emph{TWI5m}, \emph{DTW40000}, and \emph{DTW2500} rounded out the top five most important variables at Skrimfjella, while \emph{TWI20m} and \emph{TWI50m} were least important, other than \emph{DMK}.
-At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices \emph{slope}, \emph{TRI}, \emph{TPI}, and \emph{roughness}.
+At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices slope, TRI, TPI, and roughness.
 The radiometric variable \emph{radU} showed moderate importance at Skrimfjella.
@@
 The advantage of the DSM was not large in absolute terms (9 cm and 21 cm improvements in mean absolute error), but it explained much more of the variation in depth (improvements in R\textsuperscript{2} of 0.16 and 0.19).
-We attribute this result to the poor spatial and thematic resolution of DMK peat depth, which precludes a robust correlation with peat depths varying from 0-8 m at fine spatial scales.
+We attribute this result to the poor spatial and thematic resolution of DMK peat depth, which precludes a robust correlation with peat depths varying from 0--8 m at fine spatial scales.
 
@@
 More standardized reporting would help but not eliminate this consideration.
-For example, R\textsuperscript{2} is sensitive to high-leverage extreme values, so it will evaluate a right-skewed distribution differently than a symmetrical distribution.
+For example, R\textsuperscript{2} is sensitive to high leverage, extreme values, so it will evaluate a right-skewed distribution differently than a symmetrical distribution.
 We evaluated our models with respect to the explicit purpose of creating peat depth maps across the study areas, but not all studies tailored evaluation to match an explicitly formulated problem \citep{milaNearestNeighbourDistance2022}.
@@
 He found that radK alone predicted peat depth with much higher concordance than our models (CCC = 0.76) and that elevation alone produced comparable performance (CCC = 0.27).
-The RMSE from these univariate models was 46-68 cm (cf.~78 and 75 cm for Skrimfjella and Ørskogfjellet).
+The RMSE from these univariate models was 46--68 cm (cf.~78 and 75 cm for Skrimfjella and Ørskogfjellet).
 
@@
 The RF algorithm's flexibility means that radiometrics could be used for shallower peats if they provided predictive value for that part of the depth distribution, but there is no evidence of that in our results.
-In the partial dependence plot of radU at Skrimfjella, the expected negative relationship between depth and radU cannot be found by ignoring the left (highly-attenuated) side of the distribution.
+In the partial dependence plot of radU at Skrimfjella, the expected negative relationship between depth and radU cannot be found by ignoring the left (highly attenuated) side of the distribution.
 About a quarter of the peats in our study were less than a meter deep, and full attenuation is unlikely for these \citep{beamishGammaRayAttenuation2013}.
@@
 Although we do not believe full attenuation is the primary reason for poor performance in our study, it may limit peat depth mapping under other circumstances.
-First-principle calculations suggest that radiation should be 90 \% attenuated after about 50-60 cm of typical, wet peat or 85 cm of unnaturally dry peat \citep{beamishGammaRayAttenuation2013, beamishDetailedMappingPeat2024}, and some field tests support these values \citep{billenEignungGammaspektrometrieKartieren2015}.
+First-principle calculations suggest that radiation should be 90 \% attenuated after about 50--60 cm of typical, wet peat or 85 cm of unnaturally dry peat \citep{beamishGammaRayAttenuation2013, beamishDetailedMappingPeat2024}, and some field tests support these values \citep{billenEignungGammaspektrometrieKartieren2015}.
 It is remarkable that particular studies detected radiation differences up to several meters deep \citep{gatisMappingUplandPeat2019, kogantiMappingPeatDepth2023}, but these may be the exceptions rather than the rule.
@@
 Delineating a valley bottom involves ambiguity, but the MRVBF index is a pragmatic approach that considers a location a valley bottom if it is sufficiently low and flat at a particular scale \citep{gallantMultiresolutionIndexValley2003}.
-The multiresolution nature of the index allows small elevated but flat areas (including saddles) to be characterized as having high valley bottom flatness \citep{gallantMultiresolutionIndexValley2003}.
+The multiscale nature of the index allows small elevated but flat areas (including saddles) to be characterized as having high valley bottom flatness \citep{gallantMultiresolutionIndexValley2003}.
 Our results suggest that MRVBF is a robust indicator of high water tables (and peat accumulation) over millennial time scales, corroborating other studies \citep{rudiyantoOpenDigitalMapping2018, deragonMappingMaximumPeat2023}.
 
-Other terrain-derived predictors with predictive value in our study are hydrological (Topographic Wetness Index and Depth-To-Water index).
+Other terrain-derived predictors with predictive value in our study are hydrological (TWI and DTW).
 Notably, terrain slope was inferior to (Ørskogfjellet) or highly correlated with (Skrimfjella) these hydrological indices.
@@
 Wetter locations (high TWI and low DTW) were generally associated with deeper peat, but these relationships were not as strong or consistent as with MRVBF.
-The optimal scale for Topographic Wetness Index and Depth-To-Water index varied, and likely depends on both the dominant peat formation processes and the typical size of peatland features in a landscape.
+The optimal scale for TWI and DTW varied, and likely depends on both the dominant peat formation processes and the typical size of peatland features in a landscape.
 Including multiple scales of these variables allows the model to capture different hydrological mechanisms operating at different spatial scales.
@@
 That a peat at Ørskogfjellet was classified as shallow rather than deep did not uniformly lower the model's expectation of depth there.
-Rather, the veracity of the deep-indication was contingent on other factors.
 
@@
 
-High-quality digital terrain models are available for mainland Norway, making peat depth measurements the critical training data need.
+High quality DTMs are available for mainland Norway, making peat depth measurements the critical training data need.
 For a given area, some minimum number of measurements is necessary to create meaningful improvement over DMK peat depth.
@@
 The scarcity of depth measurements and their short spatial autocorrelation mean that mapping peat depth at broad scales is not an exercise in spatial interpolation \citep{henglGenericFrameworkSpatial2004}.
-Peat depths in our GPR data showed spatial autocorrelation to a range of 50-100 m, and getting measurements at such fine grain is only realistic for small areas, not across landscapes.
-Therefore, we anticipate that spatially-explicit DSM is of limited value in all but the most intensively sampled landscapes (currently nowhere in Norway).
+Peat depths in our GPR data showed spatial autocorrelation to a range of 50--100 m, and getting measurements at such fine grain is only realistic for small areas, not across landscapes.
+Therefore, we anticipate that spatially explicit DSM is of limited value in all but the most intensively sampled landscapes (currently nowhere in Norway).
 Fine-scale variation in peat depth raises the possibility of mapping at very fine resolution (1 m) --- even if land use planning and carbon accounting do not operate at this grain.
-Terrain-depth relationships might be stronger in 1m cells than in our 10 m cells, considering the hummock-hollow microtopography of many peatlands.
+Terrain--depth relationships might be stronger in 1m cells than in our 10 m cells, considering the hummock--hollow microtopography of many peatlands.
 
@@
 Bakkestuen et al.
-\citeyearpar{bakkestuenDelineationWetlandAreas2023} successfully predicted peatland extent using CNNs, and their ability to automatically learn multi-scale spatial features would reduce the need to manually engineer these.
-MRVBF's success as a predictor of peat depth demonstrates that multi-scale spatial patterns matter for peat depth and CNNs are designed to learn such patterns.
+\citeyearpar{bakkestuenDelineationWetlandAreas2023} successfully predicted peatland extent using CNNs, and their ability to automatically learn multiscale spatial features would reduce the need to manually engineer these.
+MRVBF's success as a predictor of peat depth demonstrates that multiscale spatial patterns matter for peat depth and CNNs are designed to learn such patterns.
 The kind of relationship described in Buffam et al.

### 2024-12-05 ms/ms.Rmd
ms: \unit for negative exponents only

@@
 Maps are crucial because they link high-level targets to specific management decisions, unlike spatially aggregated estimates [@oecdOECDEnvironmentalPerformance2022].
-For example, Norwegian regulations restrict conversion to farmland more strictly where peat depth exceeds \unit{1\,m}, and the distribution of peat depth on a farmer's property determines whether he is allowed to convert peatland (Forskrift om nydyrking, 1997, § 5a).
+For example, Norwegian regulations restrict conversion to farmland more strictly where peat depth exceeds 1 m, and the distribution of peat depth on a farmer's property determines whether he is allowed to convert peatland (Forskrift om nydyrking, 1997, § 5a).
 Maps also make it possible to quantify the effect of specific management decisions and thereby understand how local outcomes contribute to regional and national outcomes [@oecdOECDEnvironmentalPerformance2022].
@@
 
-At Skrimfjella we delineated a study area of \unit{34\,km^{2}} based on radiometric coverage and accessibility (Fig. \@ref(fig:sites)b).
+At Skrimfjella we delineated a study area of 34 km^2^ based on radiometric coverage and accessibility (Fig. \@ref(fig:sites)b).
 The study area has a diverse bedrock, with 32 % alkali feltspat granite, 26 % mergelstein, 10 % granite, and eight other types with > 1 % coverage (NGU, 1:250 000 dataset).
@@
 It is almost without human infrastructure, dominated by forest, and borders on a large nature reserve.
-The study area has a mean elevation of \unit{438\,m} above sea level (range 223--711, IQR 351--509), and its mean slope at \unit{10\,m} resolution is 10.8° (IQR 4.6--15.1°).
-In Norway's AR5 national land capability dataset [@ahlstromAR5Klassifikasjonssystem2019], \unit{1.5\,km^{2}} (4.5 %) of the study area is classified as mire -- defined as areas with mire vegetation and at least \unit{30\,cm} of peat depth.
+The study area has a mean elevation of 438 m above sea level (range 223--711, IQR 351--509), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).
+In Norway's AR5 national land capability dataset [@ahlstromAR5Klassifikasjonssystem2019], 1.5 km^2^ (4.5 %) of the study area is classified as mire -- defined as areas with mire vegetation and at least 30 cm of peat depth.
 
-At Ørskogfjellet we defined a study area of \unit{124\,km^{2}} which basically followed the footprint of the radiometric survey (Fig. \@ref(fig:sites)a).
+At Ørskogfjellet we defined a study area of 124 km^2^ which basically followed the footprint of the radiometric survey (Fig. \@ref(fig:sites)a).
 According to the Geological Survey of Norway, bedrock in the area is 84 % granitic gneiss,  11 % granite, and 5 % aluminium silicate gneiss (NGU, 1:250 000 dataset). 
@@
 It is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
-Its mean elevation is \unit{211\,m} above sea level (range 0--807, IQR 73--310), and its mean slope at \unit{10\,m} resolution is 13.0° (IQR 4.7--18.3°).
-The AR5 dataset counts \unit{15.3\,km^{2}} (12.4 %) of the study area as mire.
+Its mean elevation is 211 m above sea level (range 0--807, IQR 73--310), and its mean slope at 10 m resolution is 13.0° (IQR 4.7--18.3°).
+The AR5 dataset counts 15.3 km^2^ (12.4 %) of the study area as mire.
 
@@
 A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones [@brusSamplingDigitalSoil2019].
-We chose for our sampling and modelling a spatial resolution of \unit{10\,m}.
-We considered this a reasonable compromise between DTM resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.
-The point measurements of peat depth described below were ultimately aggregated to \unit{10\,m} resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
+We chose for our sampling and modelling a spatial resolution of 10 m.
+We considered this a reasonable compromise between DTM resolution (1 m) and small mires on the one hand, and airborne radiometric resolution (50 m) on the other.
+The point measurements of peat depth described below were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
 
@@
 
-We measured peat depth in selected locations (\unit{10\,m} raster cells) at Skrimfjella.
+We measured peat depth in selected locations (10 m raster cells) at Skrimfjella.
 The locations were chosen only from areas delineated as mire in the AR5 national land capability dataset.
@@
 This function defines the environmental space as a two-dimensional convex hull around the ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest [@hattabUnifiedFrameworkModel2017].
-Elevation was extracted from the \unit{10\,m} national DTM, slope calculated in degrees, and potassium ground concentration downscaled with bilinear resampling.
+Elevation was extracted from the 10 m national DTM, slope calculated in degrees, and potassium ground concentration downscaled with bilinear resampling.
 We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and with these parameters *eSample* returned 105 raster cells.
@@
 Field work at Skrimfjella was conducted in August 2020.
-We navigated to the centers of the raster cells in the depth and occurrence samples with a handheld global navigation satellite system (GNSS) receiver, checking that positional error was below \unit{3\,m}.
-For each depth sample location, we measured peat depth three times (at the vertices of a triangle with \unit{2\,m} sides) to get a more representative value for the \unit{10\,m} raster cell, and to dampen the effect of outlying measurements [@parryEvaluatingApproachesEstimating2014].
+We navigated to the centers of the raster cells in the depth and occurrence samples with a handheld global navigation satellite system (GNSS) receiver, checking that positional error was below 3 m.
+For each depth sample location, we measured peat depth three times (at the vertices of a triangle with 2 m sides) to get a more representative value for the 10 m raster cell, and to dampen the effect of outlying measurements [@parryEvaluatingApproachesEstimating2014].
 We used a metal probe pushed downward until resistance indicated the base of the peat column.
-Probe locations were adjusted up to \unit{20\,cm} if the base of the peat column seemed to be blocked by an obvious artifact.
+Probe locations were adjusted up to 20 cm if the base of the peat column seemed to be blocked by an obvious artifact.
 A single probe measurement was right-censored because the peat column was deeper than the probe.
-For each occurrence sample location, we recorded the presence or absence of peatland -- primarily by digging and examining the top \unit{20\,cm} of soil (where this was possible).
+For each occurrence sample location, we recorded the presence or absence of peatland -- primarily by digging and examining the top 20 cm of soil (where this was possible).
 We judged whether the soil was a peat soil based on its density, texture, and color.
@@
 Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively chosen, individual mires, using ground-penetrating radar (GPR).
-We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with its \unit{500\,MHz} shielded antenna mounted in a plastic sledge, and its control unit connected to a GNSS receiver.
-At each of the three mires we recorded GPR traces along walking transects that covered the extent of the mire, mostly in traversing, zigzag patterns with between \unit{5\,m} and \unit{20\,m} spacing at their vertices.
+We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with its 500 MHz} shielded antenna mounted in a plastic sledge, and its control unit connected to a GNSS receiver.
+At each of the three mires we recorded GPR traces along walking transects that covered the extent of the mire, mostly in traversing, zigzag patterns with between 5 m and 20 m spacing at their vertices.
 Along the GPR transects we also probed peat depth at marked trace locations, to be able to calibrate the GPR wave speed velocity.
@@
 We used picks at marked trace locations to calibrate wave speed velocity; we pooled calibration points across the three mires and fitted a linear regression of depth on one-way travel time with the intercept fixed at zero.
-In total we had 46 calibration points along \unit{3.5\,km} of GPR transects.
+In total we had 46 calibration points along 3.5 km of GPR transects.
 Finally, we used the calculated wave velocity (\unit{0.0387\,m\,ns^{-1}}, $R^2 = 0.874$) to convert the travel times of all picks to calibrated peat depths.
@@
 
-At Ørskogfjellet we also measured peat depth in a sample of \unit{10\,m} raster cells, selected from the part of the study area classified in the AR5 dataset as mire.
+At Ørskogfjellet we also measured peat depth in a sample of 10 m raster cells, selected from the part of the study area classified in the AR5 dataset as mire.
 Before selecting locations, we determined a minimal sample size that would adequately capture the terrain and radiometric properties of the entire mire area.
@@
 Feature space coverage sampling works best when all dimensions are important predictors of the outcome [@wadouxSamplingDesignOptimization2019], and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
-The radiometric predictors were downscaled to \unit{10\,m} resolution with cubic B-spline resampling in QGIS [v.3, @QGISsoftware].
+The radiometric predictors were downscaled to 10 m resolution with cubic B-spline resampling in QGIS [v.3, @QGISsoftware].
 We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
@@
 We navigated to the centers of the raster cells in the sample using real time kinematic differential GNSS (Topcon Positioning Systems, USA), to ensure sub-meter positional accuracy.
-At each location we measured peat depth three times by manual probing, with probe locations spaced approximately \unit{2.5\,m} apart.
+At each location we measured peat depth three times by manual probing, with probe locations spaced approximately 2.5 m apart.
 For a tiny fraction of these measurements (5 total), it was not possible to reach the base of the peat column, and a right-censored depth was recorded. 
-In areas with dense sampling locations, we also measured peat depth with GPR along snaking transects passing through the centers of the sampling cells (seven transects, \unit{6.2\,km} total length).
-We used the same GPR system as at Skrimfjella, but with a \unit{100\,MHz} Malå rough terrain antenna (Guideline Geo AB, Sweden) at some transects.
+In areas with dense sampling locations, we also measured peat depth with GPR along snaking transects passing through the centers of the sampling cells (seven transects, 6.2 km total length).
+We used the same GPR system as at Skrimfjella, but with a 100 MHz Malå rough terrain antenna (Guideline Geo AB, Sweden) at some transects.
 To navigate the GPR transects, we placed flags at the cell centers of the sample locations, and used a handheld GNSS receiver to guide the GPR operator.
@@
 We processed the GPR data with Reflexw software (v.8.5, Sandmeier Scientific Software, Germany), applying a dewow filter, time-zero correction, bandpass filter, gain filter, and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennae.
-The last correction is important for the rough terrain antenna, which has an antenna separation (\unit{2.2\,m}) -- comparable to typical peat depths.
+The last correction is important for the rough terrain antenna, which has an antenna separation (2.2 m) -- comparable to typical peat depths.
 As with the data from Skrimfjella, we picked the base of the peat column from strong reflectors in the radargram, and calibrated wave velocity with manual probe measurements in a linear regression.
-The points in the regression were created by joining to each probe measurement the travel time of the nearest pick, but only if these were within \unit{2\,m} of each other.
-In total we had 78 calibration points along \unit{7.8\,km} of interpretable GPR traces (transect length exceeded because of extra GPR data).
+The points in the regression were created by joining to each probe measurement the travel time of the nearest pick, but only if these were within 2 m of each other.
+In total we had 78 calibration points along 7.8 km of interpretable GPR traces (transect length exceeded because of extra GPR data).
 Finally, we used the calculated wave velocity (\unit{0.0427\,m\,ns^{-1}}, $R^2 = 0.946$) to convert the travel times of all picks to calibrated peat depths.
@@
 The first set was provided by the Norwegian Public Roads Administration, who commissioned GPR surveys of particular peatland areas in 2020 and 2021.
-The surveys were conducted with a dual channel system (\unit{70\,MHz} and \unit{300\,MHz}; ImpulseRadar AB, Sweden), connected to GNSS with CPOS correction.
+The surveys were conducted with a dual channel system (70 MHz and 300 MHz; ImpulseRadar AB, Sweden), connected to GNSS with CPOS correction.
 We used interpreted and calibrated traces from these surveys, and discarded some data where multiple depths were interpreted for the same locations.
-This summed to \unit{7.4\,km} of interpreted traces.
+This summed to 7.4 km of interpreted traces.
 The second set of existing depth data we extracted from a paper map made by the Norwegian Soil and Mire Company in 1984.
-This map presents 44 borehole depths (in decimeters) across a \unit{9\,ha} peatland area.
+This map presents 44 borehole depths (in decimeters) across a 9 ha peatland area.
 We georeferenced the map and digitized the borehole locations and depths.
@@
 The radiometric survey covering Skrimfjella was conducted in 2008--2011. 
-The survey was flown at an average altitude of \unit{75\,m} and average speed of \unit{108\,km\,h^{-1}}, with flight lines spaced \unit{200\,m} apart.
+The survey was flown at an average altitude of 75 m and average speed of \unit{108\,km\,h^{-1}}, with flight lines spaced 200 m apart.
 Spectrometer count rates were calibrated annually to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
-The Geological Survey of Norway processed data from the spectrometer following standard procedures outlined by the International Atomic Energy Association, and the processing included: correction for aircraft and cosmic background radiation, correction for radon in the air, window stripping of the gamma ray spectrum, correction for flying height, conversion of count rates to ground concentrations, and finally gridding to \unit{50\,m} resolution with micro-leveling.
+The Geological Survey of Norway processed data from the spectrometer following standard procedures outlined by the International Atomic Energy Association, and the processing included: correction for aircraft and cosmic background radiation, correction for radon in the air, window stripping of the gamma ray spectrum, correction for flying height, conversion of count rates to ground concentrations, and finally gridding to 50 m resolution with micro-leveling.
 Further details about the survey and data processing are provided in Baranwal et al. [-@baranwalHelicopterborneMagneticElectromagnetic2013].
-We downscaled the processed data to \unit{10\,m} resolution by cubic spline resampling, using the *terra* package in R.
+We downscaled the processed data to 10 m resolution by cubic spline resampling, using the *terra* package in R.
 
 A very similar radiometric survey covering Ørskogfjellet was conducted in December 2014 and January 2015. 
-This survey was flown at an average altitude of \unit{80\,m} and average speed of \unit{88\,km\,h^{-1}}, with flight lines also spaced \unit{200\,m} apart.
+This survey was flown at an average altitude of 80 m and average speed of \unit{88\,km\,h^{-1}}, with flight lines also spaced 200 m apart.
 Spectrometer count rates were calibrated in 2013 to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
@@
 Further details about the survey and data processing are provided in Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015].
-The \unit{10\,m} resolution predictors from this survey were identical to the layers used in the sampling design (resampled from \unit{50\,m} resolution with cubic B-splines).
+The 10 m resolution predictors from this survey were identical to the layers used in the sampling design (resampled from 50 m resolution with cubic B-splines).
 
@@
 
-For terrain-derived predictors, we obtained \unit{1\,m} resolution DTMs from the Norwegian Mapping Authority.
+For terrain-derived predictors, we obtained 1 m resolution DTMs from the Norwegian Mapping Authority.
 The DTM for Skrimfjella was produced from airborne laser scanning surveys in 2015 and 2022, with laser point density of \unit{5\,pts\,m^{-2}}.
@@
 We used the *terra* R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
-These were derived at two scales to produce eight different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
+These were derived at two scales to produce eight different predictors; we either calculated from 1 m DTM resolution and then aggregated, or aggregated to 10 m DTM resolution and then calculated the indices.
 This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties [@millerImpactMultiscalePredictor2015; @dornikOptimalScalingPredictors2022; @newmanAssessingSpatiallyHeterogeneous2023].
-We know that peat depth tends to vary at fine scales in Norway, which is why we chose \unit{1\,m} and \unit{10\,m} resolutions [@maxwellLandsurfaceParametersSpatial2022].
+We know that peat depth tends to vary at fine scales in Norway, which is why we chose 1 m and 10 m resolutions [@maxwellLandsurfaceParametersSpatial2022].
 We also calculated the MRVBF index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm [@gallantMultiresolutionIndexValley2003].
@@
 Next, we calculated the Topographic Wetness Index [@quinnPredictionHillslopeFlow1991].
-This index is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate to coarse resolution DTMs [@agrenEvaluatingDigitalTerrain2014; @riihimakiTopographicWetnessIndex2021], so we calculated it from \unit{5\,m}, \unit{10\,m}, \unit{20\,m}, and \unit{50\,m} DTM resolution.
+This index is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate to coarse resolution DTMs [@agrenEvaluatingDigitalTerrain2014; @riihimakiTopographicWetnessIndex2021], so we calculated it from 5 m, 10 m, 20 m, and 50 m DTM resolution.
 The calculations were performed with Whitebox software [@lindsayWhiteboxGATCase2016], accessed through the *whitebox* R package [v2.4, @wuWhiteboxWhiteboxToolsFrontend2022].
@@
 It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature [eq. 5 in @murphyTopographicModellingSoil2009].
-We calculated unitless slope from the \unit{1\,m} DTM using the Whitebox software.
+We calculated unitless slope from the 1 m DTM using the Whitebox software.
 Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell [@schonauerSpatiotemporalPredictionSoil2021; @schonauerRcodeCalculatingDepthwater2021].
-This catchments area layer was then thresholded at seven different levels (*flow initiation areas* from \unit{0.25\,ha} to \unit{16\,ha}) to estimate surface water features under moisture scenarios varying from wet to dry [@murphyModellingMappingTopographic2011; @agrenEvaluatingDigitalTerrain2014; @schonauerSpatiotemporalPredictionSoil2021].
+This catchments area layer was then thresholded at seven different levels (*flow initiation areas* 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry [@murphyModellingMappingTopographic2011; @agrenEvaluatingDigitalTerrain2014; @schonauerSpatiotemporalPredictionSoil2021].
 In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
@@
 We prepared one categorical predictor -- peat depth class -- from a historical national map dataset called *DMK* [@ahlstromAR5Klassifikasjonssystem2019].
-The DMK peat depth classes are: < \unit{1\,m} (*shallow*), > \unit{1\,m} (*deep*), and *unknown*.
-Mappers generally assigned peat depth classes to polygons of at least \unit{0.5\,ha}, although delineating polygons down to \unit{0.2\,ha} was allowed if peat depth showed a "particularly marked difference" [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
-We rasterized the peat depth class attribute to our \unit{10\,m} grid.
+The DMK peat depth classes are: < 1 m (*shallow*), > 1 m (*deep*), and *unknown*.
+Mappers generally assigned peat depth classes to polygons of at least 0.5 ha, although delineating polygons down to 0.2 ha was allowed if peat depth showed a "particularly marked difference" [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
+We rasterized the peat depth class attribute to our 10 m grid.
 
@@
 We obtained depth measurements for 372 cells at Skrimfjella (area equivalent to 2.4 % of mapped peatland) and 1878 cells at Ørskogfjellet (area equivalent to 1.2 % of mapped peatland).
-Roughly 80 % of these \unit{10\,m} cells were within mapped peatland, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
+Roughly 80 % of these 10 m cells were within mapped peatland, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
 Coverage of DMK peat depth classes was higher at Ørskogfjellet than at Skrimfjella (79 % versus 27 % of cells), and at Ørskogfjellet the *deep* and *shallow* classes showed a larger difference in measured depth.
-Overall mean peat depths were similar at Skrimfjella and Ørskogfjellet:  \unit{119\,cm} and \unit{126\,cm}, respectively.
+Overall mean peat depths were similar at Skrimfjella and Ørskogfjellet:  119 cm and 126 cm, respectively.
 
@@
 None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \@ref(fig:modelMetrics)).
-For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an R^2^ of 0.34, and a mean absolute error of \unit{60\,cm}.
-For Ørskogfjellet the same values were 0.39, 0.33, and \unit{56\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
+For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an R^2^ of 0.34, and a mean absolute error of 60 cm.
+For Ørskogfjellet the same values were 0.39, 0.33, and 56 cm, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
 These values were derived from kNNDM spatial cross-validation with 20 folds at Skrimfjella and 10 folds at Ørskogfjellet.
@@
 The performance gap between the *terrain and DMK* configuration and the *terrain-only* configuration was similarly small.
-Compared to the *terrain-only* configuration, the *terrain and radiometric* configuration improved concordance correlation by 0.04, R^2^ by 0.04, and mean absolute error by \unit{1\,cm}.
+Compared to the *terrain-only* configuration, the *terrain and radiometric* configuration improved concordance correlation by 0.04, R^2^ by 0.04, and mean absolute error by 1 cm.
 *DMK class* alone -- although calibrated to measured depths -- was a very poor predictor of peat depth, with a concordance correlation coefficient of 0.008.
@@
 Adding radiometric predictors to these configurations worsened model performance, especially in terms of concordance correlation and R^2^.
-By itself, *DMK class* produced a concordance correlation coefficient of 0.17 (compared to 0.008 at Skrimfjella), but the worst mean absolute error of any model at either site: \unit{77\,cm}.
+By itself, *DMK class* produced a concordance correlation coefficient of 0.17 (compared to 0.008 at Skrimfjella), but the worst mean absolute error of any model at either site: 77 cm.
 
 The best models at both sites overpredicted shallow peats and strongly underpredicted very deep peats (Fig. \@ref(fig:calPlots)).
-The mean error (bias) of these models was \unit{10\,cm} at Skrimfjella and \unit{-4\,cm} at Ørskogfjellet.
+The mean error (bias) of these models was 10 cm at Skrimfjella and -4 cm at Ørskogfjellet.
 
@@
 
-When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced worse mean absolute error than just assuming a constant \unit{30\,cm} depth (Fig.
+When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced worse mean absolute error than just assuming a constant 30 cm depth (Fig.
 \@ref(fig:modelMetricsExtrapolation)).
@@
 Many of the the most important predictors in the best performing models showed non-monotonic effects on peat depth (Fig. \@ref(fig:pdps)).
-At Ørskogfjellet for example, increasing *elevation* was predictive of deeper peat up to about 75 meters above sea level, after which a further increase was predictive of shallower peat.
-At Skrimfjella the partial dependence on *elevation* had the opposite shape, with the shallowest peats predicted at intermediate elevations, around 350 meters above sea level.
+At Ørskogfjellet for example, increasing *elevation* was predictive of deeper peat up to about 75 m above sea level, after which a further increase was predictive of shallower peat.
+At Skrimfjella the partial dependence on *elevation* had the opposite shape, with the shallowest peats predicted at intermediate elevations, around 350 m above sea level.
 *TWI50m* at Ørskogfjellet and *DTW4000* and *radU* at Skrimfjella were other predictors that showed considerable fluctuations in their predictive effects on peat depth.
@@
 Individual conditional expectation lines indicated some interactions between predictors (Fig. \@ref(fig:pdps)).
-For example, the magnitude of the increase in depth with *elevation* that the model expected at Ørskogfjellet was different for different observations; some depth predictions increased by only \unit{40\,cm} while others increased by more than \unit{100\,cm}, over the same elevation gain.
+For example, the magnitude of the increase in depth with *elevation* that the model expected at Ørskogfjellet was different for different observations; some depth predictions increased by only 40 cm while others increased by more than 100 cm, over the same elevation gain.
 Similarly, ICE lines of *radU* at Skrimfjella were non-parallel, with some locations showing monotonically increasing peat depth predictions with uranium concentration (unlike the average effect).
@@
 
-Gatis et al. [-@gatisMappingUplandPeat2019] used similar predictors and the same spatial grain, finding a much stronger correlation between predicted and observed peat depth (R^2^ = 0.68). 
+Gatis et al. [-@gatisMappingUplandPeat2019] used similar predictors and the same spatial grain, finding a much stronger correlation between predicted and observed peat depth ($R^2 = 0.68$). 
 Although their random evaluation data partition could make performance estimates too optimistic [@robertsCrossvalidationStrategiesData2017; @wadouxSpatialCrossvalidationNot2021], the confounding effect of spatial structure is probably small because they used linear regression and few predictors.

### 2024-12-05 ms/ms.tex
ms: \unit for negative exponents only

@@
 Maps are crucial because they link high-level targets to specific management decisions, unlike spatially aggregated estimates \citep{oecdOECDEnvironmentalPerformance2022}.
-For example, Norwegian regulations restrict conversion to farmland more strictly where peat depth exceeds \unit{1\,m}, and the distribution of peat depth on a farmer's property determines whether he is allowed to convert peatland (Forskrift om nydyrking, 1997, § 5a).
+For example, Norwegian regulations restrict conversion to farmland more strictly where peat depth exceeds 1 m, and the distribution of peat depth on a farmer's property determines whether he is allowed to convert peatland (Forskrift om nydyrking, 1997, § 5a).
 Maps also make it possible to quantify the effect of specific management decisions and thereby understand how local outcomes contribute to regional and national outcomes \citep{oecdOECDEnvironmentalPerformance2022}.
@@
 
-At Skrimfjella we delineated a study area of \unit{34\,km^{2}} based on radiometric coverage and accessibility (Fig. \ref{fig:sites}b).
+At Skrimfjella we delineated a study area of 34 km\textsuperscript{2} based on radiometric coverage and accessibility (Fig. \ref{fig:sites}b).
 The study area has a diverse bedrock, with 32 \% alkali feltspat granite, 26 \% mergelstein, 10 \% granite, and eight other types with \textgreater{} 1 \% coverage (NGU, 1:250 000 dataset).
@@
 It is almost without human infrastructure, dominated by forest, and borders on a large nature reserve.
-The study area has a mean elevation of \unit{438\,m} above sea level (range 223--711, IQR 351--509), and its mean slope at \unit{10\,m} resolution is 10.8° (IQR 4.6--15.1°).
-In Norway's AR5 national land capability dataset \citep{ahlstromAR5Klassifikasjonssystem2019}, \unit{1.5\,km^{2}} (4.5 \%) of the study area is classified as mire -- defined as areas with mire vegetation and at least \unit{30\,cm} of peat depth.
+The study area has a mean elevation of 438 m above sea level (range 223--711, IQR 351--509), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).
+In Norway's AR5 national land capability dataset \citep{ahlstromAR5Klassifikasjonssystem2019}, 1.5 km\textsuperscript{2} (4.5 \%) of the study area is classified as mire -- defined as areas with mire vegetation and at least 30 cm of peat depth.
 
-At Ørskogfjellet we defined a study area of \unit{124\,km^{2}} which basically followed the footprint of the radiometric survey (Fig. \ref{fig:sites}a).
+At Ørskogfjellet we defined a study area of 124 km\textsuperscript{2} which basically followed the footprint of the radiometric survey (Fig. \ref{fig:sites}a).
 According to the Geological Survey of Norway, bedrock in the area is 84 \% granitic gneiss, 11 \% granite, and 5 \% aluminium silicate gneiss (NGU, 1:250 000 dataset).
@@
 It is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
-Its mean elevation is \unit{211\,m} above sea level (range 0--807, IQR 73--310), and its mean slope at \unit{10\,m} resolution is 13.0° (IQR 4.7--18.3°).
-The AR5 dataset counts \unit{15.3\,km^{2}} (12.4 \%) of the study area as mire.
+Its mean elevation is 211 m above sea level (range 0--807, IQR 73--310), and its mean slope at 10 m resolution is 13.0° (IQR 4.7--18.3°).
+The AR5 dataset counts 15.3 km\textsuperscript{2} (12.4 \%) of the study area as mire.
 
@@
 A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones \citep{brusSamplingDigitalSoil2019}.
-We chose for our sampling and modelling a spatial resolution of \unit{10\,m}.
-We considered this a reasonable compromise between DTM resolution (\unit{1\,m}) and small mires on the one hand, and airborne radiometric resolution (\unit{50\,m}) on the other.
-The point measurements of peat depth described below were ultimately aggregated to \unit{10\,m} resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
+We chose for our sampling and modelling a spatial resolution of 10 m.
+We considered this a reasonable compromise between DTM resolution (1 m) and small mires on the one hand, and airborne radiometric resolution (50 m) on the other.
+The point measurements of peat depth described below were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
 
@@
 This function defines the environmental space as a two-dimensional convex hull around the ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest \citep{hattabUnifiedFrameworkModel2017}.
-Elevation was extracted from the \unit{10\,m} national DTM, slope calculated in degrees, and potassium ground concentration downscaled with bilinear resampling.
+Elevation was extracted from the 10 m national DTM, slope calculated in degrees, and potassium ground concentration downscaled with bilinear resampling.
 We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and with these parameters \emph{eSample} returned 105 raster cells.
@@
 Field work at Skrimfjella was conducted in August 2020.
-We navigated to the centers of the raster cells in the depth and occurrence samples with a handheld global navigation satellite system (GNSS) receiver, checking that positional error was below \unit{3\,m}.
-For each depth sample location, we measured peat depth three times (at the vertices of a triangle with \unit{2\,m} sides) to get a more representative value for the \unit{10\,m} raster cell, and to dampen the effect of outlying measurements \citep{parryEvaluatingApproachesEstimating2014}.
+We navigated to the centers of the raster cells in the depth and occurrence samples with a handheld global navigation satellite system (GNSS) receiver, checking that positional error was below 3 m.
+For each depth sample location, we measured peat depth three times (at the vertices of a triangle with 2 m sides) to get a more representative value for the 10 m raster cell, and to dampen the effect of outlying measurements \citep{parryEvaluatingApproachesEstimating2014}.
 We used a metal probe pushed downward until resistance indicated the base of the peat column.
-Probe locations were adjusted up to \unit{20\,cm} if the base of the peat column seemed to be blocked by an obvious artifact.
+Probe locations were adjusted up to 20 cm if the base of the peat column seemed to be blocked by an obvious artifact.
 A single probe measurement was right-censored because the peat column was deeper than the probe.
-For each occurrence sample location, we recorded the presence or absence of peatland -- primarily by digging and examining the top \unit{20\,cm} of soil (where this was possible).
+For each occurrence sample location, we recorded the presence or absence of peatland -- primarily by digging and examining the top 20 cm of soil (where this was possible).
 We judged whether the soil was a peat soil based on its density, texture, and color.
@@
 Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively chosen, individual mires, using ground-penetrating radar (GPR).
-We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with its \unit{500\,MHz} shielded antenna mounted in a plastic sledge, and its control unit connected to a GNSS receiver.
-At each of the three mires we recorded GPR traces along walking transects that covered the extent of the mire, mostly in traversing, zigzag patterns with between \unit{5\,m} and \unit{20\,m} spacing at their vertices.
+We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with its 500 MHz\} shielded antenna mounted in a plastic sledge, and its control unit connected to a GNSS receiver.
+At each of the three mires we recorded GPR traces along walking transects that covered the extent of the mire, mostly in traversing, zigzag patterns with between 5 m and 20 m spacing at their vertices.
 Along the GPR transects we also probed peat depth at marked trace locations, to be able to calibrate the GPR wave speed velocity.
@@
 We used picks at marked trace locations to calibrate wave speed velocity; we pooled calibration points across the three mires and fitted a linear regression of depth on one-way travel time with the intercept fixed at zero.
-In total we had 46 calibration points along \unit{3.5\,km} of GPR transects.
+In total we had 46 calibration points along 3.5 km of GPR transects.
 Finally, we used the calculated wave velocity (\unit{0.0387\,m\,ns^{-1}}, \(R^2 = 0.874\)) to convert the travel times of all picks to calibrated peat depths.
@@
 Feature space coverage sampling works best when all dimensions are important predictors of the outcome \citep{wadouxSamplingDesignOptimization2019}, and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
-The radiometric predictors were downscaled to \unit{10\,m} resolution with cubic B-spline resampling in QGIS \citep[v.3,][]{QGISsoftware}.
+The radiometric predictors were downscaled to 10 m resolution with cubic B-spline resampling in QGIS \citep[v.3,][]{QGISsoftware}.
 We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
@@
 We navigated to the centers of the raster cells in the sample using real time kinematic differential GNSS (Topcon Positioning Systems, USA), to ensure sub-meter positional accuracy.
-At each location we measured peat depth three times by manual probing, with probe locations spaced approximately \unit{2.5\,m} apart.
+At each location we measured peat depth three times by manual probing, with probe locations spaced approximately 2.5 m apart.
 For a tiny fraction of these measurements (5 total), it was not possible to reach the base of the peat column, and a right-censored depth was recorded.
-In areas with dense sampling locations, we also measured peat depth with GPR along snaking transects passing through the centers of the sampling cells (seven transects, \unit{6.2\,km} total length).
-We used the same GPR system as at Skrimfjella, but with a \unit{100\,MHz} Malå rough terrain antenna (Guideline Geo AB, Sweden) at some transects.
+In areas with dense sampling locations, we also measured peat depth with GPR along snaking transects passing through the centers of the sampling cells (seven transects, 6.2 km total length).
+We used the same GPR system as at Skrimfjella, but with a 100 MHz Malå rough terrain antenna (Guideline Geo AB, Sweden) at some transects.
 To navigate the GPR transects, we placed flags at the cell centers of the sample locations, and used a handheld GNSS receiver to guide the GPR operator.
@@
 We processed the GPR data with Reflexw software (v.8.5, Sandmeier Scientific Software, Germany), applying a dewow filter, time-zero correction, bandpass filter, gain filter, and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennae.
-The last correction is important for the rough terrain antenna, which has an antenna separation (\unit{2.2\,m}) -- comparable to typical peat depths.
+The last correction is important for the rough terrain antenna, which has an antenna separation (2.2 m) -- comparable to typical peat depths.
 As with the data from Skrimfjella, we picked the base of the peat column from strong reflectors in the radargram, and calibrated wave velocity with manual probe measurements in a linear regression.
-The points in the regression were created by joining to each probe measurement the travel time of the nearest pick, but only if these were within \unit{2\,m} of each other.
-In total we had 78 calibration points along \unit{7.8\,km} of interpretable GPR traces (transect length exceeded because of extra GPR data).
+The points in the regression were created by joining to each probe measurement the travel time of the nearest pick, but only if these were within 2 m of each other.
+In total we had 78 calibration points along 7.8 km of interpretable GPR traces (transect length exceeded because of extra GPR data).
 Finally, we used the calculated wave velocity (\unit{0.0427\,m\,ns^{-1}}, \(R^2 = 0.946\)) to convert the travel times of all picks to calibrated peat depths.
@@
 The radiometric survey covering Skrimfjella was conducted in 2008--2011.
-The survey was flown at an average altitude of \unit{75\,m} and average speed of \unit{108\,km\,h^{-1}}, with flight lines spaced \unit{200\,m} apart.
+The survey was flown at an average altitude of 75 m and average speed of \unit{108\,km\,h^{-1}}, with flight lines spaced 200 m apart.
 Spectrometer count rates were calibrated annually to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
-The Geological Survey of Norway processed data from the spectrometer following standard procedures outlined by the International Atomic Energy Association, and the processing included: correction for aircraft and cosmic background radiation, correction for radon in the air, window stripping of the gamma ray spectrum, correction for flying height, conversion of count rates to ground concentrations, and finally gridding to \unit{50\,m} resolution with micro-leveling.
+The Geological Survey of Norway processed data from the spectrometer following standard procedures outlined by the International Atomic Energy Association, and the processing included: correction for aircraft and cosmic background radiation, correction for radon in the air, window stripping of the gamma ray spectrum, correction for flying height, conversion of count rates to ground concentrations, and finally gridding to 50 m resolution with micro-leveling.
 Further details about the survey and data processing are provided in Baranwal et al. \citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013}.
-We downscaled the processed data to \unit{10\,m} resolution by cubic spline resampling, using the \emph{terra} package in R.
+We downscaled the processed data to 10 m resolution by cubic spline resampling, using the \emph{terra} package in R.
 
 A very similar radiometric survey covering Ørskogfjellet was conducted in December 2014 and January 2015.
-This survey was flown at an average altitude of \unit{80\,m} and average speed of \unit{88\,km\,h^{-1}}, with flight lines also spaced \unit{200\,m} apart.
+This survey was flown at an average altitude of 80 m and average speed of \unit{88\,km\,h^{-1}}, with flight lines also spaced 200 m apart.
 Spectrometer count rates were calibrated in 2013 to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
@@
 Further details about the survey and data processing are provided in Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.
-The \unit{10\,m} resolution predictors from this survey were identical to the layers used in the sampling design (resampled from \unit{50\,m} resolution with cubic B-splines).
+The 10 m resolution predictors from this survey were identical to the layers used in the sampling design (resampled from 50 m resolution with cubic B-splines).
 
@@
 We used the \emph{terra} R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
-These were derived at two scales to produce eight different predictors; we either calculated from \unit{1\,m} DTM resolution and then aggregated, or aggregated to \unit{10\,m} DTM resolution and then calculated the indices.
+These were derived at two scales to produce eight different predictors; we either calculated from 1 m DTM resolution and then aggregated, or aggregated to 10 m DTM resolution and then calculated the indices.
 This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, dornikOptimalScalingPredictors2022, newmanAssessingSpatiallyHeterogeneous2023}.
-We know that peat depth tends to vary at fine scales in Norway, which is why we chose \unit{1\,m} and \unit{10\,m} resolutions \citep{maxwellLandsurfaceParametersSpatial2022}.
+We know that peat depth tends to vary at fine scales in Norway, which is why we chose 1 m and 10 m resolutions \citep{maxwellLandsurfaceParametersSpatial2022}.
 We also calculated the MRVBF index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
@@
 Next, we calculated the Topographic Wetness Index \citep{quinnPredictionHillslopeFlow1991}.
-This index is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate to coarse resolution DTMs \citep{agrenEvaluatingDigitalTerrain2014, riihimakiTopographicWetnessIndex2021}, so we calculated it from \unit{5\,m}, \unit{10\,m}, \unit{20\,m}, and \unit{50\,m} DTM resolution.
+This index is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate to coarse resolution DTMs \citep{agrenEvaluatingDigitalTerrain2014, riihimakiTopographicWetnessIndex2021}, so we calculated it from 5 m, 10 m, 20 m, and 50 m DTM resolution.
 The calculations were performed with Whitebox software \citep{lindsayWhiteboxGATCase2016}, accessed through the \emph{whitebox} R package \citep[v2.4,][]{wuWhiteboxWhiteboxToolsFrontend2022}.
@@
 It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature \citep[eq. 5 in][]{murphyTopographicModellingSoil2009}.
-We calculated unitless slope from the \unit{1\,m} DTM using the Whitebox software.
+We calculated unitless slope from the 1 m DTM using the Whitebox software.
 Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell \citep{schonauerSpatiotemporalPredictionSoil2021, schonauerRcodeCalculatingDepthwater2021}.
-This catchments area layer was then thresholded at seven different levels (\emph{flow initiation areas} from \unit{0.25\,ha} to \unit{16\,ha}) to estimate surface water features under moisture scenarios varying from wet to dry \citep{murphyModellingMappingTopographic2011, agrenEvaluatingDigitalTerrain2014, schonauerSpatiotemporalPredictionSoil2021}.
+This catchments area layer was then thresholded at seven different levels (\emph{flow initiation areas} 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry \citep{murphyModellingMappingTopographic2011, agrenEvaluatingDigitalTerrain2014, schonauerSpatiotemporalPredictionSoil2021}.
 In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
@@
 We prepared one categorical predictor -- peat depth class -- from a historical national map dataset called \emph{DMK} \citep{ahlstromAR5Klassifikasjonssystem2019}.
-The DMK peat depth classes are: \textless{} \unit{1\,m} (\emph{shallow}), \textgreater{} \unit{1\,m} (\emph{deep}), and \emph{unknown}.
-Mappers generally assigned peat depth classes to polygons of at least \unit{0.5\,ha}, although delineating polygons down to \unit{0.2\,ha} was allowed if peat depth showed a ``particularly marked difference'' \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
-We rasterized the peat depth class attribute to our \unit{10\,m} grid.
+The DMK peat depth classes are: \textless{} 1 m (\emph{shallow}), \textgreater{} 1 m (\emph{deep}), and \emph{unknown}.
+Mappers generally assigned peat depth classes to polygons of at least 0.5 ha, although delineating polygons down to 0.2 ha was allowed if peat depth showed a ``particularly marked difference'' \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
+We rasterized the peat depth class attribute to our 10 m grid.
 
@@
 We obtained depth measurements for 372 cells at Skrimfjella (area equivalent to 2.4 \% of mapped peatland) and 1878 cells at Ørskogfjellet (area equivalent to 1.2 \% of mapped peatland).
-Roughly 80 \% of these \unit{10\,m} cells were within mapped peatland, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
+Roughly 80 \% of these 10 m cells were within mapped peatland, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
 Coverage of DMK peat depth classes was higher at Ørskogfjellet than at Skrimfjella (79 \% versus 27 \% of cells), and at Ørskogfjellet the \emph{deep} and \emph{shallow} classes showed a larger difference in measured depth.
-Overall mean peat depths were similar at Skrimfjella and Ørskogfjellet: \unit{119\,cm} and \unit{126\,cm}, respectively.
+Overall mean peat depths were similar at Skrimfjella and Ørskogfjellet: 119 cm and 126 cm, respectively.
 
@@
 None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \ref{fig:modelMetrics}).
-For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an R\textsuperscript{2} of 0.34, and a mean absolute error of \unit{60\,cm}.
-For Ørskogfjellet the same values were 0.39, 0.33, and \unit{56\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
+For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an R\textsuperscript{2} of 0.34, and a mean absolute error of 60 cm.
+For Ørskogfjellet the same values were 0.39, 0.33, and 56 cm, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
 These values were derived from kNNDM spatial cross-validation with 20 folds at Skrimfjella and 10 folds at Ørskogfjellet.
@@
 The performance gap between the \emph{terrain and DMK} configuration and the \emph{terrain-only} configuration was similarly small.
-Compared to the \emph{terrain-only} configuration, the \emph{terrain and radiometric} configuration improved concordance correlation by 0.04, R\textsuperscript{2} by 0.04, and mean absolute error by \unit{1\,cm}.
+Compared to the \emph{terrain-only} configuration, the \emph{terrain and radiometric} configuration improved concordance correlation by 0.04, R\textsuperscript{2} by 0.04, and mean absolute error by 1 cm.
 \emph{DMK class} alone -- although calibrated to measured depths -- was a very poor predictor of peat depth, with a concordance correlation coefficient of 0.008.
@@
 Adding radiometric predictors to these configurations worsened model performance, especially in terms of concordance correlation and R\textsuperscript{2}.
-By itself, \emph{DMK class} produced a concordance correlation coefficient of 0.17 (compared to 0.008 at Skrimfjella), but the worst mean absolute error of any model at either site: \unit{77\,cm}.
+By itself, \emph{DMK class} produced a concordance correlation coefficient of 0.17 (compared to 0.008 at Skrimfjella), but the worst mean absolute error of any model at either site: 77 cm.
 
 The best models at both sites overpredicted shallow peats and strongly underpredicted very deep peats (Fig. \ref{fig:calPlots}).
-The mean error (bias) of these models was \unit{10\,cm} at Skrimfjella and \unit{-4\,cm} at Ørskogfjellet.
+The mean error (bias) of these models was 10 cm at Skrimfjella and -4 cm at Ørskogfjellet.
 
@@
 
-When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced worse mean absolute error than just assuming a constant \unit{30\,cm} depth (Fig.
+When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced worse mean absolute error than just assuming a constant 30 cm depth (Fig.
 \ref{fig:modelMetricsExtrapolation}).
@@
 Many of the the most important predictors in the best performing models showed non-monotonic effects on peat depth (Fig. \ref{fig:pdps}).
-At Ørskogfjellet for example, increasing \emph{elevation} was predictive of deeper peat up to about 75 meters above sea level, after which a further increase was predictive of shallower peat.
-At Skrimfjella the partial dependence on \emph{elevation} had the opposite shape, with the shallowest peats predicted at intermediate elevations, around 350 meters above sea level.
+At Ørskogfjellet for example, increasing \emph{elevation} was predictive of deeper peat up to about 75 m above sea level, after which a further increase was predictive of shallower peat.
+At Skrimfjella the partial dependence on \emph{elevation} had the opposite shape, with the shallowest peats predicted at intermediate elevations, around 350 m above sea level.
 \emph{TWI50m} at Ørskogfjellet and \emph{DTW4000} and \emph{radU} at Skrimfjella were other predictors that showed considerable fluctuations in their predictive effects on peat depth.
@@
 Individual conditional expectation lines indicated some interactions between predictors (Fig. \ref{fig:pdps}).
-For example, the magnitude of the increase in depth with \emph{elevation} that the model expected at Ørskogfjellet was different for different observations; some depth predictions increased by only \unit{40\,cm} while others increased by more than \unit{100\,cm}, over the same elevation gain.
+For example, the magnitude of the increase in depth with \emph{elevation} that the model expected at Ørskogfjellet was different for different observations; some depth predictions increased by only 40 cm while others increased by more than 100 cm, over the same elevation gain.
 Similarly, ICE lines of \emph{radU} at Skrimfjella were non-parallel, with some locations showing monotonically increasing peat depth predictions with uranium concentration (unlike the average effect).
@@
 Gatis et al.
-\citeyearpar{gatisMappingUplandPeat2019} used similar predictors and the same spatial grain, finding a much stronger correlation between predicted and observed peat depth (R\textsuperscript{2} = 0.68).
+\citeyearpar{gatisMappingUplandPeat2019} used similar predictors and the same spatial grain, finding a much stronger correlation between predicted and observed peat depth (\(R^2 = 0.68\)).
 Although their random evaluation data partition could make performance estimates too optimistic \citep{robertsCrossvalidationStrategiesData2017, wadouxSpatialCrossvalidationNot2021}, the confounding effect of spatial structure is probably small because they used linear regression and few predictors.

### 2024-12-05 ms/ms.Rmd
ms: grammer
ms: code and data statement

@@
   codedata: |
-    *Code and data availability.* Use this to add a statement when having data sets and software code available
+    R code used in this study is available at <https://github.com/julienvollering/DSMdepth>. Depth measurements will be archived in a public repository upon publication.
 authorcontribution: |
@@
 
-None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \@ref(fig:modelMetrics)).
+None of the models was able to predict peat depth across the study areas with high accuracy (Fig. \@ref(fig:modelMetrics)).
 For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an R^2^ of 0.34, and a mean absolute error of 60 cm.

### 2024-12-05 ms/ms.tex
ms: grammer
ms: code and data statement

@@
 
-None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \ref{fig:modelMetrics}).
+None of the models was able to predict peat depth across the study areas with high accuracy (Fig. \ref{fig:modelMetrics}).
 For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an R\textsuperscript{2} of 0.34, and a mean absolute error of 60 cm.
@@
 
-\codedataavailability{\emph{Code and data availability.} Use this to add a statement when having data sets and software code available} %% use this section when having data sets and software code available
+\codedataavailability{R code used in this study is available at \url{https://github.com/julienvollering/DSMdepth}.
+Depth measurements will be archived in a public repository upon publication.} %% use this section when having data sets and software code available
 

### 2025-01-07 ms/ms.Rmd
addressing comments in Intro + Methods up thru Sites

@@
   This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in Norway. 
-  Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data at two contrasting sites. 
-  We measured peat depth at 372 and 1878 locations and trained random forest models to predict depth at 10 m resolution. 
-  The best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth. 
+  Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data across two, >10 km^2^ sites. 
+  We measured peat depth by probing and ground-penetrating radar at 372 and 1878 locations at the two sites, respectively.
+  Then we trained random forest models using radiometric and terrain variables, plus the national map of peat depth, to predict peat depth at 10 m resolution. 
+  The two best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth. 
   Terrain variables were better predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth. 
-  Radiometric data showed inconsistent value, improving predictions at one site while degrading performance at the other. 
-  Our models improved upon the national peat depth map, but weak relationships with remotely sensed variables make peat depth hard to predict. 
-  These findings suggest that while digital soil mapping can enhance existing peat depth maps, reliable carbon stock assessments still require strategic field sampling. 
+  Radiometric variables showed inconsistent predictive value -- improving performance at one site while degrading it at the other.
+  The accuracy of the national map of peat depth did not measure up to any of our remote sensing models, even though it was calibrated to the same data.
+  Still, weak relationships with remotely sensed variables make peat depth hard to predict overall. 
+  Based on these findings, digital soil mapping can improve existing, broad-scale maps of peat depth, but local carbon stock assessment requires additional field sampling. 
   Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat occurrence and depth.
@@
 They store about 30 % of global soil carbon despite covering only 2--3 % of Earth's land [@xuPEATMAPRefiningEstimates2018; @friedlingsteinGlobalCarbonBudget2020; @unepGlobalPeatlandsAssessment2022]. 
-In other words, they are extremely carbon dense -- denser than any other ecosystem per square meter [@temminkRecoveringWetlandBiogeomorphic2022]. 
-This makes peatlands crucial to climate change mitigation. 
-Intact peatlands sequester carbon and produce a negative temperature forcing, overall [@joostenRolePeatlandsClimate2016]. 
+Peatlands (areas with peat soil) are more carbon dense per square meter than any other ecosystem [@temminkRecoveringWetlandBiogeomorphic2022]. 
+This makes them crucial to climate change mitigation. 
+Intact peatlands sequester carbon and overall produce a negative temperature forcing [@joostenRolePeatlandsClimate2016]. 
 When disturbed, often by conversion to another land use, they can produce large greenhouse gas emissions [@maGloballyRobustRelationship2022].
@@
 In the anoxic and acidic conditions created by a high water table, plant material decay is slightly outweighed by new growth, and the surplus carbon is laid down as peat. 
-There it remains sequestered as long as the water table stays high and prevents oxidation.
+There it remains sequestered as long as the water table stays high and prevents oxidative peat decomposistion.
 
@@
 However, other land uses also displace peatlands. 
-In Norway, where 9% of the land area is peatland [@brynLandCoverNorway2018], construction and development have become important drivers of loss, after lawmakers restricted conversion to forest and farmland in recent decades.
+In Norway -- where flat lowland is relatively scarce but 9 % of the land area is peatland [@brynLandCoverNorway2018] -- lawmakers have restricted peatland afforestation and cultivation in recent decades.
+Since then, a larger proportion of peatland loss is driven by construction [@flagetMyrMellomLokal2024].
 
-The spatial distribution of peat depth is often overlooked in land use planning and carbon accounting because peat depth is not mapped with sufficient coverage, resolution, or accuracy.
+The spatial distribution of peat depth is often overlooked in land use planning and carbon accounting because peat depth is not mapped with sufficient coverage, resolution, or accuracy [@beilmanPeatCarbonStocks2008; @hastieRisksCarbonStorage2022; @unepGlobalPeatlandsAssessment2022].
 Maps are crucial because they link high-level targets to specific management decisions, unlike spatially aggregated estimates [@oecdOECDEnvironmentalPerformance2022].
-For example, Norwegian regulations restrict conversion to farmland more strictly where peat depth exceeds 1 m, and the distribution of peat depth on a farmer's property determines whether he is allowed to convert peatland (Forskrift om nydyrking, 1997, § 5a).
-Maps also make it possible to quantify the effect of specific management decisions and thereby understand how local outcomes contribute to regional and national outcomes [@oecdOECDEnvironmentalPerformance2022].
+For example, Norwegian regulations restrict conversion of peatland to farmland more strictly where peat depth exceeds 1 m, and the distribution of peat depth on a farmer's property determines whether they are allowed to convert peatland (Forskrift om nydyrking, 1997, § 5a).
+Maps of peat depth also make it possible to quantify the effect of specific management decisions and thereby understand how local outcomes contribute to regional and national outcomes [@oecdOECDEnvironmentalPerformance2022].
 
-Measuring peat depth on the ground is straightforward, and a field survey can map a small area at low cost. 
-However, surveying large areas is impractical when depth varies over short distances -- as in most peatlands. 
+Measuring peat depth on the ground by probing or ground-penetrating radar is straightforward, and a field survey can map a small area at low cost. 
+However, surveying large areas is impractical when depth varies over short distances (e.g. < 1 m) -- as in most peatlands. 
 A complementary approach from soil science is digital soil mapping (DSM). 
 DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest. 
-This approach has grown important with the availability of remotely sensed data and the advancement of methods for identifying pattern, especially through machine learning [@minasnyDigitalMappingPeatlands2019; @wadouxMachineLearningDigital2020].
+This approach has grown in importance with the availability of remotely sensed data and the advancement of methods for identifying patterns, especially through machine learning [@minasnyDigitalMappingPeatlands2019; @wadouxMachineLearningDigital2020].
 
-The crux for DSM of peat depth are the relationships between peat depth and the other, mapped variables. 
+The crux for DSM of peat depth is the relationship between peat depth and each of the other, mapped variables. 
 For DSM to work, these relationships must be strong and consistent over the area of interest.
-They can be purely correlative rather than causal, but mechanistic relationships are stronger and more consistent. 
+They can be purely correlative rather than causal, but mechanistic relationships are stronger and more consistent than non-causal ones. 
 The *scorpan* framework for DSM suggests seven predictor classes to explore: other soil properties, climate, organisms, relief (topography), parent material, age, and spatial position [@mcbratneyDigitalSoilMapping2003].
 
-The most practical and widespread *scorpan* factors for peat depth are relief and spatial position. 
+For peat depth, the most practical and widespread *scorpan* factors are relief and spatial position. 
 Spatial position is unique because it is always known (with varying accuracy). 
@@
 The degree of attenuation relates to properties of the overburden, especially composition and depth. 
-Deep soil with high water content attenuates most [@beamishGammaRayAttenuation2013; @reinhardtGammaraySpectrometryVersatile2019]. 
-Thus, gamma-ray radiometrics integrate the *scorpan* factors soil, parent material, and age.
+Deep soil with high water content will attenuate radiation most [@beamishGammaRayAttenuation2013; @reinhardtGammaraySpectrometryVersatile2019]. 
+Thus, gamma-ray radiometrics integrate two *scorpan* factors: other soil properties and parent material [@mcbratneyDigitalSoilMapping2003].
 
 Although theory suggests that one meter of peat may fully attenuate radiometric signal [@beamishGammaRayAttenuation2013; @reinhardtGammaraySpectrometryVersatile2019], empirical investigations show that the association between peat depth and radioactivity can extend beyond the first meter [@keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @kogantiMappingPeatDepth2023].
-Radiometric data are increasingly available over large areas, presenting a big opportunity insofar as they are predictive. 
-Airborne surveys are used for various applications, and some countries have high spatial coverage of such data  [@minasnyDigitalMappingPeatlands2019; @baranwalAirborneGeophysicalSurveys2020].
+Radiometric data are increasingly available over large areas, presenting a big opportunity for mapping peat depth, insofar as they are predictive of peat depth. 
+Airborne surveys are used for various applications like mineral prospecting and radon risk management, and some countries have high spatial coverage of such data [@minasnyDigitalMappingPeatlands2019; @baranwalAirborneGeophysicalSurveys2020].
 
@@
 Norway may be instructive in this respect because its peatlands vary widely across climates and topographies. 
-There are many mire massif types, with different hydrology, formation, and development -- from topogeneous or soligeneous fens to blanket or raised bogs [@lyngstadBeskrivelserAvTorvmassivenheter2023].
-These types have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric variables will vary by landscape.
+Norway has 22 peatland mesotopes, with different hydrology, formation, and development -- from topogeneous or soligeneous fens to blanket or raised bogs [@joostenWiseUseMires2002; @lyngstadBeskrivelserAvTorvmassivenheter2023].
+These mesotopes have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric variables will vary by landscape.
 
@@
 After scattered surveys in the early 20th century, a comprehensive round of surveying was completed 1964--2001 as part of a wider land cover mapping in Norway  [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
-Because of its agricultural and silvicultural focus, it covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable [@ahlstromAR5Klassifikasjonssystem2019].
-Field surveyors carried a 1 m probe, so they measured peat depth as shallow (< 1 m), or deep (> 1 m). 
-These classes were assigned to whole mire polygons, so spatial resolution is on the order of hectares.
+This survey produced the initial maps from which Norway's most detailed updated land cover datasets are derived -- including the AR5 and DMK datasets used in this study and described later.
+Because of its agricultural and silvicultural focus, the survey covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable [@ahlstromAR5Klassifikasjonssystem2019].
+Field surveyors carried a 1 m probe and they measured peat depth as a categorical variable: shallow (< 1 m), or deep (> 1 m). 
+These classes were assigned to whole polygons, so spatial resolution is on the order of hectares.
+The number of measurements per unit area was not standardized but probably low.
 
-Soil science needs more studies of relationships between peat depth and topographic or radiometric data to determine how predictive they are, which variables are most predictive, and how consistent their associations.
-The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify rich peatland carbon stocks and avoid their conversion.
-Although small areas can be mapped accurately on the ground, having landscape-scale maps *before* detailed investigation increases the option space for spatial planning and land management. 
+The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify peatlands with rich carbon stocks to avoid their conversion and prioritize their preservation or restoration.
+Although ground surveys of peat depth are accurate and feasible across small areas, having landscape-scale maps *before* detailed investigation increases the option space for spatial planning and land management.
 For example, the Norwegian Public Roads Administration routinely measures peat depth during geotechnical work, but by that time, the route of the road is already set. 
 With a digital peat depth map, planners could better compare routes and climate mitigation measures would have more leverage.
+Therefore, soil science needs more studies of relationships between peat depth and topographic or radiometric data to determine how predictive they are, which variables are most predictive, and how consistent their associations are.
 
-We assess how well remotely sensed topographic and radiometric data can predict peat depth at the landscape scale, with a view towards revising regional and national maps in Norway. 
+Here, we assess how well remotely sensed topographic and radiometric data can predict peat depth at the landscape scale, with a view towards revising regional and national maps in Norway. 
 Specifically, we: (1) quantify the accuracy of predictions from topographic and radiometric variables, (2) identify key predictive variables, and (3) compare these relationships across two different Norwegian landscapes.
 To our knowledge, this is the first study to predict peat depth from airborne radiometric data using machine learning algorithms. 
-Where airborne radiometric data have been used to predict peat depth, it has been through modeling techniques less suited for prediction and spatial extrapolation [e.g., @keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @siemonAirborneElectromagneticRadiometric2020].
-Where machine learning algorithms have been used on airborne radiometric data, it has been to predict peat extent rather than depth [e.g., @olearyDigitalSoilMapping2022].
+Where airborne radiometric data have previously been used to predict peat depth, it has been through modeling techniques less suited for prediction and spatial extrapolation [e.g., @keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @siemonAirborneElectromagneticRadiometric2020].
+Where machine learning algorithms have previously been used on airborne radiometric data, it has been to predict peat extent rather than depth [e.g., @olearyDigitalSoilMapping2022].
 
@@
 
-We assessed how well we could predict peat depth at two sites with conspicuously different physical geography: Skrimfjella in eastern Norway and Ørskogfjellet in western Norway (Fig.
+To assess the predictive value of terrain and radiometric data for mapping peat depth, we chose two sites with conspicuously different physical geography: Skrimfjella (Fig.
+\@ref(fig:sites)b) in eastern Norway and Ørskogfjellet (Fig.
+\@ref(fig:sites)a) in western Norway (Fig.
 \@ref(fig:sites)c).
-These sites were chosen because they were covered by radiometric data from airborne surveys, relatively little built-up area, and road access.
+These sites were chosen because they were covered by radiometric data from airborne surveys, had relatively little built-up area, and were accessible by road.
 
-```{r sites, out.height="90%", echo = FALSE, fig.cap = "Study areas at Ørskogfjellet (a) and Skrimfjella (b) within southern Norway (c).
-Land cover shown here is from the AR50 national land resource database and has simplified geometry with respect to the AR5 database used in the study."}
+```{r sites, out.height="82%", echo = FALSE, fig.cap = "Sites in southern Norway (a) with land cover at Skrimfjella (b) and Ørskogfjellet (c).
+To fit the scale of the maps, the land cover shown here is from the AR50 national land resource dataset, which has simplified geometry with respect to the AR5 dataset that is used in the study.
+Terrain is visualized using a hillshade with slope 45° and azimuth 225°.
+AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (<https://data.norge.no/nlod/en/1.0>) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (<https://creativecommons.org/licenses/by/4.0/>)."}
 knitr::include_graphics("figures/sites-patchwork.pdf")
@@
 At Skrimfjella we delineated a study area of 34 km^2^ based on radiometric coverage and accessibility (Fig. \@ref(fig:sites)b).
-The study area has a diverse bedrock, with 32 % alkali feltspat granite, 26 % mergelstein, 10 % granite, and eight other types with > 1 % coverage (NGU, 1:250 000 dataset).
-The landscape within our delineation is classified as *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
+The study area has a diverse bedrock, with 32 % alkali feldspar granite, 26 % marlstone, 10 % granite, 8 % mozonite, 7 % sandstone, 6 % limestone, and five other rock types with 1--5 % coverage (NGU, 1:250 000 dataset).
+The area within our delineation at Skrimfjella is classified as the landscape type *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
 It is almost without human infrastructure, dominated by forest, and borders on a large nature reserve.
-The study area has a mean elevation of 438 m above sea level (range 223--711, IQR 351--509), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).
-In Norway's AR5 national land capability dataset [@ahlstromAR5Klassifikasjonssystem2019], 1.5 km^2^ (4.5 %) of the study area is classified as mire -- defined as areas with mire vegetation and at least 30 cm of peat depth.
+The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).
+In Norway's AR5 national land cover dataset ["areal resources in scale 1:5000", @ahlstromAR5Klassifikasjonssystem2019], 1.5 km^2^ (4.5 %) of the study area is classified as 'mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.
 
-At Ørskogfjellet we defined a study area of 124 km^2^ which basically followed the footprint of the radiometric survey (Fig. \@ref(fig:sites)a).
-According to the Geological Survey of Norway, bedrock in the area is 84 % granitic gneiss,  11 % granite, and 5 % aluminium silicate gneiss (NGU, 1:250 000 dataset). 
+At Ørskogfjellet we defined a study area of 124 km^2^ which basically followed the southernmost part of the radiometric survey extent (Fig. \@ref(fig:sites)a).
+Bedrock in the area is 84 % granitic gneiss,  11 % granite, and 5 % aluminium silicate gneiss (NGU, 1:250 000 dataset). 
 This study area comprises a wide range of major landscape types: *coastal plains*, *coastal fjord*, *inland valleys*, as well as *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
 It is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
-Its mean elevation is 211 m above sea level (range 0--807, IQR 73--310), and its mean slope at 10 m resolution is 13.0° (IQR 4.7--18.3°).
-The AR5 dataset counts 15.3 km^2^ (12.4 %) of the study area as mire.
+Its mean elevation is 211 m above sea level (range 0--807 m, IQR 73--310 m), and its mean slope at 10 m resolution is 13.0° (IQR 4.7--18.3°).
+In the AR5 dataset, 15.3 km^2^ (12.4 %) of the study area is classified as mire.
 
@@
 We chose for our sampling and modelling a spatial resolution of 10 m.
-We considered this a reasonable compromise between DTM resolution (1 m) and small mires on the one hand, and airborne radiometric resolution (50 m) on the other.
+We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other.
 The point measurements of peat depth described below were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
@@
 We measured peat depth in selected locations (10 m raster cells) at Skrimfjella.
-The locations were chosen only from areas delineated as mire in the AR5 national land capability dataset.
-Within this mire area, we stratified our sample across values of elevation, slope, and potassium ground concentration [from processed airborne gamma ray spectrometry, @baranwalHelicopterborneMagneticElectromagnetic2013].
+The locations were chosen only from areas delineated as mire in the AR5 dataset.
+Within these AR5 mires, we stratified our sample across values of elevation, slope, and potassium ground concentration [from processed airborne gamma ray spectrometry, @baranwalHelicopterborneMagneticElectromagnetic2013].
 Specifically, we used the *eSample* function in the *iSDM* R package (v.1.0) to chose an environmentally systematic sample.
@@
 In addition to the peat depth locations, we had another arm of our sampling design for measuring peatland occurrence, as a binary variable.
-We wanted to measure peatland occurrence outside of mapped mire areas because the AR5 dataset is known to underestimate peatland coverage [especially in forests, @brynLandCoverNorway2018], and because airborne radiometrics may help identify unmapped peatland [@gatisMappingUplandPeat2019; @olearyDigitalSoilMapping2022].
-The occurrence locations were sampled from the part of the study area that (1) was mapped as something other than mire in the AR5 database and (2) had a slope < 20°. 
+We wanted to measure peatland occurrence outside of AR5 mires because the AR5 dataset is known to underestimate peatland coverage [especially in forests, @brynLandCoverNorway2018], and because airborne radiometrics may help identify unmapped peatland [@gatisMappingUplandPeat2019; @olearyDigitalSoilMapping2022].
+The occurrence locations were sampled from the part of the study area that (1) was mapped as something other than mire in the AR5 dataset and (2) had a slope < 20°. 
 We performed environmentally systematic sampling of this population with the same procedure as for the depth locations, and *eSample* returned 106 raster cells.
@@
 
-Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively chosen, individual mires, using ground-penetrating radar (GPR).
+Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively chosen, separate peatlands, using ground-penetrating radar (GPR).
 We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with its 500 MHz} shielded antenna mounted in a plastic sledge, and its control unit connected to a GNSS receiver.
-At each of the three mires we recorded GPR traces along walking transects that covered the extent of the mire, mostly in traversing, zigzag patterns with between 5 m and 20 m spacing at their vertices.
+At each of the three peatlands we recorded GPR traces along walking transects that covered the extent of the peatland, mostly in traversing, zigzag patterns with between 5 m and 20 m spacing at their vertices.
 Along the GPR transects we also probed peat depth at marked trace locations, to be able to calibrate the GPR wave speed velocity.
@@
 Then we picked strong reflectors in the radargrams that we interpreted as the base of the peat column.
-We used picks at marked trace locations to calibrate wave speed velocity; we pooled calibration points across the three mires and fitted a linear regression of depth on one-way travel time with the intercept fixed at zero.
+We used picks at marked trace locations to calibrate wave speed velocity; we pooled calibration points across the three peatlands and fitted a linear regression of depth on one-way travel time with the intercept fixed at zero.
 In total we had 46 calibration points along 3.5 km of GPR transects.
@@
 
-At Ørskogfjellet we also measured peat depth in a sample of 10 m raster cells, selected from the part of the study area classified in the AR5 dataset as mire.
-Before selecting locations, we determined a minimal sample size that would adequately capture the terrain and radiometric properties of the entire mire area.
+At Ørskogfjellet we also measured peat depth in a sample of 10 m raster cells, selected from the part of the study area classified as mire in the AR5 dataset.
+Before selecting locations, we determined a minimal sample size that would adequately capture the terrain and radiometric properties of the entire AR5 mire area.
 Specifically, we aimed to identify the size at which adding locations produced diminishing decreases in divergence between sample and population distributions -- i.e.
@@
 Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
-To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly accessible roads.
+To adjust, we first restricted the sampling population to AR5 mire areas that were within an arbitrary cost distance of publicly accessible roads.
 Cost distance was calculated using GRASS's *r.walk* function, with friction costs defined by AR5 land classes [@GRASSv8-2].
@@
 
-We prepared one categorical predictor -- peat depth class -- from a historical national map dataset called *DMK* [@ahlstromAR5Klassifikasjonssystem2019].
-The DMK peat depth classes are: < 1 m (*shallow*), > 1 m (*deep*), and *unknown*.
-Mappers generally assigned peat depth classes to polygons of at least 0.5 ha, although delineating polygons down to 0.2 ha was allowed if peat depth showed a "particularly marked difference" [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
+We prepared one categorical predictor -- peat depth class -- from the national map dataset called *DMK* [@ahlstromAR5Klassifikasjonssystem2019].
+The DMK dataset is derived from the same historical surveys as the AR5 dataset, and peat depth classes are: < 1 m (*shallow*), > 1 m (*deep*), and *unknown*.
+Surveyors generally assigned peat depth classes to polygons of at least 0.5 ha, although delineating polygons down to 0.2 ha was allowed if peat depth showed a "particularly marked difference" [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
 We rasterized the peat depth class attribute to our 10 m grid.
@@
 
-While the primary aim of the DSM was to predict peat depth within mapped peatlands (where peat depth > 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth or occurrence outside of mapped peatland.
-In other words, we tested whether our peatland-trained models of peat depth could indicate which other locations were most likely to have peat.
-It is generally impractical to obtain good data for this purpose, because locations with non-zero peat depth outside of mapped peatland are unknown and relatively rare.
-For Skrimfjella, we had independent data from the arm of our sampling design that measured peatland occurrence outside of mapped mire areas.
+While the primary aim of the DSM was to predict peat depth within AR5 mires (where peat depth is supposed to be > 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth or occurrence outside of AR5 mire areas.
+In other words, we tested whether our mire-trained models of peat depth could identify peatlands (with or without mire vegetation).
+It is difficult to obtain good data for this purpose, because peat soil not covered by mire vegetation frequently goes unidentified and is relatively rare at these sites.
+For Skrimfjella, we had independent data from the arm of our sampling design that measured peatland occurrence outside of AR5 mires.
 For Ørskogfjellet we did not have such data.
-Therefore, we first -- for both sites -- performed an evaluation with just our peat depth measurements, which included a fraction of locations not mapped as peatland.
-For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on locations mapped as peatland (in the training folds), and evaluated them only on locations not mapped as peatland (in the test fold).
-In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
+Therefore, we first -- for both sites -- performed an evaluation with only our peat depth measurements, which included a fraction of locations outside of AR5 mires.
+For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on AR5 mire locations (in the training folds), and evaluated them only on (test fold) locations classified in AR5 as something other than mire.
+In this way, we simulated the situation where a model trained on AR5 mires is used to predict peat depth outside AR5 mires.
 From the cross-validation we quantified mean absolute error.
-Second -- for Skrimfjella -- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
+Second -- for Skrimfjella alone -- we simply trained a model on all peat depth measurements (AR5 mire or not) and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
 In this analysis, the model's prediction of peat depth is treated as a rank index of peatland likelihood.
-Since DMK peat depth class is always undefined outside known peatlands, we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.
+Since the purpose was to test the models' ability to uncover unmapped peatland (where DMK peat depth class is undefined), we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.
 
@@
 
-We obtained depth measurements for 372 cells at Skrimfjella (area equivalent to 2.4 % of mapped peatland) and 1878 cells at Ørskogfjellet (area equivalent to 1.2 % of mapped peatland).
-Roughly 80 % of these 10 m cells were within mapped peatland, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
+We obtained depth measurements for 372 cells at Skrimfjella (2.4 % of AR5 mire area) and 1878 cells at Ørskogfjellet (1.2 % of AR5 mire area).
+Roughly 80 % of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
 Coverage of DMK peat depth classes was higher at Ørskogfjellet than at Skrimfjella (79 % versus 27 % of cells), and at Ørskogfjellet the *deep* and *shallow* classes showed a larger difference in measured depth.
@@
 
-When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced worse mean absolute error than just assuming a constant 30 cm depth (Fig.
+When we tested how well models extrapolated from within to outside of AR5 mire areas, we found that the models produced worse mean absolute error than just assuming a constant 30 cm depth (Fig.
 \@ref(fig:modelMetricsExtrapolation)).
@@
 
-```{r modelMetricsExtrapolation, echo = FALSE, fig.cap = "Performance of models that extrapolate from training data inside of mapped peatland to test data outside of mapped peatland, evaluated via spatial cross-validation.
+```{r modelMetricsExtrapolation, echo = FALSE, fig.cap = "Performance of models that extrapolate from training data inside of mapped mires to test data outside of mapped mires (AR5 dataset), evaluated via spatial cross-validation.
 Parentheses denote the number of variables in each model, and point estimates are shown +/- their standard error."}
@@
 The performance of our terrain- and radiometric-based maps could have been improved with a spatially explicit mapping approach like regression kriging.
-By ignoring spatial autocorrelation in peat depth (i.e. the n in scorpan), we have not extracted all of the information about peat depth in the study area out of the training data.
+By ignoring spatial autocorrelation in peat depth (i.e. the n in *scorpan*), we have not extracted all of the information about peat depth in the study area out of the training data.
 If we were to put our DSM approach into production for published maps, we would harness the spatial component, but the actual maps for Ørskogfjellet and Skrimfjella are not of primary interest in this study.
@@
 
-That we measured more than 30 cm of peat in areas not mapped as peatland was expected, because AR5 underrepresents peatland extent by about one third [@brynLandCoverNorway2018]. 
+That we measured more than 30 cm of peat in areas not mapped as mire was expected, because AR5 underrepresents peatland extent by about one third [@brynLandCoverNorway2018]. 
 For that reason, maps of peatland occurrence also need revising. 
-Unfortunately, our models did not predict peat occurrence outside mapped peatlands. 
+Unfortunately, our models did not predict peat occurrence outside AR5 mires. 
 In these areas, assuming a constant 30 cm of peat (the depth threshold used in AR5) produced less error than our models. 
-Moreover, the performance gap would likely have been larger had our evaluation data been representative of the non-peatland areas. 
-Instead they were near mapped peatlands. 
+Moreover, the performance gap would likely have been larger had our evaluation data been representative of all areas not classified as mire in AR5. 
+Instead they were concentrated near AR5 mires. 
 Most importantly, the independent occurrence data at Skrimfjella showed the model completely failing to predict occurrence, and these data were suited for testing this ability. 
@@
 Why wasn't peat occurrence predicted well? 
-The models were trained on fundamentally different populations of locations (~80 % mapped peatland in training vs.
+The models were trained on fundamentally different populations of locations (~80 % AR5 mire in training vs.
 0 % in prediction), so it is not surprising that the associations they learned did not transfer well.
-Moreover, the ~20 % of training data from outside mapped peatland were incidentally collected (near peatland) and not representative of other land cover classes. 
+Moreover, the ~20 % of training data from outside AR5 mires were incidentally collected (near AR5 mires) and not representative of other land cover classes. 
 In short, the models were blind to the fact that most of both study areas have no peat.
@@
 However, it is unclear how big (N) any cluster of depths should be, and the tradeoff between multiple small-extent DSM and fewer large-extent DSM needs research. 
-Bohn and Miller [-@bohnLocallyEnhancedDigital2024] advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions, where the composition of peatland types is constant [e.g.
-@moenVegetasjonstyper1998].
+Bohn and Miller [-@bohnLocallyEnhancedDigital2024] advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or 'supertopes'), where the composition of mesotopes is similar [@moenVegetasjonstyper1998; @joostenWiseUseMires2002].
 Depth varies strongly between peats formed by terrestrialization versus paludification [@buffamFillingHolesRegional2010], and probably along other axes of peatland typology. 
-Although most Norwegian landscapes contain multiple, unknown types, it makes sense to limit their diversity within a single DSM.
+Although most Norwegian landscapes contain multiple, unknown mesotopes, it makes sense to limit their diversity within a single DSM.
 
@@
 We caution against reducing continuous depth predictions to arbitrary classes [as in @ivanovsModelingGeospatialDistribution2024; @karjalainenComparisonTwoGammaray2024], since classes can be derived from continuous predictions.
-The distribution of peat depths across landscapes is zero-inflated, and research is needed to determine whether it is more efficient to parameterize a single model of peat depth (with a larger, generalized dataset) or to break down the problem by classifying zero depth and then regressing non-zero depths (with smaller, specialized datasets).
+The distribution of peat depths across full landscapes is zero-inflated, and research is needed to determine whether it is more efficient to parameterize a single model of peat depth (with a larger, generalized dataset) or to break down the problem by classifying zero depth and then regressing non-zero depths (with smaller, specialized datasets).
 Coupling occurrence and depth will reduce the prevalence of incoherence that we found: deep peat outside the peatland extent and zero depth inside it. 

### 2025-01-07 ms/ms.tex
addressing comments in Intro + Methods up thru Sites

@@
 This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in Norway.
-Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data at two contrasting sites.
-We measured peat depth at 372 and 1878 locations and trained random forest models to predict depth at 10 m resolution.
-The best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth.
+Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data across two, \textgreater10 km\textsuperscript{2} sites.
+We measured peat depth by probing and ground-penetrating radar at 372 and 1878 locations at the two sites, respectively.
+Then we trained random forest models using radiometric and terrain variables, plus the national map of peat depth, to predict peat depth at 10 m resolution.
+The two best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth.
 Terrain variables were better predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth.
-Radiometric data showed inconsistent value, improving predictions at one site while degrading performance at the other.
-Our models improved upon the national peat depth map, but weak relationships with remotely sensed variables make peat depth hard to predict.
-These findings suggest that while digital soil mapping can enhance existing peat depth maps, reliable carbon stock assessments still require strategic field sampling.
+Radiometric variables showed inconsistent predictive value -- improving performance at one site while degrading it at the other.
+The accuracy of the national map of peat depth did not measure up to any of our remote sensing models, even though it was calibrated to the same data.
+Still, weak relationships with remotely sensed variables make peat depth hard to predict overall.
+Based on these findings, digital soil mapping can improve existing, broad-scale maps of peat depth, but local carbon stock assessment requires additional field sampling.
 Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat occurrence and depth.
@@
 They store about 30 \% of global soil carbon despite covering only 2--3 \% of Earth's land \citep{xuPEATMAPRefiningEstimates2018, friedlingsteinGlobalCarbonBudget2020, unepGlobalPeatlandsAssessment2022}.
-In other words, they are extremely carbon dense -- denser than any other ecosystem per square meter \citep{temminkRecoveringWetlandBiogeomorphic2022}.
-This makes peatlands crucial to climate change mitigation.
-Intact peatlands sequester carbon and produce a negative temperature forcing, overall \citep{joostenRolePeatlandsClimate2016}.
+Peatlands (areas with peat soil) are more carbon dense per square meter than any other ecosystem \citep{temminkRecoveringWetlandBiogeomorphic2022}.
+This makes them crucial to climate change mitigation.
+Intact peatlands sequester carbon and overall produce a negative temperature forcing \citep{joostenRolePeatlandsClimate2016}.
 When disturbed, often by conversion to another land use, they can produce large greenhouse gas emissions \citep{maGloballyRobustRelationship2022}.
@@
 In the anoxic and acidic conditions created by a high water table, plant material decay is slightly outweighed by new growth, and the surplus carbon is laid down as peat.
-There it remains sequestered as long as the water table stays high and prevents oxidation.
+There it remains sequestered as long as the water table stays high and prevents oxidative peat decomposistion.
 
@@
 However, other land uses also displace peatlands.
-In Norway, where 9\% of the land area is peatland \citep{brynLandCoverNorway2018}, construction and development have become important drivers of loss, after lawmakers restricted conversion to forest and farmland in recent decades.
+In Norway -- where flat lowland is relatively scarce but 9 \% of the land area is peatland \citep{brynLandCoverNorway2018} -- lawmakers have restricted peatland afforestation and cultivation in recent decades.
+Since then, a larger proportion of peatland loss is driven by construction \citep{flagetMyrMellomLokal2024}.
 
-The spatial distribution of peat depth is often overlooked in land use planning and carbon accounting because peat depth is not mapped with sufficient coverage, resolution, or accuracy.
+The spatial distribution of peat depth is often overlooked in land use planning and carbon accounting because peat depth is not mapped with sufficient coverage, resolution, or accuracy \citep{beilmanPeatCarbonStocks2008, hastieRisksCarbonStorage2022, unepGlobalPeatlandsAssessment2022}.
 Maps are crucial because they link high-level targets to specific management decisions, unlike spatially aggregated estimates \citep{oecdOECDEnvironmentalPerformance2022}.
-For example, Norwegian regulations restrict conversion to farmland more strictly where peat depth exceeds 1 m, and the distribution of peat depth on a farmer's property determines whether he is allowed to convert peatland (Forskrift om nydyrking, 1997, § 5a).
-Maps also make it possible to quantify the effect of specific management decisions and thereby understand how local outcomes contribute to regional and national outcomes \citep{oecdOECDEnvironmentalPerformance2022}.
+For example, Norwegian regulations restrict conversion of peatland to farmland more strictly where peat depth exceeds 1 m, and the distribution of peat depth on a farmer's property determines whether they are allowed to convert peatland (Forskrift om nydyrking, 1997, § 5a).
+Maps of peat depth also make it possible to quantify the effect of specific management decisions and thereby understand how local outcomes contribute to regional and national outcomes \citep{oecdOECDEnvironmentalPerformance2022}.
 
-Measuring peat depth on the ground is straightforward, and a field survey can map a small area at low cost.
-However, surveying large areas is impractical when depth varies over short distances -- as in most peatlands.
+Measuring peat depth on the ground by probing or ground-penetrating radar is straightforward, and a field survey can map a small area at low cost.
+However, surveying large areas is impractical when depth varies over short distances (e.g.~\textless{} 1 m) -- as in most peatlands.
 A complementary approach from soil science is digital soil mapping (DSM).
 DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest.
-This approach has grown important with the availability of remotely sensed data and the advancement of methods for identifying pattern, especially through machine learning \citep{minasnyDigitalMappingPeatlands2019, wadouxMachineLearningDigital2020}.
+This approach has grown in importance with the availability of remotely sensed data and the advancement of methods for identifying patterns, especially through machine learning \citep{minasnyDigitalMappingPeatlands2019, wadouxMachineLearningDigital2020}.
 
-The crux for DSM of peat depth are the relationships between peat depth and the other, mapped variables.
+The crux for DSM of peat depth is the relationship between peat depth and each of the other, mapped variables.
 For DSM to work, these relationships must be strong and consistent over the area of interest.
-They can be purely correlative rather than causal, but mechanistic relationships are stronger and more consistent.
+They can be purely correlative rather than causal, but mechanistic relationships are stronger and more consistent than non-causal ones.
 The \emph{scorpan} framework for DSM suggests seven predictor classes to explore: other soil properties, climate, organisms, relief (topography), parent material, age, and spatial position \citep{mcbratneyDigitalSoilMapping2003}.
 
-The most practical and widespread \emph{scorpan} factors for peat depth are relief and spatial position.
+For peat depth, the most practical and widespread \emph{scorpan} factors are relief and spatial position.
 Spatial position is unique because it is always known (with varying accuracy).
@@
 The degree of attenuation relates to properties of the overburden, especially composition and depth.
-Deep soil with high water content attenuates most \citep{beamishGammaRayAttenuation2013, reinhardtGammaraySpectrometryVersatile2019}.
-Thus, gamma-ray radiometrics integrate the \emph{scorpan} factors soil, parent material, and age.
+Deep soil with high water content will attenuate radiation most \citep{beamishGammaRayAttenuation2013, reinhardtGammaraySpectrometryVersatile2019}.
+Thus, gamma-ray radiometrics integrate two \emph{scorpan} factors: other soil properties and parent material \citep{mcbratneyDigitalSoilMapping2003}.
 
 Although theory suggests that one meter of peat may fully attenuate radiometric signal \citep{beamishGammaRayAttenuation2013, reinhardtGammaraySpectrometryVersatile2019}, empirical investigations show that the association between peat depth and radioactivity can extend beyond the first meter \citep{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, kogantiMappingPeatDepth2023}.
-Radiometric data are increasingly available over large areas, presenting a big opportunity insofar as they are predictive.
-Airborne surveys are used for various applications, and some countries have high spatial coverage of such data \citep{minasnyDigitalMappingPeatlands2019, baranwalAirborneGeophysicalSurveys2020}.
+Radiometric data are increasingly available over large areas, presenting a big opportunity for mapping peat depth, insofar as they are predictive of peat depth.
+Airborne surveys are used for various applications like mineral prospecting and radon risk management, and some countries have high spatial coverage of such data \citep{minasnyDigitalMappingPeatlands2019, baranwalAirborneGeophysicalSurveys2020}.
 
@@
 Norway may be instructive in this respect because its peatlands vary widely across climates and topographies.
-There are many mire massif types, with different hydrology, formation, and development -- from topogeneous or soligeneous fens to blanket or raised bogs \citep{lyngstadBeskrivelserAvTorvmassivenheter2023}.
-These types have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric variables will vary by landscape.
+Norway has 22 peatland mesotopes, with different hydrology, formation, and development -- from topogeneous or soligeneous fens to blanket or raised bogs \citep{joostenWiseUseMires2002, lyngstadBeskrivelserAvTorvmassivenheter2023}.
+These mesotopes have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric variables will vary by landscape.
 
@@
 After scattered surveys in the early 20th century, a comprehensive round of surveying was completed 1964--2001 as part of a wider land cover mapping in Norway \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
-Because of its agricultural and silvicultural focus, it covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable \citep{ahlstromAR5Klassifikasjonssystem2019}.
-Field surveyors carried a 1 m probe, so they measured peat depth as shallow (\textless{} 1 m), or deep (\textgreater{} 1 m).
-These classes were assigned to whole mire polygons, so spatial resolution is on the order of hectares.
+This survey produced the initial maps from which Norway's most detailed updated land cover datasets are derived -- including the AR5 and DMK datasets used in this study and described later.
+Because of its agricultural and silvicultural focus, the survey covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable \citep{ahlstromAR5Klassifikasjonssystem2019}.
+Field surveyors carried a 1 m probe and they measured peat depth as a categorical variable: shallow (\textless{} 1 m), or deep (\textgreater{} 1 m).
+These classes were assigned to whole polygons, so spatial resolution is on the order of hectares.
+The number of measurements per unit area was not standardized but probably low.
 
-Soil science needs more studies of relationships between peat depth and topographic or radiometric data to determine how predictive they are, which variables are most predictive, and how consistent their associations.
-The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify rich peatland carbon stocks and avoid their conversion.
-Although small areas can be mapped accurately on the ground, having landscape-scale maps \emph{before} detailed investigation increases the option space for spatial planning and land management.
+The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify peatlands with rich carbon stocks to avoid their conversion and prioritize their preservation or restoration.
+Although ground surveys of peat depth are accurate and feasible across small areas, having landscape-scale maps \emph{before} detailed investigation increases the option space for spatial planning and land management.
 For example, the Norwegian Public Roads Administration routinely measures peat depth during geotechnical work, but by that time, the route of the road is already set.
 With a digital peat depth map, planners could better compare routes and climate mitigation measures would have more leverage.
+Therefore, soil science needs more studies of relationships between peat depth and topographic or radiometric data to determine how predictive they are, which variables are most predictive, and how consistent their associations are.
 
-We assess how well remotely sensed topographic and radiometric data can predict peat depth at the landscape scale, with a view towards revising regional and national maps in Norway.
+Here, we assess how well remotely sensed topographic and radiometric data can predict peat depth at the landscape scale, with a view towards revising regional and national maps in Norway.
 Specifically, we: (1) quantify the accuracy of predictions from topographic and radiometric variables, (2) identify key predictive variables, and (3) compare these relationships across two different Norwegian landscapes.
 To our knowledge, this is the first study to predict peat depth from airborne radiometric data using machine learning algorithms.
-Where airborne radiometric data have been used to predict peat depth, it has been through modeling techniques less suited for prediction and spatial extrapolation \citep[e.g.,][]{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, siemonAirborneElectromagneticRadiometric2020}.
-Where machine learning algorithms have been used on airborne radiometric data, it has been to predict peat extent rather than depth \citep[e.g.,][]{olearyDigitalSoilMapping2022}.
+Where airborne radiometric data have previously been used to predict peat depth, it has been through modeling techniques less suited for prediction and spatial extrapolation \citep[e.g.,][]{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, siemonAirborneElectromagneticRadiometric2020}.
+Where machine learning algorithms have previously been used on airborne radiometric data, it has been to predict peat extent rather than depth \citep[e.g.,][]{olearyDigitalSoilMapping2022}.
 
@@
 
-We assessed how well we could predict peat depth at two sites with conspicuously different physical geography: Skrimfjella in eastern Norway and Ørskogfjellet in western Norway (Fig.
+To assess the predictive value of terrain and radiometric data for mapping peat depth, we chose two sites with conspicuously different physical geography: Skrimfjella (Fig.
+\ref{fig:sites}b) in eastern Norway and Ørskogfjellet (Fig.
+\ref{fig:sites}a) in western Norway (Fig.
 \ref{fig:sites}c).
-These sites were chosen because they were covered by radiometric data from airborne surveys, relatively little built-up area, and road access.
+These sites were chosen because they were covered by radiometric data from airborne surveys, had relatively little built-up area, and were accessible by road.
 
 \begin{figure}
-\includegraphics[height=0.9\textheight]{figures/sites-patchwork} \caption{Study areas at Ørskogfjellet (a) and Skrimfjella (b) within southern Norway (c).
-Land cover shown here is from the AR50 national land resource database and has simplified geometry with respect to the AR5 database used in the study.}\label{fig:sites}
+\includegraphics[height=0.82\textheight]{figures/sites-patchwork} \caption{Sites in southern Norway (a) with land cover at Skrimfjella (b) and Ørskogfjellet (c).
+To fit the scale of the maps, the land cover shown here is from the AR50 national land resource dataset, which has simplified geometry with respect to the AR5 dataset that is used in the study.
+Terrain is visualized using a hillshade with slope 45° and azimuth 225°.
+AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (<https://data.norge.no/nlod/en/1.0>) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (<https://creativecommons.org/licenses/by/4.0/>).}\label{fig:sites}
 \end{figure}
@@
 At Skrimfjella we delineated a study area of 34 km\textsuperscript{2} based on radiometric coverage and accessibility (Fig. \ref{fig:sites}b).
-The study area has a diverse bedrock, with 32 \% alkali feltspat granite, 26 \% mergelstein, 10 \% granite, and eight other types with \textgreater{} 1 \% coverage (NGU, 1:250 000 dataset).
-The landscape within our delineation is classified as \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
+The study area has a diverse bedrock, with 32 \% alkali feldspar granite, 26 \% marlstone, 10 \% granite, 8 \% mozonite, 7 \% sandstone, 6 \% limestone, and five other rock types with 1--5 \% coverage (NGU, 1:250 000 dataset).
+The area within our delineation at Skrimfjella is classified as the landscape type \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
 It is almost without human infrastructure, dominated by forest, and borders on a large nature reserve.
-The study area has a mean elevation of 438 m above sea level (range 223--711, IQR 351--509), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).
-In Norway's AR5 national land capability dataset \citep{ahlstromAR5Klassifikasjonssystem2019}, 1.5 km\textsuperscript{2} (4.5 \%) of the study area is classified as mire -- defined as areas with mire vegetation and at least 30 cm of peat depth.
+The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).
+In Norway's AR5 national land cover dataset \citep[``areal resources in scale 1:5000'',][]{ahlstromAR5Klassifikasjonssystem2019}, 1.5 km\textsuperscript{2} (4.5 \%) of the study area is classified as `mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.
 
-At Ørskogfjellet we defined a study area of 124 km\textsuperscript{2} which basically followed the footprint of the radiometric survey (Fig. \ref{fig:sites}a).
-According to the Geological Survey of Norway, bedrock in the area is 84 \% granitic gneiss, 11 \% granite, and 5 \% aluminium silicate gneiss (NGU, 1:250 000 dataset).
+At Ørskogfjellet we defined a study area of 124 km\textsuperscript{2} which basically followed the southernmost part of the radiometric survey extent (Fig. \ref{fig:sites}a).
+Bedrock in the area is 84 \% granitic gneiss, 11 \% granite, and 5 \% aluminium silicate gneiss (NGU, 1:250 000 dataset).
 This study area comprises a wide range of major landscape types: \emph{coastal plains}, \emph{coastal fjord}, \emph{inland valleys}, as well as \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
 It is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
-Its mean elevation is 211 m above sea level (range 0--807, IQR 73--310), and its mean slope at 10 m resolution is 13.0° (IQR 4.7--18.3°).
-The AR5 dataset counts 15.3 km\textsuperscript{2} (12.4 \%) of the study area as mire.
+Its mean elevation is 211 m above sea level (range 0--807 m, IQR 73--310 m), and its mean slope at 10 m resolution is 13.0° (IQR 4.7--18.3°).
+In the AR5 dataset, 15.3 km\textsuperscript{2} (12.4 \%) of the study area is classified as mire.
 
@@
 We measured peat depth in selected locations (10 m raster cells) at Skrimfjella.
-The locations were chosen only from areas delineated as mire in the AR5 national land capability dataset.
-Within this mire area, we stratified our sample across values of elevation, slope, and potassium ground concentration \citep[from processed airborne gamma ray spectrometry,][]{baranwalHelicopterborneMagneticElectromagnetic2013}.
+The locations were chosen only from areas delineated as mire in the AR5 dataset.
+Within these AR5 mires, we stratified our sample across values of elevation, slope, and potassium ground concentration \citep[from processed airborne gamma ray spectrometry,][]{baranwalHelicopterborneMagneticElectromagnetic2013}.
 Specifically, we used the \emph{eSample} function in the \emph{iSDM} R package (v.1.0) to chose an environmentally systematic sample.
@@
 In addition to the peat depth locations, we had another arm of our sampling design for measuring peatland occurrence, as a binary variable.
-We wanted to measure peatland occurrence outside of mapped mire areas because the AR5 dataset is known to underestimate peatland coverage \citep[especially in forests,][]{brynLandCoverNorway2018}, and because airborne radiometrics may help identify unmapped peatland \citep{gatisMappingUplandPeat2019, olearyDigitalSoilMapping2022}.
-The occurrence locations were sampled from the part of the study area that (1) was mapped as something other than mire in the AR5 database and (2) had a slope \textless{} 20°.
+We wanted to measure peatland occurrence outside of AR5 mires because the AR5 dataset is known to underestimate peatland coverage \citep[especially in forests,][]{brynLandCoverNorway2018}, and because airborne radiometrics may help identify unmapped peatland \citep{gatisMappingUplandPeat2019, olearyDigitalSoilMapping2022}.
+The occurrence locations were sampled from the part of the study area that (1) was mapped as something other than mire in the AR5 dataset and (2) had a slope \textless{} 20°.
 We performed environmentally systematic sampling of this population with the same procedure as for the depth locations, and \emph{eSample} returned 106 raster cells.
@@
 
-Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively chosen, individual mires, using ground-penetrating radar (GPR).
+Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively chosen, separate peatlands, using ground-penetrating radar (GPR).
 We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with its 500 MHz\} shielded antenna mounted in a plastic sledge, and its control unit connected to a GNSS receiver.
-At each of the three mires we recorded GPR traces along walking transects that covered the extent of the mire, mostly in traversing, zigzag patterns with between 5 m and 20 m spacing at their vertices.
+At each of the three peatlands we recorded GPR traces along walking transects that covered the extent of the peatland, mostly in traversing, zigzag patterns with between 5 m and 20 m spacing at their vertices.
 Along the GPR transects we also probed peat depth at marked trace locations, to be able to calibrate the GPR wave speed velocity.
@@
 
-At Ørskogfjellet we also measured peat depth in a sample of 10 m raster cells, selected from the part of the study area classified in the AR5 dataset as mire.
-Before selecting locations, we determined a minimal sample size that would adequately capture the terrain and radiometric properties of the entire mire area.
+At Ørskogfjellet we also measured peat depth in a sample of 10 m raster cells, selected from the part of the study area classified as mire in the AR5 dataset.
+Before selecting locations, we determined a minimal sample size that would adequately capture the terrain and radiometric properties of the entire AR5 mire area.
 Specifically, we aimed to identify the size at which adding locations produced diminishing decreases in divergence between sample and population distributions -- i.e.~the elbow point in a curve of similarity between sample and population \citep{maloneMethodsImproveUtility2019}.
@@
 Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
-To adjust, we first restricted the sampling population to mire areas that were within an arbitrary cost distance of publicly accessible roads.
+To adjust, we first restricted the sampling population to AR5 mire areas that were within an arbitrary cost distance of publicly accessible roads.
 Cost distance was calculated using GRASS's \emph{r.walk} function, with friction costs defined by AR5 land classes \citep{GRASSv8-2}.
@@
 
-We prepared one categorical predictor -- peat depth class -- from a historical national map dataset called \emph{DMK} \citep{ahlstromAR5Klassifikasjonssystem2019}.
-The DMK peat depth classes are: \textless{} 1 m (\emph{shallow}), \textgreater{} 1 m (\emph{deep}), and \emph{unknown}.
-Mappers generally assigned peat depth classes to polygons of at least 0.5 ha, although delineating polygons down to 0.2 ha was allowed if peat depth showed a ``particularly marked difference'' \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
+We prepared one categorical predictor -- peat depth class -- from the national map dataset called \emph{DMK} \citep{ahlstromAR5Klassifikasjonssystem2019}.
+The DMK dataset is derived from the same historical surveys as the AR5 dataset, and peat depth classes are: \textless{} 1 m (\emph{shallow}), \textgreater{} 1 m (\emph{deep}), and \emph{unknown}.
+Surveyors generally assigned peat depth classes to polygons of at least 0.5 ha, although delineating polygons down to 0.2 ha was allowed if peat depth showed a ``particularly marked difference'' \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
 We rasterized the peat depth class attribute to our 10 m grid.
@@
 
-While the primary aim of the DSM was to predict peat depth within mapped peatlands (where peat depth \textgreater{} 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth or occurrence outside of mapped peatland.
-In other words, we tested whether our peatland-trained models of peat depth could indicate which other locations were most likely to have peat.
-It is generally impractical to obtain good data for this purpose, because locations with non-zero peat depth outside of mapped peatland are unknown and relatively rare.
-For Skrimfjella, we had independent data from the arm of our sampling design that measured peatland occurrence outside of mapped mire areas.
+While the primary aim of the DSM was to predict peat depth within AR5 mires (where peat depth is supposed to be \textgreater{} 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth or occurrence outside of AR5 mire areas.
+In other words, we tested whether our mire-trained models of peat depth could identify peatlands (with or without mire vegetation).
+It is difficult to obtain good data for this purpose, because peat soil not covered by mire vegetation frequently goes unidentified and is relatively rare at these sites.
+For Skrimfjella, we had independent data from the arm of our sampling design that measured peatland occurrence outside of AR5 mires.
 For Ørskogfjellet we did not have such data.
-Therefore, we first -- for both sites -- performed an evaluation with just our peat depth measurements, which included a fraction of locations not mapped as peatland.
-For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on locations mapped as peatland (in the training folds), and evaluated them only on locations not mapped as peatland (in the test fold).
-In this way, we simulated the situation where a model trained on known peatland is used to predict peat depth outside the known peatland.
+Therefore, we first -- for both sites -- performed an evaluation with only our peat depth measurements, which included a fraction of locations outside of AR5 mires.
+For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on AR5 mire locations (in the training folds), and evaluated them only on (test fold) locations classified in AR5 as something other than mire.
+In this way, we simulated the situation where a model trained on AR5 mires is used to predict peat depth outside AR5 mires.
 From the cross-validation we quantified mean absolute error.
-Second -- for Skrimfjella -- we simply trained a model on all peat depth measurements and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
+Second -- for Skrimfjella alone -- we simply trained a model on all peat depth measurements (AR5 mire or not) and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
 In this analysis, the model's prediction of peat depth is treated as a rank index of peatland likelihood.
-Since DMK peat depth class is always undefined outside known peatlands, we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.
+Since the purpose was to test the models' ability to uncover unmapped peatland (where DMK peat depth class is undefined), we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.
 
@@
 
-We obtained depth measurements for 372 cells at Skrimfjella (area equivalent to 2.4 \% of mapped peatland) and 1878 cells at Ørskogfjellet (area equivalent to 1.2 \% of mapped peatland).
-Roughly 80 \% of these 10 m cells were within mapped peatland, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
+We obtained depth measurements for 372 cells at Skrimfjella (2.4 \% of AR5 mire area) and 1878 cells at Ørskogfjellet (1.2 \% of AR5 mire area).
+Roughly 80 \% of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
 Coverage of DMK peat depth classes was higher at Ørskogfjellet than at Skrimfjella (79 \% versus 27 \% of cells), and at Ørskogfjellet the \emph{deep} and \emph{shallow} classes showed a larger difference in measured depth.
@@
 
-When we tested how well models extrapolated from within to outside of mapped peatland, we found that the models produced worse mean absolute error than just assuming a constant 30 cm depth (Fig.
+When we tested how well models extrapolated from within to outside of AR5 mire areas, we found that the models produced worse mean absolute error than just assuming a constant 30 cm depth (Fig.
 \ref{fig:modelMetricsExtrapolation}).
@@
 \includegraphics{figures/modelmetrics-extrapolation.pdf}
-\caption{\label{fig:modelMetricsExtrapolation}Performance of models that extrapolate from training data inside of mapped peatland to test data outside of mapped peatland, evaluated via spatial cross-validation.
+\caption{\label{fig:modelMetricsExtrapolation}Performance of models that extrapolate from training data inside of mapped mires to test data outside of mapped mires (AR5 dataset), evaluated via spatial cross-validation.
 Parentheses denote the number of variables in each model, and point estimates are shown +/- their standard error.}
@@
 The performance of our terrain- and radiometric-based maps could have been improved with a spatially explicit mapping approach like regression kriging.
-By ignoring spatial autocorrelation in peat depth (i.e.~the n in scorpan), we have not extracted all of the information about peat depth in the study area out of the training data.
+By ignoring spatial autocorrelation in peat depth (i.e.~the n in \emph{scorpan}), we have not extracted all of the information about peat depth in the study area out of the training data.
 If we were to put our DSM approach into production for published maps, we would harness the spatial component, but the actual maps for Ørskogfjellet and Skrimfjella are not of primary interest in this study.
@@
 
-That we measured more than 30 cm of peat in areas not mapped as peatland was expected, because AR5 underrepresents peatland extent by about one third \citep{brynLandCoverNorway2018}.
+That we measured more than 30 cm of peat in areas not mapped as mire was expected, because AR5 underrepresents peatland extent by about one third \citep{brynLandCoverNorway2018}.
 For that reason, maps of peatland occurrence also need revising.
-Unfortunately, our models did not predict peat occurrence outside mapped peatlands.
+Unfortunately, our models did not predict peat occurrence outside AR5 mires.
 In these areas, assuming a constant 30 cm of peat (the depth threshold used in AR5) produced less error than our models.
-Moreover, the performance gap would likely have been larger had our evaluation data been representative of the non-peatland areas.
-Instead they were near mapped peatlands.
+Moreover, the performance gap would likely have been larger had our evaluation data been representative of all areas not classified as mire in AR5.
+Instead they were concentrated near AR5 mires.
 Most importantly, the independent occurrence data at Skrimfjella showed the model completely failing to predict occurrence, and these data were suited for testing this ability.
@@
 Why wasn't peat occurrence predicted well?
-The models were trained on fundamentally different populations of locations (\textasciitilde80 \% mapped peatland in training vs.~0 \% in prediction), so it is not surprising that the associations they learned did not transfer well.
-Moreover, the \textasciitilde20 \% of training data from outside mapped peatland were incidentally collected (near peatland) and not representative of other land cover classes.
+The models were trained on fundamentally different populations of locations (\textasciitilde80 \% AR5 mire in training vs.~0 \% in prediction), so it is not surprising that the associations they learned did not transfer well.
+Moreover, the \textasciitilde20 \% of training data from outside AR5 mires were incidentally collected (near AR5 mires) and not representative of other land cover classes.
 In short, the models were blind to the fact that most of both study areas have no peat.
@@
 However, it is unclear how big (N) any cluster of depths should be, and the tradeoff between multiple small-extent DSM and fewer large-extent DSM needs research.
-Bohn and Miller \citeyearpar{bohnLocallyEnhancedDigital2024} advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions, where the composition of peatland types is constant \citep[e.g.][]{moenVegetasjonstyper1998}.
+Bohn and Miller \citeyearpar{bohnLocallyEnhancedDigital2024} advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or `supertopes'), where the composition of mesotopes is similar \citep{moenVegetasjonstyper1998, joostenWiseUseMires2002}.
 Depth varies strongly between peats formed by terrestrialization versus paludification \citep{buffamFillingHolesRegional2010}, and probably along other axes of peatland typology.
-Although most Norwegian landscapes contain multiple, unknown types, it makes sense to limit their diversity within a single DSM.
+Although most Norwegian landscapes contain multiple, unknown mesotopes, it makes sense to limit their diversity within a single DSM.
 
@@
 We caution against reducing continuous depth predictions to arbitrary classes \citep[as in][]{ivanovsModelingGeospatialDistribution2024, karjalainenComparisonTwoGammaray2024}, since classes can be derived from continuous predictions.
-The distribution of peat depths across landscapes is zero-inflated, and research is needed to determine whether it is more efficient to parameterize a single model of peat depth (with a larger, generalized dataset) or to break down the problem by classifying zero depth and then regressing non-zero depths (with smaller, specialized datasets).
+The distribution of peat depths across full landscapes is zero-inflated, and research is needed to determine whether it is more efficient to parameterize a single model of peat depth (with a larger, generalized dataset) or to break down the problem by classifying zero depth and then regressing non-zero depths (with smaller, specialized datasets).
 Coupling occurrence and depth will reduce the prevalence of incoherence that we found: deep peat outside the peatland extent and zero depth inside it.

### 2025-01-07 ms/figures.R

@@ -2,12 +2,12 @@
 
 library(sf)
 library(terra)
-library(tidyterra)
 library(tidyverse)
 library(patchwork)
 library(ggtext)
 library(ggrepel)
 library(ggspatial)
+library(ggnewscale)
 library(rnaturalearth)
 
 ## Inset ####
@@ -47,11 +47,12 @@ g1 <- ggplot() +
     nudge_y = c(-8e4, 12e4),
     nudge_x = c(13e4, 4e4),
     segment.curvature = 0.5) +
-  coord_sf(xlim = c(-7e4, 3.7e5), ylim = c(6.47e6, 7.05e6), crs = 25833) +
-  theme(legend.position = "none",
-        axis.title.x = element_blank(), axis.title.y= element_blank(),
-        axis.ticks = element_blank(), axis.text = element_blank(),
-        plot.margin = margin(1, 1, 1, 1, "pt"))
+  coord_sf(xlim = c(-6.1e4, 3.7e5), ylim = c(6.47e6, 7.05e6), crs = 25833) +
+  theme(
+    legend.position = "none",
+    axis.title.x = element_blank(), axis.title.y= element_blank(),
+    axis.ticks = element_blank(), axis.text = element_blank()
+  )
 
 ## AR50 Ørskogfjellet ####
 
@@ -69,39 +70,50 @@ dtm <- rast("ms/figures/data/940695_dtm50/data/dtm50_6900_50m_33.tif")
 slope <- terrain(dtm, "slope", unit = "radians")
 aspect <- terrain(dtm, "aspect", unit = "radians")
 hill <- shade(slope, aspect, 45, 225)
-hill <- mask(hill, ar50orskog)
-names(hill) <- "shades"
-pal_greys <- hcl.colors(1000, "Grays")
-pal_greys_index <- hill %>%
-  mutate(index_col = scales::rescale(shades, to = c(1, length(pal_greys)))) %>%
-  mutate(index_col = round(index_col)) %>%
-  pull(index_col)
+ar50orskogmask <- ar50orskog |> 
+  group_by(TRUE) %>%
+  summarize()
+hill <- mask(hill, ar50orskogmask)
+hilldf_single <- as.data.frame(hill, xy = TRUE)
 
 # https://colorbrewer2.org/?type=qualitative&scheme=Paired&n=6
[…]

### 2025-01-16 ms/ms.Rmd
addressing comments

@@
 We also thank Vikas Baranwal from the Geological Survey of Norway for helping us access the radiometric data from Skrim.
-This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, © Kartverket, (2) *Norge digitalt* license, Norwegian Institute of Bioeconomy Research (NIBIO), © Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway (NGU).
+This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, © Kartverket, (2) *Norge digitalt* license, Norwegian Institute of Bioeconomy Research (NIBIO), © Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway.
 Large language models have been used during the drafting and editing of this manuscript, with author oversight.
 We maintain full responsibility for the scientific output, as per the European Commission's *Living guidelines on the responsible use of generative AI in research*.
+appendix: |
+  \appendixtables
+  \begin{table}[tbp]
+  \caption{Attributes of the radiometric surveys, as reported in Baranwal et al., \citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013} and Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.}
+  \begin{tabular}{lll}
+  \hline
+                                           & Skrimfjella & Ørskogfjellet              \\ \cline{2-3} 
+  Survey period                            & 2008-2011   & December 2014–January 2015 \\
+  Average flight altitude (m)              & 75          & 80                         \\
+  Average flight speed (\unit{km\,h^{-1}}) & 108         & 88                         \\
+  Flight line spacing (m)                  & 200         & 200                        \\ \hline
+  \end{tabular}
+  \label{tab:radSurveys}
+  \end{table}
+  \clearpage
+  \appendixfigures
+  \begin{figure}
+  \centering
+  \includegraphics[width=0.8\textwidth]{figures/GPRwavevelocity.pdf}
+  \caption{\label{fig:GPRwavevelocity}Calibration of GPR wave velocity at Skrimfjella (a) and Ørskogfjellet (b) by regression of probed depth against wave travel time.}
+  \end{figure}
+  \clearpage
+  \appendixfigures
+  \begin{figure}
+  \centering
+  \includegraphics{figures/semivariograms.pdf}
+  \caption{\label{fig:semivariograms}Empirical semivariograms of peat depth point measurements at Skrimfjella (a) and Ørskogfjellet (b).}
+  \end{figure}
 output:
@@
 @holdenEstimatingCarbonStock2011; @parryMethodModellingPeat2012; @rudiyantoDigitalMappingCosteffective2016; @rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024].
-Also more complex derivations of topography, such as the Topographic Wetness Index (TWI) and the Multi-Resolution Valley Bottom Flatness (MRVBF) index, have shown associations with peat depth in some studies [e.g.
+Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth in some studies [e.g.
 @rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024].
@@
 
-```{r sites, out.height="82%", echo = FALSE, fig.cap = "Sites in southern Norway (a) with land cover at Skrimfjella (b) and Ørskogfjellet (c).
+```{r sites, out.height="81%", echo = FALSE, fig.cap = "Sites in southern Norway (a) with land cover at Skrimfjella (b) and Ørskogfjellet (c).
 To fit the scale of the maps, the land cover shown here is from the AR50 national land resource dataset, which has simplified geometry with respect to the AR5 dataset that is used in the study.
@@
 At Skrimfjella we delineated a study area of 34 km^2^ based on radiometric coverage and accessibility (Fig. \@ref(fig:sites)b).
-The study area has a diverse bedrock, with 32 % alkali feldspar granite, 26 % marlstone, 10 % granite, 8 % mozonite, 7 % sandstone, 6 % limestone, and five other rock types with 1--5 % coverage (NGU, 1:250 000 dataset).
+The study area has a diverse bedrock, with 32 % alkali feldspar granite, 26 % marlstone, 10 % granite, 8 % mozonite, 7 % sandstone, 6 % limestone, and five other rock types with 1--5 % coverage (Geological Survey of Norway, 1:250 000 dataset).
 The area within our delineation at Skrimfjella is classified as the landscape type *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
@@
 At Ørskogfjellet we defined a study area of 124 km^2^ which basically followed the southernmost part of the radiometric survey extent (Fig. \@ref(fig:sites)a).
-Bedrock in the area is 84 % granitic gneiss,  11 % granite, and 5 % aluminium silicate gneiss (NGU, 1:250 000 dataset). 
+Bedrock in the area is 84 % granitic gneiss,  11 % granite, and 5 % aluminium silicate gneiss (Geological Survey of Norway, 1:250 000 dataset). 
 This study area comprises a wide range of major landscape types: *coastal plains*, *coastal fjord*, *inland valleys*, as well as *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
@@
 
-## Peat depth measurements
-
-At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest (RF) model of peat depth, and we designed our sampling with this in mind [@brusSamplingDigitalSoil2019].
-Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
-A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones [@brusSamplingDigitalSoil2019].
-We chose for our sampling and modelling a spatial resolution of 10 m.
-We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other.
-The point measurements of peat depth described below were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
-
-### Skrimfjella
-
-We measured peat depth in selected locations (10 m raster cells) at Skrimfjella.
-The locations were chosen only from areas delineated as mire in the AR5 dataset.
-Within these AR5 mires, we stratified our sample across values of elevation, slope, and potassium ground concentration [from processed airborne gamma ray spectrometry, @baranwalHelicopterborneMagneticElectromagnetic2013].
-Specifically, we used the *eSample* function in the *iSDM* R package (v.1.0) to chose an environmentally systematic sample.
-This function defines the environmental space as a two-dimensional convex hull around the ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest [@hattabUnifiedFrameworkModel2017].
-Elevation was extracted from the 10 m national DTM, slope calculated in degrees, and potassium ground concentration downscaled with bilinear resampling.
-We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and with these parameters *eSample* returned 105 raster cells.
-
-In addition to the peat depth locations, we had another arm of our sampling design for measuring peatland occurrence, as a binary variable.
-We wanted to measure peatland occurrence outside of AR5 mires because the AR5 dataset is known to underestimate peatland coverage [especially in forests, @brynLandCoverNorway2018], and because airborne radiometrics may help identify unmapped peatland [@gatisMappingUplandPeat2019; @olearyDigitalSoilMapping2022].
-The occurrence locations were sampled from the part of the study area that (1) was mapped as something other than mire in the AR5 dataset and (2) had a slope < 20°. 
-We performed environmentally systematic sampling of this population with the same procedure as for the depth locations, and *eSample* returned 106 raster cells.
-
-Field work at Skrimfjella was conducted in August 2020.
-We navigated to the centers of the raster cells in the depth and occurrence samples with a handheld global navigation satellite system (GNSS) receiver, checking that positional error was below 3 m.
-For each depth sample location, we measured peat depth three times (at the vertices of a triangle with 2 m sides) to get a more representative value for the 10 m raster cell, and to dampen the effect of outlying measurements [@parryEvaluatingApproachesEstimating2014].
-We used a metal probe pushed downward until resistance indicated the base of the peat column.
-Probe locations were adjusted up to 20 cm if the base of the peat column seemed to be blocked by an obvious artifact.
-A single probe measurement was right-censored because the peat column was deeper than the probe.
-For each occurrence sample location, we recorded the presence or absence of peatland -- primarily by digging and examining the top 20 cm of soil (where this was possible).
-We judged whether the soil was a peat soil based on its density, texture, and color.
-Occasionally, when the soil itself was difficult to judge, we made our determination also based on the presence or absence of mire vegetation.
-Although peat soil is strictly defined by organic content (which we did not analyze), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland [@minasnyMappingMonitoringPeatland2023].
-
-Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively chosen, separate peatlands, using ground-penetrating radar (GPR).
-We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with its 500 MHz} shielded antenna mounted in a plastic sledge, and its control unit connected to a GNSS receiver.
-At each of the three peatlands we recorded GPR traces along walking transects that covered the extent of the peatland, mostly in traversing, zigzag patterns with between 5 m and 20 m spacing at their vertices.
-Along the GPR transects we also probed peat depth at marked trace locations, to be able to calibrate the GPR wave speed velocity.
-We processed the GPR data with Reflex2DQuick software (v.3.0, Sandmeier Scientific Software, Germany), applying a time-zero correction, a dewow filter, and a gain filter based on observed energy decay.
-Then we picked strong reflectors in the radargrams that we interpreted as the base of the peat column.
-We used picks at marked trace locations to calibrate wave speed velocity; we pooled calibration points across the three peatlands and fitted a linear regression of depth on one-way travel time with the intercept fixed at zero.
-In total we had 46 calibration points along 3.5 km of GPR transects.
-Finally, we used the calculated wave velocity (\unit{0.0387\,m\,ns^{-1}}, $R^2 = 0.874$) to convert the travel times of all picks to calibrated peat depths.
-
-### Ørskogfjellet
-
-At Ørskogfjellet we also measured peat depth in a sample of 10 m raster cells, selected from the part of the study area classified as mire in the AR5 dataset.
-Before selecting locations, we determined a minimal sample size that would adequately capture the terrain and radiometric properties of the entire AR5 mire area.
-Specifically, we aimed to identify the size at which adding locations produced diminishing decreases in divergence between sample and population distributions -- i.e.
-the elbow point in a curve of similarity between sample and population [@maloneMethodsImproveUtility2019].
-This approach has been found to identify sample sizes that correspond with diminishing returns in predictive model performance on external evaluation data [@sauretteDivergenceMetricsDetermining2023].
-We defined a sequence of sample sizes (50--500) and for ten replicate samples at each size [drawn by conditioned latin hypercube sampling, @minasnyConditionedLatinHypercube2006; @roudierClhsPackageConditioned2011], we calculated the mean Kullback--Leibler divergence between sample and population distributions [@maloneMethodsImproveUtility2019; @sauretteDivergenceMetricsDetermining2023].
-The predictors in the divergence calculation were terrain slope and four radiometrics: potassium, thorium, uranium, and total count.
-Next, we fitted a asymptotic regression of mean divergence on sample size, and identified the sample size at which the curve reached 95 % of the fitted asymptote.
-Through this procedure we found that we could adequately capture the population distribution with a sample of 160 locations.
-
-To choose 160 locations, we performed feature space coverage sampling.
-This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
-Feature space coverage sampling aims to disperse samples as uniformly as possible in multidimensional predictor space, and is implemented by choosing locations that are closest to cluster centers in a k-means clustering of the standardized predictor space [@brusSamplingDigitalSoil2019].
-Feature space coverage sampling works best when all dimensions are important predictors of the outcome [@wadouxSamplingDesignOptimization2019], and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
-The radiometric predictors were downscaled to 10 m resolution with cubic B-spline resampling in QGIS [v.3, @QGISsoftware].
-We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
-Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
-To adjust, we first restricted the sampling population to AR5 mire areas that were within an arbitrary cost distance of publicly accessible roads.
-Cost distance was calculated using GRASS's *r.walk* function, with friction costs defined by AR5 land classes [@GRASSv8-2].
-After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
-Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 % (with respect to the ideal sample), but distance in our sample was still only 46 % of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
-
-Field work at Ørskogfjellet was conducted in August 2023.
-We navigated to the centers of the raster cells in the sample using real time kinematic differential GNSS (Topcon Positioning Systems, USA), to ensure sub-meter positional accuracy.
-At each location we measured peat depth three times by manual probing, with probe locations spaced approximately 2.5 m apart.
-For a tiny fraction of these measurements (5 total), it was not possible to reach the base of the peat column, and a right-censored depth was recorded. 
-In areas with dense sampling locations, we also measured peat depth with GPR along snaking transects passing through the centers of the sampling cells (seven transects, 6.2 km total length).
-We used the same GPR system as at Skrimfjella, but with a 100 MHz Malå rough terrain antenna (Guideline Geo AB, Sweden) at some transects.
-To navigate the GPR transects, we placed flags at the cell centers of the sample locations, and used a handheld GNSS receiver to guide the GPR operator.
-At sampling locations crossed by a GPR transect, we arranged the manual probe positions along the transect (for better calibration of the GPR wave speed velocity), while other locations were probed in a triangular pattern around the cell center like at Skrimfjella.
-
-We processed the GPR data with Reflexw software (v.8.5, Sandmeier Scientific Software, Germany), applying a dewow filter, time-zero correction, bandpass filter, gain filter, and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennae.
-The last correction is important for the rough terrain antenna, which has an antenna separation (2.2 m) -- comparable to typical peat depths.
-As with the data from Skrimfjella, we picked the base of the peat column from strong reflectors in the radargram, and calibrated wave velocity with manual probe measurements in a linear regression.
-The points in the regression were created by joining to each probe measurement the travel time of the nearest pick, but only if these were within 2 m of each other.
-In total we had 78 calibration points along 7.8 km of interpretable GPR traces (transect length exceeded because of extra GPR data).
-Finally, we used the calculated wave velocity (\unit{0.0427\,m\,ns^{-1}}, $R^2 = 0.946$) to convert the travel times of all picks to calibrated peat depths.
-
-We also used two sets of existing depth measurements from Ørskogfjellet.
-The first set was provided by the Norwegian Public Roads Administration, who commissioned GPR surveys of particular peatland areas in 2020 and 2021.
-The surveys were conducted with a dual channel system (70 MHz and 300 MHz; ImpulseRadar AB, Sweden), connected to GNSS with CPOS correction.
-We used interpreted and calibrated traces from these surveys, and discarded some data where multiple depths were interpreted for the same locations.
-This summed to 7.4 km of interpreted traces.
-The second set of existing depth data we extracted from a paper map made by the Norwegian Soil and Mire Company in 1984.
-This map presents 44 borehole depths (in decimeters) across a 9 ha peatland area.
-We georeferenced the map and digitized the borehole locations and depths.
-
 ## Peat depth predictors
@@
 All continuous predictors were derived either from an airborne radiometric survey or from a DTM.
-From the radiometric surveys we simply used the four variables produced by the surveyors (Geological Survey of Norway): ground concentration of Potassium, Thorium, Uranium, as well as total count.
+From the radiometric surveys we simply used the four variables produced by the Geological Survey of Norway: ground concentration of Potassium, Thorium, Uranium, as well as total count.
 From the DTMs we calculated several land surface parameters, ranging from simple terrain indices to more complex geomorphometric and hydrological variables [@maxwellLandsurfaceParametersSpatial2022].
 The categorical predictor was peat depth class, from a national map dataset.
+We chose a spatial resolution of 10 m for our predictors, peat depth sample, and modelling.
+We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other.
 Complete descriptions of predictors follow below.
+
 ```{r preds, echo=FALSE, message=FALSE}
@@
 
-The radiometric survey covering Skrimfjella was conducted in 2008--2011. 
-The survey was flown at an average altitude of 75 m and average speed of \unit{108\,km\,h^{-1}}, with flight lines spaced 200 m apart.
-Spectrometer count rates were calibrated annually to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
-The Geological Survey of Norway processed data from the spectrometer following standard procedures outlined by the International Atomic Energy Association, and the processing included: correction for aircraft and cosmic background radiation, correction for radon in the air, window stripping of the gamma ray spectrum, correction for flying height, conversion of count rates to ground concentrations, and finally gridding to 50 m resolution with micro-leveling.
-Further details about the survey and data processing are provided in Baranwal et al. [-@baranwalHelicopterborneMagneticElectromagnetic2013].
-We downscaled the processed data to 10 m resolution by cubic spline resampling, using the *terra* package in R.
+The Geological Survey of Norway conducted and processed radiometric surveys over our study areas, as reported in Baranwal et al.
+[-@baranwalHelicopterborneMagneticElectromagnetic2013] and Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015].
+They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling, using the *terra* package in R.
 
-A very similar radiometric survey covering Ørskogfjellet was conducted in December 2014 and January 2015. 
-This survey was flown at an average altitude of 80 m and average speed of \unit{88\,km\,h^{-1}}, with flight lines also spaced 200 m apart.
-Spectrometer count rates were calibrated in 2013 to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
-The spectrometer data were processed following the same procedure as for the survey at Skrimfjella, except that a convolution filter was added to smooth the gridded data.
-Further details about the survey and data processing are provided in Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015].
-The 10 m resolution predictors from this survey were identical to the layers used in the sampling design (resampled from 50 m resolution with cubic B-splines).
+The radiometric surveys were conduced with similar flight parameters (Table \ref{tab:radSurveys}).
+Spectrometer count rates were calibrated to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
+Data were processed following standard procedures outlined by the International Atomic Energy Association.
+Processing included: correction for aircraft and cosmic background radiation, correction for radon in the air, window stripping of the gamma ray spectrum, correction for flying height, conversion of count rates to ground concentrations, and finally gridding to 50 m resolution with micro-leveling.
+At Ørskogfjellet an additional convolution filter was added to smooth the gridded data.
 
@@
 
-For terrain-derived predictors, we obtained 1 m resolution DTMs from the Norwegian Mapping Authority.
+For terrain-derived predictors, we obtained 1 m resolution rasters from the national DTM ([CC BY 4.0](https://creativecommons.org/licenses/by/4.0/) Norwegian Mapping Authority).
 The DTM for Skrimfjella was produced from airborne laser scanning surveys in 2015 and 2022, with laser point density of \unit{5\,pts\,m^{-2}}.
@@
 
-We used the *terra* R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
-These were derived at two scales to produce eight different predictors; we either calculated from 1 m DTM resolution and then aggregated, or aggregated to 10 m DTM resolution and then calculated the indices.
+We used the *terra* R package to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
+These were derived at two scales to produce eight different predictors; we either calculated the indices at 1 m DTM resolution and then aggregated to 10 m resolution, or aggregated to 10 m DTM resolution and then calculated the indices.
 This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties [@millerImpactMultiscalePredictor2015; @dornikOptimalScalingPredictors2022; @newmanAssessingSpatiallyHeterogeneous2023].
 We know that peat depth tends to vary at fine scales in Norway, which is why we chose 1 m and 10 m resolutions [@maxwellLandsurfaceParametersSpatial2022].
-We also calculated the MRVBF index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm [@gallantMultiresolutionIndexValley2003].
-We calculated this index in SAGA GIS [v.9.3.2, @conradSystemAutomatedGeoscientific2015] with default parameters.
+We also calculated the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm [@gallantMultiresolutionIndexValley2003].
+We calculated this index in SAGA GIS [v.9.3.2, Morphometry library, @conradSystemAutomatedGeoscientific2015] with default parameters (initial slope threshold = 16 %, lowness threshold = 0.4, upness threshold = 0.35, slope shape parameter = 4, elevation shape parameter = 3).
 
-Next, we calculated the Topographic Wetness Index [@quinnPredictionHillslopeFlow1991].
-This index is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate to coarse resolution DTMs [@agrenEvaluatingDigitalTerrain2014; @riihimakiTopographicWetnessIndex2021], so we calculated it from 5 m, 10 m, 20 m, and 50 m DTM resolution.
+The Topographic Wetness Index [@quinnPredictionHillslopeFlow1991] is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate to coarse resolution DTMs [@agrenEvaluatingDigitalTerrain2014; @riihimakiTopographicWetnessIndex2021], so we calculated it from 5 m, 10 m, 20 m, and 50 m DTM resolution.
 The calculations were performed with Whitebox software [@lindsayWhiteboxGATCase2016], accessed through the *whitebox* R package [v2.4, @wuWhiteboxWhiteboxToolsFrontend2022].
@@
 
-The last terrain-based predictor we included was the depth-to-water (DTW) index [@murphyMappingWetlandsComparison2007].
-This index approximates a location's vertical height above the surface water feature that it is likely to drain towards.
+The Depth-to-Water index [@murphyMappingWetlandsComparison2007] approximates a location's vertical height above the surface water feature that it is likely to drain towards.
 It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature [eq. 5 in @murphyTopographicModellingSoil2009].
@@
 Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell [@schonauerSpatiotemporalPredictionSoil2021; @schonauerRcodeCalculatingDepthwater2021].
-This catchments area layer was then thresholded at seven different levels (*flow initiation areas* 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry [@murphyModellingMappingTopographic2011; @agrenEvaluatingDigitalTerrain2014; @schonauerSpatiotemporalPredictionSoil2021].
+This catchments area layer was then thresholded at seven different levels (flow initiation area 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry [@murphyModellingMappingTopographic2011; @agrenEvaluatingDigitalTerrain2014; @schonauerSpatiotemporalPredictionSoil2021].
 In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
-For each of the seven surface water layers, we derived DTW using the *Distance Accumulation* tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.
+For each of the seven surface water layers, we derived Depth-to-Water using the *Distance Accumulation* tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.
 
@@
 
+## Peat depth sample selection
+
+The places (10 m raster cells) we chose to measure peat depth were sampled from mire areas in the AR5 dataset, and optimized for training a Random Forest (RF) model of peat depth [@brusSamplingDigitalSoil2019].
+Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
+A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population, while avoiding spurious ones [@brusSamplingDigitalSoil2019].
+Although we implemented the approach differently for Skrimfjella (in 2020) than for Ørskogfjellet (in 2023), the objective was the same.
+
+We made several point measurements of peat depth within the sampled raster cells (three if by probing only, more if by GPR as well).
+The point measurements (described below) were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
+
+### Skrimfjella
+
+At Skrimfjella, we used the *eSample* function in the *iSDM* R package (v.1.0) to stratify our sample across elevation, slope, and potassium concentration.
+This function defines the environmental space as a two-dimensional convex hull around the ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest [@hattabUnifiedFrameworkModel2017].
+We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and *eSample* returned 105 raster cells.
+
+We also measured peatland occurrence (binary variable) in a separate sample at Skrimfjella.
+We measured peatland occurrence because the AR5 dataset is known to underestimate peatland coverage [especially in forests, @brynLandCoverNorway2018], and because airborne radiometrics may help identify unmapped peatland [@gatisMappingUplandPeat2019; @olearyDigitalSoilMapping2022].
+The 10 m cells were sampled from those outside AR5 mire and with slope < 20°. 
+Following the same procedure, *eSample* returned 106 raster cells from this population.
+
+### Ørskogfjellet
+
+At Ørskogfjellet, we first determined a minimal sample size that would adequately capture the slope and radiometric properties (potassium, thorium, uranium, and total count) of the entire AR5 mire area [@sauretteDivergenceMetricsDetermining2023].
+Specifically, we identified an elbow point in a curve of similarity between sample and population [@maloneMethodsImproveUtility2019].
+For a sequence of sample sizes (50--500) [ten replicates each, drawn by conditioned latin hypercube sampling, @minasnyConditionedLatinHypercube2006; @roudierClhsPackageConditioned2011], we calculated the mean Kullback--Leibler divergence between sample and population distributions [@maloneMethodsImproveUtility2019; @sauretteDivergenceMetricsDetermining2023].
+Then, we fitted a asymptotic regression of mean divergence on sample size, and found that the curve reached 95 % of the fitted asymptote at a sample size of 160.
+
+To choose 160 locations, we performed feature space coverage sampling, implemented using the *kmeans* function in base R and the *rdist* function in the the *fields* package (v.14.2).
+Feature space coverage sampling chooses locations that are closest to cluster centers in standardized predictor space [@brusSamplingDigitalSoil2019].
+This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
+Feature space coverage sampling works best when all dimensions are important predictors of the outcome [@wadouxSamplingDesignOptimization2019], and we used the same five predictors that we used to choose sample size: slope and four radiometrics.
+
+We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
+Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
+To adjust, we first restricted the sampling population to AR5 mire areas that were within an arbitrary cost distance of publicly accessible roads.
+Cost distance was calculated using GRASS's *r.walk* function, with friction costs defined by AR5 land classes [@GRASSv8-2].
+After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
+Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 % (with respect to the ideal sample), but distance in our sample was still only 46 % of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
+
+## Depth measurements
+
+We measured peat depths at Skrimfjella in August 2020 and at Ørskogfjellet in August 2023.
+At both sites we used manual probing as the primary method of measuring peat depth, and ground-penetrating radar (GPR) as a secondary method.
+That is: peat depth was always measured by probing, and also by GPR in a subset of cells that partly overlapped with the probed cells
+We chose this approach because probing is a fast and reliable method for point measurements, while GPR can provide higher lateral density of data in the same amount of time [@parryEvaluatingApproachesEstimating2014].
+Probed depths serve to calibrate GPR measurements when calibration by common midpoint survey is not possible -- as was the case with our fixed radar antennas -- so the methods are complementary.
+
+### Peat probing
+
+We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real time kinematic (Ørskogfjellet) global navigation satellite system receivers.
+We dampened the effect of outlying measurements by probing three times at each location [@parryEvaluatingApproachesEstimating2014], at the vertices of a centered triangle with 2 m (Skrimfjella) or 4.5 m (Ørskogfjellet) sides.
+We used changes in resistance to indicate the base of the peat column.
+Probe locations were adjusted up to 20 cm if the base of the peat column seemed to be blocked by an obvious artifact, like a buried rock.
+Where the peat column was deeper than the extendable probe could be manually inserted and extracted by a pair of operators, we recorded a right-censored result (one at Skrimfjella, five at Ørskogfjellet).
+
+We also extracted a set of existing depth probe measurements for Ørskogfjellet from a paper map made by the Norwegian Soil and Mire Company in 1984.
+The map presents 44 borehole depths (in decimeters) across a 9 ha peatland area.
+We georeferenced the map and digitized the borehole locations and depths.
+
+For our sample of peatland occurrence at Skrimfjella, we recorded the presence or absence of peatland at the cell center -- primarily by digging and examining the top 20 cm of soil.
+We judged whether the soil was a peat soil based on its density, texture, and color, as well as the presence or absence of mire vegetation.
+Although peat soil is strictly defined by organic content (which we did not analyze), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland [@minasnyMappingMonitoringPeatland2023].
+
+### Ground-penetrating radar
+
+We performed GPR surveys in three subjectively chosen mires at Skrimfjella and in areas with a high density of sampled raster cells at Ørskogfjellet. 
+We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with a GNSS-enabled control unit connected to a 500 MHz shielded antenna mounted in a plastic sledge (transmitter--receiver separation 0.18 m, trace frequency 10 Hz).
+For some transects at Ørskogfjellet we substituted a 100 MHz Malå rough terrain antenna (transmitter--receiver separation 2.2 m, trace frequency 5 Hz), because the lower frequency antenna gives greater penetration depth.
+In all cases the system was towed by a walking GPR operator.
+
+GPR traces were recorded along zigzag (Skrimfjella) or snaking (Ørskogfjellet) transects.
+At Ørskogfjellet, transects were predetermined to pass through the centers of sampled raster cells, and we marked these precisely with flags to guide the GPR operator.
+A GPR records the time taken for a radio wave to travel from the transmitter to a reflector and back to the receiver, and the velocity of the wave varies with properties of the peat column.
+Therefore, wave velocity has to be calibrated to convert travel time to peat depth, and we probed peat depth along the transects.
+
+We processed the GPR data with Reflex2DQuick (v.3.0; Skrimfjella) or REFLEXW (v.8.5; Ørskogfjellet) software (Sandmeier Scientific Software, Germany). 
+We applied a time-zero correction, a dewow filter, and a gain filter based on observed energy decay. 
+With Ørskogfjellet data, we also applied a bandpass filter and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennas.
+The latter is important for the rough terrain antenna, where the antenna separation is comparable to typical peat depths.
+From the processed radargrams, we picked travel times from strong reflectors that we interpreted as the base of the peat column.
+
+We used picks near probed depths to calibrate wave speed velocity -- separately for each site.
+Calibration data were created by matching marked trace locations to a corresponding depth probe (Skrimfjella), or by a spatial join that identified interpreted traces and depth probes within 2 m of each other (Ørskogfjellet).
+We had sufficient calibration point density to avoid bias in wave velocity as a major source of error [@rosaDeterminingNumberManual2009]: 46 calibration points along 3.5 km of interpretable traces at Skrimfjella, and 78 along 7.8 km at Ørskogfjellet.
+We fitted site-specific linear regressions of probed depth on picked travel time, with the intercept fixed at zero, to estimate wave velocities.
+Notwithstanding a few outlying points, our regressions showed good fits and the resulting velocities are within the range reported for peat [@parsekianUncertaintyPeatVolume2012]: \unit{0.0387\,m\,ns^{-1}}, $R^2 = 0.874$ at Skrimfjella and \unit{0.0427\,m\,ns^{-1}}, $R^2 = 0.946$ at Ørskogfjellet.
+Finally, we used these two wave velocities to convert the travel times of all picks to calibrated peat depths.
+In total the GPR surveys produced 48579 point measurements of peat depth at Skrimfjella and 32653 at Ørskogfjellet.
+
+We also used a set of existing GPR depth measurements from Ørskogfjellet, commissioned and provided to us by the Norwegian Public Roads Administration.
+These data were collected in 2020 and 2021 with a dual channel system (70 MHz and 300 MHz; ImpulseRadar AB, Sweden), connected to GNSS with CPOS correction.
+For the work presented here, we used a total of 403440 interpreted and calibrated traces along 7.4 km of transects -- discarding data where multiple depths were interpreted for the same location.
+
 ## Predictive models of peat depth
@@
 
-We used RF to predict peat depth at both sites.
+We used random forests (RF) to predict peat depth at both sites.
 RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees [@breimanRandomForests2001].
@@
 
-For both sites we compared the performance of models with five different configurations of three groups of predictors. 
+For both sites we compared the performance of models with five different configurations of predictors. 
 Specifically, we trained models with:
@@
 
-These different configurations simulate different scenarios of data availability that are common in Norway or internationally.
+These different configurations simulate different scenarios of data availability.
 Comparing the different configurations allowed us to isolate the added value of each of the predictor groups.
@@
 
-The performance of DSM must be evaluated with reference to a specific purpose [i.e., map vs.
-model validation, interpolation vs.
-extrapolation, @robertsCrossvalidationStrategiesData2017; @milaNearestNeighbourDistance2022], and here we aimed to evaluate maps of peat depth across the study areas.
-In the absence of additional field work to collect a design-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance [@wadouxSpatialCrossvalidationNot2021; @meyerMachineLearningbasedGlobal2022].
-Specifically, we used k-Means Nearest Neighbor Distance Matching (kNNDM), which creates cross-validation folds that mimic the spatial prediction task that is defined as the goal [@linnenbrinkKNNDMCVKfold2024].
+We used a spatial cross-validation scheme to evaluate model performance [@wadouxSpatialCrossvalidationNot2021; @meyerMachineLearningbasedGlobal2022].
+To set the folds we used k-Means Nearest Neighbor Distance Matching (kNNDM), which aims to mimic the spatial prediction task that is defined as the goal [@linnenbrinkKNNDMCVKfold2024].
 In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
@@
 
-DSM products have much more value when their predictions are accompanied by uncertainty estimates, and all DSM should strive to assess uncertainty [@arrouaysImpressionsDigitalSoil2020; @wadouxMachineLearningDigital2020].
-Moreover, the quality of uncertainty estimates should be evaluated, just as the quality of predictions are [@heuvelinkSpatialStatisticsSoil2022].
+DSM products have much more value when their predictions are accompanied by uncertainty estimates, and all DSM should strive to assess uncertainty [@arrouaysImpressionsDigitalSoil2020; @wadouxMachineLearningDigital2020] and evaluate uncertainty estimates [@heuvelinkSpatialStatisticsSoil2022].
 Therefore, we produced prediction intervals with quantile regression forests [@meinshausenQuantileRegressionForests2006], and used the same spatial cross-validation to evaluate the prediction interval coverage probability [@shresthaMachineLearningApproaches2006].
@@
 
-While the primary aim of the DSM was to predict peat depth within AR5 mires (where peat depth is supposed to be > 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth or occurrence outside of AR5 mire areas.
-In other words, we tested whether our mire-trained models of peat depth could identify peatlands (with or without mire vegetation).
-It is difficult to obtain good data for this purpose, because peat soil not covered by mire vegetation frequently goes unidentified and is relatively rare at these sites.
-For Skrimfjella, we had independent data from the arm of our sampling design that measured peatland occurrence outside of AR5 mires.
-For Ørskogfjellet we did not have such data.
-Therefore, we first -- for both sites -- performed an evaluation with only our peat depth measurements, which included a fraction of locations outside of AR5 mires.
+While the primary aim of the DSM was to predict peat depth within AR5 mires (where peat depth is supposed to be > 30 cm), we also tested whether our mire-trained models of peat depth could identify peatlands outside of AR5 mires.
+For both sites we performed an evaluation with our (cell-aggregated) peat depth measurements, of which about one-fifth were outside of AR5 mires.
 For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on AR5 mire locations (in the training folds), and evaluated them only on (test fold) locations classified in AR5 as something other than mire.
-In this way, we simulated the situation where a model trained on AR5 mires is used to predict peat depth outside AR5 mires.
 From the cross-validation we quantified mean absolute error.
-Second -- for Skrimfjella alone -- we simply trained a model on all peat depth measurements (AR5 mire or not) and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
-In this analysis, the model's prediction of peat depth is treated as a rank index of peatland likelihood.
+For Skrimfjella -- where we measured peatland occurrence separately from depth -- we also trained a model on all peat depth measurements (AR5 mire or not) and then used the independent occurrence dataset to evaluate this model.
+Specifically, we calculated the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
+This analysis treats the model's prediction of peat depth as a rank index of peatland likelihood.
 Since the purpose was to test the models' ability to uncover unmapped peatland (where DMK peat depth class is undefined), we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.
@@
 
-We quantified global variable importance and examined partial dependence plots for the best-performing predictor configuration at each site.
-Global variable importance measures the influence of a given predictor on the output of the model, aggregated across all locations.
-Partial dependence plots depict the shape of the fitted relationship between a given predictor and the outcome.
+We quantified global variable importance (predictor influence across locations) and examined partial dependence plots (curves of fitted relationships) for the best-performing predictor configuration at each site.
 Both are useful for understanding the mechanisms behind the model's predictions and the roles of the predictors in the model.
@@
 *FIRM* values measure the flatness of the partial dependence plot, *permutation* values measure the decrease in model performance when the predictor is permuted, and *Shapley* values are aggregated from local, game-theoretical measures of variable importance [@greenwellVariableImportancePlots2020].
-*Permutation* values were obtained from ten iterations, with RMSE as the performance measure.
+*Permutation* values were obtained from ten iterations, with root mean square error as the performance measure.
 
 We calculated partial dependence with the *pdp* R package [v.0.8.1, @greenwellPdpPackageConstructing2017].
-For the top six most important variables, we plotted both partial dependence and individual conditional expectation (ICE), to show the average effect of the predictor on the outcome and the variation in the effect across observations, respectively [@goldsteinPeekingBlackBox2015].
-Non-parallel ICE lines indicate the presence of interactions between predictors.
+For the six most important variables, we plotted both partial dependence and individual conditional expectation, to show the average effect of the predictor on the outcome and the variation in the effect across observations, respectively [@goldsteinPeekingBlackBox2015].
+Non-parallel individual conditional expectation lines indicate the presence of interactions between predictors.
 
@@
 
-We obtained depth measurements for 372 cells at Skrimfjella (2.4 % of AR5 mire area) and 1878 cells at Ørskogfjellet (1.2 % of AR5 mire area).
+Our point measurements of peat depth (all sources) produced aggregated depths for 372 cells at Skrimfjella (2.4 % of AR5 mire area) and 1878 cells at Ørskogfjellet (1.2 % of AR5 mire area).
 Roughly 80 % of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
@@
 
-None of the models was able to predict peat depth across the study areas with high accuracy (Fig. \@ref(fig:modelMetrics)).
+None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \@ref(fig:modelMetrics)).
 For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an R^2^ of 0.34, and a mean absolute error of 60 cm.
-For Ørskogfjellet the same values were 0.39, 0.33, and 56 cm, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
+For Ørskogfjellet the concordance correlation coefficient was 0.39, R^2^ was 0.33, and mean absolute error was 56 cm, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
 These values were derived from kNNDM spatial cross-validation with 20 folds at Skrimfjella and 10 folds at Ørskogfjellet.
@@
 Compared to the *terrain-only* configuration, the *terrain and radiometric* configuration improved concordance correlation by 0.04, R^2^ by 0.04, and mean absolute error by 1 cm.
-*DMK class* alone -- although calibrated to measured depths -- was a very poor predictor of peat depth, with a concordance correlation coefficient of 0.008.
+*DMK class* alone -- although calibrated to measured depths -- was a very poor predictor of peat depth, with a concordance correlation coefficient of 0.0077.
 
@@
 ```{r calPlots, echo = FALSE, fig.cap = "Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions from spatial cross-validation.
-Blue smoother lines are local polynomial regressions.
-Marginal distributions are shaded by quartile."}
+Points are transparent to show overlap.
+Blue lines are local polynomial regressions and the red dashed line in each panel shows the 1:1 line.
+Marginal distributions (top) are shaded by quartile."}
 knitr::include_graphics("figures/calibration_plots.pdf")
@@
 Similarly, the *terrain and DMK* configuration for Ørskogfjellet was reduced from 23 variables to 11 non-collinear variables.
-At both sites, *elevation* and *MRVBF* were important predictors (Fig. \@ref(fig:varImp)).
-At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet *elevation* was more important than *MRVBF*.
-*DMK* was also important -- the shallow class in particular -- but only at Ørskogfjellet.
-Some realizations of the hydrological predictors TWI and DTW showed considerable importance, while others showed little -- with no consistency between sites.
-For example, *TWI5m*, *DTW40000*, and *DTW2500* rounded out the top five most important variables at Skrimfjella, while *TWI20m* and *TWI50m* were least important, other than *DMK*.
-At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices slope, TRI, TPI, and roughness.
-The radiometric variable *radU* showed moderate importance at Skrimfjella.
+At both sites, elevation and Multi-Resolution Valley Bottom Flatness were important predictors (Fig. \@ref(fig:varImp)).
+At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet elevation was more important than Multi-Resolution Valley Bottom Flatness.
+DMK was also important -- the shallow class in particular -- but only at Ørskogfjellet.
+Some realizations of the hydrological predictors Topographic Wetness Index and Depth-to-Water showed considerable importance, while others showed little -- with no consistency between sites.
+For example, TWI5m, DTW40000, and DTW2500 rounded out the top five most important variables at Skrimfjella, while TWI20m and TWI50m were least important, other than DMK.
+At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices slope, Terrain Ruggedness Index, Topographic Position Index, and roughness.
+The radiometric variable radU showed moderate importance at Skrimfjella.
 
 ```{r varImp, echo = FALSE, fig.cap = "Global variable importance in the best-performing models at Ørskogfjellet (a) and Skrimfjella (b), as measured by three different metrics.
-Variables removed due to collinearity are shown to the right of that with which they are most correlated."}
+Variables removed due to collinearity are shown to the right of the variable with which they are most correlated."}
 knitr::include_graphics("figures/variable_importance.pdf")
@@
 Many of the the most important predictors in the best performing models showed non-monotonic effects on peat depth (Fig. \@ref(fig:pdps)).
-At Ørskogfjellet for example, increasing *elevation* was predictive of deeper peat up to about 75 m above sea level, after which a further increase was predictive of shallower peat.
-At Skrimfjella the partial dependence on *elevation* had the opposite shape, with the shallowest peats predicted at intermediate elevations, around 350 m above sea level.
-*TWI50m* at Ørskogfjellet and *DTW4000* and *radU* at Skrimfjella were other predictors that showed considerable fluctuations in their predictive effects on peat depth.
-The radiometric predictor in particular displayed an idiosyncratic effect, with a marked dip in predicted depth at intermediate values of *radU*.
+At Ørskogfjellet for example, increasing elevation was predictive of deeper peat up to about 75 m above sea level, after which a further increase was predictive of shallower peat.
+At Skrimfjella the partial dependence on elevation had the opposite shape but covers a higher elevation range, with the shallowest peats predicted around 350 m above sea level.
+TWI50m at Ørskogfjellet and DTW4000 and radU at Skrimfjella were other predictors that showed considerable fluctuations in their predictive effects on peat depth.
+The radiometric predictor in particular displayed an idiosyncratic effect, with a marked dip in predicted depth at intermediate values of radU.
 On the other hand, the partial effects of some important predictors were more straightforward.
-The partial dependence on *MRVBF* was quite similar across sites, with the deepest peats predicted in the very flattest valley bottoms.
-Also, *TWI5m* and *DTW2500* at Skrimfjella showed monotonically positive and negative predictive effects, respectively.
+The partial dependence on Multi-Resolution Valley Bottom Flatness was quite similar across sites, with the deepest peats predicted in the very flattest valley bottoms.
+Also, TWI5m and DTW2500 at Skrimfjella showed monotonically positive and negative predictive effects, respectively.
 
 Individual conditional expectation lines indicated some interactions between predictors (Fig. \@ref(fig:pdps)).
-For example, the magnitude of the increase in depth with *elevation* that the model expected at Ørskogfjellet was different for different observations; some depth predictions increased by only 40 cm while others increased by more than 100 cm, over the same elevation gain.
-Similarly, ICE lines of *radU* at Skrimfjella were non-parallel, with some locations showing monotonically increasing peat depth predictions with uranium concentration (unlike the average effect).
-Nevertheless, most ICE lines were generally parallel -- indicating that the average effects of the predictors were good representations of their overall effects.
+For example, the magnitude of the increase in depth with elevation that the model expected at Ørskogfjellet was different for different observations; some depth predictions increased by only 40 cm while others increased by more than 100 cm, over the same elevation gain.
+Similarly, individual conditional expectation lines of radU at Skrimfjella were non-parallel, with some locations showing monotonically increasing peat depth predictions with uranium concentration (unlike the average effect).
+Nevertheless, most individual conditional expectation lines were generally parallel -- indicating that the average effects of the predictors were good representations of their overall effects.
 
@@
 
-On the other hand, we improved on the best available map of peat depth (DMK depth class), which is based on field measurements only.
+Ground-based measurements also have uncertainty [@parryEvaluatingApproachesEstimating2014], but our data show that this error was smaller than the variation in peat depth over short distances (e.g.
+100 m).
+Independent probing and GPR measurements up to 2 m apart were differed by an average of 29--42 cm (Fig. \@ref(fig:GPRwavevelocity)).
+By comparison, the average variation among measurements 100 m apart was 70--106 cm (Fig. \@ref(fig:semivariograms)).
+We also note that cell-level measurement error that is correlated with predictors (e.g.
+underestimated depth at high Multi-Resolution Valley Bottom Flatness), will go undetected by cross-validation and does not account for the limited accuracy of our models.
+Ultimately, the effect of measurement error falls mostly outside the scope of DSM, since it can only be evaluated with higher-quality, independent data.
+
+Although model performance was limited, we improved on the best available map of peat depth (DMK depth class), which is based on field measurements only.
 This highlights the general value of remotely sensed data, whose complete coverage can improve maps even when their association with the variable of interest is weak.
@@
 
-DMK peat depth classes were a worse predictor of peat depth than our models even though we calibrated them with the same data.
+DMK peat depth classes were a worse predictor of peat depth than our models even though we calibrated them with the same data (Fig. \@ref(fig:modelMetrics)).
 That is: we gave DMK and the DSM equal footing for a fair comparison.
@@
 Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth but generally produced better or comparable error magnitude (R^2^ vs.
-MAE/RMSE; @wadouxIntegratedApproachEvaluation2022).
+mean absolute error or root mean square error; @wadouxIntegratedApproachEvaluation2022).
 It is important to keep in mind that differences in peat depth distributions, spatial scales, and evaluation methods make direct performance comparisons precarious. 
@@
 [-@gatisMappingUplandPeat2019] at 100 m resolution, using splines to relax linearity between radiometry/terrain and peat depth.
-He found that radK alone predicted peat depth with much higher concordance than our models (CCC = 0.76) and that elevation alone produced comparable performance (CCC = 0.27). 
-The RMSE from these univariate models was 46--68 cm (cf. 78 and 75 cm for Skrimfjella and Ørskogfjellet).
+He found that radK alone predicted peat depth with much higher concordance than our models (concordance correlation coefficient = 0.76) and that elevation alone produced comparable performance (concordance correlation coefficient = 0.27).
+The root mean square error from these univariate models was 46--68 cm (cf. 78 and 75 cm for Skrimfjella and Ørskogfjellet).
 
@@
 [-@kogantiMappingPeatDepth2023] and our study (besides spatial scale) is that they measured radiometrics on the ground, rather than using airborne survey data.
+Thus, the footprint of their detector was much smaller and they could capture variation in radiation at finer scales.
 
@@
 
-Our models erred most for the deepest peats. 
+Our models erred most for the deepest peats (Fig. \@ref(fig:calPlots)).
 Where overprediction occurred, it was smaller. 
@@
 Unfortunately, our models did not predict peat occurrence outside AR5 mires. 
-In these areas, assuming a constant 30 cm of peat (the depth threshold used in AR5) produced less error than our models. 
+In these areas, assuming a constant 30 cm of peat (the depth threshold used in AR5) produced less error than our models (Fig. \@ref(fig:modelMetricsExtrapolation)). 
 Moreover, the performance gap would likely have been larger had our evaluation data been representative of all areas not classified as mire in AR5. 
@@
 
-At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables. 
-Elevation was the most important predictor at both sites, and peat depth showed non-monotonic responses to changes in elevation. 
+At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables (Fig. \@ref(fig:varImp)). 
+Elevation was the most important predictor at both sites, and peat depth showed non-monotonic responses to changes in elevation (Fig. \@ref(fig:pdps)). 
 We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g.
 via climate) to explain the observed patterns.
-Indeed, elevation at Ørskogfjellet seems to interact with other variables (non-parallel ICE lines), complicating its interpretation. 
+Indeed, elevation at Ørskogfjellet seems to interact with other variables (non-parallel individual conditional expectation lines), complicating its interpretation. 
 Moreover, relationships between elevation and peat depth have previously shown opposite shapes in different areas [@finlaysonEstimatingOrganicSurface2021]. 
 Nonetheless, a relationship that is not generalizable beyond the mapping area is still useful for DSM, as long as it is evaluated to demonstrate its robustness for the predictive task (e.g.
-through kNNDM spatial cross validation).
+through kNNDM spatial cross-validation).
 
@@
 
-Another influential terrain-based predictor was MRVBF. 
-Unlike elevation, it showed a monotonic effect on peat depth: greater valley bottom flatness was always associated with increases in peat depth. 
-Delineating a valley bottom involves ambiguity, but the MRVBF index is a pragmatic approach that considers a location a valley bottom if it is sufficiently low and flat at a particular scale [@gallantMultiresolutionIndexValley2003].
+Another influential terrain-based predictor was Multi-Resolution Valley Bottom Flatness (Fig. \@ref(fig:varImp)). 
+Unlike elevation, it showed a monotonic effect on peat depth: greater valley bottom flatness was always associated with increases in peat depth (Fig. \@ref(fig:pdps)). 
+Delineating a valley bottom involves ambiguity, but the Multi-Resolution Valley Bottom Flatness index is a pragmatic approach that considers a location a valley bottom if it is sufficiently low and flat at a particular scale [@gallantMultiresolutionIndexValley2003].
 The multiscale nature of the index allows small elevated but flat areas (including saddles) to be characterized as having high valley bottom flatness [@gallantMultiresolutionIndexValley2003]. 
-Our results suggest that MRVBF is a robust indicator of high water tables (and peat accumulation) over millennial time scales, corroborating other studies [@rudiyantoOpenDigitalMapping2018; @deragonMappingMaximumPeat2023].
+Our results suggest that Multi-Resolution Valley Bottom Flatness is a robust indicator of high water tables (and peat accumulation) over millennial time scales, corroborating other studies [@rudiyantoOpenDigitalMapping2018; @deragonMappingMaximumPeat2023].
 
-Other terrain-derived predictors with predictive value in our study are hydrological (TWI and DTW). 
-Notably, terrain slope was inferior to (Ørskogfjellet) or highly correlated with (Skrimfjella) these hydrological indices. 
+Other terrain-derived predictors with predictive value in our study are hydrological (Topographic Wetness Index and Depth-to-Water). 
+Notably, slope was inferior to (Ørskogfjellet) or highly correlated with (Skrimfjella) these hydrological indices. 
 Mappers of peat depth should not assume that slope is the best predictor in its class, despite its prevalence in the literature. 
-Wetter locations (high TWI and low DTW) were generally associated with deeper peat, but these relationships were not as strong or consistent as with MRVBF. 
-The optimal scale for TWI and DTW varied, and likely depends on both the dominant peat formation processes and the typical size of peatland features in a landscape. 
+Wetter locations (high Topographic Wetness Index and low Depth-to-Water) were generally associated with deeper peat, but these relationships were not as strong or consistent as with Multi-Resolution Valley Bottom Flatness (Figs.
+\@ref(fig:varImp), \@ref(fig:pdps)).
+The optimal scale for Topographic Wetness Index and Depth-to-Water varied, and likely depends on both the dominant peat formation processes and the typical size of peatland features in a landscape. 
 Including multiple scales of these variables allows the model to capture different hydrological mechanisms operating at different spatial scales.
@@
 DMK peat depth class proved an inconsistent predictor of peat depth. 
-At Skrimfjella, it barely improved model performance. 
-At Ørskogfjellet, it increased performance more, and both indicator variables were among the most important in the model. 
+At Skrimfjella, it barely improved model performance (Fig. \@ref(fig:modelMetrics)). 
+At Ørskogfjellet, it increased performance more, and both indicator variables were among the most important in the model (Fig. \@ref(fig:varImp)). 
 We suspect the discrepancy between sites is due to different levels of effort and coverage during the historical surveys; more lowland peatland near agriculture at Ørskogfjellet may have caused more purposeful surveying.
@@
 
-Depth measurements are foundational, and better infrastructure to collect, preserve, and make FAIR these data would help DSM and other applications. 
+Depth measurements are foundational, and better infrastructure to make these data findable, accessible, interoperable, and reusable would help DSM and other applications. 
 Geoportal access and data exchange standards [like Natural England’s for peat surveys, -@naturalenglandDataExchangeStandard2023] are important. 
 Peat data often fall through the cracks between geology-oriented and ecology-oriented archives, but increased awareness of peatland importance is a good impetus to remedy this situation. 
-Peat depth is quick and easy to measure, so integrating its measurement into existing national field programs, like Norway’s spatially representative nature monitoring (ANO) or national forest inventory, would be helpful (although not sufficient for regional or local mapping).
+Peat depth is quick and easy to measure, so integrating its measurement into existing national field programs, like Norway’s spatially representative nature monitoring or national forest inventory, would be helpful (although not sufficient for regional or local mapping).
 Municipalities and local actors in Norway are increasingly measuring peat depth and can contribute to a growing data foundation [@kyrkjeeideCalculatorLocalPeatland2023]. 
@@
 
+Peat depths typically vary over short distances (e.g. < 1 m), so mapping at 10 m resolution implies that the map will compress much of the fine-scale variation in depth.
+For example, large differences in maximum depth can be obscured at 10 m resolution.
+For the same reason, seemingly small improvements in mean absolute error at 10 m resolution may represent meaningful improvements for map users.
+Spatial aggregation also means that an uneven distribution of point measurements within a cell can cause its mean depth to be unrepresentative.
+Some cells had dense and evenly distributed measurements, while others had only a few in one corner.
+Although we tried to reduce the influence of uneven point coverage -- by inversely weighing point values by their distances to the cell center -- it probably contributed to model inaccuracy.
+Fine-scale variation in peat depth may favor mapping at very fine resolution (1 m) — even if land use planning and carbon accounting do not operate at this grain. 
+Terrain--depth relationships might be stronger in 1 m cells than in our 10 m cells, especially considering the hummock--hollow microtopography of many peatlands.
+
 The scarcity of depth measurements and their short spatial autocorrelation mean that mapping peat depth at broad scales is not an exercise in spatial interpolation [@henglGenericFrameworkSpatial2004].
-Peat depths in our GPR data showed spatial autocorrelation to a range of 50--100 m, and getting measurements at such fine grain is only realistic for small areas, not across landscapes. 
+Peat depths in our GPR data showed spatial autocorrelation to a range of 50--100 m, and getting measurements at such fine grain is only realistic for small areas, not across whole landscapes. 
 Therefore, we anticipate that spatially explicit DSM is of limited value in all but the most intensively sampled landscapes (currently nowhere in Norway).
-Fine-scale variation in peat depth raises the possibility of mapping at very fine resolution (1 m) — even if land use planning and carbon accounting do not operate at this grain. 
-Terrain--depth relationships might be stronger in 1m cells than in our 10 m cells, considering the hummock--hollow microtopography of many peatlands.
 
@@
 @kogantiMappingPeatDepth2023] and produces good uncertainty estimates [our study, @vaysseUsingQuantileRegression2017;@takoutsingComparingPredictionPerformance2022].
-As depth data become more abundant, we may move from pixel-based learners to deep learning approaches like convolutional neural networks (CNNs). 
+As depth data become more abundant, we may move from pixel-based learners to deep learning approaches like convolutional neural networks. 
 Bakkestuen et al.
-[-@bakkestuenDelineationWetlandAreas2023] successfully predicted peatland extent using CNNs, and their ability to automatically learn multiscale spatial features would reduce the need to manually engineer these.
-MRVBF's success as a predictor of peat depth demonstrates that multiscale spatial patterns matter for peat depth and CNNs are designed to learn such patterns. 
-The kind of relationship described in Buffam et al. [-@buffamFillingHolesRegional2010], where peat depth in basins related to terrain slope at the basin edge, is also something a CNN could learn. 
+[-@bakkestuenDelineationWetlandAreas2023] successfully predicted peatland extent using convolutional neural networks, and their ability to automatically learn multiscale spatial features would reduce the need to manually engineer these.
+The success of Multi-Resolution Valley Bottom Flatness as a predictor of peat depth demonstrates that multiscale spatial patterns matter for peat depth and convolutional neural networks are designed to learn such patterns.
+The kind of relationship described in Buffam et al.
+[-@buffamFillingHolesRegional2010], where peat depth in basins related to terrain slope at the basin edge, is also something a convolutional neural network could learn.
 However, since this approach is data-hungry, we should build soil knowledge into the DSM where we can [@minasnySoilScienceInformedMachine2024]. 

### 2025-01-16 ms/ms.tex
addressing comments

@@
 Peat depth also changes with elevation in many contexts, but with inconsistent directionality \citep[e.g.][]{holdenEstimatingCarbonStock2011, parryMethodModellingPeat2012, rudiyantoDigitalMappingCosteffective2016, rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024}.
-Also more complex derivations of topography, such as the Topographic Wetness Index (TWI) and the Multi-Resolution Valley Bottom Flatness (MRVBF) index, have shown associations with peat depth in some studies \citep[e.g.][]{rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024}.
+Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth in some studies \citep[e.g.][]{rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024}.
 Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic variables and the resolution of the peat depth analysis.
@@
 \begin{figure}
-\includegraphics[height=0.82\textheight]{figures/sites-patchwork} \caption{Sites in southern Norway (a) with land cover at Skrimfjella (b) and Ørskogfjellet (c).
+\includegraphics[height=0.81\textheight]{figures/sites-patchwork} \caption{Sites in southern Norway (a) with land cover at Skrimfjella (b) and Ørskogfjellet (c).
 To fit the scale of the maps, the land cover shown here is from the AR50 national land resource dataset, which has simplified geometry with respect to the AR5 dataset that is used in the study.
@@
 At Skrimfjella we delineated a study area of 34 km\textsuperscript{2} based on radiometric coverage and accessibility (Fig. \ref{fig:sites}b).
-The study area has a diverse bedrock, with 32 \% alkali feldspar granite, 26 \% marlstone, 10 \% granite, 8 \% mozonite, 7 \% sandstone, 6 \% limestone, and five other rock types with 1--5 \% coverage (NGU, 1:250 000 dataset).
+The study area has a diverse bedrock, with 32 \% alkali feldspar granite, 26 \% marlstone, 10 \% granite, 8 \% mozonite, 7 \% sandstone, 6 \% limestone, and five other rock types with 1--5 \% coverage (Geological Survey of Norway, 1:250 000 dataset).
 The area within our delineation at Skrimfjella is classified as the landscape type \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
@@
 At Ørskogfjellet we defined a study area of 124 km\textsuperscript{2} which basically followed the southernmost part of the radiometric survey extent (Fig. \ref{fig:sites}a).
-Bedrock in the area is 84 \% granitic gneiss, 11 \% granite, and 5 \% aluminium silicate gneiss (NGU, 1:250 000 dataset).
+Bedrock in the area is 84 \% granitic gneiss, 11 \% granite, and 5 \% aluminium silicate gneiss (Geological Survey of Norway, 1:250 000 dataset).
 This study area comprises a wide range of major landscape types: \emph{coastal plains}, \emph{coastal fjord}, \emph{inland valleys}, as well as \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
@@
 
-\subsection{Peat depth measurements}
-
-At both study sites, our measurements of peat depth were made for the purpose of training a Random Forest (RF) model of peat depth, and we designed our sampling with this in mind \citep{brusSamplingDigitalSoil2019}.
-Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
-A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population while avoiding spurious ones \citep{brusSamplingDigitalSoil2019}.
-We chose for our sampling and modelling a spatial resolution of 10 m.
-We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other.
-The point measurements of peat depth described below were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
-
-\subsubsection{Skrimfjella}
-
-We measured peat depth in selected locations (10 m raster cells) at Skrimfjella.
-The locations were chosen only from areas delineated as mire in the AR5 dataset.
-Within these AR5 mires, we stratified our sample across values of elevation, slope, and potassium ground concentration \citep[from processed airborne gamma ray spectrometry,][]{baranwalHelicopterborneMagneticElectromagnetic2013}.
-Specifically, we used the \emph{eSample} function in the \emph{iSDM} R package (v.1.0) to chose an environmentally systematic sample.
-This function defines the environmental space as a two-dimensional convex hull around the ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest \citep{hattabUnifiedFrameworkModel2017}.
-Elevation was extracted from the 10 m national DTM, slope calculated in degrees, and potassium ground concentration downscaled with bilinear resampling.
-We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and with these parameters \emph{eSample} returned 105 raster cells.
-
-In addition to the peat depth locations, we had another arm of our sampling design for measuring peatland occurrence, as a binary variable.
-We wanted to measure peatland occurrence outside of AR5 mires because the AR5 dataset is known to underestimate peatland coverage \citep[especially in forests,][]{brynLandCoverNorway2018}, and because airborne radiometrics may help identify unmapped peatland \citep{gatisMappingUplandPeat2019, olearyDigitalSoilMapping2022}.
-The occurrence locations were sampled from the part of the study area that (1) was mapped as something other than mire in the AR5 dataset and (2) had a slope \textless{} 20°.
-We performed environmentally systematic sampling of this population with the same procedure as for the depth locations, and \emph{eSample} returned 106 raster cells.
-
-Field work at Skrimfjella was conducted in August 2020.
-We navigated to the centers of the raster cells in the depth and occurrence samples with a handheld global navigation satellite system (GNSS) receiver, checking that positional error was below 3 m.
-For each depth sample location, we measured peat depth three times (at the vertices of a triangle with 2 m sides) to get a more representative value for the 10 m raster cell, and to dampen the effect of outlying measurements \citep{parryEvaluatingApproachesEstimating2014}.
-We used a metal probe pushed downward until resistance indicated the base of the peat column.
-Probe locations were adjusted up to 20 cm if the base of the peat column seemed to be blocked by an obvious artifact.
-A single probe measurement was right-censored because the peat column was deeper than the probe.
-For each occurrence sample location, we recorded the presence or absence of peatland -- primarily by digging and examining the top 20 cm of soil (where this was possible).
-We judged whether the soil was a peat soil based on its density, texture, and color.
-Occasionally, when the soil itself was difficult to judge, we made our determination also based on the presence or absence of mire vegetation.
-Although peat soil is strictly defined by organic content (which we did not analyze), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland \citep{minasnyMappingMonitoringPeatland2023}.
-
-Besides the depth and occurrence measurements described above, we also measured peat depth in three subjectively chosen, separate peatlands, using ground-penetrating radar (GPR).
-We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with its 500 MHz\} shielded antenna mounted in a plastic sledge, and its control unit connected to a GNSS receiver.
-At each of the three peatlands we recorded GPR traces along walking transects that covered the extent of the peatland, mostly in traversing, zigzag patterns with between 5 m and 20 m spacing at their vertices.
-Along the GPR transects we also probed peat depth at marked trace locations, to be able to calibrate the GPR wave speed velocity.
-We processed the GPR data with Reflex2DQuick software (v.3.0, Sandmeier Scientific Software, Germany), applying a time-zero correction, a dewow filter, and a gain filter based on observed energy decay.
-Then we picked strong reflectors in the radargrams that we interpreted as the base of the peat column.
-We used picks at marked trace locations to calibrate wave speed velocity; we pooled calibration points across the three peatlands and fitted a linear regression of depth on one-way travel time with the intercept fixed at zero.
-In total we had 46 calibration points along 3.5 km of GPR transects.
-Finally, we used the calculated wave velocity (\unit{0.0387\,m\,ns^{-1}}, \(R^2 = 0.874\)) to convert the travel times of all picks to calibrated peat depths.
-
-\subsubsection{Ørskogfjellet}
-
-At Ørskogfjellet we also measured peat depth in a sample of 10 m raster cells, selected from the part of the study area classified as mire in the AR5 dataset.
-Before selecting locations, we determined a minimal sample size that would adequately capture the terrain and radiometric properties of the entire AR5 mire area.
-Specifically, we aimed to identify the size at which adding locations produced diminishing decreases in divergence between sample and population distributions -- i.e.~the elbow point in a curve of similarity between sample and population \citep{maloneMethodsImproveUtility2019}.
-This approach has been found to identify sample sizes that correspond with diminishing returns in predictive model performance on external evaluation data \citep{sauretteDivergenceMetricsDetermining2023}.
-We defined a sequence of sample sizes (50--500) and for ten replicate samples at each size \citep[drawn by conditioned latin hypercube sampling,][]{minasnyConditionedLatinHypercube2006, roudierClhsPackageConditioned2011}, we calculated the mean Kullback--Leibler divergence between sample and population distributions \citep{maloneMethodsImproveUtility2019, sauretteDivergenceMetricsDetermining2023}.
-The predictors in the divergence calculation were terrain slope and four radiometrics: potassium, thorium, uranium, and total count.
-Next, we fitted a asymptotic regression of mean divergence on sample size, and identified the sample size at which the curve reached 95 \% of the fitted asymptote.
-Through this procedure we found that we could adequately capture the population distribution with a sample of 160 locations.
-
-To choose 160 locations, we performed feature space coverage sampling.
-This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
-Feature space coverage sampling aims to disperse samples as uniformly as possible in multidimensional predictor space, and is implemented by choosing locations that are closest to cluster centers in a k-means clustering of the standardized predictor space \citep{brusSamplingDigitalSoil2019}.
-Feature space coverage sampling works best when all dimensions are important predictors of the outcome \citep{wadouxSamplingDesignOptimization2019}, and we used the same five predictors that we used to choose sample size: terrain slope and four radiometrics.
-The radiometric predictors were downscaled to 10 m resolution with cubic B-spline resampling in QGIS \citep[v.3,][]{QGISsoftware}.
-We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
-Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
-To adjust, we first restricted the sampling population to AR5 mire areas that were within an arbitrary cost distance of publicly accessible roads.
-Cost distance was calculated using GRASS's \emph{r.walk} function, with friction costs defined by AR5 land classes \citep{GRASSv8-2}.
-After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
-Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 \% (with respect to the ideal sample), but distance in our sample was still only 46 \% of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
-
-Field work at Ørskogfjellet was conducted in August 2023.
-We navigated to the centers of the raster cells in the sample using real time kinematic differential GNSS (Topcon Positioning Systems, USA), to ensure sub-meter positional accuracy.
-At each location we measured peat depth three times by manual probing, with probe locations spaced approximately 2.5 m apart.
-For a tiny fraction of these measurements (5 total), it was not possible to reach the base of the peat column, and a right-censored depth was recorded.
-In areas with dense sampling locations, we also measured peat depth with GPR along snaking transects passing through the centers of the sampling cells (seven transects, 6.2 km total length).
-We used the same GPR system as at Skrimfjella, but with a 100 MHz Malå rough terrain antenna (Guideline Geo AB, Sweden) at some transects.
-To navigate the GPR transects, we placed flags at the cell centers of the sample locations, and used a handheld GNSS receiver to guide the GPR operator.
-At sampling locations crossed by a GPR transect, we arranged the manual probe positions along the transect (for better calibration of the GPR wave speed velocity), while other locations were probed in a triangular pattern around the cell center like at Skrimfjella.
-
-We processed the GPR data with Reflexw software (v.8.5, Sandmeier Scientific Software, Germany), applying a dewow filter, time-zero correction, bandpass filter, gain filter, and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennae.
-The last correction is important for the rough terrain antenna, which has an antenna separation (2.2 m) -- comparable to typical peat depths.
-As with the data from Skrimfjella, we picked the base of the peat column from strong reflectors in the radargram, and calibrated wave velocity with manual probe measurements in a linear regression.
-The points in the regression were created by joining to each probe measurement the travel time of the nearest pick, but only if these were within 2 m of each other.
-In total we had 78 calibration points along 7.8 km of interpretable GPR traces (transect length exceeded because of extra GPR data).
-Finally, we used the calculated wave velocity (\unit{0.0427\,m\,ns^{-1}}, \(R^2 = 0.946\)) to convert the travel times of all picks to calibrated peat depths.
-
-We also used two sets of existing depth measurements from Ørskogfjellet.
-The first set was provided by the Norwegian Public Roads Administration, who commissioned GPR surveys of particular peatland areas in 2020 and 2021.
-The surveys were conducted with a dual channel system (70 MHz and 300 MHz; ImpulseRadar AB, Sweden), connected to GNSS with CPOS correction.
-We used interpreted and calibrated traces from these surveys, and discarded some data where multiple depths were interpreted for the same locations.
-This summed to 7.4 km of interpreted traces.
-The second set of existing depth data we extracted from a paper map made by the Norwegian Soil and Mire Company in 1984.
-This map presents 44 borehole depths (in decimeters) across a 9 ha peatland area.
-We georeferenced the map and digitized the borehole locations and depths.
-
 \subsection{Peat depth predictors}
@@
 All continuous predictors were derived either from an airborne radiometric survey or from a DTM.
-From the radiometric surveys we simply used the four variables produced by the surveyors (Geological Survey of Norway): ground concentration of Potassium, Thorium, Uranium, as well as total count.
+From the radiometric surveys we simply used the four variables produced by the Geological Survey of Norway: ground concentration of Potassium, Thorium, Uranium, as well as total count.
 From the DTMs we calculated several land surface parameters, ranging from simple terrain indices to more complex geomorphometric and hydrological variables \citep{maxwellLandsurfaceParametersSpatial2022}.
 The categorical predictor was peat depth class, from a national map dataset.
+We chose a spatial resolution of 10 m for our predictors, peat depth sample, and modelling.
+We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other.
 Complete descriptions of predictors follow below.
@@
 
-The radiometric survey covering Skrimfjella was conducted in 2008--2011.
-The survey was flown at an average altitude of 75 m and average speed of \unit{108\,km\,h^{-1}}, with flight lines spaced 200 m apart.
-Spectrometer count rates were calibrated annually to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
-The Geological Survey of Norway processed data from the spectrometer following standard procedures outlined by the International Atomic Energy Association, and the processing included: correction for aircraft and cosmic background radiation, correction for radon in the air, window stripping of the gamma ray spectrum, correction for flying height, conversion of count rates to ground concentrations, and finally gridding to 50 m resolution with micro-leveling.
-Further details about the survey and data processing are provided in Baranwal et al. \citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013}.
-We downscaled the processed data to 10 m resolution by cubic spline resampling, using the \emph{terra} package in R.
+The Geological Survey of Norway conducted and processed radiometric surveys over our study areas, as reported in Baranwal et al.
+\citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013} and Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.
+They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling, using the \emph{terra} package in R.
 
-A very similar radiometric survey covering Ørskogfjellet was conducted in December 2014 and January 2015.
-This survey was flown at an average altitude of 80 m and average speed of \unit{88\,km\,h^{-1}}, with flight lines also spaced 200 m apart.
-Spectrometer count rates were calibrated in 2013 to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
-The spectrometer data were processed following the same procedure as for the survey at Skrimfjella, except that a convolution filter was added to smooth the gridded data.
-Further details about the survey and data processing are provided in Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.
-The 10 m resolution predictors from this survey were identical to the layers used in the sampling design (resampled from 50 m resolution with cubic B-splines).
+The radiometric surveys were conduced with similar flight parameters (Table \ref{tab:radSurveys}).
+Spectrometer count rates were calibrated to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
+Data were processed following standard procedures outlined by the International Atomic Energy Association.
+Processing included: correction for aircraft and cosmic background radiation, correction for radon in the air, window stripping of the gamma ray spectrum, correction for flying height, conversion of count rates to ground concentrations, and finally gridding to 50 m resolution with micro-leveling.
+At Ørskogfjellet an additional convolution filter was added to smooth the gridded data.
 
@@
 
-For terrain-derived predictors, we obtained 1 m resolution DTMs from the Norwegian Mapping Authority.
+For terrain-derived predictors, we obtained 1 m resolution rasters from the national DTM (\href{https://creativecommons.org/licenses/by/4.0/}{CC BY 4.0} Norwegian Mapping Authority).
 The DTM for Skrimfjella was produced from airborne laser scanning surveys in 2015 and 2022, with laser point density of \unit{5\,pts\,m^{-2}}.
@@
 
-We used the \emph{terra} R package to calculate from the DTMs: slope, topographic position index (difference from mean of eight neighbors), terrain ruggedness index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
-These were derived at two scales to produce eight different predictors; we either calculated from 1 m DTM resolution and then aggregated, or aggregated to 10 m DTM resolution and then calculated the indices.
+We used the \emph{terra} R package to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
+These were derived at two scales to produce eight different predictors; we either calculated the indices at 1 m DTM resolution and then aggregated to 10 m resolution, or aggregated to 10 m DTM resolution and then calculated the indices.
 This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, dornikOptimalScalingPredictors2022, newmanAssessingSpatiallyHeterogeneous2023}.
 We know that peat depth tends to vary at fine scales in Norway, which is why we chose 1 m and 10 m resolutions \citep{maxwellLandsurfaceParametersSpatial2022}.
-We also calculated the MRVBF index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
-We calculated this index in SAGA GIS \citep[v.9.3.2,][]{conradSystemAutomatedGeoscientific2015} with default parameters.
+We also calculated the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
+We calculated this index in SAGA GIS \citep[v.9.3.2, Morphometry library,][]{conradSystemAutomatedGeoscientific2015} with default parameters (initial slope threshold = 16 \%, lowness threshold = 0.4, upness threshold = 0.35, slope shape parameter = 4, elevation shape parameter = 3).
 
-Next, we calculated the Topographic Wetness Index \citep{quinnPredictionHillslopeFlow1991}.
-This index is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate to coarse resolution DTMs \citep{agrenEvaluatingDigitalTerrain2014, riihimakiTopographicWetnessIndex2021}, so we calculated it from 5 m, 10 m, 20 m, and 50 m DTM resolution.
+The Topographic Wetness Index \citep{quinnPredictionHillslopeFlow1991} is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate to coarse resolution DTMs \citep{agrenEvaluatingDigitalTerrain2014, riihimakiTopographicWetnessIndex2021}, so we calculated it from 5 m, 10 m, 20 m, and 50 m DTM resolution.
 The calculations were performed with Whitebox software \citep{lindsayWhiteboxGATCase2016}, accessed through the \emph{whitebox} R package \citep[v2.4,][]{wuWhiteboxWhiteboxToolsFrontend2022}.
@@
 
-The last terrain-based predictor we included was the depth-to-water (DTW) index \citep{murphyMappingWetlandsComparison2007}.
-This index approximates a location's vertical height above the surface water feature that it is likely to drain towards.
+The Depth-to-Water index \citep{murphyMappingWetlandsComparison2007} approximates a location's vertical height above the surface water feature that it is likely to drain towards.
 It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature \citep[eq. 5 in][]{murphyTopographicModellingSoil2009}.
@@
 Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell \citep{schonauerSpatiotemporalPredictionSoil2021, schonauerRcodeCalculatingDepthwater2021}.
-This catchments area layer was then thresholded at seven different levels (\emph{flow initiation areas} 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry \citep{murphyModellingMappingTopographic2011, agrenEvaluatingDigitalTerrain2014, schonauerSpatiotemporalPredictionSoil2021}.
+This catchments area layer was then thresholded at seven different levels (flow initiation area 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry \citep{murphyModellingMappingTopographic2011, agrenEvaluatingDigitalTerrain2014, schonauerSpatiotemporalPredictionSoil2021}.
 In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
-For each of the seven surface water layers, we derived DTW using the \emph{Distance Accumulation} tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.
+For each of the seven surface water layers, we derived Depth-to-Water using the \emph{Distance Accumulation} tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.
 
@@
 
+\subsection{Peat depth sample selection}
+
+The places (10 m raster cells) we chose to measure peat depth were sampled from mire areas in the AR5 dataset, and optimized for training a Random Forest (RF) model of peat depth \citep{brusSamplingDigitalSoil2019}.
+Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
+A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population, while avoiding spurious ones \citep{brusSamplingDigitalSoil2019}.
+Although we implemented the approach differently for Skrimfjella (in 2020) than for Ørskogfjellet (in 2023), the objective was the same.
+
+We made several point measurements of peat depth within the sampled raster cells (three if by probing only, more if by GPR as well).
+The point measurements (described below) were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
+
+\subsubsection{Skrimfjella}
+
+At Skrimfjella, we used the \emph{eSample} function in the \emph{iSDM} R package (v.1.0) to stratify our sample across elevation, slope, and potassium concentration.
+This function defines the environmental space as a two-dimensional convex hull around the ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest \citep{hattabUnifiedFrameworkModel2017}.
+We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and \emph{eSample} returned 105 raster cells.
+
+We also measured peatland occurrence (binary variable) in a separate sample at Skrimfjella.
+We measured peatland occurrence because the AR5 dataset is known to underestimate peatland coverage \citep[especially in forests,][]{brynLandCoverNorway2018}, and because airborne radiometrics may help identify unmapped peatland \citep{gatisMappingUplandPeat2019, olearyDigitalSoilMapping2022}.
+The 10 m cells were sampled from those outside AR5 mire and with slope \textless{} 20°.
+Following the same procedure, \emph{eSample} returned 106 raster cells from this population.
+
+\subsubsection{Ørskogfjellet}
+
+At Ørskogfjellet, we first determined a minimal sample size that would adequately capture the slope and radiometric properties (potassium, thorium, uranium, and total count) of the entire AR5 mire area \citep{sauretteDivergenceMetricsDetermining2023}.
+Specifically, we identified an elbow point in a curve of similarity between sample and population \citep{maloneMethodsImproveUtility2019}.
+For a sequence of sample sizes (50--500) \citep[ten replicates each, drawn by conditioned latin hypercube sampling,][]{minasnyConditionedLatinHypercube2006, roudierClhsPackageConditioned2011}, we calculated the mean Kullback--Leibler divergence between sample and population distributions \citep{maloneMethodsImproveUtility2019, sauretteDivergenceMetricsDetermining2023}.
+Then, we fitted a asymptotic regression of mean divergence on sample size, and found that the curve reached 95 \% of the fitted asymptote at a sample size of 160.
+
+To choose 160 locations, we performed feature space coverage sampling, implemented using the \emph{kmeans} function in base R and the \emph{rdist} function in the the \emph{fields} package (v.14.2).
+Feature space coverage sampling chooses locations that are closest to cluster centers in standardized predictor space \citep{brusSamplingDigitalSoil2019}.
+This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
+Feature space coverage sampling works best when all dimensions are important predictors of the outcome \citep{wadouxSamplingDesignOptimization2019}, and we used the same five predictors that we used to choose sample size: slope and four radiometrics.
+
+We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
+Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
+To adjust, we first restricted the sampling population to AR5 mire areas that were within an arbitrary cost distance of publicly accessible roads.
+Cost distance was calculated using GRASS's \emph{r.walk} function, with friction costs defined by AR5 land classes \citep{GRASSv8-2}.
+After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
+Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 \% (with respect to the ideal sample), but distance in our sample was still only 46 \% of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
+
+\subsection{Depth measurements}
+
+We measured peat depths at Skrimfjella in August 2020 and at Ørskogfjellet in August 2023.
+At both sites we used manual probing as the primary method of measuring peat depth, and ground-penetrating radar (GPR) as a secondary method.
+That is: peat depth was always measured by probing, and also by GPR in a subset of cells that partly overlapped with the probed cells
+We chose this approach because probing is a fast and reliable method for point measurements, while GPR can provide higher lateral density of data in the same amount of time \citep{parryEvaluatingApproachesEstimating2014}.
+Probed depths serve to calibrate GPR measurements when calibration by common midpoint survey is not possible -- as was the case with our fixed radar antennas -- so the methods are complementary.
+
+\subsubsection{Peat probing}
+
+We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real time kinematic (Ørskogfjellet) global navigation satellite system receivers.
+We dampened the effect of outlying measurements by probing three times at each location \citep{parryEvaluatingApproachesEstimating2014}, at the vertices of a centered triangle with 2 m (Skrimfjella) or 4.5 m (Ørskogfjellet) sides.
+We used changes in resistance to indicate the base of the peat column.
+Probe locations were adjusted up to 20 cm if the base of the peat column seemed to be blocked by an obvious artifact, like a buried rock.
+Where the peat column was deeper than the extendable probe could be manually inserted and extracted by a pair of operators, we recorded a right-censored result (one at Skrimfjella, five at Ørskogfjellet).
+
+We also extracted a set of existing depth probe measurements for Ørskogfjellet from a paper map made by the Norwegian Soil and Mire Company in 1984.
+The map presents 44 borehole depths (in decimeters) across a 9 ha peatland area.
+We georeferenced the map and digitized the borehole locations and depths.
+
+For our sample of peatland occurrence at Skrimfjella, we recorded the presence or absence of peatland at the cell center -- primarily by digging and examining the top 20 cm of soil.
+We judged whether the soil was a peat soil based on its density, texture, and color, as well as the presence or absence of mire vegetation.
+Although peat soil is strictly defined by organic content (which we did not analyze), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland \citep{minasnyMappingMonitoringPeatland2023}.
+
+\subsubsection{Ground-penetrating radar}
+
+We performed GPR surveys in three subjectively chosen mires at Skrimfjella and in areas with a high density of sampled raster cells at Ørskogfjellet.
+We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with a GNSS-enabled control unit connected to a 500 MHz shielded antenna mounted in a plastic sledge (transmitter--receiver separation 0.18 m, trace frequency 10 Hz).
+For some transects at Ørskogfjellet we substituted a 100 MHz Malå rough terrain antenna (transmitter--receiver separation 2.2 m, trace frequency 5 Hz), because the lower frequency antenna gives greater penetration depth.
+In all cases the system was towed by a walking GPR operator.
+
+GPR traces were recorded along zigzag (Skrimfjella) or snaking (Ørskogfjellet) transects.
+At Ørskogfjellet, transects were predetermined to pass through the centers of sampled raster cells, and we marked these precisely with flags to guide the GPR operator.
+A GPR records the time taken for a radio wave to travel from the transmitter to a reflector and back to the receiver, and the velocity of the wave varies with properties of the peat column.
+Therefore, wave velocity has to be calibrated to convert travel time to peat depth, and we probed peat depth along the transects.
+
+We processed the GPR data with Reflex2DQuick (v.3.0; Skrimfjella) or REFLEXW (v.8.5; Ørskogfjellet) software (Sandmeier Scientific Software, Germany).
+We applied a time-zero correction, a dewow filter, and a gain filter based on observed energy decay.
+With Ørskogfjellet data, we also applied a bandpass filter and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennas.
+The latter is important for the rough terrain antenna, where the antenna separation is comparable to typical peat depths.
+From the processed radargrams, we picked travel times from strong reflectors that we interpreted as the base of the peat column.
+
+We used picks near probed depths to calibrate wave speed velocity -- separately for each site.
+Calibration data were created by matching marked trace locations to a corresponding depth probe (Skrimfjella), or by a spatial join that identified interpreted traces and depth probes within 2 m of each other (Ørskogfjellet).
+We had sufficient calibration point density to avoid bias in wave velocity as a major source of error \citep{rosaDeterminingNumberManual2009}: 46 calibration points along 3.5 km of interpretable traces at Skrimfjella, and 78 along 7.8 km at Ørskogfjellet.
+We fitted site-specific linear regressions of probed depth on picked travel time, with the intercept fixed at zero, to estimate wave velocities.
+Notwithstanding a few outlying points, our regressions showed good fits and the resulting velocities are within the range reported for peat \citep{parsekianUncertaintyPeatVolume2012}: \unit{0.0387\,m\,ns^{-1}}, \(R^2 = 0.874\) at Skrimfjella and \unit{0.0427\,m\,ns^{-1}}, \(R^2 = 0.946\) at Ørskogfjellet.
+Finally, we used these two wave velocities to convert the travel times of all picks to calibrated peat depths.
+In total the GPR surveys produced 48579 point measurements of peat depth at Skrimfjella and 32653 at Ørskogfjellet.
+
+We also used a set of existing GPR depth measurements from Ørskogfjellet, commissioned and provided to us by the Norwegian Public Roads Administration.
+These data were collected in 2020 and 2021 with a dual channel system (70 MHz and 300 MHz; ImpulseRadar AB, Sweden), connected to GNSS with CPOS correction.
+For the work presented here, we used a total of 403440 interpreted and calibrated traces along 7.4 km of transects -- discarding data where multiple depths were interpreted for the same location.
+
 \subsection{Predictive models of peat depth}
@@
 
-We used RF to predict peat depth at both sites.
+We used random forests (RF) to predict peat depth at both sites.
 RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees \citep{breimanRandomForests2001}.
@@
 
-For both sites we compared the performance of models with five different configurations of three groups of predictors.
+For both sites we compared the performance of models with five different configurations of predictors.
 Specifically, we trained models with:
@@
 
-The performance of DSM must be evaluated with reference to a specific purpose \citep[i.e., map vs.~model validation, interpolation vs.~extrapolation,][]{robertsCrossvalidationStrategiesData2017, milaNearestNeighbourDistance2022}, and here we aimed to evaluate maps of peat depth across the study areas.
-In the absence of additional field work to collect a design-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance \citep{wadouxSpatialCrossvalidationNot2021, meyerMachineLearningbasedGlobal2022}.
-Specifically, we used k-Means Nearest Neighbor Distance Matching (kNNDM), which creates cross-validation folds that mimic the spatial prediction task that is defined as the goal \citep{linnenbrinkKNNDMCVKfold2024}.
+We used a spatial cross-validation scheme to evaluate model performance \citep{wadouxSpatialCrossvalidationNot2021, meyerMachineLearningbasedGlobal2022}.
+To set the folds we used k-Means Nearest Neighbor Distance Matching (kNNDM), which aims to mimic the spatial prediction task that is defined as the goal \citep{linnenbrinkKNNDMCVKfold2024}.
 In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
@@
 
-DSM products have much more value when their predictions are accompanied by uncertainty estimates, and all DSM should strive to assess uncertainty \citep{arrouaysImpressionsDigitalSoil2020, wadouxMachineLearningDigital2020}.
-Moreover, the quality of uncertainty estimates should be evaluated, just as the quality of predictions are \citep{heuvelinkSpatialStatisticsSoil2022}.
+DSM products have much more value when their predictions are accompanied by uncertainty estimates, and all DSM should strive to assess uncertainty \citep{arrouaysImpressionsDigitalSoil2020, wadouxMachineLearningDigital2020} and evaluate uncertainty estimates \citep{heuvelinkSpatialStatisticsSoil2022}.
 Therefore, we produced prediction intervals with quantile regression forests \citep{meinshausenQuantileRegressionForests2006}, and used the same spatial cross-validation to evaluate the prediction interval coverage probability \citep{shresthaMachineLearningApproaches2006}.
@@
 
-While the primary aim of the DSM was to predict peat depth within AR5 mires (where peat depth is supposed to be \textgreater{} 30 cm), we also performed a minimal evaluation of our ability to estimate peat depth or occurrence outside of AR5 mire areas.
-In other words, we tested whether our mire-trained models of peat depth could identify peatlands (with or without mire vegetation).
-It is difficult to obtain good data for this purpose, because peat soil not covered by mire vegetation frequently goes unidentified and is relatively rare at these sites.
-For Skrimfjella, we had independent data from the arm of our sampling design that measured peatland occurrence outside of AR5 mires.
-For Ørskogfjellet we did not have such data.
-Therefore, we first -- for both sites -- performed an evaluation with only our peat depth measurements, which included a fraction of locations outside of AR5 mires.
+While the primary aim of the DSM was to predict peat depth within AR5 mires (where peat depth is supposed to be \textgreater{} 30 cm), we also tested whether our mire-trained models of peat depth could identify peatlands outside of AR5 mires.
+For both sites we performed an evaluation with our (cell-aggregated) peat depth measurements, of which about one-fifth were outside of AR5 mires.
 For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on AR5 mire locations (in the training folds), and evaluated them only on (test fold) locations classified in AR5 as something other than mire.
-In this way, we simulated the situation where a model trained on AR5 mires is used to predict peat depth outside AR5 mires.
 From the cross-validation we quantified mean absolute error.
-Second -- for Skrimfjella alone -- we simply trained a model on all peat depth measurements (AR5 mire or not) and then used the independent occurrence dataset to calculate the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
-In this analysis, the model's prediction of peat depth is treated as a rank index of peatland likelihood.
+For Skrimfjella -- where we measured peatland occurrence separately from depth -- we also trained a model on all peat depth measurements (AR5 mire or not) and then used the independent occurrence dataset to evaluate this model.
+Specifically, we calculated the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
+This analysis treats the model's prediction of peat depth as a rank index of peatland likelihood.
 Since the purpose was to test the models' ability to uncover unmapped peatland (where DMK peat depth class is undefined), we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.
@@
 \emph{FIRM} values measure the flatness of the partial dependence plot, \emph{permutation} values measure the decrease in model performance when the predictor is permuted, and \emph{Shapley} values are aggregated from local, game-theoretical measures of variable importance \citep{greenwellVariableImportancePlots2020}.
-\emph{Permutation} values were obtained from ten iterations, with RMSE as the performance measure.
+\emph{Permutation} values were obtained from ten iterations, with root mean square error as the performance measure.
 
 We calculated partial dependence with the \emph{pdp} R package \citep[v.0.8.1,][]{greenwellPdpPackageConstructing2017}.
-For the top six most important variables, we plotted both partial dependence and individual conditional expectation (ICE), to show the average effect of the predictor on the outcome and the variation in the effect across observations, respectively \citep{goldsteinPeekingBlackBox2015}.
-Non-parallel ICE lines indicate the presence of interactions between predictors.
+For the six most important variables, we plotted both partial dependence and individual conditional expectation, to show the average effect of the predictor on the outcome and the variation in the effect across observations, respectively \citep{goldsteinPeekingBlackBox2015}.
+Non-parallel individual conditional expectation lines indicate the presence of interactions between predictors.
 
@@
 
-We obtained depth measurements for 372 cells at Skrimfjella (2.4 \% of AR5 mire area) and 1878 cells at Ørskogfjellet (1.2 \% of AR5 mire area).
+Our point measurements of peat depth (all sources) produced aggregated depths for 372 cells at Skrimfjella (2.4 \% of AR5 mire area) and 1878 cells at Ørskogfjellet (1.2 \% of AR5 mire area).
 Roughly 80 \% of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
@@
 
-None of the models was able to predict peat depth across the study areas with high accuracy (Fig. \ref{fig:modelMetrics}).
+None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \ref{fig:modelMetrics}).
 For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an R\textsuperscript{2} of 0.34, and a mean absolute error of 60 cm.
-For Ørskogfjellet the same values were 0.39, 0.33, and 56 cm, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
+For Ørskogfjellet the concordance correlation coefficient was 0.39, R\textsuperscript{2} was 0.33, and mean absolute error was 56 cm, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
 These values were derived from kNNDM spatial cross-validation with 20 folds at Skrimfjella and 10 folds at Ørskogfjellet.
@@
 Compared to the \emph{terrain-only} configuration, the \emph{terrain and radiometric} configuration improved concordance correlation by 0.04, R\textsuperscript{2} by 0.04, and mean absolute error by 1 cm.
-\emph{DMK class} alone -- although calibrated to measured depths -- was a very poor predictor of peat depth, with a concordance correlation coefficient of 0.008.
+\emph{DMK class} alone -- although calibrated to measured depths -- was a very poor predictor of peat depth, with a concordance correlation coefficient of 0.0077.
 
@@
 \caption{\label{fig:calPlots}Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions from spatial cross-validation.
-Blue smoother lines are local polynomial regressions.
-Marginal distributions are shaded by quartile.}
+Points are transparent to show overlap.
+Blue lines are local polynomial regressions and the red dashed line in each panel shows the 1:1 line.
+Marginal distributions (top) are shaded by quartile.}
 \end{figure}
@@
 Similarly, the \emph{terrain and DMK} configuration for Ørskogfjellet was reduced from 23 variables to 11 non-collinear variables.
-At both sites, \emph{elevation} and \emph{MRVBF} were important predictors (Fig. \ref{fig:varImp}).
-At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet \emph{elevation} was more important than \emph{MRVBF}.
-\emph{DMK} was also important -- the shallow class in particular -- but only at Ørskogfjellet.
-Some realizations of the hydrological predictors TWI and DTW showed considerable importance, while others showed little -- with no consistency between sites.
-For example, \emph{TWI5m}, \emph{DTW40000}, and \emph{DTW2500} rounded out the top five most important variables at Skrimfjella, while \emph{TWI20m} and \emph{TWI50m} were least important, other than \emph{DMK}.
-At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices slope, TRI, TPI, and roughness.
-The radiometric variable \emph{radU} showed moderate importance at Skrimfjella.
+At both sites, elevation and Multi-Resolution Valley Bottom Flatness were important predictors (Fig. \ref{fig:varImp}).
+At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet elevation was more important than Multi-Resolution Valley Bottom Flatness.
+DMK was also important -- the shallow class in particular -- but only at Ørskogfjellet.
+Some realizations of the hydrological predictors Topographic Wetness Index and Depth-to-Water showed considerable importance, while others showed little -- with no consistency between sites.
+For example, TWI5m, DTW40000, and DTW2500 rounded out the top five most important variables at Skrimfjella, while TWI20m and TWI50m were least important, other than DMK.
+At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices slope, Terrain Ruggedness Index, Topographic Position Index, and roughness.
+The radiometric variable radU showed moderate importance at Skrimfjella.
 
@@
 \caption{\label{fig:varImp}Global variable importance in the best-performing models at Ørskogfjellet (a) and Skrimfjella (b), as measured by three different metrics.
-Variables removed due to collinearity are shown to the right of that with which they are most correlated.}
+Variables removed due to collinearity are shown to the right of the variable with which they are most correlated.}
 \end{figure}
@@
 Many of the the most important predictors in the best performing models showed non-monotonic effects on peat depth (Fig. \ref{fig:pdps}).
-At Ørskogfjellet for example, increasing \emph{elevation} was predictive of deeper peat up to about 75 m above sea level, after which a further increase was predictive of shallower peat.
-At Skrimfjella the partial dependence on \emph{elevation} had the opposite shape, with the shallowest peats predicted at intermediate elevations, around 350 m above sea level.
-\emph{TWI50m} at Ørskogfjellet and \emph{DTW4000} and \emph{radU} at Skrimfjella were other predictors that showed considerable fluctuations in their predictive effects on peat depth.
-The radiometric predictor in particular displayed an idiosyncratic effect, with a marked dip in predicted depth at intermediate values of \emph{radU}.
+At Ørskogfjellet for example, increasing elevation was predictive of deeper peat up to about 75 m above sea level, after which a further increase was predictive of shallower peat.
+At Skrimfjella the partial dependence on elevation had the opposite shape but covers a higher elevation range, with the shallowest peats predicted around 350 m above sea level.
+TWI50m at Ørskogfjellet and DTW4000 and radU at Skrimfjella were other predictors that showed considerable fluctuations in their predictive effects on peat depth.
+The radiometric predictor in particular displayed an idiosyncratic effect, with a marked dip in predicted depth at intermediate values of radU.
 On the other hand, the partial effects of some important predictors were more straightforward.
-The partial dependence on \emph{MRVBF} was quite similar across sites, with the deepest peats predicted in the very flattest valley bottoms.
-Also, \emph{TWI5m} and \emph{DTW2500} at Skrimfjella showed monotonically positive and negative predictive effects, respectively.
+The partial dependence on Multi-Resolution Valley Bottom Flatness was quite similar across sites, with the deepest peats predicted in the very flattest valley bottoms.
+Also, TWI5m and DTW2500 at Skrimfjella showed monotonically positive and negative predictive effects, respectively.
 
 Individual conditional expectation lines indicated some interactions between predictors (Fig. \ref{fig:pdps}).
-For example, the magnitude of the increase in depth with \emph{elevation} that the model expected at Ørskogfjellet was different for different observations; some depth predictions increased by only 40 cm while others increased by more than 100 cm, over the same elevation gain.
-Similarly, ICE lines of \emph{radU} at Skrimfjella were non-parallel, with some locations showing monotonically increasing peat depth predictions with uranium concentration (unlike the average effect).
-Nevertheless, most ICE lines were generally parallel -- indicating that the average effects of the predictors were good representations of their overall effects.
+For example, the magnitude of the increase in depth with elevation that the model expected at Ørskogfjellet was different for different observations; some depth predictions increased by only 40 cm while others increased by more than 100 cm, over the same elevation gain.
+Similarly, individual conditional expectation lines of radU at Skrimfjella were non-parallel, with some locations showing monotonically increasing peat depth predictions with uranium concentration (unlike the average effect).
+Nevertheless, most individual conditional expectation lines were generally parallel -- indicating that the average effects of the predictors were good representations of their overall effects.
 
@@
 
-On the other hand, we improved on the best available map of peat depth (DMK depth class), which is based on field measurements only.
+Ground-based measurements also have uncertainty \citep{parryEvaluatingApproachesEstimating2014}, but our data show that this error was smaller than the variation in peat depth over short distances (e.g.~100 m).
+Independent probing and GPR measurements up to 2 m apart were differed by an average of 29--42 cm (Fig. \ref{fig:GPRwavevelocity}).
+By comparison, the average variation among measurements 100 m apart was 70--106 cm (Fig. \ref{fig:semivariograms}).
+We also note that cell-level measurement error that is correlated with predictors (e.g.~underestimated depth at high Multi-Resolution Valley Bottom Flatness), will go undetected by cross-validation and does not account for the limited accuracy of our models.
+Ultimately, the effect of measurement error falls mostly outside the scope of DSM, since it can only be evaluated with higher-quality, independent data.
+
+Although model performance was limited, we improved on the best available map of peat depth (DMK depth class), which is based on field measurements only.
 This highlights the general value of remotely sensed data, whose complete coverage can improve maps even when their association with the variable of interest is weak.
@@
 
-DMK peat depth classes were a worse predictor of peat depth than our models even though we calibrated them with the same data.
+DMK peat depth classes were a worse predictor of peat depth than our models even though we calibrated them with the same data (Fig. \ref{fig:modelMetrics}).
 That is: we gave DMK and the DSM equal footing for a fair comparison.
@@
 
-Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth but generally produced better or comparable error magnitude (R\textsuperscript{2} vs.~MAE/RMSE; \citet{wadouxIntegratedApproachEvaluation2022}).
+Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth but generally produced better or comparable error magnitude (R\textsuperscript{2} vs.~mean absolute error or root mean square error; \citet{wadouxIntegratedApproachEvaluation2022}).
 It is important to keep in mind that differences in peat depth distributions, spatial scales, and evaluation methods make direct performance comparisons precarious.
@@
 \citeyearpar{gatisMappingUplandPeat2019} at 100 m resolution, using splines to relax linearity between radiometry/terrain and peat depth.
-He found that radK alone predicted peat depth with much higher concordance than our models (CCC = 0.76) and that elevation alone produced comparable performance (CCC = 0.27).
-The RMSE from these univariate models was 46--68 cm (cf.~78 and 75 cm for Skrimfjella and Ørskogfjellet).
+He found that radK alone predicted peat depth with much higher concordance than our models (concordance correlation coefficient = 0.76) and that elevation alone produced comparable performance (concordance correlation coefficient = 0.27).
+The root mean square error from these univariate models was 46--68 cm (cf.~78 and 75 cm for Skrimfjella and Ørskogfjellet).
 
@@
 \citeyearpar{kogantiMappingPeatDepth2023} and our study (besides spatial scale) is that they measured radiometrics on the ground, rather than using airborne survey data.
+Thus, the footprint of their detector was much smaller and they could capture variation in radiation at finer scales.
 
@@
 
-Our models erred most for the deepest peats.
+Our models erred most for the deepest peats (Fig. \ref{fig:calPlots}).
 Where overprediction occurred, it was smaller.
@@
 Unfortunately, our models did not predict peat occurrence outside AR5 mires.
-In these areas, assuming a constant 30 cm of peat (the depth threshold used in AR5) produced less error than our models.
+In these areas, assuming a constant 30 cm of peat (the depth threshold used in AR5) produced less error than our models (Fig. \ref{fig:modelMetricsExtrapolation}).
 Moreover, the performance gap would likely have been larger had our evaluation data been representative of all areas not classified as mire in AR5.
@@
 
-At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables.
-Elevation was the most important predictor at both sites, and peat depth showed non-monotonic responses to changes in elevation.
+At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables (Fig. \ref{fig:varImp}).
+Elevation was the most important predictor at both sites, and peat depth showed non-monotonic responses to changes in elevation (Fig. \ref{fig:pdps}).
 We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g.~via climate) to explain the observed patterns.
-Indeed, elevation at Ørskogfjellet seems to interact with other variables (non-parallel ICE lines), complicating its interpretation.
+Indeed, elevation at Ørskogfjellet seems to interact with other variables (non-parallel individual conditional expectation lines), complicating its interpretation.
 Moreover, relationships between elevation and peat depth have previously shown opposite shapes in different areas \citep{finlaysonEstimatingOrganicSurface2021}.
-Nonetheless, a relationship that is not generalizable beyond the mapping area is still useful for DSM, as long as it is evaluated to demonstrate its robustness for the predictive task (e.g.~through kNNDM spatial cross validation).
+Nonetheless, a relationship that is not generalizable beyond the mapping area is still useful for DSM, as long as it is evaluated to demonstrate its robustness for the predictive task (e.g.~through kNNDM spatial cross-validation).
 
@@
 
-Another influential terrain-based predictor was MRVBF.
-Unlike elevation, it showed a monotonic effect on peat depth: greater valley bottom flatness was always associated with increases in peat depth.
-Delineating a valley bottom involves ambiguity, but the MRVBF index is a pragmatic approach that considers a location a valley bottom if it is sufficiently low and flat at a particular scale \citep{gallantMultiresolutionIndexValley2003}.
+Another influential terrain-based predictor was Multi-Resolution Valley Bottom Flatness (Fig. \ref{fig:varImp}).
+Unlike elevation, it showed a monotonic effect on peat depth: greater valley bottom flatness was always associated with increases in peat depth (Fig. \ref{fig:pdps}).
+Delineating a valley bottom involves ambiguity, but the Multi-Resolution Valley Bottom Flatness index is a pragmatic approach that considers a location a valley bottom if it is sufficiently low and flat at a particular scale \citep{gallantMultiresolutionIndexValley2003}.
 The multiscale nature of the index allows small elevated but flat areas (including saddles) to be characterized as having high valley bottom flatness \citep{gallantMultiresolutionIndexValley2003}.
-Our results suggest that MRVBF is a robust indicator of high water tables (and peat accumulation) over millennial time scales, corroborating other studies \citep{rudiyantoOpenDigitalMapping2018, deragonMappingMaximumPeat2023}.
+Our results suggest that Multi-Resolution Valley Bottom Flatness is a robust indicator of high water tables (and peat accumulation) over millennial time scales, corroborating other studies \citep{rudiyantoOpenDigitalMapping2018, deragonMappingMaximumPeat2023}.
 
-Other terrain-derived predictors with predictive value in our study are hydrological (TWI and DTW).
-Notably, terrain slope was inferior to (Ørskogfjellet) or highly correlated with (Skrimfjella) these hydrological indices.
+Other terrain-derived predictors with predictive value in our study are hydrological (Topographic Wetness Index and Depth-to-Water).
+Notably, slope was inferior to (Ørskogfjellet) or highly correlated with (Skrimfjella) these hydrological indices.
 Mappers of peat depth should not assume that slope is the best predictor in its class, despite its prevalence in the literature.
-Wetter locations (high TWI and low DTW) were generally associated with deeper peat, but these relationships were not as strong or consistent as with MRVBF.
-The optimal scale for TWI and DTW varied, and likely depends on both the dominant peat formation processes and the typical size of peatland features in a landscape.
+Wetter locations (high Topographic Wetness Index and low Depth-to-Water) were generally associated with deeper peat, but these relationships were not as strong or consistent as with Multi-Resolution Valley Bottom Flatness (Figs.
+\ref{fig:varImp}, \ref{fig:pdps}).
+The optimal scale for Topographic Wetness Index and Depth-to-Water varied, and likely depends on both the dominant peat formation processes and the typical size of peatland features in a landscape.
 Including multiple scales of these variables allows the model to capture different hydrological mechanisms operating at different spatial scales.
@@
 DMK peat depth class proved an inconsistent predictor of peat depth.
-At Skrimfjella, it barely improved model performance.
-At Ørskogfjellet, it increased performance more, and both indicator variables were among the most important in the model.
+At Skrimfjella, it barely improved model performance (Fig. \ref{fig:modelMetrics}).
+At Ørskogfjellet, it increased performance more, and both indicator variables were among the most important in the model (Fig. \ref{fig:varImp}).
 We suspect the discrepancy between sites is due to different levels of effort and coverage during the historical surveys; more lowland peatland near agriculture at Ørskogfjellet may have caused more purposeful surveying.
@@
 
-Depth measurements are foundational, and better infrastructure to collect, preserve, and make FAIR these data would help DSM and other applications.
+Depth measurements are foundational, and better infrastructure to make these data findable, accessible, interoperable, and reusable would help DSM and other applications.
 Geoportal access and data exchange standards \citeyearpar[like Natural England's for peat surveys,][]{naturalenglandDataExchangeStandard2023} are important.
 Peat data often fall through the cracks between geology-oriented and ecology-oriented archives, but increased awareness of peatland importance is a good impetus to remedy this situation.
-Peat depth is quick and easy to measure, so integrating its measurement into existing national field programs, like Norway's spatially representative nature monitoring (ANO) or national forest inventory, would be helpful (although not sufficient for regional or local mapping).
+Peat depth is quick and easy to measure, so integrating its measurement into existing national field programs, like Norway's spatially representative nature monitoring or national forest inventory, would be helpful (although not sufficient for regional or local mapping).
 Municipalities and local actors in Norway are increasingly measuring peat depth and can contribute to a growing data foundation \citep{kyrkjeeideCalculatorLocalPeatland2023}.
@@
 
+Peat depths typically vary over short distances (e.g.~\textless{} 1 m), so mapping at 10 m resolution implies that the map will compress much of the fine-scale variation in depth.
+For example, large differences in maximum depth can be obscured at 10 m resolution.
+For the same reason, seemingly small improvements in mean absolute error at 10 m resolution may represent meaningful improvements for map users.
+Spatial aggregation also means that an uneven distribution of point measurements within a cell can cause its mean depth to be unrepresentative.
+Some cells had dense and evenly distributed measurements, while others had only a few in one corner.
+Although we tried to reduce the influence of uneven point coverage -- by inversely weighing point values by their distances to the cell center -- it probably contributed to model inaccuracy.
+Fine-scale variation in peat depth may favor mapping at very fine resolution (1 m) --- even if land use planning and carbon accounting do not operate at this grain.
+Terrain--depth relationships might be stronger in 1 m cells than in our 10 m cells, especially considering the hummock--hollow microtopography of many peatlands.
+
 The scarcity of depth measurements and their short spatial autocorrelation mean that mapping peat depth at broad scales is not an exercise in spatial interpolation \citep{henglGenericFrameworkSpatial2004}.
-Peat depths in our GPR data showed spatial autocorrelation to a range of 50--100 m, and getting measurements at such fine grain is only realistic for small areas, not across landscapes.
+Peat depths in our GPR data showed spatial autocorrelation to a range of 50--100 m, and getting measurements at such fine grain is only realistic for small areas, not across whole landscapes.
 Therefore, we anticipate that spatially explicit DSM is of limited value in all but the most intensively sampled landscapes (currently nowhere in Norway).
-Fine-scale variation in peat depth raises the possibility of mapping at very fine resolution (1 m) --- even if land use planning and carbon accounting do not operate at this grain.
-Terrain--depth relationships might be stronger in 1m cells than in our 10 m cells, considering the hummock--hollow microtopography of many peatlands.
 
@@
 RF avoids negative predictions \citep[c.f.][]{kogantiMappingPeatDepth2023} and produces good uncertainty estimates \citep[our study,][]{vaysseUsingQuantileRegression2017, takoutsingComparingPredictionPerformance2022}.
-As depth data become more abundant, we may move from pixel-based learners to deep learning approaches like convolutional neural networks (CNNs).
+As depth data become more abundant, we may move from pixel-based learners to deep learning approaches like convolutional neural networks.
 Bakkestuen et al.
-\citeyearpar{bakkestuenDelineationWetlandAreas2023} successfully predicted peatland extent using CNNs, and their ability to automatically learn multiscale spatial features would reduce the need to manually engineer these.
-MRVBF's success as a predictor of peat depth demonstrates that multiscale spatial patterns matter for peat depth and CNNs are designed to learn such patterns.
+\citeyearpar{bakkestuenDelineationWetlandAreas2023} successfully predicted peatland extent using convolutional neural networks, and their ability to automatically learn multiscale spatial features would reduce the need to manually engineer these.
+The success of Multi-Resolution Valley Bottom Flatness as a predictor of peat depth demonstrates that multiscale spatial patterns matter for peat depth and convolutional neural networks are designed to learn such patterns.
 The kind of relationship described in Buffam et al.
-\citeyearpar{buffamFillingHolesRegional2010}, where peat depth in basins related to terrain slope at the basin edge, is also something a CNN could learn.
+\citeyearpar{buffamFillingHolesRegional2010}, where peat depth in basins related to terrain slope at the basin edge, is also something a convolutional neural network could learn.
 However, since this approach is data-hungry, we should build soil knowledge into the DSM where we can \citep{minasnySoilScienceInformedMachine2024}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\appendix
+\appendixtables
+\begin{table}[tbp]
+\caption{Attributes of the radiometric surveys, as reported in Baranwal et al., \citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013} and Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.}
+\begin{tabular}{lll}
+\hline
+                                         & Skrimfjella & Ørskogfjellet              \\ \cline{2-3} 
+Survey period                            & 2008-2011   & December 2014–January 2015 \\
+Average flight altitude (m)              & 75          & 80                         \\
+Average flight speed (\unit{km\,h^{-1}}) & 108         & 88                         \\
+Flight line spacing (m)                  & 200         & 200                        \\ \hline
+\end{tabular}
+\label{tab:radSurveys}
+\end{table}
+\clearpage
+\appendixfigures
+\begin{figure}
+\centering
+\includegraphics[width=0.8\textwidth]{figures/GPRwavevelocity.pdf}
+\caption{\label{fig:GPRwavevelocity}Calibration of GPR wave velocity at Skrimfjella (a) and Ørskogfjellet (b) by regression of probed depth against wave travel time.}
+\end{figure}
+\clearpage
+\appendixfigures
+\begin{figure}
+\centering
+\includegraphics{figures/semivariograms.pdf}
+\caption{\label{fig:semivariograms}Empirical semivariograms of peat depth point measurements at Skrimfjella (a) and Ørskogfjellet (b).}
+\end{figure}
+\noappendix
 
@@
 We also thank Vikas Baranwal from the Geological Survey of Norway for helping us access the radiometric data from Skrim.
-This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, © Kartverket, (2) \emph{Norge digitalt} license, Norwegian Institute of Bioeconomy Research (NIBIO), © Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway (NGU).
+This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, © Kartverket, (2) \emph{Norge digitalt} license, Norwegian Institute of Bioeconomy Research (NIBIO), © Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway.
 Large language models have been used during the drafting and editing of this manuscript, with author oversight.

### 2025-01-16 R/Skrim/concatenateDepthMeasurements.R

@@ -27,8 +27,9 @@ modelfit <- data.frame(OWTT = seq(0, max(caldata$OWTT)*1.05, length.out = 100))
 modelfit <- modelfit %>% 
   bind_cols(predict(m0, modelfit, se.fit = TRUE, interval = "confidence")$fit) %>% 
   rename(depth = fit)
+mean(abs(m0$residuals)) #MAE (m)
 
-calplot <- caldata %>% 
+caldata %>% 
   ggplot(aes(x=OWTT, y=depth)) +
   geom_point(pch=16, size=2) +
   geom_ribbon(data=modelfit, aes(ymin=lwr, ymax=upr), alpha=0.4) +
@@ -41,22 +42,41 @@ calplot <- caldata %>%
   coord_cartesian(expand = FALSE) +
   theme_bw()
 
-ggsave('output/Skrim/calibrationGPR.svg', calplot, width = 119, height = 84, units = "mm")
+write_csv(caldata, "output/Skrim/GPRcalibration-caldata.csv")
+write_csv(modelfit, "output/Skrim/GPRcalibration-modelfit.csv")
+
 v <- units::set_units(m0$coefficients[[1]], "m/ns")
 
 ## TWTT to depth ####
 mire1 <- st_read("data/Skrim/GPRpicks.gpkg", "mire1")
 mire2 <- st_read("data/Skrim/GPRpicks.gpkg", "mire2")
 mire3 <- st_read("data/Skrim/GPRpicks.gpkg", "mire3")
-gpr <- bind_rows(mire1, mire2, mire3) |> 
+gpr <- bind_rows(mire1, mire2, mire3, .id = "mire") |> 
   mutate(depth_cm = TWT/2*units::drop_units(v)*100) |> 
-  select(depth_cm)
+  select(mire, Name, depth_cm)
 
 gprlines <- bind_rows(mire1, mire2, mire3) %>% 
   group_by(Name) |> 
   arrange(Trace) |>
   select(Name) |>
   summarize(Name = first(Name), do_union=FALSE) |> 
+  st_cast("LINESTRING")
+gprlines |> 
+  mutate(length = st_length(gprlines)) |> 
+  pull(length) |> 
+  sum()
+
+# Distance threshold to divide trace sequences into separate lines
+threshold_distance <- 10
+buffers_sf <- st_buffer(gpr, dist = threshold_distance/2) |> 
+  st_union() |> 
+  st_cast("POLYGON") |> 
[…]

### 2025-01-16 R/Skrim/exploratoryDataAnalysis.R

@@ -17,6 +17,17 @@ extract(predictors, sa, ID = FALSE) |>
 depthpts <- read_csv("data/Skrim/depth_all.csv") |> 
   st_as_sf(coords = c('X', 'Y'), crs = 25833)
 
+# From all data
+depth_all <- depthpts |> 
+  as("Spatial")  
+emp_variog <- variogram(depth_cm ~ 1, cutoff = 200, data = depth_all)
+print(emp_variog)
+plot(emp_variog)
+emp_variog |> 
+  as_tibble() |> 
+  select(np, dist, gamma) |>
+  write_csv("output/Skrim/variogram-all-cutoff200.csv")
+
 # From probe data only
 depth_probe <- depthpts |> 
   filter(source == "probe") |> 

### 2025-01-16 R/concatenateDepthMeasurements.R

@@ -114,8 +114,9 @@ modelfit <- data.frame(OWTT = seq(0, max(caldata$OWTT)*1.05, length.out = 100))
 modelfit <- modelfit %>% 
   bind_cols(predict(m0, modelfit, se.fit = TRUE, interval = "confidence")$fit) %>% 
   rename(depth_m = fit)
+mean(abs(m0$residuals)) #MAE (m)
 
-calplot <- caldata %>% 
+caldata %>% 
   ggplot(aes(x=OWTT, y=depth_m)) +
   geom_point(aes(alpha = xydistance), pch=16, size=2) +
   geom_ribbon(data=modelfit, aes(ymin=lwr, ymax=upr), alpha=0.4) +
@@ -131,7 +132,9 @@ calplot <- caldata %>%
   theme(legend.position = c(0.3, 0.8),
         legend.background = element_blank())
 
-ggsave('output/calibrationGPR.svg', calplot, width = 119, height = 84, units = "mm")
+write_csv(caldata, "output/GPRcalibration-caldata.csv")
+write_csv(modelfit, "output/GPRcalibration-modelfit.csv")
+
 v <- units::set_units(m0$coefficients[[1]], "m/ns")
 
 ### TWTT to depth ####

### 2025-01-16 R/exploratoryDataAnalysis.R

@@ -16,6 +16,17 @@ extract(predictors, sa, ID = FALSE) |>
 depthpts <- read_csv("data/depth/all.csv") |> 
   st_as_sf(coords = c('X', 'Y'), crs = 25832)
 
+# From all data
+depth_all <- depthpts |> 
+  as("Spatial")  
+emp_variog <- variogram(depth_cm ~ 1, cutoff = 200, data = depth_all)
+print(emp_variog)
+plot(emp_variog)
+emp_variog |> 
+  as_tibble() |> 
+  select(np, dist, gamma) |>
+  write_csv("output/variogram-all-cutoff200.csv")
+
 # From probe data only
 depth_probe <- depthpts |> 
   filter(source == "probe") |> 

### 2025-01-16 ms/figures.R

@@ -190,7 +190,7 @@ ggsave(g123, filename = 'sites-patchwork.pdf', path = "ms/figures",
 # ggsave(filename = 'sites-patchwork.svg', path = "ms/figures",
 #        width =210-30, height = 240-40, units = 'mm')
 
-# Model metrics ####
+# Model metrics by faceting ####
 
 library(tidyverse)
 
@@ -201,13 +201,13 @@ plotting <- bind_rows(orskog = orskog, skrim = skrim, .id = 'site') %>%
   mutate(
     site = fct_relevel(site, "orskog"),
     metric = case_when(
-      .metric == "ccc" ~ "Concordance \ncorrelation coefficient",
-      .metric == "rsq" ~ "R-squared",
-      .metric == "mae" ~ "Mean \nabsolute error"),
+      .metric == "ccc" ~ "unitless\nConcordance correlation",
+      .metric == "rsq" ~ "unitless\nR-squared",
+      .metric == "mae" ~ "cm\nMean absolute error"),
     metric = fct_relevel(metric,
-                         "Concordance \ncorrelation coefficient",
-                         "R-squared",
-                         "Mean \nabsolute error"),
+                         "unitless\nConcordance correlation",
+                         "unitless\nR-squared",
+                         "cm\nMean absolute error"),
     model = case_when(
       model == "DMK" ~ "DMK class (2)",
       model == "Terrain" ~ "terrain (21)",
@@ -234,7 +234,8 @@ g1 <- ggplot(plotting) +
   geom_text(aes(y = model, x = mean, label = signif(mean, 2), group = site),
             position = position_dodge2(width= 1, reverse = TRUE),
             color = "grey50", size = 2.5) +
-  facet_wrap(~metric, nrow = 1, scales = "free_x", axes ="margins") +
+  facet_wrap(~metric, nrow = 1, scales = "free_x", axes ="margins",
+             strip.position = "bottom") +
   scale_color_discrete(labels = c(orskog = "\u00D8rskogfjellet", 
                                   skrim = "Skrimfjella")) +
   theme_bw() +
@@ -243,13 +244,114 @@ g1 <- ggplot(plotting) +
         panel.grid = element_blank(),
         axis.ticks.y = element_blank(),
         legend.position = "inside",
[…]

### 2025-02-28 ms/ms.Rmd
addressed coauthor comments

@@
   Accurate mapping of peat depth is crucial for carbon accounting and land management, yet existing maps lack the resolution and accuracy needed for these applications. 
-  This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in Norway. 
+  This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in western and southeastern Norway. 
   Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data across two, >10 km^2^ sites. 
   We measured peat depth by probing and ground-penetrating radar at 372 and 1878 locations at the two sites, respectively.
-  Then we trained random forest models using radiometric and terrain variables, plus the national map of peat depth, to predict peat depth at 10 m resolution. 
+  Then we trained Random Forest models using radiometric and terrain variables, plus the national map of peat depth, to predict peat depth at 10 m resolution. 
   The two best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth. 
@@
   The accuracy of the national map of peat depth did not measure up to any of our remote sensing models, even though it was calibrated to the same data.
-  Still, weak relationships with remotely sensed variables make peat depth hard to predict overall. 
-  Based on these findings, digital soil mapping can improve existing, broad-scale maps of peat depth, but local carbon stock assessment requires additional field sampling. 
-  Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat occurrence and depth.
+  Still, weak relationships with remotely sensed variables made peat depth hard to predict overall.
+  Based on these findings, we conclude that digital soil mapping can improve existing, broad-scale maps of peat depth in Norway, but highly localized carbon stock assessments are best made from field measurements.
+  Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat lateral extent and depth.
+  Together, these pathways promise more accurate landscape-scale carbon stock assessments and better-informed land management policies.
 bibliography: ms.bib
@@
   codedata: |
-    R code used in this study is available at <https://github.com/julienvollering/DSMdepth>. Depth measurements will be archived in a public repository upon publication.
+    R code used in this study is available at <https://github.com/julienvollering/DSMdepth>.
+The depth measurements we used are archived at \doi{10.6073/pasta/6ce440152f693f2156bf5b692a2e7917} and follow data and metadata standards.
 authorcontribution: |
@@
 Peat soils are a terrestrial carbon stock of global importance. 
-They store about 30 % of global soil carbon despite covering only 2--3 % of Earth's land [@xuPEATMAPRefiningEstimates2018; @friedlingsteinGlobalCarbonBudget2020; @unepGlobalPeatlandsAssessment2022]. 
+They store 450-650 Gt of carbon, or about 30 % of global soil carbon, despite covering only 2--3 % of Earth's land [@xuPEATMAPRefiningEstimates2018; @friedlingsteinGlobalCarbonBudget2020; @unepGlobalPeatlandsAssessment2022].
 Peatlands (areas with peat soil) are more carbon dense per square meter than any other ecosystem [@temminkRecoveringWetlandBiogeomorphic2022]. 
@@
 One of the keys to the areal density of peatland carbon stocks lies in the third dimension: their depth. 
-Peat soils range from zero to over ten meters deep, so deep peats comprise a large volume. 
+Peat soils range from zero to over ten meters deep [@widyastutiPEATGRIDSMappingThickness2024], so deep peats comprise a large volume. 
 Their depth results from the accumulation of organic matter over thousands of years [@loiselDatabaseSynthesisNorthern2014; @joostenRolePeatlandsClimate2016]. 
 In the anoxic and acidic conditions created by a high water table, plant material decay is slightly outweighed by new growth, and the surplus carbon is laid down as peat. 
-There it remains sequestered as long as the water table stays high and prevents oxidative peat decomposistion.
+There it remains sequestered as long as the water table stays high and prevents oxidative peat decomposition.
 
-Peatlands are unevenly distributed globally, and in regions with high cover they are frequently converted to human land use [@unepGlobalPeatlandsAssessment2022].
+Peatlands are most common at high latitudes, and in regions with high cover they are frequently converted to human land use [@unepGlobalPeatlandsAssessment2022].
 They are attractive for agriculture and forestry because they are flat, treeless, and have developed soils.
 However, other land uses also displace peatlands. 
-In Norway -- where flat lowland is relatively scarce but 9 % of the land area is peatland [@brynLandCoverNorway2018] -- lawmakers have restricted peatland afforestation and cultivation in recent decades.
+In Norway -- where flat lowland is relatively scarce but upward of 9 % of the land area is peatland [@brynLandCoverNorway2018; @bakkestuenDelineationWetlandAreas2023] -- lawmakers have restricted peatland afforestation and cultivation in recent decades.
 Since then, a larger proportion of peatland loss is driven by construction [@flagetMyrMellomLokal2024].
@@
 Measuring peat depth on the ground by probing or ground-penetrating radar is straightforward, and a field survey can map a small area at low cost. 
-However, surveying large areas is impractical when depth varies over short distances (e.g. < 1 m) -- as in most peatlands. 
+However, surveying large areas is impractical when depth varies widely over short distances (e.g.
+10 m) -- as in many peatlands [@torppaStatisticalApproachDetermining2011; @proulx-mcinnisReviewGroundpenetratingRadar2013; @henrionMappingMonitoringPeatlands2024].
 A complementary approach from soil science is digital soil mapping (DSM). 
@@
 
-The crux for DSM of peat depth is the relationship between peat depth and each of the other, mapped variables. 
+The crux for DSM of peat depth is the relationship between peat depth and each of the other, mapped predictors. 
 For DSM to work, these relationships must be strong and consistent over the area of interest.
@@
 Moreover, mechanisms of peat formation are linked to topography. 
-For example, a steep slope is unlikely to have a high water table, thereby limiting accumulation.
+For example, a steep slope is unlikely to have a high water table, thereby limiting accumulation of peat.
 
@@
 @rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024].
-Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic variables and the resolution of the peat depth analysis.
+Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic predictors and the resolution of the peat depth analysis.
 
-Another set of variables related to peat depth are measurements of natural radioactivity from the ground surface. 
+Another set of predictors related to peat depth are measurements of natural radioactivity from the ground surface. 
 Gamma-ray spectrometry can survey the activity (decay counts per second) from radioactive isotopes in the earth's crust: potassium-40, uranium-238, and thorium-232 [@reinhardtGammaraySpectrometryVersatile2019].
@@
 Norway has 22 peatland mesotopes, with different hydrology, formation, and development -- from topogeneous or soligeneous fens to blanket or raised bogs [@joostenWiseUseMires2002; @lyngstadBeskrivelserAvTorvmassivenheter2023].
-These mesotopes have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric variables will vary by landscape.
+These mesotopes have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric predictors will vary by landscape.
 
@@
 
-The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify peatlands with rich carbon stocks to avoid their conversion and prioritize their preservation or restoration.
+The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify peatlands with rich carbon stocks to avoid their conversion and prioritize their preservation or restoration [@strackPotentialPeatlandsNatureBased2022].
 Although ground surveys of peat depth are accurate and feasible across small areas, having landscape-scale maps *before* detailed investigation increases the option space for spatial planning and land management.
@@
 
-Here, we assess how well remotely sensed topographic and radiometric data can predict peat depth at the landscape scale, with a view towards revising regional and national maps in Norway. 
-Specifically, we: (1) quantify the accuracy of predictions from topographic and radiometric variables, (2) identify key predictive variables, and (3) compare these relationships across two different Norwegian landscapes.
+Here, we assess how well remotely sensed topographic and radiometric data can predict peat depth at two contrasting sites, with a view towards revising regional and national maps in Norway. 
+Specifically, we aim to: (1) quantify the accuracy of predictions from topographic and radiometric variables, and (2) identify key predictive variables.
+Reflecting on our results and the need for continued improvement in peat depth maps, we close by discussing implications for digital soil mapping of peat depth.
+
 To our knowledge, this is the first study to predict peat depth from airborne radiometric data using machine learning algorithms. 
@@
 At Skrimfjella we delineated a study area of 34 km^2^ based on radiometric coverage and accessibility (Fig. \@ref(fig:sites)b).
-The study area has a diverse bedrock, with 32 % alkali feldspar granite, 26 % marlstone, 10 % granite, 8 % mozonite, 7 % sandstone, 6 % limestone, and five other rock types with 1--5 % coverage (Geological Survey of Norway, 1:250 000 dataset).
+The study area has a diverse bedrock, with 32 % alkali feldspar granite, 26 % marlstone, 10 % granite, 8 % monzonite, 7 % sandstone, 6 % limestone, and five other rock types with 1--5 % coverage (Geological Survey of Norway, 1:250 000 dataset).
 The area within our delineation at Skrimfjella is classified as the landscape type *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
-It is almost without human infrastructure, dominated by forest, and borders on a large nature reserve.
+It is almost without human infrastructure, dominated by conifer forest, and borders on a nature reserve.
 The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).
@@
 We created the same suite of peat depth predictors for both sites (25 continuous and 1 categorical; Table \@ref(tab:preds)).
+Each continuous predictor equates to one variable in the model, while the categorical predictor equates to two variables in the model because its three levels become two indicator variables (one level is used as the reference).
 All continuous predictors were derived either from an airborne radiometric survey or from a DTM.
-From the radiometric surveys we simply used the four variables produced by the Geological Survey of Norway: ground concentration of Potassium, Thorium, Uranium, as well as total count.
+From the radiometric surveys, we simply used the four variables produced by the Geological Survey of Norway: ground concentration of potassium, thorium, uranium, as well as total count.
 From the DTMs we calculated several land surface parameters, ranging from simple terrain indices to more complex geomorphometric and hydrological variables [@maxwellLandsurfaceParametersSpatial2022].
@@
 We chose a spatial resolution of 10 m for our predictors, peat depth sample, and modelling.
-We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other.
+We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other hand.
 Complete descriptions of predictors follow below.
 
-```{r preds, echo=FALSE, message=FALSE}
-df <- readr::read_csv("tables/predictors.csv")
-options(knitr.kable.NA = '')
-knitr::kable(df, booktabs = FALSE, vline = "", linesep = "",
-             col.names = c("Group", "Name", "Description"),
-             caption = "Candidate predictors of peat depth.")
+```{r predictors, echo=FALSE, results='asis'}
+predictors <- readr::read_file("tables/predictors.tex")
+cat(predictors)
 ```
@@
 [-@baranwalHelicopterborneMagneticElectromagnetic2013] and Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015].
-They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling, using the *terra* package in R.
+They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling, using the *terra* package (v1.7) in R.
 
 The radiometric surveys were conduced with similar flight parameters (Table \ref{tab:radSurveys}).
-Spectrometer count rates were calibrated to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
+Spectrometer count rates were calibrated to known concentrations of potassium, thorium, and uranium in mobile pads.
 Data were processed following standard procedures outlined by the International Atomic Energy Association.
@@
 
-We used the *terra* R package to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
+We used the *terra* package to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
 These were derived at two scales to produce eight different predictors; we either calculated the indices at 1 m DTM resolution and then aggregated to 10 m resolution, or aggregated to 10 m DTM resolution and then calculated the indices.
@@
 
-The Depth-to-Water index [@murphyMappingWetlandsComparison2007] approximates a location's vertical height above the surface water feature that it is likely to drain towards.
+The Depth-to-Water index [@murphyMappingWetlandsComparison2007] approximates a location's vertical height above the surface water feature (e.g.
+stream, lake, or sea) that it is likely to drain towards.
 It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature [eq. 5 in @murphyTopographicModellingSoil2009].
@@
 Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell [@schonauerSpatiotemporalPredictionSoil2021; @schonauerRcodeCalculatingDepthwater2021].
-This catchments area layer was then thresholded at seven different levels (flow initiation area 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry [@murphyModellingMappingTopographic2011; @agrenEvaluatingDigitalTerrain2014; @schonauerSpatiotemporalPredictionSoil2021].
+This catchment area layer was then thresholded at seven different levels (flow initiation area 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry [@murphyModellingMappingTopographic2011; @agrenEvaluatingDigitalTerrain2014; @schonauerSpatiotemporalPredictionSoil2021].
 In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
@@
 
-At Skrimfjella, we used the *eSample* function in the *iSDM* R package (v.1.0) to stratify our sample across elevation, slope, and potassium concentration.
-This function defines the environmental space as a two-dimensional convex hull around the ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest [@hattabUnifiedFrameworkModel2017].
+At Skrimfjella, we used the *eSample* function in the *iSDM* R package (v1.0) to stratify our sample across elevation, slope, and potassium concentration.
+This function defines the environmental space as a two-dimensional convex hull around the PCA-ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest [@hattabUnifiedFrameworkModel2017].
 We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and *eSample* returned 105 raster cells.
@@
 We also measured peatland occurrence (binary variable) in a separate sample at Skrimfjella.
-We measured peatland occurrence because the AR5 dataset is known to underestimate peatland coverage [especially in forests, @brynLandCoverNorway2018], and because airborne radiometrics may help identify unmapped peatland [@gatisMappingUplandPeat2019; @olearyDigitalSoilMapping2022].
+We measured peatland occurrence because the AR5 dataset is known to underestimate peatland extent [especially in forests, @brynLandCoverNorway2018], and because airborne radiometrics may help identify unmapped peatland [@gatisMappingUplandPeat2019; @olearyDigitalSoilMapping2022].
 The 10 m cells were sampled from those outside AR5 mire and with slope < 20°. 
@@
 For a sequence of sample sizes (50--500) [ten replicates each, drawn by conditioned latin hypercube sampling, @minasnyConditionedLatinHypercube2006; @roudierClhsPackageConditioned2011], we calculated the mean Kullback--Leibler divergence between sample and population distributions [@maloneMethodsImproveUtility2019; @sauretteDivergenceMetricsDetermining2023].
-Then, we fitted a asymptotic regression of mean divergence on sample size, and found that the curve reached 95 % of the fitted asymptote at a sample size of 160.
+Then we fitted an asymptotic regression of mean divergence on sample size, and found that the curve reached 95 % of the fitted asymptote at a sample size of 160.
 
-To choose 160 locations, we performed feature space coverage sampling, implemented using the *kmeans* function in base R and the *rdist* function in the the *fields* package (v.14.2).
+To choose 160 locations, we performed feature space coverage sampling, implemented using the *kmeans* function in base R and the *rdist* function in the the *fields* package (v14.2).
 Feature space coverage sampling chooses locations that are closest to cluster centers in standardized predictor space [@brusSamplingDigitalSoil2019].
@@
 After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
-Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 % (with respect to the ideal sample), but distance in our sample was still only 46 % of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
+Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 % (compared to the ideal sample), but distance in our sample was still only 46 % of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
 
@@
 
-We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real time kinematic (Ørskogfjellet) global navigation satellite system receivers.
+We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real time kinematic (Ørskogfjellet) global navigation satellite system (GNSS) receivers.
 We dampened the effect of outlying measurements by probing three times at each location [@parryEvaluatingApproachesEstimating2014], at the vertices of a centered triangle with 2 m (Skrimfjella) or 4.5 m (Ørskogfjellet) sides.
@@
 
-We also extracted a set of existing depth probe measurements for Ørskogfjellet from a paper map made by the Norwegian Soil and Mire Company in 1984.
+We also extracted a set of existing depth measurements for Ørskogfjellet from a paper map made by the Norwegian Soil and Mire Company in 1984.
 The map presents 44 borehole depths (in decimeters) across a 9 ha peatland area.
@@
 We judged whether the soil was a peat soil based on its density, texture, and color, as well as the presence or absence of mire vegetation.
-Although peat soil is strictly defined by organic content (which we did not analyze), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland [@minasnyMappingMonitoringPeatland2023].
+Although peat soil is strictly defined by organic content (which we did not analyze), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland [@minasnyMappingMonitoringPeatland2024].
 
@@
 
-We performed GPR surveys in three subjectively chosen mires at Skrimfjella and in areas with a high density of sampled raster cells at Ørskogfjellet. 
+We performed GPR surveys in three subjectively chosen peatlands at Skrimfjella and in areas with a high density of sampled raster cells at Ørskogfjellet. 
 We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with a GNSS-enabled control unit connected to a 500 MHz shielded antenna mounted in a plastic sledge (transmitter--receiver separation 0.18 m, trace frequency 10 Hz).
@@
 Finally, we used these two wave velocities to convert the travel times of all picks to calibrated peat depths.
-In total the GPR surveys produced 48579 point measurements of peat depth at Skrimfjella and 32653 at Ørskogfjellet.
+In total, the GPR surveys produced 48579 point measurements of peat depth at Skrimfjella and 32653 at Ørskogfjellet.
 
@@
 
-We used random forests (RF) to predict peat depth at both sites.
+We used Random Forests (RF) to predict peat depth at both sites.
 RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees [@breimanRandomForests2001].
 We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in DSM applications than other algorithms [@beguinPredictingSoilProperties2017; @nussbaumEvaluationDigitalSoil2018; @lamichhaneDigitalSoilMapping2019].
-It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance [@khaledianSelectingAppropriateMachine2020].
+It is suited for use on relatively small training datasets and its predictions can be interrogated to learn about predictor importance [@khaledianSelectingAppropriateMachine2020].
 Evaluating variable importance in a maximally predictive model aligns with the aim of this study.
@@
 If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.
-We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that (non-spatial) RF rather regression kriging was justified at both sites.
+We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that (non-spatial) RF rather than regression kriging was justified at both sites.
 
-We implemented models in the *tidymodels* framework in R [@kuhnTidymodelsCollectionPackages2020], with the *ranger* R package for RFs [v.0.16, @wrightRangerFastImplementation2017].
+We implemented models in the *tidymodels* framework in R [@kuhnTidymodelsCollectionPackages2020], with the *ranger* R package for RFs [v0.16, @wrightRangerFastImplementation2017].
 RFs were fit with 1000 trees, minimum node size of 5, and the number of predictors randomly sampled at each split was the square root of the total number of predictors (*ranger* default).
@@
 
-For both sites we compared the performance of models with five different configurations of predictors. 
+For both sites, we compared the performance of models with five different configurations of predictors. 
 Specifically, we trained models with:
 
-  1. only DMK peat depth class (2 indicator variables from 3 levels)
-  2. only terrain (21 continuous variables)
-  3. terrain and DMK peat depth class (23 variables)
-  4. terrain and radiometric predictors (25 continuous variables)
-  5. all predictors (27 variables)
+  1. *DMK* (2 variables)
+  2. *terrain* (21 variables)
+  3. *terrain + DMK* (23 variables)
+  4. *terrain + radiometric* (25 variables)
+  5. *all predictors* (27 variables)
 
@@
 For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation [@wadouxSpatialCrossvalidationNot2021].
-We implemented the kNNDM with the *CAST* R package [v.1.0.2, @meyerCASTPackageTraining2024], setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
+We implemented the kNNDM with the *CAST* R package [v1.0, @meyerCASTPackageTraining2024], setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
 From the cross-validation we quantified mean absolute error (error magnitude, original scale), R^2^ (explained variation, standardized scale), and Lin's concordance correlation coefficient (error magnitude and explained variation, standardized scale).
@@
 The quantile regression forests were trained with predictor configuration that showed the highest performance at each site (under the assumption that these models would be put into production) and we extracted 90 % prediction intervals.
+
+#### Peatland extent
 
@@
 This analysis treats the model's prediction of peat depth as a rank index of peatland likelihood.
-Since the purpose was to test the models' ability to uncover unmapped peatland (where DMK peat depth class is undefined), we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.
+Since the purpose was to test the models' ability to uncover unmapped peatland (where DMK peat depth class is undefined), we evaluated models with two predictor configurations: *terrain* or *terrain + radiometric*.
 
@@
 
-We quantified global variable importance (predictor influence across locations) and examined partial dependence plots (curves of fitted relationships) for the best-performing predictor configuration at each site.
+We quantified global variable importance (predictor influence across all locations) and examined partial dependence plots (curves of fitted relationships) for the best-performing predictor configuration at each site.
 Both are useful for understanding the mechanisms behind the model's predictions and the roles of the predictors in the model.
 
-For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration -- because correlation between predictors degrades variable importance measures [@stroblConditionalVariableImportance2008; @biauRandomForestGuided2016] and can produce misleading visualizations of predictor--outcome relationships [@biecekExplanatoryModelAnalysis2021; @dwivediExplainableAIXAI2023].
-Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation above 0.7.
-Thus, highly correlated sets of predictors are represented by a single variable for the purposes of model interpretation.
+For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration -- because correlation between predictors degrades variable importance metrics [@stroblConditionalVariableImportance2008; @biauRandomForestGuided2016] and can produce misleading visualizations of predictor--outcome relationships [@biecekExplanatoryModelAnalysis2021; @dwivediExplainableAIXAI2023].
+Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation coefficient above 0.7.
+Thus, highly correlated sets of variables are represented by a single variable for the purposes of model interpretation.
 
-We calculated variable importance with the *vip* R package (v.0.4.1), by three different methods: *FIRM*, *permutation*, and *Shapley* [@greenwellVariableImportancePlots2020].
+We calculated variable importance with the *vip* R package (v0.4), by three different methods: *FIRM*, *permutation*, and *Shapley* [@greenwellVariableImportancePlots2020].
 *FIRM* values measure the flatness of the partial dependence plot, *permutation* values measure the decrease in model performance when the predictor is permuted, and *Shapley* values are aggregated from local, game-theoretical measures of variable importance [@greenwellVariableImportancePlots2020].
@@
 
-We calculated partial dependence with the *pdp* R package [v.0.8.1, @greenwellPdpPackageConstructing2017].
+We calculated partial dependence with the *pdp* R package [v0.8, @greenwellPdpPackageConstructing2017].
 For the six most important variables, we plotted both partial dependence and individual conditional expectation, to show the average effect of the predictor on the outcome and the variation in the effect across observations, respectively [@goldsteinPeekingBlackBox2015].
@@
 
+## Model performance
+
 None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \@ref(fig:modelMetrics)).
-For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an R^2^ of 0.34, and a mean absolute error of 60 cm.
-For Ørskogfjellet the concordance correlation coefficient was 0.39, R^2^ was 0.33, and mean absolute error was 56 cm, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
+For Skrimfjella, the best model achieved a concordance correlation coefficient of 0.30, an R^2^ of 0.34, and a mean absolute error of 60 cm.
+For Ørskogfjellet, the concordance correlation coefficient was 0.39, R^2^ was 0.33, and mean absolute error was 56 cm, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
 These values were derived from kNNDM spatial cross-validation with 20 folds at Skrimfjella and 10 folds at Ørskogfjellet.
@@
 Although the difficulty of predicting peat depth caused prediction intervals to be wide, these uncertainty estimates were well calibrated.
-At Skrimfjella the prediction interval coverage probability was 91 %, and at Ørskogfjellet it was 84 % (both compared to the target value of 90 %).
+At Skrimfjella, the prediction interval coverage probability was 91 %, and at Ørskogfjellet it was 84 % (both compared to the target value of 90 %).
 Observations outside of the prediction intervals were evenly distributed across each of the study areas.
 
-For Skrimfjella, the best predictor configuration was *all predictors*, followed closely by *terrain and radiometric*.
-The performance gap between the *terrain and DMK* configuration and the *terrain-only* configuration was similarly small.
-Compared to the *terrain-only* configuration, the *terrain and radiometric* configuration improved concordance correlation by 0.04, R^2^ by 0.04, and mean absolute error by 1 cm.
-*DMK class* alone -- although calibrated to measured depths -- was a very poor predictor of peat depth, with a concordance correlation coefficient of 0.0077.
+For Skrimfjella, the best predictor configuration was *all predictors*, followed closely by *terrain + radiometric*.
+The performance gap between the *terrain + DMK* configuration and the *terrain* configuration was similarly small.
+Compared to the *terrain* configuration, the *terrain + radiometric* configuration improved concordance correlation by 0.04, R^2^ by 0.04, and mean absolute error by 1 cm.
+*DMK* was a very poor predictor of peat depth even though it was calibrated to measured depths, with a concordance correlation coefficient of 0.0077.
 
-For Ørskogfjellet, the best predictor configuration was *terrain and DMK*, followed by *terrain-only*.
+For Ørskogfjellet, the best predictor configuration was *terrain + DMK*, followed by *terrain*.
 Adding radiometric predictors to these configurations worsened model performance, especially in terms of concordance correlation and R^2^.
@@
 
+### Peatland extent
+
 When we tested how well models extrapolated from within to outside of AR5 mire areas, we found that the models produced worse mean absolute error than just assuming a constant 30 cm depth (Fig.
 \@ref(fig:modelMetricsExtrapolation)).
-With the independent occurrence data at Skrimfjella, we found that neither the *terrain-only* nor *terrain and radiometric* configurations were able to discriminate between peat presence and absence (area under the curve of the receiver operating characteristic 0.44 and 0.52 respectively, where 0.5 indicates random guessing).
+With the independent occurrence data at Skrimfjella, we found that neither the *terrain* nor *terrain + radiometric* configurations were able to discriminate between peat presence and absence (area under the curve of the receiver operating characteristic 0.44 and 0.52 respectively, where 0.5 indicates random guessing).
 
@@
 
-For the purpose of model interpretation, the *all predictors* configuration for Skrimfjella was reduced from 27 variables to 11 non-collinear variables.
-Similarly, the *terrain and DMK* configuration for Ørskogfjellet was reduced from 23 variables to 11 non-collinear variables.
+## Model interpretation
+
+For the purpose of model interpretation, the *all predictors* configuration for Skrimfjella was reduced from 27 variables to 11 non-collinear variables, by removing one of the variables in each highly-correlated pair.
+Similarly, the *terrain + DMK* configuration for Ørskogfjellet was reduced from 23 variables to 11 non-collinear variables.
+
+### Variable importance
+
 At both sites, elevation and Multi-Resolution Valley Bottom Flatness were important predictors (Fig. \@ref(fig:varImp)).
@@
 At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices slope, Terrain Ruggedness Index, Topographic Position Index, and roughness.
-The radiometric variable radU showed moderate importance at Skrimfjella.
+The radiometric predictor uranium ground concentration showed moderate importance at Skrimfjella.
 
-```{r varImp, echo = FALSE, fig.cap = "Global variable importance in the best-performing models at Ørskogfjellet (a) and Skrimfjella (b), as measured by three different metrics.
+```{r varImp, echo = FALSE, fig.cap = "Global variable importance in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), as measured by three different metrics: Shapley values, permutation importance, and the Feature Importance Ranking Measure [@greenwellVariableImportancePlots2020].
 Variables removed due to collinearity are shown to the right of the variable with which they are most correlated."}
@@
 
-Many of the the most important predictors in the best performing models showed non-monotonic effects on peat depth (Fig. \@ref(fig:pdps)).
+### Partial dependence
+
+Many of the most important predictors in the best performing models showed non-monotonic effects on peat depth (Fig. \@ref(fig:pdps)).
 At Ørskogfjellet for example, increasing elevation was predictive of deeper peat up to about 75 m above sea level, after which a further increase was predictive of shallower peat.
 At Skrimfjella the partial dependence on elevation had the opposite shape but covers a higher elevation range, with the shallowest peats predicted around 350 m above sea level.
-TWI50m at Ørskogfjellet and DTW4000 and radU at Skrimfjella were other predictors that showed considerable fluctuations in their predictive effects on peat depth.
-The radiometric predictor in particular displayed an idiosyncratic effect, with a marked dip in predicted depth at intermediate values of radU.
+TWI50m at Ørskogfjellet and DTW4000 and uranium ground concentration at Skrimfjella were other predictors that showed considerable fluctuations in their predictive effects on peat depth.
+The radiometric predictor in particular displayed an idiosyncratic effect, with a marked dip in predicted depth at intermediate values of uranium ground concentration.
 On the other hand, the partial effects of some important predictors were more straightforward.
@@
 For example, the magnitude of the increase in depth with elevation that the model expected at Ørskogfjellet was different for different observations; some depth predictions increased by only 40 cm while others increased by more than 100 cm, over the same elevation gain.
-Similarly, individual conditional expectation lines of radU at Skrimfjella were non-parallel, with some locations showing monotonically increasing peat depth predictions with uranium concentration (unlike the average effect).
+Similarly, individual conditional expectation lines of uranium ground concentration at Skrimfjella were non-parallel, with some locations showing monotonically increasing peat depth predictions with uranium concentration (unlike the average effect).
 Nevertheless, most individual conditional expectation lines were generally parallel -- indicating that the average effects of the predictors were good representations of their overall effects.
 
-```{r pdps, echo = FALSE, fig.cap = "Partial dependence plots of the six most important variables in the best-performing models at Ørskogfjellet (a) and Skrimfjella (b).
-The average effect of the predictor on the outcome (red line) overlays the variation in the effect across observations (grey lines)."}
+```{r pdps, echo = FALSE, fig.cap = "Partial dependence plots of the six most important variables in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b).
+The average effect of the predictor on the outcome (red line) overlays the variation in the effect across observations (grey lines).
+The variables' training data distributions are indicated with rug plots along the x-axes."}
 knitr::include_graphics("figures/partial_dependence.pdf")
@@
 Although model performance was limited, we improved on the best available map of peat depth (DMK depth class), which is based on field measurements only.
-This highlights the general value of remotely sensed data, whose complete coverage can improve maps even when their association with the variable of interest is weak.
+This highlights the general value of remotely sensed data, whose complete coverage can improve maps even when their association with the variable of interest is weak [@mulderUseRemoteSensing2011].
 Since remotely sensed data are widely available, improvements to soil maps as shown here are low hanging fruit.
-This point is recognized and reflected in the rise of DSM.
+This point is recognized and reflected in the rise of DSM [@minasnyDigitalMappingPeatlands2019].
 
@@
 
-Gatis et al. [-@gatisMappingUplandPeat2019] used similar predictors and the same spatial grain, finding a much stronger correlation between predicted and observed peat depth ($R^2 = 0.68$). 
+Gatis et al. [-@gatisMappingUplandPeat2019] used similar predictors and the same spatial grain, finding a much stronger relationship between predicted and observed peat depth ($R^2 = 0.68$). 
 Although their random evaluation data partition could make performance estimates too optimistic [@robertsCrossvalidationStrategiesData2017; @wadouxSpatialCrossvalidationNot2021], the confounding effect of spatial structure is probably small because they used linear regression and few predictors.
@@
 They study a flatter area with a higher proportion of peatland cover, and their peatland is primarily blanket bog.
+The peat surface of a blanket bog is tied more closely to its underlying topography than the peat surface of raised bogs or fens [@lindsayPeatlandMireTypes2016], which may make depths in blanket bogs easier to predict.
 The predominance of fens and smaller peatland extent may have contributed to worse performance in our study.
@@
 [-@gatisMappingUplandPeat2019] at 100 m resolution, using splines to relax linearity between radiometry/terrain and peat depth.
-He found that radK alone predicted peat depth with much higher concordance than our models (concordance correlation coefficient = 0.76) and that elevation alone produced comparable performance (concordance correlation coefficient = 0.27).
+He found that potassium ground concentration alone predicted peat depth with much higher concordance than our models (concordance correlation coefficient = 0.76) and that elevation alone produced comparable performance (concordance correlation coefficient = 0.27).
 The root mean square error from these univariate models was 46--68 cm (cf. 78 and 75 cm for Skrimfjella and Ørskogfjellet).
@@
 Our models erred most for the deepest peats (Fig. \@ref(fig:calPlots)).
-Where overprediction occurred, it was smaller. 
+Where overprediction occurred, the error was smaller. 
 This is not unexpected for the right-skewed distributions of peat depth, but it has management implications. 
@@
 
-### Depth predictions do not necessarily extrapolate to peat occurrence
+### Depth predictions do not necessarily extrapolate to peat extent
 
 That we measured more than 30 cm of peat in areas not mapped as mire was expected, because AR5 underrepresents peatland extent by about one third [@brynLandCoverNorway2018]. 
-For that reason, maps of peatland occurrence also need revising. 
+For that reason, maps of peatland extent also need revising. 
 Unfortunately, our models did not predict peat occurrence outside AR5 mires. 
@@
 Peatland extent is probably best mapped using different remotely sensed data than we used here [@bakkestuenDelineationWetlandAreas2023], and this study’s purpose was not to map extent. 
-Nevertheless — as long as the relevant peatland definition includes a depth component — depth predictions (or predictors with subsurface information) should help delineate peatland extent [@olearyDigitalSoilMapping2022; @beamishDetailedMappingPeat2024].
+Nevertheless — as long as the relevant peatland definition includes a depth component — depth predictions (or predictors thereof) should help delineate peatland extent [@olearyDigitalSoilMapping2022; @beamishDetailedMappingPeat2024].
 We return to this point in our discussion of implications for digital soil mapping.
@@
 The bedrock is more homogeneous at Ørskogfjellet than at Skrimfjella, so uneven radiogenesis is not a viable explanation for the differences between sites nor the poor performance in general [@beamishEnvironmentalRadioactivityUK2014; @reinhardtGammaraySpectrometryVersatile2019].
-To the extent that radiometrics had predictive value at Skrimfjella, it appears that they were most valuable near the extremes of the depth distribution, since their inclusion improved R^2^ more than mean absolute error.
+To the degree that radiometrics had predictive value at Skrimfjella, it appears that they were most valuable near the extremes of the depth distribution, since their inclusion improved R^2^ more than mean absolute error.
 All four variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value. 
-This contrasts with Koganti et al. [-@kogantiMappingPeatDepth2023], who found that radTC was a much better predictor than radK.
+This contrasts with Koganti et al. [-@kogantiMappingPeatDepth2023], who found that radiometric total count was a much better predictor than potassium ground concentration.
 
@@
 
-Studies comparing airborne and ground radiometric surveys confirm that they are poorly correlated in low-activity areas like peatlands [@kockComparisonAirborneTerrestrial2011; @karjalainenComparisonTwoGammaray2024].
-Karjalainen et al. [-@karjalainenComparisonTwoGammaray2024] found that ground-based measurements predicted peat depth better than airborne measurements. 
+Studies comparing airborne and ground radiometric surveys confirm that they are poorly correlated in low-activity areas like peatlands [@kockComparisonAirborneTerrestrial2011; @karjalainenComparisonTwoGammaray2025].
+Karjalainen et al. [-@karjalainenComparisonTwoGammaray2025] found that ground-based measurements predicted peat depth better than airborne measurements. 
 Nevertheless, a large radiometric footprint did not prevent strong associations with peat depth in Gatis et al.
-[-@gatisMappingUplandPeat2019] and Marchant [-@marchantUsingRemoteSensors2021], likely due to the extensive blanket bog landscape with more gradual changes in depth [@lindsayBogsEcologyClassification1995; @lyngstadBeskrivelserAvTorvmassivenheter2023].
+[-@gatisMappingUplandPeat2019] and Marchant [-@marchantUsingRemoteSensors2021], perhaps because the extensive blanket bog landscape in these studies has more gradual changes in depth [@lindsayBogsEcologyClassification1995].
 We are unsure whether short-range depth changes explain the weak associations that Siemon et al. [-@siemonAirborneElectromagneticRadiometric2020] found in a large raised bog.
@@
 The RF algorithm’s flexibility means that radiometrics could be used for shallower peats if they provided predictive value for that part of the depth distribution, but there is no evidence of that in our results.
-In the partial dependence plot of radU at Skrimfjella, the expected negative relationship between depth and radU cannot be found by ignoring the left (highly attenuated) side of the distribution. 
+In the partial dependence plot of uranium ground concentration at Skrimfjella, the expected negative relationship between depth and uranium ground concentration cannot be found by ignoring the left (highly attenuated) side of the distribution.
 About a quarter of the peats in our study were less than a meter deep, and full attenuation is unlikely for these [@beamishGammaRayAttenuation2013].
@@
 via climate) to explain the observed patterns.
-Indeed, elevation at Ørskogfjellet seems to interact with other variables (non-parallel individual conditional expectation lines), complicating its interpretation. 
-Moreover, relationships between elevation and peat depth have previously shown opposite shapes in different areas [@finlaysonEstimatingOrganicSurface2021]. 
+For example, the increase in peat depth from 350 to 700 m.a.s.l.
+at Skrimfjella is opposite to the general pattern of deeper peats in lowland than upland Norway [@lyngstadBeskrivelserAvTorvmassivenheter2023].
+Moreover, elevation at Ørskogfjellet seems to interact with other variables (non-parallel individual conditional expectation lines), complicating its interpretation. 
+Relationships between elevation and peat depth have previously shown opposite shapes in different areas [@finlaysonEstimatingOrganicSurface2021]. 
 Nonetheless, a relationship that is not generalizable beyond the mapping area is still useful for DSM, as long as it is evaluated to demonstrate its robustness for the predictive task (e.g.
@@
 
-One interesting feature of the elevational relationships we found may be generalizable: a steep increase in peat depth near the upper marine limit after the last ice age. 
-At Ørskogfjellet, the upper marine limit is about 75 meters above today’s sea level [Geological Survey of Norway, @hogaasDatabaseRegistreringAv2012], where the partial dependence plot of elevation shows a sharp increase in peat depth.
-In areas under the upper marine limit there has been less time for peat accumulation since the ice sheets retreated, and it is plausible that this makes peats there shallower, all else being equal. 
-We cannot evaluate this effect at Skrimfjella, where the upper marine limit is below our study area (at 175 m.a.s.l.).
+One interesting feature of the elevational relationships we found may be generalizable: a steep increase in peat depth near the marine limit after the last ice age. 
+At Ørskogfjellet, the marine limit is about 75 meters above today’s sea level [Geological Survey of Norway, @hogaasDatabaseRegistreringAv2012], where the partial dependence plot of elevation shows a sharp increase in peat depth.
+In areas under the marine limit there has been less time for peat accumulation since the ice sheets retreated, and it is plausible that this makes peats there shallower, all else being equal. 
+This hypothesis is supported by similar findings at a coarser scale in the Hudson Bay Lowlands, where a strong positive relationship between peat depth and distance from the coast can be explained by isostatic uplift and time since peat initiation [@liPeatDepthCarbon2025].
+We cannot evaluate this effect at Skrimfjella, where the marine limit is below our study area (at 175 m.a.s.l.).
 
@@
 We suspect the discrepancy between sites is due to different levels of effort and coverage during the historical surveys; more lowland peatland near agriculture at Ørskogfjellet may have caused more purposeful surveying.
-73 % of cells measured at Skrimfjella had unknown depth in DMK, compared to 20 % at Ørskogfjellet. 
-Still, the 25 % of cells classified as deep at Skrimfjella were not meaningfully deeper than the rest, according to our model. 
-Interactions between DMK and other variables underline the inconsistency of DMK depth maps, even within a site. 
-That a peat at Ørskogfjellet was classified as shallow rather than deep did not uniformly lower the model’s expectation of depth there.
+This is evidenced by the fact that 73 % of the cells measured at Skrimfjella had unknown depth in DMK, compared to 20 % at Ørskogfjellet.
+Our results show that classification as deep peat at Skrimfjella had no predictive value -- these cells were no different than those classified as shallow or unknown, all else being equal. 
+Interactions between DMK and other variables also underline the inconsistency of DMK depth maps, even within a site. 
+For example, that a peat at Ørskogfjellet was classified as shallow rather than deep did not always cause the model to predict shallower peat -- sometimes it increased the predicted depth.
 
@@
 
-The performance gap between the best models and the DMK-only models shows that peat depth in Norway should be mapped digitally. 
+The performance gap between the best models and the DMK models shows that peat depth in Norway should be mapped digitally. 
 Anywhere we have some calibrating measurements, we can get better maps than DMK peat depth classes, at low cost. 
@@
 Fine-scale variation in peat depth may favor mapping at very fine resolution (1 m) — even if land use planning and carbon accounting do not operate at this grain. 
-Terrain--depth relationships might be stronger in 1 m cells than in our 10 m cells, especially considering the hummock--hollow microtopography of many peatlands.
+Terrain--depth relationships might be stronger in 1 m cells than in our 10 m cells, especially considering the hummock--hollow microtopography of many peatlands [@rydin7Mires1999; @lindsayPeatbogsCarbonCritical2010].
 
@@
 A natural starting point is the bounding area around a spatial cluster of depth measurements. 
-For a given cluster, kNNDM can define how far the boundary can extend beyond the point measurements; if a cross-validation nearest neighbor distribution cannot match the sample-to-prediction nearest neighbor distribution, the boundary is too expansive [@meyerMachineLearningbasedGlobal2022; @linnenbrinkKNNDMCVKfold2024].
+For a given cluster, kNNDM can define how far the boundary can extend beyond the point measurements; if the sample-to-prediction nearest neighbor distribution cannot be simulated by any set of cross-validation folds, then the extent is too expansive [@meyerMachineLearningbasedGlobal2022; @linnenbrinkKNNDMCVKfold2024].
 However, it is unclear how big (N) any cluster of depths should be, and the tradeoff between multiple small-extent DSM and fewer large-extent DSM needs research. 
-Bohn and Miller [-@bohnLocallyEnhancedDigital2024] advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or 'supertopes'), where the composition of mesotopes is similar [@moenVegetasjonstyper1998; @joostenWiseUseMires2002].
-Depth varies strongly between peats formed by terrestrialization versus paludification [@buffamFillingHolesRegional2010], and probably along other axes of peatland typology. 
-Although most Norwegian landscapes contain multiple, unknown mesotopes, it makes sense to limit their diversity within a single DSM.
+Bohn and Miller [-@bohnLocallyEnhancedDigital2024] advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or 'supertopes'), where the composition of mesotopes is similar -- e.g.
+regions dominated by raised bogs versus regions dominated by sloping fens [@moenNationalAtlasNorway1999; @joostenWiseUseMires2002].
+Depth varies systematically between bogs and fens [@lindsayPeatlandMireTypes2016], between peats formed by terrestrialization versus paludification [@buffamFillingHolesRegional2010], and probably along other axes of peatland typology.
+Therefore, DSM is more likely to uncover consistent predictor--depth relationships within peatland regions than across them.
 
@@
 However, since this approach is data-hungry, we should build soil knowledge into the DSM where we can [@minasnySoilScienceInformedMachine2024]. 
-For example, if further research confirms the effect of the upper marine boundary that we found, then it is better to include the marine boundary as a predictor than to make the algorithm learn this pattern from elevation independently.
+For example, if further research confirms the effect of the marine limit that we found, then it is better to include the marine limit as a predictor than to make the algorithm learn this pattern from elevation independently.
 
-### Peat occurrence and depth should be mapped together
+### Peat extent and depth should be mapped together
 
-Finally, we want to highlight the need for research on peatland occurrence and peat depth to be better integrated. 
-Since peatland occurrence/extent requires non-zero peat depth [the specific threshold varies by definition, @minasnyMappingMonitoringPeatland2023], they are fundamentally linked. 
-The goal, therefore, should be a unified prediction framework for occurrence and depth. 
-We caution against reducing continuous depth predictions to arbitrary classes [as in @ivanovsModelingGeospatialDistribution2024; @karjalainenComparisonTwoGammaray2024], since classes can be derived from continuous predictions.
-The distribution of peat depths across full landscapes is zero-inflated, and research is needed to determine whether it is more efficient to parameterize a single model of peat depth (with a larger, generalized dataset) or to break down the problem by classifying zero depth and then regressing non-zero depths (with smaller, specialized datasets).
-Coupling occurrence and depth will reduce the prevalence of incoherence that we found: deep peat outside the peatland extent and zero depth inside it. 
+Finally, we want to highlight the need for research on peatland extent and peat depth to be better integrated. 
+Since peatland extent is defined by non-zero peat depth [the specific threshold varies by definition, @minasnyMappingMonitoringPeatland2024], they are fundamentally linked. 
+The goal, therefore, should be a unified prediction framework for extent and depth. 
+We caution against reducing continuous depth predictions to arbitrary classes [as in @ivanovsModelingGeospatialDistribution2024; @karjalainenComparisonTwoGammaray2025], since classes can be derived from continuous predictions.
+The distribution of peat depths across full landscapes is zero-inflated, and research is needed to determine whether it is more efficient to parameterize a single model of peat depth (with a larger, generalized dataset) or to break down the problem into a hurdle model by classifying zero depth and then regressing non-zero depths (with smaller, specialized datasets).
+Coupling extent and depth will reduce the prevalence of incoherence that we found: deep peat outside the peatland extent and zero depth inside it. 
 A key challenge going forward will be obtaining training data that represents both zero and non-zero components of the depth distribution, since sampling designs often focus on known peatlands.

### 2025-02-28 ms/ms.tex
addressed coauthor comments

@@
 Accurate mapping of peat depth is crucial for carbon accounting and land management, yet existing maps lack the resolution and accuracy needed for these applications.
-This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in Norway.
+This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in western and southeastern Norway.
 Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data across two, \textgreater10 km\textsuperscript{2} sites.
 We measured peat depth by probing and ground-penetrating radar at 372 and 1878 locations at the two sites, respectively.
-Then we trained random forest models using radiometric and terrain variables, plus the national map of peat depth, to predict peat depth at 10 m resolution.
+Then we trained Random Forest models using radiometric and terrain variables, plus the national map of peat depth, to predict peat depth at 10 m resolution.
 The two best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth.
@@
 The accuracy of the national map of peat depth did not measure up to any of our remote sensing models, even though it was calibrated to the same data.
-Still, weak relationships with remotely sensed variables make peat depth hard to predict overall.
-Based on these findings, digital soil mapping can improve existing, broad-scale maps of peat depth, but local carbon stock assessment requires additional field sampling.
-Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat occurrence and depth.
+Still, weak relationships with remotely sensed variables made peat depth hard to predict overall.
+Based on these findings, we conclude that digital soil mapping can improve existing, broad-scale maps of peat depth in Norway, but highly localized carbon stock assessments are best made from field measurements.
+Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat lateral extent and depth.
+Together, these pathways promise more accurate landscape-scale carbon stock assessments and better-informed land management policies.
 \end{abstract}
@@
 Peat soils are a terrestrial carbon stock of global importance.
-They store about 30 \% of global soil carbon despite covering only 2--3 \% of Earth's land \citep{xuPEATMAPRefiningEstimates2018, friedlingsteinGlobalCarbonBudget2020, unepGlobalPeatlandsAssessment2022}.
+They store 450-650 Gt of carbon, or about 30 \% of global soil carbon, despite covering only 2--3 \% of Earth's land \citep{xuPEATMAPRefiningEstimates2018, friedlingsteinGlobalCarbonBudget2020, unepGlobalPeatlandsAssessment2022}.
 Peatlands (areas with peat soil) are more carbon dense per square meter than any other ecosystem \citep{temminkRecoveringWetlandBiogeomorphic2022}.
@@
 One of the keys to the areal density of peatland carbon stocks lies in the third dimension: their depth.
-Peat soils range from zero to over ten meters deep, so deep peats comprise a large volume.
+Peat soils range from zero to over ten meters deep \citep{widyastutiPEATGRIDSMappingThickness2024}, so deep peats comprise a large volume.
 Their depth results from the accumulation of organic matter over thousands of years \citep{loiselDatabaseSynthesisNorthern2014, joostenRolePeatlandsClimate2016}.
 In the anoxic and acidic conditions created by a high water table, plant material decay is slightly outweighed by new growth, and the surplus carbon is laid down as peat.
-There it remains sequestered as long as the water table stays high and prevents oxidative peat decomposistion.
+There it remains sequestered as long as the water table stays high and prevents oxidative peat decomposition.
 
-Peatlands are unevenly distributed globally, and in regions with high cover they are frequently converted to human land use \citep{unepGlobalPeatlandsAssessment2022}.
+Peatlands are most common at high latitudes, and in regions with high cover they are frequently converted to human land use \citep{unepGlobalPeatlandsAssessment2022}.
 They are attractive for agriculture and forestry because they are flat, treeless, and have developed soils.
 However, other land uses also displace peatlands.
-In Norway -- where flat lowland is relatively scarce but 9 \% of the land area is peatland \citep{brynLandCoverNorway2018} -- lawmakers have restricted peatland afforestation and cultivation in recent decades.
+In Norway -- where flat lowland is relatively scarce but upward of 9 \% of the land area is peatland \citep{brynLandCoverNorway2018, bakkestuenDelineationWetlandAreas2023} -- lawmakers have restricted peatland afforestation and cultivation in recent decades.
 Since then, a larger proportion of peatland loss is driven by construction \citep{flagetMyrMellomLokal2024}.
@@
 Measuring peat depth on the ground by probing or ground-penetrating radar is straightforward, and a field survey can map a small area at low cost.
-However, surveying large areas is impractical when depth varies over short distances (e.g.~\textless{} 1 m) -- as in most peatlands.
+However, surveying large areas is impractical when depth varies widely over short distances (e.g.~10 m) -- as in many peatlands \citep{torppaStatisticalApproachDetermining2011, proulx-mcinnisReviewGroundpenetratingRadar2013, henrionMappingMonitoringPeatlands2024}.
 A complementary approach from soil science is digital soil mapping (DSM).
@@
 
-The crux for DSM of peat depth is the relationship between peat depth and each of the other, mapped variables.
+The crux for DSM of peat depth is the relationship between peat depth and each of the other, mapped predictors.
 For DSM to work, these relationships must be strong and consistent over the area of interest.
@@
 Moreover, mechanisms of peat formation are linked to topography.
-For example, a steep slope is unlikely to have a high water table, thereby limiting accumulation.
+For example, a steep slope is unlikely to have a high water table, thereby limiting accumulation of peat.
 
@@
 Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth in some studies \citep[e.g.][]{rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024}.
-Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic variables and the resolution of the peat depth analysis.
+Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic predictors and the resolution of the peat depth analysis.
 
-Another set of variables related to peat depth are measurements of natural radioactivity from the ground surface.
+Another set of predictors related to peat depth are measurements of natural radioactivity from the ground surface.
 Gamma-ray spectrometry can survey the activity (decay counts per second) from radioactive isotopes in the earth's crust: potassium-40, uranium-238, and thorium-232 \citep{reinhardtGammaraySpectrometryVersatile2019}.
@@
 Norway has 22 peatland mesotopes, with different hydrology, formation, and development -- from topogeneous or soligeneous fens to blanket or raised bogs \citep{joostenWiseUseMires2002, lyngstadBeskrivelserAvTorvmassivenheter2023}.
-These mesotopes have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric variables will vary by landscape.
+These mesotopes have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric predictors will vary by landscape.
 
@@
 
-The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify peatlands with rich carbon stocks to avoid their conversion and prioritize their preservation or restoration.
+The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify peatlands with rich carbon stocks to avoid their conversion and prioritize their preservation or restoration \citep{strackPotentialPeatlandsNatureBased2022}.
 Although ground surveys of peat depth are accurate and feasible across small areas, having landscape-scale maps \emph{before} detailed investigation increases the option space for spatial planning and land management.
@@
 
-Here, we assess how well remotely sensed topographic and radiometric data can predict peat depth at the landscape scale, with a view towards revising regional and national maps in Norway.
-Specifically, we: (1) quantify the accuracy of predictions from topographic and radiometric variables, (2) identify key predictive variables, and (3) compare these relationships across two different Norwegian landscapes.
+Here, we assess how well remotely sensed topographic and radiometric data can predict peat depth at two contrasting sites, with a view towards revising regional and national maps in Norway.
+Specifically, we aim to: (1) quantify the accuracy of predictions from topographic and radiometric variables, and (2) identify key predictive variables.
+Reflecting on our results and the need for continued improvement in peat depth maps, we close by discussing implications for digital soil mapping of peat depth.
+
 To our knowledge, this is the first study to predict peat depth from airborne radiometric data using machine learning algorithms.
@@
 At Skrimfjella we delineated a study area of 34 km\textsuperscript{2} based on radiometric coverage and accessibility (Fig. \ref{fig:sites}b).
-The study area has a diverse bedrock, with 32 \% alkali feldspar granite, 26 \% marlstone, 10 \% granite, 8 \% mozonite, 7 \% sandstone, 6 \% limestone, and five other rock types with 1--5 \% coverage (Geological Survey of Norway, 1:250 000 dataset).
+The study area has a diverse bedrock, with 32 \% alkali feldspar granite, 26 \% marlstone, 10 \% granite, 8 \% monzonite, 7 \% sandstone, 6 \% limestone, and five other rock types with 1--5 \% coverage (Geological Survey of Norway, 1:250 000 dataset).
 The area within our delineation at Skrimfjella is classified as the landscape type \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
-It is almost without human infrastructure, dominated by forest, and borders on a large nature reserve.
+It is almost without human infrastructure, dominated by conifer forest, and borders on a nature reserve.
 The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).
@@
 We created the same suite of peat depth predictors for both sites (25 continuous and 1 categorical; Table \ref{tab:preds}).
+Each continuous predictor equates to one variable in the model, while the categorical predictor equates to two variables in the model because its three levels become two indicator variables (one level is used as the reference).
 All continuous predictors were derived either from an airborne radiometric survey or from a DTM.
-From the radiometric surveys we simply used the four variables produced by the Geological Survey of Norway: ground concentration of Potassium, Thorium, Uranium, as well as total count.
+From the radiometric surveys, we simply used the four variables produced by the Geological Survey of Norway: ground concentration of potassium, thorium, uranium, as well as total count.
 From the DTMs we calculated several land surface parameters, ranging from simple terrain indices to more complex geomorphometric and hydrological variables \citep{maxwellLandsurfaceParametersSpatial2022}.
@@
 We chose a spatial resolution of 10 m for our predictors, peat depth sample, and modelling.
-We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other.
+We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other hand.
 Complete descriptions of predictors follow below.
 
-\begin{table}
-
-\caption{\label{tab:preds}Candidate predictors of peat depth.}
-\centering
-\begin{tabular}[t]{lll}
+\begin{table}[tbp]
+\caption{Twenty-six candidate predictors of peat depth.
+Note that each continuous predictor contributes one variable to the model while the single categorical predictor (DMK) contributes two binary variables, corresponding to its three levels (one level is used as the reference).
+This results in a total of 27 variables.}
+\begin{tabular}{llll}
 \hline
-Group & Name & Description\\
-\hline
-radiometric & radK & Potassium ground concentration\\
- & radTh & Thorium ground concentration\\
- & radU & Uranium ground concentration\\
- & radTC & Total count of gamma radiation\\
-terrain & elevation & Mean elevation\\
- & slope1m & Mean of 1 m slope\\
- & TPI1m & Mean of 1 m topographic position index\\
- & TRI1m & Mean of 1 m terrain ruggedness index\\
- & roughness1m & Mean of 1 m roughness\\
- & slope10m & 10 m slope\\
- & TPI10m & 10 m topographic position index\\
- & TRI10m & 10 m terrain ruggedness index\\
- & roughness10m & 10 m roughness\\
- & MRVBF & Multi-resolution valley bottom flatness\\
- & TWI5m & Mean of 5 m topographic wetness index\\
- & TWI10m & 10 m topographic wetness index\\
- & TWI20m & Bilinear interpolation of 20 m topographic wetness index\\
- & TWI50m & Bilinear interpolation of 50 m topographic wetness index\\
- & DTW2500 & Depth-to-water index, flow initiation area of 0.25 ha\\
- & DTW5000 & Depth-to-water index, flow initiation area of 0.5 ha\\
- & DTW10000 & Depth-to-water index, flow initiation area of 1 ha\\
- & DTW20000 & Depth-to-water index, flow initiation area of 2 ha\\
- & DTW40000 & Depth-to-water index, flow initiation area of 4 ha\\
- & DTW80000 & Depth-to-water index, flow initiation area of 8 ha\\
- & DTW160000 & Depth-to-water index, flow initiation area of 16 ha\\
-national map & DMK & DMK peat depth class, categorical with 3 levels\\
-\hline
+Group        & Code         & Units                 & Description                                                            \\ \hline
+radiometric  & radK         & \%                    & Potassium ground concentration                                         \\ 
+             & radTh        & ppm                   & Thorium ground concentration                                           \\
+             & radU         & ppm                   & Uranium ground concentration                                           \\
+             & radTC        & \unit{counts\,s^{-1}} & Total Count of gamma radiation                                         \\
+terrain      & elevation    & m                     & mean elevation from DTM with 1 m resolution                            \\
+             & slope1m      & degrees               & mean of slope at 1 m resolution                                        \\
+             & TPI1m        & m                     & mean of Topographic Position Index at 1 m resolution                   \\
+             & TRI1m        & m                     & mean of Terrain Ruggedness Index at 1 m resolution                     \\
+             & roughness1m  & m                     & mean of roughness at 1 m resolution                                    \\
+             & slope10m     & degrees               & slope from DTM with 10 m resolution                                    \\
+             & TPI10m       & m                     & Topographic Position Index from DTM with 10 m resolution               \\
+             & TRI10m       & m                     & Terrain Ruggedness Index from DTM with 10 m resolution                 \\
+             & roughness10m & m                     & roughness from DTM with 10 m resolution                                \\
+             & MRVBF        & dimensionless         & Multi-Resolution Valley Bottom Flatness                                \\
+             & TWI5m        & dimensionless         & mean of Topographic Wetness Index at 5 m resolution                    \\
+             & TWI10m       & dimensionless         & Topographic Wetness Index at 10 m resolution                           \\
+             & TWI20m       & dimensionless         & bilinear interpolation of Topographic Wetness Index at 20 m resolution \\
+             & TWI50m       & dimensionless         & bilinear interpolation of Topographic Wetness Index at 50 m resolution \\
+             & DTW2500      & m                     & Depth-To-Water index, flow initiation area of 0.25 ha                  \\
+             & DTW5000      & m                     & Depth-To-Water index, flow initiation area of 0.5 ha                   \\
+             & DTW10000     & m                     & Depth-To-Water index, flow initiation area of 1 ha                     \\
+             & DTW20000     & m                     & Depth-To-Water index, flow initiation area of 2 ha                     \\
+             & DTW40000     & m                     & Depth-To-Water index, flow initiation area of 4 ha                     \\
+             & DTW80000     & m                     & Depth-To-Water index, flow initiation area of 8 ha                     \\
+             & DTW160000    & m                     & Depth-To-Water index, flow initiation area of 16 ha                    \\
+DMK          & DMK          & categorical           & peat depth class in the DMK national map dataset: shallow/deep/unknown \\ \hline
 \end{tabular}
+\label{tab:preds}
 \end{table}
@@
 \citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013} and Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.
-They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling, using the \emph{terra} package in R.
+They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling, using the \emph{terra} package (v1.7) in R.
 
 The radiometric surveys were conduced with similar flight parameters (Table \ref{tab:radSurveys}).
-Spectrometer count rates were calibrated to known concentrations of Potassium, Thorium, and Uranium in mobile pads.
+Spectrometer count rates were calibrated to known concentrations of potassium, thorium, and uranium in mobile pads.
 Data were processed following standard procedures outlined by the International Atomic Energy Association.
@@
 
-We used the \emph{terra} R package to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
+We used the \emph{terra} package to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
 These were derived at two scales to produce eight different predictors; we either calculated the indices at 1 m DTM resolution and then aggregated to 10 m resolution, or aggregated to 10 m DTM resolution and then calculated the indices.
@@
 
-The Depth-to-Water index \citep{murphyMappingWetlandsComparison2007} approximates a location's vertical height above the surface water feature that it is likely to drain towards.
+The Depth-to-Water index \citep{murphyMappingWetlandsComparison2007} approximates a location's vertical height above the surface water feature (e.g.~stream, lake, or sea) that it is likely to drain towards.
 It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature \citep[eq. 5 in][]{murphyTopographicModellingSoil2009}.
@@
 Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell \citep{schonauerSpatiotemporalPredictionSoil2021, schonauerRcodeCalculatingDepthwater2021}.
-This catchments area layer was then thresholded at seven different levels (flow initiation area 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry \citep{murphyModellingMappingTopographic2011, agrenEvaluatingDigitalTerrain2014, schonauerSpatiotemporalPredictionSoil2021}.
+This catchment area layer was then thresholded at seven different levels (flow initiation area 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry \citep{murphyModellingMappingTopographic2011, agrenEvaluatingDigitalTerrain2014, schonauerSpatiotemporalPredictionSoil2021}.
 In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
@@
 
-At Skrimfjella, we used the \emph{eSample} function in the \emph{iSDM} R package (v.1.0) to stratify our sample across elevation, slope, and potassium concentration.
-This function defines the environmental space as a two-dimensional convex hull around the ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest \citep{hattabUnifiedFrameworkModel2017}.
+At Skrimfjella, we used the \emph{eSample} function in the \emph{iSDM} R package (v1.0) to stratify our sample across elevation, slope, and potassium concentration.
+This function defines the environmental space as a two-dimensional convex hull around the PCA-ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest \citep{hattabUnifiedFrameworkModel2017}.
 We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and \emph{eSample} returned 105 raster cells.
@@
 We also measured peatland occurrence (binary variable) in a separate sample at Skrimfjella.
-We measured peatland occurrence because the AR5 dataset is known to underestimate peatland coverage \citep[especially in forests,][]{brynLandCoverNorway2018}, and because airborne radiometrics may help identify unmapped peatland \citep{gatisMappingUplandPeat2019, olearyDigitalSoilMapping2022}.
+We measured peatland occurrence because the AR5 dataset is known to underestimate peatland extent \citep[especially in forests,][]{brynLandCoverNorway2018}, and because airborne radiometrics may help identify unmapped peatland \citep{gatisMappingUplandPeat2019, olearyDigitalSoilMapping2022}.
 The 10 m cells were sampled from those outside AR5 mire and with slope \textless{} 20°.
@@
 For a sequence of sample sizes (50--500) \citep[ten replicates each, drawn by conditioned latin hypercube sampling,][]{minasnyConditionedLatinHypercube2006, roudierClhsPackageConditioned2011}, we calculated the mean Kullback--Leibler divergence between sample and population distributions \citep{maloneMethodsImproveUtility2019, sauretteDivergenceMetricsDetermining2023}.
-Then, we fitted a asymptotic regression of mean divergence on sample size, and found that the curve reached 95 \% of the fitted asymptote at a sample size of 160.
+Then we fitted an asymptotic regression of mean divergence on sample size, and found that the curve reached 95 \% of the fitted asymptote at a sample size of 160.
 
-To choose 160 locations, we performed feature space coverage sampling, implemented using the \emph{kmeans} function in base R and the \emph{rdist} function in the the \emph{fields} package (v.14.2).
+To choose 160 locations, we performed feature space coverage sampling, implemented using the \emph{kmeans} function in base R and the \emph{rdist} function in the the \emph{fields} package (v14.2).
 Feature space coverage sampling chooses locations that are closest to cluster centers in standardized predictor space \citep{brusSamplingDigitalSoil2019}.
@@
 After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
-Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 \% (with respect to the ideal sample), but distance in our sample was still only 46 \% of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
+Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 \% (compared to the ideal sample), but distance in our sample was still only 46 \% of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
 
@@
 
-We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real time kinematic (Ørskogfjellet) global navigation satellite system receivers.
+We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real time kinematic (Ørskogfjellet) global navigation satellite system (GNSS) receivers.
 We dampened the effect of outlying measurements by probing three times at each location \citep{parryEvaluatingApproachesEstimating2014}, at the vertices of a centered triangle with 2 m (Skrimfjella) or 4.5 m (Ørskogfjellet) sides.
@@
 We judged whether the soil was a peat soil based on its density, texture, and color, as well as the presence or absence of mire vegetation.
-Although peat soil is strictly defined by organic content (which we did not analyze), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland \citep{minasnyMappingMonitoringPeatland2023}.
+Although peat soil is strictly defined by organic content (which we did not analyze), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland \citep{minasnyMappingMonitoringPeatland2024}.
 
@@
 
-We performed GPR surveys in three subjectively chosen mires at Skrimfjella and in areas with a high density of sampled raster cells at Ørskogfjellet.
+We performed GPR surveys in three subjectively chosen peatlands at Skrimfjella and in areas with a high density of sampled raster cells at Ørskogfjellet.
 We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with a GNSS-enabled control unit connected to a 500 MHz shielded antenna mounted in a plastic sledge (transmitter--receiver separation 0.18 m, trace frequency 10 Hz).
@@
 
-We used random forests (RF) to predict peat depth at both sites.
+We used Random Forests (RF) to predict peat depth at both sites.
 RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees \citep{breimanRandomForests2001}.
 We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in DSM applications than other algorithms \citep{beguinPredictingSoilProperties2017, nussbaumEvaluationDigitalSoil2018, lamichhaneDigitalSoilMapping2019}.
-It is suited for use on relatively small training data sets and its predictions can be interrogated to learn about predictor importance \citep{khaledianSelectingAppropriateMachine2020}.
+It is suited for use on relatively small training datasets and its predictions can be interrogated to learn about predictor importance \citep{khaledianSelectingAppropriateMachine2020}.
 Evaluating variable importance in a maximally predictive model aligns with the aim of this study.
@@
 If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.
-We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that (non-spatial) RF rather regression kriging was justified at both sites.
+We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that (non-spatial) RF rather than regression kriging was justified at both sites.
 
-We implemented models in the \emph{tidymodels} framework in R \citep{kuhnTidymodelsCollectionPackages2020}, with the \emph{ranger} R package for RFs \citep[v.0.16,][]{wrightRangerFastImplementation2017}.
+We implemented models in the \emph{tidymodels} framework in R \citep{kuhnTidymodelsCollectionPackages2020}, with the \emph{ranger} R package for RFs \citep[v0.16,][]{wrightRangerFastImplementation2017}.
 RFs were fit with 1000 trees, minimum node size of 5, and the number of predictors randomly sampled at each split was the square root of the total number of predictors (\emph{ranger} default).
@@
 
-For both sites we compared the performance of models with five different configurations of predictors.
+For both sites, we compared the performance of models with five different configurations of predictors.
 Specifically, we trained models with:
@@
 \item
-  only DMK peat depth class (2 indicator variables from 3 levels)
+  \emph{DMK} (2 variables)
 \item
-  only terrain (21 continuous variables)
+  \emph{terrain} (21 variables)
 \item
-  terrain and DMK peat depth class (23 variables)
+  \emph{terrain + DMK} (23 variables)
 \item
-  terrain and radiometric predictors (25 continuous variables)
+  \emph{terrain + radiometric} (25 variables)
 \item
-  all predictors (27 variables)
+  \emph{all predictors} (27 variables)
 \end{enumerate}
@@
 For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation \citep{wadouxSpatialCrossvalidationNot2021}.
-We implemented the kNNDM with the \emph{CAST} R package \citep[v.1.0.2,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
+We implemented the kNNDM with the \emph{CAST} R package \citep[v1.0,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
 From the cross-validation we quantified mean absolute error (error magnitude, original scale), R\textsuperscript{2} (explained variation, standardized scale), and Lin's concordance correlation coefficient (error magnitude and explained variation, standardized scale).
@@
 The quantile regression forests were trained with predictor configuration that showed the highest performance at each site (under the assumption that these models would be put into production) and we extracted 90 \% prediction intervals.
+
+\paragraph{Peatland extent}
 
@@
 This analysis treats the model's prediction of peat depth as a rank index of peatland likelihood.
-Since the purpose was to test the models' ability to uncover unmapped peatland (where DMK peat depth class is undefined), we evaluated models with two predictor configurations: terrain predictors alone or terrain and radiometric predictors together.
+Since the purpose was to test the models' ability to uncover unmapped peatland (where DMK peat depth class is undefined), we evaluated models with two predictor configurations: \emph{terrain} or \emph{terrain + radiometric}.
 
@@
 
-We quantified global variable importance (predictor influence across locations) and examined partial dependence plots (curves of fitted relationships) for the best-performing predictor configuration at each site.
+We quantified global variable importance (predictor influence across all locations) and examined partial dependence plots (curves of fitted relationships) for the best-performing predictor configuration at each site.
 Both are useful for understanding the mechanisms behind the model's predictions and the roles of the predictors in the model.
 
-For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration -- because correlation between predictors degrades variable importance measures \citep{stroblConditionalVariableImportance2008, biauRandomForestGuided2016} and can produce misleading visualizations of predictor--outcome relationships \citep{biecekExplanatoryModelAnalysis2021, dwivediExplainableAIXAI2023}.
-Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation above 0.7.
-Thus, highly correlated sets of predictors are represented by a single variable for the purposes of model interpretation.
+For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration -- because correlation between predictors degrades variable importance metrics \citep{stroblConditionalVariableImportance2008, biauRandomForestGuided2016} and can produce misleading visualizations of predictor--outcome relationships \citep{biecekExplanatoryModelAnalysis2021, dwivediExplainableAIXAI2023}.
+Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation coefficient above 0.7.
+Thus, highly correlated sets of variables are represented by a single variable for the purposes of model interpretation.
 
-We calculated variable importance with the \emph{vip} R package (v.0.4.1), by three different methods: \emph{FIRM}, \emph{permutation}, and \emph{Shapley} \citep{greenwellVariableImportancePlots2020}.
+We calculated variable importance with the \emph{vip} R package (v0.4), by three different methods: \emph{FIRM}, \emph{permutation}, and \emph{Shapley} \citep{greenwellVariableImportancePlots2020}.
 \emph{FIRM} values measure the flatness of the partial dependence plot, \emph{permutation} values measure the decrease in model performance when the predictor is permuted, and \emph{Shapley} values are aggregated from local, game-theoretical measures of variable importance \citep{greenwellVariableImportancePlots2020}.
@@
 
-We calculated partial dependence with the \emph{pdp} R package \citep[v.0.8.1,][]{greenwellPdpPackageConstructing2017}.
+We calculated partial dependence with the \emph{pdp} R package \citep[v0.8,][]{greenwellPdpPackageConstructing2017}.
 For the six most important variables, we plotted both partial dependence and individual conditional expectation, to show the average effect of the predictor on the outcome and the variation in the effect across observations, respectively \citep{goldsteinPeekingBlackBox2015}.
@@
                     &                            & n        & \%      & depth      & n          & \%      & depth      \\ \hline
-    &                            & 372      & 100     & 119        & 1878       & 100     & 126        \\
+all measured cells  &                            & 372      & 100     & 119        & 1878       & 100     & 126        \\
 AR5                 & Agricultural               &          &         &            & 21         & 1       & 30         \\
@@
 
+\subsection{Model performance}
+
 None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \ref{fig:modelMetrics}).
-For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an R\textsuperscript{2} of 0.34, and a mean absolute error of 60 cm.
-For Ørskogfjellet the concordance correlation coefficient was 0.39, R\textsuperscript{2} was 0.33, and mean absolute error was 56 cm, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
+For Skrimfjella, the best model achieved a concordance correlation coefficient of 0.30, an R\textsuperscript{2} of 0.34, and a mean absolute error of 60 cm.
+For Ørskogfjellet, the concordance correlation coefficient was 0.39, R\textsuperscript{2} was 0.33, and mean absolute error was 56 cm, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
 These values were derived from kNNDM spatial cross-validation with 20 folds at Skrimfjella and 10 folds at Ørskogfjellet.
@@
 Although the difficulty of predicting peat depth caused prediction intervals to be wide, these uncertainty estimates were well calibrated.
-At Skrimfjella the prediction interval coverage probability was 91 \%, and at Ørskogfjellet it was 84 \% (both compared to the target value of 90 \%).
+At Skrimfjella, the prediction interval coverage probability was 91 \%, and at Ørskogfjellet it was 84 \% (both compared to the target value of 90 \%).
 Observations outside of the prediction intervals were evenly distributed across each of the study areas.
 
-For Skrimfjella, the best predictor configuration was \emph{all predictors}, followed closely by \emph{terrain and radiometric}.
-The performance gap between the \emph{terrain and DMK} configuration and the \emph{terrain-only} configuration was similarly small.
-Compared to the \emph{terrain-only} configuration, the \emph{terrain and radiometric} configuration improved concordance correlation by 0.04, R\textsuperscript{2} by 0.04, and mean absolute error by 1 cm.
-\emph{DMK class} alone -- although calibrated to measured depths -- was a very poor predictor of peat depth, with a concordance correlation coefficient of 0.0077.
+For Skrimfjella, the best predictor configuration was \emph{all predictors}, followed closely by \emph{terrain + radiometric}.
+The performance gap between the \emph{terrain + DMK} configuration and the \emph{terrain} configuration was similarly small.
+Compared to the \emph{terrain} configuration, the \emph{terrain + radiometric} configuration improved concordance correlation by 0.04, R\textsuperscript{2} by 0.04, and mean absolute error by 1 cm.
+\emph{DMK} was a very poor predictor of peat depth even though it was calibrated to measured depths, with a concordance correlation coefficient of 0.0077.
 
-For Ørskogfjellet, the best predictor configuration was \emph{terrain and DMK}, followed by \emph{terrain-only}.
+For Ørskogfjellet, the best predictor configuration was \emph{terrain + DMK}, followed by \emph{terrain}.
 Adding radiometric predictors to these configurations worsened model performance, especially in terms of concordance correlation and R\textsuperscript{2}.
@@
 
+\subsubsection{Peatland extent}
+
 When we tested how well models extrapolated from within to outside of AR5 mire areas, we found that the models produced worse mean absolute error than just assuming a constant 30 cm depth (Fig.
 \ref{fig:modelMetricsExtrapolation}).
-With the independent occurrence data at Skrimfjella, we found that neither the \emph{terrain-only} nor \emph{terrain and radiometric} configurations were able to discriminate between peat presence and absence (area under the curve of the receiver operating characteristic 0.44 and 0.52 respectively, where 0.5 indicates random guessing).
+With the independent occurrence data at Skrimfjella, we found that neither the \emph{terrain} nor \emph{terrain + radiometric} configurations were able to discriminate between peat presence and absence (area under the curve of the receiver operating characteristic 0.44 and 0.52 respectively, where 0.5 indicates random guessing).
 
@@
 
-For the purpose of model interpretation, the \emph{all predictors} configuration for Skrimfjella was reduced from 27 variables to 11 non-collinear variables.
-Similarly, the \emph{terrain and DMK} configuration for Ørskogfjellet was reduced from 23 variables to 11 non-collinear variables.
+\subsection{Model interpretation}
+
+For the purpose of model interpretation, the \emph{all predictors} configuration for Skrimfjella was reduced from 27 variables to 11 non-collinear variables, by removing one of the variables in each highly-correlated pair.
+Similarly, the \emph{terrain + DMK} configuration for Ørskogfjellet was reduced from 23 variables to 11 non-collinear variables.
+
+\subsubsection{Variable importance}
+
 At both sites, elevation and Multi-Resolution Valley Bottom Flatness were important predictors (Fig. \ref{fig:varImp}).
@@
 At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices slope, Terrain Ruggedness Index, Topographic Position Index, and roughness.
-The radiometric variable radU showed moderate importance at Skrimfjella.
+The radiometric predictor uranium ground concentration showed moderate importance at Skrimfjella.
 
@@
 \includegraphics{figures/variable_importance.pdf}
-\caption{\label{fig:varImp}Global variable importance in the best-performing models at Ørskogfjellet (a) and Skrimfjella (b), as measured by three different metrics.
+\caption{\label{fig:varImp}Global variable importance in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), as measured by three different metrics: Shapley values, permutation importance, and the Feature Importance Ranking Measure \citep{greenwellVariableImportancePlots2020}.
 Variables removed due to collinearity are shown to the right of the variable with which they are most correlated.}
@@
 
-Many of the the most important predictors in the best performing models showed non-monotonic effects on peat depth (Fig. \ref{fig:pdps}).
+\subsubsection{Partial dependence}
+
+Many of the most important predictors in the best performing models showed non-monotonic effects on peat depth (Fig. \ref{fig:pdps}).
 At Ørskogfjellet for example, increasing elevation was predictive of deeper peat up to about 75 m above sea level, after which a further increase was predictive of shallower peat.
 At Skrimfjella the partial dependence on elevation had the opposite shape but covers a higher elevation range, with the shallowest peats predicted around 350 m above sea level.
-TWI50m at Ørskogfjellet and DTW4000 and radU at Skrimfjella were other predictors that showed considerable fluctuations in their predictive effects on peat depth.
-The radiometric predictor in particular displayed an idiosyncratic effect, with a marked dip in predicted depth at intermediate values of radU.
+TWI50m at Ørskogfjellet and DTW4000 and uranium ground concentration at Skrimfjella were other predictors that showed considerable fluctuations in their predictive effects on peat depth.
+The radiometric predictor in particular displayed an idiosyncratic effect, with a marked dip in predicted depth at intermediate values of uranium ground concentration.
 On the other hand, the partial effects of some important predictors were more straightforward.
@@
 For example, the magnitude of the increase in depth with elevation that the model expected at Ørskogfjellet was different for different observations; some depth predictions increased by only 40 cm while others increased by more than 100 cm, over the same elevation gain.
-Similarly, individual conditional expectation lines of radU at Skrimfjella were non-parallel, with some locations showing monotonically increasing peat depth predictions with uranium concentration (unlike the average effect).
+Similarly, individual conditional expectation lines of uranium ground concentration at Skrimfjella were non-parallel, with some locations showing monotonically increasing peat depth predictions with uranium concentration (unlike the average effect).
 Nevertheless, most individual conditional expectation lines were generally parallel -- indicating that the average effects of the predictors were good representations of their overall effects.
@@
 \includegraphics{figures/partial_dependence.pdf}
-\caption{\label{fig:pdps}Partial dependence plots of the six most important variables in the best-performing models at Ørskogfjellet (a) and Skrimfjella (b).
-The average effect of the predictor on the outcome (red line) overlays the variation in the effect across observations (grey lines).}
+\caption{\label{fig:pdps}Partial dependence plots of the six most important variables in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b).
+The average effect of the predictor on the outcome (red line) overlays the variation in the effect across observations (grey lines).
+The variables' training data distributions are indicated with rug plots along the x-axes.}
 \end{figure}
@@
 Although model performance was limited, we improved on the best available map of peat depth (DMK depth class), which is based on field measurements only.
-This highlights the general value of remotely sensed data, whose complete coverage can improve maps even when their association with the variable of interest is weak.
+This highlights the general value of remotely sensed data, whose complete coverage can improve maps even when their association with the variable of interest is weak \citep{mulderUseRemoteSensing2011}.
 Since remotely sensed data are widely available, improvements to soil maps as shown here are low hanging fruit.
-This point is recognized and reflected in the rise of DSM.
+This point is recognized and reflected in the rise of DSM \citep{minasnyDigitalMappingPeatlands2019}.
 
@@
 Gatis et al.
-\citeyearpar{gatisMappingUplandPeat2019} used similar predictors and the same spatial grain, finding a much stronger correlation between predicted and observed peat depth (\(R^2 = 0.68\)).
+\citeyearpar{gatisMappingUplandPeat2019} used similar predictors and the same spatial grain, finding a much stronger relationship between predicted and observed peat depth (\(R^2 = 0.68\)).
 Although their random evaluation data partition could make performance estimates too optimistic \citep{robertsCrossvalidationStrategiesData2017, wadouxSpatialCrossvalidationNot2021}, the confounding effect of spatial structure is probably small because they used linear regression and few predictors.
@@
 They study a flatter area with a higher proportion of peatland cover, and their peatland is primarily blanket bog.
+The peat surface of a blanket bog is tied more closely to its underlying topography than the peat surface of raised bogs or fens \citep{lindsayPeatlandMireTypes2016}, which may make depths in blanket bogs easier to predict.
 The predominance of fens and smaller peatland extent may have contributed to worse performance in our study.
@@
 \citeyearpar{gatisMappingUplandPeat2019} at 100 m resolution, using splines to relax linearity between radiometry/terrain and peat depth.
-He found that radK alone predicted peat depth with much higher concordance than our models (concordance correlation coefficient = 0.76) and that elevation alone produced comparable performance (concordance correlation coefficient = 0.27).
+He found that potassium ground concentration alone predicted peat depth with much higher concordance than our models (concordance correlation coefficient = 0.76) and that elevation alone produced comparable performance (concordance correlation coefficient = 0.27).
 The root mean square error from these univariate models was 46--68 cm (cf.~78 and 75 cm for Skrimfjella and Ørskogfjellet).
@@
 Our models erred most for the deepest peats (Fig. \ref{fig:calPlots}).
-Where overprediction occurred, it was smaller.
+Where overprediction occurred, the error was smaller.
 This is not unexpected for the right-skewed distributions of peat depth, but it has management implications.
@@
 
-\subsubsection{Depth predictions do not necessarily extrapolate to peat occurrence}
+\subsubsection{Depth predictions do not necessarily extrapolate to peat extent}
 
 That we measured more than 30 cm of peat in areas not mapped as mire was expected, because AR5 underrepresents peatland extent by about one third \citep{brynLandCoverNorway2018}.
-For that reason, maps of peatland occurrence also need revising.
+For that reason, maps of peatland extent also need revising.
 Unfortunately, our models did not predict peat occurrence outside AR5 mires.
@@
 Peatland extent is probably best mapped using different remotely sensed data than we used here \citep{bakkestuenDelineationWetlandAreas2023}, and this study's purpose was not to map extent.
-Nevertheless --- as long as the relevant peatland definition includes a depth component --- depth predictions (or predictors with subsurface information) should help delineate peatland extent \citep{olearyDigitalSoilMapping2022, beamishDetailedMappingPeat2024}.
+Nevertheless --- as long as the relevant peatland definition includes a depth component --- depth predictions (or predictors thereof) should help delineate peatland extent \citep{olearyDigitalSoilMapping2022, beamishDetailedMappingPeat2024}.
 We return to this point in our discussion of implications for digital soil mapping.
@@
 The bedrock is more homogeneous at Ørskogfjellet than at Skrimfjella, so uneven radiogenesis is not a viable explanation for the differences between sites nor the poor performance in general \citep{beamishEnvironmentalRadioactivityUK2014, reinhardtGammaraySpectrometryVersatile2019}.
-To the extent that radiometrics had predictive value at Skrimfjella, it appears that they were most valuable near the extremes of the depth distribution, since their inclusion improved R\textsuperscript{2} more than mean absolute error.
+To the degree that radiometrics had predictive value at Skrimfjella, it appears that they were most valuable near the extremes of the depth distribution, since their inclusion improved R\textsuperscript{2} more than mean absolute error.
 All four variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value.
-This contrasts with Koganti et al. \citeyearpar{kogantiMappingPeatDepth2023}, who found that radTC was a much better predictor than radK.
+This contrasts with Koganti et al. \citeyearpar{kogantiMappingPeatDepth2023}, who found that radiometric total count was a much better predictor than potassium ground concentration.
 
@@
 
-Studies comparing airborne and ground radiometric surveys confirm that they are poorly correlated in low-activity areas like peatlands \citep{kockComparisonAirborneTerrestrial2011, karjalainenComparisonTwoGammaray2024}.
-Karjalainen et al. \citeyearpar{karjalainenComparisonTwoGammaray2024} found that ground-based measurements predicted peat depth better than airborne measurements.
+Studies comparing airborne and ground radiometric surveys confirm that they are poorly correlated in low-activity areas like peatlands \citep{kockComparisonAirborneTerrestrial2011, karjalainenComparisonTwoGammaray2025}.
+Karjalainen et al. \citeyearpar{karjalainenComparisonTwoGammaray2025} found that ground-based measurements predicted peat depth better than airborne measurements.
 Nevertheless, a large radiometric footprint did not prevent strong associations with peat depth in Gatis et al.
-\citeyearpar{gatisMappingUplandPeat2019} and Marchant \citeyearpar{marchantUsingRemoteSensors2021}, likely due to the extensive blanket bog landscape with more gradual changes in depth \citep{lindsayBogsEcologyClassification1995, lyngstadBeskrivelserAvTorvmassivenheter2023}.
+\citeyearpar{gatisMappingUplandPeat2019} and Marchant \citeyearpar{marchantUsingRemoteSensors2021}, perhaps because the extensive blanket bog landscape in these studies has more gradual changes in depth \citep{lindsayBogsEcologyClassification1995}.
 We are unsure whether short-range depth changes explain the weak associations that Siemon et al. \citeyearpar{siemonAirborneElectromagneticRadiometric2020} found in a large raised bog.
@@
 The RF algorithm's flexibility means that radiometrics could be used for shallower peats if they provided predictive value for that part of the depth distribution, but there is no evidence of that in our results.
-In the partial dependence plot of radU at Skrimfjella, the expected negative relationship between depth and radU cannot be found by ignoring the left (highly attenuated) side of the distribution.
+In the partial dependence plot of uranium ground concentration at Skrimfjella, the expected negative relationship between depth and uranium ground concentration cannot be found by ignoring the left (highly attenuated) side of the distribution.
 About a quarter of the peats in our study were less than a meter deep, and full attenuation is unlikely for these \citep{beamishGammaRayAttenuation2013}.
@@
 We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g.~via climate) to explain the observed patterns.
-Indeed, elevation at Ørskogfjellet seems to interact with other variables (non-parallel individual conditional expectation lines), complicating its interpretation.
-Moreover, relationships between elevation and peat depth have previously shown opposite shapes in different areas \citep{finlaysonEstimatingOrganicSurface2021}.
+For example, the increase in peat depth from 350 to 700 m.a.s.l.
+at Skrimfjella is opposite to the general pattern of deeper peats in lowland than upland Norway \citep{lyngstadBeskrivelserAvTorvmassivenheter2023}.
+Moreover, elevation at Ørskogfjellet seems to interact with other variables (non-parallel individual conditional expectation lines), complicating its interpretation.
+Relationships between elevation and peat depth have previously shown opposite shapes in different areas \citep{finlaysonEstimatingOrganicSurface2021}.
 Nonetheless, a relationship that is not generalizable beyond the mapping area is still useful for DSM, as long as it is evaluated to demonstrate its robustness for the predictive task (e.g.~through kNNDM spatial cross-validation).
 
-One interesting feature of the elevational relationships we found may be generalizable: a steep increase in peat depth near the upper marine limit after the last ice age.
-At Ørskogfjellet, the upper marine limit is about 75 meters above today's sea level \citep[Geological Survey of Norway,][]{hogaasDatabaseRegistreringAv2012}, where the partial dependence plot of elevation shows a sharp increase in peat depth.
-In areas under the upper marine limit there has been less time for peat accumulation since the ice sheets retreated, and it is plausible that this makes peats there shallower, all else being equal.
-We cannot evaluate this effect at Skrimfjella, where the upper marine limit is below our study area (at 175 m.a.s.l.).
+One interesting feature of the elevational relationships we found may be generalizable: a steep increase in peat depth near the marine limit after the last ice age.
+At Ørskogfjellet, the marine limit is about 75 meters above today's sea level \citep[Geological Survey of Norway,][]{hogaasDatabaseRegistreringAv2012}, where the partial dependence plot of elevation shows a sharp increase in peat depth.
+In areas under the marine limit there has been less time for peat accumulation since the ice sheets retreated, and it is plausible that this makes peats there shallower, all else being equal.
+This hypothesis is supported by similar findings at a coarser scale in the Hudson Bay Lowlands, where a strong positive relationship between peat depth and distance from the coast can be explained by isostatic uplift and time since peat initiation \citep{liPeatDepthCarbon2025}.
+We cannot evaluate this effect at Skrimfjella, where the marine limit is below our study area (at 175 m.a.s.l.).
 
@@
 We suspect the discrepancy between sites is due to different levels of effort and coverage during the historical surveys; more lowland peatland near agriculture at Ørskogfjellet may have caused more purposeful surveying.
-73 \% of cells measured at Skrimfjella had unknown depth in DMK, compared to 20 \% at Ørskogfjellet.
-Still, the 25 \% of cells classified as deep at Skrimfjella were not meaningfully deeper than the rest, according to our model.
-Interactions between DMK and other variables underline the inconsistency of DMK depth maps, even within a site.
-That a peat at Ørskogfjellet was classified as shallow rather than deep did not uniformly lower the model's expectation of depth there.
+This is evidenced by the fact that 73 \% of the cells measured at Skrimfjella had unknown depth in DMK, compared to 20 \% at Ørskogfjellet.
+Our results show that classification as deep peat at Skrimfjella had no predictive value -- these cells were no different than those classified as shallow or unknown, all else being equal.
+Interactions between DMK and other variables also underline the inconsistency of DMK depth maps, even within a site.
+For example, that a peat at Ørskogfjellet was classified as shallow rather than deep did not always cause the model to predict shallower peat -- sometimes it increased the predicted depth.
 
@@
 
-The performance gap between the best models and the DMK-only models shows that peat depth in Norway should be mapped digitally.
+The performance gap between the best models and the DMK models shows that peat depth in Norway should be mapped digitally.
 Anywhere we have some calibrating measurements, we can get better maps than DMK peat depth classes, at low cost.
@@
 Fine-scale variation in peat depth may favor mapping at very fine resolution (1 m) --- even if land use planning and carbon accounting do not operate at this grain.
-Terrain--depth relationships might be stronger in 1 m cells than in our 10 m cells, especially considering the hummock--hollow microtopography of many peatlands.
+Terrain--depth relationships might be stronger in 1 m cells than in our 10 m cells, especially considering the hummock--hollow microtopography of many peatlands \citep{rydin7Mires1999, lindsayPeatbogsCarbonCritical2010}.
 
@@
 A natural starting point is the bounding area around a spatial cluster of depth measurements.
-For a given cluster, kNNDM can define how far the boundary can extend beyond the point measurements; if a cross-validation nearest neighbor distribution cannot match the sample-to-prediction nearest neighbor distribution, the boundary is too expansive \citep{meyerMachineLearningbasedGlobal2022, linnenbrinkKNNDMCVKfold2024}.
+For a given cluster, kNNDM can define how far the boundary can extend beyond the point measurements; if the sample-to-prediction nearest neighbor distribution cannot be simulated by any set of cross-validation folds, then the extent is too expansive \citep{meyerMachineLearningbasedGlobal2022, linnenbrinkKNNDMCVKfold2024}.
 However, it is unclear how big (N) any cluster of depths should be, and the tradeoff between multiple small-extent DSM and fewer large-extent DSM needs research.
-Bohn and Miller \citeyearpar{bohnLocallyEnhancedDigital2024} advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or `supertopes'), where the composition of mesotopes is similar \citep{moenVegetasjonstyper1998, joostenWiseUseMires2002}.
-Depth varies strongly between peats formed by terrestrialization versus paludification \citep{buffamFillingHolesRegional2010}, and probably along other axes of peatland typology.
-Although most Norwegian landscapes contain multiple, unknown mesotopes, it makes sense to limit their diversity within a single DSM.
+Bohn and Miller \citeyearpar{bohnLocallyEnhancedDigital2024} advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or `supertopes'), where the composition of mesotopes is similar -- e.g.~regions dominated by raised bogs versus regions dominated by sloping fens \citep{moenNationalAtlasNorway1999, joostenWiseUseMires2002}.
+Depth varies systematically between bogs and fens \citep{lindsayPeatlandMireTypes2016}, between peats formed by terrestrialization versus paludification \citep{buffamFillingHolesRegional2010}, and probably along other axes of peatland typology.
+Therefore, DSM is more likely to uncover consistent predictor--depth relationships within peatland regions than across them.
 
@@
 However, since this approach is data-hungry, we should build soil knowledge into the DSM where we can \citep{minasnySoilScienceInformedMachine2024}.
-For example, if further research confirms the effect of the upper marine boundary that we found, then it is better to include the marine boundary as a predictor than to make the algorithm learn this pattern from elevation independently.
+For example, if further research confirms the effect of the marine limit that we found, then it is better to include the marine limit as a predictor than to make the algorithm learn this pattern from elevation independently.
 
-\subsubsection{Peat occurrence and depth should be mapped together}
+\subsubsection{Peat extent and depth should be mapped together}
 
-Finally, we want to highlight the need for research on peatland occurrence and peat depth to be better integrated.
-Since peatland occurrence/extent requires non-zero peat depth \citep[the specific threshold varies by definition,][]{minasnyMappingMonitoringPeatland2023}, they are fundamentally linked.
-The goal, therefore, should be a unified prediction framework for occurrence and depth.
-We caution against reducing continuous depth predictions to arbitrary classes \citep[as in][]{ivanovsModelingGeospatialDistribution2024, karjalainenComparisonTwoGammaray2024}, since classes can be derived from continuous predictions.
-The distribution of peat depths across full landscapes is zero-inflated, and research is needed to determine whether it is more efficient to parameterize a single model of peat depth (with a larger, generalized dataset) or to break down the problem by classifying zero depth and then regressing non-zero depths (with smaller, specialized datasets).
-Coupling occurrence and depth will reduce the prevalence of incoherence that we found: deep peat outside the peatland extent and zero depth inside it.
+Finally, we want to highlight the need for research on peatland extent and peat depth to be better integrated.
+Since peatland extent is defined by non-zero peat depth \citep[the specific threshold varies by definition,][]{minasnyMappingMonitoringPeatland2024}, they are fundamentally linked.
+The goal, therefore, should be a unified prediction framework for extent and depth.
+We caution against reducing continuous depth predictions to arbitrary classes \citep[as in][]{ivanovsModelingGeospatialDistribution2024, karjalainenComparisonTwoGammaray2025}, since classes can be derived from continuous predictions.
+The distribution of peat depths across full landscapes is zero-inflated, and research is needed to determine whether it is more efficient to parameterize a single model of peat depth (with a larger, generalized dataset) or to break down the problem into a hurdle model by classifying zero depth and then regressing non-zero depths (with smaller, specialized datasets).
+Coupling extent and depth will reduce the prevalence of incoherence that we found: deep peat outside the peatland extent and zero depth inside it.
 A key challenge going forward will be obtaining training data that represents both zero and non-zero components of the depth distribution, since sampling designs often focus on known peatlands.
@@
 \codedataavailability{R code used in this study is available at \url{https://github.com/julienvollering/DSMdepth}.
-Depth measurements will be archived in a public repository upon publication.} %% use this section when having data sets and software code available
+The depth measurements we used are archived at \doi{10.6073/pasta/6ce440152f693f2156bf5b692a2e7917} and follow data and metadata standards.} %% use this section when having data sets and software code available
 

### 2025-02-28 ms/tables/predictors.tex
addressed coauthor comments

--- 2025-02-28 ms/ms.tex
@@
-%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
-%% ---------------------------------
-%% This template should be used for copernicus.cls
-%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
-%% For further assistance please contact Copernicus Publications at: production@copernicus.org
-%% https://publications.copernicus.org/for_authors/manuscript_preparation.html
-
-%% copernicus_rticles_template (flag for rticles template detection - do not remove!)
-
-%% Please use the following documentclass and journal abbreviations for discussion papers and final revised papers.
-
-%% 2-column papers and discussion papers
-\documentclass[soil, manuscript]{copernicus}
-
-%% Journal abbreviations (please use the same for preprints and final revised papers)
-
-% Advances in Geosciences (adgeo)
-% Advances in Radio Science (ars)
-% Advances in Science and Research (asr)
-% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
-% Aerosol Research (ar)
-% Annales Geophysicae (angeo)
-% Archives Animal Breeding (aab)
-% Atmospheric Chemistry and Physics (acp)
-% Atmospheric Measurement Techniques (amt)
-% Biogeosciences (bg)
-% Climate of the Past (cp)
-% DEUQUA Special Publications (deuquasp)
-% Earth Surface Dynamics (esurf)
-% Earth System Dynamics (esd)
-% Earth System Science Data (essd)
-% E&G Quaternary Science Journal (egqsj)
-% EGUsphere (egusphere) | This is only for EGUsphere preprints submitted without relation to an EGU journal.
-% European Journal of Mineralogy (ejm)
-% Fossil Record (fr)
-% Geochronology (gchron)
-% Geographica Helvetica (gh)
-% Geoscience Communication (gc)
-% Geoscientific Instrumentation, Methods and Data Systems (gi)
-% Geoscientific Model Development (gmd)
-% History of Geo- and Space Sciences (hgss)
-% Hydrology and Earth System Sciences (hess)
-% Journal of Bone and Joint Infection (jbji)
-% Journal of Micropalaeontology (jm)
-% Journal of Sensors and Sensor Systems (jsss)
-% Magnetic Resonance (mr)
-% Mechanical Sciences (ms)
-% Natural Hazards and Earth System Sciences (nhess)
-% Nonlinear Processes in Geophysics (npg)
-% Ocean Science (os)
-% Polarforschung - Journal of the German Society for Polar Research (polf)
-% Primate Biology (pb)
-% Proceedings of the International Association of Hydrological Sciences (piahs)
-% Safety of Nuclear Waste Disposal (sand)
-% Scientific Drilling (sd)
-% SOIL (soil)
-% Solid Earth (se)
-% State of the Planet (sp)
-% The Cryosphere (tc)
-% Weather and Climate Dynamics (wcd)
-% Web Ecology (we)
-% Wind Energy Science (wes)
-
-% Pandoc citation processing
-
-% The "Technical instructions for LaTex" by Copernicus require _not_ to insert any additional packages.
-% % % From pandoc table feature
-% \usepackage{longtable,booktabs,array}
-% % \usepackage{calc} % for calculating minipage widths
-% % Correct order of tables after \paragraph or \subparagraph
-% \usepackage{etoolbox}
-% \makeatletter
-% \patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-% \makeatother
-% % Allow footnotes in longtable head/foot
-% \IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-% \makesavenoteenv{longtable}
-% 
-% tightlist command for lists without linebreak
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-%
-%% \usepackage commands included in the copernicus.cls:
-%\usepackage[german, english]{babel}
-%\usepackage{tabularx}
-%\usepackage{cancel}
-%\usepackage{multirow}
-%\usepackage{supertabular}
-%\usepackage{algorithmic}
-%\usepackage{algorithm}
-%\usepackage{amsthm}
-%\usepackage{float}
-%\usepackage{subfig}
-%\usepackage{rotating}
-
-\begin{document}
-
-\title{Terrain is a stronger predictor of peat depth than airborne radiometrics in Norwegian landscapes}
-
-\Author[1][julien.vollering@hvl.no]{Julien}{Vollering}
-\Author[2]{Naomi}{Gatis}
-\Author[1]{Mette}{Kusk Gillespie}
-\Author[1]{Karl-Kristian}{Muggerud}
-\Author[1]{Sigurd Daniel}{Nerhus}
-\Author[1]{Knut}{Rydgren}
-\Author[1]{Mikko}{Sparf}
-
-\affil[1]{Department of Civil Engineering and Environmental Sciences, Western Norway University of Applied Sciences, Norway}
-\affil[2]{Department of Geography, University of Exeter, United Kingdom}
-
-\runningtitle{Peat depth from terrain and radiometrics}
-
-\runningauthor{Vollering et al.}
-
-\correspondence{Julien\ Vollering\ (julien.vollering@hvl.no)}
-
-\received{}
-\pubdiscuss{} %% only important for two-stage journals
-\revised{}
-\accepted{}
-\published{}
-
-%% These dates will be inserted by Copernicus Publications during the typesetting process.
-
-\firstpage{1}
-
-\maketitle
-
-\begin{abstract}
-Peatlands are Earth's most carbon-dense terrestrial ecosystems and their carbon density varies with the depth of the peat layer.
-Accurate mapping of peat depth is crucial for carbon accounting and land management, yet existing maps lack the resolution and accuracy needed for these applications.
-This study evaluates whether digital soil mapping using remotely sensed data can improve existing maps of peat depth in western and southeastern Norway.
-Specifically, we assessed the predictive value of LiDAR-derived terrain variables and airborne radiometric data across two, \textgreater10 km\textsuperscript{2} sites.
-We measured peat depth by probing and ground-penetrating radar at 372 and 1878 locations at the two sites, respectively.
-Then we trained Random Forest models using radiometric and terrain variables, plus the national map of peat depth, to predict peat depth at 10 m resolution.
-The two best models achieved mean absolute errors of 60 and 56 cm, explaining one-third of the variation in peat depth.
-Terrain variables were better predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth.
-Radiometric variables showed inconsistent predictive value -- improving performance at one site while degrading it at the other.
-The accuracy of the national map of peat depth did not measure up to any of our remote sensing models, even though it was calibrated to the same data.
-Still, weak relationships with remotely sensed variables made peat depth hard to predict overall.
-Based on these findings, we conclude that digital soil mapping can improve existing, broad-scale maps of peat depth in Norway, but highly localized carbon stock assessments are best made from field measurements.
-Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat lateral extent and depth.
-Together, these pathways promise more accurate landscape-scale carbon stock assessments and better-informed land management policies.
-\end{abstract}
-
-\section{Introduction}
-
-Peat soils are a terrestrial carbon stock of global importance.
-They store 450-650 Gt of carbon, or about 30 \% of global soil carbon, despite covering only 2--3 \% of Earth's land \citep{xuPEATMAPRefiningEstimates2018, friedlingsteinGlobalCarbonBudget2020, unepGlobalPeatlandsAssessment2022}.
-Peatlands (areas with peat soil) are more carbon dense per square meter than any other ecosystem \citep{temminkRecoveringWetlandBiogeomorphic2022}.
-This makes them crucial to climate change mitigation.
-Intact peatlands sequester carbon and overall produce a negative temperature forcing \citep{joostenRolePeatlandsClimate2016}.
-When disturbed, often by conversion to another land use, they can produce large greenhouse gas emissions \citep{maGloballyRobustRelationship2022}.
-
-One of the keys to the areal density of peatland carbon stocks lies in the third dimension: their depth.
-Peat soils range from zero to over ten meters deep \citep{widyastutiPEATGRIDSMappingThickness2024}, so deep peats comprise a large volume.
-Their depth results from the accumulation of organic matter over thousands of years \citep{loiselDatabaseSynthesisNorthern2014, joostenRolePeatlandsClimate2016}.
-In the anoxic and acidic conditions created by a high water table, plant material decay is slightly outweighed by new growth, and the surplus carbon is laid down as peat.
-There it remains sequestered as long as the water table stays high and prevents oxidative peat decomposition.
-
-Peatlands are most common at high latitudes, and in regions with high cover they are frequently converted to human land use \citep{unepGlobalPeatlandsAssessment2022}.
-They are attractive for agriculture and forestry because they are flat, treeless, and have developed soils.
-However, other land uses also displace peatlands.
-In Norway -- where flat lowland is relatively scarce but upward of 9 \% of the land area is peatland \citep{brynLandCoverNorway2018, bakkestuenDelineationWetlandAreas2023} -- lawmakers have restricted peatland afforestation and cultivation in recent decades.
-Since then, a larger proportion of peatland loss is driven by construction \citep{flagetMyrMellomLokal2024}.
-
-The spatial distribution of peat depth is often overlooked in land use planning and carbon accounting because peat depth is not mapped with sufficient coverage, resolution, or accuracy \citep{beilmanPeatCarbonStocks2008, hastieRisksCarbonStorage2022, unepGlobalPeatlandsAssessment2022}.
-Maps are crucial because they link high-level targets to specific management decisions, unlike spatially aggregated estimates \citep{oecdOECDEnvironmentalPerformance2022}.
-For example, Norwegian regulations restrict conversion of peatland to farmland more strictly where peat depth exceeds 1 m, and the distribution of peat depth on a farmer's property determines whether they are allowed to convert peatland (Forskrift om nydyrking, 1997, § 5a).
-Maps of peat depth also make it possible to quantify the effect of specific management decisions and thereby understand how local outcomes contribute to regional and national outcomes \citep{oecdOECDEnvironmentalPerformance2022}.
-
-Measuring peat depth on the ground by probing or ground-penetrating radar is straightforward, and a field survey can map a small area at low cost.
-However, surveying large areas is impractical when depth varies widely over short distances (e.g.~10 m) -- as in many peatlands \citep{torppaStatisticalApproachDetermining2011, proulx-mcinnisReviewGroundpenetratingRadar2013, henrionMappingMonitoringPeatlands2024}.
-A complementary approach from soil science is digital soil mapping (DSM).
-DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest.
-This approach has grown in importance with the availability of remotely sensed data and the advancement of methods for identifying patterns, especially through machine learning \citep{minasnyDigitalMappingPeatlands2019, wadouxMachineLearningDigital2020}.
-
-The crux for DSM of peat depth is the relationship between peat depth and each of the other, mapped predictors.
-For DSM to work, these relationships must be strong and consistent over the area of interest.
-They can be purely correlative rather than causal, but mechanistic relationships are stronger and more consistent than non-causal ones.
-The \emph{scorpan} framework for DSM suggests seven predictor classes to explore: other soil properties, climate, organisms, relief (topography), parent material, age, and spatial position \citep{mcbratneyDigitalSoilMapping2003}.
-
-For peat depth, the most practical and widespread \emph{scorpan} factors are relief and spatial position.
-Spatial position is unique because it is always known (with varying accuracy).
-However, the short range of spatial autocorrelation in peat depth limits its mapping value \citep{henglGenericFrameworkSpatial2004}.
-Relief, or topography, is widely and accurately mapped in digital terrain models (DTMs).
-For example, most of mainland Norway has at least one elevation measurement per square meter from airborne Light Detection and Ranging (LiDAR) surveys.
-Moreover, mechanisms of peat formation are linked to topography.
-For example, a steep slope is unlikely to have a high water table, thereby limiting accumulation of peat.
-
-Studies have consistently shown relationships between peat depth and topography, although the specific patterns vary.
-One of the most robust relationships is a negative correlation between depth and terrain slope \citep[e.g.][]{holdenEstimatingCarbonStock2011, parryMethodModellingPeat2012, gatisMappingUplandPeat2019}.
-Peat depth also changes with elevation in many contexts, but with inconsistent directionality \citep[e.g.][]{holdenEstimatingCarbonStock2011, parryMethodModellingPeat2012, rudiyantoDigitalMappingCosteffective2016, rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024}.
-Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth in some studies \citep[e.g.][]{rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024}.
-Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic predictors and the resolution of the peat depth analysis.
-
-Another set of predictors related to peat depth are measurements of natural radioactivity from the ground surface.
-Gamma-ray spectrometry can survey the activity (decay counts per second) from radioactive isotopes in the earth's crust: potassium-40, uranium-238, and thorium-232 \citep{reinhardtGammaraySpectrometryVersatile2019}.
-These exist in bedrock (and mineral soils) and a peat overburden attenuates the radiation intensity at the surface.
-The degree of attenuation relates to properties of the overburden, especially composition and depth.
-Deep soil with high water content will attenuate radiation most \citep{beamishGammaRayAttenuation2013, reinhardtGammaraySpectrometryVersatile2019}.
-Thus, gamma-ray radiometrics integrate two \emph{scorpan} factors: other soil properties and parent material \citep{mcbratneyDigitalSoilMapping2003}.
-
-Although theory suggests that one meter of peat may fully attenuate radiometric signal \citep{beamishGammaRayAttenuation2013, reinhardtGammaraySpectrometryVersatile2019}, empirical investigations show that the association between peat depth and radioactivity can extend beyond the first meter \citep{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, kogantiMappingPeatDepth2023}.
-Radiometric data are increasingly available over large areas, presenting a big opportunity for mapping peat depth, insofar as they are predictive of peat depth.
-Airborne surveys are used for various applications like mineral prospecting and radon risk management, and some countries have high spatial coverage of such data \citep{minasnyDigitalMappingPeatlands2019, baranwalAirborneGeophysicalSurveys2020}.
-
-The effectiveness of DSM depends not only on the methods and data, but also on the characteristics of the mapping area.
-Norway may be instructive in this respect because its peatlands vary widely across climates and topographies.
-Norway has 22 peatland mesotopes, with different hydrology, formation, and development -- from topogeneous or soligeneous fens to blanket or raised bogs \citep{joostenWiseUseMires2002, lyngstadBeskrivelserAvTorvmassivenheter2023}.
-These mesotopes have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric predictors will vary by landscape.
-
-The little we know about the depths of Norwegian peatlands comes from surveys meant to identify arable land.
-After scattered surveys in the early 20th century, a comprehensive round of surveying was completed 1964--2001 as part of a wider land cover mapping in Norway \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
-This survey produced the initial maps from which Norway's most detailed updated land cover datasets are derived -- including the AR5 and DMK datasets used in this study and described later.
-Because of its agricultural and silvicultural focus, the survey covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable \citep{ahlstromAR5Klassifikasjonssystem2019}.
-Field surveyors carried a 1 m probe and they measured peat depth as a categorical variable: shallow (\textless{} 1 m), or deep (\textgreater{} 1 m).
-These classes were assigned to whole polygons, so spatial resolution is on the order of hectares.
-The number of measurements per unit area was not standardized but probably low.
-
-The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify peatlands with rich carbon stocks to avoid their conversion and prioritize their preservation or restoration \citep{strackPotentialPeatlandsNatureBased2022}.
-Although ground surveys of peat depth are accurate and feasible across small areas, having landscape-scale maps \emph{before} detailed investigation increases the option space for spatial planning and land management.
-For example, the Norwegian Public Roads Administration routinely measures peat depth during geotechnical work, but by that time, the route of the road is already set.
-With a digital peat depth map, planners could better compare routes and climate mitigation measures would have more leverage.
-Therefore, soil science needs more studies of relationships between peat depth and topographic or radiometric data to determine how predictive they are, which variables are most predictive, and how consistent their associations are.
-
-Here, we assess how well remotely sensed topographic and radiometric data can predict peat depth at two contrasting sites, with a view towards revising regional and national maps in Norway.
-Specifically, we aim to: (1) quantify the accuracy of predictions from topographic and radiometric variables, and (2) identify key predictive variables.
-Reflecting on our results and the need for continued improvement in peat depth maps, we close by discussing implications for digital soil mapping of peat depth.
-
-To our knowledge, this is the first study to predict peat depth from airborne radiometric data using machine learning algorithms.
-Where airborne radiometric data have previously been used to predict peat depth, it has been through modeling techniques less suited for prediction and spatial extrapolation \citep[e.g.,][]{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, siemonAirborneElectromagneticRadiometric2020}.
-Where machine learning algorithms have previously been used on airborne radiometric data, it has been to predict peat extent rather than depth \citep[e.g.,][]{olearyDigitalSoilMapping2022}.
-
-\section{Materials and methods}
-
-\subsection{Sites}
-
-To assess the predictive value of terrain and radiometric data for mapping peat depth, we chose two sites with conspicuously different physical geography: Skrimfjella (Fig.
-\ref{fig:sites}b) in eastern Norway and Ørskogfjellet (Fig.
-\ref{fig:sites}a) in western Norway (Fig.
-\ref{fig:sites}c).
-These sites were chosen because they were covered by radiometric data from airborne surveys, had relatively little built-up area, and were accessible by road.
-
-\begin{figure}
-\includegraphics[height=0.81\textheight]{figures/sites-patchwork} \caption{Sites in southern Norway (a) with land cover at Skrimfjella (b) and Ørskogfjellet (c).
-To fit the scale of the maps, the land cover shown here is from the AR50 national land resource dataset, which has simplified geometry with respect to the AR5 dataset that is used in the study.
-Terrain is visualized using a hillshade with slope 45° and azimuth 225°.
-AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (<https://data.norge.no/nlod/en/1.0>) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (<https://creativecommons.org/licenses/by/4.0/>).}\label{fig:sites}
-\end{figure}
-
-At Skrimfjella we delineated a study area of 34 km\textsuperscript{2} based on radiometric coverage and accessibility (Fig. \ref{fig:sites}b).
-The study area has a diverse bedrock, with 32 \% alkali feldspar granite, 26 \% marlstone, 10 \% granite, 8 \% monzonite, 7 \% sandstone, 6 \% limestone, and five other rock types with 1--5 \% coverage (Geological Survey of Norway, 1:250 000 dataset).
-The area within our delineation at Skrimfjella is classified as the landscape type \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
-It is almost without human infrastructure, dominated by conifer forest, and borders on a nature reserve.
-The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).
-In Norway's AR5 national land cover dataset \citep[``areal resources in scale 1:5000'',][]{ahlstromAR5Klassifikasjonssystem2019}, 1.5 km\textsuperscript{2} (4.5 \%) of the study area is classified as `mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.
-
-At Ørskogfjellet we defined a study area of 124 km\textsuperscript{2} which basically followed the southernmost part of the radiometric survey extent (Fig. \ref{fig:sites}a).
-Bedrock in the area is 84 \% granitic gneiss, 11 \% granite, and 5 \% aluminium silicate gneiss (Geological Survey of Norway, 1:250 000 dataset).
-This study area comprises a wide range of major landscape types: \emph{coastal plains}, \emph{coastal fjord}, \emph{inland valleys}, as well as \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
-It is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
-Its mean elevation is 211 m above sea level (range 0--807 m, IQR 73--310 m), and its mean slope at 10 m resolution is 13.0° (IQR 4.7--18.3°).
-In the AR5 dataset, 15.3 km\textsuperscript{2} (12.4 \%) of the study area is classified as mire.
-
-\subsection{Peat depth predictors}
-
-We created the same suite of peat depth predictors for both sites (25 continuous and 1 categorical; Table \ref{tab:preds}).
-Each continuous predictor equates to one variable in the model, while the categorical predictor equates to two variables in the model because its three levels become two indicator variables (one level is used as the reference).
-All continuous predictors were derived either from an airborne radiometric survey or from a DTM.
-From the radiometric surveys, we simply used the four variables produced by the Geological Survey of Norway: ground concentration of potassium, thorium, uranium, as well as total count.
-From the DTMs we calculated several land surface parameters, ranging from simple terrain indices to more complex geomorphometric and hydrological variables \citep{maxwellLandsurfaceParametersSpatial2022}.
-The categorical predictor was peat depth class, from a national map dataset.
-We chose a spatial resolution of 10 m for our predictors, peat depth sample, and modelling.
-We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other hand.
-Complete descriptions of predictors follow below.
-
 \begin{table}[tbp]
@@
 \end{table}
-
-\subsubsection{Radiometric}
-
-The Geological Survey of Norway conducted and processed radiometric surveys over our study areas, as reported in Baranwal et al.
-\citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013} and Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.
-They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling, using the \emph{terra} package (v1.7) in R.
-
-The radiometric surveys were conduced with similar flight parameters (Table \ref{tab:radSurveys}).
-Spectrometer count rates were calibrated to known concentrations of potassium, thorium, and uranium in mobile pads.
-Data were processed following standard procedures outlined by the International Atomic Energy Association.
-Processing included: correction for aircraft and cosmic background radiation, correction for radon in the air, window stripping of the gamma ray spectrum, correction for flying height, conversion of count rates to ground concentrations, and finally gridding to 50 m resolution with micro-leveling.
-At Ørskogfjellet an additional convolution filter was added to smooth the gridded data.
-
-\subsubsection{Terrain}
-
-For terrain-derived predictors, we obtained 1 m resolution rasters from the national DTM (\href{https://creativecommons.org/licenses/by/4.0/}{CC BY 4.0} Norwegian Mapping Authority).
-The DTM for Skrimfjella was produced from airborne laser scanning surveys in 2015 and 2022, with laser point density of \unit{5\,pts\,m^{-2}}.
-For Ørskogfjellet, the DTM was produced from a 2015 survey with \unit{2\,pts\,m^{-2}}.
-Where necessary, DTMs were resampled to the coordinate reference system of the radiometric data.
-
-We used the \emph{terra} package to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
-These were derived at two scales to produce eight different predictors; we either calculated the indices at 1 m DTM resolution and then aggregated to 10 m resolution, or aggregated to 10 m DTM resolution and then calculated the indices.
-This kind of multiscale feature engineering of land surface parameters has been found to improve machine learning predictions of soil properties \citep{millerImpactMultiscalePredictor2015, dornikOptimalScalingPredictors2022, newmanAssessingSpatiallyHeterogeneous2023}.
-We know that peat depth tends to vary at fine scales in Norway, which is why we chose 1 m and 10 m resolutions \citep{maxwellLandsurfaceParametersSpatial2022}.
-We also calculated the Multi-Resolution Valley Bottom Flatness index, which indicates the degree of valley bottom flatness at a given location via a multiscale algorithm \citep{gallantMultiresolutionIndexValley2003}.
-We calculated this index in SAGA GIS \citep[v.9.3.2, Morphometry library,][]{conradSystemAutomatedGeoscientific2015} with default parameters (initial slope threshold = 16 \%, lowness threshold = 0.4, upness threshold = 0.35, slope shape parameter = 4, elevation shape parameter = 3).
-
-The Topographic Wetness Index \citep{quinnPredictionHillslopeFlow1991} is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate to coarse resolution DTMs \citep{agrenEvaluatingDigitalTerrain2014, riihimakiTopographicWetnessIndex2021}, so we calculated it from 5 m, 10 m, 20 m, and 50 m DTM resolution.
-The calculations were performed with Whitebox software \citep{lindsayWhiteboxGATCase2016}, accessed through the \emph{whitebox} R package \citep[v2.4,][]{wuWhiteboxWhiteboxToolsFrontend2022}.
-We filled depressions in the DTM with the algorithm in Wang \& Liu \citeyearpar{wangEfficientMethodIdentifying2006}, and used the deterministic infinity flow accumulation algorithm \citep{tarbotonNewMethodDetermination1997}.
-
-The Depth-to-Water index \citep{murphyMappingWetlandsComparison2007} approximates a location's vertical height above the surface water feature (e.g.~stream, lake, or sea) that it is likely to drain towards.
-It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature \citep[eq. 5 in][]{murphyTopographicModellingSoil2009}.
-We calculated unitless slope from the 1 m DTM using the Whitebox software.
-Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell \citep{schonauerSpatiotemporalPredictionSoil2021, schonauerRcodeCalculatingDepthwater2021}.
-This catchment area layer was then thresholded at seven different levels (flow initiation area 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry \citep{murphyModellingMappingTopographic2011, agrenEvaluatingDigitalTerrain2014, schonauerSpatiotemporalPredictionSoil2021}.
-In addition, all surface water features mapped in the AR5 dataset were also transferred to the raster layer.
-For each of the seven surface water layers, we derived Depth-to-Water using the \emph{Distance Accumulation} tool in ArcGIS Pro (v.3.1, ESRI, USA), which has an efficient algorithm to find the cumulative distance over a cost surface to the least-cost source.
-
-\subsubsection{Peat depth class}
-
-We prepared one categorical predictor -- peat depth class -- from the national map dataset called \emph{DMK} \citep{ahlstromAR5Klassifikasjonssystem2019}.
-The DMK dataset is derived from the same historical surveys as the AR5 dataset, and peat depth classes are: \textless{} 1 m (\emph{shallow}), \textgreater{} 1 m (\emph{deep}), and \emph{unknown}.
-Surveyors generally assigned peat depth classes to polygons of at least 0.5 ha, although delineating polygons down to 0.2 ha was allowed if peat depth showed a ``particularly marked difference'' \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
-We rasterized the peat depth class attribute to our 10 m grid.
-
-\subsection{Peat depth sample selection}
-
-The places (10 m raster cells) we chose to measure peat depth were sampled from mire areas in the AR5 dataset, and optimized for training a Random Forest (RF) model of peat depth \citep{brusSamplingDigitalSoil2019}.
-Broadly, we aimed for a sample that was representative of the predictor space defined by the most important predictors of peat depth \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
-A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population, while avoiding spurious ones \citep{brusSamplingDigitalSoil2019}.
-Although we implemented the approach differently for Skrimfjella (in 2020) than for Ørskogfjellet (in 2023), the objective was the same.
-
-We made several point measurements of peat depth within the sampled raster cells (three if by probing only, more if by GPR as well).
-The point measurements (described below) were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
-
-\subsubsection{Skrimfjella}
-
-At Skrimfjella, we used the \emph{eSample} function in the \emph{iSDM} R package (v1.0) to stratify our sample across elevation, slope, and potassium concentration.
-This function defines the environmental space as a two-dimensional convex hull around the PCA-ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest \citep{hattabUnifiedFrameworkModel2017}.
-We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and \emph{eSample} returned 105 raster cells.
-
-We also measured peatland occurrence (binary variable) in a separate sample at Skrimfjella.
-We measured peatland occurrence because the AR5 dataset is known to underestimate peatland extent \citep[especially in forests,][]{brynLandCoverNorway2018}, and because airborne radiometrics may help identify unmapped peatland \citep{gatisMappingUplandPeat2019, olearyDigitalSoilMapping2022}.
-The 10 m cells were sampled from those outside AR5 mire and with slope \textless{} 20°.
-Following the same procedure, \emph{eSample} returned 106 raster cells from this population.
-
-\subsubsection{Ørskogfjellet}
-
-At Ørskogfjellet, we first determined a minimal sample size that would adequately capture the slope and radiometric properties (potassium, thorium, uranium, and total count) of the entire AR5 mire area \citep{sauretteDivergenceMetricsDetermining2023}.
-Specifically, we identified an elbow point in a curve of similarity between sample and population \citep{maloneMethodsImproveUtility2019}.
-For a sequence of sample sizes (50--500) \citep[ten replicates each, drawn by conditioned latin hypercube sampling,][]{minasnyConditionedLatinHypercube2006, roudierClhsPackageConditioned2011}, we calculated the mean Kullback--Leibler divergence between sample and population distributions \citep{maloneMethodsImproveUtility2019, sauretteDivergenceMetricsDetermining2023}.
-Then we fitted an asymptotic regression of mean divergence on sample size, and found that the curve reached 95 \% of the fitted asymptote at a sample size of 160.
-
-To choose 160 locations, we performed feature space coverage sampling, implemented using the \emph{kmeans} function in base R and the \emph{rdist} function in the the \emph{fields} package (v14.2).
-Feature space coverage sampling chooses locations that are closest to cluster centers in standardized predictor space \citep{brusSamplingDigitalSoil2019}.
-This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
-Feature space coverage sampling works best when all dimensions are important predictors of the outcome \citep{wadouxSamplingDesignOptimization2019}, and we used the same five predictors that we used to choose sample size: slope and four radiometrics.
-
-We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
-Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
-To adjust, we first restricted the sampling population to AR5 mire areas that were within an arbitrary cost distance of publicly accessible roads.
-Cost distance was calculated using GRASS's \emph{r.walk} function, with friction costs defined by AR5 land classes \citep{GRASSv8-2}.
-After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
-Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 \% (compared to the ideal sample), but distance in our sample was still only 46 \% of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
-
-\subsection{Depth measurements}
-
-We measured peat depths at Skrimfjella in August 2020 and at Ørskogfjellet in August 2023.
-At both sites we used manual probing as the primary method of measuring peat depth, and ground-penetrating radar (GPR) as a secondary method.
-That is: peat depth was always measured by probing, and also by GPR in a subset of cells that partly overlapped with the probed cells
-We chose this approach because probing is a fast and reliable method for point measurements, while GPR can provide higher lateral density of data in the same amount of time \citep{parryEvaluatingApproachesEstimating2014}.
-Probed depths serve to calibrate GPR measurements when calibration by common midpoint survey is not possible -- as was the case with our fixed radar antennas -- so the methods are complementary.
-
-\subsubsection{Peat probing}
-
-We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real time kinematic (Ørskogfjellet) global navigation satellite system (GNSS) receivers.
-We dampened the effect of outlying measurements by probing three times at each location \citep{parryEvaluatingApproachesEstimating2014}, at the vertices of a centered triangle with 2 m (Skrimfjella) or 4.5 m (Ørskogfjellet) sides.
-We used changes in resistance to indicate the base of the peat column.
-Probe locations were adjusted up to 20 cm if the base of the peat column seemed to be blocked by an obvious artifact, like a buried rock.
-Where the peat column was deeper than the extendable probe could be manually inserted and extracted by a pair of operators, we recorded a right-censored result (one at Skrimfjella, five at Ørskogfjellet).
-
-We also extracted a set of existing depth measurements for Ørskogfjellet from a paper map made by the Norwegian Soil and Mire Company in 1984.
-The map presents 44 borehole depths (in decimeters) across a 9 ha peatland area.
-We georeferenced the map and digitized the borehole locations and depths.
-
-For our sample of peatland occurrence at Skrimfjella, we recorded the presence or absence of peatland at the cell center -- primarily by digging and examining the top 20 cm of soil.
-We judged whether the soil was a peat soil based on its density, texture, and color, as well as the presence or absence of mire vegetation.
-Although peat soil is strictly defined by organic content (which we did not analyze), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland \citep{minasnyMappingMonitoringPeatland2024}.
-
-\subsubsection{Ground-penetrating radar}
-
-We performed GPR surveys in three subjectively chosen peatlands at Skrimfjella and in areas with a high density of sampled raster cells at Ørskogfjellet.
-We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with a GNSS-enabled control unit connected to a 500 MHz shielded antenna mounted in a plastic sledge (transmitter--receiver separation 0.18 m, trace frequency 10 Hz).
-For some transects at Ørskogfjellet we substituted a 100 MHz Malå rough terrain antenna (transmitter--receiver separation 2.2 m, trace frequency 5 Hz), because the lower frequency antenna gives greater penetration depth.
-In all cases the system was towed by a walking GPR operator.
-
-GPR traces were recorded along zigzag (Skrimfjella) or snaking (Ørskogfjellet) transects.
-At Ørskogfjellet, transects were predetermined to pass through the centers of sampled raster cells, and we marked these precisely with flags to guide the GPR operator.
-A GPR records the time taken for a radio wave to travel from the transmitter to a reflector and back to the receiver, and the velocity of the wave varies with properties of the peat column.
-Therefore, wave velocity has to be calibrated to convert travel time to peat depth, and we probed peat depth along the transects.
-
-We processed the GPR data with Reflex2DQuick (v.3.0; Skrimfjella) or REFLEXW (v.8.5; Ørskogfjellet) software (Sandmeier Scientific Software, Germany).
-We applied a time-zero correction, a dewow filter, and a gain filter based on observed energy decay.
-With Ørskogfjellet data, we also applied a bandpass filter and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennas.
-The latter is important for the rough terrain antenna, where the antenna separation is comparable to typical peat depths.
-From the processed radargrams, we picked travel times from strong reflectors that we interpreted as the base of the peat column.
-
-We used picks near probed depths to calibrate wave speed velocity -- separately for each site.
-Calibration data were created by matching marked trace locations to a corresponding depth probe (Skrimfjella), or by a spatial join that identified interpreted traces and depth probes within 2 m of each other (Ørskogfjellet).
-We had sufficient calibration point density to avoid bias in wave velocity as a major source of error \citep{rosaDeterminingNumberManual2009}: 46 calibration points along 3.5 km of interpretable traces at Skrimfjella, and 78 along 7.8 km at Ørskogfjellet.
-We fitted site-specific linear regressions of probed depth on picked travel time, with the intercept fixed at zero, to estimate wave velocities.
-Notwithstanding a few outlying points, our regressions showed good fits and the resulting velocities are within the range reported for peat \citep{parsekianUncertaintyPeatVolume2012}: \unit{0.0387\,m\,ns^{-1}}, \(R^2 = 0.874\) at Skrimfjella and \unit{0.0427\,m\,ns^{-1}}, \(R^2 = 0.946\) at Ørskogfjellet.
-Finally, we used these two wave velocities to convert the travel times of all picks to calibrated peat depths.
-In total, the GPR surveys produced 48579 point measurements of peat depth at Skrimfjella and 32653 at Ørskogfjellet.
-
-We also used a set of existing GPR depth measurements from Ørskogfjellet, commissioned and provided to us by the Norwegian Public Roads Administration.
-These data were collected in 2020 and 2021 with a dual channel system (70 MHz and 300 MHz; ImpulseRadar AB, Sweden), connected to GNSS with CPOS correction.
-For the work presented here, we used a total of 403440 interpreted and calibrated traces along 7.4 km of transects -- discarding data where multiple depths were interpreted for the same location.
-
-\subsection{Predictive models of peat depth}
-
-\subsubsection{Modelling approach}
-
-We used Random Forests (RF) to predict peat depth at both sites.
-RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees \citep{breimanRandomForests2001}.
-We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in DSM applications than other algorithms \citep{beguinPredictingSoilProperties2017, nussbaumEvaluationDigitalSoil2018, lamichhaneDigitalSoilMapping2019}.
-It is suited for use on relatively small training datasets and its predictions can be interrogated to learn about predictor importance \citep{khaledianSelectingAppropriateMachine2020}.
-Evaluating variable importance in a maximally predictive model aligns with the aim of this study.
-
-RF by itself is not a spatial model, and it will only predict spatial structure in the outcome to the degree that the structure is captured by predictors.
-We considered using regression kriging -- a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically interpolated surface of RF residuals \citep{henglGenericFrameworkSpatial2004}.
-The spatial component in regression kriging often improves map accuracy compared to a non-spatial model \citep{beguinPredictingSoilProperties2017, lamichhaneDigitalSoilMapping2019, mollaMachineLearningGeostatistical2023}, but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations \citep{henglGenericFrameworkSpatial2004, szaboMappingSoilHydraulic2019, takoutsingComparingPredictionPerformance2022}.
-If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.
-We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that (non-spatial) RF rather than regression kriging was justified at both sites.
-
-We implemented models in the \emph{tidymodels} framework in R \citep{kuhnTidymodelsCollectionPackages2020}, with the \emph{ranger} R package for RFs \citep[v0.16,][]{wrightRangerFastImplementation2017}.
-RFs were fit with 1000 trees, minimum node size of 5, and the number of predictors randomly sampled at each split was the square root of the total number of predictors (\emph{ranger} default).
-We did not tune these hyperparameters because RFs are relatively insensitive to tuning \citep{probstHyperparametersTuningStrategies2019}, and because it would require nested spatial cross-validation to prevent data leakage \citep{schratzHyperparameterTuningPerformance2019}.
-
-\subsubsection{Model performance}
-
-For both sites, we compared the performance of models with five different configurations of predictors.
-Specifically, we trained models with:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \emph{DMK} (2 variables)
-\item
-  \emph{terrain} (21 variables)
-\item
-  \emph{terrain + DMK} (23 variables)
-\item
-  \emph{terrain + radiometric} (25 variables)
-\item
-  \emph{all predictors} (27 variables)
-\end{enumerate}
-
-These different configurations simulate different scenarios of data availability.
-Comparing the different configurations allowed us to isolate the added value of each of the predictor groups.
-We did not explore configurations comprising radiometrics without terrain, because LiDAR terrain surveys typically precede airborne radiometric surveys.
-The models with only DMK peat depth class were simple linear models rather than RFs, and served to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth, calibrated on the same data.
-
-We used a spatial cross-validation scheme to evaluate model performance \citep{wadouxSpatialCrossvalidationNot2021, meyerMachineLearningbasedGlobal2022}.
-To set the folds we used k-Means Nearest Neighbor Distance Matching (kNNDM), which aims to mimic the spatial prediction task that is defined as the goal \citep{linnenbrinkKNNDMCVKfold2024}.
-In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
-That way, the spatial separation between folds is similar to the separation between training and prediction locations -- which increases the quality of the map accuracy estimate \citep{linnenbrinkKNNDMCVKfold2024}.
-For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation \citep{wadouxSpatialCrossvalidationNot2021}.
-We implemented the kNNDM with the \emph{CAST} R package \citep[v1.0,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
-From the cross-validation we quantified mean absolute error (error magnitude, original scale), R\textsuperscript{2} (explained variation, standardized scale), and Lin's concordance correlation coefficient (error magnitude and explained variation, standardized scale).
-
-DSM products have much more value when their predictions are accompanied by uncertainty estimates, and all DSM should strive to assess uncertainty \citep{arrouaysImpressionsDigitalSoil2020, wadouxMachineLearningDigital2020} and evaluate uncertainty estimates \citep{heuvelinkSpatialStatisticsSoil2022}.
-Therefore, we produced prediction intervals with quantile regression forests \citep{meinshausenQuantileRegressionForests2006}, and used the same spatial cross-validation to evaluate the prediction interval coverage probability \citep{shresthaMachineLearningApproaches2006}.
-The quantile regression forests were trained with predictor configuration that showed the highest performance at each site (under the assumption that these models would be put into production) and we extracted 90 \% prediction intervals.
-
-\paragraph{Peatland extent}
-
-While the primary aim of the DSM was to predict peat depth within AR5 mires (where peat depth is supposed to be \textgreater{} 30 cm), we also tested whether our mire-trained models of peat depth could identify peatlands outside of AR5 mires.
-For both sites we performed an evaluation with our (cell-aggregated) peat depth measurements, of which about one-fifth were outside of AR5 mires.
-For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on AR5 mire locations (in the training folds), and evaluated them only on (test fold) locations classified in AR5 as something other than mire.
-From the cross-validation we quantified mean absolute error.
-For Skrimfjella -- where we measured peatland occurrence separately from depth -- we also trained a model on all peat depth measurements (AR5 mire or not) and then used the independent occurrence dataset to evaluate this model.
-Specifically, we calculated the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
-This analysis treats the model's prediction of peat depth as a rank index of peatland likelihood.
-Since the purpose was to test the models' ability to uncover unmapped peatland (where DMK peat depth class is undefined), we evaluated models with two predictor configurations: \emph{terrain} or \emph{terrain + radiometric}.
-
-\subsubsection{Model interpretation}
-
-We quantified global variable importance (predictor influence across all locations) and examined partial dependence plots (curves of fitted relationships) for the best-performing predictor configuration at each site.
-Both are useful for understanding the mechanisms behind the model's predictions and the roles of the predictors in the model.
-
-For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration -- because correlation between predictors degrades variable importance metrics \citep{stroblConditionalVariableImportance2008, biauRandomForestGuided2016} and can produce misleading visualizations of predictor--outcome relationships \citep{biecekExplanatoryModelAnalysis2021, dwivediExplainableAIXAI2023}.
-Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation coefficient above 0.7.
-Thus, highly correlated sets of variables are represented by a single variable for the purposes of model interpretation.
-
-We calculated variable importance with the \emph{vip} R package (v0.4), by three different methods: \emph{FIRM}, \emph{permutation}, and \emph{Shapley} \citep{greenwellVariableImportancePlots2020}.
-\emph{FIRM} values measure the flatness of the partial dependence plot, \emph{permutation} values measure the decrease in model performance when the predictor is permuted, and \emph{Shapley} values are aggregated from local, game-theoretical measures of variable importance \citep{greenwellVariableImportancePlots2020}.
-\emph{Permutation} values were obtained from ten iterations, with root mean square error as the performance measure.
-
-We calculated partial dependence with the \emph{pdp} R package \citep[v0.8,][]{greenwellPdpPackageConstructing2017}.
-For the six most important variables, we plotted both partial dependence and individual conditional expectation, to show the average effect of the predictor on the outcome and the variation in the effect across observations, respectively \citep{goldsteinPeekingBlackBox2015}.
-Non-parallel individual conditional expectation lines indicate the presence of interactions between predictors.
-
-\section{Results}
-
-Our point measurements of peat depth (all sources) produced aggregated depths for 372 cells at Skrimfjella (2.4 \% of AR5 mire area) and 1878 cells at Ørskogfjellet (1.2 \% of AR5 mire area).
-Roughly 80 \% of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
-Coverage of DMK peat depth classes was higher at Ørskogfjellet than at Skrimfjella (79 \% versus 27 \% of cells), and at Ørskogfjellet the \emph{deep} and \emph{shallow} classes showed a larger difference in measured depth.
-Overall mean peat depths were similar at Skrimfjella and Ørskogfjellet: 119 cm and 126 cm, respectively.
-
-\begin{table}[tbp]
-\caption{Mean peat depth (cm) in \unit{10\,m} cells at Skrimfjella and Ørskogfjellet. The cells are also shown stratified by AR5 land class and DMK peat depth class.}
-\begin{tabular}{llllllll}
-\hline
-                    &                            & \multicolumn{3}{c}{Skrimfjella} & \multicolumn{3}{c}{Ørskogfjellet} \\ \cline{3-8} 
-                    &                            & n        & \%      & depth      & n          & \%      & depth      \\ \hline
-all measured cells  &                            & 372      & 100     & 119        & 1878       & 100     & 126        \\
-AR5                 & Agricultural               &          &         &            & 21         & 1       & 30         \\
-                    & Forest                     & 52       & 14      & 65         & 272        & 15      & 71         \\
-                    & Open upland                & 19       & 5       & 98         & 134        & 7       & 39         \\
-                    & Peatland                   & 301      & 81      & 130        & 1451       & 77      & 145        \\
-DMK                 & deep (\textgreater 100 cm) & 94       & 25      & 100        & 659        & 35      & 219        \\
-                    & shallow (\textless 100 cm) & 6        & 2       & 50         & 838        & 45      & 82         \\
-                    & unknown                    & 272      & 73      & 127        & 381        & 20      & 60         \\ \hline
-\end{tabular}
-\label{tab:depthsByClass}
-\end{table}
-
-\subsection{Model performance}
-
-None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \ref{fig:modelMetrics}).
-For Skrimfjella, the best model achieved a concordance correlation coefficient of 0.30, an R\textsuperscript{2} of 0.34, and a mean absolute error of 60 cm.
-For Ørskogfjellet, the concordance correlation coefficient was 0.39, R\textsuperscript{2} was 0.33, and mean absolute error was 56 cm, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
-These values were derived from kNNDM spatial cross-validation with 20 folds at Skrimfjella and 10 folds at Ørskogfjellet.
-
-\begin{figure}
-\centering
-\includegraphics{figures/modelmetrics.pdf}
-\caption{\label{fig:modelMetrics}Performance of peatland depth models with different predictor configurations, evaluated via spatial cross-validation.
-Parentheses denote the number of variables in each predictor configuration, and point estimates are shown +/- their standard error.}
-\end{figure}
-
-Although the difficulty of predicting peat depth caused prediction intervals to be wide, these uncertainty estimates were well calibrated.
-At Skrimfjella, the prediction interval coverage probability was 91 \%, and at Ørskogfjellet it was 84 \% (both compared to the target value of 90 \%).
-Observations outside of the prediction intervals were evenly distributed across each of the study areas.
-
-For Skrimfjella, the best predictor configuration was \emph{all predictors}, followed closely by \emph{terrain + radiometric}.
-The performance gap between the \emph{terrain + DMK} configuration and the \emph{terrain} configuration was similarly small.
-Compared to the \emph{terrain} configuration, the \emph{terrain + radiometric} configuration improved concordance correlation by 0.04, R\textsuperscript{2} by 0.04, and mean absolute error by 1 cm.
-\emph{DMK} was a very poor predictor of peat depth even though it was calibrated to measured depths, with a concordance correlation coefficient of 0.0077.
-
-For Ørskogfjellet, the best predictor configuration was \emph{terrain + DMK}, followed by \emph{terrain}.
-Adding radiometric predictors to these configurations worsened model performance, especially in terms of concordance correlation and R\textsuperscript{2}.
-By itself, \emph{DMK class} produced a concordance correlation coefficient of 0.17 (compared to 0.008 at Skrimfjella), but the worst mean absolute error of any model at either site: 77 cm.
-
-The best models at both sites overpredicted shallow peats and strongly underpredicted very deep peats (Fig. \ref{fig:calPlots}).
-The mean error (bias) of these models was 10 cm at Skrimfjella and -4 cm at Ørskogfjellet.
-
-\begin{figure}
-\centering
-\includegraphics{figures/calibration_plots.pdf}
-\caption{\label{fig:calPlots}Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions from spatial cross-validation.
-Points are transparent to show overlap.
-Blue lines are local polynomial regressions and the red dashed line in each panel shows the 1:1 line.
-Marginal distributions (top) are shaded by quartile.}
-\end{figure}
-
-\subsubsection{Peatland extent}
-
-When we tested how well models extrapolated from within to outside of AR5 mire areas, we found that the models produced worse mean absolute error than just assuming a constant 30 cm depth (Fig.
-\ref{fig:modelMetricsExtrapolation}).
-With the independent occurrence data at Skrimfjella, we found that neither the \emph{terrain} nor \emph{terrain + radiometric} configurations were able to discriminate between peat presence and absence (area under the curve of the receiver operating characteristic 0.44 and 0.52 respectively, where 0.5 indicates random guessing).
-
-\begin{figure}
-\centering
-\includegraphics{figures/modelmetrics-extrapolation.pdf}
-\caption{\label{fig:modelMetricsExtrapolation}Performance of models that extrapolate from training data inside of mapped mires to test data outside of mapped mires (AR5 dataset), evaluated via spatial cross-validation.
-Parentheses denote the number of variables in each model, and point estimates are shown +/- their standard error.}
-\end{figure}
-
-\subsection{Model interpretation}
-
-For the purpose of model interpretation, the \emph{all predictors} configuration for Skrimfjella was reduced from 27 variables to 11 non-collinear variables, by removing one of the variables in each highly-correlated pair.
-Similarly, the \emph{terrain + DMK} configuration for Ørskogfjellet was reduced from 23 variables to 11 non-collinear variables.
-
-\subsubsection{Variable importance}
-
-At both sites, elevation and Multi-Resolution Valley Bottom Flatness were important predictors (Fig. \ref{fig:varImp}).
-At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet elevation was more important than Multi-Resolution Valley Bottom Flatness.
-DMK was also important -- the shallow class in particular -- but only at Ørskogfjellet.
-Some realizations of the hydrological predictors Topographic Wetness Index and Depth-to-Water showed considerable importance, while others showed little -- with no consistency between sites.
-For example, TWI5m, DTW40000, and DTW2500 rounded out the top five most important variables at Skrimfjella, while TWI20m and TWI50m were least important, other than DMK.
-At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices slope, Terrain Ruggedness Index, Topographic Position Index, and roughness.
-The radiometric predictor uranium ground concentration showed moderate importance at Skrimfjella.
-
-\begin{figure}
-\centering
-\includegraphics{figures/variable_importance.pdf}
-\caption{\label{fig:varImp}Global variable importance in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), as measured by three different metrics: Shapley values, permutation importance, and the Feature Importance Ranking Measure \citep{greenwellVariableImportancePlots2020}.
-Variables removed due to collinearity are shown to the right of the variable with which they are most correlated.}
-\end{figure}
-
-\subsubsection{Partial dependence}
-
-Many of the most important predictors in the best performing models showed non-monotonic effects on peat depth (Fig. \ref{fig:pdps}).
-At Ørskogfjellet for example, increasing elevation was predictive of deeper peat up to about 75 m above sea level, after which a further increase was predictive of shallower peat.
-At Skrimfjella the partial dependence on elevation had the opposite shape but covers a higher elevation range, with the shallowest peats predicted around 350 m above sea level.
-TWI50m at Ørskogfjellet and DTW4000 and uranium ground concentration at Skrimfjella were other predictors that showed considerable fluctuations in their predictive effects on peat depth.
-The radiometric predictor in particular displayed an idiosyncratic effect, with a marked dip in predicted depth at intermediate values of uranium ground concentration.
-On the other hand, the partial effects of some important predictors were more straightforward.
-The partial dependence on Multi-Resolution Valley Bottom Flatness was quite similar across sites, with the deepest peats predicted in the very flattest valley bottoms.
-Also, TWI5m and DTW2500 at Skrimfjella showed monotonically positive and negative predictive effects, respectively.
-
-Individual conditional expectation lines indicated some interactions between predictors (Fig. \ref{fig:pdps}).
-For example, the magnitude of the increase in depth with elevation that the model expected at Ørskogfjellet was different for different observations; some depth predictions increased by only 40 cm while others increased by more than 100 cm, over the same elevation gain.
-Similarly, individual conditional expectation lines of uranium ground concentration at Skrimfjella were non-parallel, with some locations showing monotonically increasing peat depth predictions with uranium concentration (unlike the average effect).
-Nevertheless, most individual conditional expectation lines were generally parallel -- indicating that the average effects of the predictors were good representations of their overall effects.
-
-\begin{figure}
-\centering
-\includegraphics{figures/partial_dependence.pdf}
-\caption{\label{fig:pdps}Partial dependence plots of the six most important variables in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b).
-The average effect of the predictor on the outcome (red line) overlays the variation in the effect across observations (grey lines).
-The variables' training data distributions are indicated with rug plots along the x-axes.}
-\end{figure}
-
-\section{Discussion}
-
-\subsection{Can we improve Norway's peat depth maps?}
-
-\subsubsection{Remotely sensed variables are weak (but not useless!) predictors of peat depth}
-
-Our ability to predict peat depth in the study sites based on terrain and radiometric data was limited.
-Mean absolute errors of 60 and 56 cm at the two sites --- relative to mean depths of 119 and 126 cm --- illustrate the practical limitations of these maps.
-Since any given 10 m cell will miss by about 60 cm, applications requiring detailed peat depth in a small area (e.g.~\textless{} 1 ha) would benefit from measuring depth on the ground rather than relying on the DSM alone.
-
-Ground-based measurements also have uncertainty \citep{parryEvaluatingApproachesEstimating2014}, but our data show that this error was smaller than the variation in peat depth over short distances (e.g.~100 m).
-Independent probing and GPR measurements up to 2 m apart were differed by an average of 29--42 cm (Fig. \ref{fig:GPRwavevelocity}).
-By comparison, the average variation among measurements 100 m apart was 70--106 cm (Fig. \ref{fig:semivariograms}).
-We also note that cell-level measurement error that is correlated with predictors (e.g.~underestimated depth at high Multi-Resolution Valley Bottom Flatness), will go undetected by cross-validation and does not account for the limited accuracy of our models.
-Ultimately, the effect of measurement error falls mostly outside the scope of DSM, since it can only be evaluated with higher-quality, independent data.
-
-Although model performance was limited, we improved on the best available map of peat depth (DMK depth class), which is based on field measurements only.
-This highlights the general value of remotely sensed data, whose complete coverage can improve maps even when their association with the variable of interest is weak \citep{mulderUseRemoteSensing2011}.
-Since remotely sensed data are widely available, improvements to soil maps as shown here are low hanging fruit.
-This point is recognized and reflected in the rise of DSM \citep{minasnyDigitalMappingPeatlands2019}.
-
-DMK peat depth classes were a worse predictor of peat depth than our models even though we calibrated them with the same data (Fig. \ref{fig:modelMetrics}).
-That is: we gave DMK and the DSM equal footing for a fair comparison.
-If we had taken the DMK peat depth classes at face value and assumed depths according to their class definitions (\textless{} 1 m, \textgreater{} 1 m), they would have performed worse and the advantage of the DSM would be greater.
-The advantage of the DSM was not large in absolute terms (9 cm and 21 cm improvements in mean absolute error), but it explained much more of the variation in depth (improvements in R\textsuperscript{2} of 0.16 and 0.19).
-We attribute this result to the poor spatial and thematic resolution of DMK peat depth, which precludes a robust correlation with peat depths varying from 0--8 m at fine spatial scales.
-
-The performance of our terrain- and radiometric-based maps could have been improved with a spatially explicit mapping approach like regression kriging.
-By ignoring spatial autocorrelation in peat depth (i.e.~the n in \emph{scorpan}), we have not extracted all of the information about peat depth in the study area out of the training data.
-If we were to put our DSM approach into production for published maps, we would harness the spatial component, but the actual maps for Ørskogfjellet and Skrimfjella are not of primary interest in this study.
-Moreover, residuals of the RF predictions showed weak spatial structure at Skrimfjella, and only up to a range of 150 m at Ørskogfjellet.
-Therefore, the improvement from regression kriging would be small overall and limited to parts of the maps close to measurements.
-
-\subsubsection{Similar error but less explanatory power in Norwegian peatlands}
-
-Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth but generally produced better or comparable error magnitude (R\textsuperscript{2} vs.~mean absolute error or root mean square error; \citet{wadouxIntegratedApproachEvaluation2022}).
-It is important to keep in mind that differences in peat depth distributions, spatial scales, and evaluation methods make direct performance comparisons precarious.
-More standardized reporting would help but not eliminate this consideration.
-For example, R\textsuperscript{2} is sensitive to high leverage, extreme values, so it will evaluate a right-skewed distribution differently than a symmetrical distribution.
-We evaluated our models with respect to the explicit purpose of creating peat depth maps across the study areas, but not all studies tailored evaluation to match an explicitly formulated problem \citep{milaNearestNeighbourDistance2022}.
-
-Gatis et al.
-\citeyearpar{gatisMappingUplandPeat2019} used similar predictors and the same spatial grain, finding a much stronger relationship between predicted and observed peat depth (\(R^2 = 0.68\)).
-Although their random evaluation data partition could make performance estimates too optimistic \citep{robertsCrossvalidationStrategiesData2017, wadouxSpatialCrossvalidationNot2021}, the confounding effect of spatial structure is probably small because they used linear regression and few predictors.
-Their model had limited opportunity to overfit to the spatial structure in peat depth.
-The most salient difference in Gatis et al. \citeyearpar{gatisMappingUplandPeat2019} compared to our study is the character of the study area.
-They study a flatter area with a higher proportion of peatland cover, and their peatland is primarily blanket bog.
-The peat surface of a blanket bog is tied more closely to its underlying topography than the peat surface of raised bogs or fens \citep{lindsayPeatlandMireTypes2016}, which may make depths in blanket bogs easier to predict.
-The predominance of fens and smaller peatland extent may have contributed to worse performance in our study.
-
-Marchant \citeyearpar{marchantUsingRemoteSensors2021} examined a subset of the area studied in Gatis et al.
-\citeyearpar{gatisMappingUplandPeat2019} at 100 m resolution, using splines to relax linearity between radiometry/terrain and peat depth.
-He found that potassium ground concentration alone predicted peat depth with much higher concordance than our models (concordance correlation coefficient = 0.76) and that elevation alone produced comparable performance (concordance correlation coefficient = 0.27).
-The root mean square error from these univariate models was 46--68 cm (cf.~78 and 75 cm for Skrimfjella and Ørskogfjellet).
-
-Koganti et al. \citeyearpar{kogantiMappingPeatDepth2023} had a peat depth distribution and predictors similar to ours, but at much smaller spatial grain and extent.
-Their training and validation points are closer than the range of spatial autocorrelation in peat depth, so our results are best compared to their non-spatial models.
-They accounted for more variability in peat depth (adjusted \(R^2 = 0.71\)) but had larger errors (RSME = 110 cm).
-Koganti et al.'s \citeyearpar{kogantiMappingPeatDepth2023} linear regression models produced negative predictions, and it is unclear whether the values quoted above include these.
-If we disregard the negative predictions, their model showed the same pattern as ours in overpredicting shallow peats and underpredicting deep peats, although their underprediction was less severe.
-An important difference between Koganti et al.
-\citeyearpar{kogantiMappingPeatDepth2023} and our study (besides spatial scale) is that they measured radiometrics on the ground, rather than using airborne survey data.
-Thus, the footprint of their detector was much smaller and they could capture variation in radiation at finer scales.
-
-\subsubsection{Asymmetries in depth predictions for land use planning and carbon accounting}
-
-Our models erred most for the deepest peats (Fig. \ref{fig:calPlots}).
-Where overprediction occurred, the error was smaller.
-This is not unexpected for the right-skewed distributions of peat depth, but it has management implications.
-Identifying the deepest peats will require additional field work in candidate areas, which could be defined by an upper quantile of predicted depth.
-Map users should not trust the maps to identify all large carbon stocks, but they can trust that identified large stocks really are large.
-That makes the map more suited for ``red-lighting'' than ``green-lighting'' peatland conversion, for example.
-Where it does not prohibit conversion (i.e.~predicts shallow peat), a ground survey should be done before conversion is allowed (assuming depth is the only consideration).
-Although this recommendation aligns with the precautionary principle, here we make it on technical grounds based on the maps' characteristics.
-
-\subsubsection{Depth predictions do not necessarily extrapolate to peat extent}
-
-That we measured more than 30 cm of peat in areas not mapped as mire was expected, because AR5 underrepresents peatland extent by about one third \citep{brynLandCoverNorway2018}.
-For that reason, maps of peatland extent also need revising.
-Unfortunately, our models did not predict peat occurrence outside AR5 mires.
-In these areas, assuming a constant 30 cm of peat (the depth threshold used in AR5) produced less error than our models (Fig. \ref{fig:modelMetricsExtrapolation}).
-Moreover, the performance gap would likely have been larger had our evaluation data been representative of all areas not classified as mire in AR5.
-Instead they were concentrated near AR5 mires.
-Most importantly, the independent occurrence data at Skrimfjella showed the model completely failing to predict occurrence, and these data were suited for testing this ability.
-Although the Ørskogfjellet model may have done better than the Skrimfjella model on a similar independent evaluation set (since it predicted depth better), the improvement would probably be small.
-
-Why wasn't peat occurrence predicted well?
-The models were trained on fundamentally different populations of locations (\textasciitilde80 \% AR5 mire in training vs.~0 \% in prediction), so it is not surprising that the associations they learned did not transfer well.
-Moreover, the \textasciitilde20 \% of training data from outside AR5 mires were incidentally collected (near AR5 mires) and not representative of other land cover classes.
-In short, the models were blind to the fact that most of both study areas have no peat.
-
-Peatland extent is probably best mapped using different remotely sensed data than we used here \citep{bakkestuenDelineationWetlandAreas2023}, and this study's purpose was not to map extent.
-Nevertheless --- as long as the relevant peatland definition includes a depth component --- depth predictions (or predictors thereof) should help delineate peatland extent \citep{olearyDigitalSoilMapping2022, beamishDetailedMappingPeat2024}.
-We return to this point in our discussion of implications for digital soil mapping.
-
-\subsection{Which variables predict peat depth?}
-
-\subsubsection{Airborne radiometrics do not predict Norwegian peat depth}
-
-Radiometric data had no predictive value at Ørskogfjellet, while at Skrimfjella they had minor influence in a relatively weak model.
-The bedrock is more homogeneous at Ørskogfjellet than at Skrimfjella, so uneven radiogenesis is not a viable explanation for the differences between sites nor the poor performance in general \citep{beamishEnvironmentalRadioactivityUK2014, reinhardtGammaraySpectrometryVersatile2019}.
-To the degree that radiometrics had predictive value at Skrimfjella, it appears that they were most valuable near the extremes of the depth distribution, since their inclusion improved R\textsuperscript{2} more than mean absolute error.
-All four variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value.
-This contrasts with Koganti et al. \citeyearpar{kogantiMappingPeatDepth2023}, who found that radiometric total count was a much better predictor than potassium ground concentration.
-
-We suspect that the primary reason for the poor predictive value of the radiometric data was the large footprint of the detector in the airborne survey.
-With an average flight altitude of 75 m, less than half of the radiation reaching the detector comes from inside the 100 m diameter circle directly below it \citep{beamishEnhancingResolutionAirborne2016, beamishDetailedMappingPeat2024}.
-The rest of the measured activity integrates a much wider area.
-For comparison, empirical variograms of peat depth at Skrimfjella and Ørskogfjellet showed no spatial autocorrelation beyond 50 and 75 m (among GPR data) or 110 and 230 m (among 10 m cells).
-Basically, the airborne radiometric data will not capture large variation over short (\textless{} 100 m) distances; the instrument's field of view has a large smoothing effect on the data \citep{beamishEnhancingResolutionAirborne2016, reinhardtGammaraySpectrometryVersatile2019}.
-Different landforms and the changes they cause in the geometry between the radioactive source and the detector can also distort airborne measurements \citep{reinhardtGammaraySpectrometryVersatile2019}.
-
-Studies comparing airborne and ground radiometric surveys confirm that they are poorly correlated in low-activity areas like peatlands \citep{kockComparisonAirborneTerrestrial2011, karjalainenComparisonTwoGammaray2025}.
-Karjalainen et al. \citeyearpar{karjalainenComparisonTwoGammaray2025} found that ground-based measurements predicted peat depth better than airborne measurements.
-Nevertheless, a large radiometric footprint did not prevent strong associations with peat depth in Gatis et al.
-\citeyearpar{gatisMappingUplandPeat2019} and Marchant \citeyearpar{marchantUsingRemoteSensors2021}, perhaps because the extensive blanket bog landscape in these studies has more gradual changes in depth \citep{lindsayBogsEcologyClassification1995}.
-We are unsure whether short-range depth changes explain the weak associations that Siemon et al. \citeyearpar{siemonAirborneElectromagneticRadiometric2020} found in a large raised bog.
-
-Weather conditions varied during the Ørskogfjellet radiometric survey and affected its data \citep{ofstadHelicopterborneMagneticRadiometric2015}.
-Thus, uneven snow cover and air moisture may also have masked the soil signal in these data.
-
-We do not believe that the poor predictive value of the radiometric data in this study was caused by fully attenuated radioactivity --- at least not in large part.
-The RF algorithm's flexibility means that radiometrics could be used for shallower peats if they provided predictive value for that part of the depth distribution, but there is no evidence of that in our results.
-In the partial dependence plot of uranium ground concentration at Skrimfjella, the expected negative relationship between depth and uranium ground concentration cannot be found by ignoring the left (highly attenuated) side of the distribution.
-About a quarter of the peats in our study were less than a meter deep, and full attenuation is unlikely for these \citep{beamishGammaRayAttenuation2013}.
-
-Although we do not believe full attenuation is the primary reason for poor performance in our study, it may limit peat depth mapping under other circumstances.
-First-principle calculations suggest that radiation should be 90 \% attenuated after about 50--60 cm of typical, wet peat or 85 cm of unnaturally dry peat \citep{beamishGammaRayAttenuation2013, beamishDetailedMappingPeat2024}, and some field tests support these values \citep{billenEignungGammaspektrometrieKartieren2015}.
-It is remarkable that particular studies detected radiation differences up to several meters deep \citep{gatisMappingUplandPeat2019, kogantiMappingPeatDepth2023}, but these may be the exceptions rather than the rule.
-Perhaps relatively deeper water tables in these study sites \citep[blanket bog, drained fen,][]{pricePeatlandRestorationHydrology2016} contributed to better penetration.
-
-\subsubsection{Terrain-based variables can predict peat depth}
-
-At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables (Fig. \ref{fig:varImp}).
-Elevation was the most important predictor at both sites, and peat depth showed non-monotonic responses to changes in elevation (Fig. \ref{fig:pdps}).
-We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g.~via climate) to explain the observed patterns.
-For example, the increase in peat depth from 350 to 700 m.a.s.l.
-at Skrimfjella is opposite to the general pattern of deeper peats in lowland than upland Norway \citep{lyngstadBeskrivelserAvTorvmassivenheter2023}.
-Moreover, elevation at Ørskogfjellet seems to interact with other variables (non-parallel individual conditional expectation lines), complicating its interpretation.
-Relationships between elevation and peat depth have previously shown opposite shapes in different areas \citep{finlaysonEstimatingOrganicSurface2021}.
-Nonetheless, a relationship that is not generalizable beyond the mapping area is still useful for DSM, as long as it is evaluated to demonstrate its robustness for the predictive task (e.g.~through kNNDM spatial cross-validation).
-
-One interesting feature of the elevational relationships we found may be generalizable: a steep increase in peat depth near the marine limit after the last ice age.
-At Ørskogfjellet, the marine limit is about 75 meters above today's sea level \citep[Geological Survey of Norway,][]{hogaasDatabaseRegistreringAv2012}, where the partial dependence plot of elevation shows a sharp increase in peat depth.
-In areas under the marine limit there has been less time for peat accumulation since the ice sheets retreated, and it is plausible that this makes peats there shallower, all else being equal.
-This hypothesis is supported by similar findings at a coarser scale in the Hudson Bay Lowlands, where a strong positive relationship between peat depth and distance from the coast can be explained by isostatic uplift and time since peat initiation \citep{liPeatDepthCarbon2025}.
-We cannot evaluate this effect at Skrimfjella, where the marine limit is below our study area (at 175 m.a.s.l.).
-
-Another influential terrain-based predictor was Multi-Resolution Valley Bottom Flatness (Fig. \ref{fig:varImp}).
-Unlike elevation, it showed a monotonic effect on peat depth: greater valley bottom flatness was always associated with increases in peat depth (Fig. \ref{fig:pdps}).
-Delineating a valley bottom involves ambiguity, but the Multi-Resolution Valley Bottom Flatness index is a pragmatic approach that considers a location a valley bottom if it is sufficiently low and flat at a particular scale \citep{gallantMultiresolutionIndexValley2003}.
-The multiscale nature of the index allows small elevated but flat areas (including saddles) to be characterized as having high valley bottom flatness \citep{gallantMultiresolutionIndexValley2003}.
-Our results suggest that Multi-Resolution Valley Bottom Flatness is a robust indicator of high water tables (and peat accumulation) over millennial time scales, corroborating other studies \citep{rudiyantoOpenDigitalMapping2018, deragonMappingMaximumPeat2023}.
-
-Other terrain-derived predictors with predictive value in our study are hydrological (Topographic Wetness Index and Depth-to-Water).
-Notably, slope was inferior to (Ørskogfjellet) or highly correlated with (Skrimfjella) these hydrological indices.
-Mappers of peat depth should not assume that slope is the best predictor in its class, despite its prevalence in the literature.
-Wetter locations (high Topographic Wetness Index and low Depth-to-Water) were generally associated with deeper peat, but these relationships were not as strong or consistent as with Multi-Resolution Valley Bottom Flatness (Figs.
-\ref{fig:varImp}, \ref{fig:pdps}).
-The optimal scale for Topographic Wetness Index and Depth-to-Water varied, and likely depends on both the dominant peat formation processes and the typical size of peatland features in a landscape.
-Including multiple scales of these variables allows the model to capture different hydrological mechanisms operating at different spatial scales.
-
-\subsubsection{Legacy depth maps have inconsistent predictive value}
-
-DMK peat depth class proved an inconsistent predictor of peat depth.
-At Skrimfjella, it barely improved model performance (Fig. \ref{fig:modelMetrics}).
-At Ørskogfjellet, it increased performance more, and both indicator variables were among the most important in the model (Fig. \ref{fig:varImp}).
-We suspect the discrepancy between sites is due to different levels of effort and coverage during the historical surveys; more lowland peatland near agriculture at Ørskogfjellet may have caused more purposeful surveying.
-This is evidenced by the fact that 73 \% of the cells measured at Skrimfjella had unknown depth in DMK, compared to 20 \% at Ørskogfjellet.
-Our results show that classification as deep peat at Skrimfjella had no predictive value -- these cells were no different than those classified as shallow or unknown, all else being equal.
-Interactions between DMK and other variables also underline the inconsistency of DMK depth maps, even within a site.
-For example, that a peat at Ørskogfjellet was classified as shallow rather than deep did not always cause the model to predict shallower peat -- sometimes it increased the predicted depth.
-
-\subsection{Implications for digital soil mapping of peat depth}
-
-The performance gap between the best models and the DMK models shows that peat depth in Norway should be mapped digitally.
-Anywhere we have some calibrating measurements, we can get better maps than DMK peat depth classes, at low cost.
-Moreover, DSM methodology can align map products with open science principles by making their production transparent, reproducible, and updatable.
-The large difference we found between the coverage and quality of DMK peat depth at Skrimfjella versus Ørskogfjellet underlines these advantages.
-With DSM we can apply the same approach across different areas and make maps with full spatial coverage, continuous values, and validated uncertainty.
-The rest of this section discusses recommendations and needs for more extensive peat depth DSM.
-
-\subsubsection{Peat depth measurements should be organized}
-
-High quality DTMs are available for mainland Norway, making peat depth measurements the critical training data need.
-For a given area, some minimum number of measurements is necessary to create meaningful improvement over DMK peat depth.
-The proportion of peatlands sampled at Skrimfjella was twice that at Ørskogfjellet, but the Skrimfjella model performed worse, which shows that the size of the depth dataset is not all-important.
-We had the luxury of stratifying our measurements over candidate predictors, and performance may suffer where locations are opportunistic or tied to a sampling design with a different purpose.
-The magnitude of this penalty will depend on the dataset characteristics, but having enough depth measurements is probably more consequential \citep{wadouxSamplingDesignOptimization2019}.
-
-Depth measurements are foundational, and better infrastructure to make these data findable, accessible, interoperable, and reusable would help DSM and other applications.
-Geoportal access and data exchange standards \citeyearpar[like Natural England's for peat surveys,][]{naturalenglandDataExchangeStandard2023} are important.
-Peat data often fall through the cracks between geology-oriented and ecology-oriented archives, but increased awareness of peatland importance is a good impetus to remedy this situation.
-Peat depth is quick and easy to measure, so integrating its measurement into existing national field programs, like Norway's spatially representative nature monitoring or national forest inventory, would be helpful (although not sufficient for regional or local mapping).
-Municipalities and local actors in Norway are increasingly measuring peat depth and can contribute to a growing data foundation \citep{kyrkjeeideCalculatorLocalPeatland2023}.
-Low-altitude, drone-mounted GPR may prove an efficient approach for collecting many, accurate depth data in a landscape, by combining the advantages of airborne deployment and active sensing \citep{pelletierPeatAnalysesHudson1991, ruolsDevelopmentDronebasedGroundpenetrating2023}.
-All of the above can lay the groundwork for renewed peatland maps.
-
-\subsubsection{Spatial scale affects model performance and utility}
-
-Peat depths typically vary over short distances (e.g.~\textless{} 1 m), so mapping at 10 m resolution implies that the map will compress much of the fine-scale variation in depth.
-For example, large differences in maximum depth can be obscured at 10 m resolution.
-For the same reason, seemingly small improvements in mean absolute error at 10 m resolution may represent meaningful improvements for map users.
-Spatial aggregation also means that an uneven distribution of point measurements within a cell can cause its mean depth to be unrepresentative.
-Some cells had dense and evenly distributed measurements, while others had only a few in one corner.
-Although we tried to reduce the influence of uneven point coverage -- by inversely weighing point values by their distances to the cell center -- it probably contributed to model inaccuracy.
-Fine-scale variation in peat depth may favor mapping at very fine resolution (1 m) --- even if land use planning and carbon accounting do not operate at this grain.
-Terrain--depth relationships might be stronger in 1 m cells than in our 10 m cells, especially considering the hummock--hollow microtopography of many peatlands \citep{rydin7Mires1999, lindsayPeatbogsCarbonCritical2010}.
-
-The scarcity of depth measurements and their short spatial autocorrelation mean that mapping peat depth at broad scales is not an exercise in spatial interpolation \citep{henglGenericFrameworkSpatial2004}.
-Peat depths in our GPR data showed spatial autocorrelation to a range of 50--100 m, and getting measurements at such fine grain is only realistic for small areas, not across whole landscapes.
-Therefore, we anticipate that spatially explicit DSM is of limited value in all but the most intensively sampled landscapes (currently nowhere in Norway).
-
-Choosing a spatial extent for DSM can be tricky.
-A natural starting point is the bounding area around a spatial cluster of depth measurements.
-For a given cluster, kNNDM can define how far the boundary can extend beyond the point measurements; if the sample-to-prediction nearest neighbor distribution cannot be simulated by any set of cross-validation folds, then the extent is too expansive \citep{meyerMachineLearningbasedGlobal2022, linnenbrinkKNNDMCVKfold2024}.
-However, it is unclear how big (N) any cluster of depths should be, and the tradeoff between multiple small-extent DSM and fewer large-extent DSM needs research.
-Bohn and Miller \citeyearpar{bohnLocallyEnhancedDigital2024} advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or `supertopes'), where the composition of mesotopes is similar -- e.g.~regions dominated by raised bogs versus regions dominated by sloping fens \citep{moenNationalAtlasNorway1999, joostenWiseUseMires2002}.
-Depth varies systematically between bogs and fens \citep{lindsayPeatlandMireTypes2016}, between peats formed by terrestrialization versus paludification \citep{buffamFillingHolesRegional2010}, and probably along other axes of peatland typology.
-Therefore, DSM is more likely to uncover consistent predictor--depth relationships within peatland regions than across them.
-
-\subsubsection{Machine learning approaches can build on success}
-
-The DSM literature and our results support using flexible machine learning algorithms like RF to predict peat depth.
-RF avoids negative predictions \citep[c.f.][]{kogantiMappingPeatDepth2023} and produces good uncertainty estimates \citep[our study,][]{vaysseUsingQuantileRegression2017, takoutsingComparingPredictionPerformance2022}.
-As depth data become more abundant, we may move from pixel-based learners to deep learning approaches like convolutional neural networks.
-Bakkestuen et al.
-\citeyearpar{bakkestuenDelineationWetlandAreas2023} successfully predicted peatland extent using convolutional neural networks, and their ability to automatically learn multiscale spatial features would reduce the need to manually engineer these.
-The success of Multi-Resolution Valley Bottom Flatness as a predictor of peat depth demonstrates that multiscale spatial patterns matter for peat depth and convolutional neural networks are designed to learn such patterns.
-The kind of relationship described in Buffam et al.
-\citeyearpar{buffamFillingHolesRegional2010}, where peat depth in basins related to terrain slope at the basin edge, is also something a convolutional neural network could learn.
-However, since this approach is data-hungry, we should build soil knowledge into the DSM where we can \citep{minasnySoilScienceInformedMachine2024}.
-For example, if further research confirms the effect of the marine limit that we found, then it is better to include the marine limit as a predictor than to make the algorithm learn this pattern from elevation independently.
-
-\subsubsection{Peat extent and depth should be mapped together}
-
-Finally, we want to highlight the need for research on peatland extent and peat depth to be better integrated.
-Since peatland extent is defined by non-zero peat depth \citep[the specific threshold varies by definition,][]{minasnyMappingMonitoringPeatland2024}, they are fundamentally linked.
-The goal, therefore, should be a unified prediction framework for extent and depth.
-We caution against reducing continuous depth predictions to arbitrary classes \citep[as in][]{ivanovsModelingGeospatialDistribution2024, karjalainenComparisonTwoGammaray2025}, since classes can be derived from continuous predictions.
-The distribution of peat depths across full landscapes is zero-inflated, and research is needed to determine whether it is more efficient to parameterize a single model of peat depth (with a larger, generalized dataset) or to break down the problem into a hurdle model by classifying zero depth and then regressing non-zero depths (with smaller, specialized datasets).
-Coupling extent and depth will reduce the prevalence of incoherence that we found: deep peat outside the peatland extent and zero depth inside it.
-A key challenge going forward will be obtaining training data that represents both zero and non-zero components of the depth distribution, since sampling designs often focus on known peatlands.
-
-\codedataavailability{R code used in this study is available at \url{https://github.com/julienvollering/DSMdepth}.
-The depth measurements we used are archived at \doi{10.6073/pasta/6ce440152f693f2156bf5b692a2e7917} and follow data and metadata standards.} %% use this section when having data sets and software code available
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% optional
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-\appendixtables
-\begin{table}[tbp]
-\caption{Attributes of the radiometric surveys, as reported in Baranwal et al., \citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013} and Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.}
-\begin{tabular}{lll}
-\hline
-                                         & Skrimfjella & Ørskogfjellet              \\ \cline{2-3} 
-Survey period                            & 2008-2011   & December 2014–January 2015 \\
-Average flight altitude (m)              & 75          & 80                         \\
-Average flight speed (\unit{km\,h^{-1}}) & 108         & 88                         \\
-Flight line spacing (m)                  & 200         & 200                        \\ \hline
-\end{tabular}
-\label{tab:radSurveys}
-\end{table}
-\clearpage
-\appendixfigures
-\begin{figure}
-\centering
-\includegraphics[width=0.8\textwidth]{figures/GPRwavevelocity.pdf}
-\caption{\label{fig:GPRwavevelocity}Calibration of GPR wave velocity at Skrimfjella (a) and Ørskogfjellet (b) by regression of probed depth against wave travel time.}
-\end{figure}
-\clearpage
-\appendixfigures
-\begin{figure}
-\centering
-\includegraphics{figures/semivariograms.pdf}
-\caption{\label{fig:semivariograms}Empirical semivariograms of peat depth point measurements at Skrimfjella (a) and Ørskogfjellet (b).}
-\end{figure}
-\noappendix
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\authorcontribution{\emph{Author contributions.}
-JV: Conceptualization, Investigation, Data curation, Formal analysis, Writing -- original draft.
-NG: Conceptualization, Methodology, Writing - review \& editing.
-MKG: Investigation, Writing - review \& editing.
-KKM: Investigation, Data curation, Writing - review \& editing.
-SDN: Investigation, Writing - review \& editing.
-KR: Conceptualization, Investigation, Writing - review \& editing.
-MS: Investigation, Data curation, Writing - review \& editing.} %% optional section
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\competinginterests{\emph{Competing interests.} The authors declare that they have no conflict of interest.} %% this section is mandatory even if you declare that no competing interests are present
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\disclaimer{\emph{Disclaimer.} The authors declare that the results, discussions, and interpretations presented in this study are solely their own.
-The views expressed herein do not necessarily reflect those of their respective institutions or funding agencies.} %% optional section
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{acknowledgements}
-\emph{Acknowledgements.} We thank the Norwegian Public Roads Administration for sharing data from ground-penetrating radar surveys.
-We also thank Vikas Baranwal from the Geological Survey of Norway for helping us access the radiometric data from Skrim.
-This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, © Kartverket, (2) \emph{Norge digitalt} license, Norwegian Institute of Bioeconomy Research (NIBIO), © Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway.
-Large language models have been used during the drafting and editing of this manuscript, with author oversight.
-We maintain full responsibility for the scientific output, as per the European Commission's \emph{Living guidelines on the responsible use of generative AI in research}.
-\end{acknowledgements}
-
-%% REFERENCES
-%% DN: pre-configured to BibTeX for rticles
-
-%% The reference list is compiled as follows:
-%%
-%% \begin{thebibliography}{}
-%%
-%% \bibitem[AUTHOR(YEAR)]{LABEL1}
-%% REFERENCE 1
-%%
-%% \bibitem[AUTHOR(YEAR)]{LABEL2}
-%% REFERENCE 2
-%%
-%% \end{thebibliography}
-
-%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
-%% authors experienced with BibTeX only have to include the following two lines:
-%%
-\bibliographystyle{copernicus}
-\bibliography{ms.bib}
-%%
-%% URLs and DOIs can be entered in your BibTeX file as:
-%%
-%% URL = {http://www.xyz.org/~jones/idx_g.htm}
-%% DOI = {10.5194/xyz}
-
-%% LITERATURE CITATIONS
-%%
-%% command                        & example result
-%% \citet{jones90}|               & Jones et al. (1990)
-%% \citep{jones90}|               & (Jones et al., 1990)
-%% \citep{jones90,jones93}|       & (Jones et al., 1990, 1993)
-%% \citep[p.~32]{jones90}|        & (Jones et al., 1990, p.~32)
-%% \citep[e.g.,][]{jones90}|      & (e.g., Jones et al., 1990)
-%% \citep[e.g.,][p.~32]{jones90}| & (e.g., Jones et al., 1990, p.~32)
-%% \citeauthor{jones90}|          & Jones et al.
-%% \citeyear{jones90}|            & 1990
-
-\end{document}

### 2025-02-28 ms/figures.R

@@ -189,6 +189,8 @@ ggsave(g123, filename = 'sites-patchwork.pdf', path = "ms/figures",
        width =210-30, height = 240-40, units = 'mm') #copernicus.cls page 210x240
 # ggsave(filename = 'sites-patchwork.svg', path = "ms/figures",
 #        width =210-30, height = 240-40, units = 'mm')
+# ggsave(filename = 'sites-patchwork.png', path = "ms/figures",
+#        width =210-30, height = 240-40, units = 'mm', dpi = 300)
 
 # Model metrics by faceting ####
 
@@ -209,15 +211,15 @@ plotting <- bind_rows(orskog = orskog, skrim = skrim, .id = 'site') %>%
                          "unitless\nR-squared",
                          "cm\nMean absolute error"),
     model = case_when(
-      model == "DMK" ~ "DMK class (2)",
+      model == "DMK" ~ "DMK (2)",
       model == "Terrain" ~ "terrain (21)",
-      model == "TerrainDMK" ~ "terrain + DMK class (23)",
+      model == "TerrainDMK" ~ "terrain + DMK (23)",
       model == "RadiometricTerrain" ~ "terrain + radiometric (25)",
       model == "RadiometricTerrainDMK" ~ "all predictors (27)"),
     model = fct_relevel(model,
-                        "DMK class (2)",
+                        "DMK (2)",
                         "terrain (21)",
-                        "terrain + DMK class (23)",
+                        "terrain + DMK (23)",
                         "terrain + radiometric (25)",
                         "all predictors (27)"))
 str(plotting)
@@ -248,7 +250,7 @@ g1 <- ggplot(plotting) +
         legend.key.spacing.y = unit(-2, "mm"),
         strip.background = element_blank(), 
         strip.placement = "outside")
-ggsave(filename = 'modelmetrics-faceting.pdf', path = "ms/figures",
+ggsave(g1, filename = 'modelmetrics-faceting.pdf', path = "ms/figures",
        width =210-30, height = (240-40)/2.5, units = 'mm') #copernicus.cls page 210x240
 
 # Model metrics ####
@@ -263,15 +265,15 @@ plotting <- bind_rows(orskog = orskog, skrim = skrim, .id = 'site') %>%
   mutate(
     site = fct_relevel(site, "skrim"),
[…]

### 2025-02-28 ms/tables.R

@@ -4,35 +4,9 @@ library(tidyverse)
 
 rast <- terra::rast("output/predictors.tif")
 names(rast)
-tribble(
-  ~name, ~description,
-  "radK", "Potassium ground concentration",        
-  "radTh", "Thorium ground concentration",
-  "radU" , "Uranium ground concentration",        
-  "radTC", "Total count of gamma radiation",        
-  "elevation", "Mean elevation",
-  "slope1m", "Mean of 1 m slope",      
-  "TPI1m", "Mean of 1 m topographic position index",
-  "TRI1m", "Mean of 1 m terrain ruggedness index",        
-  "roughness1m", "Mean of 1 m roughness",  
-  "slope10m", "10 m slope",    
-  "TPI10m", "10 m topographic position index",
-  "TRI10m", "10 m terrain ruggedness index",
-  "roughness10m" , "10 m roughness",
-  "MRVBF", "Multi-resolution valley bottom flatness",
-  "TWI5m", "Mean of 5 m topographic wetness index",
-  "TWI10m", "10 m topographic wetness index",
-  "TWI20m", "Bilinear interpolation of 20 m topographic wetness index",
-  "TWI50m", "Bilinear interpolation of 50 m topographic wetness index",
-  "DTW2500", "Depth-to-water index, flow initiation area of 0.25 ha",
-  "DTW5000", "Depth-to-water index, flow initiation area of 0.5 ha",
-  "DTW10000", "Depth-to-water index, flow initiation area of 1 ha",
-  "DTW20000", "Depth-to-water index, flow initiation area of 2 ha",
-  "DTW40000", "Depth-to-water index, flow initiation area of 4 ha",
-  "DTW80000", "Depth-to-water index, flow initiation area of 8 ha",
-  "DTW160000", "Depth-to-water index, flow initiation area of 16 ha",
-  "DMK", "DMK peat depth class, categorical with 3 levels") |> 
-  write_csv("ms/tables/predictors.csv")
+
+preds.table <- read_csv("ms/tables/predictors.csv")
+View(preds.table)
 
 # Summary cell depths by attributes ####
 

### 2025-03-05 ms/ms.Rmd
prep submission: figure licenses

@@
 Terrain is visualized using a hillshade with slope 45° and azimuth 225°.
-AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (<https://data.norge.no/nlod/en/1.0>) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (<https://creativecommons.org/licenses/by/4.0/>)."}
+AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (https://data.norge.no/nlod/en/1.0) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/)."}
 knitr::include_graphics("figures/sites-patchwork.pdf")

### 2025-03-05 ms/ms.tex
prep submission: figure licenses

@@
 Terrain is visualized using a hillshade with slope 45° and azimuth 225°.
-AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (<https://data.norge.no/nlod/en/1.0>) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (<https://creativecommons.org/licenses/by/4.0/>).}\label{fig:sites}
+AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (https://data.norge.no/nlod/en/1.0) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/).}\label{fig:sites}
 \end{figure}

### 2025-04-24 ms/submission/AuthorComments-egusphere-2025-1046.qmd
prepare response template

---
title: "Response to Referees"
author: "Julien Vollering, on behalf of all co-authors"
format:
  html:
    embed-resources: true
toc: true
---

```{=html}
<style>
.added {
  color: blue;
  font-weight: bold;
}
.deleted {
  color: red;
  text-decoration: line-through;
}
</style>
```

<https://www.soil-journal.net/peer_review/interactive_review_process.html>:

> Author comments (ACs) can only be posted by the contact authors of the manuscript in discussion on behalf of all co-authors. Co-authors can post CCs but not ACs. After the open discussion, during the final response, the author comments should be structured in a clear and easy-to-follow sequence: (1) comments from referees/community, (2) author's response, and (3) author's changes in manuscript.

# Response to Referee 1

<https://doi.org/10.5194/egusphere-2025-1046-RC1>

## General Comments

### Comment 1.1

> This article comprehensively addresses the modelling challenges of predicting peat depth from terrain variables. It takes high resolution terrain variables (derived from LiDAR) and low resolution (comparatively) airborne radiometric data across two individual peat landscapes in Norway and used Random Forest machine learning algorithm to train combinations of these variables to a multitude of peat depth probes and GPR peat depth measurements taken in order to establish the predict power of such variables for peat depth mapping.

### Comment 1.2

> Overall, I found the article to be generally well written, with a very comprehensive and detailed description of modelling mechanism, error derivation, and feature choice. It is a very long article, going into great detail in several areas, and below I suggest at least one section/topic that could be removed entirely to reduce length and increase overall readability. I suggest the authors review all sections for conciseness and reduce the article length where possible. The level of detail may mean a reader less familiar with machine learning modelling may find the article hard to follow. While I do find the article to be within the scope of SOIL, I would be concerned that it focuses heavily on the modelling methodology.

### Comment 1.3

> Additionally, there were several areas where the language used was quite casual for a scientific article. Several are highlighted under minor considerations below.

### Comment 1.4

> Finally, the main concern noted is the imbalance between the consideration given to radiometric data compared to terrain variables. Considering the title is stating terrain being “better” than radiometrics, and given the emerging understanding of the use of radiometric data in peat land mapping, there is are some fundamental errors in the methods presented, which may be biasing the conclusion alluded to in the title.

## Specific Comments

### Comment 1.5

> The first concern in the comparison of radiometric data to lidar terrain variables is related to the choice of radiometric variables. The authors opt to use potassium, uranium and thorium ground concentration units alongside the Total Count data from the full energy spectrum. These ground concentration measurements are derived from counts per second measurements on the airplane which are calibrated, usually using pads of known concentrations at a calibration facility. Therefore, concentration of any radioelement is a measurement of the concentration of that element in the top ~ 60 cm – 1 m of the soil. However, peat soils are different. Being organic, they don’t contain the typical geological material that make up soils. Therefore, they act as an attenuative environment to gamma rays. As the potential source of gamma rays in peat areas is blocked and attenuated by the peats, the concentration calibration is no longer physically valid. While these concentration data are indeed provided by the contractors of such surveys, it is now recognized that the counts per second measurement is a more appropriate unit when considering attenuation of gamma rays in peat soils (O'Leary et al, 2022, 2024). In particular considering depth, the deeper the peat, the greater the attenuation of gamma rays. Similarly, the wetter the peat the greater the attenuation of gamma rays. The use of concentration data is not valid for a study in predicting peat depth. I recommend either the authors convert these concentrations to counts per second, or remove all but the Total Count data from their analysis and consider take the next concern into account.

### Comment 1.6

> There is an additional argument missing from within the authors discussion, namely the fact that we never know the initial source strength, or counts, of the gamma rays for a given footprint. The measurement at the airplane is an attenuated version of this initial source. This attenuation is controlled by the attenuation coefficient for a given element and depth, soil moisture, bulk density and porosity (Beamish 2013) of the peat soils. From a purely physics/modelling point of view, this makes the prediction of peat depth an underdetermined problem. Even if the soil moisture, bulk density and porosity was known absolutely, the initial source is never known and so any number of peat depths may result in any given gamma count at the airplane. Additionally, this modelling exercise cannot be performed on Total Count data as this is summed from the entire measurement energy spectrum, which contains multiple element specific attenuation coefficient, meaning the Total Count data is only ever indicative of attenuation variability across a site, with no ability to model anything quantitative. This puts Radiometrics in a natural disadvantage for a quantitative prediction of peat depth. Given the title of this article, I find that the discussion around the radiometrics lacked sufficient detail to make a fair comparison, which naturally results in such a bias towards terrain variables in such a modelling context. This is therefore not a result per say, but more a perfectly expected outcome. For this concern, I suggest a more comprehensive discussion around the physical limitations of radiometrics in the prediction of peat depth.

### Comment 1.7

> The main focus of this article is on the prediction of peat depth. However, the authors include several sections of the possibility of peatland extent mapping. This is not mentioned at all in the abstract, or the introduction in great detail. As this article is already quite long and complex, I suggest the removal of any sections and text related to mapping peat land extent as it is not the focus of the article and only acts to add unnecessary complexity to an already very technical methodology. The authors even mention in Line 535 that their aim was not to map extent. I suggest the removal of all reference to peatland extent prediction and instead focus on the prediction of peat depth. A much shorter reference to the importance of peatland extent knowledge could perhaps be mentioned in the conclusions, but a full analysis and discussion (section 4.1.4) is not appropriate in this article.

## Technical Corrections

> Line 58- 59. There is no need to include this sentence with an example here, as the next paragraph goes into the necessary detail on Slope. This is an example of how the authors might reduce the overall size of the article.

> Section 2.2 I would suggest moving this opening paragraph to the end of this section as it acts more to sum up how the authors are using the various predictors. It mentions several of the predictors directly, but the are not described until later sections (for example 2.2.1). This would increase the readability of this section.

> Line 179: Remove “also using White Box” as this is obvious.

> Line 258: The authors mention density; however, they do not expand on this. Was this measured in the field, or an operator’s observation and subjective interpretation of density?

>  371: What is the relevance of the Persons correlation coefficient of 0.7. Was this tested at all? Readers may not be familiar with this so it should be explained a bit more.

> Table 2:  I recommend putting a vertical line between the results for both sites so as to easier distinguish between them.

> Heading Section 4.1.1 – “but not useless!” is very casual language to be using in a scientific article. This is just one example of this casual language. I suggest the authors review the article for this throughout.

> Line 464: “low hanging fruit” is also casual and colloquial and may not be understood by all cultures.

> Line 515: “large stocks really are large” – very vague and non-scientific comment. What is large?

> Line 530: remove the question at the start of this section.

> Line 549: This section is the first mention of the radiometric survey parameters. I would recommend moving some of this section to the Methods and Material section as it is useful descriptors of how the data came to be.

> Line 564: Typically, airborne radiometric surveys have strict conditions that they must fly under. One is related to rainfall occurrence and airborne surveys should not happen directly after rainfall. The authors statement with regards to air moisture should be clarified as otherwise the contractor may have been at fault by provide incorrect data, which would have implications for the usage of radiometric data in this area in this study.

> Line 628 – 629: remove the sentence starting “The rest of this section….” As it is unnecessary.

> Line 634: “luxury” is again a very casual phrasing within a scientific article.

> Line 664: “tricky” – casual

> Finally, there is no definite conclusion to this article, nor a heading stating same. I suggest the authors either add a section at the end and move some text here to highlight the main conclusions as currently the “discussion” section is quite large and probably not appropriate to act as a combined discussion and conclusion section.

# Response to Referee 2

<https://doi.org/10.5194/egusphere-2025-1046-RC2>

## General Comments

### Comment 2.1

> This article focuses on improving peat depth mapping in two distinct peatland landscapes in Norway using a digital soil mapping approach. The study employed the Random Forest (RF) algorithm to model both peat presence and depth, using high-resolution terrain data, remotely sensed radiometric data, and polygonised peat depth data from an existing map. The RF models were calibrated using field-measured data points collected across both regions, and variable importance was analysed.

### Comment 2.2

> Overall, this research addresses a significant challenge in peat depth mapping. The title is clear and likely to attract attention from readers interested in this topic. However, I have some concerns regarding the writing style. Certain sections use jargon excessively or are phrased in a way that feels less scientific, which may hinder clarity and accessibility for a broader audience.

## Specifics Comments

### Comment 2.3

> While the title is clear, I personally find it somewhat overconfident in summarising the results. This is mainly due to the relatively low performance of the RF models and the minimal difference observed between the terrain-only and terrain-plus-radiometric models. To better support the claims, it would be helpful to include a statistical test (e.g., a t-test) to assess whether the performance differences between models are statistically significant. Furthermore, incorporating a stand-alone model that uses only radiometric data as predictors would allow for a more balanced evaluation of the variable groups and help clarify the specific contribution of radiometric inputs in predicting peat depth in Norway.

### Comment 2.4

> Figure 1: While the terrain-look map provides a general view of the study area, the relief of the regions remains unclear. Consider adding a clearer topographic representation (e.g., contour profile) to better illustrate landscape variation. Additionally, it would be helpful to explain the rationale for selecting Skrimfjella, which appears to have relatively limited peat coverage, especially given that the adjacent region seems to contain a larger peatland area.

### Comment 2.5

> It is unclear how the peat depth data are distributed, both statistically and spatially. While the sampling design is described in the text, I recommend including a figure showing the spatial distribution of the data points to help readers better understand the coverage and representativeness of the dataset. Additionally, a basic statistical summary (e.g., mean, median, range, standard deviation) of the peat depth values would be beneficial to provide context and a clearer picture of the conditions being modelled.

### Comment 2.6

> Since the radiometric data are originally at 50 m resolution and the terrain data are at 1 m resolution, the method used to resample these datasets to a common 10 m resolution could influence model performance. In particular, the use of cubic spline resampling for the radiometric data may affect how well its spatial variability is represented, especially when compared to the aggregated 10 m topographic data. I would be interested to hear your thoughts on how this resampling approach might have impacted the results.

### Comment 2.7

> I found the section on sampling design and peat depth measurement to be overly detailed for the main text. While this information is valuable, it may be more appropriate to move some of it to the appendix or supplementary material, as it is less central to the analysis and interpretation of results. This would help improve the flow and focus of the main manuscript.

### Comment 2.8

> In the model interpretation section, it appears that three different methods were used to assess variable importance: FIRM, permutation importance, and Shapley values. However, the implications of using these different approaches are not clearly discussed. Each method captures different aspects of variable influence and may lead to different interpretations. Could you clarify how the results from these methods align or differ, and what their respective implications are for understanding the drivers of peat depth in your study?

### Comment 2.9

> I recommend adding a dedicated section to summarise the key results and findings. This would help readers clearly see how the study addresses its original research aims and questions. A concise summary at the end of the results or discussion section would also improve the overall structure and coherence of the paper.

## Technical Corrections

> Table 2: I think you need to move this table somewhere after it is mentioned in the main text. At first, I was quite confused with this table. What does that percentage stand for?

> Line 394-397: I don’t quite get the point of this section. To my understanding, you compared between the quartile predictions and the observed data to see the prediction interval coverage probability. If so, I think it would be better to plot this prediction interval together with the observed data in scatter plot, like in Figure 3.

> Line 413-414: which curve? The information inside the parentheses is not clear to me.

> Figure 4: is the null derived from the calculation between observations and assumptions of 30 cm peat depth? Where is the standard error come from?

> Line 464: “low hanging fruit” ?

> Line 515: This sentence is unclear to me.

> Line 531: “0% in prediction” what does this sentence mean?

> Line 614: What are the ‘both indicators variables’ referring to?

# Response to Referee 3

### Comment 3.1

> Referee's comment goes here.

We thank the Referee for this insightful comment. We have revised the manuscript accordingly:

::: {.callout appearance="simple"}
**Original text:**

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.

**Revised text:**

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Sed do eiusmod tempor incididunt ut labore et **dolore magna aliqua** [and additional explanation of the concept]{.added}.
:::

### Comment 3.2

> Referee's comment goes here.

Thank you for highlighting this issue. We have made the following changes:

::: {.callout appearance="simple"}
**Original text:**

Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.

**Revised text:**

[We have completely revised this section]{.added} [to address the Referee's concerns.]{.added} [Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.]{.deleted}
:::

# Response to editorial support team

### Comment E.1

> Please add the section headlines "Appendix A" (in front of Table A1) and "Appendix B" (in front of Figure B1).

### 2025-05-12 ms/ms.Rmd
address RC1.7: remove extent models
also added out.width to figures to avoid missing \pandocbounded macro from the custom template

@@
   Based on these findings, we conclude that digital soil mapping can improve existing, broad-scale maps of peat depth in Norway, but highly localized carbon stock assessments are best made from field measurements.
-  Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat lateral extent and depth.
   Together, these pathways promise more accurate landscape-scale carbon stock assessments and better-informed land management policies.
@@
 
-We also measured peatland occurrence (binary variable) in a separate sample at Skrimfjella.
-We measured peatland occurrence because the AR5 dataset is known to underestimate peatland extent [especially in forests, @brynLandCoverNorway2018], and because airborne radiometrics may help identify unmapped peatland [@gatisMappingUplandPeat2019; @olearyDigitalSoilMapping2022].
-The 10 m cells were sampled from those outside AR5 mire and with slope < 20°. 
-Following the same procedure, *eSample* returned 106 raster cells from this population.
-
 ### Ørskogfjellet
@@
 We georeferenced the map and digitized the borehole locations and depths.
-
-For our sample of peatland occurrence at Skrimfjella, we recorded the presence or absence of peatland at the cell center -- primarily by digging and examining the top 20 cm of soil.
-We judged whether the soil was a peat soil based on its density, texture, and color, as well as the presence or absence of mire vegetation.
-Although peat soil is strictly defined by organic content (which we did not analyze), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland [@minasnyMappingMonitoringPeatland2024].
 
@@
 
-#### Peatland extent
-
-While the primary aim of the DSM was to predict peat depth within AR5 mires (where peat depth is supposed to be > 30 cm), we also tested whether our mire-trained models of peat depth could identify peatlands outside of AR5 mires.
-For both sites we performed an evaluation with our (cell-aggregated) peat depth measurements, of which about one-fifth were outside of AR5 mires.
-For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on AR5 mire locations (in the training folds), and evaluated them only on (test fold) locations classified in AR5 as something other than mire.
-From the cross-validation we quantified mean absolute error.
-For Skrimfjella -- where we measured peatland occurrence separately from depth -- we also trained a model on all peat depth measurements (AR5 mire or not) and then used the independent occurrence dataset to evaluate this model.
-Specifically, we calculated the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
-This analysis treats the model's prediction of peat depth as a rank index of peatland likelihood.
-Since the purpose was to test the models' ability to uncover unmapped peatland (where DMK peat depth class is undefined), we evaluated models with two predictor configurations: *terrain* or *terrain + radiometric*.
-
 ### Model interpretation
@@
 
-```{r modelMetrics, echo = FALSE, fig.cap = "Performance of peatland depth models with different predictor configurations, evaluated via spatial cross-validation.
+```{r modelMetrics, echo = FALSE, out.width="100%", fig.cap = "Performance of peatland depth models with different predictor configurations, evaluated via spatial cross-validation.
 Parentheses denote the number of variables in each predictor configuration, and point estimates are shown +/- their standard error."}
@@
 
-```{r calPlots, echo = FALSE, fig.cap = "Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions from spatial cross-validation.
+```{r calPlots, echo = FALSE, out.width="100%", fig.cap = "Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions from spatial cross-validation.
 Points are transparent to show overlap.
@@
 knitr::include_graphics("figures/calibration_plots.pdf")
-```
-
-### Peatland extent
-
-When we tested how well models extrapolated from within to outside of AR5 mire areas, we found that the models produced worse mean absolute error than just assuming a constant 30 cm depth (Fig.
-\@ref(fig:modelMetricsExtrapolation)).
-With the independent occurrence data at Skrimfjella, we found that neither the *terrain* nor *terrain + radiometric* configurations were able to discriminate between peat presence and absence (area under the curve of the receiver operating characteristic 0.44 and 0.52 respectively, where 0.5 indicates random guessing).
-
-```{r modelMetricsExtrapolation, echo = FALSE, fig.cap = "Performance of models that extrapolate from training data inside of mapped mires to test data outside of mapped mires (AR5 dataset), evaluated via spatial cross-validation.
-Parentheses denote the number of variables in each model, and point estimates are shown +/- their standard error."}
-knitr::include_graphics("figures/modelmetrics-extrapolation.pdf")
 ```
@@
 
-```{r varImp, echo = FALSE, fig.cap = "Global variable importance in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), as measured by three different metrics: Shapley values, permutation importance, and the Feature Importance Ranking Measure [@greenwellVariableImportancePlots2020].
+```{r varImp, echo = FALSE, out.width="100%", fig.cap = "Global variable importance in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), as measured by three different metrics: Shapley values, permutation importance, and the Feature Importance Ranking Measure [@greenwellVariableImportancePlots2020].
 Variables removed due to collinearity are shown to the right of the variable with which they are most correlated."}
@@
 
-```{r pdps, echo = FALSE, fig.cap = "Partial dependence plots of the six most important variables in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b).
+```{r pdps, echo = FALSE, out.width="100%", fig.cap = "Partial dependence plots of the six most important variables in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b).
 The average effect of the predictor on the outcome (red line) overlays the variation in the effect across observations (grey lines).
@@
 Although this recommendation aligns with the precautionary principle, here we make it on technical grounds based on the maps’ characteristics.
-
-### Depth predictions do not necessarily extrapolate to peat extent
-
-That we measured more than 30 cm of peat in areas not mapped as mire was expected, because AR5 underrepresents peatland extent by about one third [@brynLandCoverNorway2018]. 
-For that reason, maps of peatland extent also need revising. 
-Unfortunately, our models did not predict peat occurrence outside AR5 mires. 
-In these areas, assuming a constant 30 cm of peat (the depth threshold used in AR5) produced less error than our models (Fig. \@ref(fig:modelMetricsExtrapolation)). 
-Moreover, the performance gap would likely have been larger had our evaluation data been representative of all areas not classified as mire in AR5. 
-Instead they were concentrated near AR5 mires. 
-Most importantly, the independent occurrence data at Skrimfjella showed the model completely failing to predict occurrence, and these data were suited for testing this ability. 
-Although the Ørskogfjellet model may have done better than the Skrimfjella model on a similar independent evaluation set (since it predicted depth better), the improvement would probably be small.
-
-Why wasn't peat occurrence predicted well? 
-The models were trained on fundamentally different populations of locations (~80 % AR5 mire in training vs.
-0 % in prediction), so it is not surprising that the associations they learned did not transfer well.
-Moreover, the ~20 % of training data from outside AR5 mires were incidentally collected (near AR5 mires) and not representative of other land cover classes. 
-In short, the models were blind to the fact that most of both study areas have no peat.
-
-Peatland extent is probably best mapped using different remotely sensed data than we used here [@bakkestuenDelineationWetlandAreas2023], and this study’s purpose was not to map extent. 
-Nevertheless — as long as the relevant peatland definition includes a depth component — depth predictions (or predictors thereof) should help delineate peatland extent [@olearyDigitalSoilMapping2022; @beamishDetailedMappingPeat2024].
-We return to this point in our discussion of implications for digital soil mapping.
 
@@
 
-Finally, we want to highlight the need for research on peatland extent and peat depth to be better integrated. 
-Since peatland extent is defined by non-zero peat depth [the specific threshold varies by definition, @minasnyMappingMonitoringPeatland2024], they are fundamentally linked. 
+Finally, we briefly want to highlight the need for peatland extent mapping and peat depth mapping to be better integrated. 
+Since peatland extent is defined by non-zero peat depth [the specific threshold varies by definition, @minasnyMappingMonitoringPeatland2024], the lateral and vertical dimensions are fundamentally linked.
 The goal, therefore, should be a unified prediction framework for extent and depth.
-We caution against reducing continuous depth predictions to arbitrary classes [as in @ivanovsModelingGeospatialDistribution2024; @karjalainenComparisonTwoGammaray2025], since classes can be derived from continuous predictions.
-The distribution of peat depths across full landscapes is zero-inflated, and research is needed to determine whether it is more efficient to parameterize a single model of peat depth (with a larger, generalized dataset) or to break down the problem into a hurdle model by classifying zero depth and then regressing non-zero depths (with smaller, specialized datasets).
-Coupling extent and depth will reduce the prevalence of incoherence that we found: deep peat outside the peatland extent and zero depth inside it. 
-A key challenge going forward will be obtaining training data that represents both zero and non-zero components of the depth distribution, since sampling designs often focus on known peatlands.
+Research is needed to determine whether it is better to parameterize a single model of peat depth across a full landscape, or to break down the problem into a hurdle model by classifying zero depth and then regressing non-zero depths.
+Though peatland definitions may encourage reducing continuous depth predictions to arbitrary classes, we caution against this practice [as in @ivanovsModelingGeospatialDistribution2024; @karjalainenComparisonTwoGammaray2025].

### 2025-05-12 ms/ms.tex
address RC1.7: remove extent models
also added out.width to figures to avoid missing \pandocbounded macro from the custom template

@@
 Based on these findings, we conclude that digital soil mapping can improve existing, broad-scale maps of peat depth in Norway, but highly localized carbon stock assessments are best made from field measurements.
-Furthermore, the inability of models to identify peat presence outside known peatlands highlights the need for integrated mapping of peat lateral extent and depth.
 Together, these pathways promise more accurate landscape-scale carbon stock assessments and better-informed land management policies.
@@
 
-We also measured peatland occurrence (binary variable) in a separate sample at Skrimfjella.
-We measured peatland occurrence because the AR5 dataset is known to underestimate peatland extent \citep[especially in forests,][]{brynLandCoverNorway2018}, and because airborne radiometrics may help identify unmapped peatland \citep{gatisMappingUplandPeat2019, olearyDigitalSoilMapping2022}.
-The 10 m cells were sampled from those outside AR5 mire and with slope \textless{} 20°.
-Following the same procedure, \emph{eSample} returned 106 raster cells from this population.
-
 \subsubsection{Ørskogfjellet}
@@
 We georeferenced the map and digitized the borehole locations and depths.
-
-For our sample of peatland occurrence at Skrimfjella, we recorded the presence or absence of peatland at the cell center -- primarily by digging and examining the top 20 cm of soil.
-We judged whether the soil was a peat soil based on its density, texture, and color, as well as the presence or absence of mire vegetation.
-Although peat soil is strictly defined by organic content (which we did not analyze), we believe our protocol produced reasonable determinations of presence or absence that would generally satisfy most of the varying definitions of peatland \citep{minasnyMappingMonitoringPeatland2024}.
 
@@
 
-\paragraph{Peatland extent}
-
-While the primary aim of the DSM was to predict peat depth within AR5 mires (where peat depth is supposed to be \textgreater{} 30 cm), we also tested whether our mire-trained models of peat depth could identify peatlands outside of AR5 mires.
-For both sites we performed an evaluation with our (cell-aggregated) peat depth measurements, of which about one-fifth were outside of AR5 mires.
-For this evaluation we used the same spatial cross-validation folds as previously, but now we trained the models only on AR5 mire locations (in the training folds), and evaluated them only on (test fold) locations classified in AR5 as something other than mire.
-From the cross-validation we quantified mean absolute error.
-For Skrimfjella -- where we measured peatland occurrence separately from depth -- we also trained a model on all peat depth measurements (AR5 mire or not) and then used the independent occurrence dataset to evaluate this model.
-Specifically, we calculated the area under the curve of the receiver operating characteristic, to evaluate the model's ability to discriminate between peat presence and absence.
-This analysis treats the model's prediction of peat depth as a rank index of peatland likelihood.
-Since the purpose was to test the models' ability to uncover unmapped peatland (where DMK peat depth class is undefined), we evaluated models with two predictor configurations: \emph{terrain} or \emph{terrain + radiometric}.
-
 \subsubsection{Model interpretation}
@@
 \begin{figure}
-\centering
-\includegraphics{figures/modelmetrics.pdf}
-\caption{\label{fig:modelMetrics}Performance of peatland depth models with different predictor configurations, evaluated via spatial cross-validation.
-Parentheses denote the number of variables in each predictor configuration, and point estimates are shown +/- their standard error.}
+\includegraphics[width=1\linewidth]{figures/modelmetrics} \caption{Performance of peatland depth models with different predictor configurations, evaluated via spatial cross-validation.
+Parentheses denote the number of variables in each predictor configuration, and point estimates are shown +/- their standard error.}\label{fig:modelMetrics}
 \end{figure}
@@
 \begin{figure}
-\centering
-\includegraphics{figures/calibration_plots.pdf}
-\caption{\label{fig:calPlots}Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions from spatial cross-validation.
+\includegraphics[width=1\linewidth]{figures/calibration_plots} \caption{Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions from spatial cross-validation.
 Points are transparent to show overlap.
 Blue lines are local polynomial regressions and the red dashed line in each panel shows the 1:1 line.
-Marginal distributions (top) are shaded by quartile.}
-\end{figure}
-
-\subsubsection{Peatland extent}
-
-When we tested how well models extrapolated from within to outside of AR5 mire areas, we found that the models produced worse mean absolute error than just assuming a constant 30 cm depth (Fig.
-\ref{fig:modelMetricsExtrapolation}).
-With the independent occurrence data at Skrimfjella, we found that neither the \emph{terrain} nor \emph{terrain + radiometric} configurations were able to discriminate between peat presence and absence (area under the curve of the receiver operating characteristic 0.44 and 0.52 respectively, where 0.5 indicates random guessing).
-
-\begin{figure}
-\centering
-\includegraphics{figures/modelmetrics-extrapolation.pdf}
-\caption{\label{fig:modelMetricsExtrapolation}Performance of models that extrapolate from training data inside of mapped mires to test data outside of mapped mires (AR5 dataset), evaluated via spatial cross-validation.
-Parentheses denote the number of variables in each model, and point estimates are shown +/- their standard error.}
+Marginal distributions (top) are shaded by quartile.}\label{fig:calPlots}
 \end{figure}
@@
 \begin{figure}
-\centering
-\includegraphics{figures/variable_importance.pdf}
-\caption{\label{fig:varImp}Global variable importance in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), as measured by three different metrics: Shapley values, permutation importance, and the Feature Importance Ranking Measure \citep{greenwellVariableImportancePlots2020}.
-Variables removed due to collinearity are shown to the right of the variable with which they are most correlated.}
+\includegraphics[width=1\linewidth]{figures/variable_importance} \caption{Global variable importance in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), as measured by three different metrics: Shapley values, permutation importance, and the Feature Importance Ranking Measure [@greenwellVariableImportancePlots2020].
+Variables removed due to collinearity are shown to the right of the variable with which they are most correlated.}\label{fig:varImp}
 \end{figure}
@@
 \begin{figure}
-\centering
-\includegraphics{figures/partial_dependence.pdf}
-\caption{\label{fig:pdps}Partial dependence plots of the six most important variables in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b).
+\includegraphics[width=1\linewidth]{figures/partial_dependence} \caption{Partial dependence plots of the six most important variables in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b).
 The average effect of the predictor on the outcome (red line) overlays the variation in the effect across observations (grey lines).
-The variables' training data distributions are indicated with rug plots along the x-axes.}
+The variables' training data distributions are indicated with rug plots along the x-axes.}\label{fig:pdps}
 \end{figure}
@@
 Although this recommendation aligns with the precautionary principle, here we make it on technical grounds based on the maps' characteristics.
-
-\subsubsection{Depth predictions do not necessarily extrapolate to peat extent}
-
-That we measured more than 30 cm of peat in areas not mapped as mire was expected, because AR5 underrepresents peatland extent by about one third \citep{brynLandCoverNorway2018}.
-For that reason, maps of peatland extent also need revising.
-Unfortunately, our models did not predict peat occurrence outside AR5 mires.
-In these areas, assuming a constant 30 cm of peat (the depth threshold used in AR5) produced less error than our models (Fig. \ref{fig:modelMetricsExtrapolation}).
-Moreover, the performance gap would likely have been larger had our evaluation data been representative of all areas not classified as mire in AR5.
-Instead they were concentrated near AR5 mires.
-Most importantly, the independent occurrence data at Skrimfjella showed the model completely failing to predict occurrence, and these data were suited for testing this ability.
-Although the Ørskogfjellet model may have done better than the Skrimfjella model on a similar independent evaluation set (since it predicted depth better), the improvement would probably be small.
-
-Why wasn't peat occurrence predicted well?
-The models were trained on fundamentally different populations of locations (\textasciitilde80 \% AR5 mire in training vs.~0 \% in prediction), so it is not surprising that the associations they learned did not transfer well.
-Moreover, the \textasciitilde20 \% of training data from outside AR5 mires were incidentally collected (near AR5 mires) and not representative of other land cover classes.
-In short, the models were blind to the fact that most of both study areas have no peat.
-
-Peatland extent is probably best mapped using different remotely sensed data than we used here \citep{bakkestuenDelineationWetlandAreas2023}, and this study's purpose was not to map extent.
-Nevertheless --- as long as the relevant peatland definition includes a depth component --- depth predictions (or predictors thereof) should help delineate peatland extent \citep{olearyDigitalSoilMapping2022, beamishDetailedMappingPeat2024}.
-We return to this point in our discussion of implications for digital soil mapping.
 
@@
 
-Finally, we want to highlight the need for research on peatland extent and peat depth to be better integrated.
-Since peatland extent is defined by non-zero peat depth \citep[the specific threshold varies by definition,][]{minasnyMappingMonitoringPeatland2024}, they are fundamentally linked.
+Finally, we briefly want to highlight the need for peatland extent mapping and peat depth mapping to be better integrated.
+Since peatland extent is defined by non-zero peat depth \citep[the specific threshold varies by definition,][]{minasnyMappingMonitoringPeatland2024}, the lateral and vertical dimensions are fundamentally linked.
 The goal, therefore, should be a unified prediction framework for extent and depth.
+The distribution of peat depths across full landscapes is zero-inflated, and research is needed to determine whether it is more efficient to parameterize a single model of peat depth (with a larger, generalized dataset) or to break down the problem into a hurdle model by classifying zero depth and then regressing non-zero depths (with smaller, specialized datasets).
 We caution against reducing continuous depth predictions to arbitrary classes \citep[as in][]{ivanovsModelingGeospatialDistribution2024, karjalainenComparisonTwoGammaray2025}, since classes can be derived from continuous predictions.
-The distribution of peat depths across full landscapes is zero-inflated, and research is needed to determine whether it is more efficient to parameterize a single model of peat depth (with a larger, generalized dataset) or to break down the problem into a hurdle model by classifying zero depth and then regressing non-zero depths (with smaller, specialized datasets).
-Coupling extent and depth will reduce the prevalence of incoherence that we found: deep peat outside the peatland extent and zero depth inside it.
-A key challenge going forward will be obtaining training data that represents both zero and non-zero components of the depth distribution, since sampling designs often focus on known peatlands.
 

### 2025-05-12 ms/submission/AuthorComments-egusphere-2025-1046.qmd
address RC1.7: remove extent models
also added out.width to figures to avoid missing \pandocbounded macro from the custom template

@@
 A much shorter reference to the importance of peatland extent knowledge could perhaps be mentioned in the conclusions, but a full analysis and discussion (section 4.1.4) is not appropriate in this article.
+
+Thank you for this helpful suggestion.
+We initially thought it would be worth including the analysis on extent mapping---especially since our occurrence data at Skrimfjella is a good test set---but based on the comments of both referees we recognize that it distracts too much from the main focus.
+We have removed all sections related to mapping peatland extent (including section 4.1.4), to shorten the manuscript and improve its readability.
+We still discuss briefly the importance of coupling peat depth and extent mapping, since we consider this a research need directly related to the present study, but we have shortened this section as well.
 

### 2025-05-12 R/RUNALL.R

@@ -12,4 +12,3 @@ render("R/Skrim/createPredictorStack.R",output_format="html_document", knit_root
 render("R/Skrim/createDataFrame.R",output_format="html_document", knit_root_dir = "../../")
 render("R/Skrim/exploratoryDataAnalysis.R",output_format="html_document", knit_root_dir = "../../")
 render("R/Skrim/modelDepth.R", output_format="html_document", knit_root_dir = "../../")
-render("R/Skrim/modelDepth-evaluateOccurrence.R", output_format="html_document", knit_root_dir = "../../")

### 2025-05-12 R/Skrim/modelDepth-evaluateOccurrence.R

@@ -1,89 +1,0 @@
-library(tidyverse)
-library(sf)
-library(tidymodels)
-
-# Reading data ####
-
-frame.depth <- st_read("output/Skrim/modeling.gpkg", layer="dataframe") |> 
-  mutate(across(.cols = c(ar5cover, ar5soil, dmkdepth), .fns = as.factor)) |> 
-  filter(!(ar5cover %in% c(11,12)))
-frame.occ <- st_read("output/Skrim/modeling.gpkg", layer="dataframe.occurrence") |> 
-  mutate(across(.cols = c(occurrence, ar5cover, ar5soil), .fns = as.factor)) |> 
-  filter(!(ar5cover %in% c(11,12)))
-
-# Evaluation with independent data ####
-
-mod_rf <- 
-  rand_forest(mtry = NULL, min_n = 5, trees = 1000) %>% 
-  set_engine("ranger", importance = "permutation",
-             scale.permutation.importance	= TRUE) %>% 
-  set_mode("regression")
-
-## test ####
-
-### preds: radiometric + terrain ####
-
-recipe_RT <- 
-  recipe(formula = depth_cm ~ 
-           radK + radTh + radU + radTC +
-           elevation + 
-           slope1m + TPI1m + TRI1m + roughness1m + 
-           slope10m + TPI10m + TRI10m + roughness10m + 
-           MRVBF + 
-           TWI5m + TWI10m + TWI20m + TWI50m + 
-           DTW2500 + DTW5000 + DTW10000 + DTW20000 + DTW40000 + DTW80000 + DTW160000,
-         data = frame.depth)
-
-workflow_RT <- 
-  workflow() %>% 
-  add_model(mod_rf) %>% 
-  add_recipe(recipe_RT)
-
-set.seed(456)
-fit_RT <- 
-  workflow_RT %>% 
-  fit(frame.depth)
-
-eval_RT <- fit_RT %>% 
-  augment(new_data = frame.occ)
-
-eval_RT %>% 
-  roc_curve(truth = occurrence, .pred) %>% 
-  autoplot()
-eval_RT %>% 
-  roc_auc(truth = occurrence, .pred)
-
-### preds: terrain ####
-
-recipe_T <- 
-  recipe(formula = depth_cm ~ 
-           elevation + 
-           slope1m + TPI1m + TRI1m + roughness1m + 
[…]

### 2025-05-12 R/Skrim/modelDepth.R

@@ -229,84 +229,6 @@ preds %>%
 mean(preds$.pred - preds$depth_cm)
 readr::write_csv(preds, "output/Skrim/calibrationplot.csv")
 
-## Extrapolating beyond mire ####
-
-frame |> 
-  st_drop_geometry() |> 
-  group_by(ar5cover == 60) |> 
-  summarise(n = n(), mean(depth_cm))
-
-# Creating rsample splits
-custom_splits_extrap <- list()
-for (k in seq_along(unique(knndm$clusters))) {
-  custom_splits_extrap[[k]] <- make_splits(
-    list(analysis = intersect(knndm$indx_train[[k]], 
-                              which(frame$ar5cover == 60)), 
-         assessment = intersect(knndm$indx_test[[k]], 
-                                which(frame$ar5cover != 60))), 
-    frame)
-}
-names(custom_splits_extrap) <- paste0("Fold", unique(knndm$clusters))
-custom_splits_extrap <- purrr::discard(custom_splits_extrap, \(x) length(x$out_id) == 0)
-  
-folds_extrap <- manual_rset(splits = custom_splits_extrap, 
-                            ids = names(custom_splits_extrap))
-folds_extrap
-
-### preds: radiometric + terrain ####
-
-set.seed(456)
-fit_RT_extrap <- 
-  workflow_RT %>% 
-  fit_resamples(
-    resamples = folds_extrap,
-    metrics = metric_set(rmse, mae),
-    control = control_resamples(save_pred = TRUE))
-# Warnings for calculating rsq on assessment folds of 1.
-collect_metrics(fit_RT_extrap)
-
-RT_extrap <- collect_predictions(fit_RT_extrap) |> 
-  select(.pred, depth_cm)
-cor.test(RT_extrap$depth_cm, RT_extrap$.pred)
-
-### preds: terrain ####
-
-set.seed(456)
-fit_T_extrap <- 
-  workflow_T %>% 
-  fit_resamples(
-    resamples = folds_extrap,
-    metrics = metric_set(rmse, mae),
-    control = control_resamples(save_pred = TRUE))
-# Warnings for calculating rsq on assessment folds of 1.
-collect_metrics(fit_T_extrap)
-
-T_extrap <- collect_predictions(fit_T_extrap) |> 
-  select(.pred, depth_cm)
-cor.test(T_extrap$depth_cm, T_extrap$.pred)
-
[…]

### 2025-05-12 R/modelDepth.R

@@ -229,84 +229,6 @@ preds %>%
 mean(preds$.pred - preds$depth_cm)
 readr::write_csv(preds, "output/calibrationplot.csv")
 
-## Extrapolating beyond mire ####
-
-frame |> 
-  st_drop_geometry() |> 
-  group_by(ar5cover == 60) |> 
-  summarise(n = n(), mean(depth_cm))
-
-# Creating rsample splits
-custom_splits_extrap <- list()
-for (k in seq_along(unique(knndm$clusters))) {
-  custom_splits_extrap[[k]] <- make_splits(
-    list(analysis = intersect(knndm$indx_train[[k]], 
-                              which(frame$ar5cover == 60)), 
-         assessment = intersect(knndm$indx_test[[k]], 
-                                which(frame$ar5cover != 60))), 
-    frame)
-}
-names(custom_splits_extrap) <- paste0("Fold", unique(knndm$clusters))
-custom_splits_extrap <- purrr::discard(custom_splits_extrap, \(x) length(x$out_id) == 0)
-  
-folds_extrap <- manual_rset(splits = custom_splits_extrap, 
-                            ids = names(custom_splits_extrap))
-folds_extrap
-
-### preds: radiometric + terrain ####
-
-set.seed(456)
-fit_RT_extrap <- 
-  workflow_RT %>% 
-  fit_resamples(
-    resamples = folds_extrap,
-    metrics = metric_set(rmse, mae),
-    control = control_resamples(save_pred = TRUE))
-# Warnings for calculating rsq on assessment folds of 1.
-collect_metrics(fit_RT_extrap)
-
-RT_extrap <- collect_predictions(fit_RT_extrap) |> 
-  select(.pred, depth_cm)
-cor.test(RT_extrap$depth_cm, RT_extrap$.pred)
-
-### preds: terrain ####
-
-set.seed(456)
-fit_T_extrap <- 
-  workflow_T %>% 
-  fit_resamples(
-    resamples = folds_extrap,
-    metrics = metric_set(rmse, mae),
-    control = control_resamples(save_pred = TRUE))
-# Warnings for calculating rsq on assessment folds of 1.
-collect_metrics(fit_T_extrap)
-
-T_extrap <- collect_predictions(fit_T_extrap) |> 
-  select(.pred, depth_cm)
-cor.test(T_extrap$depth_cm, T_extrap$.pred)
-
-### preds: none (assume 30cm) ####
-
[…]

### 2025-05-12 ms/figures.R

@@ -357,55 +357,6 @@ ggsave(filename = 'modelmetrics.pdf', path = "ms/figures",
 ggsave(filename = 'modelmetrics.png', path = "ms/figures",
        width =210-30, height = (240-40)/2.5, units = 'mm', dpi = 300) #copernicus.cls page 210x240
 
-# Model metrics extrapolation####
-
-library(tidyverse)
-
-orskog <- read_csv("output/modelmetrics-extrapolation.csv") 
-skrim <- read_csv("output/Skrim/modelmetrics-extrapolation.csv")
-plotting <- bind_rows(orskog = orskog, skrim = skrim, .id = 'site') %>% 
-  filter(.metric != 'rmse') %>% 
-  mutate(
-    site = fct_relevel(site, "skrim"),
-    metric = case_when(
-      .metric == "mae" ~ "Mean \nabsolute error"),
-    model = case_when(
-      model == "null" ~ "null, assume 30 cm",
-      model == "Terrain" ~ "terrain (21)",
-      model == "RadiometricTerrain" ~ "terrain + radiometric (25)"),
-    model = fct_relevel(model,
-                        "null, assume 30 cm",
-                        "terrain (21)",
-                        "terrain + radiometric (25)"))
-str(plotting)
-
-cols <- c("orskog" = "#d95f02", "skrim" = "#1b9e77") #https://colorbrewer2.org/#type=qualitative&scheme=Dark2&n=3
-
-g1 <- ggplot(plotting) +
-  geom_linerange(aes(y = model, 
-                     x = mean, 
-                     xmin = mean - std_err, 
-                     xmax = mean + std_err, 
-                     color = site),
-                 position = position_dodge2(width= 0.3, reverse = TRUE)) +
-  geom_point(aes(y = model, x = mean, color = site),
-             position = position_dodge2(width= 0.3, reverse = TRUE)) +
-  geom_text(aes(y = model, x = mean, label = signif(mean, 2), group = site),
-            position = position_dodge2(width=1, reverse = TRUE),
-            color = "grey50", size = 2.5) +
-  scale_colour_manual(values = cols) +
-  annotate(geom='text', x=70, y=0.85, size=3, label='\u00D8rskogfjellet', color = "#d95f02") +
-  annotate(geom='text', x=70, y=1.15, size=3, label='Skrimfjella', color = "#1b9e77") +
-  guides(color = "none") +
[…]

### 2025-05-27 ms/ms.Rmd
address RC2.3: add configurations and configuration comparisons

@@
   codedata: |
-    R code used in this study is available at <https://github.com/julienvollering/DSMdepth>.
+    *Code and data availability.* R code used in this study is available at <https://github.com/julienvollering/DSMdepth>.
 The depth measurements we used are archived at \doi{10.6073/pasta/6ce440152f693f2156bf5b692a2e7917} and follow data and metadata standards.
@@
 We maintain full responsibility for the scientific output, as per the European Commission's *Living guidelines on the responsible use of generative AI in research*.
-appendix: |
-  \appendixtables
-  \begin{table}[tbp]
-  \caption{Attributes of the radiometric surveys, as reported in Baranwal et al., \citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013} and Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.}
-  \begin{tabular}{lll}
-  \hline
-                                           & Skrimfjella & Ørskogfjellet              \\ \cline{2-3} 
-  Survey period                            & 2008-2011   & December 2014–January 2015 \\
-  Average flight altitude (m)              & 75          & 80                         \\
-  Average flight speed (\unit{km\,h^{-1}}) & 108         & 88                         \\
-  Flight line spacing (m)                  & 200         & 200                        \\ \hline
-  \end{tabular}
-  \label{tab:radSurveys}
-  \end{table}
-  \clearpage
-  \appendixfigures
-  \begin{figure}
-  \centering
-  \includegraphics[width=0.8\textwidth]{figures/GPRwavevelocity.pdf}
-  \caption{\label{fig:GPRwavevelocity}Calibration of GPR wave velocity at Skrimfjella (a) and Ørskogfjellet (b) by regression of probed depth against wave travel time.}
-  \end{figure}
-  \clearpage
-  \appendixfigures
-  \begin{figure}
-  \centering
-  \includegraphics{figures/semivariograms.pdf}
-  \caption{\label{fig:semivariograms}Empirical semivariograms of peat depth point measurements at Skrimfjella (a) and Ørskogfjellet (b).}
-  \end{figure}
 output:
@@
 
-For both sites, we compared the performance of models with five different configurations of predictors. 
-Specifically, we trained models with:
-
-  1. *DMK* (2 variables)
-  2. *terrain* (21 variables)
-  3. *terrain + DMK* (23 variables)
-  4. *terrain + radiometric* (25 variables)
-  5. *all predictors* (27 variables)
-
-These different configurations simulate different scenarios of data availability.
+For both sites, we compared the performance of models with different predictor configurations, where each configuration was one of the seven combinations of the three predictor groups: (1) radiometric (2) terrain and (3) DMK.
 Comparing the different configurations allowed us to isolate the added value of each of the predictor groups.
-We did not explore configurations comprising radiometrics without terrain, because LiDAR terrain surveys typically precede airborne radiometric surveys.
 The models with only DMK peat depth class were simple linear models rather than RFs, and served to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth, calibrated on the same data.
@@
 From the cross-validation we quantified mean absolute error (error magnitude, original scale), R^2^ (explained variation, standardized scale), and Lin's concordance correlation coefficient (error magnitude and explained variation, standardized scale).
+We formally assessed the effect of predictor configuration on performance metrics using mixed-effects models to account for the cross-validation fold structure, and testing pairwise differences between configurations.
 
@@
 
-For Skrimfjella, the best predictor configuration was *all predictors*, followed closely by *terrain + radiometric*.
-The performance gap between the *terrain + DMK* configuration and the *terrain* configuration was similarly small.
-Compared to the *terrain* configuration, the *terrain + radiometric* configuration improved concordance correlation by 0.04, R^2^ by 0.04, and mean absolute error by 1 cm.
-*DMK* was a very poor predictor of peat depth even though it was calibrated to measured depths, with a concordance correlation coefficient of 0.0077.
+For Skrimfjella, the best predictor configuration was *all predictors*, followed closely by *terrain + radiometric* (Fig. \@ref(fig:modelMetrics)).
+The *terrain* configuration outperformed the *radiometric* configuration by 0.27 in concordance correlation coefficient ($p < 0.001$; Appendix Table \@ref(tab:pairwiseSkrimCCC)), by 0.16 in R^2^ ($p = 0.190$; Appendix Table \@ref(tab:pairwiseSkrimRsq)), and by 10 cm in mean absolute error ($p = 0.038$; Appendix Table \@ref(tab:pairwiseSkrimMAE)).
+The *terrainDMK* configuration outperformed the *radiometricDMK* configuration by 0.32 in concordance correlation coefficient ($p < 0.001$; Appendix Table \@ref(tab:pairwiseSkrimCCC)), by 0.11 in R^2^ ($p = 0.605$; Appendix Table \@ref(tab:pairwiseSkrimRsq)), and by 11 cm in mean absolute error ($p = 0.011$; Appendix Table \@ref(tab:pairwiseSkrimMAE)).
 
-For Ørskogfjellet, the best predictor configuration was *terrain + DMK*, followed by *terrain*.
-Adding radiometric predictors to these configurations worsened model performance, especially in terms of concordance correlation and R^2^.
-By itself, *DMK class* produced a concordance correlation coefficient of 0.17 (compared to 0.008 at Skrimfjella), but the worst mean absolute error of any model at either site: 77 cm.
+For Ørskogfjellet, the best predictor configuration was *terrain + DMK*, followed by *terrain* (Fig. \@ref(fig:modelMetrics)).
+Adding radiometric predictors to these configurations slightly worsened model performance.
+The *terrain* configuration outperformed the *radiometric* configuration by 0.24 in concordance correlation coefficient ($p = 0.068$; Appendix Table \@ref(tab:pairwiseOrskogCCC)), by 0.24 in R^2^ ($p = 0.004$; Appendix Table \@ref(tab:pairwiseOrskogRsq)), and by 25 cm in mean absolute error ($p = 0.086$; Appendix Table \@ref(tab:pairwiseOrskogMAE)).
+The *terrainDMK* configuration outperformed the *radiometricDMK* configuration by 0.28 in concordance correlation coefficient ($p = 0.015$; Appendix Table \@ref(tab:pairwiseOrskogCCC)), by 0.27 in R^2^ ($p < 0.001$; Appendix Table \@ref(tab:pairwiseOrskogRsq)), and by 28 cm in mean absolute error ($p = 0.031$; Appendix Table \@ref(tab:pairwiseOrskogMAE)).
 
@@
 100 m).
-Independent probing and GPR measurements up to 2 m apart were differed by an average of 29--42 cm (Fig. \@ref(fig:GPRwavevelocity)).
+Independent probing and GPR measurements up to 2 m apart differed by an average of 29--42 cm (Fig. \@ref(fig:GPRwavevelocity)).
 By comparison, the average variation among measurements 100 m apart was 70--106 cm (Fig. \@ref(fig:semivariograms)).
@@
 
-Radiometric data had no predictive value at Ørskogfjellet, while at Skrimfjella they had minor influence in a relatively weak model. 
+Radiometric data had little to no predictive value at either site (poor performance of *radiometric* configuration), although they did contribute to the best model at Skrimfjella (marginal improvement in *all predictors* configuration compared to *terrain + DMK*).
+These results contrast with earlier studies that found that radiometrics were useful predictors of peat depth [@keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @kogantiMappingPeatDepth2023].
 The bedrock is more homogeneous at Ørskogfjellet than at Skrimfjella, so uneven radiogenesis is not a viable explanation for the differences between sites nor the poor performance in general [@beamishEnvironmentalRadioactivityUK2014; @reinhardtGammaraySpectrometryVersatile2019].
-To the degree that radiometrics had predictive value at Skrimfjella, it appears that they were most valuable near the extremes of the depth distribution, since their inclusion improved R^2^ more than mean absolute error.
-All four variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value. 
+All four radiometric variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value. 
 This contrasts with Koganti et al. [-@kogantiMappingPeatDepth2023], who found that radiometric total count was a much better predictor than potassium ground concentration.
@@
 
-At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables (Fig. \@ref(fig:varImp)). 
-Elevation was the most important predictor at both sites, and peat depth showed non-monotonic responses to changes in elevation (Fig. \@ref(fig:pdps)). 
+At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables (Fig. \@ref(fig:modelMetrics)). 
+Elevation was the most important predictor at both sites, and peat depth showed non-monotonic responses to changes in elevation (Figs. \@ref(fig:varImp)-\@ref(fig:pdps)). 
 We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g.
@@
 Though peatland definitions may encourage reducing continuous depth predictions to arbitrary classes, we caution against this practice [as in @ivanovsModelingGeospatialDistribution2024; @karjalainenComparisonTwoGammaray2025].
+
+\newpage
+\appendix
+\section{}
+
+\appendixtables
+\begin{table}[h]
+\caption{Attributes of the radiometric surveys, as reported in Baranwal et al., \citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013} and Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.}
+\begin{tabular}{lll}
+\hline
+& Skrimfjella & Ørskogfjellet              \\ \cline{2-3} 
+Survey period                            & 2008-2011   & December 2014–January 2015 \\
+Average flight altitude (m)              & 75          & 80                         \\
+Average flight speed (\unit{km\,h^{-1}}) & 108         & 88                         \\
+Flight line spacing (m)                  & 200         & 200                        \\ \hline
+\end{tabular}
+\label{tab:radSurveys}
+\end{table}
+\clearpage
+
+\appendixfigures
+\begin{figure}[ht]
+\centering
+\includegraphics[width=0.8\textwidth]{figures/GPRwavevelocity.pdf}
+\caption{\label{fig:GPRwavevelocity}Calibration of GPR wave velocity at Skrimfjella (a) and Ørskogfjellet (b) by regression of probed depth against wave travel time.}
+\end{figure}
+\clearpage
+
+\newpage
+\section{}
+
+```{r pairwiseSkrimCCC, echo=FALSE, message=FALSE}
+tab <- readr::read_csv("tables/pairwise-skrim-Concordance correlation.csv")
+options(knitr.kable.NA = '')
+kableExtra::kbl(tab, format = 'latex', vline = "", linesep = "", position = "h",
+                caption = "Pairwise comparisons of the concordance correlation coefficient among models for Skrimfjella, based on 10-fold cross-validation.
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package) on mixed-effects models (\\texttt{lme4} package) to account for the cross-validation fold structure.")
+```
+\clearpage
+
+```{r pairwiseSkrimRsq, echo=FALSE, message=FALSE}
+tab <- readr::read_csv("tables/pairwise-skrim-R-squared.csv")
+options(knitr.kable.NA = '')
+kableExtra::kbl(tab, format = 'latex', vline = "", linesep = "", position = "h",
+                caption = "Pairwise comparisons of $R^2$ among models for Skrimfjella, based on 10-fold cross-validation.
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package) on mixed-effects models (\\texttt{lme4} package) to account for the cross-validation fold structure.")
+```
+\clearpage
+
+```{r pairwiseSkrimMAE, echo=FALSE, message=FALSE}
+tab <- readr::read_csv("tables/pairwise-skrim-Mean absolute error.csv")
+options(knitr.kable.NA = '')
+kableExtra::kbl(tab, format = 'latex', vline = "", linesep = "", position = "h",
+                caption = "Pairwise comparisons of mean absolute error among models for Skrimfjella, based on 10-fold cross-validation.
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package) on mixed-effects models (\\texttt{lme4} package) to account for the cross-validation fold structure.")
+```
+\clearpage
+
+```{r pairwiseOrskogCCC, echo=FALSE, message=FALSE}
+tab <- readr::read_csv("tables/pairwise-orskog-Concordance correlation.csv")
+options(knitr.kable.NA = '')
+kableExtra::kbl(tab, format = 'latex', vline = "", linesep = "", position = "h",
+                caption = "Pairwise comparisons of the concordance correlation coefficient among models for Ørskogfjellet, based on 10-fold cross-validation.
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package) on mixed-effects models (\\texttt{lme4} package) to account for the cross-validation fold structure.")
+```
+\clearpage
+
+```{r pairwiseOrskogRsq, echo=FALSE, message=FALSE}
+tab <- readr::read_csv("tables/pairwise-orskog-R-squared.csv")
+options(knitr.kable.NA = '')
+kableExtra::kbl(tab, format = 'latex', vline = "", linesep = "", position = "h",
+                caption = "Pairwise comparisons of $R^2$ among models for Ørskogfjellet, based on 10-fold cross-validation.
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package) on mixed-effects models (\\texttt{lme4} package) to account for the cross-validation fold structure.")
+```
+\clearpage
+
+```{r pairwiseOrskogMAE, echo=FALSE, message=FALSE}
+tab <- readr::read_csv("tables/pairwise-orskog-Mean absolute error.csv")
+options(knitr.kable.NA = '')
+kableExtra::kbl(tab, format = 'latex', vline = "", linesep = "", position = "h",
+                caption = "Pairwise comparisons of mean absolute error among models for Ørskogfjellet, based on 10-fold cross-validation.
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package) on mixed-effects models (\\texttt{lme4} package) to account for the cross-validation fold structure.")
+```
+\clearpage
+
+\begin{figure}[ht]
+\centering
+\includegraphics{figures/semivariograms.pdf}
+\caption{\label{fig:semivariograms}Empirical semivariograms of peat depth point measurements at Skrimfjella (a) and Ørskogfjellet (b).}
+\end{figure}
+\clearpage

### 2025-05-27 ms/ms.tex
address RC2.3: add configurations and configuration comparisons

@@
 
+%%\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 %
@@
 
-For both sites, we compared the performance of models with five different configurations of predictors.
-Specifically, we trained models with:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \emph{DMK} (2 variables)
-\item
-  \emph{terrain} (21 variables)
-\item
-  \emph{terrain + DMK} (23 variables)
-\item
-  \emph{terrain + radiometric} (25 variables)
-\item
-  \emph{all predictors} (27 variables)
-\end{enumerate}
-
-These different configurations simulate different scenarios of data availability.
+For both sites, we compared the performance of models with different predictor configurations, where each configuration was one of the seven combinations of the three predictor groups: (1) radiometric (2) terrain and (3) DMK.
 Comparing the different configurations allowed us to isolate the added value of each of the predictor groups.
-We did not explore configurations comprising radiometrics without terrain, because LiDAR terrain surveys typically precede airborne radiometric surveys.
 The models with only DMK peat depth class were simple linear models rather than RFs, and served to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth, calibrated on the same data.
@@
 From the cross-validation we quantified mean absolute error (error magnitude, original scale), R\textsuperscript{2} (explained variation, standardized scale), and Lin's concordance correlation coefficient (error magnitude and explained variation, standardized scale).
+We formally assessed the effect of predictor configuration on performance metrics using mixed-effects models to account for the cross-validation fold structure, and testing pairwise differences between configurations.
 
@@
 
-For Skrimfjella, the best predictor configuration was \emph{all predictors}, followed closely by \emph{terrain + radiometric}.
-The performance gap between the \emph{terrain + DMK} configuration and the \emph{terrain} configuration was similarly small.
-Compared to the \emph{terrain} configuration, the \emph{terrain + radiometric} configuration improved concordance correlation by 0.04, R\textsuperscript{2} by 0.04, and mean absolute error by 1 cm.
-\emph{DMK} was a very poor predictor of peat depth even though it was calibrated to measured depths, with a concordance correlation coefficient of 0.0077.
-
-For Ørskogfjellet, the best predictor configuration was \emph{terrain + DMK}, followed by \emph{terrain}.
-Adding radiometric predictors to these configurations worsened model performance, especially in terms of concordance correlation and R\textsuperscript{2}.
-By itself, \emph{DMK class} produced a concordance correlation coefficient of 0.17 (compared to 0.008 at Skrimfjella), but the worst mean absolute error of any model at either site: 77 cm.
+For Skrimfjella, the best predictor configuration was \emph{all predictors}, followed closely by \emph{terrain + radiometric} (Fig. \ref{fig:modelMetrics}).
+The \emph{terrain} configuration outperformed the \emph{radiometric} configuration by 0.27 in concordance correlation coefficient (\(p < 0.001\); Appendix Table \ref{tab:pairwiseSkrimCCC}), by 0.16 in R\textsuperscript{2} (\(p = 0.190\); Appendix Table \ref{tab:pairwiseSkrimRsq}), and by 10 cm in mean absolute error (\(p = 0.038\); Appendix Table \ref{tab:pairwiseSkrimMAE}).
+The \emph{terrainDMK} configuration outperformed the \emph{radiometricDMK} configuration by 0.32 in concordance correlation coefficient (\(p < 0.001\); Appendix Table \ref{tab:pairwiseSkrimCCC}), by 0.11 in R\textsuperscript{2} (\(p = 0.605\); Appendix Table \ref{tab:pairwiseSkrimRsq}), and by 11 cm in mean absolute error (\(p = 0.011\); Appendix Table \ref{tab:pairwiseSkrimMAE}).
+
+For Ørskogfjellet, the best predictor configuration was \emph{terrain + DMK}, followed by \emph{terrain} (Fig. \ref{fig:modelMetrics}).
+Adding radiometric predictors to these configurations slightly worsened model performance.
+The \emph{terrain} configuration outperformed the \emph{radiometric} configuration by 0.24 in concordance correlation coefficient (\(p = 0.068\); Appendix Table \ref{tab:pairwiseOrskogCCC}), by 0.24 in R\textsuperscript{2} (\(p = 0.004\); Appendix Table \ref{tab:pairwiseOrskogRsq}), and by 25 cm in mean absolute error (\(p = 0.086\); Appendix Table \ref{tab:pairwiseOrskogMAE}).
+The \emph{terrainDMK} configuration outperformed the \emph{radiometricDMK} configuration by 0.28 in concordance correlation coefficient (\(p = 0.015\); Appendix Table \ref{tab:pairwiseOrskogCCC}), by 0.27 in R\textsuperscript{2} (\(p < 0.001\); Appendix Table \ref{tab:pairwiseOrskogRsq}), and by 28 cm in mean absolute error (\(p = 0.031\); Appendix Table \ref{tab:pairwiseOrskogMAE}).
 
@@
 Ground-based measurements also have uncertainty \citep{parryEvaluatingApproachesEstimating2014}, but our data show that this error was smaller than the variation in peat depth over short distances (e.g.~100 m).
-Independent probing and GPR measurements up to 2 m apart were differed by an average of 29--42 cm (Fig. \ref{fig:GPRwavevelocity}).
+Independent probing and GPR measurements up to 2 m apart differed by an average of 29--42 cm (Fig. \ref{fig:GPRwavevelocity}).
 By comparison, the average variation among measurements 100 m apart was 70--106 cm (Fig. \ref{fig:semivariograms}).
@@
 
-Radiometric data had no predictive value at Ørskogfjellet, while at Skrimfjella they had minor influence in a relatively weak model.
+Radiometric data had little to no predictive value at either site (poor performance of \emph{radiometric} configuration), although they did contribute to the best model at Skrimfjella (marginal improvement in \emph{all predictors} configuration compared to \emph{terrain + DMK}).
+These results contrast with earlier studies that found that radiometrics were useful predictors of peat depth \citep{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, kogantiMappingPeatDepth2023}.
 The bedrock is more homogeneous at Ørskogfjellet than at Skrimfjella, so uneven radiogenesis is not a viable explanation for the differences between sites nor the poor performance in general \citep{beamishEnvironmentalRadioactivityUK2014, reinhardtGammaraySpectrometryVersatile2019}.
-To the degree that radiometrics had predictive value at Skrimfjella, it appears that they were most valuable near the extremes of the depth distribution, since their inclusion improved R\textsuperscript{2} more than mean absolute error.
-All four variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value.
+All four radiometric variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value.
 This contrasts with Koganti et al. \citeyearpar{kogantiMappingPeatDepth2023}, who found that radiometric total count was a much better predictor than potassium ground concentration.
@@
 
-At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables (Fig. \ref{fig:varImp}).
-Elevation was the most important predictor at both sites, and peat depth showed non-monotonic responses to changes in elevation (Fig. \ref{fig:pdps}).
+At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables (Fig. \ref{fig:modelMetrics}).
+Elevation was the most important predictor at both sites, and peat depth showed non-monotonic responses to changes in elevation (Figs. \ref{fig:varImp}-\ref{fig:pdps}).
 We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g.~via climate) to explain the observed patterns.
@@
 The goal, therefore, should be a unified prediction framework for extent and depth.
-The distribution of peat depths across full landscapes is zero-inflated, and research is needed to determine whether it is more efficient to parameterize a single model of peat depth (with a larger, generalized dataset) or to break down the problem into a hurdle model by classifying zero depth and then regressing non-zero depths (with smaller, specialized datasets).
-We caution against reducing continuous depth predictions to arbitrary classes \citep[as in][]{ivanovsModelingGeospatialDistribution2024, karjalainenComparisonTwoGammaray2025}, since classes can be derived from continuous predictions.
-
-\codedataavailability{R code used in this study is available at \url{https://github.com/julienvollering/DSMdepth}.
-The depth measurements we used are archived at \doi{10.6073/pasta/6ce440152f693f2156bf5b692a2e7917} and follow data and metadata standards.} %% use this section when having data sets and software code available
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% optional
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Research is needed to determine whether it is better to parameterize a single model of peat depth across a full landscape, or to break down the problem into a hurdle model by classifying zero depth and then regressing non-zero depths.
+Though peatland definitions may encourage reducing continuous depth predictions to arbitrary classes, we caution against this practice \citep[as in][]{ivanovsModelingGeospatialDistribution2024, karjalainenComparisonTwoGammaray2025}.
+
+\newpage
 \appendix
+\section{}
+
 \appendixtables
-\begin{table}[tbp]
+\begin{table}[h]
 \caption{Attributes of the radiometric surveys, as reported in Baranwal et al., \citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013} and Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.}
@@
 \clearpage
+
 \appendixfigures
-\begin{figure}
+\begin{figure}[ht]
 \centering
@@
 \clearpage
-\appendixfigures
-\begin{figure}
+
+\newpage
+\section{}
+
+\begin{table}[h]
+
+\caption{\label{tab:pairwiseSkrimCCC}Pairwise comparisons of the concordance correlation coefficient among models for Skrimfjella, based on 10-fold cross-validation.
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package) on mixed-effects models (\texttt{lme4} package) to account for the cross-validation fold structure.}
+\centering
+\begin{tabular}[t]{lrrrrll}
+\hline
+Comparison & Estimate & Std.error & Df & Statistic & Adj.p.value & Significance\\
+\hline
+DMK - Radiometric & 0.02 & 0.05 & 114 & 0.44 & 0.999 & \\
+DMK - RadiometricDMK & 0.06 & 0.05 & 114 & 1.19 & 0.896 & \\
+DMK - RadiometricTerrain & -0.29 & 0.05 & 114 & -5.84 & < 0.001 & ***\\
+DMK - RadiometricTerrainDMK & -0.29 & 0.05 & 114 & -5.81 & < 0.001 & ***\\
+DMK - Terrain & -0.25 & 0.05 & 114 & -5.03 & < 0.001 & ***\\
+DMK - TerrainDMK & -0.26 & 0.05 & 114 & -5.17 & < 0.001 & ***\\
+Radiometric - RadiometricDMK & 0.04 & 0.05 & 114 & 0.75 & 0.989 & \\
+Radiometric - RadiometricTerrain & -0.31 & 0.05 & 114 & -6.28 & < 0.001 & ***\\
+Radiometric - RadiometricTerrainDMK & -0.31 & 0.05 & 114 & -6.25 & < 0.001 & ***\\
+Radiometric - Terrain & -0.27 & 0.05 & 114 & -5.47 & < 0.001 & ***\\
+Radiometric - TerrainDMK & -0.28 & 0.05 & 114 & -5.62 & < 0.001 & ***\\
+RadiometricDMK - RadiometricTerrain & -0.35 & 0.05 & 114 & -7.03 & < 0.001 & ***\\
+RadiometricDMK - RadiometricTerrainDMK & -0.35 & 0.05 & 114 & -7.00 & < 0.001 & ***\\
+RadiometricDMK - Terrain & -0.31 & 0.05 & 114 & -6.22 & < 0.001 & ***\\
+RadiometricDMK - TerrainDMK & -0.32 & 0.05 & 114 & -6.37 & < 0.001 & ***\\
+RadiometricTerrain - RadiometricTerrainDMK & 0.00 & 0.05 & 114 & 0.03 & 1 & \\
+RadiometricTerrain - Terrain & 0.04 & 0.05 & 114 & 0.81 & 0.983 & \\
+RadiometricTerrain - TerrainDMK & 0.03 & 0.05 & 114 & 0.67 & 0.994 & \\
+RadiometricTerrainDMK - Terrain & 0.04 & 0.05 & 114 & 0.79 & 0.986 & \\
+RadiometricTerrainDMK - TerrainDMK & 0.03 & 0.05 & 114 & 0.64 & 0.995 & \\
+Terrain - TerrainDMK & -0.01 & 0.05 & 114 & -0.15 & 1 & \\
+\hline
+\end{tabular}
+\end{table}
+\clearpage
+
+\begin{table}[h]
+
+\caption{\label{tab:pairwiseSkrimRsq}Pairwise comparisons of $R^2$ among models for Skrimfjella, based on 10-fold cross-validation.
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package) on mixed-effects models (\texttt{lme4} package) to account for the cross-validation fold structure.}
+\centering
+\begin{tabular}[t]{lrrrrrl}
+\hline
+Comparison & Estimate & Std.error & Df & Statistic & Adj.p.value & Significance\\
+\hline
+DMK - Radiometric & 0.04 & 0.08 & 108.23 & 0.55 & 0.998 & \\
+DMK - RadiometricDMK & 0.01 & 0.08 & 108.23 & 0.08 & 1.000 & \\
+DMK - RadiometricTerrain & -0.16 & 0.08 & 108.23 & -2.07 & 0.378 & \\
+DMK - RadiometricTerrainDMK & -0.16 & 0.08 & 108.23 & -2.11 & 0.357 & \\
+DMK - Terrain & -0.11 & 0.08 & 108.23 & -1.48 & 0.755 & \\
+DMK - TerrainDMK & -0.10 & 0.08 & 108.23 & -1.34 & 0.830 & \\
+Radiometric - RadiometricDMK & -0.04 & 0.06 & 105.01 & -0.56 & 0.998 & \\
+Radiometric - RadiometricTerrain & -0.20 & 0.06 & 105.01 & -3.16 & 0.033 & *\\
+Radiometric - RadiometricTerrainDMK & -0.20 & 0.06 & 105.01 & -3.20 & 0.029 & *\\
+Radiometric - Terrain & -0.16 & 0.06 & 105.01 & -2.45 & 0.190 & \\
+Radiometric - TerrainDMK & -0.15 & 0.06 & 105.01 & -2.28 & 0.263 & \\
+RadiometricDMK - RadiometricTerrain & -0.17 & 0.06 & 105.01 & -2.60 & 0.137 & \\
+RadiometricDMK - RadiometricTerrainDMK & -0.17 & 0.06 & 105.01 & -2.64 & 0.124 & \\
+RadiometricDMK - Terrain & -0.12 & 0.06 & 105.01 & -1.89 & 0.494 & \\
+RadiometricDMK - TerrainDMK & -0.11 & 0.06 & 105.01 & -1.72 & 0.605 & \\
+RadiometricTerrain - RadiometricTerrainDMK & 0.00 & 0.06 & 105.01 & -0.04 & 1.000 & \\
+RadiometricTerrain - Terrain & 0.05 & 0.06 & 105.01 & 0.71 & 0.992 & \\
+RadiometricTerrain - TerrainDMK & 0.06 & 0.06 & 105.01 & 0.88 & 0.975 & \\
+RadiometricTerrainDMK - Terrain & 0.05 & 0.06 & 105.01 & 0.75 & 0.989 & \\
+RadiometricTerrainDMK - TerrainDMK & 0.06 & 0.06 & 105.01 & 0.92 & 0.968 & \\
+Terrain - TerrainDMK & 0.01 & 0.06 & 105.01 & 0.17 & 1.000 & \\
+\hline
+\end{tabular}
+\end{table}
+\clearpage
+
+\begin{table}[h]
+
+\caption{\label{tab:pairwiseSkrimMAE}Pairwise comparisons of mean absolute error among models for Skrimfjella, based on 10-fold cross-validation.
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package) on mixed-effects models (\texttt{lme4} package) to account for the cross-validation fold structure.}
+\centering
+\begin{tabular}[t]{lrrrrrl}
+\hline
+Comparison & Estimate & Std.error & Df & Statistic & Adj.p.value & Significance\\
+\hline
+DMK - Radiometric & -2.91 & 3.13 & 114 & -0.93 & 0.967 & \\
+DMK - RadiometricDMK & -3.69 & 3.13 & 114 & -1.18 & 0.902 & \\
+DMK - RadiometricTerrain & 8.10 & 3.13 & 114 & 2.58 & 0.140 & \\
+DMK - RadiometricTerrainDMK & 8.21 & 3.13 & 114 & 2.62 & 0.130 & \\
+DMK - Terrain & 6.81 & 3.13 & 114 & 2.17 & 0.319 & \\
+DMK - TerrainDMK & 7.33 & 3.13 & 114 & 2.34 & 0.235 & \\
+Radiometric - RadiometricDMK & -0.77 & 3.13 & 114 & -0.25 & 1.000 & \\
+Radiometric - RadiometricTerrain & 11.02 & 3.13 & 114 & 3.51 & 0.011 & *\\
+Radiometric - RadiometricTerrainDMK & 11.13 & 3.13 & 114 & 3.55 & 0.010 & *\\
+Radiometric - Terrain & 9.72 & 3.13 & 114 & 3.10 & 0.038 & *\\
+Radiometric - TerrainDMK & 10.24 & 3.13 & 114 & 3.27 & 0.024 & *\\
+RadiometricDMK - RadiometricTerrain & 11.79 & 3.13 & 114 & 3.76 & 0.005 & **\\
+RadiometricDMK - RadiometricTerrainDMK & 11.90 & 3.13 & 114 & 3.80 & 0.004 & **\\
+RadiometricDMK - Terrain & 10.50 & 3.13 & 114 & 3.35 & 0.018 & *\\
+RadiometricDMK - TerrainDMK & 11.01 & 3.13 & 114 & 3.51 & 0.011 & *\\
+RadiometricTerrain - RadiometricTerrainDMK & 0.11 & 3.13 & 114 & 0.04 & 1.000 & \\
+RadiometricTerrain - Terrain & -1.29 & 3.13 & 114 & -0.41 & 1.000 & \\
+RadiometricTerrain - TerrainDMK & -0.77 & 3.13 & 114 & -0.25 & 1.000 & \\
+RadiometricTerrainDMK - Terrain & -1.40 & 3.13 & 114 & -0.45 & 0.999 & \\
+RadiometricTerrainDMK - TerrainDMK & -0.88 & 3.13 & 114 & -0.28 & 1.000 & \\
+Terrain - TerrainDMK & 0.52 & 3.13 & 114 & 0.17 & 1.000 & \\
+\hline
+\end{tabular}
+\end{table}
+\clearpage
+
+\begin{table}[h]
+
+\caption{\label{tab:pairwiseOrskogCCC}Pairwise comparisons of the concordance correlation coefficient among models for Ørskogfjellet, based on 10-fold cross-validation.
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package) on mixed-effects models (\texttt{lme4} package) to account for the cross-validation fold structure.}
+\centering
+\begin{tabular}[t]{lrrrrrl}
+\hline
+Comparison & Estimate & Std.error & Df & Statistic & Adj.p.value & Significance\\
+\hline
+DMK - Radiometric & 0.05 & 0.08 & 54 & 0.61 & 0.996 & \\
+DMK - RadiometricDMK & 0.07 & 0.08 & 54 & 0.84 & 0.979 & \\
+DMK - RadiometricTerrain & -0.14 & 0.08 & 54 & -1.79 & 0.559 & \\
+DMK - RadiometricTerrainDMK & -0.16 & 0.08 & 54 & -1.95 & 0.458 & \\
+DMK - Terrain & -0.19 & 0.08 & 54 & -2.33 & 0.248 & \\
+DMK - TerrainDMK & -0.22 & 0.08 & 54 & -2.67 & 0.127 & \\
+Radiometric - RadiometricDMK & 0.02 & 0.08 & 54 & 0.24 & 1.000 & \\
+Radiometric - RadiometricTerrain & -0.19 & 0.08 & 54 & -2.40 & 0.220 & \\
+Radiometric - RadiometricTerrainDMK & -0.21 & 0.08 & 54 & -2.55 & 0.161 & \\
+Radiometric - Terrain & -0.24 & 0.08 & 54 & -2.94 & 0.068 & \\
+Radiometric - TerrainDMK & -0.26 & 0.08 & 54 & -3.27 & 0.029 & *\\
+RadiometricDMK - RadiometricTerrain & -0.21 & 0.08 & 54 & -2.63 & 0.136 & \\
+RadiometricDMK - RadiometricTerrainDMK & -0.23 & 0.08 & 54 & -2.79 & 0.096 & \\
+RadiometricDMK - Terrain & -0.26 & 0.08 & 54 & -3.17 & 0.037 & *\\
+RadiometricDMK - TerrainDMK & -0.28 & 0.08 & 54 & -3.51 & 0.015 & *\\
+RadiometricTerrain - RadiometricTerrainDMK & -0.01 & 0.08 & 54 & -0.16 & 1.000 & \\
+RadiometricTerrain - Terrain & -0.04 & 0.08 & 54 & -0.54 & 0.998 & \\
+RadiometricTerrain - TerrainDMK & -0.07 & 0.08 & 54 & -0.88 & 0.975 & \\
+RadiometricTerrainDMK - Terrain & -0.03 & 0.08 & 54 & -0.38 & 1.000 & \\
+RadiometricTerrainDMK - TerrainDMK & -0.06 & 0.08 & 54 & -0.72 & 0.991 & \\
+Terrain - TerrainDMK & -0.03 & 0.08 & 54 & -0.33 & 1.000 & \\
+\hline
+\end{tabular}
+\end{table}
+\clearpage
+
+\begin{table}[h]
+
+\caption{\label{tab:pairwiseOrskogRsq}Pairwise comparisons of $R^2$ among models for Ørskogfjellet, based on 10-fold cross-validation.
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package) on mixed-effects models (\texttt{lme4} package) to account for the cross-validation fold structure.}
+\centering
+\begin{tabular}[t]{lrrrrll}
+\hline
+Comparison & Estimate & Std.error & Df & Statistic & Adj.p.value & Significance\\
+\hline
+DMK - Radiometric & 0.08 & 0.06 & 54 & 1.36 & 0.818 & \\
+DMK - RadiometricDMK & 0.08 & 0.06 & 54 & 1.35 & 0.823 & \\
+DMK - RadiometricTerrain & -0.10 & 0.06 & 54 & -1.72 & 0.607 & \\
+DMK - RadiometricTerrainDMK & -0.11 & 0.06 & 54 & -1.91 & 0.485 & \\
+DMK - Terrain & -0.15 & 0.06 & 54 & -2.62 & 0.14 & \\
+DMK - TerrainDMK & -0.19 & 0.06 & 54 & -3.29 & 0.028 & *\\
+Radiometric - RadiometricDMK & 0.00 & 0.06 & 54 & -0.01 & 1 & \\
+Radiometric - RadiometricTerrain & -0.18 & 0.06 & 54 & -3.08 & 0.048 & *\\
+Radiometric - RadiometricTerrainDMK & -0.19 & 0.06 & 54 & -3.27 & 0.029 & *\\
+Radiometric - Terrain & -0.24 & 0.06 & 54 & -3.98 & 0.004 & **\\
+Radiometric - TerrainDMK & -0.28 & 0.06 & 54 & -4.65 & < 0.001 & ***\\
+RadiometricDMK - RadiometricTerrain & -0.18 & 0.06 & 54 & -3.07 & 0.049 & *\\
+RadiometricDMK - RadiometricTerrainDMK & -0.19 & 0.06 & 54 & -3.26 & 0.03 & *\\
+RadiometricDMK - Terrain & -0.24 & 0.06 & 54 & -3.97 & 0.004 & **\\
+RadiometricDMK - TerrainDMK & -0.27 & 0.06 & 54 & -4.64 & < 0.001 & ***\\
+RadiometricTerrain - RadiometricTerrainDMK & -0.01 & 0.06 & 54 & -0.19 & 1 & \\
+RadiometricTerrain - Terrain & -0.05 & 0.06 & 54 & -0.90 & 0.971 & \\
+RadiometricTerrain - TerrainDMK & -0.09 & 0.06 & 54 & -1.57 & 0.702 & \\
+RadiometricTerrainDMK - Terrain & -0.04 & 0.06 & 54 & -0.71 & 0.991 & \\
+RadiometricTerrainDMK - TerrainDMK & -0.08 & 0.06 & 54 & -1.38 & 0.809 & \\
+Terrain - TerrainDMK & -0.04 & 0.06 & 54 & -0.67 & 0.994 & \\
+\hline
+\end{tabular}
+\end{table}
+\clearpage
+
+\begin{table}[h]
+
+\caption{\label{tab:pairwiseOrskogMAE}Pairwise comparisons of mean absolute error among models for Ørskogfjellet, based on 10-fold cross-validation.
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package) on mixed-effects models (\texttt{lme4} package) to account for the cross-validation fold structure.}
+\centering
+\begin{tabular}[t]{lrrrrrl}
+\hline
+Comparison & Estimate & Std.error & Df & Statistic & Adj.p.value & Significance\\
+\hline
+DMK - Radiometric & -6.93 & 8.69 & 54 & -0.80 & 0.984 & \\
+DMK - RadiometricDMK & -7.63 & 8.69 & 54 & -0.88 & 0.974 & \\
+DMK - RadiometricTerrain & 14.31 & 8.69 & 54 & 1.65 & 0.653 & \\
+DMK - RadiometricTerrainDMK & 15.83 & 8.69 & 54 & 1.82 & 0.539 & \\
+DMK - Terrain & 17.71 & 8.69 & 54 & 2.04 & 0.403 & \\
+DMK - TerrainDMK & 20.55 & 8.69 & 54 & 2.37 & 0.233 & \\
+Radiometric - RadiometricDMK & -0.70 & 8.69 & 54 & -0.08 & 1.000 & \\
+Radiometric - RadiometricTerrain & 21.24 & 8.69 & 54 & 2.45 & 0.200 & \\
+Radiometric - RadiometricTerrainDMK & 22.76 & 8.69 & 54 & 2.62 & 0.140 & \\
+Radiometric - Terrain & 24.64 & 8.69 & 54 & 2.84 & 0.086 & \\
+Radiometric - TerrainDMK & 27.48 & 8.69 & 54 & 3.16 & 0.039 & *\\
+RadiometricDMK - RadiometricTerrain & 21.95 & 8.69 & 54 & 2.53 & 0.170 & \\
+RadiometricDMK - RadiometricTerrainDMK & 23.47 & 8.69 & 54 & 2.70 & 0.118 & \\
+RadiometricDMK - Terrain & 25.35 & 8.69 & 54 & 2.92 & 0.071 & \\
+RadiometricDMK - TerrainDMK & 28.19 & 8.69 & 54 & 3.24 & 0.031 & *\\
+RadiometricTerrain - RadiometricTerrainDMK & 1.52 & 8.69 & 54 & 0.17 & 1.000 & \\
+RadiometricTerrain - Terrain & 3.40 & 8.69 & 54 & 0.39 & 1.000 & \\
+RadiometricTerrain - TerrainDMK & 6.24 & 8.69 & 54 & 0.72 & 0.991 & \\
+RadiometricTerrainDMK - Terrain & 1.88 & 8.69 & 54 & 0.22 & 1.000 & \\
+RadiometricTerrainDMK - TerrainDMK & 4.72 & 8.69 & 54 & 0.54 & 0.998 & \\
+Terrain - TerrainDMK & 2.84 & 8.69 & 54 & 0.33 & 1.000 & \\
+\hline
+\end{tabular}
+\end{table}
+\clearpage
+
+\begin{figure}[ht]
 \centering
@@
 \end{figure}
-\noappendix
+\clearpage
+
+\codedataavailability{\emph{Code and data availability.} R code used in this study is available at \url{https://github.com/julienvollering/DSMdepth}.
+The depth measurements we used are archived at \doi{10.6073/pasta/6ce440152f693f2156bf5b692a2e7917} and follow data and metadata standards.} %% use this section when having data sets and software code available
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% optional
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2025-05-27 ms/submission/AuthorComments-egusphere-2025-1046.qmd
address RC2.3: add configurations and configuration comparisons

@@
 
+Thank you for this suggestion.
+To clarify: the minimal marginal improvement in the terrain-plus-radiometric models compared to the terrain-only models is in fact evidence for the terrain variables being stronger predictors.
+Nevertheless, you are right that adding a radiometric-only model would provide a clearer comparison of the predictive contribution of radiometric versus terrain data.
+We now present models with all seven possible combinations of the three variable groups.
+These expanded results make clearer the predictive gap that we highlight in our title.
+
+We have also added statistical tests to assess the significance of the differences in model performance.
+The results of these tests are now presented in tables in the appendix.
+The results show that the differences in model performance are statistically significant in most cases, which supports our claims about the relative importance of terrain variables in predicting peat depth.
+Specifically, 8 of the 12 direct comparisons between radiometric and terrain variables (Radiometric--Terrain and RadiometricDMK--TerrainDMK configuration pairs * 3 metrics * 2 sites) are statistically significant at the 0.05 level.
+We now describe these results in the Results section, to better support the claim in the title.
+
+Some of the most important text revisions related to this comment are:
+
+::: {.callout appearance="simple"}
+
+**In _Methods_**
+
+[We formally assessed the effect of predictor configuration on performance metrics using mixed-effects models to account for the cross-validation fold structure, and testing pairwise differences between configurations.]{.added}
+
+**In _Results_**
+
+[Compared to the *terrain* configuration, the *terrain + radiometric* configuration improved concordance correlation by 0.04, R^2^ by 0.04, and mean absolute error by 1 cm.]{.deleted}
+
+[[For Skrimfjella] The *terrain* configuration outperformed the *radiometric* configuration by 0.27 in concordance correlation coefficient ($p < 0.001$; Appendix Table \@ref(tab:pairwiseSkrimCCC)), by 0.16 in R^2^ ($p = 0.190$; Appendix Table \@ref(tab:pairwiseSkrimRsq)), and by 10 cm in mean absolute error ($p = 0.038$; Appendix Table \@ref(tab:pairwiseSkrimMAE)).
+]{.added}
+
+[[For Ørskogfjellet] The *terrain* configuration outperformed the *radiometric* configuration by 0.24 in concordance correlation coefficient ($p = 0.068$; Appendix Table \@ref(tab:pairwiseOrskogCCC)), by 0.24 in R^2^ ($p = 0.004$; Appendix Table \@ref(tab:pairwiseOrskogRsq)), and by 25 cm in mean absolute error ($p = 0.086$; Appendix Table \@ref(tab:pairwiseOrskogMAE)).]{.added}
+
+**In _Discussion_**
+
+[Radiometric data had no predictive value at Ørskogfjellet, while at Skrimfjella they had minor influence in a relatively weak model.]{.deleted}
+
+[Radiometric data had little to no predictive value at either site (poor performance of *radiometric* configuration), although they did contribute to the best model at Skrimfjella (marginal improvement in *all predictors* configuration compared to *terrain + DMK*).
+These results contrast with earlier studies that found that radiometrics were useful predictors of peat depth [@keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @kogantiMappingPeatDepth2023].]{.added}
+
+:::
+
 ### Comment 2.4

### 2025-05-27 R/Skrim/modelDepth.R

@@ -59,13 +59,31 @@ mod_rf <-
 
 evaluation_metrics <- metric_set(rmse, mae, rsq, ccc)
 
-## Within mires ####
+## preds: radiometric ####
 
-### preds: radiometric + terrain ####
-
-recipe_RT <- 
+recipe_R <- 
   recipe(formula = depth_cm ~ 
-           radK + radTh + radU + radTC +
+           radK + radTh + radU + radTC,
+         data = frame)
+
+workflow_R <- 
+  workflow() %>% 
+  add_model(mod_rf) %>% 
+  add_recipe(recipe_R)
+
+set.seed(456)
+fit_R_knndm <- 
+  workflow_R %>% 
+  fit_resamples(
+    resamples = folds,
+    metrics = evaluation_metrics,
+    control = control_resamples(save_pred = TRUE))
+collect_metrics(fit_R_knndm)
+
+## preds: terrain ####
+
+recipe_T <- 
+  recipe(formula = depth_cm ~ 
            elevation + 
            slope1m + TPI1m + TRI1m + roughness1m + 
            slope10m + TPI10m + TRI10m + roughness10m + 
@@ -74,21 +92,21 @@ recipe_RT <-
            DTW2500 + DTW5000 + DTW10000 + DTW20000 + DTW40000 + DTW80000 + DTW160000,
          data = frame)
 
-workflow_RT <- 
+workflow_T <- 
   workflow() %>% 
   add_model(mod_rf) %>% 
-  add_recipe(recipe_RT)
+  add_recipe(recipe_T)
 
 set.seed(456)
-fit_RT_knndm <- 
-  workflow_RT %>% 
+fit_T_knndm <- 
+  workflow_T %>% 
   fit_resamples(
     resamples = folds,
     metrics = evaluation_metrics,
     control = control_resamples(save_pred = TRUE))
-collect_metrics(fit_RT_knndm)
+collect_metrics(fit_T_knndm)
 
-### preds: DMK ####
+## preds: DMK ####
 
 recipe_D <- 
   recipe(formula = depth_cm ~ dmkdepth, data = frame) |> 
@@ -116,10 +134,11 @@ frame |>
   group_by(dmkdepth) |> 
   summarize(n = n(), depth_cm = mean(depth_cm)) # sanity check
 
-### preds: terrain ####
+## preds: radiometric + terrain ####
 
-recipe_T <- 
+recipe_RT <- 
   recipe(formula = depth_cm ~ 
+           radK + radTh + radU + radTC +
[…]

### 2025-05-27 R/modelDepth.R

@@ -59,13 +59,31 @@ mod_rf <-
 
 evaluation_metrics <- metric_set(rmse, mae, rsq, ccc)
 
-## Within mires ####
+## preds: radiometric ####
 
-### preds: radiometric + terrain ####
-
-recipe_RT <- 
+recipe_R <- 
   recipe(formula = depth_cm ~ 
-           radK + radTh + radU + radTC +
+           radK + radTh + radU + radTC,
+         data = frame)
+
+workflow_R <- 
+  workflow() %>% 
+  add_model(mod_rf) %>% 
+  add_recipe(recipe_R)
+
+set.seed(456)
+fit_R_knndm <- 
+  workflow_R %>% 
+  fit_resamples(
+    resamples = folds,
+    metrics = evaluation_metrics,
+    control = control_resamples(save_pred = TRUE))
+collect_metrics(fit_R_knndm)
+
+## preds: terrain ####
+
+recipe_T <- 
+  recipe(formula = depth_cm ~ 
            elevation + 
            slope1m + TPI1m + TRI1m + roughness1m + 
            slope10m + TPI10m + TRI10m + roughness10m + 
@@ -74,21 +92,21 @@ recipe_RT <-
            DTW2500 + DTW5000 + DTW10000 + DTW20000 + DTW40000 + DTW80000 + DTW160000,
          data = frame)
 
-workflow_RT <- 
+workflow_T <- 
   workflow() %>% 
   add_model(mod_rf) %>% 
-  add_recipe(recipe_RT)
+  add_recipe(recipe_T)
 
 set.seed(456)
-fit_RT_knndm <- 
-  workflow_RT %>% 
+fit_T_knndm <- 
+  workflow_T %>% 
   fit_resamples(
     resamples = folds,
     metrics = evaluation_metrics,
     control = control_resamples(save_pred = TRUE))
-collect_metrics(fit_RT_knndm)
+collect_metrics(fit_T_knndm)
 
-### preds: DMK ####
+## preds: DMK ####
 
 recipe_D <- 
   recipe(formula = depth_cm ~ dmkdepth, data = frame) |> 
@@ -116,10 +134,11 @@ frame |>
   group_by(dmkdepth) |> 
   summarize(n = n(), depth_cm = mean(depth_cm)) # sanity check
 
-### preds: terrain ####
+## preds: radiometric + terrain ####
 
-recipe_T <- 
+recipe_RT <- 
   recipe(formula = depth_cm ~ 
+           radK + radTh + radU + radTC +
[…]

### 2025-05-27 ms/figures.R

@@ -265,13 +265,17 @@ plotting <- bind_rows(orskog = orskog, skrim = skrim, .id = 'site') %>%
   mutate(
     site = fct_relevel(site, "skrim"),
     model = case_when(
-      model == "DMK" ~ "DMK (2)",
+      model == "Radiometric" ~ "radiometric (4)",
       model == "Terrain" ~ "terrain (21)",
-      model == "TerrainDMK" ~ "terrain + DMK (23)",
+      model == "DMK" ~ "DMK (2)",
       model == "RadiometricTerrain" ~ "terrain + radiometric (25)",
+      model == "RadiometricDMK" ~ "radiometric + DMK (6)",
+      model == "TerrainDMK" ~ "terrain + DMK (23)",
       model == "RadiometricTerrainDMK" ~ "all predictors (27)"),
     model = fct_relevel(model,
                         "DMK (2)",
+                        "radiometric (4)",
+                        "radiometric + DMK (6)",
                         "terrain (21)",
                         "terrain + DMK (23)",
                         "terrain + radiometric (25)",
@@ -288,15 +292,15 @@ g1 <- plotting |>
                      xmin = mean - std_err, 
                      xmax = mean + std_err, 
                      color = site),
-                 position = position_dodge2(width= 0.3, reverse = TRUE)) +
+                 position = position_dodge2(width = 0.35, reverse = TRUE)) +
   geom_point(aes(y = model, x = mean, color = site),
-             position = position_dodge2(width= 0.3, reverse = TRUE)) +
+             position = position_dodge2(width = 0.35, reverse = TRUE)) +
   geom_text(aes(y = model, x = mean, label = signif(mean, 2), group = site),
-            position = position_dodge2(width= 1, reverse = TRUE),
+            position = position_dodge2(width = 1, reverse = TRUE),
             color = "grey50", size = 2.5) +
   scale_colour_manual(values = cols) +
-  annotate(geom='text', x=0.1, y=4.85, size=3, label='\u00D8rskogfjellet', color = "#d95f02") +
-  annotate(geom='text', x=0.1, y=5.15, size=3, label='Skrimfjella', color = "#1b9e77") +
+  annotate(geom='text', x=0.05, y=6.85, size=3, label='\u00D8rskogfjellet', color = "#d95f02") +
+  annotate(geom='text', x=0.05, y=7.15, size=3, label='Skrimfjella', color = "#1b9e77") +
   guides(color = "none") +
   labs(subtitle = "Concordance correlation") +
[…]

### 2025-05-27 ms/tables.R

@@ -45,6 +45,69 @@ bind_rows(overall, byAR5, byDMK) |>
          n_orskog, prop_orskog, depth_cm_orskog) |> 
   write_csv("ms/tables/celldepth-attribute.csv")
 
+# Compare predictive differences ####
+
+library(tidyverse)
+
+orskog <- read_csv("output/pairwise_comparisons.csv") 
+skrim <- read_csv("output/Skrim/pairwise_comparisons.csv") 
+
+orskog |> 
+  filter(.metric %in% c("rsq", "mae", "ccc")) |> 
+  mutate(
+    across(c(estimate,std.error,df,statistic), ~ round(.x, 2)),
+    adj.p.value = ifelse(adj.p.value < 0.001, "< 0.001", round(adj.p.value, 3)),
+    .metric = case_match(.metric, 
+                     "rsq" ~ "R-squared", 
+                     "mae" ~ "Mean absolute error", 
+                     "ccc" ~ "Concordance correlation"),
+    significance = case_when(
+      adj.p.value < 0.001 ~ "***",
+      adj.p.value < 0.01 ~ "**",
+      adj.p.value < 0.05 ~ "*",
+      TRUE ~ ""
+    )
+  ) |>
+  rename(metric = .metric, comparison = contrast) |>
+  rename_with(str_to_title) |>
+  nest_by(Metric) %>%
+  pwalk(~ write_csv(.y, paste0("ms/tables/pairwise-orskog-", .x, ".csv")))
+
+skrim |> 
+  filter(.metric %in% c("rsq", "mae", "ccc")) |> 
+  mutate(
+    across(c(estimate,std.error,df,statistic), ~ round(.x, 2)),
+    adj.p.value = ifelse(adj.p.value < 0.001, "< 0.001", round(adj.p.value, 3)),
+    .metric = case_match(.metric, 
+                         "rsq" ~ "R-squared", 
+                         "mae" ~ "Mean absolute error", 
+                         "ccc" ~ "Concordance correlation"),
+    significance = case_when(
+      adj.p.value < 0.001 ~ "***",
+      adj.p.value < 0.01 ~ "**",
+      adj.p.value < 0.05 ~ "*",
+      TRUE ~ ""
+    )
+  ) |>
+  rename(metric = .metric, comparison = contrast) |>
+  rename_with(str_to_title) |>
+  nest_by(Metric) %>%
[…]

### 2025-05-28 ms/ms.Rmd
address RC2.4: add contours to maps and explain Skrimfjella delineation

@@
 To fit the scale of the maps, the land cover shown here is from the AR50 national land resource dataset, which has simplified geometry with respect to the AR5 dataset that is used in the study.
-Terrain is visualized using a hillshade with slope 45° and azimuth 225°.
+Terrain is visualized with light orange contour lines at intervals of 100 m, and a hillshade with slope 45° and azimuth 225°.
 AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (https://data.norge.no/nlod/en/1.0) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/)."}
@@
 
-At Skrimfjella we delineated a study area of 34 km^2^ based on radiometric coverage and accessibility (Fig. \@ref(fig:sites)b).
+At Skrimfjella we delineated a study area of 34 km^2^ based on radiometric coverage (limiting to the west) and accessibility (limiting to the south), as part of a pilot project (Fig.
+\@ref(fig:sites)b).
+In Norway's AR5 national land cover dataset ["areal resources in scale 1:5000", @ahlstromAR5Klassifikasjonssystem2019], 1.5 km^2^ (4.5 %) of the study area is classified as 'mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.
+Relatively sparse peatland cover did not disqualify the area for our purposes, since we were also interested in peatland extent mapping in the pilot project. 
 The study area has a diverse bedrock, with 32 % alkali feldspar granite, 26 % marlstone, 10 % granite, 8 % monzonite, 7 % sandstone, 6 % limestone, and five other rock types with 1--5 % coverage (Geological Survey of Norway, 1:250 000 dataset).
@@
 The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).
-In Norway's AR5 national land cover dataset ["areal resources in scale 1:5000", @ahlstromAR5Klassifikasjonssystem2019], 1.5 km^2^ (4.5 %) of the study area is classified as 'mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.
 
 At Ørskogfjellet we defined a study area of 124 km^2^ which basically followed the southernmost part of the radiometric survey extent (Fig. \@ref(fig:sites)a).
+In the AR5 dataset, 15.3 km^2^ (12.4 %) of the study area is classified as mire.
 Bedrock in the area is 84 % granitic gneiss,  11 % granite, and 5 % aluminium silicate gneiss (Geological Survey of Norway, 1:250 000 dataset). 
@@
 Its mean elevation is 211 m above sea level (range 0--807 m, IQR 73--310 m), and its mean slope at 10 m resolution is 13.0° (IQR 4.7--18.3°).
-In the AR5 dataset, 15.3 km^2^ (12.4 %) of the study area is classified as mire.
 

### 2025-05-28 ms/ms.tex
address RC2.4: add contours to maps and explain Skrimfjella delineation

@@
 To fit the scale of the maps, the land cover shown here is from the AR50 national land resource dataset, which has simplified geometry with respect to the AR5 dataset that is used in the study.
-Terrain is visualized using a hillshade with slope 45° and azimuth 225°.
+Terrain is visualized with light orange contour lines at intervals of 100 m, and a hillshade with slope 45° and azimuth 225°.
 AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (https://data.norge.no/nlod/en/1.0) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/).}\label{fig:sites}
@@
 
-At Skrimfjella we delineated a study area of 34 km\textsuperscript{2} based on radiometric coverage and accessibility (Fig. \ref{fig:sites}b).
+At Skrimfjella we delineated a study area of 34 km\textsuperscript{2} based on radiometric coverage (limiting to the west) and accessibility (limiting to the south), as part of a pilot project (Fig.
+\ref{fig:sites}b).
+In Norway's AR5 national land cover dataset \citep[``areal resources in scale 1:5000'',][]{ahlstromAR5Klassifikasjonssystem2019}, 1.5 km\textsuperscript{2} (4.5 \%) of the study area is classified as `mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.
+Relatively sparse peatland cover did not disqualify the area for our purposes, since we were also interested in peatland extent mapping in the pilot project.
 The study area has a diverse bedrock, with 32 \% alkali feldspar granite, 26 \% marlstone, 10 \% granite, 8 \% monzonite, 7 \% sandstone, 6 \% limestone, and five other rock types with 1--5 \% coverage (Geological Survey of Norway, 1:250 000 dataset).
@@
 The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).
-In Norway's AR5 national land cover dataset \citep[``areal resources in scale 1:5000'',][]{ahlstromAR5Klassifikasjonssystem2019}, 1.5 km\textsuperscript{2} (4.5 \%) of the study area is classified as `mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.
 
 At Ørskogfjellet we defined a study area of 124 km\textsuperscript{2} which basically followed the southernmost part of the radiometric survey extent (Fig. \ref{fig:sites}a).
+In the AR5 dataset, 15.3 km\textsuperscript{2} (12.4 \%) of the study area is classified as mire.
 Bedrock in the area is 84 \% granitic gneiss, 11 \% granite, and 5 \% aluminium silicate gneiss (Geological Survey of Norway, 1:250 000 dataset).
@@
 Its mean elevation is 211 m above sea level (range 0--807 m, IQR 73--310 m), and its mean slope at 10 m resolution is 13.0° (IQR 4.7--18.3°).
-In the AR5 dataset, 15.3 km\textsuperscript{2} (12.4 \%) of the study area is classified as mire.
 

### 2025-05-28 ms/submission/AuthorComments-egusphere-2025-1046.qmd
address RC2.4: add contours to maps and explain Skrimfjella delineation

@@
 ::: {.callout appearance="simple"}
-
 **In _Methods_**
@@
 These results contrast with earlier studies that found that radiometrics were useful predictors of peat depth [@keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @kogantiMappingPeatDepth2023].]{.added}
-
 :::
@@
 Additionally, it would be helpful to explain the rationale for selecting Skrimfjella, which appears to have relatively limited peat coverage, especially given that the adjacent region seems to contain a larger peatland area.
+
+Thank you for the suggestion to make the relief of the study areas clearer.
+We have added contour lines the maps of the study areas.
+We have not added contour profiles between specific points since we think that they may not be as useful for readers as the contour lines, but we are happy to add profiles if the contour lines are not enough.
+
+We have also clarified the rationale for the delineation of the Skrimfjella study area, especially with respect to its limited peat coverage.
+In short: we were also investigating peatland extent mapping so low coverage was not disqualifying, and accessibility was also an important factor for this field work.
+
+::: {.callout appearance="simple"}
+**In _Materials and methods_**
+
+At Skrimfjella we delineated a study area of 34 km^2^ based on radiometric coverage [(limiting to the west)]{.added} and accessibility [(limiting to the south), as part of a pilot project]{.added} (Fig.
+\@ref(fig:sites)b).
+In Norway's AR5 national land cover dataset ["areal resources in scale 1:5000", @ahlstromAR5Klassifikasjonssystem2019], 1.5 km^2^ (4.5 %) of the study area is classified as 'mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.
+[Relatively sparse peatland cover did not disqualify the area for our purposes, since we were also interested in peatland extent mapping in the pilot project.]{.added}
+:::
 

### 2025-05-28 ms/figures.R

@@ -76,6 +76,19 @@ ar50orskogmask <- ar50orskog |>
 hill <- mask(hill, ar50orskogmask)
 hilldf_single <- as.data.frame(hill, xy = TRUE)
 
+# Creating contours
+elev_range <- dtm |> 
+  mask(ar50orskogmask) |> 
+  values() |> 
+  range(na.rm = TRUE)
+contourlines <- dtm |>
+  mask(ar50orskogmask) |>
+  stars::st_as_stars() |> 
+  stars::st_contour(contour_lines = TRUE,
+                    breaks = seq(from = floor(elev_range[1]), 
+                                 to = ceiling(elev_range[2]), 
+                                 by = 100))
+
 # https://colorbrewer2.org/?type=qualitative&scheme=Paired&n=6
 CBpaired.6class <- c('#e31a1c','#fb9a99','#33a02c','#b2df8a','#a6cee3','#1f78b4')
 
@@ -85,13 +98,15 @@ g2 <- ggplot(ar50orskog) +
   scale_fill_distiller(palette = "Greys") +
   new_scale_fill() +
   geom_sf(data = ar50orskog,
-          mapping = aes(fill = artype), alpha = 0.65, color = NA) +
+          mapping = aes(fill = artype), alpha = 0.75, color = NA) +
   scale_fill_discrete(labels = c("Built-up",
                                  "Agricultural",
                                  "Forest",
                                  "Open upland",
                                  "Peatland",
                                  "Freshwater"), type = CBpaired.6class) +
+  geom_sf(data = contourlines, linewidth = 0.2, color = "#fdbf6f",
+          show.legend = FALSE) +
   geom_sf(data = orskog,
           mapping = aes(color = 'Study area   '), fill = NA,
           linewidth = 1, linetype = 1) +
@@ -142,14 +157,29 @@ ar50skrimmask <- ar50skrim |>
 hill <- mask(hill, ar50skrimmask)
 hilldf_single <- as.data.frame(hill, xy = TRUE)
 
+# Creating contours
+elev_range <- dtm |> 
+  mask(ar50skrimmask) |> 
+  values() |> 
+  range(na.rm = TRUE)
+contourlines <- dtm |>
+  mask(ar50skrimmask) |>
+  stars::st_as_stars() |> 
+  stars::st_contour(contour_lines = TRUE,
+                    breaks = seq(from = floor(elev_range[1]), 
+                                 to = ceiling(elev_range[2]), 
[…]

### 2025-06-03 ms/ms.Rmd
address RC2.5: add peat depth distribution figure

@@
 
-To assess the predictive value of terrain and radiometric data for mapping peat depth, we chose two sites with conspicuously different physical geography: Skrimfjella (Fig.
-\@ref(fig:sites)b) in eastern Norway and Ørskogfjellet (Fig.
-\@ref(fig:sites)a) in western Norway (Fig.
-\@ref(fig:sites)c).
+To assess the predictive value of terrain and radiometric data for mapping peat depth, we chose two sites with conspicuously different physical geography: Skrimfjella in eastern Norway (Fig.
+\@ref(fig:map-sites)a) and Ørskogfjellet in western Norway (Fig.
+\@ref(fig:map-sites)b).
 These sites were chosen because they were covered by radiometric data from airborne surveys, had relatively little built-up area, and were accessible by road.
 
-```{r sites, out.height="81%", echo = FALSE, fig.cap = "Sites in southern Norway (a) with land cover at Skrimfjella (b) and Ørskogfjellet (c).
+```{r map-sites, out.height="81%", echo = FALSE, fig.cap = "Study areas with land cover at Skrimfjella (a) and Ørskogfjellet (b).
 To fit the scale of the maps, the land cover shown here is from the AR50 national land resource dataset, which has simplified geometry with respect to the AR5 dataset that is used in the study.
@@
 AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (https://data.norge.no/nlod/en/1.0) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/)."}
-knitr::include_graphics("figures/sites-patchwork.pdf")
+knitr::include_graphics("figures/map-sites.pdf")
 ```
@@
 At Skrimfjella we delineated a study area of 34 km^2^ based on radiometric coverage (limiting to the west) and accessibility (limiting to the south), as part of a pilot project (Fig.
-\@ref(fig:sites)b).
+\@ref(fig:map-sites)a).
 In Norway's AR5 national land cover dataset ["areal resources in scale 1:5000", @ahlstromAR5Klassifikasjonssystem2019], 1.5 km^2^ (4.5 %) of the study area is classified as 'mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.
@@
 
-At Ørskogfjellet we defined a study area of 124 km^2^ which basically followed the southernmost part of the radiometric survey extent (Fig. \@ref(fig:sites)a).
+At Ørskogfjellet we defined a study area of 124 km^2^ which basically followed the southernmost part of the radiometric survey extent (Fig. \@ref(fig:map-sites)b).
 In the AR5 dataset, 15.3 km^2^ (12.4 %) of the study area is classified as mire.
@@
 
-Our point measurements of peat depth (all sources) produced aggregated depths for 372 cells at Skrimfjella (2.4 % of AR5 mire area) and 1878 cells at Ørskogfjellet (1.2 % of AR5 mire area).
-Roughly 80 % of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
-Coverage of DMK peat depth classes was higher at Ørskogfjellet than at Skrimfjella (79 % versus 27 % of cells), and at Ørskogfjellet the *deep* and *shallow* classes showed a larger difference in measured depth.
-Overall mean peat depths were similar at Skrimfjella and Ørskogfjellet:  119 cm and 126 cm, respectively.
+Our point measurements of peat depth (all sources) produced aggregated depths for 372 cells or 2.4 % of AR5 mire area at Skrimfjella, and 1878 cells or 1.2 % of AR5 mire area at Ørskogfjellet (Table \ref{tab:depthsByClass}).
+Roughly 80 % of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland.
+DMK peat depth had higher coverage at Ørskogfjellet than at Skrimfjella (80 % versus 27 % of cells not *unknown*), and at Ørskogfjellet the *deep* and *shallow* classes showed a larger difference in measured depth.
+Overall, mean peat depths were similar at Skrimfjella and Ørskogfjellet (119 cm versus 126 cm) but the distribution was more right-skewed at Ørskogfjellet (Fig. \@ref(fig:map-distribution)).
 
@@
 cat(depthsByClass)
+```
+
+```{r map-distribution, out.height="81%", echo = FALSE, fig.cap = "Spatial and statistical distributions of mean peat depth in 10 m cells at Skrimfjella, N = 372 (a) and Ørskogfjellet, N = 1878 (b).
+Terrain is visualized with a hillshade with slope 45° and azimuth 225°.
+Terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/)."}
+knitr::include_graphics("figures/map-distribution.pdf")
 ```

### 2025-06-03 ms/ms.tex
address RC2.5: add peat depth distribution figure

@@
 
-To assess the predictive value of terrain and radiometric data for mapping peat depth, we chose two sites with conspicuously different physical geography: Skrimfjella (Fig.
-\ref{fig:sites}b) in eastern Norway and Ørskogfjellet (Fig.
-\ref{fig:sites}a) in western Norway (Fig.
-\ref{fig:sites}c).
+To assess the predictive value of terrain and radiometric data for mapping peat depth, we chose two sites with conspicuously different physical geography: Skrimfjella in eastern Norway (Fig.
+\ref{fig:map-sites}a) and Ørskogfjellet in western Norway (Fig.
+\ref{fig:map-sites}b).
 These sites were chosen because they were covered by radiometric data from airborne surveys, had relatively little built-up area, and were accessible by road.
@@
 \begin{figure}
-\includegraphics[height=0.81\textheight]{figures/sites-patchwork} \caption{Sites in southern Norway (a) with land cover at Skrimfjella (b) and Ørskogfjellet (c).
+\includegraphics[height=0.81\textheight]{figures/map-sites} \caption{Study areas with land cover at Skrimfjella (a) and Ørskogfjellet (b).
 To fit the scale of the maps, the land cover shown here is from the AR50 national land resource dataset, which has simplified geometry with respect to the AR5 dataset that is used in the study.
 Terrain is visualized with light orange contour lines at intervals of 100 m, and a hillshade with slope 45° and azimuth 225°.
-AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (https://data.norge.no/nlod/en/1.0) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/).}\label{fig:sites}
+AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (https://data.norge.no/nlod/en/1.0) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/).}\label{fig:map-sites}
 \end{figure}
@@
 At Skrimfjella we delineated a study area of 34 km\textsuperscript{2} based on radiometric coverage (limiting to the west) and accessibility (limiting to the south), as part of a pilot project (Fig.
-\ref{fig:sites}b).
+\ref{fig:map-sites}a).
 In Norway's AR5 national land cover dataset \citep[``areal resources in scale 1:5000'',][]{ahlstromAR5Klassifikasjonssystem2019}, 1.5 km\textsuperscript{2} (4.5 \%) of the study area is classified as `mire' -- defined as areas with mire vegetation and at least 30 cm of peat depth.
@@
 
-At Ørskogfjellet we defined a study area of 124 km\textsuperscript{2} which basically followed the southernmost part of the radiometric survey extent (Fig. \ref{fig:sites}a).
+At Ørskogfjellet we defined a study area of 124 km\textsuperscript{2} which basically followed the southernmost part of the radiometric survey extent (Fig. \ref{fig:map-sites}b).
 In the AR5 dataset, 15.3 km\textsuperscript{2} (12.4 \%) of the study area is classified as mire.
@@
 
-Our point measurements of peat depth (all sources) produced aggregated depths for 372 cells at Skrimfjella (2.4 \% of AR5 mire area) and 1878 cells at Ørskogfjellet (1.2 \% of AR5 mire area).
-Roughly 80 \% of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
-Coverage of DMK peat depth classes was higher at Ørskogfjellet than at Skrimfjella (79 \% versus 27 \% of cells), and at Ørskogfjellet the \emph{deep} and \emph{shallow} classes showed a larger difference in measured depth.
-Overall mean peat depths were similar at Skrimfjella and Ørskogfjellet: 119 cm and 126 cm, respectively.
+Our point measurements of peat depth (all sources) produced aggregated depths for 372 cells or 2.4 \% of AR5 mire area at Skrimfjella, and 1878 cells or 1.2 \% of AR5 mire area at Ørskogfjellet (Table \ref{tab:depthsByClass}).
+Roughly 80 \% of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland.
+DMK peat depth had higher coverage at Ørskogfjellet than at Skrimfjella (80 \% versus 27 \% of cells not \emph{unknown}), and at Ørskogfjellet the \emph{deep} and \emph{shallow} classes showed a larger difference in measured depth.
+Overall, mean peat depths were similar at Skrimfjella and Ørskogfjellet (119 cm versus 126 cm) but the distribution was more right-skewed at Ørskogfjellet (Fig. \ref{fig:map-distribution}).
 
@@
 \end{table}
+
+\begin{figure}
+\includegraphics[height=0.81\textheight]{figures/map-distribution} \caption{Spatial and statistical distributions of mean peat depth in 10 m cells at Skrimfjella, N = 372 (a) and Ørskogfjellet, N = 1878 (b).
+Terrain is visualized with a hillshade with slope 45° and azimuth 225°.
+Terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/).}\label{fig:map-distribution}
+\end{figure}
 

### 2025-06-03 ms/submission/AuthorComments-egusphere-2025-1046.qmd
address RC2.5: add peat depth distribution figure

@@
 
+Thank you for this suggestion.
+We have added a figure showing the statistical and spatial distribution of the peat depths in the Results section.
+We have taken care to use the same map format as in the figure that presents land cover in the study areas, which we hope will help readers see the spatial representativeness of the dataset.
+
 ### Comment 2.6

### 2025-06-03 ms/figures.R

@@ -1,5 +1,6 @@
-# Sites map ####
+# Sites map 2-panel ####
 
+library(rnaturalearth)
 library(sf)
 library(terra)
 library(tidyverse)
@@ -8,7 +9,7 @@ library(ggtext)
 library(ggrepel)
 library(ggspatial)
 library(ggnewscale)
-library(rnaturalearth)
+library(scales)
 
 ## Inset ####
 
@@ -29,7 +30,7 @@ labels <- both |>
   st_centroid() |> 
   mutate(site = c("\u00D8rskogfjellet", "Skrimfjella"))
 
-g1 <- ggplot() +
+pNorway <- ggplot() +
   geom_sf(data = nor50, fill = "white") + 
   geom_sf(data = norcities, color = "grey50") +
   geom_sf_text(data = norcities, aes(label = NAME), color = "grey50",
@@ -42,19 +43,26 @@ g1 <- ggplot() +
     aes(label = site, geometry = geom),
     stat = "sf_coordinates",
     min.segment.length = 0,
-    arrow = arrow(length = unit(0.02, "npc")),
+    arrow = arrow(length = unit(0.05, "npc")),
+    direction = "y",
     force_pull = 0.1,
-    nudge_y = c(-8e4, 12e4),
-    nudge_x = c(13e4, 4e4),
-    segment.curvature = 0.5) +
+    nudge_y = c(-10e4, 12e4),
+    nudge_x = c(3e4, 4e4),
+    segment.curvature = 0.2) +
   coord_sf(xlim = c(-6.1e4, 3.7e5), ylim = c(6.47e6, 7.05e6), crs = 25833) +
   theme(
     legend.position = "none",
     axis.title.x = element_blank(), axis.title.y= element_blank(),
-    axis.ticks = element_blank(), axis.text = element_blank()
-  )
+    axis.ticks = element_blank(), axis.text = element_blank(),
+    plot.background = element_blank(),
+    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
+    )
 
-## AR50 Ørskogfjellet ####
+## Ørskogfjellet ####
+
+orskog <- st_read("data/Orskogfjellet-site.gpkg", "mask_studyarea") |> 
+  st_transform(25833)
+bboxOrskog <- st_bbox(st_buffer(orskog, 500))
 
 ar50aalesund <- st_read("ms/figures/data/1508_25832_ar50_gdb/1507_25832_ar50_gdb.gdb")
[…]

### 2025-06-04 ms/ms.Rmd
address R2 Technical corrections - depths table

@@
 Because of its agricultural and silvicultural focus, the survey covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable [@ahlstromAR5Klassifikasjonssystem2019].
-Field surveyors carried a 1 m probe and they measured peat depth as a categorical variable: shallow (< 1 m), or deep (> 1 m). 
+Field surveyors carried a 1 m probe and they measured peat depth as a categorical variable: shallow (<1 m), or deep (>1 m). 
 These classes were assigned to whole polygons, so spatial resolution is on the order of hectares.
@@
 We prepared one categorical predictor -- peat depth class -- from the national map dataset called *DMK* [@ahlstromAR5Klassifikasjonssystem2019].
-The DMK dataset is derived from the same historical surveys as the AR5 dataset, and peat depth classes are: < 1 m (*shallow*), > 1 m (*deep*), and *unknown*.
+The DMK dataset is derived from the same historical surveys as the AR5 dataset, and peat depth classes are: <1 m (*shallow*), >1 m (*deep*), and *unknown*.
 Surveyors generally assigned peat depth classes to polygons of at least 0.5 ha, although delineating polygons down to 0.2 ha was allowed if peat depth showed a "particularly marked difference" [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
@@
 That is: we gave DMK and the DSM equal footing for a fair comparison.
-If we had taken the DMK peat depth classes at face value and assumed depths according to their class definitions (< 1 m, > 1 m), they would have performed worse and the advantage of the DSM would be greater.
+If we had taken the DMK peat depth classes at face value and assumed depths according to their class definitions (<1 m, >1 m), they would have performed worse and the advantage of the DSM would be greater.
 The advantage of the DSM was not large in absolute terms (9 cm and 21 cm improvements in mean absolute error), but it explained much more of the variation in depth (improvements in R^2^ of 0.16 and 0.19).

### 2025-06-04 ms/ms.tex
address R2 Technical corrections - depths table

@@
 Because of its agricultural and silvicultural focus, the survey covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable \citep{ahlstromAR5Klassifikasjonssystem2019}.
-Field surveyors carried a 1 m probe and they measured peat depth as a categorical variable: shallow (\textless{} 1 m), or deep (\textgreater{} 1 m).
+Field surveyors carried a 1 m probe and they measured peat depth as a categorical variable: shallow (\textless1 m), or deep (\textgreater1 m).
 These classes were assigned to whole polygons, so spatial resolution is on the order of hectares.
@@
 We prepared one categorical predictor -- peat depth class -- from the national map dataset called \emph{DMK} \citep{ahlstromAR5Klassifikasjonssystem2019}.
-The DMK dataset is derived from the same historical surveys as the AR5 dataset, and peat depth classes are: \textless{} 1 m (\emph{shallow}), \textgreater{} 1 m (\emph{deep}), and \emph{unknown}.
+The DMK dataset is derived from the same historical surveys as the AR5 dataset, and peat depth classes are: \textless1 m (\emph{shallow}), \textgreater1 m (\emph{deep}), and \emph{unknown}.
 Surveyors generally assigned peat depth classes to polygons of at least 0.5 ha, although delineating polygons down to 0.2 ha was allowed if peat depth showed a ``particularly marked difference'' \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
@@
 
-\begin{table}[tbp]
-\caption{Mean peat depth (cm) in \unit{10\,m} cells at Skrimfjella and Ørskogfjellet. The cells are also shown stratified by AR5 land class and DMK peat depth class.}
+\begin{table}[h]
+\caption{Selected attributes of the peat depth datasets at Skrimfjella and Ørskogfjellet.
+The first row summarizes all \unit{10\,m} cells, while subsequent rows stratify the dataset by AR5 land class and DMK peat depth class.}
 \begin{tabular}{llllllll}
 \hline
-                    &                            & \multicolumn{3}{c}{Skrimfjella} & \multicolumn{3}{c}{Ørskogfjellet} \\ \cline{3-8} 
-                    &                            & n        & \%      & depth      & n          & \%      & depth      \\ \hline
-all measured cells  &                            & 372      & 100     & 119        & 1878       & 100     & 126        \\
-AR5                 & Agricultural               &          &         &            & 21         & 1       & 30         \\
-                    & Forest                     & 52       & 14      & 65         & 272        & 15      & 71         \\
-                    & Open upland                & 19       & 5       & 98         & 134        & 7       & 39         \\
-                    & Peatland                   & 301      & 81      & 130        & 1451       & 77      & 145        \\
-DMK                 & deep (\textgreater 100 cm) & 94       & 25      & 100        & 659        & 35      & 219        \\
-                    & shallow (\textless 100 cm) & 6        & 2       & 50         & 838        & 45      & 82         \\
-                    & unknown                    & 272      & 73      & 127        & 381        & 20      & 60         \\ \hline
+    &                             & \multicolumn{3}{l}{Skrimfjella} & \multicolumn{3}{l}{Ørskogfjellet} \\ \cline{3-8} 
+    &                             & n     & \%   & mean (SD) depth  & n      & \%    & mean (SD) depth  \\ \cline{3-8} 
+all &                             & 372   & 100  & 119 (83) cm      & 1878   & 100   & 126 (119) cm     \\
+AR5 & Agricultural                &       &      &                  & 21     & 1     & 30 (8) cm        \\
+    & Forest                      & 52    & 14   & 65 (63) cm       & 272    & 14    & 71 (68) cm       \\
+    & Open upland                 & 19    & 5    & 98 (83) cm       & 134    & 7     & 39 (21) cm       \\
+    & Peatland                    & 301   & 81   & 130 (82) cm      & 1451   & 77    & 145 (125) cm     \\
+DMK & deep (\textgreater{}100 cm) & 94    & 25   & 100 (53) cm      & 659    & 35    & 219 (143) cm     \\
+    & shallow (\textless{}100 cm) & 6     & 2    & 50 (31) cm       & 838    & 45    & 82 (54) cm       \\
+    & unknown                     & 272   & 73   & 127 (90) cm      & 381    & 20    & 60 (62) cm       \\ \hline
 \end{tabular}
@@
 That is: we gave DMK and the DSM equal footing for a fair comparison.
-If we had taken the DMK peat depth classes at face value and assumed depths according to their class definitions (\textless{} 1 m, \textgreater{} 1 m), they would have performed worse and the advantage of the DSM would be greater.
+If we had taken the DMK peat depth classes at face value and assumed depths according to their class definitions (\textless1 m, \textgreater1 m), they would have performed worse and the advantage of the DSM would be greater.
 The advantage of the DSM was not large in absolute terms (9 cm and 21 cm improvements in mean absolute error), but it explained much more of the variation in depth (improvements in R\textsuperscript{2} of 0.16 and 0.19).

### 2025-06-04 ms/tables.R

@@ -18,31 +18,49 @@ orskog <- sf::st_read("output/modeling.gpkg", layer="dataframe") |>
   sf::st_drop_geometry()
 sites <- bind_rows(skrim = skrim, orskog = orskog, .id = "site") |> 
   mutate(dmkdepth = case_when(
-    dmkdepth == "djup myr" ~ "djup",
-    dmkdepth == "grunn myr" ~ "grunn",
-    TRUE ~ dmkdepth
+    grepl("djup", dmkdepth, ignore.case = TRUE) ~ "deep (>100 cm)",
+    grepl("grunn", dmkdepth, ignore.case = TRUE) ~ "shallow (<100 cm)",
+    TRUE ~ "unknown"
   )) |> 
-  filter(!(ar5cover %in% c("11", "12")))
+  filter(!(ar5cover %in% c("11", "12"))) |> 
+  mutate(ar5cover = case_when(
+    ar5cover == "21" ~ "Agricultural",
+    ar5cover == "30" ~ "Forest",
+    ar5cover == "50" ~ "Open upland",
+    ar5cover == "60" ~ "Peatland",
+    TRUE ~ "unknown"
+  ))
 overall <- sites |> 
-  group_by(site) |> 
-  summarize(n = n(), depth_cm = mean(depth_cm)) |> 
-  pivot_wider(names_from = 'site', values_from = c(n, depth_cm), 
+  group_by(site, TRUE) |> 
+  summarize(n = n(), depth_mean = mean(depth_cm), depth_sd = sd(depth_cm)) |> 
+  mutate(percent = (n / sum(n))*100) |>
+  pivot_wider(names_from = 'site', 
+              values_from = c(n, percent, depth_mean, depth_sd), 
               names_vary = "slowest")
 byAR5 <- sites |> 
   group_by(site, ar5cover) |> 
-  summarize(n = n(), depth_cm = mean(depth_cm)) |> 
-  mutate(prop = n / sum(n)) |>
-  pivot_wider(names_from = 'site', values_from = c(n, prop, depth_cm), 
+  summarize(n = n(), depth_mean = mean(depth_cm), depth_sd = sd(depth_cm)) |> 
+  mutate(percent = (n / sum(n))*100) |>
+  pivot_wider(names_from = 'site', 
+              values_from = c(n, percent, depth_mean, depth_sd), 
               names_vary = "slowest")
 byDMK <- sites |> 
   group_by(site, dmkdepth) |> 
[…]

### 2025-06-05 ms/ms.Rmd
address RC1.5: radiometric cps/concentration

@@
 At Ørskogfjellet an additional convolution filter was added to smooth the gridded data.
+
+Although radiometric data must be in units of counts per second to model attenuation directly [@olearyDigitalSoilMapping2022], we used the radiometric data as provided to us: in units of concentration for potassium, thorium, and uranium (converted from counts per second by scalar calibration factors).
+The monotonic transformation between counts per second and concentration does not affect the tree-based machine learning algorithm that we used to model peat depth [@hastieElementsStatisticalLearning2009].
 

### 2025-06-05 ms/ms.tex
address RC1.5: radiometric cps/concentration

@@
 
+Although radiometric data must be in units of counts per second to model attenuation directly \citep{olearyDigitalSoilMapping2022}, we used the radiometric data as provided to us: in units of concentration for potassium, thorium, and uranium (converted from counts per second by scalar calibration factors).
+The monotonic transformation between counts per second and concentration does not affect the tree-based machine learning algorithm that we used to model peat depth \citep{hastieElementsStatisticalLearning2009}.
+
 \subsubsection{Terrain}

### 2025-06-05 ms/submission/AuthorComments-egusphere-2025-1046.qmd
address R2 Technical corrections - depths table
address RC1.5: radiometric cps/concentration

@@
 toc: true
+bibliography: ../ms.bib
 ---
@@
 I recommend either the authors convert these concentrations to counts per second, or remove all but the Total Count data from their analysis and consider take the next concern into account.
+
+Thank you to the referee for pointing out this concern and providing a thorough explanation of their reasoning. We appreciate the references, which we have consulted.
+
+After careful consideration, we disagree that the radiometrics variables we used are not valid for predicting peat depth.
+In particular, we assert that converting the concentrations we use to counts per second -- as the referee suggests -- would produce identical results.
+That is because the method we use to predict peat depth (Random Forest) is insensitive to any monotonic transformation of the predictors, like the scalar conversion between concentration and counts per second [see Table 10.1 on p.351 in @hastieElementsStatisticalLearning2009].
+The Random Forest algorithm is based on decision trees, which partition the predictor space into regions based on the values of the predictors.
+As a result, any monotonic transformation of the predictors will not change the partitioning of the predictor space, and therefore will not change the predictions made by the model [see Fig.
+9.2 on p.306 and Algorithm 15.1 on p.588 in @hastieElementsStatisticalLearning2009].
+Appendix B3 in @baranwalHelicopterborneMagneticElectromagnetic2013 and Appendix A2 in @ofstadHelicopterborneMagneticRadiometric2015 confirm that the concentration values we use in our models are simply the counts per second multiplied by a constant conversion factor.
+
+To try to avoid confusion on this point, we have added a short explanation in the Methods section that the distinction between counts per second and concentration is not relevant for our analysis and it does not affect the predictive power of the radiometric variables in our models.
+
+::: {.callout appearance="simple"}
+**In _Materials and methods_**
+
+[Although radiometric data must be in units of counts per second to model attenuation directly [@olearyDigitalSoilMapping2022], we used the radiometric data as provided to us: in units of concentration for potassium, thorium, and uranium (converted from counts per second by scalar calibration factors).
+The monotonic transformation between counts per second and concentration does not affect the tree-based machine learning algorithm that we used to model peat depth [@hastieElementsStatisticalLearning2009].]{.added}
+:::
 
@@
 
+We have moved the table after its first mention and adjusted the table caption to clarify that the percentages represent the proportion of the total number of 10 m cells across the stratifications.
+We have also added standard deviations to the mean depths, in accordance with comment 2.5.
+
 > Line 394-397: I don’t quite get the point of this section.

### 2025-06-09 ms/ms.Rmd
address RC1.6: statistical not physical modeling of radiometric attenuation

@@
 A complementary approach from soil science is digital soil mapping (DSM). 
-DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest. 
-This approach has grown in importance with the availability of remotely sensed data and the advancement of methods for identifying patterns, especially through machine learning [@minasnyDigitalMappingPeatlands2019; @wadouxMachineLearningDigital2020].
+DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest, through a statistical model. 
+This approach has grown in importance with the availability of remotely sensed data and the advancement of methods for identifying statistical patterns, especially through machine learning [@minasnyDigitalMappingPeatlands2019; @wadouxMachineLearningDigital2020].
 
@@
 These sites were chosen because they were covered by radiometric data from airborne surveys, had relatively little built-up area, and were accessible by road.  
+Ørskogfjellet was additionally chosen for its relatively uniform bedrock geology, which should improve the peat depth signal in radiometric data, all else being equal [@minasnyDigitalMappingPeatlands2019].
 
@@
 These results contrast with earlier studies that found that radiometrics were useful predictors of peat depth [@keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @kogantiMappingPeatDepth2023].
-The bedrock is more homogeneous at Ørskogfjellet than at Skrimfjella, so uneven radiogenesis is not a viable explanation for the differences between sites nor the poor performance in general [@beamishEnvironmentalRadioactivityUK2014; @reinhardtGammaraySpectrometryVersatile2019].
+The bedrock is more homogeneous at Ørskogfjellet than at Skrimfjella, so uneven radiogenesis is not a good explanation for the higher predictive value of radiometric variables at Skrimfjella [@beamishEnvironmentalRadioactivityUK2014; @reinhardtGammaraySpectrometryVersatile2019].
 All four radiometric variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value. 
@@
 
-Weather conditions varied during the Ørskogfjellet radiometric survey and affected its data [@ofstadHelicopterborneMagneticRadiometric2015]. 
-Thus, uneven snow cover and air moisture may also have masked the soil signal in these data.
+Another possible reason for the poor predictive value of the radiometric data could be that other physical parameters influencing the amount of intercepted radiation varied too much within sites.
+Initial source strength, soil moisture, bulk density, and porosity all affect the amount of radiation that reaches the detector [@beamishGammaRayAttenuation2013; @reinhardtGammaraySpectrometryVersatile2019].
+Therefore, variation in these parameters could have masked the relationship between peat depth and radiometric data.
+We chose the Ørskogfjellet site in part because it has a relatively homogeneous bedrock, which should reduce the variation in initial source strength.
+Soil moisture, bulk density, and porosity, however, are not easily measured across landscape scales and were assumed to be homogeneous.
+Uneven snow cover and air moisture during the Ørskogfjellet radiometric survey may also have masked the soil signal these data, as Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015] reports large variation in weather conditions.
+If maps of these other physical parameters at the time of the radiometric survey were available and included in the model, the predictive value of radiometric data might improve, but this is not a practical solution for digital soil mapping of peat depth.
 
@@
 In the partial dependence plot of uranium ground concentration at Skrimfjella, the expected negative relationship between depth and uranium ground concentration cannot be found by ignoring the left (highly attenuated) side of the distribution.
-About a quarter of the peats in our study were less than a meter deep, and full attenuation is unlikely for these [@beamishGammaRayAttenuation2013].
+More than a quarter of the peats in each our study areas were less than 60 cm deep, and full attenuation is unlikely for these [@beamishGammaRayAttenuation2013].
 

### 2025-06-09 ms/ms.tex
address RC1.6: statistical not physical modeling of radiometric attenuation

@@
 A complementary approach from soil science is digital soil mapping (DSM).
-DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest.
-This approach has grown in importance with the availability of remotely sensed data and the advancement of methods for identifying patterns, especially through machine learning \citep{minasnyDigitalMappingPeatlands2019, wadouxMachineLearningDigital2020}.
+DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest, through a statistical model.
+This approach has grown in importance with the availability of remotely sensed data and the advancement of methods for identifying statistical patterns, especially through machine learning \citep{minasnyDigitalMappingPeatlands2019, wadouxMachineLearningDigital2020}.
 
@@
 \ref{fig:map-sites}b).
-These sites were chosen because they were covered by radiometric data from airborne surveys, had relatively little built-up area, and were accessible by road.
+These sites were chosen because they were covered by radiometric data from airborne surveys, had relatively little built-up area, and were accessible by road.\\
+Ørskogfjellet was additionally chosen for its relatively uniform bedrock geology, which should improve the peat depth signal in radiometric data, all else being equal \citep{minasnyDigitalMappingPeatlands2019}.
 
@@
 These results contrast with earlier studies that found that radiometrics were useful predictors of peat depth \citep{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, kogantiMappingPeatDepth2023}.
-The bedrock is more homogeneous at Ørskogfjellet than at Skrimfjella, so uneven radiogenesis is not a viable explanation for the differences between sites nor the poor performance in general \citep{beamishEnvironmentalRadioactivityUK2014, reinhardtGammaraySpectrometryVersatile2019}.
+The bedrock is more homogeneous at Ørskogfjellet than at Skrimfjella, so uneven radiogenesis is not a good explanation for the higher predictive value of radiometric variables at Skrimfjella \citep{beamishEnvironmentalRadioactivityUK2014, reinhardtGammaraySpectrometryVersatile2019}.
 All four radiometric variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value.
@@
 
-Weather conditions varied during the Ørskogfjellet radiometric survey and affected its data \citep{ofstadHelicopterborneMagneticRadiometric2015}.
-Thus, uneven snow cover and air moisture may also have masked the soil signal in these data.
+Another possible reason for the poor predictive value of the radiometric data could be that other physical parameters influencing the amount of intercepted radiation varied too much within sites.
+Initial source strength, soil moisture, bulk density, and porosity all affect the amount of radiation that reaches the detector \citep{beamishGammaRayAttenuation2013, reinhardtGammaraySpectrometryVersatile2019}.
+Therefore, variation in these parameters could have masked the relationship between peat depth and radiometric data.
+We chose the Ørskogfjellet site in part because it has a relatively homogeneous bedrock, which should reduce the variation in initial source strength.
+Soil moisture, bulk density, and porosity, however, are not easily measured across landscape scales and were assumed to be homogeneous.
+Uneven snow cover and air moisture during the Ørskogfjellet radiometric survey may also have masked the soil signal these data, as Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015} reports large variation in weather conditions.
+If maps of these other physical parameters at the time of the radiometric survey were available and included in the model, the predictive value of radiometric data might improve, but this is not a practical solution for digital soil mapping of peat depth.
 
@@
 In the partial dependence plot of uranium ground concentration at Skrimfjella, the expected negative relationship between depth and uranium ground concentration cannot be found by ignoring the left (highly attenuated) side of the distribution.
-About a quarter of the peats in our study were less than a meter deep, and full attenuation is unlikely for these \citep{beamishGammaRayAttenuation2013}.
+More than a quarter of the peats in each our study areas were less than 60 cm deep, and full attenuation is unlikely for these \citep{beamishGammaRayAttenuation2013}.
 

### 2025-06-09 ms/submission/AuthorComments-egusphere-2025-1046.qmd
address RC1.6: statistical not physical modeling of radiometric attenuation

@@
 
+Again, we thank the referee for a clear and comprehensive explanation of their concern.
+
+We agree that the physical modelling of peat depth from radiometric data is an underdetermined problem, for the reasons that the referee states.
+However, our prediction of peat depth is not based on a physical model, but rather on a statistical model that learns the relationship between the predictors (radiometric and terrain variables) and the response variable (peat depth) from the data.
+A mechanistic relationship -- in this case the physical attenuation of radiation through peat -- is what leads us to expect a statistical relationship in the first place, but the model is otherwise agnostic to the nature and origin of that statistical relationship.
+That is also why we are able to treat Total Count data the same as the other radiometric variables, even though it can only indicate attenuation variability.
+Indeed, it is exactly attenuation variability across a site that we expect to be useful for predicting peat depth, and our methods allow the model to decide which of the radiometric variables are most useful and the functional form of the statistical relationship.
+
+It is important to note here that the radiometric and terrain predictors are on equal footing -- it is the strength of their statistical relationship with peat depth (including non-linear relationships and interactions with other predictors) that is compared in our analyses and title.
+Moreover, we assert that there is a clearer mechanistic explanation for a relationship between radiometric variables and peat depth than there is for terrain variables and peat depth.
+The referee outlines this physical basis for the radiometrics, whereas for terrain variables potential mechanisms are more tenuous, or at least less consistent.
+For example, terrain slope in peatlands may reflect the topography that underlies the peat layer (e.g.
+in blanket bogs) or the position along the microtope (e.g.
+the rand in raised bogs) -- neither of which is necessarily strongly related to the thickness of the peat layer.
+The fact that radiometric measurements integrate across some part of the soil profile is a reason to expect, *a priori*, that these variables might predict peat depth better than measurements that technically only reflect the land surface, like LiDAR and the terrain models derived from it [see for example section 7.2 in @minasnyDigitalMappingPeatlands2019; @reinhardtGammaraySpectrometryVersatile2019; @beamishDetailedMappingPeat2024].
+We do not think it is correct, therefore, to conclude that there was a bias towards terrain variables in our analysis, or that the results are a perfectly expected outcome.
+
+The referee's comment does expand the list of possible explanations for the poor performance of radiometric variables in our models, and we have added to our discussion based on the comment, as suggested.
+
+::: {.callout appearance="simple"}
+**In _Introduction_**
+
+A complementary approach from soil science is digital soil mapping (DSM). 
+DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest[, through a statistical model]{.added}. 
+This approach has grown in importance with the availability of remotely sensed data and the advancement of methods for identifying [statistical]{.added} patterns, especially through machine learning [@minasnyDigitalMappingPeatlands2019; @wadouxMachineLearningDigital2020].
+
+**In _Discussion_**
+
+[Weather conditions varied during the Ørskogfjellet radiometric survey and affected its data [@ofstadHelicopterborneMagneticRadiometric2015].
+Thus, uneven snow cover and air moisture may also have masked the soil signal in these data.]{.deleted}
+[Another possible reason for the poor predictive value of the radiometric data could be that other physical parameters influencing the amount of intercepted radiation varied too much within sites.
+Initial source strength, soil moisture, bulk density, and porosity all affect the amount of radiation that reaches the detector [@beamishGammaRayAttenuation2013; @reinhardtGammaraySpectrometryVersatile2019].
+Therefore, variation in these parameters could have masked the relationship between peat depth and radiometric data.
+We chose the Ørskogfjellet site in part because it has a relatively homogeneous bedrock, which should reduce the variation in initial source strength.
+Soil moisture, bulk density, and porosity, however, are not easily measured across landscape scales and were assumed to be homogeneous.
+Uneven snow cover and air moisture during the Ørskogfjellet radiometric survey may also have masked the soil signal these data, as Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015] reports large variation in weather conditions.
+If maps of these other physical parameters at the time of the radiometric survey were available and included in the model, the predictive value of radiometric data might improve, but this is not a practical solution for digital soil mapping of peat depth.
+]{.added}
+:::
+
 ### Comment 1.7
@@
 Thank you for this helpful suggestion.
-We initially thought it would be worth including the analysis on extent mapping---especially since our occurrence data at Skrimfjella is a good test set---but based on the comments of both referees we recognize that it distracts too much from the main focus.
+We initially thought it would be worth including the analysis on extent mapping -- especially since our occurrence data at Skrimfjella is a good test set -- but based on the comments of both referees, we recognize that it distracts too much from the main focus.
 We have removed all sections related to mapping peatland extent (including section 4.1.4), to shorten the manuscript and improve its readability.

### 2025-06-09 ms/ms.Rmd
address RC2.6: choice of downscaling method

@@
 [-@baranwalHelicopterborneMagneticElectromagnetic2013] and Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015].
-They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling, using the *terra* package (v1.7) in R.
+They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling. 
+Sensitivity analysis showed that Pearson correlations between cubic spline and bilinear resampling methods exceeded 0.99 for all radiometric variables, so we are confident that the choice of resampling method did not affect our results.
 

### 2025-06-09 ms/ms.tex
address RC2.6: choice of downscaling method

@@
 \citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013} and Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.
-They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling, using the \emph{terra} package (v1.7) in R.
+They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling.
+Sensitivity analysis showed that Pearson correlations between cubic spline and bilinear resampling methods exceeded 0.99 for all radiometric variables, so we are confident that the choice of resampling method did not affect our results.
 

### 2025-06-09 ms/submission/AuthorComments-egusphere-2025-1046.qmd
address RC2.6: choice of downscaling method

@@
 
+We appreciate this attention to the potential impact of radiometric resampling methodology on our results.
+To address this concern, we added to our code a sensitivity analysis comparing cubic spline and bilinear resampling methods for downscaling the 50 m radiometric data.
+Using field measurement locations from both study sites, we extracted radiometric values resampled by both methods and calculated correlations for each radiometric variable (K, Th, U, TC).
+The correlations between cubic spline and bilinear resampling methods were very high, with the lowest correlation being 0.995.
+This near-perfect correlation demonstrates that the choice of resampling method has negligible impact on the spatial representation of radiometric variability at our target resolution.
+
+The consistently high correlations across all radiometric variables indicate that our conclusions about radiometric predictive performance would remain unchanged regardless of resampling method.
+Since we now mention this fact in the _Materials and methods_, we do not raise this point specifically in the _Discussion_ -- in the interest of brevity.
+
+::: {.callout appearance="simple"}
+**In _Materials and methods_**
+
+They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling[, using the *terra* package (v1.7) in R]{.deleted}. 
+[Sensitivity analysis showed that Pearson correlations between cubic spline and bilinear resampling methods exceeded 0.99 for all radiometric variables, so we are confident that the choice of resampling method did not affect our results.]{.added}
+:::
+
 ### Comment 2.7

### 2025-06-09 R/Skrim/createPredictorStack.R

@@ -18,12 +18,51 @@ radTC <- rast("data/Skrim/NGU-2013-029/Kong_Rad_Area3_TC/Kong_Rad_Area3_TC.ERS")
 
 rad <- c(radK, radTh, radU, radTC) |> 
   crop(y = st_transform(st_buffer(sa, 1000), st_crs(radK)))
-rad10m <- rad |> 
+
+# Sensitivity analysis: Compare cubic spline vs bilinear resampling methods
+rad10m_cubic <- rad |> 
   project(y = "epsg:25833",  
           method = "cubicspline",
           res = 10,
           origin = origin(dtm1m))
+
+rad10m_bilinear <- rad |> 
+  project(y = "epsg:25833",  
+          method = "bilinear",
+          res = 10,
+          origin = origin(dtm1m))
+
+# Use cubic spline as default (original method)
+rad10m <- rad10m_cubic
 plot(rad10m)
+
+# Correlation analysis between cubic spline and bilinear resampling methods
+# Sample points for comparison (using field measurement locations)
+probe <- read_csv("data/Skrim/depth_all.csv")
+probe <- st_as_sf(probe, coords = c('X', 'Y'), crs = 25833)
+probecells <- extract(rad10m_cubic, probe, cells = TRUE, xy=TRUE, ID=FALSE) |> 
+  select(cell, x, y) |> 
+  distinct(cell, .keep_all = TRUE) |> 
+  st_as_sf(coords = c('x','y'), crs=crs(rad10m_cubic))
+
+# Extract values at measurement locations for both methods
+rad_cubic_values <- extract(rad10m_cubic, probecells, ID=FALSE)
+rad_bilinear_values <- extract(rad10m_bilinear, probecells, ID=FALSE)
+
+# Calculate correlations for each radiometric variable
+map2_dfr(rad_cubic_values, rad_bilinear_values, 
+         ~tibble(correlation = cor(.x, .y, use="complete.obs")),
+         .id = "variable") |> 
+  mutate(variable = names(rad10m_cubic))
+
+# Visual comparison
+par(mfrow=c(2,2))
+walk2(rad_cubic_values, rad_bilinear_values, 
+      ~plot(.x, .y, 
+            xlab="Cubic spline", ylab="Bilinear",
+            main=paste(names(rad_cubic_values)[which(map_lgl(rad_cubic_values, identical, .x))],
+                      "\nr =", round(cor(.x, .y, use="complete.obs"), 3))))
+par(mfrow=c(1,1))
 
 # Simple terrain ####
 

### 2025-06-09 R/createPredictorStack.R

@@ -3,28 +3,68 @@ library(sf)
 library(terra)
 
 # Radiometrics #### 
+# Load original 50m resolution radiometric data for resampling comparison
+radK50m <- rast("data/NGU-2015-015/Romsdalsfjorden_RAD_miclev_K.ers") 
+radTh50m <- rast("data/NGU-2015-015/Romsdalsfjorden_RAD_miclev_Th.ers") 
+radU50m <- rast("data/NGU-2015-015/Romsdalsfjorden_RAD_miclev_U.ers") 
+radTC50m <- rast("data/NGU-2015-015/Romsdalsfjorden_RAD_miclev_TC.ers") 
+
+rad50m <- c(radK50m, radTh50m, radU50m, radTC50m)
+rad50m[rad50m < 0] <- NA
+
+# Sensitivity analysis: Compare cubic spline vs bilinear resampling methods
+rad10m_cubic <- rad50m |> 
+  project(y = "epsg:25832",  
+          method = "cubicspline",
+          res = 10)
+
+rad10m_bilinear <- rad50m |> 
+  project(y = "epsg:25832",  
+          method = "bilinear",
+          res = 10)
+
+# Use pre-processed 10m data as default (original method)
 radK10m <- rast("data/RAD_miclev_K_10m.tif") 
 radK10m[radK10m < 0] <- NA
-plot(radK10m)
 
 radTh10m <- rast("data/RAD_miclev_Th_10m.tif") 
 radTh10m[radTh10m < 0] <- NA
-plot(radTh10m)
 
 radU10m <- rast("data/RAD_miclev_U_10m.tif") 
 radU10m[radU10m < 0] <- NA
-plot(radU10m)
 
 radTC10m <- rast("data/RAD_miclev_TC_10m.tif") 
 radTC10m[radTC10m < 0] <- NA
-plot(radTC10m)
 
+plot(radK10m)
+
+# Correlation analysis between cubic spline and bilinear resampling methods
+# Sample points for comparison (using field measurement locations)
 probe <- read_csv("data/depth/concatenatedRawProbe.csv")
 probe <- st_as_sf(probe, coords = c('utm32E', 'utm32N'), crs = 25832)
 probecells <- extract(radK10m, probe, cells = TRUE, xy=TRUE, ID=FALSE) |> 
   select(cell, x, y) |> 
   distinct(cell, .keep_all = TRUE) |> 
   st_as_sf(coords = c('x','y'), crs=crs(radK10m))
+
+# Extract values at measurement locations for both methods
[…]

### 2025-06-09 ms/ms.Rmd
address RC2.7: move most of Sampling Design and Depth Measurements sections to appendix

@@
   bookdown::pdf_book:
+    latex_engine: lualatex
     keep_md: true
@@
 Although we implemented the approach differently for Skrimfjella (in 2020) than for Ørskogfjellet (in 2023), the objective was the same.
+We drew 105 raster cells at Skrimfjella and 160 raster cells at Ørskogfjellet.
+A complete description of our sample selection is in Appendix \ref{sec:sample-selection}.
 
@@
 The point measurements (described below) were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
-
-### Skrimfjella
-
-At Skrimfjella, we used the *eSample* function in the *iSDM* R package (v1.0) to stratify our sample across elevation, slope, and potassium concentration.
-This function defines the environmental space as a two-dimensional convex hull around the PCA-ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest [@hattabUnifiedFrameworkModel2017].
-We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and *eSample* returned 105 raster cells.
-
-### Ørskogfjellet
-
-At Ørskogfjellet, we first determined a minimal sample size that would adequately capture the slope and radiometric properties (potassium, thorium, uranium, and total count) of the entire AR5 mire area [@sauretteDivergenceMetricsDetermining2023].
-Specifically, we identified an elbow point in a curve of similarity between sample and population [@maloneMethodsImproveUtility2019].
-For a sequence of sample sizes (50--500) [ten replicates each, drawn by conditioned latin hypercube sampling, @minasnyConditionedLatinHypercube2006; @roudierClhsPackageConditioned2011], we calculated the mean Kullback--Leibler divergence between sample and population distributions [@maloneMethodsImproveUtility2019; @sauretteDivergenceMetricsDetermining2023].
-Then we fitted an asymptotic regression of mean divergence on sample size, and found that the curve reached 95 % of the fitted asymptote at a sample size of 160.
-
-To choose 160 locations, we performed feature space coverage sampling, implemented using the *kmeans* function in base R and the *rdist* function in the the *fields* package (v14.2).
-Feature space coverage sampling chooses locations that are closest to cluster centers in standardized predictor space [@brusSamplingDigitalSoil2019].
-This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
-Feature space coverage sampling works best when all dimensions are important predictors of the outcome [@wadouxSamplingDesignOptimization2019], and we used the same five predictors that we used to choose sample size: slope and four radiometrics.
-
-We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
-Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
-To adjust, we first restricted the sampling population to AR5 mire areas that were within an arbitrary cost distance of publicly accessible roads.
-Cost distance was calculated using GRASS's *r.walk* function, with friction costs defined by AR5 land classes [@GRASSv8-2].
-After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
-Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 % (compared to the ideal sample), but distance in our sample was still only 46 % of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
 
@@
 Probed depths serve to calibrate GPR measurements when calibration by common midpoint survey is not possible -- as was the case with our fixed radar antennas -- so the methods are complementary.
-
-### Peat probing
-
-We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real time kinematic (Ørskogfjellet) global navigation satellite system (GNSS) receivers.
-We dampened the effect of outlying measurements by probing three times at each location [@parryEvaluatingApproachesEstimating2014], at the vertices of a centered triangle with 2 m (Skrimfjella) or 4.5 m (Ørskogfjellet) sides.
-We used changes in resistance to indicate the base of the peat column.
-Probe locations were adjusted up to 20 cm if the base of the peat column seemed to be blocked by an obvious artifact, like a buried rock.
-Where the peat column was deeper than the extendable probe could be manually inserted and extracted by a pair of operators, we recorded a right-censored result (one at Skrimfjella, five at Ørskogfjellet).
+A complete description of our peat depth measurements is in Appendix \ref{sec:depth-measurements}.
 
@@
 We georeferenced the map and digitized the borehole locations and depths.
-
-### Ground-penetrating radar
-
-We performed GPR surveys in three subjectively chosen peatlands at Skrimfjella and in areas with a high density of sampled raster cells at Ørskogfjellet. 
-We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with a GNSS-enabled control unit connected to a 500 MHz shielded antenna mounted in a plastic sledge (transmitter--receiver separation 0.18 m, trace frequency 10 Hz).
-For some transects at Ørskogfjellet we substituted a 100 MHz Malå rough terrain antenna (transmitter--receiver separation 2.2 m, trace frequency 5 Hz), because the lower frequency antenna gives greater penetration depth.
-In all cases the system was towed by a walking GPR operator.
-
-GPR traces were recorded along zigzag (Skrimfjella) or snaking (Ørskogfjellet) transects.
-At Ørskogfjellet, transects were predetermined to pass through the centers of sampled raster cells, and we marked these precisely with flags to guide the GPR operator.
-A GPR records the time taken for a radio wave to travel from the transmitter to a reflector and back to the receiver, and the velocity of the wave varies with properties of the peat column.
-Therefore, wave velocity has to be calibrated to convert travel time to peat depth, and we probed peat depth along the transects.
-
-We processed the GPR data with Reflex2DQuick (v.3.0; Skrimfjella) or REFLEXW (v.8.5; Ørskogfjellet) software (Sandmeier Scientific Software, Germany). 
-We applied a time-zero correction, a dewow filter, and a gain filter based on observed energy decay. 
-With Ørskogfjellet data, we also applied a bandpass filter and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennas.
-The latter is important for the rough terrain antenna, where the antenna separation is comparable to typical peat depths.
-From the processed radargrams, we picked travel times from strong reflectors that we interpreted as the base of the peat column.
-
-We used picks near probed depths to calibrate wave speed velocity -- separately for each site.
-Calibration data were created by matching marked trace locations to a corresponding depth probe (Skrimfjella), or by a spatial join that identified interpreted traces and depth probes within 2 m of each other (Ørskogfjellet).
-We had sufficient calibration point density to avoid bias in wave velocity as a major source of error [@rosaDeterminingNumberManual2009]: 46 calibration points along 3.5 km of interpretable traces at Skrimfjella, and 78 along 7.8 km at Ørskogfjellet.
-We fitted site-specific linear regressions of probed depth on picked travel time, with the intercept fixed at zero, to estimate wave velocities.
-Notwithstanding a few outlying points, our regressions showed good fits and the resulting velocities are within the range reported for peat [@parsekianUncertaintyPeatVolume2012]: \unit{0.0387\,m\,ns^{-1}}, $R^2 = 0.874$ at Skrimfjella and \unit{0.0427\,m\,ns^{-1}}, $R^2 = 0.946$ at Ørskogfjellet.
-Finally, we used these two wave velocities to convert the travel times of all picks to calibrated peat depths.
-In total, the GPR surveys produced 48579 point measurements of peat depth at Skrimfjella and 32653 at Ørskogfjellet.
 
@@
 \appendixtables
-\begin{table}[h]
+\begin{table}[ht]
 \caption{Attributes of the radiometric surveys, as reported in Baranwal et al., \citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013} and Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.}
@@
 & Skrimfjella & Ørskogfjellet              \\ \cline{2-3} 
-Survey period                            & 2008-2011   & December 2014–January 2015 \\
+Survey period                            & 2008--2011   & 2014.12-–2015.01 \\
 Average flight altitude (m)              & 75           & 80               \\
@@
 
-\appendixfigures
+\section{}
+\subsection{Peat depth sample selection} \label{sec:sample-selection}
+\subsubsection{Skrimfjella}
+
+At Skrimfjella, we used the *eSample* function in the *iSDM* R package (v1.0) to stratify our sample across elevation, slope, and potassium concentration.
+This function defines the environmental space as a two-dimensional convex hull around the PCA-ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest [@hattabUnifiedFrameworkModel2017].
+We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and *eSample* returned 105 raster cells.
+
+\subsubsection{Ørskogfjellet}
+
+At Ørskogfjellet, we first determined a minimal sample size that would adequately capture the slope and radiometric properties (potassium, thorium, uranium, and total count) of the entire AR5 mire area [@sauretteDivergenceMetricsDetermining2023].
+Specifically, we identified an elbow point in a curve of similarity between sample and population [@maloneMethodsImproveUtility2019].
+For a sequence of sample sizes (50--500) [ten replicates each, drawn by conditioned latin hypercube sampling, @minasnyConditionedLatinHypercube2006; @roudierClhsPackageConditioned2011], we calculated the mean Kullback--Leibler divergence between sample and population distributions [@maloneMethodsImproveUtility2019; @sauretteDivergenceMetricsDetermining2023].
+Then we fitted an asymptotic regression of mean divergence on sample size, and found that the curve reached 95 % of the fitted asymptote at a sample size of 160.
+
+To choose 160 locations, we performed feature space coverage sampling, implemented using the *kmeans* function in base R and the *rdist* function in the the *fields* package (v14.2).
+Feature space coverage sampling chooses locations that are closest to cluster centers in standardized predictor space [@brusSamplingDigitalSoil2019].
+This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
+Feature space coverage sampling works best when all dimensions are important predictors of the outcome [@wadouxSamplingDesignOptimization2019], and we used the same five predictors that we used to choose sample size: slope and four radiometrics.
+
+We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
+Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs [@wadouxSamplingDesignOptimization2019; @maComparisonConditionedLatin2020].
+To adjust, we first restricted the sampling population to AR5 mire areas that were within an arbitrary cost distance of publicly accessible roads.
+Cost distance was calculated using GRASS's *r.walk* function, with friction costs defined by AR5 land classes [@GRASSv8-2].
+After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
+Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 % (compared to the ideal sample), but distance in our sample was still only 46 % of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
+
+\subsection{Depth measurements} \label{sec:depth-measurements}
+
+\subsubsection{Peat probing}
+
+We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real time kinematic (Ørskogfjellet) global navigation satellite system (GNSS) receivers.
+We dampened the effect of outlying measurements by probing three times at each location [@parryEvaluatingApproachesEstimating2014], at the vertices of a centered triangle with 2 m (Skrimfjella) or 4.5 m (Ørskogfjellet) sides.
+We used changes in resistance to indicate the base of the peat column.
+Probe locations were adjusted up to 20 cm if the base of the peat column seemed to be blocked by an obvious artifact, like a buried rock.
+Where the peat column was deeper than the extendable probe could be manually inserted and extracted by a pair of operators, we recorded a right-censored result (one at Skrimfjella, five at Ørskogfjellet).
+
+\subsubsection{Ground-penetrating radar}
+
+We performed GPR surveys in three subjectively chosen peatlands at Skrimfjella and in areas with a high density of sampled raster cells at Ørskogfjellet. 
+We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with a GNSS-enabled control unit connected to a 500 MHz shielded antenna mounted in a plastic sledge (transmitter--receiver separation 0.18 m, trace frequency 10 Hz).
+For some transects at Ørskogfjellet we substituted a 100 MHz Malå rough terrain antenna (transmitter--receiver separation 2.2 m, trace frequency 5 Hz), because the lower frequency antenna gives greater penetration depth.
+In all cases the system was towed by a walking GPR operator.
+
+GPR traces were recorded along zigzag (Skrimfjella) or snaking (Ørskogfjellet) transects.
+At Ørskogfjellet, transects were predetermined to pass through the centers of sampled raster cells, and we marked these precisely with flags to guide the GPR operator.
+A GPR records the time taken for a radio wave to travel from the transmitter to a reflector and back to the receiver, and the velocity of the wave varies with properties of the peat column.
+Therefore, wave velocity has to be calibrated to convert travel time to peat depth, and we probed peat depth along the transects.
+
+We processed the GPR data with Reflex2DQuick (v.3.0; Skrimfjella) or REFLEXW (v.8.5; Ørskogfjellet) software (Sandmeier Scientific Software, Germany). 
+We applied a time-zero correction, a dewow filter, and a gain filter based on observed energy decay. 
+With Ørskogfjellet data, we also applied a bandpass filter and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennas.
+The latter is important for the rough terrain antenna, where the antenna separation is comparable to typical peat depths.
+From the processed radargrams, we picked travel times from strong reflectors that we interpreted as the base of the peat column.
+
+We used picks near probed depths to calibrate wave speed velocity -- separately for each site.
+Calibration data were created by matching marked trace locations to a corresponding depth probe (Skrimfjella), or by a spatial join that identified interpreted traces and depth probes within 2 m of each other (Ørskogfjellet).
+We had sufficient calibration point density to avoid bias in wave velocity as a major source of error [@rosaDeterminingNumberManual2009]: 46 calibration points along 3.5 km of interpretable traces at Skrimfjella, and 78 along 7.8 km at Ørskogfjellet.
+We fitted site-specific linear regressions of probed depth on picked travel time, with the intercept fixed at zero, to estimate wave velocities.
+Notwithstanding a few outlying points, our regressions showed good fits and the resulting velocities are within the range reported for peat [@parsekianUncertaintyPeatVolume2012]: \unit{0.0387\,m\,ns^{-1}}, $R^2 = 0.874$ at Skrimfjella and \unit{0.0427\,m\,ns^{-1}}, $R^2 = 0.946$ at Ørskogfjellet (Fig.
+\ref{fig:GPRwavevelocity}).
+Finally, we used these two wave velocities to convert the travel times of all picks to calibrated peat depths.
+In total, the GPR surveys produced 48579 point measurements of peat depth at Skrimfjella and 32653 at Ørskogfjellet.
+
 \begin{figure}[ht]
 \centering
-\includegraphics[width=0.8\textwidth]{figures/GPRwavevelocity.pdf}
+\includegraphics[height=0.7\textheight]{figures/GPRwavevelocity.pdf}
 \caption{\label{fig:GPRwavevelocity}Calibration of GPR wave velocity at Skrimfjella (a) and Ørskogfjellet (b) by regression of probed depth against wave travel time.}
+\end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics{figures/semivariograms.pdf}
+\caption{\label{fig:semivariograms}Empirical semivariograms of peat depth point measurements at Skrimfjella (a) and Ørskogfjellet (b).}
 \end{figure}
@@
 
-\newpage
 \section{}
@@
 
-\begin{figure}[ht]
-\centering
-\includegraphics{figures/semivariograms.pdf}
-\caption{\label{fig:semivariograms}Empirical semivariograms of peat depth point measurements at Skrimfjella (a) and Ørskogfjellet (b).}
-\end{figure}
-\clearpage
+\noappendix

### 2025-06-09 ms/ms.tex
address RC2.7: move most of Sampling Design and Depth Measurements sections to appendix

@@
 The point measurements (described below) were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
-
-\subsubsection{Skrimfjella}
-
-At Skrimfjella, we used the \emph{eSample} function in the \emph{iSDM} R package (v1.0) to stratify our sample across elevation, slope, and potassium concentration.
-This function defines the environmental space as a two-dimensional convex hull around the PCA-ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest \citep{hattabUnifiedFrameworkModel2017}.
-We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and \emph{eSample} returned 105 raster cells.
-
-\subsubsection{Ørskogfjellet}
-
-At Ørskogfjellet, we first determined a minimal sample size that would adequately capture the slope and radiometric properties (potassium, thorium, uranium, and total count) of the entire AR5 mire area \citep{sauretteDivergenceMetricsDetermining2023}.
-Specifically, we identified an elbow point in a curve of similarity between sample and population \citep{maloneMethodsImproveUtility2019}.
-For a sequence of sample sizes (50--500) \citep[ten replicates each, drawn by conditioned latin hypercube sampling,][]{minasnyConditionedLatinHypercube2006, roudierClhsPackageConditioned2011}, we calculated the mean Kullback--Leibler divergence between sample and population distributions \citep{maloneMethodsImproveUtility2019, sauretteDivergenceMetricsDetermining2023}.
-Then we fitted an asymptotic regression of mean divergence on sample size, and found that the curve reached 95 \% of the fitted asymptote at a sample size of 160.
-
-To choose 160 locations, we performed feature space coverage sampling, implemented using the \emph{kmeans} function in base R and the \emph{rdist} function in the the \emph{fields} package (v14.2).
-Feature space coverage sampling chooses locations that are closest to cluster centers in standardized predictor space \citep{brusSamplingDigitalSoil2019}.
-This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
-Feature space coverage sampling works best when all dimensions are important predictors of the outcome \citep{wadouxSamplingDesignOptimization2019}, and we used the same five predictors that we used to choose sample size: slope and four radiometrics.
-
-We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
-Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
-To adjust, we first restricted the sampling population to AR5 mire areas that were within an arbitrary cost distance of publicly accessible roads.
-Cost distance was calculated using GRASS's \emph{r.walk} function, with friction costs defined by AR5 land classes \citep{GRASSv8-2}.
-After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
-Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 \% (compared to the ideal sample), but distance in our sample was still only 46 \% of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
 
@@
 Probed depths serve to calibrate GPR measurements when calibration by common midpoint survey is not possible -- as was the case with our fixed radar antennas -- so the methods are complementary.
-
-\subsubsection{Peat probing}
-
-We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real time kinematic (Ørskogfjellet) global navigation satellite system (GNSS) receivers.
-We dampened the effect of outlying measurements by probing three times at each location \citep{parryEvaluatingApproachesEstimating2014}, at the vertices of a centered triangle with 2 m (Skrimfjella) or 4.5 m (Ørskogfjellet) sides.
-We used changes in resistance to indicate the base of the peat column.
-Probe locations were adjusted up to 20 cm if the base of the peat column seemed to be blocked by an obvious artifact, like a buried rock.
-Where the peat column was deeper than the extendable probe could be manually inserted and extracted by a pair of operators, we recorded a right-censored result (one at Skrimfjella, five at Ørskogfjellet).
+A complete description of our peat depth measurements is in Appendix \ref{sec:depth-measurements}.
 
@@
 We georeferenced the map and digitized the borehole locations and depths.
-
-\subsubsection{Ground-penetrating radar}
-
-We performed GPR surveys in three subjectively chosen peatlands at Skrimfjella and in areas with a high density of sampled raster cells at Ørskogfjellet.
-We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with a GNSS-enabled control unit connected to a 500 MHz shielded antenna mounted in a plastic sledge (transmitter--receiver separation 0.18 m, trace frequency 10 Hz).
-For some transects at Ørskogfjellet we substituted a 100 MHz Malå rough terrain antenna (transmitter--receiver separation 2.2 m, trace frequency 5 Hz), because the lower frequency antenna gives greater penetration depth.
-In all cases the system was towed by a walking GPR operator.
-
-GPR traces were recorded along zigzag (Skrimfjella) or snaking (Ørskogfjellet) transects.
-At Ørskogfjellet, transects were predetermined to pass through the centers of sampled raster cells, and we marked these precisely with flags to guide the GPR operator.
-A GPR records the time taken for a radio wave to travel from the transmitter to a reflector and back to the receiver, and the velocity of the wave varies with properties of the peat column.
-Therefore, wave velocity has to be calibrated to convert travel time to peat depth, and we probed peat depth along the transects.
-
-We processed the GPR data with Reflex2DQuick (v.3.0; Skrimfjella) or REFLEXW (v.8.5; Ørskogfjellet) software (Sandmeier Scientific Software, Germany).
-We applied a time-zero correction, a dewow filter, and a gain filter based on observed energy decay.
-With Ørskogfjellet data, we also applied a bandpass filter and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennas.
-The latter is important for the rough terrain antenna, where the antenna separation is comparable to typical peat depths.
-From the processed radargrams, we picked travel times from strong reflectors that we interpreted as the base of the peat column.
-
-We used picks near probed depths to calibrate wave speed velocity -- separately for each site.
-Calibration data were created by matching marked trace locations to a corresponding depth probe (Skrimfjella), or by a spatial join that identified interpreted traces and depth probes within 2 m of each other (Ørskogfjellet).
-We had sufficient calibration point density to avoid bias in wave velocity as a major source of error \citep{rosaDeterminingNumberManual2009}: 46 calibration points along 3.5 km of interpretable traces at Skrimfjella, and 78 along 7.8 km at Ørskogfjellet.
-We fitted site-specific linear regressions of probed depth on picked travel time, with the intercept fixed at zero, to estimate wave velocities.
-Notwithstanding a few outlying points, our regressions showed good fits and the resulting velocities are within the range reported for peat \citep{parsekianUncertaintyPeatVolume2012}: \unit{0.0387\,m\,ns^{-1}}, \(R^2 = 0.874\) at Skrimfjella and \unit{0.0427\,m\,ns^{-1}}, \(R^2 = 0.946\) at Ørskogfjellet.
-Finally, we used these two wave velocities to convert the travel times of all picks to calibrated peat depths.
-In total, the GPR surveys produced 48579 point measurements of peat depth at Skrimfjella and 32653 at Ørskogfjellet.
 
@@
 
-\begin{table}[h]
+\begin{table}[ht]
 \caption{Selected attributes of the peat depth datasets at Skrimfjella and Ørskogfjellet.
@@
 
-\appendixfigures
+\section{}
+\subsection{Peat depth sample selection} \label{sec:sample-selection}
+\subsubsection{Skrimfjella}
+
+At Skrimfjella, we used the \emph{eSample} function in the \emph{iSDM} R package (v1.0) to stratify our sample across elevation, slope, and potassium concentration.
+This function defines the environmental space as a two-dimensional convex hull around the PCA-ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest \citep{hattabUnifiedFrameworkModel2017}.
+We set a target sample size of 100, excluded the top and bottom percentile from the convex hull, and \emph{eSample} returned 105 raster cells.
+
+\subsubsection{Ørskogfjellet}
+
+At Ørskogfjellet, we first determined a minimal sample size that would adequately capture the slope and radiometric properties (potassium, thorium, uranium, and total count) of the entire AR5 mire area \citep{sauretteDivergenceMetricsDetermining2023}.
+Specifically, we identified an elbow point in a curve of similarity between sample and population \citep{maloneMethodsImproveUtility2019}.
+For a sequence of sample sizes (50--500) \citep[ten replicates each, drawn by conditioned latin hypercube sampling,][]{minasnyConditionedLatinHypercube2006, roudierClhsPackageConditioned2011}, we calculated the mean Kullback--Leibler divergence between sample and population distributions \citep{maloneMethodsImproveUtility2019, sauretteDivergenceMetricsDetermining2023}.
+Then we fitted an asymptotic regression of mean divergence on sample size, and found that the curve reached 95 \% of the fitted asymptote at a sample size of 160.
+
+To choose 160 locations, we performed feature space coverage sampling, implemented using the \emph{kmeans} function in base R and the \emph{rdist} function in the the \emph{fields} package (v14.2).
+Feature space coverage sampling chooses locations that are closest to cluster centers in standardized predictor space \citep{brusSamplingDigitalSoil2019}.
+This approach has been found to produce higher accuracy in RFs than conditioned latin hypercube sampling \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
+Feature space coverage sampling works best when all dimensions are important predictors of the outcome \citep{wadouxSamplingDesignOptimization2019}, and we used the same five predictors that we used to choose sample size: slope and four radiometrics.
+
+We adjusted the feature space coverage sampling to ensure that locations were accessible within time constraints, and assessed how this changed our sample from an ideal feature space coverage sample.
+Adjusting for accessibility is justified because the smaller sample size that would result if accessibility were ignored can degrade model accuracy as much or more as deviations from ideal sampling designs \citep{wadouxSamplingDesignOptimization2019, maComparisonConditionedLatin2020}.
+To adjust, we first restricted the sampling population to AR5 mire areas that were within an arbitrary cost distance of publicly accessible roads.
+Cost distance was calculated using GRASS's \emph{r.walk} function, with friction costs defined by AR5 land classes \citep{GRASSv8-2}.
+After creating a feature space coverage sample with this restriction, we also inspected a map of the sample and substituted 16 inaccessible locations with accessible locations from the same or a nearby cluster.
+Our two accessibility adjustments increased the distance in standardized predictor space between sample locations and cluster centers by 78 \% (compared to the ideal sample), but distance in our sample was still only 46 \% of the mean distance to cluster centers -- i.e., accessibility did not force locations far from cluster centers relative to the size of the clusters.
+
+\subsection{Depth measurements} \label{sec:depth-measurements}
+
+\subsubsection{Peat probing}
+
+We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real time kinematic (Ørskogfjellet) global navigation satellite system (GNSS) receivers.
+We dampened the effect of outlying measurements by probing three times at each location \citep{parryEvaluatingApproachesEstimating2014}, at the vertices of a centered triangle with 2 m (Skrimfjella) or 4.5 m (Ørskogfjellet) sides.
+We used changes in resistance to indicate the base of the peat column.
+Probe locations were adjusted up to 20 cm if the base of the peat column seemed to be blocked by an obvious artifact, like a buried rock.
+Where the peat column was deeper than the extendable probe could be manually inserted and extracted by a pair of operators, we recorded a right-censored result (one at Skrimfjella, five at Ørskogfjellet).
+
+\subsubsection{Ground-penetrating radar}
+
+We performed GPR surveys in three subjectively chosen peatlands at Skrimfjella and in areas with a high density of sampled raster cells at Ørskogfjellet.
+We used the Malå ProEx GPR system (Guideline Geo AB, Sweden) with a GNSS-enabled control unit connected to a 500 MHz shielded antenna mounted in a plastic sledge (transmitter--receiver separation 0.18 m, trace frequency 10 Hz).
+For some transects at Ørskogfjellet we substituted a 100 MHz Malå rough terrain antenna (transmitter--receiver separation 2.2 m, trace frequency 5 Hz), because the lower frequency antenna gives greater penetration depth.
+In all cases the system was towed by a walking GPR operator.
+
+GPR traces were recorded along zigzag (Skrimfjella) or snaking (Ørskogfjellet) transects.
+At Ørskogfjellet, transects were predetermined to pass through the centers of sampled raster cells, and we marked these precisely with flags to guide the GPR operator.
+A GPR records the time taken for a radio wave to travel from the transmitter to a reflector and back to the receiver, and the velocity of the wave varies with properties of the peat column.
+Therefore, wave velocity has to be calibrated to convert travel time to peat depth, and we probed peat depth along the transects.
+
+We processed the GPR data with Reflex2DQuick (v.3.0; Skrimfjella) or REFLEXW (v.8.5; Ørskogfjellet) software (Sandmeier Scientific Software, Germany).
+We applied a time-zero correction, a dewow filter, and a gain filter based on observed energy decay.
+With Ørskogfjellet data, we also applied a bandpass filter and a dynamic correction that accounts for the non-vertical wave path between offset transmitter and receiver antennas.
+The latter is important for the rough terrain antenna, where the antenna separation is comparable to typical peat depths.
+From the processed radargrams, we picked travel times from strong reflectors that we interpreted as the base of the peat column.
+
+We used picks near probed depths to calibrate wave speed velocity -- separately for each site.
+Calibration data were created by matching marked trace locations to a corresponding depth probe (Skrimfjella), or by a spatial join that identified interpreted traces and depth probes within 2 m of each other (Ørskogfjellet).
+We had sufficient calibration point density to avoid bias in wave velocity as a major source of error \citep{rosaDeterminingNumberManual2009}: 46 calibration points along 3.5 km of interpretable traces at Skrimfjella, and 78 along 7.8 km at Ørskogfjellet.
+We fitted site-specific linear regressions of probed depth on picked travel time, with the intercept fixed at zero, to estimate wave velocities.
+Notwithstanding a few outlying points, our regressions showed good fits and the resulting velocities are within the range reported for peat \citep{parsekianUncertaintyPeatVolume2012}: \unit{0.0387\,m\,ns^{-1}}, \(R^2 = 0.874\) at Skrimfjella and \unit{0.0427\,m\,ns^{-1}}, \(R^2 = 0.946\) at Ørskogfjellet (Fig.
+\ref{fig:GPRwavevelocity}).
+Finally, we used these two wave velocities to convert the travel times of all picks to calibrated peat depths.
+In total, the GPR surveys produced 48579 point measurements of peat depth at Skrimfjella and 32653 at Ørskogfjellet.
+
 \begin{figure}[ht]
 \centering
-\includegraphics[width=0.8\textwidth]{figures/GPRwavevelocity.pdf}
+\includegraphics[height=0.7\textheight]{figures/GPRwavevelocity.pdf}
 \caption{\label{fig:GPRwavevelocity}Calibration of GPR wave velocity at Skrimfjella (a) and Ørskogfjellet (b) by regression of probed depth against wave travel time.}
 \end{figure}
+
+\begin{figure}[ht]
+\centering
+\includegraphics{figures/semivariograms.pdf}
+\caption{\label{fig:semivariograms}Empirical semivariograms of peat depth point measurements at Skrimfjella (a) and Ørskogfjellet (b).}
+\end{figure}
 \clearpage
 
-\newpage
 \section{}
@@
 
-\begin{figure}[ht]
-\centering
-\includegraphics{figures/semivariograms.pdf}
-\caption{\label{fig:semivariograms}Empirical semivariograms of peat depth point measurements at Skrimfjella (a) and Ørskogfjellet (b).}
-\end{figure}
-\clearpage
+\noappendix
 

### 2025-06-09 ms/submission/AuthorComments-egusphere-2025-1046.qmd
address RC2.7: move most of Sampling Design and Depth Measurements sections to appendix

@@
 
+Thank you for this helpful suggestion.
+We agree that this an acceptable way to reduce the length of the main text and improve its readability, especially with the added figure showing the spatial distribution of depth measurements (in response to comment 2.5).
+We have moved the bulk of the sections on _Peat depth sample selection_ and _Depth measurements_ to the appendix, while retaining brief summaries of these in the main text.
+This solution seems to be allowed under the journal's guidelines for manuscript composition.
+
 ### Comment 2.8

### 2025-06-10 ms/ms.Rmd
address RC2.8: concordance between different VI methods

@@
 *FIRM* values measure the flatness of the partial dependence plot, *permutation* values measure the decrease in model performance when the predictor is permuted, and *Shapley* values are aggregated from local, game-theoretical measures of variable importance [@greenwellVariableImportancePlots2020].
+Since *FIRM* reflects the flatness of the partial dependence plot, it captures functional complexity rather than overall predictive impact.
 *Permutation* values were obtained from ten iterations, with root mean square error as the performance measure.
@@
 
+The *permutation* and *Shapley* methods of variable importance were highly correlated at both sites, while the *FIRM* method was less correlated with the other two (Fig. \@ref(fig:varImp)).
+
 ### Variable importance
@@
 At both sites, elevation and Multi-Resolution Valley Bottom Flatness were important predictors (Fig. \@ref(fig:varImp)).
-At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet elevation was more important than Multi-Resolution Valley Bottom Flatness.
+At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet elevation had higher predictive impact (*permutation* and *Shapley*) but lower functional complexity (*FIRM*) than Multi-Resolution Valley Bottom Flatness.
 DMK was also important -- the shallow class in particular -- but only at Ørskogfjellet.
@@
 
-```{r varImp, echo = FALSE, out.width="100%", fig.cap = "Global variable importance in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), as measured by three different metrics: Shapley values, permutation importance, and the Feature Importance Ranking Measure [@greenwellVariableImportancePlots2020].
-Variables removed due to collinearity are shown to the right of the variable with which they are most correlated."}
+```{r varImp, echo = FALSE, out.width="100%", fig.cap = "Global variable importance in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), as measured by three different metrics: Shapley values, permutation importance, and the Feature Importance Ranking Measure \\citep{greenwellVariableImportancePlots2020}.
+Variables removed due to collinearity are shown to the right of the variable with which they are most correlated.
+Spearman rank correlations between the three metrics are displayed in matrices."}
 knitr::include_graphics("figures/variable_importance.pdf")

### 2025-06-10 ms/ms.tex
address RC2.8: concordance between different VI methods

@@
 \emph{FIRM} values measure the flatness of the partial dependence plot, \emph{permutation} values measure the decrease in model performance when the predictor is permuted, and \emph{Shapley} values are aggregated from local, game-theoretical measures of variable importance \citep{greenwellVariableImportancePlots2020}.
+Since \emph{FIRM} reflects the flatness of the partial dependence plot, it captures functional complexity rather than overall predictive impact.
 \emph{Permutation} values were obtained from ten iterations, with root mean square error as the performance measure.
@@
 
+The \emph{permutation} and \emph{Shapley} methods of variable importance were highly correlated at both sites, while the \emph{FIRM} method was less correlated with the other two (Fig.
+\ref{fig:varImp}).
+
 \subsubsection{Variable importance}
@@
 At both sites, elevation and Multi-Resolution Valley Bottom Flatness were important predictors (Fig. \ref{fig:varImp}).
-At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet elevation was more important than Multi-Resolution Valley Bottom Flatness.
+At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet elevation had higher predictive impact (\emph{permutation} and \emph{Shapley}) but lower functional complexity (\emph{FIRM}) than Multi-Resolution Valley Bottom Flatness.
 DMK was also important -- the shallow class in particular -- but only at Ørskogfjellet.
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{figures/variable_importance} \caption{Global variable importance in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), as measured by three different metrics: Shapley values, permutation importance, and the Feature Importance Ranking Measure [@greenwellVariableImportancePlots2020].
-Variables removed due to collinearity are shown to the right of the variable with which they are most correlated.}\label{fig:varImp}
+\includegraphics[width=1\linewidth]{figures/variable_importance} \caption{Global variable importance in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), as measured by three different metrics: Shapley values, permutation importance, and the Feature Importance Ranking Measure \citep{greenwellVariableImportancePlots2020}.
+Variables removed due to collinearity are shown to the right of the variable with which they are most correlated.
+Spearman rank correlations between the three metrics are displayed in matrices.}\label{fig:varImp}
 \end{figure}

### 2025-06-10 ms/submission/AuthorComments-egusphere-2025-1046.qmd
address RC2.8: concordance between different VI methods

@@
 
+Thank you for this suggestion.
+We have added text to the _Materials and methods_ and _Results_ sections to clarify how the three variable importance methods relate to each other.
+We have also added rank correlations to our figure on variable importance, to highlight the similarities and differences.
+In short, we find that Shapley and permutation importance are very similar for our application, while FIRM values capture a different aspect of variable importance.
+
+Since the different methods are broadly congruent and their differences do not affect our interpretation much -- and in the interest of brevity -- we do not go into this point in our _Discussion_.
+
+::: {.callout appearance="simple"}
+**In _Materials and methods_**
+
+We calculated variable importance with the *vip* R package (v0.4), by three different methods: *FIRM*, *permutation*, and *Shapley* [@greenwellVariableImportancePlots2020].
+*FIRM* values measure the flatness of the partial dependence plot, *permutation* values measure the decrease in model performance when the predictor is permuted, and *Shapley* values are aggregated from local, game-theoretical measures of variable importance [@greenwellVariableImportancePlots2020].
+[Since *FIRM* reflects the flatness of the partial dependence plot, it captures functional complexity rather than overall predictive impact.]{.added}
+*Permutation* values were obtained from ten iterations, with root mean square error as the performance measure.
+
+**In _Results_**
+
+[The *permutation* and *Shapley* methods of variable importance were highly correlated at both sites, while the *FIRM* method was less correlated with the other two (Fig. \@ref(fig:varImp)).]{.added}
+
+At both sites, elevation and Multi-Resolution Valley Bottom Flatness were important predictors (Fig. \@ref(fig:varImp)).
+At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet elevation [was more important than Multi-Resolution Valley Bottom Flatness]{.deleted} [had higher predictive impact (*permutation* and *Shapley*) but lower functional complexity (*FIRM*) than Multi-Resolution Valley Bottom Flatness]{.added}.
+:::
+
 ### Comment 2.9

### 2025-06-10 R/Skrim/modelDepth.R

@@ -483,12 +483,32 @@ import_shap <- fit_RTD_uncorr %>%
           scale = TRUE)
 vip::vip(import_shap, num_features = 30)
 
-bind_rows(perm.vip = import_perm_vip, 
+vi_combined <- bind_rows(perm.vip = import_perm_vip, 
           firm = import_firm, 
           shap = import_shap,
           .id = 'type') |> 
-  left_join(corrKey, by = c("Variable" = "correlated")) |>
+  left_join(corrKey, by = c("Variable" = "correlated"))
+
+vi_combined |> 
   readr::write_csv("output/Skrim/variable_importance.csv")
+
+### Variable importance rank correlations ####
+
+vi_ranks <- vi_combined |>
+  group_by(type) |>
+  mutate(rank = rank(-Importance, ties.method = "average")) |>
+  select(type, Variable, rank) |>
+  pivot_wider(names_from = type, values_from = rank)
+
+# Pairwise rank correlations
+rank_correlations <- vi_ranks |>
+  select(-Variable) |>
+  cor(method = "spearman", use = "complete.obs") |>
+  as_tibble(rownames = "method1") |>
+  pivot_longer(-method1, names_to = "method2", values_to = "correlation")
+rank_correlations
+rank_correlations |>
+  readr::write_csv("output/Skrim/variable_importance_rank_correlations.csv")
 
 ## Partial dependence plots ####
 # https://www.tmwr.org/explain#building-global-explanations-from-local-explanations

### 2025-06-10 R/modelDepth.R

@@ -483,12 +483,32 @@ import_shap <- fit_TD_uncorr %>%
           scale = TRUE)
 vip::vip(import_shap, num_features = 30)
 
-bind_rows(perm.vip = import_perm_vip, 
+vi_combined <- bind_rows(perm.vip = import_perm_vip, 
           firm = import_firm, 
           shap = import_shap,
           .id = 'type') |> 
-  left_join(corrKey, by = c("Variable" = "correlated")) |>
+  left_join(corrKey, by = c("Variable" = "correlated"))
+
+vi_combined |> 
   readr::write_csv("output/variable_importance.csv")
+
+### Variable importance rank correlations ####
+
+vi_ranks <- vi_combined |>
+  group_by(type) |>
+  mutate(rank = rank(-Importance, ties.method = "average")) |>
+  select(type, Variable, rank) |>
+  pivot_wider(names_from = type, values_from = rank)
+
+# Pairwise rank correlations
+rank_correlations <- vi_ranks |>
+  select(-Variable) |>
+  cor(method = "spearman", use = "complete.obs") |>
+  as_tibble(rownames = "method1") |>
+  pivot_longer(-method1, names_to = "method2", values_to = "correlation")
+rank_correlations
+rank_correlations |>
+  readr::write_csv("output/variable_importance_rank_correlations.csv")
 
 ## Partial dependence plots ####
 # https://www.tmwr.org/explain#building-global-explanations-from-local-explanations

### 2025-06-10 ms/figures.R

@@ -690,6 +690,39 @@ orskog <- read_csv("output/variable_importance.csv") %>%
 skrim <- read_csv("output/Skrim/variable_importance.csv") %>% 
   filter(type != "perm.ranger")
 
+# Read correlation matrices for insets
+corr_orskog <- read_csv("output/variable_importance_rank_correlations.csv")
+corr_skrim <- read_csv("output/Skrim/variable_importance_rank_correlations.csv")
+
+# Create simple correlation matrix plots for insets
+create_simple_corr_plot <- function(corr_data) {
+  corr_data %>%
+    mutate(
+      method1 = case_when(
+        method1 == "firm" ~ "FIRM",
+        method1 == "perm.vip" ~ "permutation", 
+        method1 == "shap" ~ "Shapley",
+        TRUE ~ method1
+      ),
+      method2 = case_when(
+        method2 == "firm" ~ "FIRM",
+        method2 == "perm.vip" ~ "permutation", 
+        method2 == "shap" ~ "Shapley",
+        TRUE ~ method2
+      )
+    ) %>%
+    ggplot(aes(method1, method2)) +
+    geom_tile(fill = "white", color = "black") +
+    geom_text(aes(label = round(correlation, 2)), size = 2.5, color = "black") +
+    coord_flip() +
+    theme_void() +
+    theme(axis.text = element_text(size = 6),
+          plot.margin = margin(2, 2, 2, 2))
+}
+
+corr_plot_orskog <- create_simple_corr_plot(corr_orskog)
+corr_plot_skrim <- create_simple_corr_plot(corr_skrim)
+
 orskog.rank <- orskog %>% 
   group_by(Variable) %>% 
   summarize(Imp.median = median(Importance))
@@ -704,7 +737,7 @@ labs.orskog <- plot.orskog |>
   arrange(Imp.median) |> 
   pull(removed) |> 
   stringr::str_wrap(width = 40)
-g1 <- plot.orskog %>% 
+pOrskog <- plot.orskog %>% 
   mutate(Variable = case_when(
     Variable == "dmkdepth_grunn.myr" ~ "DMK_shallow",
     Variable == "dmkdepth_unknown" ~ "DMK_unknown",
@@ -723,11 +756,14 @@ g1 <- plot.orskog %>%
   theme_minimal() +
   theme(axis.title.y = element_blank(),
         legend.title = element_blank(),
[…]

### 2025-06-25 ms/ms.Rmd
address RC1 Technical corrections

@@
 Maps are crucial because they link high-level targets to specific management decisions, unlike spatially aggregated estimates [@oecdOECDEnvironmentalPerformance2022].
-For example, Norwegian regulations restrict conversion of peatland to farmland more strictly where peat depth exceeds 1 m, and the distribution of peat depth on a farmer's property determines whether they are allowed to convert peatland (Forskrift om nydyrking, 1997, § 5a).
 Maps of peat depth also make it possible to quantify the effect of specific management decisions and thereby understand how local outcomes contribute to regional and national outcomes [@oecdOECDEnvironmentalPerformance2022].
@@
 
-The crux for DSM of peat depth is the relationship between peat depth and each of the other, mapped predictors. 
-For DSM to work, these relationships must be strong and consistent over the area of interest.
+The crux for DSM of peat depth is the relationship between peat depth and the other, mapped predictors. 
+For DSM to be effective, these relationships must be strong and consistent over the area of interest.
 They can be purely correlative rather than causal, but mechanistic relationships are stronger and more consistent than non-causal ones. 
@@
 Moreover, mechanisms of peat formation are linked to topography.
-For example, a steep slope is unlikely to have a high water table, thereby limiting accumulation of peat.
 
@@
 We created the same suite of peat depth predictors for both sites (25 continuous and 1 categorical; Table \@ref(tab:preds)).
-Each continuous predictor equates to one variable in the model, while the categorical predictor equates to two variables in the model because its three levels become two indicator variables (one level is used as the reference).
 All continuous predictors were derived either from an airborne radiometric survey or from a DTM.
@@
 The categorical predictor was peat depth class, from a national map dataset.
-We chose a spatial resolution of 10 m for our predictors, peat depth sample, and modelling.
+Predictor preparation is described in more detail below.
+
+We chose a spatial resolution of 10 m for our predictors, depth sampling, and modelling.
 We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other hand.
-Complete descriptions of predictors follow below.
 
@@
 
-The radiometric surveys were conduced with similar flight parameters (Table \ref{tab:radSurveys}).
+The radiometric surveys were conduced with 75--80 m average flight altitude, 88–108 \unit{km\,h^{-1}} average flight speed, and 200 m flight line spacing (Table \ref{tab:radSurveys}).
 Spectrometer count rates were calibrated to known concentrations of potassium, thorium, and uranium in mobile pads.
@@
 We calculated unitless slope from the 1 m DTM using the Whitebox software.
-Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell [@schonauerSpatiotemporalPredictionSoil2021; @schonauerRcodeCalculatingDepthwater2021].
+We defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell [@schonauerSpatiotemporalPredictionSoil2021; @schonauerRcodeCalculatingDepthwater2021].
 This catchment area layer was then thresholded at seven different levels (flow initiation area 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry [@murphyModellingMappingTopographic2011; @agrenEvaluatingDigitalTerrain2014; @schonauerSpatiotemporalPredictionSoil2021].
@@
 Surveyors generally assigned peat depth classes to polygons of at least 0.5 ha, although delineating polygons down to 0.2 ha was allowed if peat depth showed a "particularly marked difference" [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
-We rasterized the peat depth class attribute to our 10 m grid.
+We rasterized the peat depth class attribute to our 10 m grid and this predictor equates to two variables in the model because its three levels become two indicator variables (one level is used as the reference).
 
@@
 For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration -- because correlation between predictors degrades variable importance metrics [@stroblConditionalVariableImportance2008; @biauRandomForestGuided2016] and can produce misleading visualizations of predictor--outcome relationships [@biecekExplanatoryModelAnalysis2021; @dwivediExplainableAIXAI2023].
-Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation coefficient above 0.7.
+Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation coefficient above 0.7 (an arbitrary but conventional threshold for this purpose).
 Thus, highly correlated sets of variables are represented by a single variable for the purposes of model interpretation.
@@
 Our point measurements of peat depth (all sources) produced aggregated depths for 372 cells or 2.4 % of AR5 mire area at Skrimfjella, and 1878 cells or 1.2 % of AR5 mire area at Ørskogfjellet (Table \ref{tab:depthsByClass}).
-Roughly 80 % of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland.
+Approximately 80 % of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland.
 DMK peat depth had higher coverage at Ørskogfjellet than at Skrimfjella (80 % versus 27 % of cells not *unknown*), and at Ørskogfjellet the *deep* and *shallow* classes showed a larger difference in measured depth.
@@
 Some realizations of the hydrological predictors Topographic Wetness Index and Depth-to-Water showed considerable importance, while others showed little -- with no consistency between sites.
-For example, TWI5m, DTW40000, and DTW2500 rounded out the top five most important variables at Skrimfjella, while TWI20m and TWI50m were least important, other than DMK.
 At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices slope, Terrain Ruggedness Index, Topographic Position Index, and roughness.
@@
 Individual conditional expectation lines indicated some interactions between predictors (Fig. \@ref(fig:pdps)).
-For example, the magnitude of the increase in depth with elevation that the model expected at Ørskogfjellet was different for different observations; some depth predictions increased by only 40 cm while others increased by more than 100 cm, over the same elevation gain.
-Similarly, individual conditional expectation lines of uranium ground concentration at Skrimfjella were non-parallel, with some locations showing monotonically increasing peat depth predictions with uranium concentration (unlike the average effect).
-Nevertheless, most individual conditional expectation lines were generally parallel -- indicating that the average effects of the predictors were good representations of their overall effects.
+For example, the positive association of depth with elevation at Ørskogfjellet was larger for some observations than others; depth increases ranged 40--100 cm over the same elevation gain.
+Similarly, individual conditional expectations along uranium ground concentration at Skrimfjella were non-parallel, with some locations showing decreasing peat depth predictions (unlike the average effect).
+Nevertheless, most individual conditional expectations were approximately parallel, so the average effects of the predictors are representative of their overall effects.
 
 ```{r pdps, echo = FALSE, out.width="100%", fig.cap = "Partial dependence plots of the six most important variables in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b).
-The average effect of the predictor on the outcome (red line) overlays the variation in the effect across observations (grey lines).
+The average effect of the predictor on the outcome (red line) overlays the individual conditional expectations, which show the variation in the effect across observations (grey lines).
 The variables' training data distributions are indicated with rug plots along the x-axes."}
@@
 
-### Remotely sensed variables are weak (but not useless!) predictors of peat depth
+### Remotely sensed variables are weak predictors of peat depth
 
@@
 This highlights the general value of remotely sensed data, whose complete coverage can improve maps even when their association with the variable of interest is weak [@mulderUseRemoteSensing2011].
-Since remotely sensed data are widely available, improvements to soil maps as shown here are low hanging fruit.
+Since remotely sensed data are widely available, improvements to soil maps as shown here are readily achievable at low cost.
 This point is recognized and reflected in the rise of DSM [@minasnyDigitalMappingPeatlands2019].
@@
 Identifying the deepest peats will require additional field work in candidate areas, which could be defined by an upper quantile of predicted depth.
-Map users should not trust the maps to identify all large carbon stocks, but they can trust that identified large stocks really are large. 
-That makes the map more suited for "red-lighting" than "green-lighting" peatland conversion, for example. 
+That makes the maps more suited for "red-lighting" than "green-lighting" peatland conversion, for example. 
 Where it does not prohibit conversion (i.e. predicts shallow peat), a ground survey should be done before conversion is allowed (assuming depth is the only consideration). 
@@
 Other terrain-derived predictors with predictive value in our study are hydrological (Topographic Wetness Index and Depth-to-Water). 
-Notably, slope was inferior to (Ørskogfjellet) or highly correlated with (Skrimfjella) these hydrological indices. 
+Notably, slope was inferior to (at Ørskogfjellet) or highly correlated with (at Skrimfjella) these hydrological indices. 
 Mappers of peat depth should not assume that slope is the best predictor in its class, despite its prevalence in the literature. 
@@
 This is evidenced by the fact that 73 % of the cells measured at Skrimfjella had unknown depth in DMK, compared to 20 % at Ørskogfjellet.
-Our results show that classification as deep peat at Skrimfjella had no predictive value -- these cells were no different than those classified as shallow or unknown, all else being equal. 
 Interactions between DMK and other variables also underline the inconsistency of DMK depth maps, even within a site. 
-For example, that a peat at Ørskogfjellet was classified as shallow rather than deep did not always cause the model to predict shallower peat -- sometimes it increased the predicted depth.
+For example, shallow peat classification in DMK sometimes increased rather than decreased predicted depth at Ørskogfjellet.
 
@@
 
-The performance gap between the best models and the DMK models shows that peat depth in Norway should be mapped digitally. 
-Anywhere we have some calibrating measurements, we can get better maps than DMK peat depth classes, at low cost. 
-Moreover, DSM methodology can align map products with open science principles by making their production transparent, reproducible, and updatable. 
-The large difference we found between the coverage and quality of DMK peat depth at Skrimfjella versus Ørskogfjellet underlines these advantages. 
-With DSM we can apply the same approach across different areas and make maps with full spatial coverage, continuous values, and validated uncertainty. 
-The rest of this section discusses recommendations and needs for more extensive peat depth DSM.
+The performance gap between the best models and the DMK models shows that digital soil mapping of peat depth has value in Norway. 
+Where calibrating measurements are available, better maps than DMK peat depth class can be produced at low cost. 
+Moreover, DSM can make the production of maps transparent, reproducible, and updatable. 
+We can apply the same approach across different areas and make maps with full spatial coverage, continuous values, and validated uncertainty. 
+The large difference in coverage and quality of DMK peat depth at Skrimfjella versus Ørskogfjellet underlines these advantages.
 
@@
 
-High quality DTMs are available for mainland Norway, making peat depth measurements the critical training data need. 
-For a given area, some minimum number of measurements is necessary to create meaningful improvement over DMK peat depth. 
-The proportion of peatlands sampled at Skrimfjella was twice that at Ørskogfjellet, but the Skrimfjella model performed worse, which shows that the size of the depth dataset is not all-important.
-We had the luxury of stratifying our measurements over candidate predictors, and performance may suffer where locations are opportunistic or tied to a sampling design with a different purpose. 
-The magnitude of this penalty will depend on the dataset characteristics, but having enough depth measurements is probably more consequential [@wadouxSamplingDesignOptimization2019].
-
-Depth measurements are foundational, and better infrastructure to make these data findable, accessible, interoperable, and reusable would help DSM and other applications. 
+High quality DTMs are available for mainland Norway, which makes peat depth measurements the critical training data need for DSM. 
+Better infrastructure to make depth measurements findable, accessible, interoperable, and reusable would help DSM and other applications. 
 Geoportal access and data exchange standards [like Natural England’s for peat surveys, -@naturalenglandDataExchangeStandard2023] are important. 
-Peat data often fall through the cracks between geology-oriented and ecology-oriented archives, but increased awareness of peatland importance is a good impetus to remedy this situation. 
 Peat depth is quick and easy to measure, so integrating its measurement into existing national field programs, like Norway’s spatially representative nature monitoring or national forest inventory, would be helpful (although not sufficient for regional or local mapping).
-Municipalities and local actors in Norway are increasingly measuring peat depth and can contribute to a growing data foundation [@kyrkjeeideCalculatorLocalPeatland2023]. 
 Low-altitude, drone-mounted GPR may prove an efficient approach for collecting many, accurate depth data in a landscape, by combining the advantages of airborne deployment and active sensing [@pelletierPeatAnalysesHudson1991; @ruolsDevelopmentDronebasedGroundpenetrating2023].
-All of the above can lay the groundwork for renewed peatland maps.
 
@@
 
-Peat depths typically vary over short distances (e.g. < 1 m), so mapping at 10 m resolution implies that the map will compress much of the fine-scale variation in depth.
-For example, large differences in maximum depth can be obscured at 10 m resolution.
-For the same reason, seemingly small improvements in mean absolute error at 10 m resolution may represent meaningful improvements for map users.
-Spatial aggregation also means that an uneven distribution of point measurements within a cell can cause its mean depth to be unrepresentative.
-Some cells had dense and evenly distributed measurements, while others had only a few in one corner.
-Although we tried to reduce the influence of uneven point coverage -- by inversely weighing point values by their distances to the cell center -- it probably contributed to model inaccuracy.
-Fine-scale variation in peat depth may favor mapping at very fine resolution (1 m) — even if land use planning and carbon accounting do not operate at this grain. 
-Terrain--depth relationships might be stronger in 1 m cells than in our 10 m cells, especially considering the hummock--hollow microtopography of many peatlands [@rydin7Mires1999; @lindsayPeatbogsCarbonCritical2010].
+Peat depths typically vary over short distances (e.g. < 1 m), so mapping at 10 m resolution implies that the map will compress much of the fine-scale variation.
+Terrain--depth relationships might be stronger at finer resolution than 10 m, especially considering the hummock--hollow microtopography of many peatlands [@rydin7Mires1999; @lindsayPeatbogsCarbonCritical2010].
+Therefore, it may be advantageous to model peat depth at 1 m resolution -- even if land use planning and carbon accounting do not operate at such fine scale.
 
-The scarcity of depth measurements and their short spatial autocorrelation mean that mapping peat depth at broad scales is not an exercise in spatial interpolation [@henglGenericFrameworkSpatial2004].
-Peat depths in our GPR data showed spatial autocorrelation to a range of 50--100 m, and getting measurements at such fine grain is only realistic for small areas, not across whole landscapes. 
-Therefore, we anticipate that spatially explicit DSM is of limited value in all but the most intensively sampled landscapes (currently nowhere in Norway).
-
-Choosing a spatial extent for DSM can be tricky. 
-A natural starting point is the bounding area around a spatial cluster of depth measurements. 
-For a given cluster, kNNDM can define how far the boundary can extend beyond the point measurements; if the sample-to-prediction nearest neighbor distribution cannot be simulated by any set of cross-validation folds, then the extent is too expansive [@meyerMachineLearningbasedGlobal2022; @linnenbrinkKNNDMCVKfold2024].
-However, it is unclear how big (N) any cluster of depths should be, and the tradeoff between multiple small-extent DSM and fewer large-extent DSM needs research. 
+To define a spatial extent for a particular DSM, kNNDM can be used to investigate how far the mapping can extend beyond a set of point measurements; if the sample-to-prediction nearest neighbor distribution cannot be simulated by any set of cross-validation folds, then the extent is too expansive [@meyerMachineLearningbasedGlobal2022; @linnenbrinkKNNDMCVKfold2024].
+However, the tradeoff between multiple small-extent DSM and fewer large-extent DSM needs research. 
 Bohn and Miller [-@bohnLocallyEnhancedDigital2024] advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or 'supertopes'), where the composition of mesotopes is similar -- e.g.
@@
 @kogantiMappingPeatDepth2023] and produces good uncertainty estimates [our study, @vaysseUsingQuantileRegression2017;@takoutsingComparingPredictionPerformance2022].
-As depth data become more abundant, we may move from pixel-based learners to deep learning approaches like convolutional neural networks. 
-Bakkestuen et al.
-[-@bakkestuenDelineationWetlandAreas2023] successfully predicted peatland extent using convolutional neural networks, and their ability to automatically learn multiscale spatial features would reduce the need to manually engineer these.
-The success of Multi-Resolution Valley Bottom Flatness as a predictor of peat depth demonstrates that multiscale spatial patterns matter for peat depth and convolutional neural networks are designed to learn such patterns.
+As depth data become more abundant, pixel-based learners may be surpassed by deep learning approaches like convolutional neural networks. 
+The success of Multi-Resolution Valley Bottom Flatness as a predictor of peat depth demonstrates that multiscale spatial patterns matter for peat depth, and convolutional neural networks are designed to learn such patterns [@borowiecDeepLearningTool2022].
 The kind of relationship described in Buffam et al.
@@
 However, since this approach is data-hungry, we should build soil knowledge into the DSM where we can [@minasnySoilScienceInformedMachine2024].
-For example, if further research confirms the effect of the marine limit that we found, then it is better to include the marine limit as a predictor than to make the algorithm learn this pattern from elevation independently.
 
@@
 Though peatland definitions may encourage reducing continuous depth predictions to arbitrary classes, we caution against this practice [as in @ivanovsModelingGeospatialDistribution2024; @karjalainenComparisonTwoGammaray2025].
+
+# Conclusions
+
+This study demonstrates that digital soil mapping at 10 m resolution can improve upon existing peat depth maps in Norway, though the strength of the relationship between available predictors and peat depth remains limited.
+Our findings show that terrain-derived variables, particularly elevation and Multi-Resolution Valley Bottom Flatness, provide predictive value for peat depth mapping within peatland extents. 
+In contrast, airborne radiometric data showed little to no predictive value at either of two study sites, likely due to the large footprint of airborne sensors relative to the fine-scale variation in peat depths.
+
+The best models achieved mean absolute errors of 56--60 cm against mean depths of 119--126 cm, and explained approximately one-third of the variation in peat depth across landscapes with with aggregate peatland areas of 1.5--15.3 km^2^.
+Though field measurements remain necessary for local, detailed assessments of peat depth, digital soil maps at 10 m resolution can provide valuable information for landscape-scale planning and regional carbon assessments.
+The models' tendency to underpredict the deepest peats has important practical implications, making them more suitable for precautionary screening than comprehensive coverage.
+As Norway and other nations pursue nature-based climate solutions, these findings highlight both the potential and limitations of remote sensing for peatland carbon mapping.
 

### 2025-06-25 ms/ms.tex
address RC1 Technical corrections

@@
 Maps are crucial because they link high-level targets to specific management decisions, unlike spatially aggregated estimates \citep{oecdOECDEnvironmentalPerformance2022}.
-For example, Norwegian regulations restrict conversion of peatland to farmland more strictly where peat depth exceeds 1 m, and the distribution of peat depth on a farmer's property determines whether they are allowed to convert peatland (Forskrift om nydyrking, 1997, § 5a).
 Maps of peat depth also make it possible to quantify the effect of specific management decisions and thereby understand how local outcomes contribute to regional and national outcomes \citep{oecdOECDEnvironmentalPerformance2022}.
@@
 
-The crux for DSM of peat depth is the relationship between peat depth and each of the other, mapped predictors.
-For DSM to work, these relationships must be strong and consistent over the area of interest.
+The crux for DSM of peat depth is the relationship between peat depth and the other, mapped predictors.
+For DSM to be effective, these relationships must be strong and consistent over the area of interest.
 They can be purely correlative rather than causal, but mechanistic relationships are stronger and more consistent than non-causal ones.
@@
 We created the same suite of peat depth predictors for both sites (25 continuous and 1 categorical; Table \ref{tab:preds}).
-Each continuous predictor equates to one variable in the model, while the categorical predictor equates to two variables in the model because its three levels become two indicator variables (one level is used as the reference).
 All continuous predictors were derived either from an airborne radiometric survey or from a DTM.
@@
 
-The radiometric surveys were conduced with similar flight parameters (Table \ref{tab:radSurveys}).
+The radiometric surveys were conduced with 75--80 m average flight altitude, 88--108 \unit{km\,h^{-1}} average flight speed, and 200 m flight line spacing (Table \ref{tab:radSurveys}).
 Spectrometer count rates were calibrated to known concentrations of potassium, thorium, and uranium in mobile pads.
@@
 We calculated unitless slope from the 1 m DTM using the Whitebox software.
-Also using Whitebox, we defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell \citep{schonauerSpatiotemporalPredictionSoil2021, schonauerRcodeCalculatingDepthwater2021}.
+We defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell \citep{schonauerSpatiotemporalPredictionSoil2021, schonauerRcodeCalculatingDepthwater2021}.
 This catchment area layer was then thresholded at seven different levels (flow initiation area 0.25--16 ha) to estimate surface water features under moisture scenarios varying from wet to dry \citep{murphyModellingMappingTopographic2011, agrenEvaluatingDigitalTerrain2014, schonauerSpatiotemporalPredictionSoil2021}.
@@
 Surveyors generally assigned peat depth classes to polygons of at least 0.5 ha, although delineating polygons down to 0.2 ha was allowed if peat depth showed a ``particularly marked difference'' \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
-We rasterized the peat depth class attribute to our 10 m grid.
+We rasterized the peat depth class attribute to our 10 m grid and this predictor equates to two variables in the model because its three levels become two indicator variables (one level is used as the reference).
 
@@
 For both sites, we interpreted a model trained on a non-collinear subset of variables from the best performing predictor configuration -- because correlation between predictors degrades variable importance metrics \citep{stroblConditionalVariableImportance2008, biauRandomForestGuided2016} and can produce misleading visualizations of predictor--outcome relationships \citep{biecekExplanatoryModelAnalysis2021, dwivediExplainableAIXAI2023}.
-Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation coefficient above 0.7.
+Specifically, we eliminated variables from the best performing predictor configuration to obtain a set with no pairwise Pearson correlation coefficient above 0.7 (an arbitrary but conventional threshold for this purpose).
 Thus, highly correlated sets of variables are represented by a single variable for the purposes of model interpretation.
@@
 Our point measurements of peat depth (all sources) produced aggregated depths for 372 cells or 2.4 \% of AR5 mire area at Skrimfjella, and 1878 cells or 1.2 \% of AR5 mire area at Ørskogfjellet (Table \ref{tab:depthsByClass}).
-Roughly 80 \% of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland.
+Approximately 80 \% of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland.
 DMK peat depth had higher coverage at Ørskogfjellet than at Skrimfjella (80 \% versus 27 \% of cells not \emph{unknown}), and at Ørskogfjellet the \emph{deep} and \emph{shallow} classes showed a larger difference in measured depth.
@@
 The first row summarizes all \unit{10\,m} cells, while subsequent rows stratify the dataset by AR5 land class and DMK peat depth class.}
-\begin{tabular}{llllllll}
-\hline
-    &                             & \multicolumn{3}{l}{Skrimfjella} & \multicolumn{3}{l}{Ørskogfjellet} \\ \cline{3-8} 
-    &                             & n     & \%   & mean (SD) depth  & n      & \%    & mean (SD) depth  \\ \cline{3-8} 
+\begin{tabular}{llrrr@{\hspace{5em}}rrr}
+\hline
+    &                             & \multicolumn{3}{l}{Skrimfjella} & \multicolumn{3}{l}{Ørskogfjellet} \\ \cline{3-5} \cline{6-8}
+    &                             & n     & \%   & mean (SD) depth  & n      & \%    & mean (SD) depth  \\ \hline 
 all &                             & 372   & 100  & 119 (83) cm      & 1878   & 100   & 126 (119) cm     \\
@@
 Individual conditional expectation lines indicated some interactions between predictors (Fig. \ref{fig:pdps}).
-For example, the magnitude of the increase in depth with elevation that the model expected at Ørskogfjellet was different for different observations; some depth predictions increased by only 40 cm while others increased by more than 100 cm, over the same elevation gain.
-Similarly, individual conditional expectation lines of uranium ground concentration at Skrimfjella were non-parallel, with some locations showing monotonically increasing peat depth predictions with uranium concentration (unlike the average effect).
-Nevertheless, most individual conditional expectation lines were generally parallel -- indicating that the average effects of the predictors were good representations of their overall effects.
+For example, the positive association of depth with elevation at Ørskogfjellet was larger for some observations than others; depth increases ranged 40--100 cm over the same elevation gain.
+Similarly, individual conditional expectations along uranium ground concentration at Skrimfjella were non-parallel, with some locations showing decreasing peat depth predictions (unlike the average effect).
+Nevertheless, most individual conditional expectations were approximately parallel, so the average effects of the predictors are representative of their overall effects.
 
@@
 \includegraphics[width=1\linewidth]{figures/partial_dependence} \caption{Partial dependence plots of the six most important variables in the best-performing models at Skrimfjella (a) and Ørskogfjellet (b).
-The average effect of the predictor on the outcome (red line) overlays the variation in the effect across observations (grey lines).
+The average effect of the predictor on the outcome (red line) overlays the individual conditional expectations, which show the variation in the effect across observations (grey lines).
 The variables' training data distributions are indicated with rug plots along the x-axes.}\label{fig:pdps}
@@
 
-\subsubsection{Remotely sensed variables are weak (but not useless!) predictors of peat depth}
+\subsubsection{Remotely sensed variables are weak predictors of peat depth}
 
@@
 This highlights the general value of remotely sensed data, whose complete coverage can improve maps even when their association with the variable of interest is weak \citep{mulderUseRemoteSensing2011}.
-Since remotely sensed data are widely available, improvements to soil maps as shown here are low hanging fruit.
+Since remotely sensed data are widely available, improvements to soil maps as shown here are readily achievable at low cost.
 This point is recognized and reflected in the rise of DSM \citep{minasnyDigitalMappingPeatlands2019}.
@@
 Identifying the deepest peats will require additional field work in candidate areas, which could be defined by an upper quantile of predicted depth.
-Map users should not trust the maps to identify all large carbon stocks, but they can trust that identified large stocks really are large.
-That makes the map more suited for ``red-lighting'' than ``green-lighting'' peatland conversion, for example.
+That makes the maps more suited for ``red-lighting'' than ``green-lighting'' peatland conversion, for example.
 Where it does not prohibit conversion (i.e.~predicts shallow peat), a ground survey should be done before conversion is allowed (assuming depth is the only consideration).
@@
 Other terrain-derived predictors with predictive value in our study are hydrological (Topographic Wetness Index and Depth-to-Water).
-Notably, slope was inferior to (Ørskogfjellet) or highly correlated with (Skrimfjella) these hydrological indices.
+Notably, slope was inferior to (at Ørskogfjellet) or highly correlated with (at Skrimfjella) these hydrological indices.
 Mappers of peat depth should not assume that slope is the best predictor in its class, despite its prevalence in the literature.
@@
 This is evidenced by the fact that 73 \% of the cells measured at Skrimfjella had unknown depth in DMK, compared to 20 \% at Ørskogfjellet.
-Our results show that classification as deep peat at Skrimfjella had no predictive value -- these cells were no different than those classified as shallow or unknown, all else being equal.
 Interactions between DMK and other variables also underline the inconsistency of DMK depth maps, even within a site.
-For example, that a peat at Ørskogfjellet was classified as shallow rather than deep did not always cause the model to predict shallower peat -- sometimes it increased the predicted depth.
+For example, shallow peat classification in DMK sometimes increased rather than decreased predicted depth at Ørskogfjellet.
 
@@
 
-The performance gap between the best models and the DMK models shows that peat depth in Norway should be mapped digitally.
-Anywhere we have some calibrating measurements, we can get better maps than DMK peat depth classes, at low cost.
-Moreover, DSM methodology can align map products with open science principles by making their production transparent, reproducible, and updatable.
-The large difference we found between the coverage and quality of DMK peat depth at Skrimfjella versus Ørskogfjellet underlines these advantages.
-With DSM we can apply the same approach across different areas and make maps with full spatial coverage, continuous values, and validated uncertainty.
-The rest of this section discusses recommendations and needs for more extensive peat depth DSM.
+The performance gap between the best models and the DMK models shows that digital soil mapping of peat depth has value in Norway.
+Where calibrating measurements are available, better maps than DMK peat depth class can be produced at low cost.
+Moreover, DSM can make the production of maps transparent, reproducible, and updatable.
+We can apply the same approach across different areas and make maps with full spatial coverage, continuous values, and validated uncertainty.
+The large difference in coverage and quality of DMK peat depth at Skrimfjella versus Ørskogfjellet underlines these advantages.
 
@@
 
-High quality DTMs are available for mainland Norway, making peat depth measurements the critical training data need.
-For a given area, some minimum number of measurements is necessary to create meaningful improvement over DMK peat depth.
-The proportion of peatlands sampled at Skrimfjella was twice that at Ørskogfjellet, but the Skrimfjella model performed worse, which shows that the size of the depth dataset is not all-important.
-We had the luxury of stratifying our measurements over candidate predictors, and performance may suffer where locations are opportunistic or tied to a sampling design with a different purpose.
-The magnitude of this penalty will depend on the dataset characteristics, but having enough depth measurements is probably more consequential \citep{wadouxSamplingDesignOptimization2019}.
-
-Depth measurements are foundational, and better infrastructure to make these data findable, accessible, interoperable, and reusable would help DSM and other applications.
+High quality DTMs are available for mainland Norway, which makes peat depth measurements the critical training data need for DSM.
+Better infrastructure to make depth measurements findable, accessible, interoperable, and reusable would help DSM and other applications.
 Geoportal access and data exchange standards \citeyearpar[like Natural England's for peat surveys,][]{naturalenglandDataExchangeStandard2023} are important.
-Peat data often fall through the cracks between geology-oriented and ecology-oriented archives, but increased awareness of peatland importance is a good impetus to remedy this situation.
 Peat depth is quick and easy to measure, so integrating its measurement into existing national field programs, like Norway's spatially representative nature monitoring or national forest inventory, would be helpful (although not sufficient for regional or local mapping).
-Municipalities and local actors in Norway are increasingly measuring peat depth and can contribute to a growing data foundation \citep{kyrkjeeideCalculatorLocalPeatland2023}.
 Low-altitude, drone-mounted GPR may prove an efficient approach for collecting many, accurate depth data in a landscape, by combining the advantages of airborne deployment and active sensing \citep{pelletierPeatAnalysesHudson1991, ruolsDevelopmentDronebasedGroundpenetrating2023}.
-All of the above can lay the groundwork for renewed peatland maps.
 
@@
 
-Peat depths typically vary over short distances (e.g.~\textless{} 1 m), so mapping at 10 m resolution implies that the map will compress much of the fine-scale variation in depth.
-For example, large differences in maximum depth can be obscured at 10 m resolution.
-For the same reason, seemingly small improvements in mean absolute error at 10 m resolution may represent meaningful improvements for map users.
-Spatial aggregation also means that an uneven distribution of point measurements within a cell can cause its mean depth to be unrepresentative.
-Some cells had dense and evenly distributed measurements, while others had only a few in one corner.
-Although we tried to reduce the influence of uneven point coverage -- by inversely weighing point values by their distances to the cell center -- it probably contributed to model inaccuracy.
-Fine-scale variation in peat depth may favor mapping at very fine resolution (1 m) --- even if land use planning and carbon accounting do not operate at this grain.
-Terrain--depth relationships might be stronger in 1 m cells than in our 10 m cells, especially considering the hummock--hollow microtopography of many peatlands \citep{rydin7Mires1999, lindsayPeatbogsCarbonCritical2010}.
-
-The scarcity of depth measurements and their short spatial autocorrelation mean that mapping peat depth at broad scales is not an exercise in spatial interpolation \citep{henglGenericFrameworkSpatial2004}.
-Peat depths in our GPR data showed spatial autocorrelation to a range of 50--100 m, and getting measurements at such fine grain is only realistic for small areas, not across whole landscapes.
-Therefore, we anticipate that spatially explicit DSM is of limited value in all but the most intensively sampled landscapes (currently nowhere in Norway).
-
-Choosing a spatial extent for DSM can be tricky.
-A natural starting point is the bounding area around a spatial cluster of depth measurements.
-For a given cluster, kNNDM can define how far the boundary can extend beyond the point measurements; if the sample-to-prediction nearest neighbor distribution cannot be simulated by any set of cross-validation folds, then the extent is too expansive \citep{meyerMachineLearningbasedGlobal2022, linnenbrinkKNNDMCVKfold2024}.
-However, it is unclear how big (N) any cluster of depths should be, and the tradeoff between multiple small-extent DSM and fewer large-extent DSM needs research.
+Peat depths typically vary over short distances (e.g.~\textless{} 1 m), so mapping at 10 m resolution implies that the map will compress much of the fine-scale variation.
+Terrain--depth relationships might be stronger at finer resolution than 10 m, especially considering the hummock--hollow microtopography of many peatlands \citep{rydin7Mires1999, lindsayPeatbogsCarbonCritical2010}.
+Therefore, it may be advantageous to model peat depth at 1 m resolution -- even if land use planning and carbon accounting do not operate at such fine scale.
+
+To define a spatial extent for a particular DSM, kNNDM can be used to investigate how far the mapping can extend beyond a set of point measurements; if the sample-to-prediction nearest neighbor distribution cannot be simulated by any set of cross-validation folds, then the extent is too expansive \citep{meyerMachineLearningbasedGlobal2022, linnenbrinkKNNDMCVKfold2024}.
+However, the tradeoff between multiple small-extent DSM and fewer large-extent DSM needs research.
 Bohn and Miller \citeyearpar{bohnLocallyEnhancedDigital2024} advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or `supertopes'), where the composition of mesotopes is similar -- e.g.~regions dominated by raised bogs versus regions dominated by sloping fens \citep{moenNationalAtlasNorway1999, joostenWiseUseMires2002}.
@@
 RF avoids negative predictions \citep[c.f.][]{kogantiMappingPeatDepth2023} and produces good uncertainty estimates \citep[our study,][]{vaysseUsingQuantileRegression2017, takoutsingComparingPredictionPerformance2022}.
-As depth data become more abundant, we may move from pixel-based learners to deep learning approaches like convolutional neural networks.
-Bakkestuen et al.
-\citeyearpar{bakkestuenDelineationWetlandAreas2023} successfully predicted peatland extent using convolutional neural networks, and their ability to automatically learn multiscale spatial features would reduce the need to manually engineer these.
-The success of Multi-Resolution Valley Bottom Flatness as a predictor of peat depth demonstrates that multiscale spatial patterns matter for peat depth and convolutional neural networks are designed to learn such patterns.
+As depth data become more abundant, pixel-based learners may be surpassed by deep learning approaches like convolutional neural networks.
+The success of Multi-Resolution Valley Bottom Flatness as a predictor of peat depth demonstrates that multiscale spatial patterns matter for peat depth, and convolutional neural networks are designed to learn such patterns \citep{borowiecDeepLearningTool2022}.
 The kind of relationship described in Buffam et al.
@@
 However, since this approach is data-hungry, we should build soil knowledge into the DSM where we can \citep{minasnySoilScienceInformedMachine2024}.
-For example, if further research confirms the effect of the marine limit that we found, then it is better to include the marine limit as a predictor than to make the algorithm learn this pattern from elevation independently.
 
@@
 Though peatland definitions may encourage reducing continuous depth predictions to arbitrary classes, we caution against this practice \citep[as in][]{ivanovsModelingGeospatialDistribution2024, karjalainenComparisonTwoGammaray2025}.
+
+\section{Conclusions}
+
+This study demonstrates that digital soil mapping at 10 m resolution can improve upon existing peat depth maps in Norway, though the strength of the relationship between available predictors and peat depth remains limited.
+Our findings show that terrain-derived variables, particularly elevation and Multi-Resolution Valley Bottom Flatness, provide predictive value for peat depth mapping within peatland extents.
+In contrast, airborne radiometric data showed little to no predictive value at either of two study sites, likely due to the large footprint of airborne sensors relative to the fine-scale variation in peat depths.
+
+The best models achieved mean absolute errors of 56--60 cm against mean depths of 119--126 cm, and explained approximately one-third of the variation in peat depth across landscapes with with aggregate peatland areas of 1.5--15.3 km\textsuperscript{2}.
+Though field measurements remain necessary for local, detailed assessments of peat depth, digital soil maps at 10 m resolution can provide valuable information for landscape-scale planning and regional carbon assessments.
+The models' tendency to underpredict the deepest peats has important practical implications, making them more suitable for precautionary screening than comprehensive coverage.
+As Norway and other nations pursue nature-based climate solutions, these findings highlight both the potential and limitations of remote sensing for peatland carbon mapping.
 

### 2025-06-25 ms/submission/AuthorComments-egusphere-2025-1046.qmd
address RC1 Technical corrections

@@
 
+Thank you. We have removed this sentence and looked to remove similar cases.
+
 > Section 2.2 I would suggest moving this opening paragraph to the end of this section as it acts more to sum up how the authors are using the various predictors.
@@
 
+Thank you for this suggestion.
+This opening paragraph is intended to give a brief overview of the full suite of predictors before going into a complete description of how each predictor was derived.
+We think therefore that it is best placed at the start of the section, but we have tried to improve its readability within the structure (including the mentions of specific predictors).
+
 > Line 179: Remove “also using White Box” as this is obvious.
+
+Thank you.
 
@@
 
+Thank you. We have clarified that this was the operator's observation and subjective interpretation.
+
 >  371: What is the relevance of the Persons correlation coefficient of 0.7. Was this tested at all? Readers may not be familiar with this so it should be explained a bit more.
 
+Thank you. We have added some more context to explain the 0.7-cutoff.
+
 > Table 2:  I recommend putting a vertical line between the results for both sites so as to easier distinguish between them.
+
+Thank you. We have reexamined the formatting of the table.
 
@@
 
+Thank you for this example.
+We have gone through the manuscript and revised the language to make it more formal and scientific, while aiming to retain clarity and accessibility.
+We have removed this phrase and similar casual language throughout the manuscript.
+
 > Line 464: “low hanging fruit” is also casual and colloquial and may not be understood by all cultures.
+
+Thank you. Removed.
 
@@
 
+Thank you. Clarified.
+
 > Line 530: remove the question at the start of this section.
+
+Thank you. Removed.
 
@@
 I would recommend moving some of this section to the Methods and Material section as it is useful descriptors of how the data came to be.
+
+Thank you. We have added a very brief summary of the radiometric surveys to the _Materials and methods_ section, while retaining the reference to an appendix for the full details.
 
@@
 
+Thank you. We have clarified our statement based on the information in the report by the Geological Survey of Norway.
+
 > Line 628 – 629: remove the sentence starting “The rest of this section….” As it is unnecessary.
+
+Thank you. Removed.
 
@@
 
+Revised.
+
 > Line 664: “tricky” – casual
+
+Revised.
 
@@
 I suggest the authors either add a section at the end and move some text here to highlight the main conclusions as currently the “discussion” section is quite large and probably not appropriate to act as a combined discussion and conclusion section.
+
+Thank you for this helpful recommendation.
+We have added a _Conclusions_ section to summarize the key findings of the study.
+This summary highlights the main results and their implications, which we hope will help readers more easily see how the study addresses its research aims of quantifying predictive accuracy and identifying key predictors.
+
+We have also used this addition to reduce the length of the _Discussion_ section, by reserving some of the more general discussion points for the _Conclusions_.
 
@@
 
+Thank you for this helpful recommendation.
+We have added a _Conclusions_ section to summarize the key findings of the study.
+This summary highlights the main results and their implications, which we hope will help readers more easily see how the study addresses its research aims of quantifying predictive accuracy and identifying key predictors.
+
 ## Technical Corrections

### 2025-06-25 ms/ms.Rmd
address RC2 Technical corrections

@@
 Because of its agricultural and silvicultural focus, the survey covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable [@ahlstromAR5Klassifikasjonssystem2019].
-Field surveyors carried a 1 m probe and they measured peat depth as a categorical variable: shallow (<1 m), or deep (>1 m). 
+Field surveyors carried a 1 m probe and they measured peat depth as a categorical variable: *shallow* (<1 m), or *deep* (>1 m). 
 These classes were assigned to whole polygons, so spatial resolution is on the order of hectares.
@@
 Therefore, we produced prediction intervals with quantile regression forests [@meinshausenQuantileRegressionForests2006], and used the same spatial cross-validation to evaluate the prediction interval coverage probability [@shresthaMachineLearningApproaches2006].
-The quantile regression forests were trained with predictor configuration that showed the highest performance at each site (under the assumption that these models would be put into production) and we extracted 90 % prediction intervals.
+The quantile regression forests were trained with predictor configuration that showed the highest performance at each site, and we extracted 90 % prediction intervals.
 
@@
 
-Although the difficulty of predicting peat depth caused prediction intervals to be wide, these uncertainty estimates were well calibrated.
-At Skrimfjella, the prediction interval coverage probability was 91 %, and at Ørskogfjellet it was 84 % (both compared to the target value of 90 %).
-Observations outside of the prediction intervals were evenly distributed across each of the study areas.
-
 For Skrimfjella, the best predictor configuration was *all predictors*, followed closely by *terrain + radiometric* (Fig. \@ref(fig:modelMetrics)).
@@
 The mean error (bias) of these models was 10 cm at Skrimfjella and -4 cm at Ørskogfjellet.
+Although the prediction intervals from the quantile regression forests were wide, they were well calibrated.
+At Skrimfjella, the prediction interval coverage probability was 92 %, and at Ørskogfjellet it was 84 % (both compared to the target value of 90 %).
+Observations outside of the prediction intervals showed no obvious spatial pattern.
 
-```{r calPlots, echo = FALSE, out.width="100%", fig.cap = "Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions from spatial cross-validation.
+```{r calPlots, echo = FALSE, out.width="100%", fig.cap = "Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions and 90 % prediction intervals (grey veritical lines) from spatial cross-validation.
 Points are transparent to show overlap.
@@
 At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet elevation had higher predictive impact (*permutation* and *Shapley*) but lower functional complexity (*FIRM*) than Multi-Resolution Valley Bottom Flatness.
-DMK was also important -- the shallow class in particular -- but only at Ørskogfjellet.
+DMK was also important -- the *shallow* class in particular -- but only at Ørskogfjellet.
 Some realizations of the hydrological predictors Topographic Wetness Index and Depth-to-Water showed considerable importance, while others showed little -- with no consistency between sites.
@@
 At Skrimfjella, it barely improved model performance (Fig. \@ref(fig:modelMetrics)). 
-At Ørskogfjellet, it increased performance more, and both indicator variables were among the most important in the model (Fig. \@ref(fig:varImp)). 
+At Ørskogfjellet, it increased performance more, and both indicator variables (*shallow* and *unknown*) were among the most important in the model (Fig. \@ref(fig:varImp)). 
 We suspect the discrepancy between sites is due to different levels of effort and coverage during the historical surveys; more lowland peatland near agriculture at Ørskogfjellet may have caused more purposeful surveying.
@@
 Interactions between DMK and other variables also underline the inconsistency of DMK depth maps, even within a site. 
-For example, shallow peat classification in DMK sometimes increased rather than decreased predicted depth at Ørskogfjellet.
+For example, *shallow* peat classification in DMK sometimes increased rather than decreased predicted depth at Ørskogfjellet.
 

### 2025-06-25 ms/ms.tex
address RC2 Technical corrections

@@
 Because of its agricultural and silvicultural focus, the survey covered only productive areas below the tree line, and peat depth was measured only in places judged to be potentially arable or afforestable \citep{ahlstromAR5Klassifikasjonssystem2019}.
-Field surveyors carried a 1 m probe and they measured peat depth as a categorical variable: shallow (\textless1 m), or deep (\textgreater1 m).
+Field surveyors carried a 1 m probe and they measured peat depth as a categorical variable: \emph{shallow} (\textless1 m), or \emph{deep} (\textgreater1 m).
 These classes were assigned to whole polygons, so spatial resolution is on the order of hectares.
@@
 Therefore, we produced prediction intervals with quantile regression forests \citep{meinshausenQuantileRegressionForests2006}, and used the same spatial cross-validation to evaluate the prediction interval coverage probability \citep{shresthaMachineLearningApproaches2006}.
-The quantile regression forests were trained with predictor configuration that showed the highest performance at each site (under the assumption that these models would be put into production) and we extracted 90 \% prediction intervals.
+The quantile regression forests were trained with predictor configuration that showed the highest performance at each site, and we extracted 90 \% prediction intervals.
 
@@
 
-Although the difficulty of predicting peat depth caused prediction intervals to be wide, these uncertainty estimates were well calibrated.
-At Skrimfjella, the prediction interval coverage probability was 91 \%, and at Ørskogfjellet it was 84 \% (both compared to the target value of 90 \%).
-Observations outside of the prediction intervals were evenly distributed across each of the study areas.
-
 For Skrimfjella, the best predictor configuration was \emph{all predictors}, followed closely by \emph{terrain + radiometric} (Fig. \ref{fig:modelMetrics}).
@@
 The mean error (bias) of these models was 10 cm at Skrimfjella and -4 cm at Ørskogfjellet.
+Although the prediction intervals from the quantile regression forests were wide, they were well calibrated.
+At Skrimfjella, the prediction interval coverage probability was 92 \%, and at Ørskogfjellet it was 84 \% (both compared to the target value of 90 \%).
+Observations outside of the prediction intervals showed no obvious spatial pattern.
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{figures/calibration_plots} \caption{Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions from spatial cross-validation.
+\includegraphics[width=1\linewidth]{figures/calibration_plots} \caption{Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions and 90 \% prediction intervals (grey veritical lines) from spatial cross-validation.
 Points are transparent to show overlap.
@@
 At Skrimfjella these two predictors were of similar importance, while at Ørskogfjellet elevation had higher predictive impact (\emph{permutation} and \emph{Shapley}) but lower functional complexity (\emph{FIRM}) than Multi-Resolution Valley Bottom Flatness.
-DMK was also important -- the shallow class in particular -- but only at Ørskogfjellet.
+DMK was also important -- the \emph{shallow} class in particular -- but only at Ørskogfjellet.
 Some realizations of the hydrological predictors Topographic Wetness Index and Depth-to-Water showed considerable importance, while others showed little -- with no consistency between sites.
@@
 At Skrimfjella, it barely improved model performance (Fig. \ref{fig:modelMetrics}).
-At Ørskogfjellet, it increased performance more, and both indicator variables were among the most important in the model (Fig. \ref{fig:varImp}).
+At Ørskogfjellet, it increased performance more, and both indicator variables (\emph{shallow} and \emph{unknown}) were among the most important in the model (Fig. \ref{fig:varImp}).
 We suspect the discrepancy between sites is due to different levels of effort and coverage during the historical surveys; more lowland peatland near agriculture at Ørskogfjellet may have caused more purposeful surveying.
@@
 Interactions between DMK and other variables also underline the inconsistency of DMK depth maps, even within a site.
-For example, shallow peat classification in DMK sometimes increased rather than decreased predicted depth at Ørskogfjellet.
+For example, \emph{shallow} peat classification in DMK sometimes increased rather than decreased predicted depth at Ørskogfjellet.
 

### 2025-06-25 ms/submission/AuthorComments-egusphere-2025-1046.qmd
address RC2 Technical corrections

@@
 
+Thank you for identifying this confusion.
+We have revised the text to clarify that we are comparing the quartile predictions to the observed data, to follow best practices in digital soil mapping.
+We have also added the prediction intervals to figure 3, and moved the text description to where we also present the figure.
+We think that this makes this particular result clearer to readers.
+
 > Line 413-414: which curve? The information inside the parentheses is not clear to me.
+
+Thank you. Revised for clarity.
 
@@
 
+We have removed this figure along with all sections about peat extent mapping.
+The null did reflect the error between observations and assuming 30 cm depth.
+The standard errors on the other error estimates were from the cross-validation folds.
+
 > Line 464: “low hanging fruit” ?
+
+Revised.
 
@@
 
+Revised. The sentence was intended to reiterate what came before, so we removed it for clarity conciseness.
+
 > Line 531: “0% in prediction” what does this sentence mean?
+
+Removed along with all sections about peat extent (occurrence) mapping.
 
@@
 
-# Response to Referee 3
-
-### Comment 3.1
-
-> Referee's comment goes here.
-
-We thank the Referee for this insightful comment. We have revised the manuscript accordingly:
-
-::: {.callout appearance="simple"}
-**Original text:**
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Sed do eiusmod tempor incididunt ut labore et dolore magna aliqua.
-
-**Revised text:**
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit. Sed do eiusmod tempor incididunt ut labore et **dolore magna aliqua** [and additional explanation of the concept]{.added}.
-:::
-
-### Comment 3.2
-
-> Referee's comment goes here.
-
-Thank you for highlighting this issue. We have made the following changes:
-
-::: {.callout appearance="simple"}
-**Original text:**
-
-Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.
-
-**Revised text:**
-
-[We have completely revised this section]{.added} [to address the Referee's concerns.]{.added} [Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat.]{.deleted}
-:::
+Revised for clarity.
 
@@
 > Please add the section headlines "Appendix A" (in front of Table A1) and "Appendix B" (in front of Figure B1).
+
+Thank you. Revised.

### 2025-06-25 R/Skrim/modelDepth.R

@@ -335,20 +335,23 @@ quant_predict <- function(fit, new_data, level = 0.9) {
   quant_pred
 }
 # Fit the model and get predictions using resampling
-quantiles.cv <- map(folds$splits, function(spliti) {
-  train_data <- analysis(spliti)
-  test_data <- assessment(spliti)
-  # Fit QRF model
-  set.seed(345)
-  qrf_wf <- workflow_RTD_uncertainty %>% 
-    fit(train_data)
-  # Make predictions
-  test_data_baked <- workflows::extract_recipe(qrf_wf) %>% 
-    bake(test_data)
-  predictions <- quant_predict(qrf_wf$fit$fit$fit, test_data_baked)
-  # Return results
-  bind_cols(predictions, test_data)
-}) %>% 
+quantiles.cv <- folds |> 
+  arrange(id) |> 
+  pull(splits) |> 
+  map(function(spliti) {
+    train_data <- analysis(spliti)
+    test_data <- assessment(spliti)
+    # Fit QRF model
+    set.seed(345)
+    qrf_wf <- workflow_RTD_uncertainty %>% 
+      fit(train_data)
+    # Make predictions
+    test_data_baked <- workflows::extract_recipe(qrf_wf) %>% 
+      bake(test_data)
+    predictions <- quant_predict(qrf_wf$fit$fit$fit, test_data_baked)
+    # Return results
+    bind_cols(predictions, test_data)
+  }) %>% 
   bind_rows()
 
 # Evaluate prediction intervals

### 2025-06-25 R/modelDepth.R

@@ -335,20 +335,23 @@ quant_predict <- function(fit, new_data, level = 0.9) {
   quant_pred
 }
 # Fit the model and get predictions using resampling
-quantiles.cv <- map(folds$splits, function(spliti) {
-  train_data <- analysis(spliti)
-  test_data <- assessment(spliti)
-  # Fit QRF model
-  set.seed(345)
-  qrf_wf <- workflow_TD_uncertainty %>% 
-    fit(train_data)
-  # Make predictions
-  test_data_baked <- workflows::extract_recipe(qrf_wf) %>% 
-    bake(test_data)
-  predictions <- quant_predict(qrf_wf$fit$fit$fit, test_data_baked)
-  # Return results
-  bind_cols(predictions, test_data)
-}) %>% 
+quantiles.cv <- folds |> 
+  arrange(id) |> 
+  pull(splits) |> 
+  map(function(spliti) {
+    train_data <- analysis(spliti)
+    test_data <- assessment(spliti)
+    # Fit QRF model
+    set.seed(345)
+    qrf_wf <- workflow_TD_uncertainty %>% 
+      fit(train_data)
+    # Make predictions
+    test_data_baked <- workflows::extract_recipe(qrf_wf) %>% 
+      bake(test_data)
+    predictions <- quant_predict(qrf_wf$fit$fit$fit, test_data_baked)
+    # Return results
+    bind_cols(predictions, test_data)
+  }) %>% 
   bind_rows()
 
 # Evaluate prediction intervals

### 2025-06-25 ms/figures.R

@@ -634,17 +634,33 @@ library(tidyverse)
 library(patchwork)
 
 orskog <- read_csv("output/calibrationplot.csv")
+orskogQuantiles <- sf::st_read("output/modeling.gpkg", layer = "quantiles.cv") |> 
+  sf::st_drop_geometry() |> 
+  select(coverage, .pred_lower, .pred_upper, depth_cm) |> 
+  as_tibble()
+all(orskog$depth_cm == orskogQuantiles$depth_cm) # Check that depth_cm matches
+orskog <- bind_cols(orskog, select(orskogQuantiles, -depth_cm))
+
 skrim <- read_csv("output/Skrim/calibrationplot.csv")
+skrimQuantiles <- sf::st_read("output/Skrim/modeling.gpkg", layer = "quantiles.cv") |> 
+  sf::st_drop_geometry() |> 
+  select(coverage, .pred_lower, .pred_upper, depth_cm) |> 
+  as_tibble()
+all(skrim$depth_cm == skrimQuantiles$depth_cm) # Check that depth_cm matches
+skrim <- bind_cols(skrim, select(skrimQuantiles, -depth_cm))
 
 probs <- c(0.25, 0.5, 0.75)
 
 scatter1 <- ggplot(orskog, aes(x = depth_cm, y = .pred)) +
-  geom_point(size = 0.8, alpha=0.5) +
+  geom_segment(aes(x = depth_cm, xend = depth_cm, 
+                   y = .pred_lower, yend = .pred_upper), 
+               color = "grey70", alpha = 0.2, linewidth = 0.2) +
+  geom_point(size = 0.7, alpha=0.5) +
   geom_smooth(method = 'loess', formula= y ~ x, se = FALSE) +
   geom_abline(intercept = 0, slope = 1, linetype = 2, color = "red") +
   tune::coord_obs_pred(ratio = 1) +
   labs(x = "Observed depth (cm)", y = "Predicted depth (cm)") +
-  theme_bw()
+  theme_classic()
 dens <- density(orskog$depth_cm)
 df <- data.frame(x = dens$x, y = dens$y)
 quantiles <- quantile(orskog$depth_cm, prob=probs)
@@ -656,12 +672,15 @@ density1 <- ggplot(df, aes(x,y)) +
   theme_void()
 
 scatter2 <- ggplot(skrim, aes(x = depth_cm, y = .pred)) +
-  geom_point(size = 0.8, alpha=0.5) +
[…]

### 2025-06-25 ms/ms.Rmd
address RC1.2, RC1.3, RC2.2

@@
 We also thank Vikas Baranwal from the Geological Survey of Norway for helping us access the radiometric data from Skrim.
-This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, © Kartverket, (2) *Norge digitalt* license, Norwegian Institute of Bioeconomy Research (NIBIO), © Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway.
-Large language models have been used during the drafting and editing of this manuscript, with author oversight.
+This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, \copyright Kartverket, (2) *Norge digitalt* license, Norwegian Institute of Bioeconomy Research (NIBIO), \copyright Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway.
+Large language models have been used to expedite coding of the analysis, with author oversight.
+Large language models have also been used during the drafting and editing of this manuscript, with author oversight.
 We maintain full responsibility for the scientific output, as per the European Commission's *Living guidelines on the responsible use of generative AI in research*.
@@
 Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth in some studies [e.g.
-@rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024].
+@rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024; @pohjankukkaDigitalMappingPeat2025].
 Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic predictors and the resolution of the peat depth analysis.
@@
 
-To our knowledge, this is the first study to predict peat depth from airborne radiometric data using machine learning algorithms. 
-Where airborne radiometric data have previously been used to predict peat depth, it has been through modeling techniques less suited for prediction and spatial extrapolation [e.g., @keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @siemonAirborneElectromagneticRadiometric2020].
-Where machine learning algorithms have previously been used on airborne radiometric data, it has been to predict peat extent rather than depth [e.g., @olearyDigitalSoilMapping2022].
+To our knowledge, this is the first study to predict continuous peat depth from airborne radiometric data using machine learning. 
+Where airborne radiometric data have previously been used to predict continuous peat depth, it has been through modeling techniques less suited for prediction and spatial extrapolation [e.g., @keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @siemonAirborneElectromagneticRadiometric2020].
+Where machine learning has previously been used with airborne radiometric data, it has been to predict peat extent or depth classes rather than continuous depth [e.g., @olearyDigitalSoilMapping2022; @pohjankukkaDigitalMappingPeat2025].
 
@@
 To fit the scale of the maps, the land cover shown here is from the AR50 national land resource dataset, which has simplified geometry with respect to the AR5 dataset that is used in the study.
-Terrain is visualized with light orange contour lines at intervals of 100 m, and a hillshade with slope 45° and azimuth 225°.
+Terrain is visualized with light orange contour lines at intervals of 100 m, and a hillshade with slope $45^\\circ$ and azimuth $225^\\circ$.
 AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (https://data.norge.no/nlod/en/1.0) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/)."}
@@
 It is almost without human infrastructure, dominated by conifer forest, and borders on a nature reserve.
-The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).
+The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is $10.8^\circ$ (IQR $4.6-15.1^\circ$).
 
@@
 It is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
-Its mean elevation is 211 m above sea level (range 0--807 m, IQR 73--310 m), and its mean slope at 10 m resolution is 13.0° (IQR 4.7--18.3°).
+Its mean elevation is 211 m above sea level (range 0--807 m, IQR 73--310 m), and its mean slope at 10 m resolution is $13.0^\circ$ (IQR $4.7-18.3^\circ$).
 
@@
 
-We chose a spatial resolution of 10 m for our predictors, depth sampling, and modelling.
+We chose a spatial resolution of 10 m for our predictors, depth sampling, and modeling.
 We considered this a reasonable compromise between DTM resolution (1 m) and small peatlands on the one hand, and airborne radiometric resolution (50 m) on the other hand.
@@
 
-The radiometric surveys were conduced with 75--80 m average flight altitude, 88–108 \unit{km\,h^{-1}} average flight speed, and 200 m flight line spacing (Table \ref{tab:radSurveys}).
+The radiometric surveys were conducted with 75--80 m average flight altitude, 88–108 \unit{km\,h^{-1}} average flight speed, and 200 m flight line spacing (Table \ref{tab:radSurveys}).
 Spectrometer count rates were calibrated to known concentrations of potassium, thorium, and uranium in mobile pads.
@@
 We measured peat depths at Skrimfjella in August 2020 and at Ørskogfjellet in August 2023.
-At both sites we used manual probing as the primary method of measuring peat depth, and ground-penetrating radar (GPR) as a secondary method.
-That is: peat depth was always measured by probing, and also by GPR in a subset of cells that partly overlapped with the probed cells
+At both sites, we used manual probing as the primary method of measuring peat depth and ground-penetrating radar (GPR) as a secondary method.
+That is: peat depth was always measured by probing, and also by GPR in a subset of cells that partly overlapped with the probed cells.
 We chose this approach because probing is a fast and reliable method for point measurements, while GPR can provide higher lateral density of data in the same amount of time [@parryEvaluatingApproachesEstimating2014].
@@
 
-### Modelling approach
+### Modeling approach
 
@@
 ```{r map-distribution, out.height="81%", echo = FALSE, fig.cap = "Spatial and statistical distributions of mean peat depth in 10 m cells at Skrimfjella, N = 372 (a) and Ørskogfjellet, N = 1878 (b).
-Terrain is visualized with a hillshade with slope 45° and azimuth 225°.
+Terrain is visualized with a hillshade with slope $45^\\circ$ and azimuth $225^\\circ$.
 Terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/)."}
@@
 
-```{r calPlots, echo = FALSE, out.width="100%", fig.cap = "Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions and 90 % prediction intervals (grey veritical lines) from spatial cross-validation.
-Points are transparent to show overlap.
+```{r calPlots, echo = FALSE, out.width="100%", fig.cap = "Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions and 90 % prediction intervals (grey vertical lines) from spatial cross-validation.
+Points have transparency to show overlap.
 Blue lines are local polynomial regressions and the red dashed line in each panel shows the 1:1 line.
@@
 
-Ground-based measurements also have uncertainty [@parryEvaluatingApproachesEstimating2014], but our data show that this error was smaller than the variation in peat depth over short distances (e.g.
-100 m).
-Independent probing and GPR measurements up to 2 m apart differed by an average of 29--42 cm (Fig. \@ref(fig:GPRwavevelocity)).
-By comparison, the average variation among measurements 100 m apart was 70--106 cm (Fig. \@ref(fig:semivariograms)).
-We also note that cell-level measurement error that is correlated with predictors (e.g.
-underestimated depth at high Multi-Resolution Valley Bottom Flatness), will go undetected by cross-validation and does not account for the limited accuracy of our models.
-Ultimately, the effect of measurement error falls mostly outside the scope of DSM, since it can only be evaluated with higher-quality, independent data.
-
 Although model performance was limited, we improved on the best available map of peat depth (DMK depth class), which is based on field measurements only.
 This highlights the general value of remotely sensed data, whose complete coverage can improve maps even when their association with the variable of interest is weak [@mulderUseRemoteSensing2011].
-Since remotely sensed data are widely available, improvements to soil maps as shown here are readily achievable at low cost.
-This point is recognized and reflected in the rise of DSM [@minasnyDigitalMappingPeatlands2019].
+Since remotely sensed data are widely available, improvements to soil maps as shown here are readily achievable at low cost [@minasnyDigitalMappingPeatlands2019].
 
 DMK peat depth classes were a worse predictor of peat depth than our models even though we calibrated them with the same data (Fig. \@ref(fig:modelMetrics)).
-That is: we gave DMK and the DSM equal footing for a fair comparison.
 If we had taken the DMK peat depth classes at face value and assumed depths according to their class definitions (<1 m, >1 m), they would have performed worse and the advantage of the DSM would be greater.
 The advantage of the DSM was not large in absolute terms (9 cm and 21 cm improvements in mean absolute error), but it explained much more of the variation in depth (improvements in R^2^ of 0.16 and 0.19).
-We attribute this result to the poor spatial and thematic resolution of DMK peat depth, which precludes a robust correlation with peat depths varying from 0--8 m at fine spatial scales.
+We attribute this result to the poor spatial and thematic resolution of DMK peat depth, which precludes a robust correlation with peat depths.
 
-The performance of our terrain- and radiometric-based maps could have been improved with a spatially explicit mapping approach like regression kriging.
-By ignoring spatial autocorrelation in peat depth (i.e. the n in *scorpan*), we have not extracted all of the information about peat depth in the study area out of the training data.
-If we were to put our DSM approach into production for published maps, we would harness the spatial component, but the actual maps for Ørskogfjellet and Skrimfjella are not of primary interest in this study.
-Moreover, residuals of the RF predictions showed weak spatial structure at Skrimfjella, and only up to a range of 150 m at Ørskogfjellet.
+The accuracy of our mapping could have been improved with a spatially explicit approach like regression kriging.
+However, residuals of the RF predictions showed weak spatial structure at Skrimfjella, and only up to a range of 150 m at Ørskogfjellet.
 Therefore, the improvement from regression kriging would be small overall and limited to parts of the maps close to measurements.
@@
 
-Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth but generally produced better or comparable error magnitude (R^2^ vs.
-mean absolute error or root mean square error; @wadouxIntegratedApproachEvaluation2022).
+Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth but generally produced better or comparable error magnitude [R^2^ vs.
+mean absolute error or root mean square error, @wadouxIntegratedApproachEvaluation2022].
 It is important to keep in mind that differences in peat depth distributions, spatial scales, and evaluation methods make direct performance comparisons precarious. 
@@
 [-@kogantiMappingPeatDepth2023] and our study (besides spatial scale) is that they measured radiometrics on the ground, rather than using airborne survey data.
-Thus, the footprint of their detector was much smaller and they could capture variation in radiation at finer scales.
+Thus, the footprint of their detector was smaller and they captured variation in radiation at finer scales.
 
@@
 This is not unexpected for the right-skewed distributions of peat depth, but it has management implications. 
+For example, it makes the maps more suited for "red-lighting" than "green-lighting" peatland conversion (assuming depth is the determinative factor). 
 Identifying the deepest peats will require additional field work in candidate areas, which could be defined by an upper quantile of predicted depth.
-That makes the maps more suited for "red-lighting" than "green-lighting" peatland conversion, for example. 
-Where it does not prohibit conversion (i.e. predicts shallow peat), a ground survey should be done before conversion is allowed (assuming depth is the only consideration). 
-Although this recommendation aligns with the precautionary principle, here we make it on technical grounds based on the maps’ characteristics.
 
@@
 Radiometric data had little to no predictive value at either site (poor performance of *radiometric* configuration), although they did contribute to the best model at Skrimfjella (marginal improvement in *all predictors* configuration compared to *terrain + DMK*).
-These results contrast with earlier studies that found that radiometrics were useful predictors of peat depth [@keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @kogantiMappingPeatDepth2023].
+These results contrast with earlier studies that found that radiometrics were useful predictors of peat depth [@keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @kogantiMappingPeatDepth2023; @pohjankukkaDigitalMappingPeat2025].
 The bedrock is more homogeneous at Ørskogfjellet than at Skrimfjella, so uneven radiogenesis is not a good explanation for the higher predictive value of radiometric variables at Skrimfjella [@beamishEnvironmentalRadioactivityUK2014; @reinhardtGammaraySpectrometryVersatile2019].
@@
 Soil moisture, bulk density, and porosity, however, are not easily measured across landscape scales and were assumed to be homogeneous.
-Uneven snow cover and air moisture during the Ørskogfjellet radiometric survey may also have masked the soil signal these data, as Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015] reports large variation in weather conditions.
+Uneven snow cover and air moisture during the Ørskogfjellet radiometric survey may also have masked the soil signal in these data, as Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015] reports large variation in weather conditions.
 If maps of these other physical parameters at the time of the radiometric survey were available and included in the model, the predictive value of radiometric data might improve, but this is not a practical solution for digital soil mapping of peat depth.
 
-We do not believe that the poor predictive value of the radiometric data in this study was caused by fully attenuated radioactivity — at least not in large part. 
+We do not believe that the poor predictive value of the radiometric data in this study was caused primarily by full attenuation of radioactivity. 
 The RF algorithm’s flexibility means that radiometrics could be used for shallower peats if they provided predictive value for that part of the depth distribution, but there is no evidence of that in our results.
 In the partial dependence plot of uranium ground concentration at Skrimfjella, the expected negative relationship between depth and uranium ground concentration cannot be found by ignoring the left (highly attenuated) side of the distribution.
-More than a quarter of the peats in each our study areas were less than 60 cm deep, and full attenuation is unlikely for these [@beamishGammaRayAttenuation2013].
+More than a quarter of the peats in each of our study areas were less than 60 cm deep, and full attenuation is unlikely for these [@beamishGammaRayAttenuation2013].
 
@@
 At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables (Fig. \@ref(fig:modelMetrics)).
-Elevation was the most important predictor at both sites, and peat depth showed non-monotonic responses to changes in elevation (Figs. \@ref(fig:varImp)-\@ref(fig:pdps)). 
+This is consistent with the findings of Pohjankukka et al. [-@pohjankukkaDigitalMappingPeat2025], who mapped peat depth classes at 50 m resolution across Finland.
+
+At both our sites, elevation was the most important predictor, and peat depth showed non-monotonic responses to changes in elevation (Figs. \@ref(fig:varImp)-\@ref(fig:pdps)).
 We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g.
@@
 This hypothesis is supported by similar findings at a coarser scale in the Hudson Bay Lowlands, where a strong positive relationship between peat depth and distance from the coast can be explained by isostatic uplift and time since peat initiation [@liPeatDepthCarbon2025].
+The same has been reported for Finland [@pohjankukkaDigitalMappingPeat2025]. 
 We cannot evaluate this effect at Skrimfjella, where the marine limit is below our study area (at 175 m.a.s.l.).
@@
 Notably, slope was inferior to (at Ørskogfjellet) or highly correlated with (at Skrimfjella) these hydrological indices. 
-Mappers of peat depth should not assume that slope is the best predictor in its class, despite its prevalence in the literature. 
+Mappers of peat depth should not assume that slope is the best terrain-derived predictor, despite its prevalence in the literature [@pohjankukkaDigitalMappingPeat2025]. 
 Wetter locations (high Topographic Wetness Index and low Depth-to-Water) were generally associated with deeper peat, but these relationships were not as strong or consistent as with Multi-Resolution Valley Bottom Flatness (Figs.
@@
 The optimal scale for Topographic Wetness Index and Depth-to-Water varied, and likely depends on both the dominant peat formation processes and the typical size of peatland features in a landscape. 
-Including multiple scales of these variables allows the model to capture different hydrological mechanisms operating at different spatial scales.
+Including multiple scales of these variables allows the model to capture different hydrological mechanisms or patterns operating at different spatial scales.
 
@@
 
-High quality DTMs are available for mainland Norway, which makes peat depth measurements the critical training data need for DSM. 
+High-quality DTMs are available for mainland Norway, which makes peat depth measurements the critical training data needed for DSM. 
 Better infrastructure to make depth measurements findable, accessible, interoperable, and reusable would help DSM and other applications. 
@@
 
-The best models achieved mean absolute errors of 56--60 cm against mean depths of 119--126 cm, and explained approximately one-third of the variation in peat depth across landscapes with with aggregate peatland areas of 1.5--15.3 km^2^.
+The best models achieved mean absolute errors of 56--60 cm against mean depths of 119--126 cm, and explained approximately one-third of the variation in peat depth across landscapes with aggregate peatland areas of 1.5--15.3 km^2^.
 Though field measurements remain necessary for local, detailed assessments of peat depth, digital soil maps at 10 m resolution can provide valuable information for landscape-scale planning and regional carbon assessments.
@@
 
-To choose 160 locations, we performed feature space coverage sampling, implemented using the *kmeans* function in base R and the *rdist* function in the the *fields* package (v14.2).
+To choose 160 locations, we performed feature space coverage sampling, implemented using the *kmeans* function in base R and the *rdist* function in the *fields* package (v14.2).
 Feature space coverage sampling chooses locations that are closest to cluster centers in standardized predictor space [@brusSamplingDigitalSoil2019].
@@
 
-We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real time kinematic (Ørskogfjellet) global navigation satellite system (GNSS) receivers.
+We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real-time kinematic (Ørskogfjellet) global navigation satellite system (GNSS) receivers.
 We dampened the effect of outlying measurements by probing three times at each location [@parryEvaluatingApproachesEstimating2014], at the vertices of a centered triangle with 2 m (Skrimfjella) or 4.5 m (Ørskogfjellet) sides.

### 2025-06-25 ms/ms.tex
address RC1.2, RC1.3, RC2.2

@@
 Peat depth also changes with elevation in many contexts, but with inconsistent directionality \citep[e.g.][]{holdenEstimatingCarbonStock2011, parryMethodModellingPeat2012, rudiyantoDigitalMappingCosteffective2016, rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024}.
-Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth in some studies \citep[e.g.][]{rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024}.
+Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth in some studies \citep[e.g.][]{rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024, pohjankukkaDigitalMappingPeat2025}.
 Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic predictors and the resolution of the peat depth analysis.
@@
 
-To our knowledge, this is the first study to predict peat depth from airborne radiometric data using machine learning algorithms.
-Where airborne radiometric data have previously been used to predict peat depth, it has been through modeling techniques less suited for prediction and spatial extrapolation \citep[e.g.,][]{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, siemonAirborneElectromagneticRadiometric2020}.
-Where machine learning algorithms have previously been used on airborne radiometric data, it has been to predict peat extent rather than depth \citep[e.g.,][]{olearyDigitalSoilMapping2022}.
+To our knowledge, this is the first study to predict continuous peat depth from airborne radiometric data using machine learning.
+Where airborne radiometric data have previously been used to predict continuous peat depth, it has been through modeling techniques less suited for prediction and spatial extrapolation \citep[e.g.,][]{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, siemonAirborneElectromagneticRadiometric2020}.
+Where machine learning has previously been used with airborne radiometric data, it has been to predict peat extent or depth classes rather than continuous depth \citep[e.g.,][]{olearyDigitalSoilMapping2022, pohjankukkaDigitalMappingPeat2025}.
 
@@
 \ref{fig:map-sites}b).
-These sites were chosen because they were covered by radiometric data from airborne surveys, had relatively little built-up area, and were accessible by road.\\
+These sites were chosen because they were covered by radiometric data from airborne surveys, had relatively little built-up area, and were accessible by road.
 Ørskogfjellet was additionally chosen for its relatively uniform bedrock geology, which should improve the peat depth signal in radiometric data, all else being equal \citep{minasnyDigitalMappingPeatlands2019}.
@@
 To fit the scale of the maps, the land cover shown here is from the AR50 national land resource dataset, which has simplified geometry with respect to the AR5 dataset that is used in the study.
-Terrain is visualized with light orange contour lines at intervals of 100 m, and a hillshade with slope 45° and azimuth 225°.
+Terrain is visualized with light orange contour lines at intervals of 100 m, and a hillshade with slope $45^\circ$ and azimuth $225^\circ$.
 AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (https://data.norge.no/nlod/en/1.0) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/).}\label{fig:map-sites}
@@
 It is almost without human infrastructure, dominated by conifer forest, and borders on a nature reserve.
-The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is 10.8° (IQR 4.6--15.1°).
+The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is \(10.8^\circ\) (IQR \(4.6-15.1^\circ\)).
 
@@
 It is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
-Its mean elevation is 211 m above sea level (range 0--807 m, IQR 73--310 m), and its mean slope at 10 m resolution is 13.0° (IQR 4.7--18.3°).
+Its mean elevation is 211 m above sea level (range 0--807 m, IQR 73--310 m), and its mean slope at 10 m resolution is \(13.0^\circ\) (IQR \(4.7-18.3^\circ\)).
 
@@
 
-The radiometric surveys were conduced with 75--80 m average flight altitude, 88--108 \unit{km\,h^{-1}} average flight speed, and 200 m flight line spacing (Table \ref{tab:radSurveys}).
+The radiometric surveys were conducted with 75--80 m average flight altitude, 88--108 \unit{km\,h^{-1}} average flight speed, and 200 m flight line spacing (Table \ref{tab:radSurveys}).
 Spectrometer count rates were calibrated to known concentrations of potassium, thorium, and uranium in mobile pads.
@@
 We measured peat depths at Skrimfjella in August 2020 and at Ørskogfjellet in August 2023.
-At both sites we used manual probing as the primary method of measuring peat depth, and ground-penetrating radar (GPR) as a secondary method.
-That is: peat depth was always measured by probing, and also by GPR in a subset of cells that partly overlapped with the probed cells
+At both sites, we used manual probing as the primary method of measuring peat depth and ground-penetrating radar (GPR) as a secondary method.
+That is: peat depth was always measured by probing, and also by GPR in a subset of cells that partly overlapped with the probed cells.
 We chose this approach because probing is a fast and reliable method for point measurements, while GPR can provide higher lateral density of data in the same amount of time \citep{parryEvaluatingApproachesEstimating2014}.
@@
 
-\subsubsection{Modelling approach}
+\subsubsection{Modeling approach}
 
@@
 \includegraphics[height=0.81\textheight]{figures/map-distribution} \caption{Spatial and statistical distributions of mean peat depth in 10 m cells at Skrimfjella, N = 372 (a) and Ørskogfjellet, N = 1878 (b).
-Terrain is visualized with a hillshade with slope 45° and azimuth 225°.
+Terrain is visualized with a hillshade with slope $45^\circ$ and azimuth $225^\circ$.
 Terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/).}\label{fig:map-distribution}
@@
 \begin{figure}
-\includegraphics[width=1\linewidth]{figures/calibration_plots} \caption{Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions and 90 \% prediction intervals (grey veritical lines) from spatial cross-validation.
-Points are transparent to show overlap.
+\includegraphics[width=1\linewidth]{figures/calibration_plots} \caption{Calibration plots for the best-performing models at Skrimfjella (a) and Ørskogfjellet (b), with predictions and 90 \% prediction intervals (grey vertical lines) from spatial cross-validation.
+Points have transparency to show overlap.
 Blue lines are local polynomial regressions and the red dashed line in each panel shows the 1:1 line.
@@
 
-Ground-based measurements also have uncertainty \citep{parryEvaluatingApproachesEstimating2014}, but our data show that this error was smaller than the variation in peat depth over short distances (e.g.~100 m).
-Independent probing and GPR measurements up to 2 m apart differed by an average of 29--42 cm (Fig. \ref{fig:GPRwavevelocity}).
-By comparison, the average variation among measurements 100 m apart was 70--106 cm (Fig. \ref{fig:semivariograms}).
-We also note that cell-level measurement error that is correlated with predictors (e.g.~underestimated depth at high Multi-Resolution Valley Bottom Flatness), will go undetected by cross-validation and does not account for the limited accuracy of our models.
-Ultimately, the effect of measurement error falls mostly outside the scope of DSM, since it can only be evaluated with higher-quality, independent data.
-
 Although model performance was limited, we improved on the best available map of peat depth (DMK depth class), which is based on field measurements only.
 This highlights the general value of remotely sensed data, whose complete coverage can improve maps even when their association with the variable of interest is weak \citep{mulderUseRemoteSensing2011}.
-Since remotely sensed data are widely available, improvements to soil maps as shown here are readily achievable at low cost.
-This point is recognized and reflected in the rise of DSM \citep{minasnyDigitalMappingPeatlands2019}.
+Since remotely sensed data are widely available, improvements to soil maps as shown here are readily achievable at low cost \citep{minasnyDigitalMappingPeatlands2019}.
 
 DMK peat depth classes were a worse predictor of peat depth than our models even though we calibrated them with the same data (Fig. \ref{fig:modelMetrics}).
-That is: we gave DMK and the DSM equal footing for a fair comparison.
 If we had taken the DMK peat depth classes at face value and assumed depths according to their class definitions (\textless1 m, \textgreater1 m), they would have performed worse and the advantage of the DSM would be greater.
 The advantage of the DSM was not large in absolute terms (9 cm and 21 cm improvements in mean absolute error), but it explained much more of the variation in depth (improvements in R\textsuperscript{2} of 0.16 and 0.19).
-We attribute this result to the poor spatial and thematic resolution of DMK peat depth, which precludes a robust correlation with peat depths varying from 0--8 m at fine spatial scales.
-
-The performance of our terrain- and radiometric-based maps could have been improved with a spatially explicit mapping approach like regression kriging.
-By ignoring spatial autocorrelation in peat depth (i.e.~the n in \emph{scorpan}), we have not extracted all of the information about peat depth in the study area out of the training data.
-If we were to put our DSM approach into production for published maps, we would harness the spatial component, but the actual maps for Ørskogfjellet and Skrimfjella are not of primary interest in this study.
-Moreover, residuals of the RF predictions showed weak spatial structure at Skrimfjella, and only up to a range of 150 m at Ørskogfjellet.
+We attribute this result to the poor spatial and thematic resolution of DMK peat depth, which precludes a robust correlation with peat depths.
+
+The accuracy of our mapping could have been improved with a spatially explicit approach like regression kriging.
+However, residuals of the RF predictions showed weak spatial structure at Skrimfjella, and only up to a range of 150 m at Ørskogfjellet.
 Therefore, the improvement from regression kriging would be small overall and limited to parts of the maps close to measurements.
@@
 
-Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth but generally produced better or comparable error magnitude (R\textsuperscript{2} vs.~mean absolute error or root mean square error; \citet{wadouxIntegratedApproachEvaluation2022}).
+Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth but generally produced better or comparable error magnitude \citep[R\textsuperscript{2} vs.~mean absolute error or root mean square error,][]{wadouxIntegratedApproachEvaluation2022}.
 It is important to keep in mind that differences in peat depth distributions, spatial scales, and evaluation methods make direct performance comparisons precarious.
@@
 \citeyearpar{kogantiMappingPeatDepth2023} and our study (besides spatial scale) is that they measured radiometrics on the ground, rather than using airborne survey data.
-Thus, the footprint of their detector was much smaller and they could capture variation in radiation at finer scales.
+Thus, the footprint of their detector was smaller and they captured variation in radiation at finer scales.
 
@@
 This is not unexpected for the right-skewed distributions of peat depth, but it has management implications.
+For example, it makes the maps more suited for ``red-lighting'' than ``green-lighting'' peatland conversion (assuming depth is the determinative factor).
 Identifying the deepest peats will require additional field work in candidate areas, which could be defined by an upper quantile of predicted depth.
-That makes the maps more suited for ``red-lighting'' than ``green-lighting'' peatland conversion, for example.
-Where it does not prohibit conversion (i.e.~predicts shallow peat), a ground survey should be done before conversion is allowed (assuming depth is the only consideration).
-Although this recommendation aligns with the precautionary principle, here we make it on technical grounds based on the maps' characteristics.
 
@@
 Radiometric data had little to no predictive value at either site (poor performance of \emph{radiometric} configuration), although they did contribute to the best model at Skrimfjella (marginal improvement in \emph{all predictors} configuration compared to \emph{terrain + DMK}).
-These results contrast with earlier studies that found that radiometrics were useful predictors of peat depth \citep{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, kogantiMappingPeatDepth2023}.
+These results contrast with earlier studies that found that radiometrics were useful predictors of peat depth \citep{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, kogantiMappingPeatDepth2023, pohjankukkaDigitalMappingPeat2025}.
 The bedrock is more homogeneous at Ørskogfjellet than at Skrimfjella, so uneven radiogenesis is not a good explanation for the higher predictive value of radiometric variables at Skrimfjella \citep{beamishEnvironmentalRadioactivityUK2014, reinhardtGammaraySpectrometryVersatile2019}.
@@
 Soil moisture, bulk density, and porosity, however, are not easily measured across landscape scales and were assumed to be homogeneous.
-Uneven snow cover and air moisture during the Ørskogfjellet radiometric survey may also have masked the soil signal these data, as Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015} reports large variation in weather conditions.
+Uneven snow cover and air moisture during the Ørskogfjellet radiometric survey may also have masked the soil signal in these data, as Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015} reports large variation in weather conditions.
 If maps of these other physical parameters at the time of the radiometric survey were available and included in the model, the predictive value of radiometric data might improve, but this is not a practical solution for digital soil mapping of peat depth.
 
-We do not believe that the poor predictive value of the radiometric data in this study was caused by fully attenuated radioactivity --- at least not in large part.
+We do not believe that the poor predictive value of the radiometric data in this study was caused primarily by full attenuation of radioactivity.
 The RF algorithm's flexibility means that radiometrics could be used for shallower peats if they provided predictive value for that part of the depth distribution, but there is no evidence of that in our results.
 In the partial dependence plot of uranium ground concentration at Skrimfjella, the expected negative relationship between depth and uranium ground concentration cannot be found by ignoring the left (highly attenuated) side of the distribution.
-More than a quarter of the peats in each our study areas were less than 60 cm deep, and full attenuation is unlikely for these \citep{beamishGammaRayAttenuation2013}.
+More than a quarter of the peats in each of our study areas were less than 60 cm deep, and full attenuation is unlikely for these \citep{beamishGammaRayAttenuation2013}.
 
@@
 At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables (Fig. \ref{fig:modelMetrics}).
-Elevation was the most important predictor at both sites, and peat depth showed non-monotonic responses to changes in elevation (Figs. \ref{fig:varImp}-\ref{fig:pdps}).
+This is consistent with the findings of Pohjankukka et al. \citeyearpar{pohjankukkaDigitalMappingPeat2025}, who mapped peat depth classes at 50 m resolution across Finland.
+
+At both our sites, elevation was the most important predictor, and peat depth showed non-monotonic responses to changes in elevation (Figs. \ref{fig:varImp}-\ref{fig:pdps}).
 We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g.~via climate) to explain the observed patterns.
@@
 This hypothesis is supported by similar findings at a coarser scale in the Hudson Bay Lowlands, where a strong positive relationship between peat depth and distance from the coast can be explained by isostatic uplift and time since peat initiation \citep{liPeatDepthCarbon2025}.
+The same has been reported for Finland \citep{pohjankukkaDigitalMappingPeat2025}.
 We cannot evaluate this effect at Skrimfjella, where the marine limit is below our study area (at 175 m.a.s.l.).
@@
 Notably, slope was inferior to (at Ørskogfjellet) or highly correlated with (at Skrimfjella) these hydrological indices.
-Mappers of peat depth should not assume that slope is the best predictor in its class, despite its prevalence in the literature.
+Mappers of peat depth should not assume that slope is the best terrain-derived predictor, despite its prevalence in the literature \citep{pohjankukkaDigitalMappingPeat2025}.
 Wetter locations (high Topographic Wetness Index and low Depth-to-Water) were generally associated with deeper peat, but these relationships were not as strong or consistent as with Multi-Resolution Valley Bottom Flatness (Figs.
@@
 The optimal scale for Topographic Wetness Index and Depth-to-Water varied, and likely depends on both the dominant peat formation processes and the typical size of peatland features in a landscape.
-Including multiple scales of these variables allows the model to capture different hydrological mechanisms operating at different spatial scales.
+Including multiple scales of these variables allows the model to capture different hydrological mechanisms or patterns operating at different spatial scales.
 
@@
 
-High quality DTMs are available for mainland Norway, which makes peat depth measurements the critical training data need for DSM.
+High-quality DTMs are available for mainland Norway, which makes peat depth measurements the critical training data needed for DSM.
 Better infrastructure to make depth measurements findable, accessible, interoperable, and reusable would help DSM and other applications.
@@
 
-The best models achieved mean absolute errors of 56--60 cm against mean depths of 119--126 cm, and explained approximately one-third of the variation in peat depth across landscapes with with aggregate peatland areas of 1.5--15.3 km\textsuperscript{2}.
+The best models achieved mean absolute errors of 56--60 cm against mean depths of 119--126 cm, and explained approximately one-third of the variation in peat depth across landscapes with aggregate peatland areas of 1.5--15.3 km\textsuperscript{2}.
 Though field measurements remain necessary for local, detailed assessments of peat depth, digital soil maps at 10 m resolution can provide valuable information for landscape-scale planning and regional carbon assessments.
@@
 
-To choose 160 locations, we performed feature space coverage sampling, implemented using the \emph{kmeans} function in base R and the \emph{rdist} function in the the \emph{fields} package (v14.2).
+To choose 160 locations, we performed feature space coverage sampling, implemented using the \emph{kmeans} function in base R and the \emph{rdist} function in the \emph{fields} package (v14.2).
 Feature space coverage sampling chooses locations that are closest to cluster centers in standardized predictor space \citep{brusSamplingDigitalSoil2019}.
@@
 
-We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real time kinematic (Ørskogfjellet) global navigation satellite system (GNSS) receivers.
+We navigated to the centers of the raster cells in our samples using handheld (Skrimfjella) or real-time kinematic (Ørskogfjellet) global navigation satellite system (GNSS) receivers.
 We dampened the effect of outlying measurements by probing three times at each location \citep{parryEvaluatingApproachesEstimating2014}, at the vertices of a centered triangle with 2 m (Skrimfjella) or 4.5 m (Ørskogfjellet) sides.
@@
 We also thank Vikas Baranwal from the Geological Survey of Norway for helping us access the radiometric data from Skrim.
-This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, © Kartverket, (2) \emph{Norge digitalt} license, Norwegian Institute of Bioeconomy Research (NIBIO), © Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway.
-Large language models have been used during the drafting and editing of this manuscript, with author oversight.
+This work contains data under the following licenses: (1) Creative Commons Attribution 4.0 International, \copyright Kartverket, (2) \emph{Norge digitalt} license, Norwegian Institute of Bioeconomy Research (NIBIO), \copyright Geovekst, and (3) the Norwegian License for Public Data (NLOD), made available by the Geological Survey of Norway.
+Large language models have been used to expedite coding of the analysis, with author oversight.
+Large language models have also been used during the drafting and editing of this manuscript, with author oversight.
 We maintain full responsibility for the scientific output, as per the European Commission's \emph{Living guidelines on the responsible use of generative AI in research}.

### 2025-06-25 ms/submission/AuthorComments-egusphere-2025-1046.qmd
address RC1.2, RC1.3, RC2.2

@@
 
+Thank you for your positive comments on the article and for your suggestions to improve its readability.
+We have taken your advice to review the article for conciseness and have made several changes to reduce its length, while retaining the key details of the modelling methodology.
+In particular, we have removed the sections on peatland extent mapping, which was not central to the main focus of the study, and we have shortened some of the other sections as well.
+We have moved some of the more detailed descriptions of the methodology to the appendix, to make the main text more accessible and focused.
+
 ### Comment 1.3
@@
 > Additionally, there were several areas where the language used was quite casual for a scientific article. Several are highlighted under minor considerations below.
+
+Thank you for this advice and the specific examples you provided.
+We have gone through the manuscript and revised the language to make it more formal and scientific, while aiming to retain clarity and accessibility.
 
@@
 Considering the title is stating terrain being “better” than radiometrics, and given the emerging understanding of the use of radiometric data in peat land mapping, there is are some fundamental errors in the methods presented, which may be biasing the conclusion alluded to in the title.
+
+Thank you for raising this concern.
+We address the specific comments on this theme below, and believe that our responses show why we disagree that there are fundamental errors in the methods.
+For the same reasons outlined below, we do not think that the conclusion the title alludes to is biased.
 
@@
 
-## Specifics Comments
+Thank you for your positive comments on the research and title, and for your suggestions to improve the writing style.
+We have gone through the manuscript and revised the language to make it more formal and scientific, while aiming to retain clarity and accessibility.
+We have also tried to reduce jargon where possible, while retaining the necessary technical terms.
+
+## Specific Comments
 

### 2025-06-25 ms/figures.R

@@ -655,7 +655,7 @@ scatter1 <- ggplot(orskog, aes(x = depth_cm, y = .pred)) +
   geom_segment(aes(x = depth_cm, xend = depth_cm, 
                    y = .pred_lower, yend = .pred_upper), 
                color = "grey70", alpha = 0.2, linewidth = 0.2) +
-  geom_point(size = 0.7, alpha=0.5) +
+  geom_point(size = 0.7, alpha=0.5, pch = 16) +
   geom_smooth(method = 'loess', formula= y ~ x, se = FALSE) +
   geom_abline(intercept = 0, slope = 1, linetype = 2, color = "red") +
   tune::coord_obs_pred(ratio = 1) +
@@ -675,7 +675,7 @@ scatter2 <- ggplot(skrim, aes(x = depth_cm, y = .pred)) +
   geom_segment(aes(x = depth_cm, xend = depth_cm, 
                    y = .pred_lower, yend = .pred_upper), 
                color = "grey70", alpha = 0.2, linewidth = 0.2) +
-  geom_point(size = 0.7, alpha=0.5) +
+  geom_point(size = 0.7, alpha=0.5, pch = 16) +
   geom_smooth(method = 'loess', formula= y ~ x, se = FALSE) +
   geom_abline(intercept = 0, slope = 1, linetype = 2, color = "red") +
   tune::coord_obs_pred(ratio = 1) +

### 2025-06-27 ms/ms.Rmd
read through on paper

@@
   Terrain variables were better predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth. 
-  Radiometric variables showed inconsistent predictive value -- improving performance at one site while degrading it at the other.
-  The accuracy of the national map of peat depth did not measure up to any of our remote sensing models, even though it was calibrated to the same data.
+  Radiometric variables showed inconsistent and weak predictive value -- improving performance at one site while degrading it at the other.
+  Our remote sensing models had better accuracy than the national map of peat depth, even when we calibrated the national map to the same depth data.
   Still, weak relationships with remotely sensed variables made peat depth hard to predict overall.
-  Based on these findings, we conclude that digital soil mapping can improve existing, broad-scale maps of peat depth in Norway, but highly localized carbon stock assessments are best made from field measurements.
+  Based on these findings, we conclude that digital soil mapping can improve the existing, national map of peat depth in Norway, but detailed local maps are best made from tailored field measurements.
   Together, these pathways promise more accurate landscape-scale carbon stock assessments and better-informed land management policies.
@@
 Intact peatlands sequester carbon and overall produce a negative temperature forcing [@joostenRolePeatlandsClimate2016]. 
-When disturbed, often by conversion to another land use, they can produce large greenhouse gas emissions [@maGloballyRobustRelationship2022].
+When disturbed -- often by conversion to another land use -- they can produce large greenhouse gas emissions [@maGloballyRobustRelationship2022].
 
-One of the keys to the areal density of peatland carbon stocks lies in the third dimension: their depth. 
-Peat soils range from zero to over ten meters deep [@widyastutiPEATGRIDSMappingThickness2024], so deep peats comprise a large volume. 
-Their depth results from the accumulation of organic matter over thousands of years [@loiselDatabaseSynthesisNorthern2014; @joostenRolePeatlandsClimate2016]. 
+One of the keys to the areal density of peatland carbon stocks (and variation therein) lies in the third dimension: their depth. 
+Peat soils range from zero to over ten meters deep [@widyastutiPEATGRIDSMappingThickness2024]. 
+Their depth depends on the rate and duration of organic matter accumulation, potentially over thousands of years [@loiselDatabaseSynthesisNorthern2014; @joostenRolePeatlandsClimate2016]. 
 In the anoxic and acidic conditions created by a high water table, plant material decay is slightly outweighed by new growth, and the surplus carbon is laid down as peat.
-There it remains sequestered as long as the water table stays high and prevents oxidative peat decomposition.
 
 Peatlands are most common at high latitudes, and in regions with high cover they are frequently converted to human land use [@unepGlobalPeatlandsAssessment2022].
-They are attractive for agriculture and forestry because they are flat, treeless, and have developed soils.
-However, other land uses also displace peatlands. 
+They are attractive for agriculture and forestry because they are flat, treeless, and have developed soils, but other land uses also displace peatlands. 
 In Norway -- where flat lowland is relatively scarce but upward of 9 % of the land area is peatland [@brynLandCoverNorway2018; @bakkestuenDelineationWetlandAreas2023] -- lawmakers have restricted peatland afforestation and cultivation in recent decades.
@@
 
-The spatial distribution of peat depth is often overlooked in land use planning and carbon accounting because peat depth is not mapped with sufficient coverage, resolution, or accuracy [@beilmanPeatCarbonStocks2008; @hastieRisksCarbonStorage2022; @unepGlobalPeatlandsAssessment2022].
-Maps are crucial because they link high-level targets to specific management decisions, unlike spatially aggregated estimates [@oecdOECDEnvironmentalPerformance2022].
+Variation in peat depth and its spatial distribution is often overlooked in land use planning and carbon accounting because peat depth is not mapped with sufficient coverage, resolution, or accuracy [@beilmanPeatCarbonStocks2008; @hastieRisksCarbonStorage2022; @unepGlobalPeatlandsAssessment2022].
+Maps are crucial because, unlike spatially aggregated estimates, they link high-level targets to specific management decisions [@oecdOECDEnvironmentalPerformance2022].
 Maps of peat depth also make it possible to quantify the effect of specific management decisions and thereby understand how local outcomes contribute to regional and national outcomes [@oecdOECDEnvironmentalPerformance2022].
 
-Measuring peat depth on the ground by probing or ground-penetrating radar is straightforward, and a field survey can map a small area at low cost. 
-However, surveying large areas is impractical when depth varies widely over short distances (e.g.
-10 m) -- as in many peatlands [@torppaStatisticalApproachDetermining2011; @proulx-mcinnisReviewGroundpenetratingRadar2013; @henrionMappingMonitoringPeatlands2024].
+Measuring peat depth on the ground is straightforward, and a field survey can map a small area at low cost. 
+However, surveying large areas is impractical when depth varies widely over short distances (e.g., 10 m) -- as in many peatlands [@torppaStatisticalApproachDetermining2011; @proulx-mcinnisReviewGroundpenetratingRadar2013; @henrionMappingMonitoringPeatlands2024].
 A complementary approach from soil science is digital soil mapping (DSM). 
 DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest, through a statistical model. 
-This approach has grown in importance with the availability of remotely sensed data and the advancement of methods for identifying statistical patterns, especially through machine learning [@minasnyDigitalMappingPeatlands2019; @wadouxMachineLearningDigital2020].
+This approach has grown in importance with the availability of remotely sensed data and the advancement of machine learning methods [@minasnyDigitalMappingPeatlands2019; @wadouxMachineLearningDigital2020].
 
@@
 
-Studies have consistently shown relationships between peat depth and topography, although the specific patterns vary. 
-One of the most robust relationships is a negative correlation between depth and terrain slope [e.g. @holdenEstimatingCarbonStock2011; @parryMethodModellingPeat2012; @gatisMappingUplandPeat2019]. 
-Peat depth also changes with elevation in many contexts, but with inconsistent directionality [e.g.
-@holdenEstimatingCarbonStock2011; @parryMethodModellingPeat2012; @rudiyantoDigitalMappingCosteffective2016; @rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024].
-Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth in some studies [e.g.
-@rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024; @pohjankukkaDigitalMappingPeat2025].
+Studies have consistently shown relationships between peat depth and topography, though the specific patterns vary. 
+One of the most robust relationships is a negative correlation between depth and terrain slope [e.g., @holdenEstimatingCarbonStock2011; @parryMethodModellingPeat2012; @gatisMappingUplandPeat2019]. 
+Peat depth also changes with elevation in many contexts, but with inconsistent directionality [e.g., @holdenEstimatingCarbonStock2011; @parryMethodModellingPeat2012; @rudiyantoDigitalMappingCosteffective2016; @rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024].
+Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth [e.g., @rudiyantoOpenDigitalMapping2018; @kogantiMappingPeatDepth2023; @liFactorsControllingPeat2024; @pohjankukkaDigitalMappingPeat2025].
 Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic predictors and the resolution of the peat depth analysis.
@@
 Although theory suggests that one meter of peat may fully attenuate radiometric signal [@beamishGammaRayAttenuation2013; @reinhardtGammaraySpectrometryVersatile2019], empirical investigations show that the association between peat depth and radioactivity can extend beyond the first meter [@keaneySpatialStatisticsEstimate2013; @gatisMappingUplandPeat2019; @kogantiMappingPeatDepth2023].
-Radiometric data are increasingly available over large areas, presenting a big opportunity for mapping peat depth, insofar as they are predictive of peat depth. 
-Airborne surveys are used for various applications like mineral prospecting and radon risk management, and some countries have high spatial coverage of such data [@minasnyDigitalMappingPeatlands2019; @baranwalAirborneGeophysicalSurveys2020].
+Radiometric data are increasingly available over large areas [@minasnyDigitalMappingPeatlands2019; @baranwalAirborneGeophysicalSurveys2020], and -- insofar as they are predictive of peat depth -- this presents an opportunity for DSM.
 
@@
 Norway has 22 peatland mesotopes, with different hydrology, formation, and development -- from topogeneous or soligeneous fens to blanket or raised bogs [@joostenWiseUseMires2002; @lyngstadBeskrivelserAvTorvmassivenheter2023].
-These mesotopes have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric predictors will vary by landscape.
+These mesotopes have fundamentally different geomorphology, which suggests that relationships between peat depth and terrain or radiometric predictors may vary by landscape.
 
-The little we know about the depths of Norwegian peatlands comes from surveys meant to identify arable land. 
-After scattered surveys in the early 20th century, a comprehensive round of surveying was completed 1964--2001 as part of a wider land cover mapping in Norway  [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
+The only systematic mapping of peat depth at the national scale in Norway comes from surveys meant to identify arable land. 
+After scattered efforts in the early 20th century, a comprehensive round of surveying was completed 1964--2001 as part of a wider land cover mapping in Norway  [@bjordalMarkslagsklassifikasjonOkonomiskKartverk2007].
 This survey produced the initial maps from which Norway's most detailed updated land cover datasets are derived -- including the AR5 and DMK datasets used in this study and described later.
@@
 
-The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify peatlands with rich carbon stocks to avoid their conversion and prioritize their preservation or restoration [@strackPotentialPeatlandsNatureBased2022].
-Although ground surveys of peat depth are accurate and feasible across small areas, having landscape-scale maps *before* detailed investigation increases the option space for spatial planning and land management.
-For example, the Norwegian Public Roads Administration routinely measures peat depth during geotechnical work, but by that time, the route of the road is already set. 
-With a digital peat depth map, planners could better compare routes and climate mitigation measures would have more leverage.
-Therefore, soil science needs more studies of relationships between peat depth and topographic or radiometric data to determine how predictive they are, which variables are most predictive, and how consistent their associations are.
+The push for nature-based climate solutions motivates for extensive mapping of peat depth, to identify peatlands with rich carbon stocks to avoid their conversion and prioritize their preservation or restoration [@strackPotentialPeatlandsNatureBased2022].
+Although ground surveys of peat depth are accurate and feasible across small areas, having landscape-scale maps available before detailed investigation increases the option space for spatial planning and land management.
+For example, the Norwegian Public Roads Administration routinely measures peat depth during geotechnical work before road construction, but by that time, the route of the road is already set. 
+With a landscape-scale peat depth map, planners could better compare the climate impact of different routes.
+To make landscape-scale maps possible, more studies need to investigate the relationships between peat depth and topographic or radiometric variables, to determine how strong and consistent these are.
 
-Here, we assess how well remotely sensed topographic and radiometric data can predict peat depth at two contrasting sites, with a view towards revising regional and national maps in Norway. 
+Here, we assess how well remotely sensed topographic and radiometric data can predict peat depth at two contrasting sites, with a view toward revising regional and national maps in Norway. 
 Specifically, we aim to: (1) quantify the accuracy of predictions from topographic and radiometric variables, and (2) identify key predictive variables.
@@
 The study area has a diverse bedrock, with 32 % alkali feldspar granite, 26 % marlstone, 10 % granite, 8 % monzonite, 7 % sandstone, 6 % limestone, and five other rock types with 1--5 % coverage (Geological Survey of Norway, 1:250 000 dataset).
-The area within our delineation at Skrimfjella is classified as the landscape type *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
 It is almost without human infrastructure, dominated by conifer forest, and borders on a nature reserve.
-The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is $10.8^\circ$ (IQR $4.6-15.1^\circ$).
+Its mean elevation is 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is $10.8^\circ$ (IQR $4.6-15.1^\circ$).
 
@@
 Bedrock in the area is 84 % granitic gneiss,  11 % granite, and 5 % aluminium silicate gneiss (Geological Survey of Norway, 1:250 000 dataset). 
-This study area comprises a wide range of major landscape types: *coastal plains*, *coastal fjord*, *inland valleys*, as well as *inland hills and mountains* [@simensenDiversityDistributionLandscape2021].
-It is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
+This study area is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
 Its mean elevation is 211 m above sea level (range 0--807 m, IQR 73--310 m), and its mean slope at 10 m resolution is $13.0^\circ$ (IQR $4.7-18.3^\circ$).
@@
 Although radiometric data must be in units of counts per second to model attenuation directly [@olearyDigitalSoilMapping2022], we used the radiometric data as provided to us: in units of concentration for potassium, thorium, and uranium (converted from counts per second by scalar calibration factors).
-The monotonic transformation between counts per second and concentration does not affect the tree-based machine learning algorithm that we used to model peat depth [@hastieElementsStatisticalLearning2009].
+The monotonic transformation between counts per second and concentration has no effect on the tree-based machine learning algorithm that we used to model peat depth [@hastieElementsStatisticalLearning2009].
 
@@
 
-We used the *terra* package to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
+We used the *terra* R package (v1.8) to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
 These were derived at two scales to produce eight different predictors; we either calculated the indices at 1 m DTM resolution and then aggregated to 10 m resolution, or aggregated to 10 m DTM resolution and then calculated the indices.
@@
 
-The Depth-to-Water index [@murphyMappingWetlandsComparison2007] approximates a location's vertical height above the surface water feature (e.g.
-stream, lake, or sea) that it is likely to drain towards.
+The Depth-to-Water index [@murphyMappingWetlandsComparison2007] approximates a location's vertical height above the surface water feature (e.g., stream, lake, or sea) that it is likely to drain towards.
 It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature [eq. 5 in @murphyTopographicModellingSoil2009].
-We calculated unitless slope from the 1 m DTM using the Whitebox software.
+We calculated unitless slope from the 1 m DTM using Whitebox.
 We defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell [@schonauerSpatiotemporalPredictionSoil2021; @schonauerRcodeCalculatingDepthwater2021].
@@
 A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population, while avoiding spurious ones [@brusSamplingDigitalSoil2019].
-Although we implemented the approach differently for Skrimfjella (in 2020) than for Ørskogfjellet (in 2023), the objective was the same.
-We drew 105 raster cells at Skrimfjella and 160 raster cells at Ørskogfjellet.
+Although we implemented the sampling design differently at Skrimfjella (in 2020) than at Ørskogfjellet (in 2023), the overall approach was the same.
+We chose 105 raster cells at Skrimfjella and 160 raster cells at Ørskogfjellet as our designed samples.
 A complete description of our sample selection is in Appendix \ref{sec:sample-selection}.
-
-We made several point measurements of peat depth within the sampled raster cells (three if by probing only, more if by GPR as well).
-The point measurements (described below) were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
 
@@
 We measured peat depths at Skrimfjella in August 2020 and at Ørskogfjellet in August 2023.
+We made at least three point measurements of peat depth within each of the raster cells in the designed samples.
 At both sites, we used manual probing as the primary method of measuring peat depth and ground-penetrating radar (GPR) as a secondary method.
-That is: peat depth was always measured by probing, and also by GPR in a subset of cells that partly overlapped with the probed cells.
-We chose this approach because probing is a fast and reliable method for point measurements, while GPR can provide higher lateral density of data in the same amount of time [@parryEvaluatingApproachesEstimating2014].
+That is: peat depth was always measured by probing in our designed samples, and we also used GPR in a set of cells that partly overlapped with the designed samples.
+We chose this combined approach because probing is a fast and reliable method for point measurements, while GPR can provide higher lateral density of data in the same amount of time [@parryEvaluatingApproachesEstimating2014].
 Probed depths serve to calibrate GPR measurements when calibration by common midpoint survey is not possible -- as was the case with our fixed radar antennas -- so the methods are complementary.
@@
 
-We also extracted a set of existing depth measurements for Ørskogfjellet from a paper map made by the Norwegian Soil and Mire Company in 1984.
-The map presents 44 borehole depths (in decimeters) across a 9 ha peatland area.
-We georeferenced the map and digitized the borehole locations and depths.
+At Ørskogfjellet we were also able to use two existing sets of peat depth measurements in addition to our own.
+We extracted a set of 44 borehole depths (in decimeters) across a 9 ha peatland area, from a paper map made by the Norwegian Soil and Mire Company in 1984.
+We also used a set of GPR-based depth measurements, commissioned and provided to us by the Norwegian Public Roads Administration.
+These data were collected in 2020 and 2021 with a dual channel system (70 MHz and 300 MHz; ImpulseRadar AB, Sweden), connected to GNSS with CPOS correction.
+We used a total of 403440 interpreted and calibrated traces along 7.4 km of transects from this work -- discarding data where multiple depths were interpreted for the same location.
 
-We also used a set of existing GPR depth measurements from Ørskogfjellet, commissioned and provided to us by the Norwegian Public Roads Administration.
-These data were collected in 2020 and 2021 with a dual channel system (70 MHz and 300 MHz; ImpulseRadar AB, Sweden), connected to GNSS with CPOS correction.
-For the work presented here, we used a total of 403440 interpreted and calibrated traces along 7.4 km of transects -- discarding data where multiple depths were interpreted for the same location.
+All point measurements described above were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
 
@@
 We used Random Forests (RF) to predict peat depth at both sites.
-RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees [@breimanRandomForests2001].
+RF is an ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees [@breimanRandomForests2001].
 We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in DSM applications than other algorithms [@beguinPredictingSoilProperties2017; @nussbaumEvaluationDigitalSoil2018; @lamichhaneDigitalSoilMapping2019].
 It is suited for use on relatively small training datasets and its predictions can be interrogated to learn about predictor importance [@khaledianSelectingAppropriateMachine2020].
-Evaluating variable importance in a maximally predictive model aligns with the aim of this study.
 
-RF by itself is not a spatial model, and it will only predict spatial structure in the outcome to the degree that the structure is captured by predictors.
-We considered using regression kriging -- a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically interpolated surface of RF residuals [@henglGenericFrameworkSpatial2004].
-The spatial component in regression kriging often improves map accuracy compared to a non-spatial model [@beguinPredictingSoilProperties2017; @lamichhaneDigitalSoilMapping2019; @mollaMachineLearningGeostatistical2023], but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations [@henglGenericFrameworkSpatial2004; @szaboMappingSoilHydraulic2019; @takoutsingComparingPredictionPerformance2022].
+RF by itself is not a spatial model, and it will only predict spatial structure in the peat depth to the degree that spatial structure is captured by predictors.
+We considered using regression kriging -- a hybrid between non-spatial and spatial techniques that adds to the RF predictions a geostatistically interpolated surface of RF residuals [@henglGenericFrameworkSpatial2004].
+The spatial component in regression kriging often improves map accuracy [@beguinPredictingSoilProperties2017; @lamichhaneDigitalSoilMapping2019; @mollaMachineLearningGeostatistical2023], but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations [@henglGenericFrameworkSpatial2004; @szaboMappingSoilHydraulic2019; @takoutsingComparingPredictionPerformance2022].
 If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.
-We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that (non-spatial) RF rather than regression kriging was justified at both sites.
+We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that (non-spatial) RF was justified at both sites.
 
@@
 Comparing the different configurations allowed us to isolate the added value of each of the predictor groups.
-The models with only DMK peat depth class were simple linear models rather than RFs, and served to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth, calibrated on the same data.
+The models with only DMK peat depth class served to provide a fair comparison between the accuracy of the RF models and the existing national map of peat depth, calibrated on the same data.
 
@@
 To set the folds we used k-Means Nearest Neighbor Distance Matching (kNNDM), which aims to mimic the spatial prediction task that is defined as the goal [@linnenbrinkKNNDMCVKfold2024].
-In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
+In particular, kNNDM searches for a fold assignment that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
 That way, the spatial separation between folds is similar to the separation between training and prediction locations -- which increases the quality of the map accuracy estimate [@linnenbrinkKNNDMCVKfold2024].
 For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation [@wadouxSpatialCrossvalidationNot2021].
-We implemented the kNNDM with the *CAST* R package [v1.0, @meyerCASTPackageTraining2024], setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
+We implemented the kNNDM with the *CAST* R package [v1.0, @meyerCASTPackageTraining2024], setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (k = 5--20) that produced the best match between the two NND distributions.
 From the cross-validation we quantified mean absolute error (error magnitude, original scale), R^2^ (explained variation, standardized scale), and Lin's concordance correlation coefficient (error magnitude and explained variation, standardized scale).
@@
 Therefore, we produced prediction intervals with quantile regression forests [@meinshausenQuantileRegressionForests2006], and used the same spatial cross-validation to evaluate the prediction interval coverage probability [@shresthaMachineLearningApproaches2006].
-The quantile regression forests were trained with predictor configuration that showed the highest performance at each site, and we extracted 90 % prediction intervals.
+The quantile regression forests were trained with the predictor configuration that showed the highest performance at each site, and we extracted 90 % prediction intervals.
 
@@
 We calculated partial dependence with the *pdp* R package [v0.8, @greenwellPdpPackageConstructing2017].
-For the six most important variables, we plotted both partial dependence and individual conditional expectation, to show the average effect of the predictor on the outcome and the variation in the effect across observations, respectively [@goldsteinPeekingBlackBox2015].
+For the six most important variables, we plotted partial dependence to show the average effect of the predictor on the outcome and individual conditional expectations to show variation in the effect across observations [@goldsteinPeekingBlackBox2015].
 Non-parallel individual conditional expectation lines indicate the presence of interactions between predictors.
@@
 
-Our point measurements of peat depth (all sources) produced aggregated depths for 372 cells or 2.4 % of AR5 mire area at Skrimfjella, and 1878 cells or 1.2 % of AR5 mire area at Ørskogfjellet (Table \ref{tab:depthsByClass}).
-Approximately 80 % of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland.
+Our point measurements of peat depth (all sources) produced aggregated depths for 372 cells or 2.4 % of AR5 mire area at Skrimfjella, and 1878 cells or 1.2 % of AR5 mire area at Ørskogfjellet.
+Approximately 80 % of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
 DMK peat depth had higher coverage at Ørskogfjellet than at Skrimfjella (80 % versus 27 % of cells not *unknown*), and at Ørskogfjellet the *deep* and *shallow* classes showed a larger difference in measured depth.
@@
 The *terrain* configuration outperformed the *radiometric* configuration by 0.27 in concordance correlation coefficient ($p < 0.001$; Appendix Table \@ref(tab:pairwiseSkrimCCC)), by 0.16 in R^2^ ($p = 0.190$; Appendix Table \@ref(tab:pairwiseSkrimRsq)), and by 10 cm in mean absolute error ($p = 0.038$; Appendix Table \@ref(tab:pairwiseSkrimMAE)).
-The *terrainDMK* configuration outperformed the *radiometricDMK* configuration by 0.32 in concordance correlation coefficient ($p < 0.001$; Appendix Table \@ref(tab:pairwiseSkrimCCC)), by 0.11 in R^2^ ($p = 0.605$; Appendix Table \@ref(tab:pairwiseSkrimRsq)), and by 11 cm in mean absolute error ($p = 0.011$; Appendix Table \@ref(tab:pairwiseSkrimMAE)).
+The *terrain + DMK* configuration outperformed the *radiometric + DMK* configuration by 0.32 in concordance correlation coefficient ($p < 0.001$; Appendix Table \@ref(tab:pairwiseSkrimCCC)), by 0.11 in R^2^ ($p = 0.605$; Appendix Table \@ref(tab:pairwiseSkrimRsq)), and by 11 cm in mean absolute error ($p = 0.011$; Appendix Table \@ref(tab:pairwiseSkrimMAE)).
 
@@
 The *terrain* configuration outperformed the *radiometric* configuration by 0.24 in concordance correlation coefficient ($p = 0.068$; Appendix Table \@ref(tab:pairwiseOrskogCCC)), by 0.24 in R^2^ ($p = 0.004$; Appendix Table \@ref(tab:pairwiseOrskogRsq)), and by 25 cm in mean absolute error ($p = 0.086$; Appendix Table \@ref(tab:pairwiseOrskogMAE)).
-The *terrainDMK* configuration outperformed the *radiometricDMK* configuration by 0.28 in concordance correlation coefficient ($p = 0.015$; Appendix Table \@ref(tab:pairwiseOrskogCCC)), by 0.27 in R^2^ ($p < 0.001$; Appendix Table \@ref(tab:pairwiseOrskogRsq)), and by 28 cm in mean absolute error ($p = 0.031$; Appendix Table \@ref(tab:pairwiseOrskogMAE)).
+The *terrain + DMK* configuration outperformed the *radiometric + DMK* configuration by 0.28 in concordance correlation coefficient ($p = 0.015$; Appendix Table \@ref(tab:pairwiseOrskogCCC)), by 0.27 in R^2^ ($p < 0.001$; Appendix Table \@ref(tab:pairwiseOrskogRsq)), and by 28 cm in mean absolute error ($p = 0.031$; Appendix Table \@ref(tab:pairwiseOrskogMAE)).
 
@@
 Similarly, the *terrain + DMK* configuration for Ørskogfjellet was reduced from 23 variables to 11 non-collinear variables.
-
-The *permutation* and *Shapley* methods of variable importance were highly correlated at both sites, while the *FIRM* method was less correlated with the other two (Fig. \@ref(fig:varImp)).
+The *permutation* and *Shapley* methods of variable importance showed high rank correlation at both sites, while the *FIRM* method ranked variable importance differently (Fig. \@ref(fig:varImp)).
 
@@
 DMK was also important -- the *shallow* class in particular -- but only at Ørskogfjellet.
-Some realizations of the hydrological predictors Topographic Wetness Index and Depth-to-Water showed considerable importance, while others showed little -- with no consistency between sites.
-At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices slope, Terrain Ruggedness Index, Topographic Position Index, and roughness.
-The radiometric predictor uranium ground concentration showed moderate importance at Skrimfjella.
+Some realizations of the hydrological predictors Topographic Wetness Index and Depth-to-Water showed considerable importance, while others showed little -- with no clear consistency between sites.
+At both sites, the most important hydrological predictors were more important than the simple terrain indices slope, Terrain Ruggedness Index, Topographic Position Index, and roughness.
+At Skrimfjella, the radiometric predictor uranium ground concentration -- which was highly correlated with all other radiometric variables -- showed moderate importance.
 
@@
 Variables removed due to collinearity are shown to the right of the variable with which they are most correlated.
-Spearman rank correlations between the three metrics are displayed in matrices."}
+Spearman rank correlations between the three metrics at each site are displayed in matrices."}
 knitr::include_graphics("figures/variable_importance.pdf")
@@
 Many of the most important predictors in the best performing models showed non-monotonic effects on peat depth (Fig. \@ref(fig:pdps)).
-At Ørskogfjellet for example, increasing elevation was predictive of deeper peat up to about 75 m above sea level, after which a further increase was predictive of shallower peat.
+At Ørskogfjellet, for example, increasing elevation was predictive of deeper peat up to about 75 m above sea level, after which a further increase was predictive of shallower peat.
 At Skrimfjella the partial dependence on elevation had the opposite shape but covers a higher elevation range, with the shallowest peats predicted around 350 m above sea level.
-TWI50m at Ørskogfjellet and DTW4000 and uranium ground concentration at Skrimfjella were other predictors that showed considerable fluctuations in their predictive effects on peat depth.
+TWI50m at Ørskogfjellet and DTW4000 and uranium ground concentration at Skrimfjella were other predictors that showed non-monotonic associations with peat depth.
 The radiometric predictor in particular displayed an idiosyncratic effect, with a marked dip in predicted depth at intermediate values of uranium ground concentration.
@@
 Similarly, individual conditional expectations along uranium ground concentration at Skrimfjella were non-parallel, with some locations showing decreasing peat depth predictions (unlike the average effect).
-Nevertheless, most individual conditional expectations were approximately parallel, so the average effects of the predictors are representative of their overall effects.
+Nevertheless, most individual conditional expectations were approximately parallel, so the average effects of the predictors are mostly representative of their overall effects.
 
@@
 Mean absolute errors of 60 and 56 cm at the two sites — relative to mean depths of 119 and 126 cm — illustrate the practical limitations of these maps.
-Since any given 10 m cell will miss by about 60 cm, applications requiring detailed peat depth in a small area (e.g.
-< 1 ha) would benefit from measuring depth on the ground rather than relying on the DSM alone.
+Since any given 10 m cell will miss by about 60 cm, applications requiring detailed peat depth in a small area (e.g., < 1 ha) would benefit from measuring depth on the ground rather than relying on the DSM alone.
 
@@
 
-Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth but generally produced better or comparable error magnitude [R^2^ vs.
-mean absolute error or root mean square error, @wadouxIntegratedApproachEvaluation2022].
+Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth (R^2^) but generally produced better or comparable error magnitude [mean absolute error or root mean square error, @wadouxIntegratedApproachEvaluation2022].
 It is important to keep in mind that differences in peat depth distributions, spatial scales, and evaluation methods make direct performance comparisons precarious. 
-More standardized reporting would help but not eliminate this consideration. 
 For example, R^2^ is sensitive to high leverage, extreme values, so it will evaluate a right-skewed distribution differently than a symmetrical distribution. 
@@
 Gatis et al. [-@gatisMappingUplandPeat2019] used similar predictors and the same spatial grain, finding a much stronger relationship between predicted and observed peat depth ($R^2 = 0.68$). 
-Although their random evaluation data partition could make performance estimates too optimistic [@robertsCrossvalidationStrategiesData2017; @wadouxSpatialCrossvalidationNot2021], the confounding effect of spatial structure is probably small because they used linear regression and few predictors.
-Their model had limited opportunity to overfit to the spatial structure in peat depth. 
+Although their random evaluation data partition could make performance estimates too optimistic [@robertsCrossvalidationStrategiesData2017; @wadouxSpatialCrossvalidationNot2021], the confounding effect of spatial structure is probably small because they used linear regression and few predictors, so their model had limited opportunity to overfit to the spatial structure in peat depth.
 The most salient difference in Gatis et al. [-@gatisMappingUplandPeat2019] compared to our study is the character of the study area. 
 They study a flatter area with a higher proportion of peatland cover, and their peatland is primarily blanket bog.
-The peat surface of a blanket bog is tied more closely to its underlying topography than the peat surface of raised bogs or fens [@lindsayPeatlandMireTypes2016], which may make depths in blanket bogs easier to predict.
+The peat surface of a blanket bog is tied more closely to its underlying topography than the peat surface of raised bogs or fens [@lindsayPeatlandMireTypes2016], which may make blanket bog depth easier to predict.
 The predominance of fens and smaller peatland extent may have contributed to worse performance in our study.
@@
 Where overprediction occurred, the error was smaller. 
-This is not unexpected for the right-skewed distributions of peat depth, but it has management implications. 
+This is not unexpected for the right-skewed distributions of peat depth, but it has implications for potential users of the maps. 
 For example, it makes the maps more suited for "red-lighting" than "green-lighting" peatland conversion (assuming depth is the determinative factor). 
-Identifying the deepest peats will require additional field work in candidate areas, which could be defined by an upper quantile of predicted depth.
+Identifying the deepest peats will require additional field work in candidate areas -- where candidate areas could be defined by some upper quantile of predicted depth.
 
@@
 The RF algorithm’s flexibility means that radiometrics could be used for shallower peats if they provided predictive value for that part of the depth distribution, but there is no evidence of that in our results.
-In the partial dependence plot of uranium ground concentration at Skrimfjella, the expected negative relationship between depth and uranium ground concentration cannot be found by ignoring the left (highly attenuated) side of the distribution.
+In the partial dependence plot of uranium ground concentration at Skrimfjella, the expected negative relationship between depth and uranium ground concentration cannot be found by ignoring the left (highly attenuated) side of the distribution (Fig.
+\@ref(fig:pdps)).
 More than a quarter of the peats in each of our study areas were less than 60 cm deep, and full attenuation is unlikely for these [@beamishGammaRayAttenuation2013].
@@
 
-At both our sites, elevation was the most important predictor, and peat depth showed non-monotonic responses to changes in elevation (Figs. \@ref(fig:varImp)-\@ref(fig:pdps)).
-We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g.
-via climate) to explain the observed patterns.
+At both our sites, elevation was the most important predictor, and peat depth showed non-monotonic responses to changes in elevation (Figs. \@ref(fig:varImp), \@ref(fig:pdps)).
+We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g., via climate) to explain the observed patterns.
 For example, the increase in peat depth from 350 to 700 m.a.s.l.
@@
 Relationships between elevation and peat depth have previously shown opposite shapes in different areas [@finlaysonEstimatingOrganicSurface2021]. 
-Nonetheless, a relationship that is not generalizable beyond the mapping area is still useful for DSM, as long as it is evaluated to demonstrate its robustness for the predictive task (e.g.
-through kNNDM spatial cross-validation).
+Nonetheless, a relationship that is not generalizable beyond the mapping area is still useful for DSM, as long as it is evaluated to demonstrate its robustness for the predictive task (e.g., through kNNDM spatial cross-validation).
 
@@
 
-The performance gap between the best models and the DMK models shows that digital soil mapping of peat depth has value in Norway. 
+The performance gap between the best models and the DMK models shows that DSM of peat depth has value in Norway. 
 Where calibrating measurements are available, better maps than DMK peat depth class can be produced at low cost. 
@@
 
-Peat depths typically vary over short distances (e.g. < 1 m), so mapping at 10 m resolution implies that the map will compress much of the fine-scale variation.
+Peat depths typically vary over short distances (e.g., < 1 m), so mapping at 10 m resolution implies that the map will compress much of the fine-scale variation.
 Terrain--depth relationships might be stronger at finer resolution than 10 m, especially considering the hummock--hollow microtopography of many peatlands [@rydin7Mires1999; @lindsayPeatbogsCarbonCritical2010].
@@
 
-To define a spatial extent for a particular DSM, kNNDM can be used to investigate how far the mapping can extend beyond a set of point measurements; if the sample-to-prediction nearest neighbor distribution cannot be simulated by any set of cross-validation folds, then the extent is too expansive [@meyerMachineLearningbasedGlobal2022; @linnenbrinkKNNDMCVKfold2024].
+To define a spatial extent for a particular DSM, kNNDM can be used to investigate how far the map can extend beyond a set of point measurements; if the sample-to-prediction nearest neighbor distribution cannot be simulated by any set of cross-validation folds, then the extent is too expansive [@meyerMachineLearningbasedGlobal2022; @linnenbrinkKNNDMCVKfold2024].
 However, the tradeoff between multiple small-extent DSM and fewer large-extent DSM needs research. 
-Bohn and Miller [-@bohnLocallyEnhancedDigital2024] advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or 'supertopes'), where the composition of mesotopes is similar -- e.g.
-regions dominated by raised bogs versus regions dominated by sloping fens [@moenNationalAtlasNorway1999; @joostenWiseUseMires2002].
+Bohn and Miller [-@bohnLocallyEnhancedDigital2024] advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or 'supertopes'), where the composition of mesotopes is similar -- e.g., regions dominated by raised bogs versus regions dominated by sloping fens [@moenNationalAtlasNorway1999; @joostenWiseUseMires2002].
 Depth varies systematically between bogs and fens [@lindsayPeatlandMireTypes2016], between peats formed by terrestrialization versus paludification [@buffamFillingHolesRegional2010], and probably along other axes of peatland typology.
@@
 
-Finally, we briefly want to highlight the need for peatland extent mapping and peat depth mapping to be better integrated. 
+Finally, we would like to highlight briefly the need for peatland extent mapping and peat depth mapping to be better integrated. 
 Since peatland extent is defined by non-zero peat depth [the specific threshold varies by definition, @minasnyMappingMonitoringPeatland2024], the lateral and vertical dimensions are fundamentally linked.
@@
 Our findings show that terrain-derived variables, particularly elevation and Multi-Resolution Valley Bottom Flatness, provide predictive value for peat depth mapping within peatland extents. 
-In contrast, airborne radiometric data showed little to no predictive value at either of two study sites, likely due to the large footprint of airborne sensors relative to the fine-scale variation in peat depths.
+In contrast, airborne radiometric data showed little to no predictive value at either of two study sites, possibly because of the large footprint of airborne spectrometers relative to the fine-scale variation in peat depths.
 

### 2025-06-27 ms/ms.tex
read through on paper

@@
 Terrain variables were better predictors than radiometric variables, with elevation and valley bottom flatness showing the strongest relationships to depth.
-Radiometric variables showed inconsistent predictive value -- improving performance at one site while degrading it at the other.
-The accuracy of the national map of peat depth did not measure up to any of our remote sensing models, even though it was calibrated to the same data.
+Radiometric variables showed inconsistent and weak predictive value -- improving performance at one site while degrading it at the other.
+Our remote sensing models had better accuracy than the national map of peat depth, even when we calibrated the national map to the same depth data.
 Still, weak relationships with remotely sensed variables made peat depth hard to predict overall.
-Based on these findings, we conclude that digital soil mapping can improve existing, broad-scale maps of peat depth in Norway, but highly localized carbon stock assessments are best made from field measurements.
+Based on these findings, we conclude that digital soil mapping can improve the existing, national map of peat depth in Norway, but detailed local maps are best made from tailored field measurements.
 Together, these pathways promise more accurate landscape-scale carbon stock assessments and better-informed land management policies.
@@
 Intact peatlands sequester carbon and overall produce a negative temperature forcing \citep{joostenRolePeatlandsClimate2016}.
-When disturbed, often by conversion to another land use, they can produce large greenhouse gas emissions \citep{maGloballyRobustRelationship2022}.
-
-One of the keys to the areal density of peatland carbon stocks lies in the third dimension: their depth.
-Peat soils range from zero to over ten meters deep \citep{widyastutiPEATGRIDSMappingThickness2024}, so deep peats comprise a large volume.
-Their depth results from the accumulation of organic matter over thousands of years \citep{loiselDatabaseSynthesisNorthern2014, joostenRolePeatlandsClimate2016}.
+When disturbed -- often by conversion to another land use -- they can produce large greenhouse gas emissions \citep{maGloballyRobustRelationship2022}.
+
+One of the keys to the areal density of peatland carbon stocks (and variation therein) lies in the third dimension: their depth.
+Peat soils range from zero to over ten meters deep \citep{widyastutiPEATGRIDSMappingThickness2024}.
+Their depth depends on the rate and duration of organic matter accumulation, potentially over thousands of years \citep{loiselDatabaseSynthesisNorthern2014, joostenRolePeatlandsClimate2016}.
 In the anoxic and acidic conditions created by a high water table, plant material decay is slightly outweighed by new growth, and the surplus carbon is laid down as peat.
-There it remains sequestered as long as the water table stays high and prevents oxidative peat decomposition.
 
 Peatlands are most common at high latitudes, and in regions with high cover they are frequently converted to human land use \citep{unepGlobalPeatlandsAssessment2022}.
-They are attractive for agriculture and forestry because they are flat, treeless, and have developed soils.
-However, other land uses also displace peatlands.
+They are attractive for agriculture and forestry because they are flat, treeless, and have developed soils, but other land uses also displace peatlands.
 In Norway -- where flat lowland is relatively scarce but upward of 9 \% of the land area is peatland \citep{brynLandCoverNorway2018, bakkestuenDelineationWetlandAreas2023} -- lawmakers have restricted peatland afforestation and cultivation in recent decades.
@@
 
-The spatial distribution of peat depth is often overlooked in land use planning and carbon accounting because peat depth is not mapped with sufficient coverage, resolution, or accuracy \citep{beilmanPeatCarbonStocks2008, hastieRisksCarbonStorage2022, unepGlobalPeatlandsAssessment2022}.
-Maps are crucial because they link high-level targets to specific management decisions, unlike spatially aggregated estimates \citep{oecdOECDEnvironmentalPerformance2022}.
+Variation in peat depth and its spatial distribution is often overlooked in land use planning and carbon accounting because peat depth is not mapped with sufficient coverage, resolution, or accuracy \citep{beilmanPeatCarbonStocks2008, hastieRisksCarbonStorage2022, unepGlobalPeatlandsAssessment2022}.
+Maps are crucial because, unlike spatially aggregated estimates, they link high-level targets to specific management decisions \citep{oecdOECDEnvironmentalPerformance2022}.
 Maps of peat depth also make it possible to quantify the effect of specific management decisions and thereby understand how local outcomes contribute to regional and national outcomes \citep{oecdOECDEnvironmentalPerformance2022}.
 
-Measuring peat depth on the ground by probing or ground-penetrating radar is straightforward, and a field survey can map a small area at low cost.
-However, surveying large areas is impractical when depth varies widely over short distances (e.g.~10 m) -- as in many peatlands \citep{torppaStatisticalApproachDetermining2011, proulx-mcinnisReviewGroundpenetratingRadar2013, henrionMappingMonitoringPeatlands2024}.
+Measuring peat depth on the ground is straightforward, and a field survey can map a small area at low cost.
+However, surveying large areas is impractical when depth varies widely over short distances (e.g., 10 m) -- as in many peatlands \citep{torppaStatisticalApproachDetermining2011, proulx-mcinnisReviewGroundpenetratingRadar2013, henrionMappingMonitoringPeatlands2024}.
 A complementary approach from soil science is digital soil mapping (DSM).
 DSM scales up field measurements from a set of locations to a wider area, by relating the measured values to other variables mapped over the area of interest, through a statistical model.
-This approach has grown in importance with the availability of remotely sensed data and the advancement of methods for identifying statistical patterns, especially through machine learning \citep{minasnyDigitalMappingPeatlands2019, wadouxMachineLearningDigital2020}.
+This approach has grown in importance with the availability of remotely sensed data and the advancement of machine learning methods \citep{minasnyDigitalMappingPeatlands2019, wadouxMachineLearningDigital2020}.
 
@@
 
-Studies have consistently shown relationships between peat depth and topography, although the specific patterns vary.
-One of the most robust relationships is a negative correlation between depth and terrain slope \citep[e.g.][]{holdenEstimatingCarbonStock2011, parryMethodModellingPeat2012, gatisMappingUplandPeat2019}.
-Peat depth also changes with elevation in many contexts, but with inconsistent directionality \citep[e.g.][]{holdenEstimatingCarbonStock2011, parryMethodModellingPeat2012, rudiyantoDigitalMappingCosteffective2016, rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024}.
-Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth in some studies \citep[e.g.][]{rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024, pohjankukkaDigitalMappingPeat2025}.
+Studies have consistently shown relationships between peat depth and topography, though the specific patterns vary.
+One of the most robust relationships is a negative correlation between depth and terrain slope \citep[e.g.,][]{holdenEstimatingCarbonStock2011, parryMethodModellingPeat2012, gatisMappingUplandPeat2019}.
+Peat depth also changes with elevation in many contexts, but with inconsistent directionality \citep[e.g.,][]{holdenEstimatingCarbonStock2011, parryMethodModellingPeat2012, rudiyantoDigitalMappingCosteffective2016, rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024}.
+Also more complex derivations of topography, such as the Topographic Wetness Index and the Multi-Resolution Valley Bottom Flatness index, have shown associations with peat depth \citep[e.g.,][]{rudiyantoOpenDigitalMapping2018, kogantiMappingPeatDepth2023, liFactorsControllingPeat2024, pohjankukkaDigitalMappingPeat2025}.
 Some of the variation between studies in quantifying these relationships is undoubtedly attributable to issues of spatial scale -- both the scaling of the topographic predictors and the resolution of the peat depth analysis.
@@
 Although theory suggests that one meter of peat may fully attenuate radiometric signal \citep{beamishGammaRayAttenuation2013, reinhardtGammaraySpectrometryVersatile2019}, empirical investigations show that the association between peat depth and radioactivity can extend beyond the first meter \citep{keaneySpatialStatisticsEstimate2013, gatisMappingUplandPeat2019, kogantiMappingPeatDepth2023}.
-Radiometric data are increasingly available over large areas, presenting a big opportunity for mapping peat depth, insofar as they are predictive of peat depth.
-Airborne surveys are used for various applications like mineral prospecting and radon risk management, and some countries have high spatial coverage of such data \citep{minasnyDigitalMappingPeatlands2019, baranwalAirborneGeophysicalSurveys2020}.
+Radiometric data are increasingly available over large areas \citep{minasnyDigitalMappingPeatlands2019, baranwalAirborneGeophysicalSurveys2020}, and -- insofar as they are predictive of peat depth -- this presents an opportunity for DSM.
 
@@
 Norway has 22 peatland mesotopes, with different hydrology, formation, and development -- from topogeneous or soligeneous fens to blanket or raised bogs \citep{joostenWiseUseMires2002, lyngstadBeskrivelserAvTorvmassivenheter2023}.
-These mesotopes have fundamentally different geomorphology, suggesting that peat depth relationships with terrain or radiometric predictors will vary by landscape.
-
-The little we know about the depths of Norwegian peatlands comes from surveys meant to identify arable land.
-After scattered surveys in the early 20th century, a comprehensive round of surveying was completed 1964--2001 as part of a wider land cover mapping in Norway \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
+These mesotopes have fundamentally different geomorphology, which suggests that relationships between peat depth and terrain or radiometric predictors may vary by landscape.
+
+The only systematic mapping of peat depth at the national scale in Norway comes from surveys meant to identify arable land.
+After scattered efforts in the early 20th century, a comprehensive round of surveying was completed 1964--2001 as part of a wider land cover mapping in Norway \citep{bjordalMarkslagsklassifikasjonOkonomiskKartverk2007}.
 This survey produced the initial maps from which Norway's most detailed updated land cover datasets are derived -- including the AR5 and DMK datasets used in this study and described later.
@@
 
-The push for nature-based climate solutions motivates for broad mapping of peat depth, to identify peatlands with rich carbon stocks to avoid their conversion and prioritize their preservation or restoration \citep{strackPotentialPeatlandsNatureBased2022}.
-Although ground surveys of peat depth are accurate and feasible across small areas, having landscape-scale maps \emph{before} detailed investigation increases the option space for spatial planning and land management.
-For example, the Norwegian Public Roads Administration routinely measures peat depth during geotechnical work, but by that time, the route of the road is already set.
-With a digital peat depth map, planners could better compare routes and climate mitigation measures would have more leverage.
-Therefore, soil science needs more studies of relationships between peat depth and topographic or radiometric data to determine how predictive they are, which variables are most predictive, and how consistent their associations are.
-
-Here, we assess how well remotely sensed topographic and radiometric data can predict peat depth at two contrasting sites, with a view towards revising regional and national maps in Norway.
+The push for nature-based climate solutions motivates for extensive mapping of peat depth, to identify peatlands with rich carbon stocks to avoid their conversion and prioritize their preservation or restoration \citep{strackPotentialPeatlandsNatureBased2022}.
+Although ground surveys of peat depth are accurate and feasible across small areas, having landscape-scale maps available before detailed investigation increases the option space for spatial planning and land management.
+For example, the Norwegian Public Roads Administration routinely measures peat depth during geotechnical work before road construction, but by that time, the route of the road is already set.
+With a landscape-scale peat depth map, planners could better compare the climate impact of different routes.
+To make landscape-scale maps possible, more studies need to investigate the relationships between peat depth and topographic or radiometric variables, to determine how strong and consistent these are.
+
+Here, we assess how well remotely sensed topographic and radiometric data can predict peat depth at two contrasting sites, with a view toward revising regional and national maps in Norway.
 Specifically, we aim to: (1) quantify the accuracy of predictions from topographic and radiometric variables, and (2) identify key predictive variables.
@@
 The study area has a diverse bedrock, with 32 \% alkali feldspar granite, 26 \% marlstone, 10 \% granite, 8 \% monzonite, 7 \% sandstone, 6 \% limestone, and five other rock types with 1--5 \% coverage (Geological Survey of Norway, 1:250 000 dataset).
-The area within our delineation at Skrimfjella is classified as the landscape type \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
 It is almost without human infrastructure, dominated by conifer forest, and borders on a nature reserve.
-The study area has a mean elevation of 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is \(10.8^\circ\) (IQR \(4.6-15.1^\circ\)).
+Its mean elevation is 438 m a.s.l. (range 223--711 m, IQR 351--509 m), and its mean slope at 10 m resolution is \(10.8^\circ\) (IQR \(4.6-15.1^\circ\)).
 
@@
 Bedrock in the area is 84 \% granitic gneiss, 11 \% granite, and 5 \% aluminium silicate gneiss (Geological Survey of Norway, 1:250 000 dataset).
-This study area comprises a wide range of major landscape types: \emph{coastal plains}, \emph{coastal fjord}, \emph{inland valleys}, as well as \emph{inland hills and mountains} \citep{simensenDiversityDistributionLandscape2021}.
-It is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
+This study area is mostly forested, but also contains considerable farmland and open upland, and has several large lakes.
 Its mean elevation is 211 m above sea level (range 0--807 m, IQR 73--310 m), and its mean slope at 10 m resolution is \(13.0^\circ\) (IQR \(4.7-18.3^\circ\)).
@@
 Although radiometric data must be in units of counts per second to model attenuation directly \citep{olearyDigitalSoilMapping2022}, we used the radiometric data as provided to us: in units of concentration for potassium, thorium, and uranium (converted from counts per second by scalar calibration factors).
-The monotonic transformation between counts per second and concentration does not affect the tree-based machine learning algorithm that we used to model peat depth \citep{hastieElementsStatisticalLearning2009}.
+The monotonic transformation between counts per second and concentration has no effect on the tree-based machine learning algorithm that we used to model peat depth \citep{hastieElementsStatisticalLearning2009}.
 
@@
 
-We used the \emph{terra} package to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
+We used the \emph{terra} R package (v1.8) to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
 These were derived at two scales to produce eight different predictors; we either calculated the indices at 1 m DTM resolution and then aggregated to 10 m resolution, or aggregated to 10 m DTM resolution and then calculated the indices.
@@
 
-The Depth-to-Water index \citep{murphyMappingWetlandsComparison2007} approximates a location's vertical height above the surface water feature (e.g.~stream, lake, or sea) that it is likely to drain towards.
+The Depth-to-Water index \citep{murphyMappingWetlandsComparison2007} approximates a location's vertical height above the surface water feature (e.g., stream, lake, or sea) that it is likely to drain towards.
 It is calculated as the minimum cumulative slope (scaled by cell size) to a surface water feature \citep[eq. 5 in][]{murphyTopographicModellingSoil2009}.
-We calculated unitless slope from the 1 m DTM using the Whitebox software.
+We calculated unitless slope from the 1 m DTM using Whitebox.
 We defined surface water features from the DTM by filling depressions and then calculating flow accumulation to define catchment areas for each cell \citep{schonauerSpatiotemporalPredictionSoil2021, schonauerRcodeCalculatingDepthwater2021}.
@@
 A sample that preserves the properties of the multivariate distribution of predictor and outcome variables is most likely to maintain any complex, non-linear relationships that exist in the population, while avoiding spurious ones \citep{brusSamplingDigitalSoil2019}.
-Although we implemented the approach differently for Skrimfjella (in 2020) than for Ørskogfjellet (in 2023), the objective was the same.
-We drew 105 raster cells at Skrimfjella and 160 raster cells at Ørskogfjellet.
+Although we implemented the sampling design differently at Skrimfjella (in 2020) than at Ørskogfjellet (in 2023), the overall approach was the same.
+We chose 105 raster cells at Skrimfjella and 160 raster cells at Ørskogfjellet as our designed samples.
 A complete description of our sample selection is in Appendix \ref{sec:sample-selection}.
 
-We made several point measurements of peat depth within the sampled raster cells (three if by probing only, more if by GPR as well).
-The point measurements (described below) were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
-
 \subsection{Depth measurements}
@@
 We measured peat depths at Skrimfjella in August 2020 and at Ørskogfjellet in August 2023.
+We made at least three point measurements of peat depth within each of the raster cells in the designed samples.
 At both sites, we used manual probing as the primary method of measuring peat depth and ground-penetrating radar (GPR) as a secondary method.
-That is: peat depth was always measured by probing, and also by GPR in a subset of cells that partly overlapped with the probed cells.
-We chose this approach because probing is a fast and reliable method for point measurements, while GPR can provide higher lateral density of data in the same amount of time \citep{parryEvaluatingApproachesEstimating2014}.
+That is: peat depth was always measured by probing in our designed samples, and we also used GPR in a set of cells that partly overlapped with the designed samples.
+We chose this combined approach because probing is a fast and reliable method for point measurements, while GPR can provide higher lateral density of data in the same amount of time \citep{parryEvaluatingApproachesEstimating2014}.
 Probed depths serve to calibrate GPR measurements when calibration by common midpoint survey is not possible -- as was the case with our fixed radar antennas -- so the methods are complementary.
@@
 
-We also extracted a set of existing depth measurements for Ørskogfjellet from a paper map made by the Norwegian Soil and Mire Company in 1984.
-The map presents 44 borehole depths (in decimeters) across a 9 ha peatland area.
-We georeferenced the map and digitized the borehole locations and depths.
-
-We also used a set of existing GPR depth measurements from Ørskogfjellet, commissioned and provided to us by the Norwegian Public Roads Administration.
+At Ørskogfjellet we were also able to use two existing sets of peat depth measurements in addition to our own.
+We extracted a set of 44 borehole depths (in decimeters) across a 9 ha peatland area, from a paper map made by the Norwegian Soil and Mire Company in 1984.
+We also used a set of GPR-based depth measurements, commissioned and provided to us by the Norwegian Public Roads Administration.
 These data were collected in 2020 and 2021 with a dual channel system (70 MHz and 300 MHz; ImpulseRadar AB, Sweden), connected to GNSS with CPOS correction.
-For the work presented here, we used a total of 403440 interpreted and calibrated traces along 7.4 km of transects -- discarding data where multiple depths were interpreted for the same location.
+We used a total of 403440 interpreted and calibrated traces along 7.4 km of transects from this work -- discarding data where multiple depths were interpreted for the same location.
+
+All point measurements described above were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
 
@@
 We used Random Forests (RF) to predict peat depth at both sites.
-RF is a tree-based ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees \citep{breimanRandomForests2001}.
+RF is an ensemble machine learning algorithm that builds many decision trees on bootstrapped samples of the training data, randomly subsets predictors in the trees, and averages the predictions of the trees \citep{breimanRandomForests2001}.
 We chose RF because it can handle complex interactions between predictors, is robust to overfitting, and generally shows higher performance in DSM applications than other algorithms \citep{beguinPredictingSoilProperties2017, nussbaumEvaluationDigitalSoil2018, lamichhaneDigitalSoilMapping2019}.
 It is suited for use on relatively small training datasets and its predictions can be interrogated to learn about predictor importance \citep{khaledianSelectingAppropriateMachine2020}.
-Evaluating variable importance in a maximally predictive model aligns with the aim of this study.
-
-RF by itself is not a spatial model, and it will only predict spatial structure in the outcome to the degree that the structure is captured by predictors.
-We considered using regression kriging -- a hybrid between non-spatial and spatial techniques that would be achieved by adding to the RF predictions a geostatistically interpolated surface of RF residuals \citep{henglGenericFrameworkSpatial2004}.
-The spatial component in regression kriging often improves map accuracy compared to a non-spatial model \citep{beguinPredictingSoilProperties2017, lamichhaneDigitalSoilMapping2019, mollaMachineLearningGeostatistical2023}, but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations \citep{henglGenericFrameworkSpatial2004, szaboMappingSoilHydraulic2019, takoutsingComparingPredictionPerformance2022}.
+
+RF by itself is not a spatial model, and it will only predict spatial structure in the peat depth to the degree that spatial structure is captured by predictors.
+We considered using regression kriging -- a hybrid between non-spatial and spatial techniques that adds to the RF predictions a geostatistically interpolated surface of RF residuals \citep{henglGenericFrameworkSpatial2004}.
+The spatial component in regression kriging often improves map accuracy \citep{beguinPredictingSoilProperties2017, lamichhaneDigitalSoilMapping2019, mollaMachineLearningGeostatistical2023}, but it can do so only if the spatial autocorrelation range in the non-spatial residuals is large compared to distances between samples and prediction locations \citep{henglGenericFrameworkSpatial2004, szaboMappingSoilHydraulic2019, takoutsingComparingPredictionPerformance2022}.
 If the outcome varies at fine scales and the samples are clustered in small parts of the study area, a spatial component will hardly improve overall map accuracy.
-We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that (non-spatial) RF rather than regression kriging was justified at both sites.
+We used semivariograms to assess the spatial structure in the residuals of the RF predictions, and found that (non-spatial) RF was justified at both sites.
 
@@
 To set the folds we used k-Means Nearest Neighbor Distance Matching (kNNDM), which aims to mimic the spatial prediction task that is defined as the goal \citep{linnenbrinkKNNDMCVKfold2024}.
-In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
+In particular, kNNDM searches for a fold assignment that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
 That way, the spatial separation between folds is similar to the separation between training and prediction locations -- which increases the quality of the map accuracy estimate \citep{linnenbrinkKNNDMCVKfold2024}.
 For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation \citep{wadouxSpatialCrossvalidationNot2021}.
-We implemented the kNNDM with the \emph{CAST} R package \citep[v1.0,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
+We implemented the kNNDM with the \emph{CAST} R package \citep[v1.0,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (k = 5--20) that produced the best match between the two NND distributions.
 From the cross-validation we quantified mean absolute error (error magnitude, original scale), R\textsuperscript{2} (explained variation, standardized scale), and Lin's concordance correlation coefficient (error magnitude and explained variation, standardized scale).
@@
 Therefore, we produced prediction intervals with quantile regression forests \citep{meinshausenQuantileRegressionForests2006}, and used the same spatial cross-validation to evaluate the prediction interval coverage probability \citep{shresthaMachineLearningApproaches2006}.
-The quantile regression forests were trained with predictor configuration that showed the highest performance at each site, and we extracted 90 \% prediction intervals.
+The quantile regression forests were trained with the predictor configuration that showed the highest performance at each site, and we extracted 90 \% prediction intervals.
 
@@
 We calculated partial dependence with the \emph{pdp} R package \citep[v0.8,][]{greenwellPdpPackageConstructing2017}.
-For the six most important variables, we plotted both partial dependence and individual conditional expectation, to show the average effect of the predictor on the outcome and the variation in the effect across observations, respectively \citep{goldsteinPeekingBlackBox2015}.
+For the six most important variables, we plotted partial dependence to show the average effect of the predictor on the outcome and individual conditional expectations to show variation in the effect across observations \citep{goldsteinPeekingBlackBox2015}.
 Non-parallel individual conditional expectation lines indicate the presence of interactions between predictors.
@@
 
-Our point measurements of peat depth (all sources) produced aggregated depths for 372 cells or 2.4 \% of AR5 mire area at Skrimfjella, and 1878 cells or 1.2 \% of AR5 mire area at Ørskogfjellet (Table \ref{tab:depthsByClass}).
-Approximately 80 \% of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland.
+Our point measurements of peat depth (all sources) produced aggregated depths for 372 cells or 2.4 \% of AR5 mire area at Skrimfjella, and 1878 cells or 1.2 \% of AR5 mire area at Ørskogfjellet.
+Approximately 80 \% of these 10 m cells were within AR5 mires, and the remainder in forest, open upland, or farmland (Table \ref{tab:depthsByClass}).
 DMK peat depth had higher coverage at Ørskogfjellet than at Skrimfjella (80 \% versus 27 \% of cells not \emph{unknown}), and at Ørskogfjellet the \emph{deep} and \emph{shallow} classes showed a larger difference in measured depth.
@@
 The \emph{terrain} configuration outperformed the \emph{radiometric} configuration by 0.27 in concordance correlation coefficient (\(p < 0.001\); Appendix Table \ref{tab:pairwiseSkrimCCC}), by 0.16 in R\textsuperscript{2} (\(p = 0.190\); Appendix Table \ref{tab:pairwiseSkrimRsq}), and by 10 cm in mean absolute error (\(p = 0.038\); Appendix Table \ref{tab:pairwiseSkrimMAE}).
-The \emph{terrainDMK} configuration outperformed the \emph{radiometricDMK} configuration by 0.32 in concordance correlation coefficient (\(p < 0.001\); Appendix Table \ref{tab:pairwiseSkrimCCC}), by 0.11 in R\textsuperscript{2} (\(p = 0.605\); Appendix Table \ref{tab:pairwiseSkrimRsq}), and by 11 cm in mean absolute error (\(p = 0.011\); Appendix Table \ref{tab:pairwiseSkrimMAE}).
+The \emph{terrain + DMK} configuration outperformed the \emph{radiometric + DMK} configuration by 0.32 in concordance correlation coefficient (\(p < 0.001\); Appendix Table \ref{tab:pairwiseSkrimCCC}), by 0.11 in R\textsuperscript{2} (\(p = 0.605\); Appendix Table \ref{tab:pairwiseSkrimRsq}), and by 11 cm in mean absolute error (\(p = 0.011\); Appendix Table \ref{tab:pairwiseSkrimMAE}).
 
@@
 The \emph{terrain} configuration outperformed the \emph{radiometric} configuration by 0.24 in concordance correlation coefficient (\(p = 0.068\); Appendix Table \ref{tab:pairwiseOrskogCCC}), by 0.24 in R\textsuperscript{2} (\(p = 0.004\); Appendix Table \ref{tab:pairwiseOrskogRsq}), and by 25 cm in mean absolute error (\(p = 0.086\); Appendix Table \ref{tab:pairwiseOrskogMAE}).
-The \emph{terrainDMK} configuration outperformed the \emph{radiometricDMK} configuration by 0.28 in concordance correlation coefficient (\(p = 0.015\); Appendix Table \ref{tab:pairwiseOrskogCCC}), by 0.27 in R\textsuperscript{2} (\(p < 0.001\); Appendix Table \ref{tab:pairwiseOrskogRsq}), and by 28 cm in mean absolute error (\(p = 0.031\); Appendix Table \ref{tab:pairwiseOrskogMAE}).
+The \emph{terrain + DMK} configuration outperformed the \emph{radiometric + DMK} configuration by 0.28 in concordance correlation coefficient (\(p = 0.015\); Appendix Table \ref{tab:pairwiseOrskogCCC}), by 0.27 in R\textsuperscript{2} (\(p < 0.001\); Appendix Table \ref{tab:pairwiseOrskogRsq}), and by 28 cm in mean absolute error (\(p = 0.031\); Appendix Table \ref{tab:pairwiseOrskogMAE}).
 
@@
 Similarly, the \emph{terrain + DMK} configuration for Ørskogfjellet was reduced from 23 variables to 11 non-collinear variables.
-
-The \emph{permutation} and \emph{Shapley} methods of variable importance were highly correlated at both sites, while the \emph{FIRM} method was less correlated with the other two (Fig.
+The \emph{permutation} and \emph{Shapley} methods of variable importance showed high rank correlation at both sites, while the \emph{FIRM} method ranked variable importance differently (Fig.
 \ref{fig:varImp}).
@@
 DMK was also important -- the \emph{shallow} class in particular -- but only at Ørskogfjellet.
-Some realizations of the hydrological predictors Topographic Wetness Index and Depth-to-Water showed considerable importance, while others showed little -- with no consistency between sites.
-At both sites, the most important realizations of hydrological predictors were more important than the simple terrain indices slope, Terrain Ruggedness Index, Topographic Position Index, and roughness.
-The radiometric predictor uranium ground concentration showed moderate importance at Skrimfjella.
+Some realizations of the hydrological predictors Topographic Wetness Index and Depth-to-Water showed considerable importance, while others showed little -- with no clear consistency between sites.
+At both sites, the most important hydrological predictors were more important than the simple terrain indices slope, Terrain Ruggedness Index, Topographic Position Index, and roughness.
+At Skrimfjella, the radiometric predictor uranium ground concentration -- which was highly correlated with all other radiometric variables -- showed moderate importance.
 
@@
 Variables removed due to collinearity are shown to the right of the variable with which they are most correlated.
-Spearman rank correlations between the three metrics are displayed in matrices.}\label{fig:varImp}
+Spearman rank correlations between the three metrics at each site are displayed in matrices.}\label{fig:varImp}
 \end{figure}
@@
 Many of the most important predictors in the best performing models showed non-monotonic effects on peat depth (Fig. \ref{fig:pdps}).
-At Ørskogfjellet for example, increasing elevation was predictive of deeper peat up to about 75 m above sea level, after which a further increase was predictive of shallower peat.
+At Ørskogfjellet, for example, increasing elevation was predictive of deeper peat up to about 75 m above sea level, after which a further increase was predictive of shallower peat.
 At Skrimfjella the partial dependence on elevation had the opposite shape but covers a higher elevation range, with the shallowest peats predicted around 350 m above sea level.
-TWI50m at Ørskogfjellet and DTW4000 and uranium ground concentration at Skrimfjella were other predictors that showed considerable fluctuations in their predictive effects on peat depth.
+TWI50m at Ørskogfjellet and DTW4000 and uranium ground concentration at Skrimfjella were other predictors that showed non-monotonic associations with peat depth.
 The radiometric predictor in particular displayed an idiosyncratic effect, with a marked dip in predicted depth at intermediate values of uranium ground concentration.
@@
 Mean absolute errors of 60 and 56 cm at the two sites --- relative to mean depths of 119 and 126 cm --- illustrate the practical limitations of these maps.
-Since any given 10 m cell will miss by about 60 cm, applications requiring detailed peat depth in a small area (e.g.~\textless{} 1 ha) would benefit from measuring depth on the ground rather than relying on the DSM alone.
+Since any given 10 m cell will miss by about 60 cm, applications requiring detailed peat depth in a small area (e.g., \textless{} 1 ha) would benefit from measuring depth on the ground rather than relying on the DSM alone.
 
@@
 
-Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth but generally produced better or comparable error magnitude \citep[R\textsuperscript{2} vs.~mean absolute error or root mean square error,][]{wadouxIntegratedApproachEvaluation2022}.
+Compared to other studies using terrain and radiometric data to predict peat depth, our models explained less variability in peat depth (R\textsuperscript{2}) but generally produced better or comparable error magnitude \citep[mean absolute error or root mean square error,][]{wadouxIntegratedApproachEvaluation2022}.
 It is important to keep in mind that differences in peat depth distributions, spatial scales, and evaluation methods make direct performance comparisons precarious.
-More standardized reporting would help but not eliminate this consideration.
 For example, R\textsuperscript{2} is sensitive to high leverage, extreme values, so it will evaluate a right-skewed distribution differently than a symmetrical distribution.
@@
 \citeyearpar{gatisMappingUplandPeat2019} used similar predictors and the same spatial grain, finding a much stronger relationship between predicted and observed peat depth (\(R^2 = 0.68\)).
-Although their random evaluation data partition could make performance estimates too optimistic \citep{robertsCrossvalidationStrategiesData2017, wadouxSpatialCrossvalidationNot2021}, the confounding effect of spatial structure is probably small because they used linear regression and few predictors.
-Their model had limited opportunity to overfit to the spatial structure in peat depth.
+Although their random evaluation data partition could make performance estimates too optimistic \citep{robertsCrossvalidationStrategiesData2017, wadouxSpatialCrossvalidationNot2021}, the confounding effect of spatial structure is probably small because they used linear regression and few predictors, so their model had limited opportunity to overfit to the spatial structure in peat depth.
 The most salient difference in Gatis et al. \citeyearpar{gatisMappingUplandPeat2019} compared to our study is the character of the study area.
 They study a flatter area with a higher proportion of peatland cover, and their peatland is primarily blanket bog.
-The peat surface of a blanket bog is tied more closely to its underlying topography than the peat surface of raised bogs or fens \citep{lindsayPeatlandMireTypes2016}, which may make depths in blanket bogs easier to predict.
+The peat surface of a blanket bog is tied more closely to its underlying topography than the peat surface of raised bogs or fens \citep{lindsayPeatlandMireTypes2016}, which may make blanket bog depth easier to predict.
 The predominance of fens and smaller peatland extent may have contributed to worse performance in our study.
@@
 Where overprediction occurred, the error was smaller.
-This is not unexpected for the right-skewed distributions of peat depth, but it has management implications.
+This is not unexpected for the right-skewed distributions of peat depth, but it has implications for potential users of the maps.
 For example, it makes the maps more suited for ``red-lighting'' than ``green-lighting'' peatland conversion (assuming depth is the determinative factor).
-Identifying the deepest peats will require additional field work in candidate areas, which could be defined by an upper quantile of predicted depth.
+Identifying the deepest peats will require additional field work in candidate areas -- where candidate areas could be defined by some upper quantile of predicted depth.
 
@@
 The RF algorithm's flexibility means that radiometrics could be used for shallower peats if they provided predictive value for that part of the depth distribution, but there is no evidence of that in our results.
-In the partial dependence plot of uranium ground concentration at Skrimfjella, the expected negative relationship between depth and uranium ground concentration cannot be found by ignoring the left (highly attenuated) side of the distribution.
+In the partial dependence plot of uranium ground concentration at Skrimfjella, the expected negative relationship between depth and uranium ground concentration cannot be found by ignoring the left (highly attenuated) side of the distribution (Fig.
+\ref{fig:pdps}).
 More than a quarter of the peats in each of our study areas were less than 60 cm deep, and full attenuation is unlikely for these \citep{beamishGammaRayAttenuation2013}.
@@
 
-At both our sites, elevation was the most important predictor, and peat depth showed non-monotonic responses to changes in elevation (Figs. \ref{fig:varImp}-\ref{fig:pdps}).
-We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g.~via climate) to explain the observed patterns.
+At both our sites, elevation was the most important predictor, and peat depth showed non-monotonic responses to changes in elevation (Figs. \ref{fig:varImp}, \ref{fig:pdps}).
+We believe that the idiosyncratic elevational relationships we detected are mostly not generalizable beyond the study areas, because we see no simple mechanism (e.g., via climate) to explain the observed patterns.
 For example, the increase in peat depth from 350 to 700 m.a.s.l.
@@
 Relationships between elevation and peat depth have previously shown opposite shapes in different areas \citep{finlaysonEstimatingOrganicSurface2021}.
-Nonetheless, a relationship that is not generalizable beyond the mapping area is still useful for DSM, as long as it is evaluated to demonstrate its robustness for the predictive task (e.g.~through kNNDM spatial cross-validation).
+Nonetheless, a relationship that is not generalizable beyond the mapping area is still useful for DSM, as long as it is evaluated to demonstrate its robustness for the predictive task (e.g., through kNNDM spatial cross-validation).
 
@@
 
-The performance gap between the best models and the DMK models shows that digital soil mapping of peat depth has value in Norway.
+The performance gap between the best models and the DMK models shows that DSM of peat depth has value in Norway.
 Where calibrating measurements are available, better maps than DMK peat depth class can be produced at low cost.
@@
 
-Peat depths typically vary over short distances (e.g.~\textless{} 1 m), so mapping at 10 m resolution implies that the map will compress much of the fine-scale variation.
+Peat depths typically vary over short distances (e.g., \textless{} 1 m), so mapping at 10 m resolution implies that the map will compress much of the fine-scale variation.
 Terrain--depth relationships might be stronger at finer resolution than 10 m, especially considering the hummock--hollow microtopography of many peatlands \citep{rydin7Mires1999, lindsayPeatbogsCarbonCritical2010}.
@@
 
-To define a spatial extent for a particular DSM, kNNDM can be used to investigate how far the mapping can extend beyond a set of point measurements; if the sample-to-prediction nearest neighbor distribution cannot be simulated by any set of cross-validation folds, then the extent is too expansive \citep{meyerMachineLearningbasedGlobal2022, linnenbrinkKNNDMCVKfold2024}.
+To define a spatial extent for a particular DSM, kNNDM can be used to investigate how far the map can extend beyond a set of point measurements; if the sample-to-prediction nearest neighbor distribution cannot be simulated by any set of cross-validation folds, then the extent is too expansive \citep{meyerMachineLearningbasedGlobal2022, linnenbrinkKNNDMCVKfold2024}.
 However, the tradeoff between multiple small-extent DSM and fewer large-extent DSM needs research.
-Bohn and Miller \citeyearpar{bohnLocallyEnhancedDigital2024} advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or `supertopes'), where the composition of mesotopes is similar -- e.g.~regions dominated by raised bogs versus regions dominated by sloping fens \citep{moenNationalAtlasNorway1999, joostenWiseUseMires2002}.
+Bohn and Miller \citeyearpar{bohnLocallyEnhancedDigital2024} advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or `supertopes'), where the composition of mesotopes is similar -- e.g., regions dominated by raised bogs versus regions dominated by sloping fens \citep{moenNationalAtlasNorway1999, joostenWiseUseMires2002}.
 Depth varies systematically between bogs and fens \citep{lindsayPeatlandMireTypes2016}, between peats formed by terrestrialization versus paludification \citep{buffamFillingHolesRegional2010}, and probably along other axes of peatland typology.
@@
 
-Finally, we briefly want to highlight the need for peatland extent mapping and peat depth mapping to be better integrated.
+Finally, we would like to highlight briefly the need for peatland extent mapping and peat depth mapping to be better integrated.
 Since peatland extent is defined by non-zero peat depth \citep[the specific threshold varies by definition,][]{minasnyMappingMonitoringPeatland2024}, the lateral and vertical dimensions are fundamentally linked.
@@
 Our findings show that terrain-derived variables, particularly elevation and Multi-Resolution Valley Bottom Flatness, provide predictive value for peat depth mapping within peatland extents.
-In contrast, airborne radiometric data showed little to no predictive value at either of two study sites, likely due to the large footprint of airborne sensors relative to the fine-scale variation in peat depths.
+In contrast, airborne radiometric data showed little to no predictive value at either of two study sites, possibly because of the large footprint of airborne spectrometers relative to the fine-scale variation in peat depths.
 

### 2025-08-05 ms/ms.Rmd
address RC1.5 and RC1.6 with coauthor input

@@
 Another set of predictors related to peat depth are measurements of natural radioactivity from the ground surface. 
-Gamma-ray spectrometry can survey the activity (decay counts per second) from radioactive isotopes in the earth's crust: potassium-40, uranium-238, and thorium-232 [@reinhardtGammaraySpectrometryVersatile2019].
-These exist in bedrock (and mineral soils) and a peat overburden attenuates the radiation intensity at the surface. 
-The degree of attenuation relates to properties of the overburden, especially composition and depth. 
+Gamma-ray spectrometry can survey the activity (decay counts per second) from radioactive isotopes in bedrock and mineral soils: potassium-40, uranium-238, and thorium-232 [@beamishEnvironmentalRadioactivityUK2014; @reinhardtGammaraySpectrometryVersatile2019].
+Although survey measurements are commonly reported as ground concentrations (linearly scaled from decay counts per second), in peatland environments these predictors do not reflect the concentration of radionuclides near the ground surface, but rather the radiation intensity after attenuation by the peat overburden.
 Deep soil with high water content will attenuate radiation most [@beamishGammaRayAttenuation2013; @reinhardtGammaraySpectrometryVersatile2019]. 
-Thus, gamma-ray radiometrics integrate two *scorpan* factors: other soil properties and parent material [@mcbratneyDigitalSoilMapping2003].
+Thus, reported ground concentrations can be statistically informative about peat soils, even if not physically correct, with respect to two *scorpan* factors: other soil properties and parent material [@mcbratneyDigitalSoilMapping2003].
 
@@
 Therefore, variation in these parameters could have masked the relationship between peat depth and radiometric data.
+This makes physical modeling of peat depth from radiometric data an undetermined problem.
 We chose the Ørskogfjellet site in part because it has a relatively homogeneous bedrock, which should reduce the variation in initial source strength.

### 2025-08-05 ms/ms.tex
address RC1.5 and RC1.6 with coauthor input

@@
 Another set of predictors related to peat depth are measurements of natural radioactivity from the ground surface.
-Gamma-ray spectrometry can survey the activity (decay counts per second) from radioactive isotopes in the earth's crust: potassium-40, uranium-238, and thorium-232 \citep{reinhardtGammaraySpectrometryVersatile2019}.
-These exist in bedrock (and mineral soils) and a peat overburden attenuates the radiation intensity at the surface.
-The degree of attenuation relates to properties of the overburden, especially composition and depth.
+Gamma-ray spectrometry can survey the activity (decay counts per second) from radioactive isotopes in bedrock and mineral soils: potassium-40, uranium-238, and thorium-232 \citep{beamishEnvironmentalRadioactivityUK2014, reinhardtGammaraySpectrometryVersatile2019}.
+Although survey measurements are commonly reported as ground concentrations (linearly scaled from decay counts per second), in peatland environments these predictors do not reflect the concentration of radionuclides near the ground surface, but rather the radiation intensity after attenuation by the peat overburden.
 Deep soil with high water content will attenuate radiation most \citep{beamishGammaRayAttenuation2013, reinhardtGammaraySpectrometryVersatile2019}.
-Thus, gamma-ray radiometrics integrate two \emph{scorpan} factors: other soil properties and parent material \citep{mcbratneyDigitalSoilMapping2003}.
+Thus, reported ground concentrations can be statistically informative about peat soils, even if not physically correct, with respect to two \emph{scorpan} factors: other soil properties and parent material \citep{mcbratneyDigitalSoilMapping2003}.
 

### 2025-08-05 ms/ms.Rmd
address RC1.5, RC1.6, RC2.3, RC2.6, RC2.8 with coauthor input

@@
 They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling. 
-Sensitivity analysis showed that Pearson correlations between cubic spline and bilinear resampling methods exceeded 0.99 for all radiometric variables, so we are confident that the choice of resampling method did not affect our results.
+Sensitivity analysis showed that Pearson correlations between cubic spline and bilinear resampling methods exceeded 0.995 for all radiometric variables, so we are confident that the choice of resampling method did not affect our results.
 
@@
 From the cross-validation we quantified mean absolute error (error magnitude, original scale), R^2^ (explained variation, standardized scale), and Lin's concordance correlation coefficient (error magnitude and explained variation, standardized scale).
-We formally assessed the effect of predictor configuration on performance metrics using mixed-effects models to account for the cross-validation fold structure, and testing pairwise differences between configurations.
+We formally assessed the effect of predictor configuration on performance metrics using mixed-effects models to account for the cross-validation fold structure (folds as random effects), and testing pairwise differences between configurations.
 
@@
 Similarly, the *terrain + DMK* configuration for Ørskogfjellet was reduced from 23 variables to 11 non-collinear variables.
-The *permutation* and *Shapley* methods of variable importance showed high rank correlation at both sites, while the *FIRM* method ranked variable importance differently (Fig. \@ref(fig:varImp)).
+The *permutation* and *Shapley* methods of variable importance showed high Spearman rank correlation at both sites, while the *FIRM* method ranked variable importance differently (Fig.
+\@ref(fig:varImp)).
 
@@
 All four radiometric variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value. 
-This contrasts with Koganti et al. [-@kogantiMappingPeatDepth2023], who found that radiometric total count was a much better predictor than potassium ground concentration.
+This contrasts with Koganti et al.
+[-@kogantiMappingPeatDepth2023], who found that radiometric total count and potassium ground concentration were better predictors than thorium or uranium ground concentration.
 

### 2025-08-05 ms/ms.tex
address RC1.5, RC1.6, RC2.3, RC2.6, RC2.8 with coauthor input

@@
 They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling.
-Sensitivity analysis showed that Pearson correlations between cubic spline and bilinear resampling methods exceeded 0.99 for all radiometric variables, so we are confident that the choice of resampling method did not affect our results.
+Sensitivity analysis showed that Pearson correlations between cubic spline and bilinear resampling methods exceeded 0.995 for all radiometric variables, so we are confident that the choice of resampling method did not affect our results.
 
@@
 From the cross-validation we quantified mean absolute error (error magnitude, original scale), R\textsuperscript{2} (explained variation, standardized scale), and Lin's concordance correlation coefficient (error magnitude and explained variation, standardized scale).
-We formally assessed the effect of predictor configuration on performance metrics using mixed-effects models to account for the cross-validation fold structure, and testing pairwise differences between configurations.
+We formally assessed the effect of predictor configuration on performance metrics using mixed-effects models to account for the cross-validation fold structure (folds as random effects), and testing pairwise differences between configurations.
 
@@
 Similarly, the \emph{terrain + DMK} configuration for Ørskogfjellet was reduced from 23 variables to 11 non-collinear variables.
-The \emph{permutation} and \emph{Shapley} methods of variable importance showed high rank correlation at both sites, while the \emph{FIRM} method ranked variable importance differently (Fig.
+The \emph{permutation} and \emph{Shapley} methods of variable importance showed high Spearman rank correlation at both sites, while the \emph{FIRM} method ranked variable importance differently (Fig.
 \ref{fig:varImp}).
@@
 All four radiometric variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value.
-This contrasts with Koganti et al. \citeyearpar{kogantiMappingPeatDepth2023}, who found that radiometric total count was a much better predictor than potassium ground concentration.
+This contrasts with Koganti et al.
+\citeyearpar{kogantiMappingPeatDepth2023}, who found that radiometric total count and potassium ground concentration were better predictors than thorium or uranium ground concentration.
 

### 2025-08-05 ms/ms.Rmd
address TE.3

@@
 The monotonic transformation between counts per second and concentration has no effect on the tree-based machine learning algorithm that we used to model peat depth [@hastieElementsStatisticalLearning2009].
+We also used the total count variable as provided, rather than calculating a gamma dose rate based on the potassium, thorium, and uranium [as was done in @gatisMappingUplandPeat2019], because these were highly correlated at both study sites ($\rho = 0.989, 0.986$), and because conversions to dose rates are approximations [@iaeaGuidelinesRadioelementMapping2003].
 

### 2025-08-05 ms/ms.tex
address TE.3

@@
 The monotonic transformation between counts per second and concentration has no effect on the tree-based machine learning algorithm that we used to model peat depth \citep{hastieElementsStatisticalLearning2009}.
+We also used the total count variable as provided, rather than calculating a gamma dose rate based on the potassium, thorium, and uranium \citep[as was done in][]{gatisMappingUplandPeat2019}, because these were highly correlated at both study sites (\(\rho = 0.989, 0.986\)), and because conversions to dose rates are approximations \citep{iaeaGuidelinesRadioelementMapping2003}.
 

### 2025-08-05 R/Skrim/createPredictorStack.R

@@ -19,7 +19,7 @@ radTC <- rast("data/Skrim/NGU-2013-029/Kong_Rad_Area3_TC/Kong_Rad_Area3_TC.ERS")
 rad <- c(radK, radTh, radU, radTC) |> 
   crop(y = st_transform(st_buffer(sa, 1000), st_crs(radK)))
 
-# Sensitivity analysis: Compare cubic spline vs bilinear resampling methods
+## Sensitivity analysis: Compare cubic spline vs bilinear resampling methods ####
 rad10m_cubic <- rad |> 
   project(y = "epsg:25833",  
           method = "cubicspline",
@@ -63,6 +63,16 @@ walk2(rad_cubic_values, rad_bilinear_values,
             main=paste(names(rad_cubic_values)[which(map_lgl(rad_cubic_values, identical, .x))],
                       "\nr =", round(cor(.x, .y, use="complete.obs"), 3))))
 par(mfrow=c(1,1))
+
+## Dose rate ####
+radDR10m <- 
+  13.078 * rad10m$Kong_Rad_Area3_K +
+   5.675 * rad10m$Kong_Rad_Area3_U +
+   2.494 * rad10m$Kong_Rad_Area3_Th
+names(radDR10m) <- "rad_DR"
+c(radDR10m, rad10m) |> 
+  pairs(hist=FALSE, cor=TRUE, maxcells=1e4)
+cor(values(radDR10m), values(rad10m$Kong_Rad_Area3_TC), use = "pairwise.complete.obs")
 
 # Simple terrain ####
 

### 2025-08-05 R/createPredictorStack.R

@@ -12,7 +12,7 @@ radTC50m <- rast("data/NGU-2015-015/Romsdalsfjorden_RAD_miclev_TC.ers")
 rad50m <- c(radK50m, radTh50m, radU50m, radTC50m)
 rad50m[rad50m < 0] <- NA
 
-# Sensitivity analysis: Compare cubic spline vs bilinear resampling methods
+## Sensitivity analysis: Compare cubic spline vs bilinear resampling methods ####
 rad10m_cubic <- rad50m |> 
   project(y = "epsg:25832",  
           method = "cubicspline",
@@ -65,6 +65,16 @@ walk2(rad_cubic_values, rad_bilinear_values,
             main=paste(names(rad_cubic_values)[which(map_lgl(rad_cubic_values, identical, .x))],
                       "\nr =", round(cor(.x, .y, use="complete.obs"), 3))))
 par(mfrow=c(1,1))
+
+## Dose rate ####
+radDR10m <- 
+  13.078 * radK10m +
+   5.675 * radU10m +
+   2.494 * radTh10m
+names(radDR10m) <- "rad_DR_10m"
+c(radDR10m, radTC10m, radK10m, radU10m, radTh10m) |> 
+  pairs(hist=FALSE, cor=TRUE, maxcells=1e4)
+cor(values(radDR10m), values(radTC10m), use = "pairwise.complete.obs")
 
 # Simple terrain ####
 

### 2025-08-29 ms/ms.Rmd
add data archive to reference list
as requested by Copernicus production office

@@
 
-All point measurements described above were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
+All point measurements described above [@volleringPeatDepthOccurrence2025] were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
 

### 2025-08-29 ms/ms.tex
add data archive to reference list
as requested by Copernicus production office

@@
 
-All point measurements described above were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
+All point measurements described above \citep{volleringPeatDepthOccurrence2025} were ultimately aggregated to 10 m resolution by taking the mean of point values within each cell, inversely weighted by their distances to the cell center.
 

### 2025-08-29 ms/submission/AuthorComments2-egusphere-2025-1046.qmd
author response to acceptance

---
title: "Authors' response #2"
author: "Julien Vollering, on behalf of all co-authors"
format: pdf
---

We thank the topic editor for accepting our manuscript. We also thank the the reviewers again for their efforts.

We are not aware of any comments on our revised submission, or specific requested changes. The only change we have made in the text with respect to the revised submission is that we have included an in-text citation to our archived data, as requested in the email from the editorial support team. We have also added a new affiliation for one author.

### 2025-09-05 ms/ms.Rmd
add author affiliation
addressing Editorial support requests
- add copyright symbol to Kartverket as per Kartverket web page
- avoid calls to \citeyearpar

@@
     surname: Kusk Gillespie
-    affiliation: 1
+    affiliation: 1 3
   - given_name: Karl-Kristian
@@
     address: Department of Geography, University of Exeter, United Kingdom
+  - code: 3
+    address: University College VIA, Denmark
 abstract: |
@@
 Terrain is visualized with light orange contour lines at intervals of 100 m, and a hillshade with slope $45^\\circ$ and azimuth $225^\\circ$.
-AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (https://data.norge.no/nlod/en/1.0) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/)."}
+AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (https://data.norge.no/nlod/en/1.0) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/) \\textcopyright\\ Kartverket."}
 knitr::include_graphics("figures/map-sites.pdf")
@@
 
-The Geological Survey of Norway conducted and processed radiometric surveys over our study areas, as reported in Baranwal et al.
-[-@baranwalHelicopterborneMagneticElectromagnetic2013] and Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015].
+The Geological Survey of Norway conducted and processed radiometric surveys over our study areas, as reported in @baranwalHelicopterborneMagneticElectromagnetic2013 and @ofstadHelicopterborneMagneticRadiometric2015.
 They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling. 
@@
 The calculations were performed with Whitebox software [@lindsayWhiteboxGATCase2016], accessed through the *whitebox* R package [v2.4, @wuWhiteboxWhiteboxToolsFrontend2022].
-We filled depressions in the DTM with the algorithm in Wang & Liu [-@wangEfficientMethodIdentifying2006], and used the deterministic infinity flow accumulation algorithm [@tarbotonNewMethodDetermination1997].
+We filled depressions in the DTM with the algorithm in @wangEfficientMethodIdentifying2006, and used the deterministic infinity flow accumulation algorithm [@tarbotonNewMethodDetermination1997].
 
@@
 Terrain is visualized with a hillshade with slope $45^\\circ$ and azimuth $225^\\circ$.
-Terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/)."}
+Terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/) \\textcopyright\\ Kartverket."}
 knitr::include_graphics("figures/map-distribution.pdf")
@@
 
-Gatis et al. [-@gatisMappingUplandPeat2019] used similar predictors and the same spatial grain, finding a much stronger relationship between predicted and observed peat depth ($R^2 = 0.68$). 
+@gatisMappingUplandPeat2019 used similar predictors and the same spatial grain, finding a much stronger relationship between predicted and observed peat depth ($R^2 = 0.68$). 
 Although their random evaluation data partition could make performance estimates too optimistic [@robertsCrossvalidationStrategiesData2017; @wadouxSpatialCrossvalidationNot2021], the confounding effect of spatial structure is probably small because they used linear regression and few predictors, so their model had limited opportunity to overfit to the spatial structure in peat depth.
-The most salient difference in Gatis et al. [-@gatisMappingUplandPeat2019] compared to our study is the character of the study area. 
+The most salient difference in @gatisMappingUplandPeat2019 compared to our study is the character of the study area. 
 They study a flatter area with a higher proportion of peatland cover, and their peatland is primarily blanket bog.
@@
 
-Marchant [-@marchantUsingRemoteSensors2021] examined a subset of the area studied in Gatis et al.
-[-@gatisMappingUplandPeat2019] at 100 m resolution, using splines to relax linearity between radiometry/terrain and peat depth.
+@marchantUsingRemoteSensors2021 examined a subset of the area studied in @gatisMappingUplandPeat2019 at 100 m resolution, using splines to relax linearity between radiometry/terrain and peat depth. 
 He found that potassium ground concentration alone predicted peat depth with much higher concordance than our models (concordance correlation coefficient = 0.76) and that elevation alone produced comparable performance (concordance correlation coefficient = 0.27).
@@
 
-Koganti et al. [-@kogantiMappingPeatDepth2023] had a peat depth distribution and predictors similar to ours, but at much smaller spatial grain and extent. 
+@kogantiMappingPeatDepth2023 had a peat depth distribution and predictors similar to ours, but at much smaller spatial grain and extent. 
 Their training and validation points are closer than the range of spatial autocorrelation in peat depth, so our results are best compared to their non-spatial models. 
 They accounted for more variability in peat depth (adjusted $R^2 = 0.71$) but had larger errors (RSME = 110 cm). 
-Koganti et al.’s [-@kogantiMappingPeatDepth2023] linear regression models produced negative predictions, and it is unclear whether the values quoted above include these. 
+The linear regression models in @kogantiMappingPeatDepth2023 produced negative predictions, and it is unclear whether the values quoted above include these. 
 If we disregard the negative predictions, their model showed the same pattern as ours in overpredicting shallow peats and underpredicting deep peats, although their underprediction was less severe. 
-An important difference between Koganti et al.
-[-@kogantiMappingPeatDepth2023] and our study (besides spatial scale) is that they measured radiometrics on the ground, rather than using airborne survey data.
+An important difference between @kogantiMappingPeatDepth2023 and our study (besides spatial scale) is that they measured radiometrics on the ground, rather than using airborne survey data.
 Thus, the footprint of their detector was smaller and they captured variation in radiation at finer scales.
@@
 All four radiometric variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value. 
-This contrasts with Koganti et al.
-[-@kogantiMappingPeatDepth2023], who found that radiometric total count and potassium ground concentration were better predictors than thorium or uranium ground concentration.
+This contrasts with @kogantiMappingPeatDepth2023, who found that radiometric total count and potassium ground concentration were better predictors than thorium or uranium ground concentration.
 
@@
 Studies comparing airborne and ground radiometric surveys confirm that they are poorly correlated in low-activity areas like peatlands [@kockComparisonAirborneTerrestrial2011; @karjalainenComparisonTwoGammaray2025].
-Karjalainen et al. [-@karjalainenComparisonTwoGammaray2025] found that ground-based measurements predicted peat depth better than airborne measurements. 
-Nevertheless, a large radiometric footprint did not prevent strong associations with peat depth in Gatis et al.
-[-@gatisMappingUplandPeat2019] and Marchant [-@marchantUsingRemoteSensors2021], perhaps because the extensive blanket bog landscape in these studies has more gradual changes in depth [@lindsayBogsEcologyClassification1995].
-We are unsure whether short-range depth changes explain the weak associations that Siemon et al. [-@siemonAirborneElectromagneticRadiometric2020] found in a large raised bog.
+@karjalainenComparisonTwoGammaray2025 found that ground-based measurements predicted peat depth better than airborne measurements. 
+Nevertheless, a large radiometric footprint did not prevent strong associations with peat depth in @gatisMappingUplandPeat2019 and @marchantUsingRemoteSensors2021, perhaps because the extensive blanket bog landscape in these studies has more gradual changes in depth [@lindsayBogsEcologyClassification1995].
+We are unsure whether short-range depth changes explain the weak associations that @siemonAirborneElectromagneticRadiometric2020 found in a large raised bog.
 
@@
 Soil moisture, bulk density, and porosity, however, are not easily measured across landscape scales and were assumed to be homogeneous.
-Uneven snow cover and air moisture during the Ørskogfjellet radiometric survey may also have masked the soil signal in these data, as Ofstad [-@ofstadHelicopterborneMagneticRadiometric2015] reports large variation in weather conditions.
+Uneven snow cover and air moisture during the Ørskogfjellet radiometric survey may also have masked the soil signal in these data, as @ofstadHelicopterborneMagneticRadiometric2015 reports large variation in weather conditions.
 If maps of these other physical parameters at the time of the radiometric survey were available and included in the model, the predictive value of radiometric data might improve, but this is not a practical solution for digital soil mapping of peat depth.
@@
 At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables (Fig. \@ref(fig:modelMetrics)).
-This is consistent with the findings of Pohjankukka et al. [-@pohjankukkaDigitalMappingPeat2025], who mapped peat depth classes at 50 m resolution across Finland.
+This is consistent with the findings of @pohjankukkaDigitalMappingPeat2025, who mapped peat depth classes at 50 m resolution across Finland.
 
@@
 Better infrastructure to make depth measurements findable, accessible, interoperable, and reusable would help DSM and other applications. 
-Geoportal access and data exchange standards [like Natural England’s for peat surveys, -@naturalenglandDataExchangeStandard2023] are important. 
+Geoportal access and data exchange standards, such as those developed in @naturalenglandDataExchangeStandard2023 for peat surveys, are important. 
 Peat depth is quick and easy to measure, so integrating its measurement into existing national field programs, like Norway’s spatially representative nature monitoring or national forest inventory, would be helpful (although not sufficient for regional or local mapping).
@@
 However, the tradeoff between multiple small-extent DSM and fewer large-extent DSM needs research. 
-Bohn and Miller [-@bohnLocallyEnhancedDigital2024] advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or 'supertopes'), where the composition of mesotopes is similar -- e.g., regions dominated by raised bogs versus regions dominated by sloping fens [@moenNationalAtlasNorway1999; @joostenWiseUseMires2002].
+@bohnLocallyEnhancedDigital2024 advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or 'supertopes'), where the composition of mesotopes is similar -- e.g., regions dominated by raised bogs versus regions dominated by sloping fens [@moenNationalAtlasNorway1999; @joostenWiseUseMires2002].
 Depth varies systematically between bogs and fens [@lindsayPeatlandMireTypes2016], between peats formed by terrestrialization versus paludification [@buffamFillingHolesRegional2010], and probably along other axes of peatland typology.
@@
 The success of Multi-Resolution Valley Bottom Flatness as a predictor of peat depth demonstrates that multiscale spatial patterns matter for peat depth, and convolutional neural networks are designed to learn such patterns [@borowiecDeepLearningTool2022].
-The kind of relationship described in Buffam et al.
-[-@buffamFillingHolesRegional2010], where peat depth in basins related to terrain slope at the basin edge, is also something a convolutional neural network could learn.
+The kind of relationship described in @buffamFillingHolesRegional2010, where peat depth in basins related to terrain slope at the basin edge, is also something a convolutional neural network could learn.
 However, since this approach is data-hungry, we should build soil knowledge into the DSM where we can [@minasnySoilScienceInformedMachine2024].
@@
 \begin{table}[ht]
-\caption{Attributes of the radiometric surveys, as reported in Baranwal et al., \citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013} and Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.}
+\caption{Attributes of the radiometric surveys, as reported in \citet{baranwalHelicopterborneMagneticElectromagnetic2013} and \citet{ofstadHelicopterborneMagneticRadiometric2015}.}
 \begin{tabular}{lll}

### 2025-09-05 ms/ms.tex
add author affiliation
addressing Editorial support requests
- add copyright symbol to Kartverket as per Kartverket web page
- avoid calls to \citeyearpar

@@
 \Author[2]{Naomi}{Gatis}
-\Author[1]{Mette}{Kusk Gillespie}
+\Author[1 3]{Mette}{Kusk Gillespie}
 \Author[1]{Karl-Kristian}{Muggerud}
@@
 \affil[2]{Department of Geography, University of Exeter, United Kingdom}
+\affil[3]{University College VIA, Denmark}
 
@@
 Terrain is visualized with light orange contour lines at intervals of 100 m, and a hillshade with slope $45^\circ$ and azimuth $225^\circ$.
-AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (https://data.norge.no/nlod/en/1.0) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/).}\label{fig:map-sites}
+AR50 data are from Norwegian Institute of Bioeconomy Research under the Norwegian Licence for Open Government Data (https://data.norge.no/nlod/en/1.0) and terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/) \textcopyright\ Kartverket.}\label{fig:map-sites}
 \end{figure}
@@
 
-The Geological Survey of Norway conducted and processed radiometric surveys over our study areas, as reported in Baranwal et al.
-\citeyearpar{baranwalHelicopterborneMagneticElectromagnetic2013} and Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015}.
+The Geological Survey of Norway conducted and processed radiometric surveys over our study areas, as reported in \citet{baranwalHelicopterborneMagneticElectromagnetic2013} and \citet{ofstadHelicopterborneMagneticRadiometric2015}.
 They provided us for each site four variables at 50 m resolution, which we downscaled to 10 m resolution by cubic spline resampling.
@@
 The calculations were performed with Whitebox software \citep{lindsayWhiteboxGATCase2016}, accessed through the \emph{whitebox} R package \citep[v2.4,][]{wuWhiteboxWhiteboxToolsFrontend2022}.
-We filled depressions in the DTM with the algorithm in Wang \& Liu \citeyearpar{wangEfficientMethodIdentifying2006}, and used the deterministic infinity flow accumulation algorithm \citep{tarbotonNewMethodDetermination1997}.
+We filled depressions in the DTM with the algorithm in \citet{wangEfficientMethodIdentifying2006}, and used the deterministic infinity flow accumulation algorithm \citep{tarbotonNewMethodDetermination1997}.
 
@@
 Terrain is visualized with a hillshade with slope $45^\circ$ and azimuth $225^\circ$.
-Terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/).}\label{fig:map-distribution}
+Terrain data are from the Norwegian Mapping Authority under the Creative Commons Attribution 4.0 International license (https://creativecommons.org/licenses/by/4.0/) \textcopyright\ Kartverket.}\label{fig:map-distribution}
 \end{figure}
@@
 
-Gatis et al.
-\citeyearpar{gatisMappingUplandPeat2019} used similar predictors and the same spatial grain, finding a much stronger relationship between predicted and observed peat depth (\(R^2 = 0.68\)).
+\citet{gatisMappingUplandPeat2019} used similar predictors and the same spatial grain, finding a much stronger relationship between predicted and observed peat depth (\(R^2 = 0.68\)).
 Although their random evaluation data partition could make performance estimates too optimistic \citep{robertsCrossvalidationStrategiesData2017, wadouxSpatialCrossvalidationNot2021}, the confounding effect of spatial structure is probably small because they used linear regression and few predictors, so their model had limited opportunity to overfit to the spatial structure in peat depth.
-The most salient difference in Gatis et al. \citeyearpar{gatisMappingUplandPeat2019} compared to our study is the character of the study area.
+The most salient difference in \citet{gatisMappingUplandPeat2019} compared to our study is the character of the study area.
 They study a flatter area with a higher proportion of peatland cover, and their peatland is primarily blanket bog.
@@
 
-Marchant \citeyearpar{marchantUsingRemoteSensors2021} examined a subset of the area studied in Gatis et al.
-\citeyearpar{gatisMappingUplandPeat2019} at 100 m resolution, using splines to relax linearity between radiometry/terrain and peat depth.
+\citet{marchantUsingRemoteSensors2021} examined a subset of the area studied in \citet{gatisMappingUplandPeat2019} at 100 m resolution, using splines to relax linearity between radiometry/terrain and peat depth.
 He found that potassium ground concentration alone predicted peat depth with much higher concordance than our models (concordance correlation coefficient = 0.76) and that elevation alone produced comparable performance (concordance correlation coefficient = 0.27).
@@
 
-Koganti et al. \citeyearpar{kogantiMappingPeatDepth2023} had a peat depth distribution and predictors similar to ours, but at much smaller spatial grain and extent.
+\citet{kogantiMappingPeatDepth2023} had a peat depth distribution and predictors similar to ours, but at much smaller spatial grain and extent.
 Their training and validation points are closer than the range of spatial autocorrelation in peat depth, so our results are best compared to their non-spatial models.
 They accounted for more variability in peat depth (adjusted \(R^2 = 0.71\)) but had larger errors (RSME = 110 cm).
-Koganti et al.'s \citeyearpar{kogantiMappingPeatDepth2023} linear regression models produced negative predictions, and it is unclear whether the values quoted above include these.
+The linear regression models in \citet{kogantiMappingPeatDepth2023} produced negative predictions, and it is unclear whether the values quoted above include these.
 If we disregard the negative predictions, their model showed the same pattern as ours in overpredicting shallow peats and underpredicting deep peats, although their underprediction was less severe.
-An important difference between Koganti et al.
-\citeyearpar{kogantiMappingPeatDepth2023} and our study (besides spatial scale) is that they measured radiometrics on the ground, rather than using airborne survey data.
+An important difference between \citet{kogantiMappingPeatDepth2023} and our study (besides spatial scale) is that they measured radiometrics on the ground, rather than using airborne survey data.
 Thus, the footprint of their detector was smaller and they captured variation in radiation at finer scales.
@@
 All four radiometric variables were highly correlated within the peatland parts of our study sites, so there could be no large differences in their predictive value.
-This contrasts with Koganti et al.
-\citeyearpar{kogantiMappingPeatDepth2023}, who found that radiometric total count and potassium ground concentration were better predictors than thorium or uranium ground concentration.
+This contrasts with \citet{kogantiMappingPeatDepth2023}, who found that radiometric total count and potassium ground concentration were better predictors than thorium or uranium ground concentration.
 
@@
 Studies comparing airborne and ground radiometric surveys confirm that they are poorly correlated in low-activity areas like peatlands \citep{kockComparisonAirborneTerrestrial2011, karjalainenComparisonTwoGammaray2025}.
-Karjalainen et al. \citeyearpar{karjalainenComparisonTwoGammaray2025} found that ground-based measurements predicted peat depth better than airborne measurements.
-Nevertheless, a large radiometric footprint did not prevent strong associations with peat depth in Gatis et al.
-\citeyearpar{gatisMappingUplandPeat2019} and Marchant \citeyearpar{marchantUsingRemoteSensors2021}, perhaps because the extensive blanket bog landscape in these studies has more gradual changes in depth \citep{lindsayBogsEcologyClassification1995}.
-We are unsure whether short-range depth changes explain the weak associations that Siemon et al. \citeyearpar{siemonAirborneElectromagneticRadiometric2020} found in a large raised bog.
+\citet{karjalainenComparisonTwoGammaray2025} found that ground-based measurements predicted peat depth better than airborne measurements.
+Nevertheless, a large radiometric footprint did not prevent strong associations with peat depth in \citet{gatisMappingUplandPeat2019} and \citet{marchantUsingRemoteSensors2021}, perhaps because the extensive blanket bog landscape in these studies has more gradual changes in depth \citep{lindsayBogsEcologyClassification1995}.
+We are unsure whether short-range depth changes explain the weak associations that \citet{siemonAirborneElectromagneticRadiometric2020} found in a large raised bog.
 
@@
 Soil moisture, bulk density, and porosity, however, are not easily measured across landscape scales and were assumed to be homogeneous.
-Uneven snow cover and air moisture during the Ørskogfjellet radiometric survey may also have masked the soil signal in these data, as Ofstad \citeyearpar{ofstadHelicopterborneMagneticRadiometric2015} reports large variation in weather conditions.
+Uneven snow cover and air moisture during the Ørskogfjellet radiometric survey may also have masked the soil signal in these data, as \citet{ofstadHelicopterborneMagneticRadiometric2015} reports large variation in weather conditions.
 If maps of these other physical parameters at the time of the radiometric survey were available and included in the model, the predictive value of radiometric data might improve, but this is not a practical solution for digital soil mapping of peat depth.
@@
 At both our sites, LiDAR-derived terrain variables predicted peat depth much better than radiometric variables (Fig. \ref{fig:modelMetrics}).
-This is consistent with the findings of Pohjankukka et al. \citeyearpar{pohjankukkaDigitalMappingPeat2025}, who mapped peat depth classes at 50 m resolution across Finland.
+This is consistent with the findings of \citet{pohjankukkaDigitalMappingPeat2025}, who mapped peat depth classes at 50 m resolution across Finland.
 
@@
 Better infrastructure to make depth measurements findable, accessible, interoperable, and reusable would help DSM and other applications.
-Geoportal access and data exchange standards \citeyearpar[like Natural England's for peat surveys,][]{naturalenglandDataExchangeStandard2023} are important.
+Geoportal access and data exchange standards, such as those developed in \citet{naturalenglandDataExchangeStandard2023} for peat surveys, are important.
 Peat depth is quick and easy to measure, so integrating its measurement into existing national field programs, like Norway's spatially representative nature monitoring or national forest inventory, would be helpful (although not sufficient for regional or local mapping).
@@
 However, the tradeoff between multiple small-extent DSM and fewer large-extent DSM needs research.
-Bohn and Miller \citeyearpar{bohnLocallyEnhancedDigital2024} advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or `supertopes'), where the composition of mesotopes is similar -- e.g., regions dominated by raised bogs versus regions dominated by sloping fens \citep{moenNationalAtlasNorway1999, joostenWiseUseMires2002}.
+\citet{bohnLocallyEnhancedDigital2024} advocate for bottom-up stitching of local DSM, and for peat depth we assert that these should at least stay within peatland regions (or `supertopes'), where the composition of mesotopes is similar -- e.g., regions dominated by raised bogs versus regions dominated by sloping fens \citep{moenNationalAtlasNorway1999, joostenWiseUseMires2002}.
 Depth varies systematically between bogs and fens \citep{lindsayPeatlandMireTypes2016}, between peats formed by terrestrialization versus paludification \citep{buffamFillingHolesRegional2010}, and probably along other axes of peatland typology.
@@
 The success of Multi-Resolution Valley Bottom Flatness as a predictor of peat depth demonstrates that multiscale spatial patterns matter for peat depth, and convolutional neural networks are designed to learn such patterns \citep{borowiecDeepLearningTool2022}.
-The kind of relationship described in Buffam et al.
-\citeyearpar{buffamFillingHolesRegional2010}, where peat depth in basins related to terrain slope at the basin edge, is also something a convolutional neural network could learn.
+The kind of relationship described in \citet{buffamFillingHolesRegional2010}, where peat depth in basins related to terrain slope at the basin edge, is also something a convolutional neural network could learn.
 However, since this approach is data-hungry, we should build soil knowledge into the DSM where we can \citep{minasnySoilScienceInformedMachine2024}.

### 2025-09-29 ms/ms.Rmd
proof-stage corrections

@@
 
-We used the *terra* R package (v1.8) to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
+We used the *terra* R package [v.1.8, @hijmansTerraSpatialData2025] to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
 These were derived at two scales to produce eight different predictors; we either calculated the indices at 1 m DTM resolution and then aggregated to 10 m resolution, or aggregated to 10 m DTM resolution and then calculated the indices.
@@
 The Topographic Wetness Index [@quinnPredictionHillslopeFlow1991] is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate to coarse resolution DTMs [@agrenEvaluatingDigitalTerrain2014; @riihimakiTopographicWetnessIndex2021], so we calculated it from 5 m, 10 m, 20 m, and 50 m DTM resolution.
-The calculations were performed with Whitebox software [@lindsayWhiteboxGATCase2016], accessed through the *whitebox* R package [v2.4, @wuWhiteboxWhiteboxToolsFrontend2022].
+The calculations were performed with Whitebox software [@lindsayWhiteboxGATCase2016], accessed through the *whitebox* R package [v.2.4, @wuWhiteboxWhiteboxToolsFrontend2022].
 We filled depressions in the DTM with the algorithm in @wangEfficientMethodIdentifying2006, and used the deterministic infinity flow accumulation algorithm [@tarbotonNewMethodDetermination1997].
@@
 
-We implemented models in the *tidymodels* framework in R [@kuhnTidymodelsCollectionPackages2020], with the *ranger* R package for RFs [v0.16, @wrightRangerFastImplementation2017].
+We implemented models in the *tidymodels* framework in R [@kuhnTidymodelsCollectionPackages2020], with the *ranger* R package for RFs [v.0.16, @wrightRangerFastImplementation2017].
 RFs were fit with 1000 trees, minimum node size of 5, and the number of predictors randomly sampled at each split was the square root of the total number of predictors (*ranger* default).
@@
 For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation [@wadouxSpatialCrossvalidationNot2021].
-We implemented the kNNDM with the *CAST* R package [v1.0, @meyerCASTPackageTraining2024], setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (k = 5--20) that produced the best match between the two NND distributions.
+We implemented the kNNDM with the *CAST* R package [v.1.0, @meyerCASTPackageTraining2024], setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (k = 5--20) that produced the best match between the two NND distributions.
 From the cross-validation we quantified mean absolute error (error magnitude, original scale), R^2^ (explained variation, standardized scale), and Lin's concordance correlation coefficient (error magnitude and explained variation, standardized scale).
@@
 
-We calculated variable importance with the *vip* R package (v0.4), by three different methods: *FIRM*, *permutation*, and *Shapley* [@greenwellVariableImportancePlots2020].
+We calculated variable importance with the *vip* R package (v.0.4), by three different methods: *FIRM*, *permutation*, and *Shapley* [@greenwellVariableImportancePlots2020].
 *FIRM* values measure the flatness of the partial dependence plot, *permutation* values measure the decrease in model performance when the predictor is permuted, and *Shapley* values are aggregated from local, game-theoretical measures of variable importance [@greenwellVariableImportancePlots2020].
@@
 
-We calculated partial dependence with the *pdp* R package [v0.8, @greenwellPdpPackageConstructing2017].
+We calculated partial dependence with the *pdp* R package [v.0.8, @greenwellPdpPackageConstructing2017].
 For the six most important variables, we plotted partial dependence to show the average effect of the predictor on the outcome and individual conditional expectations to show variation in the effect across observations [@goldsteinPeekingBlackBox2015].
@@
 
-At Skrimfjella, we used the *eSample* function in the *iSDM* R package (v1.0) to stratify our sample across elevation, slope, and potassium concentration.
+At Skrimfjella, we used the *eSample* function in the *iSDM* R package (v.1.0) to stratify our sample across elevation, slope, and potassium concentration.
 This function defines the environmental space as a two-dimensional convex hull around the PCA-ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest [@hattabUnifiedFrameworkModel2017].
@@
 
-To choose 160 locations, we performed feature space coverage sampling, implemented using the *kmeans* function in base R and the *rdist* function in the *fields* package (v14.2).
+To choose 160 locations, we performed feature space coverage sampling, implemented using the *kmeans* function in base R and the *rdist* function in the *fields* package [v.14.2, @nychkaFieldsToolsSpatial2025].
 Feature space coverage sampling chooses locations that are closest to cluster centers in standardized predictor space [@brusSamplingDigitalSoil2019].
@@
                 caption = "Pairwise comparisons of the concordance correlation coefficient among models for Skrimfjella, based on 10-fold cross-validation.
-The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package) on mixed-effects models (\\texttt{lme4} package) to account for the cross-validation fold structure.")
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package, v.1.11.1, \\citealp{lenthEmmeansEstimatedMarginal2025}) on mixed-effects models (\\texttt{lme4} package, v.1.1-36, \\citealp{batesFittingLinearMixedEffects2015}) to account for the cross-validation fold structure.")
 ```
@@
                 caption = "Pairwise comparisons of $R^2$ among models for Skrimfjella, based on 10-fold cross-validation.
-The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package) on mixed-effects models (\\texttt{lme4} package) to account for the cross-validation fold structure.")
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package, v.1.11.1, \\citealp{lenthEmmeansEstimatedMarginal2025}) on mixed-effects models (\\texttt{lme4} package, v.1.1-36, \\citealp{batesFittingLinearMixedEffects2015}) to account for the cross-validation fold structure.")
 ```
@@
                 caption = "Pairwise comparisons of mean absolute error among models for Skrimfjella, based on 10-fold cross-validation.
-The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package) on mixed-effects models (\\texttt{lme4} package) to account for the cross-validation fold structure.")
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package, v.1.11.1, \\citealp{lenthEmmeansEstimatedMarginal2025}) on mixed-effects models (\\texttt{lme4} package, v.1.1-36, \\citealp{batesFittingLinearMixedEffects2015}) to account for the cross-validation fold structure.")
 ```
@@
                 caption = "Pairwise comparisons of the concordance correlation coefficient among models for Ørskogfjellet, based on 10-fold cross-validation.
-The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package) on mixed-effects models (\\texttt{lme4} package) to account for the cross-validation fold structure.")
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package, v.1.11.1, \\citealp{lenthEmmeansEstimatedMarginal2025}) on mixed-effects models (\\texttt{lme4} package, v.1.1-36, \\citealp{batesFittingLinearMixedEffects2015}) to account for the cross-validation fold structure.")
 ```
@@
                 caption = "Pairwise comparisons of $R^2$ among models for Ørskogfjellet, based on 10-fold cross-validation.
-The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package) on mixed-effects models (\\texttt{lme4} package) to account for the cross-validation fold structure.")
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package, v.1.11.1, \\citealp{lenthEmmeansEstimatedMarginal2025}) on mixed-effects models (\\texttt{lme4} package, v.1.1-36, \\citealp{batesFittingLinearMixedEffects2015}) to account for the cross-validation fold structure.")
 ```
@@
                 caption = "Pairwise comparisons of mean absolute error among models for Ørskogfjellet, based on 10-fold cross-validation.
-The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package) on mixed-effects models (\\texttt{lme4} package) to account for the cross-validation fold structure.")
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\\texttt{emmeans} package, v.1.11.1, \\citealp{lenthEmmeansEstimatedMarginal2025}) on mixed-effects models (\\texttt{lme4} package, v.1.1-36, \\citealp{batesFittingLinearMixedEffects2015}) to account for the cross-validation fold structure.")
 ```

### 2025-09-29 ms/ms.tex
proof-stage corrections

@@
 
-We used the \emph{terra} R package (v1.8) to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
+We used the \emph{terra} R package \citep[v.1.8,][]{hijmansTerraSpatialData2025} to calculate from the DTMs: slope, Topographic Position Index (difference from mean of eight neighbors), Terrain Ruggedness Index (mean of absolute differences from eight neighbors), and roughness (range in the nine-cell neighborhood).
 These were derived at two scales to produce eight different predictors; we either calculated the indices at 1 m DTM resolution and then aggregated to 10 m resolution, or aggregated to 10 m DTM resolution and then calculated the indices.
@@
 The Topographic Wetness Index \citep{quinnPredictionHillslopeFlow1991} is notoriously scale-dependent and often matches real hydrological conditions best when calculated from moderate to coarse resolution DTMs \citep{agrenEvaluatingDigitalTerrain2014, riihimakiTopographicWetnessIndex2021}, so we calculated it from 5 m, 10 m, 20 m, and 50 m DTM resolution.
-The calculations were performed with Whitebox software \citep{lindsayWhiteboxGATCase2016}, accessed through the \emph{whitebox} R package \citep[v2.4,][]{wuWhiteboxWhiteboxToolsFrontend2022}.
+The calculations were performed with Whitebox software \citep{lindsayWhiteboxGATCase2016}, accessed through the \emph{whitebox} R package \citep[v.2.4,][]{wuWhiteboxWhiteboxToolsFrontend2022}.
 We filled depressions in the DTM with the algorithm in \citet{wangEfficientMethodIdentifying2006}, and used the deterministic infinity flow accumulation algorithm \citep{tarbotonNewMethodDetermination1997}.
@@
 
-We implemented models in the \emph{tidymodels} framework in R \citep{kuhnTidymodelsCollectionPackages2020}, with the \emph{ranger} R package for RFs \citep[v0.16,][]{wrightRangerFastImplementation2017}.
+We implemented models in the \emph{tidymodels} framework in R \citep{kuhnTidymodelsCollectionPackages2020}, with the \emph{ranger} R package for RFs \citep[v.0.16,][]{wrightRangerFastImplementation2017}.
 RFs were fit with 1000 trees, minimum node size of 5, and the number of predictors randomly sampled at each split was the square root of the total number of predictors (\emph{ranger} default).
@@
 For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation \citep{wadouxSpatialCrossvalidationNot2021}.
-We implemented the kNNDM with the \emph{CAST} R package \citep[v1.0,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (k = 5--20) that produced the best match between the two NND distributions.
+We implemented the kNNDM with the \emph{CAST} R package \citep[v.1.0,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (k = 5--20) that produced the best match between the two NND distributions.
 From the cross-validation we quantified mean absolute error (error magnitude, original scale), R\textsuperscript{2} (explained variation, standardized scale), and Lin's concordance correlation coefficient (error magnitude and explained variation, standardized scale).
@@
 
-We calculated variable importance with the \emph{vip} R package (v0.4), by three different methods: \emph{FIRM}, \emph{permutation}, and \emph{Shapley} \citep{greenwellVariableImportancePlots2020}.
+We calculated variable importance with the \emph{vip} R package (v.0.4), by three different methods: \emph{FIRM}, \emph{permutation}, and \emph{Shapley} \citep{greenwellVariableImportancePlots2020}.
 \emph{FIRM} values measure the flatness of the partial dependence plot, \emph{permutation} values measure the decrease in model performance when the predictor is permuted, and \emph{Shapley} values are aggregated from local, game-theoretical measures of variable importance \citep{greenwellVariableImportancePlots2020}.
@@
 
-We calculated partial dependence with the \emph{pdp} R package \citep[v0.8,][]{greenwellPdpPackageConstructing2017}.
+We calculated partial dependence with the \emph{pdp} R package \citep[v.0.8,][]{greenwellPdpPackageConstructing2017}.
 For the six most important variables, we plotted partial dependence to show the average effect of the predictor on the outcome and individual conditional expectations to show variation in the effect across observations \citep{goldsteinPeekingBlackBox2015}.
@@
 
-At Skrimfjella, we used the \emph{eSample} function in the \emph{iSDM} R package (v1.0) to stratify our sample across elevation, slope, and potassium concentration.
+At Skrimfjella, we used the \emph{eSample} function in the \emph{iSDM} R package (v.1.0) to stratify our sample across elevation, slope, and potassium concentration.
 This function defines the environmental space as a two-dimensional convex hull around the PCA-ordinated data, then creates a regular grid across that space, and lastly finds for each grid cell the datum that is nearest \citep{hattabUnifiedFrameworkModel2017}.
@@
 
-To choose 160 locations, we performed feature space coverage sampling, implemented using the \emph{kmeans} function in base R and the \emph{rdist} function in the \emph{fields} package (v14.2).
+To choose 160 locations, we performed feature space coverage sampling, implemented using the \emph{kmeans} function in base R and the \emph{rdist} function in the \emph{fields} package \citep[v.14.2,][]{nychkaFieldsToolsSpatial2025}.
 Feature space coverage sampling chooses locations that are closest to cluster centers in standardized predictor space \citep{brusSamplingDigitalSoil2019}.
@@
 \caption{\label{tab:pairwiseSkrimCCC}Pairwise comparisons of the concordance correlation coefficient among models for Skrimfjella, based on 10-fold cross-validation.
-The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package) on mixed-effects models (\texttt{lme4} package) to account for the cross-validation fold structure.}
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package, v.1.11.1, \citealp{lenthEmmeansEstimatedMarginal2025}) on mixed-effects models (\texttt{lme4} package, v.1.1-36, \citealp{batesFittingLinearMixedEffects2015}) to account for the cross-validation fold structure.}
 \centering
@@
 \caption{\label{tab:pairwiseSkrimRsq}Pairwise comparisons of $R^2$ among models for Skrimfjella, based on 10-fold cross-validation.
-The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package) on mixed-effects models (\texttt{lme4} package) to account for the cross-validation fold structure.}
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package, v.1.11.1, \citealp{lenthEmmeansEstimatedMarginal2025}) on mixed-effects models (\texttt{lme4} package, v.1.1-36, \citealp{batesFittingLinearMixedEffects2015}) to account for the cross-validation fold structure.}
 \centering
@@
 \caption{\label{tab:pairwiseSkrimMAE}Pairwise comparisons of mean absolute error among models for Skrimfjella, based on 10-fold cross-validation.
-The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package) on mixed-effects models (\texttt{lme4} package) to account for the cross-validation fold structure.}
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package, v.1.11.1, \citealp{lenthEmmeansEstimatedMarginal2025}) on mixed-effects models (\texttt{lme4} package, v.1.1-36, \citealp{batesFittingLinearMixedEffects2015}) to account for the cross-validation fold structure.}
 \centering
@@
 \caption{\label{tab:pairwiseOrskogCCC}Pairwise comparisons of the concordance correlation coefficient among models for Ørskogfjellet, based on 10-fold cross-validation.
-The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package) on mixed-effects models (\texttt{lme4} package) to account for the cross-validation fold structure.}
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package, v.1.11.1, \citealp{lenthEmmeansEstimatedMarginal2025}) on mixed-effects models (\texttt{lme4} package, v.1.1-36, \citealp{batesFittingLinearMixedEffects2015}) to account for the cross-validation fold structure.}
 \centering
@@
 \caption{\label{tab:pairwiseOrskogRsq}Pairwise comparisons of $R^2$ among models for Ørskogfjellet, based on 10-fold cross-validation.
-The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package) on mixed-effects models (\texttt{lme4} package) to account for the cross-validation fold structure.}
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package, v.1.11.1, \citealp{lenthEmmeansEstimatedMarginal2025}) on mixed-effects models (\texttt{lme4} package, v.1.1-36, \citealp{batesFittingLinearMixedEffects2015}) to account for the cross-validation fold structure.}
 \centering
@@
 \caption{\label{tab:pairwiseOrskogMAE}Pairwise comparisons of mean absolute error among models for Ørskogfjellet, based on 10-fold cross-validation.
-The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package) on mixed-effects models (\texttt{lme4} package) to account for the cross-validation fold structure.}
+The comparisons use estimated marginal means with Tukey HSD correction and Kenward-Roger degrees of freedom approximation (\texttt{emmeans} package, v.1.11.1, \citealp{lenthEmmeansEstimatedMarginal2025}) on mixed-effects models (\texttt{lme4} package, v.1.1-36, \citealp{batesFittingLinearMixedEffects2015}) to account for the cross-validation fold structure.}
 \centering
